### 2022-10-31 test.qmd
moved folder to 'thesis'

---
title: "test"
format: html
editor: visual
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2022-11-15 test.Rmd
edited test.qmd
added new joined csv

---
title: "test"
output: html_document
date: "2022-11-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Read in data

```{r}
# Read in soundscape data, round to nearest 20 min
BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv")
BB_08 <- BB_08_csv %>% 
  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
BB_08 <- BB_08 %>%
  mutate(dateTime = round_date(ymd_hms(BB_08$dateTime),"20 minutes"))
```

### Read in Tracks
```{r}
# Read in tracks, filter out lost buoys
tracks_csv <- read_csv("Data/AllTracks.csv")
tracks <- tracks_csv %>%
  dplyr::filter(station == "4"| station == "7"| station =="8" |station == "10"| 
                station == "12"|station == "13"| station == "14"| station == "16"|
                station == "17"|station == "18"| station == "19"| station == "20"| 
                station == "21"|station == "22"| station == "23")

# Filter by CCES_08 buoy
tracks_08 <- tracks %>% dplyr::filter(station == "8")

# Round CCES_08 to nearest 20 min
tracks_08 <- tracks_08 %>% 
  mutate(dateTime = ceiling_date(ymd_hms(tracks_08$dateTime), "20 minutes"))

```

### Read in OL
```{r}
# Read in OL data, round to nearest 20 min
OL_08_csv <- read_csv("Data/CCES_08_1Hz_1s_OL_2min.csv")
OL_08 <- OL_08_csv %>% 
  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
OL_08 <- OL_08 %>%
  mutate(dateTime = round_date(ymd_hms(OL_08$dateTime),"20 minutes"))
OL_08

```

Joining tables
```{r}
# Join tables
track_BB_08 <- left_join(tracks_08_rnd, BB_08, by = "dateTime")
track_BB_OL_08 <- left_join(track_BB_08, OL_08, by="dateTime")

write_csv(track_BB_OL_08, "Data/track_bb_OL_08.csv")
```

### 2023-02-07 coastAbstract.qmd
removed files, made map
removed large soundscape files, created animated map

---
title: "COAST Award Abstract"
format: docx
editor: visual
---

### Abstract

The ocean is becoming noisier with increased shipping and offshore human activity. Beaked whales (Ziphiidae) and sperm whales (Physeteroidea) are among the species at risk of harm from noise pollution. They rely on acoustic signaling for basic life functions like communication and foraging, but the extent of harm from anthropogenic activity is unclear as little is known about these deep diving whales and their distribution. Passive acoustic monitoring (PAM) is a non-invasive tool that can provide insight to these species where visual data collection methods fall short. This research will characterize the extent of spatiotemporal noise exposure in relation to acoustic beaked and sperm whale detections in the California Current Ecosystem. Drifting passive acoustic recorders were deployed in twenty-two offshore locations from southern Canada to northern Mexico in the summer of 2018 to record whale echolocation pulses, GPS point data, broadband levels, octave band levels, third octave levels, and power spectral density for a total of 1,910 recording hours. Soundscape metrics were analyzed in space and time in relation to whale detections. Preliminary results show that the presence of whales is not necessarily associated with a louder soundscape, and higher intensities of sound do not only occur near shipping routes. I will further investigate the variables that are associated with various soundscapes, such as environmental factors, species composition, and distance from shore. In exploring these data, I can better characterize the spatiotemporal relationship between noise pollution and beaked and sperm whale habitats using soundscape metrics.

### 2023-02-07 manuscript.qmd
removed files, made map
removed large soundscape files, created animated map

---
title: "manuscript"
format: html
editor: visual
---

# Manuscript Draft

## Introduction

## Methods

Acoustic detections were collected with the use of free-drifitng Drifting Acoustic Spar Buoy Recorders (DASBRs). These were deployed as part of the California Current Ecosystem Survey (CCES) spanning the offshore waters from southern Canada to northern Mexico in summer of 2018. DASBRs offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species. These DASBRs contained two satellite geo-location trackers, two hydrophones, etc and collected a total of 1910 hours of data. SWFSC ran acoustic detections in PAMGuard. Eight drifting recorders were lost at sea and therefore removed from the analysis. These contain 1, 5, 10, 25, 75, 95, and 99 percentile distribution from each metric

```{r}

```

## Results

## Discussion

```{r}
#| echo: false

```

The `echo: false` option disables the printing of code (only output is displayed).

### 2023-02-07 test.Rmd
removed files, made map
removed large soundscape files, created animated map

@@
 library(lubridate)
+library(sf)
+library(rerddap)
+library(terra)
 ```
 
-## Read in data
+## Read in CCES08 data
 
@@
 # Read in soundscape data, round to nearest 20 min
-BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv")
+BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
 BB_08 <- BB_08_csv %>% 
@@
   mutate(dateTime = round_date(ymd_hms(BB_08$dateTime),"20 minutes"))
+
 ```
@@
 ### Read in Tracks
+
 ```{r}
 # Read in tracks, filter out lost buoys
-tracks_csv <- read_csv("Data/AllTracks.csv")
+tracks_csv <- read_csv("Data/AllTracks.csv", show_col_types = FALSE)
 tracks <- tracks_csv %>%
@@
 # Round CCES_08 to nearest 20 min
-tracks_08 <- tracks_08 %>% 
-  mutate(dateTime = ceiling_date(ymd_hms(tracks_08$dateTime), "20 minutes"))
-
+tracks_08_rnd <- tracks_08 %>% 
+  mutate(dateTime = round_date(ymd_hms(tracks_08$dateTime), "20 minutes"))
+tracks_08_rnd
 ```
 
-### Read in OL
+### Read in OL, round to nearest 20 min
+
 ```{r}
-# Read in OL data, round to nearest 20 min
+# CCES_08
 OL_08_csv <- read_csv("Data/CCES_08_1Hz_1s_OL_2min.csv")
@@
 
+# CCES_10
+OL_10_csv <- read_csv("Data/CCES_10_1Hz_1s_OL_2min.csv")
+OL_10 <- OL_10_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+OL_10 <- OL_10 %>%
+  mutate(dateTime = round_date(ymd_hms(OL_10$dateTime),"20 minutes"))
+OL_10
 ```
 
-Joining tables
+### Read in CCES_10
+
 ```{r}
-# Join tables
+# Read in CCES_10 soundscape metrics
+BB_10_csv <- read_csv("Data/CCES_10_1Hz_1s_BB_2min.csv")
+BB_10 <- BB_10_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+BB_10 <- BB_10 %>%
+  mutate(dateTime = round_date(ymd_hms(BB_10$dateTime),"20 minutes"))
+
+# Filter by CCES_10 buoy
+tracks_10 <- tracks %>% dplyr::filter(station == "10")
+
+# Round CCES_10 to nearest 20 min
+tracks_10_rnd <- tracks_10 %>% 
+  mutate(dateTime = ceiling_date(ymd_hms(tracks_10$dateTime), "20 minutes"))
+
+```
+
+### Read in whale detections
+
+```{r}
+# Read in whale detections from CCES_08 and CCES_10
+whales <- read_csv("Data/EventInfo.csv")
+whales_sf <- whales %>%
+  dplyr::filter(Deployment == "8"| Deployment == "10") %>%
+  st_as_sf(coords = c("Longitude","Latitude"), crs=4326) %>%
+  mutate(bestNumber = as.numeric(bestNumber)) %>%
+  mutate(nClicks = as.numeric(nClicks))
+
+```
+
+## Read in AIS Data
+```{r}
+rawAIS<- read_csv("Data/AIS_072518_072918.csv")
+# need to figure out where to store data
+```
+
+### Joining tables with soundscape metrics
+
+```{r}
+# CCES_08
 track_BB_08 <- left_join(tracks_08_rnd, BB_08, by = "dateTime")
 track_BB_OL_08 <- left_join(track_BB_08, OL_08, by="dateTime")
+write_csv(track_BB_OL_08, "Data/track_bb_OL_08.csv")
 
-write_csv(track_BB_OL_08, "Data/track_bb_OL_08.csv")
+# CCES_10
+track_BB_10 <- left_join(tracks_10_rnd, BB_10, by = "dateTime")
+track_BB_OL_10 <- left_join(track_BB_10, OL_10, by="dateTime")
+write_csv(track_BB_OL_10, "Data/track_bb_OL_10.csv")
 ```
+
+### Map Animations
+```{r}
+# Install
+#devtools::install_github("ropenscilabs/rnaturalearth")
+#install.packages("rnaturalearth",
+                 #repos = "http://packages.ropensci.org",
+                 #type = "source")
+#install.packages("gganimate")
+#install.packages("gifski")
+#install.packages("marmap")
+
+# Libraries
+library(sf)
+library(tidyverse)
+library(marmap)
+library(gifski)
+library(gganimate)
+library(rnaturalearth)
+```
+
+### New ggplot wtih bplot as basemap 
+```{r}
+library(mapproj)
+# Get bathymetry (code from Ocean Tracking Network https://www.youtube.com/watch?v=xT2F-Y9bKLk)
+bgo <- getNOAA.bathy(lon1 = min(df_all$lon-0.7), lon2 = max(df_all$lon + 0.7),
+                     lat1 = min(df_all$lat-0.7), lat2=max(df_all$lat + 0.7), resolution = 1)
+
+# Create basemap
+bf <- fortify.bathy(bgo, geom=c("r", "c")) + scale_fill_etopo() # converts bathy object to data frame
+bplot <- autoplot(bgo, geom = c("raster", "contour"), coast = TRUE) + 
+  scale_fill_etopo() #creates ggplot object
+
+# Plot
+p <- bplot + 
+  geom_path(data = track_BB_OL_10, 
+            aes(x=long,y=lat,group=spotID), 
+            alpha = 0.5,
+            linewidth = 3) + # line width parameter is ignored for some reason
+  geom_point(data = track_BB_OL_10, 
+             aes(x=long,y=lat,group=spotID, size = OL_125, color = OL_125),
+             shape=19, stroke = 4) +
+  #formatting  
+  scale_color_viridis_c(option = "inferno") +
+  scale_size_continuous(range = c(0.1,10)) +
+  labs(x=NULL, y=NULL, 
+       fill = 'Depth',
+       color = 'OL 125')+
+  theme(panel.grid = element_blank())
+```
+
+```{r}
+# Animate
+anim = p +
+  transition_reveal(along = dateTime)+
+  ease_aes('linear')+
+  shadow_wake(0.1) +
+  ggtitle("Date: {frame_along}")
+
+# Animate with control over frames per second and number of frames
+animate(anim, nframes = 70, fps = 3)
+
+```
+
+Linear Regression Model
+
+```{r}
+# Does sound vary with latitude?
+lm(formula = OL_63 ~ lat, data = track_BB_OL_08)
+fit10 <- lm(formula = lat ~ OL_63, data = track_BB_OL_10)
+plot(fit10)
+ggplot(track_BB_OL_08, aes(x = lat, y = `BB_20-24000`)) + scale_y_continuous() +
+   geom_point() + geom_smooth()
+
+# Does sound vary with date/time?
+summary(lm(formula = `BB_20-24000`~ dateTime, data = track_BB_OL_10))
+
+ggplot(track_BB_OL_08, aes(x = dateTime, y = `BB_20-24000`)) +
+  geom_point() +
+  geom_smooth() +
+  ggtitle("Buoy 08 - Sound levels by date")
+
+ggplot(track_BB_OL_10, aes(x = dateTime, y = `BB_20-24000`)) +
+  geom_point() +
+  geom_smooth() +
+  ggtitle("Buoy 10 - Sound levels by date")
+
+plot(lm(formula = `BB_20-24000`~ dateTime, data = track_BB_OL_08))
+
+ggplot() +
+  geom_point(data = track_BB_OL_08, aes(x = dateTime, y = `BB_20-24000`, color = "red")) +
+  geom_point(data = track_BB_OL_10, aes(x = dateTime, y = `BB_20-24000`, color = "blue"))
+
+```
+
+CCES Broadband and OL Map
+
+```{r}
+library(tmap); library(maptiles); library(sf)
+tmap_mode("plot")
+trackssf <- st_as_sf(tracks, coords = c("long", "lat"), crs=4326)
+tracks08sf <- st_as_sf(track_BB_OL_08, coords = c("long", "lat"), crs=4326)
+tracks10sf <- st_as_sf(track_BB_OL_10, coords = c("long", "lat"), crs=4326)
+trackBase <- get_tiles(whales_sf, provider = "Esri.OceanBasemap")
+allTrackbase <- get_tiles(trackssf, provider = "Esri.OceanBasemap")
+
+#Map all tracks
+tm_shape(allTrackbase) + tm_rgb() +
+  tm_shape(trackssf) + 
+  tm_dots(col = "black")
+
+# Map tracks 08 and 10 only
+tm_shape(trackBase) + tm_rgb() + 
+  tm_shape(tracks08sf) + tm_dots(col = "red", 
+                                 legend.show = TRUE) +
+  tm_shape(tracks10sf) + tm_dots(col = "green")
+
+# CCES_08 + CCES_10 Broadband
+tm_shape(trackBase) + tm_rgb() + 
+  tm_shape(tracks08sf) + 
+  tm_symbols(col = "BB_20-24000", 
+             border.lwd = NA, 
+             size = 0.4, 
+             sizes.legend = 0.7, 
+             n = 10) +
+  tm_shape(tracks10sf) + 
+  tm_symbols(col = "BB_20-24000",
+             border.lwd = NA, 
+             size = 0.4,
+             n = 10) +
+  tm_layout(legend.width = 0.5, 
+            legend.position = c("right","top"),
+            legend.bg.color = "lightblue")
+
+#CCES_08 Octave Level
+tm_shape(trackBase) + tm_rgb() + 
+  tm_shape(tracks08sf) + 
+  tm_symbols(col = "OL_63", 
+             border.lwd = NA, 
+             size = 0.5, 
+             n = 10) + 
+  tm_layout(legend.width = 0.5, 
+            legend.position = c("right","top")) + 
+  tm_legend()
+
+```
+
+### Sea Surface Temp with rerddap
+
+```{r}
+# Looking at one day, Aug 23 2018
+
+library(rerddap)
+SSTInfo <- info('jplMURSST41')
+murSST <- griddap(sstInfo, latitude = c(22., 51.), longitude = c(-140., -105), time = c("2018-08-23", "2018-08-23"), fields = 'analysed_sst')
+ggplot(data = murSST$data, aes(x = longitude, y = latitude, fill = analysed_sst))
+
+# needs fixing!
+
+# Trying SST here again...
+library(terra)
+SST <- terra::rast("Data/SST.tif")
+track_rast <- rasterize(tracks08sf, SST)
+
+plot(SST)
+plot(track_rast)
+```
+
+### Bathymetry Analysis
+
+```{r}
+# Increase cell size by 30x to be visible
+bathy <- rast("Data/gebco_bathy.tif")
+agg <- terra::aggregate(bathy, 30)
+plot(agg)
+
+# Rasterize track with agg as the template
+track_rast <- rasterize(tracks08sf, agg, field = "BB_20-24000")
+plot(track_rast)
+
+bathy_df <- as.data.frame(bathy, xy = TRUE)
+
+# Map the coastline
+cont <- st_as_sf(as.contour(bathy))
+coast <- cont %>% filter(level == 0)
+cont_slope <- cont %>% filter(level == -2000)
+cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
+plot(cont_slope_rast, col = "red")
+cont_slope_rast <- rasterize(cont_slope, agg)
+plot(cont_slope_rast, col = "red")
+
+ggplot() + 
+  geom_raster(data = bathy_df, 
+              mapping = aes(x=x, y=y, 
+                            alpha = 0.8,
+                            fill = `gebco_bathy`))+
+  geom_sf(data = cont, col="black") + 
+  geom_sf(data = coast, col = "red") +
+  geom_sf(data = cont_slope, col = "purple") +
+  geom_sf(data = tracks08sf, mapping = aes(col= `BB_20-24000`)) +
+  geom_sf(data = tracks10sf, mapping = aes(col= `BB_20-24000`)) +
+  scale_color_gradient(low="green", high="red")
+
+# Distance to continental slope
+
+BB_dist <- terra::distance(track_rast, vect(cont_slope))
+dist_ex08 <- terra::extract(BB_dist, vect(tracks08sf))
+pts08 <- cbind(tracks08sf,dist_ex08) %>%
+  dplyr::select(-ID) %>%
+  rename("Distance to Continental Slope" = gebco_bathy)
+
+dist_ex10 <- terra::extract(BB_dist, vect(tracks10sf))
+pts10 <- cbind(tracks10sf,dist_ex10) %>%
+  dplyr::select(-ID) %>%
+  rename("Distance to Continental Slope" = gebco_bathy)
+
+# Map distance to continental slope by color
+ggplot() +
+  geom_sf(data = pts08, 
+          mapping = aes(col = `Distance to Continental Slope`)) +
+  geom_sf(data = pts10, 
+          mapping = aes(col = `Distance to Continental Slope`)) +
+  geom_sf(data = cont_slope)
+
+# Plot distance by BB 
+ggplot(data = pts08, aes(x = `Distance to Continental Slope`, y = `BB_20.24000`)) + geom_boxplot()
+
+summary(lm(formula = BB_20.24000 ~ `Distance to Continental Slope`, data = pts08))
+
+```
+
+Does hour of day affect sound levels?
+
+```{r}
+# CCES_08
+hr_08 <- track_BB_OL_08 %>%
+  mutate(Hr = hour(track_BB_OL_08$dateTime)) %>%
+  group_by(Hr) %>%
+  summarize(meanBB = mean(`BB_20-24000`)) %>%
+  rename(`Buoy 08 Broadband` = `meanBB`)
+ggplot(data = hr_08, mapping = aes(x = Hr, y = `Buoy 08 Broadband`)) + 
+  geom_point() + geom_smooth()
+
+#CCES_10
+hr_10 <- track_BB_OL_10 %>%
+  mutate(Hr = hour(track_BB_OL_10$dateTime)) %>%
+  group_by(Hr) %>%
+  summarize(meanBB = mean(`BB_20-24000`)) %>%
+  rename(`Buoy 10 Broadband` = `meanBB`)
+ggplot(data = hr_10, mapping = aes(x = Hr, y = `Buoy 10 Broadband`)) + 
+  geom_point() + geom_smooth()
+
+# Join and graph
+hr_join <- left_join(hr_08, hr_10, by = "Hr")
+hr_piv <- hr_join %>%
+  pivot_longer(col=c(`Buoy 08 Broadband`,`Buoy 10 Broadband`), names_to = "buoys", values_to = "BB")
+
+ggplot(data = hr_piv, mapping = aes(Hr, BB, col=buoys)) +
+  geom_point() + geom_smooth()
+```
+
+### Are sound levels affected by species composition or number of clicks?
+
+```{r}
+# Plot
+library(tmap); library(maptiles)
+tmap_mode("plot")
+oceanBase <- get_tiles(whales_sf, provider="Esri.OceanBasemap", crop = TRUE)
+tm_shape(oceanBase) + tm_rgb() +
+  tm_shape(whales_sf) + 
+  tm_symbols(col="species", size = "nClicks") +
+  tm_layout(legend.width = 0.5, 
+            legend.position = c("right","top"))
+
+# CCES 08
+BB_whales08 <- st_join(pts08, whales_sf) %>%
+  filter(!is.na(`species`))
+
+summary(glm(`BB_20.24000` ~ species + `Distance to Continental Slope` + nClicks, data = BB_whales08))
+
+# CCES 10
+BB_whales10 <- st_join(pts10, whales_sf) %>%
+  filter(!is.na(`species`))
+
+summary(glm(`BB_20.24000` ~ species + `Distance to Continental Slope` + nClicks, data = BB_whales10))
+
+#Raster base map
+tmap_mode("plot")
+oceanBase <- get_tiles(BB_whales10, provider="Esri.OceanBasemap")
+tm_shape(bathy) + tm_raster() +
+  tm_shape(BB_whales10) + 
+  tm_symbols(col="species", size = "nClicks", style = "jenks") +
+  tm_layout(legend.width = 0.5)
+
+```

### 2023-02-08 test.Rmd
added AIS plot
still needs editing

@@
 ```{r}
-rawAIS<- read_csv("Data/AIS_072518_072918.csv")
-# need to figure out where to store data
+options(readr.show_col_types = FALSE) # quiet the message
+rawAIS<- read_csv("Data/AIS_100118_101018.csv") # Vessel data from 10-01-18 to 10-10-18
+
+# Filter by vessels over 100m in length, remove unnecessary columns
+AIS <- rawAIS %>%
+  filter(Length >= 100) %>%
+  dplyr::select(-SOG,-COG, -CallSign, -IMO, -Cargo, -Draft, -Status)
+
+bbox <- st_bbox(c(xmin = min(drift$lat-0.7), 
+            xmax = max(drift$lat + 0.7), 
+            ymax = max(drift$long + 0.7), 
+            ymin = min(drift$long-0.7)), crs = st_crs(4326))
+
+AIS_sf <- AIS %>%
+  st_as_sf(coords = c("LON", "LAT"), crs=4326, remove=F) #%>%
+  #st_crop(y = bbox)
+
+# Plot + crop to drift area
+ggplot() + 
+  geom_point(data = AIS_sf, mapping = aes(x=LON,y=LAT)) + 
+  scale_x_continuous(limits = c(min(drift$long-0.7), max(drift$long + 0.7))) + 
+  scale_y_continuous(limits = c(min(drift$lat-0.7), max(drift$lat + 0.7)))
 ```
@@
 library(mapproj)
+drift <- track_BB_OL_10
+
 # Get bathymetry (code from Ocean Tracking Network https://www.youtube.com/watch?v=xT2F-Y9bKLk)
-bgo <- getNOAA.bathy(lon1 = min(df_all$lon-0.7), lon2 = max(df_all$lon + 0.7),
-                     lat1 = min(df_all$lat-0.7), lat2=max(df_all$lat + 0.7), resolution = 1)
+bgo <- getNOAA.bathy(lon1 = min(drift$long-0.7), lon2 = max(drift$long + 0.7),
+                     lat1 = min(drift$lat-0.7), lat2=max(drift$lat + 0.7), resolution = 1)
 
@@
 
-# Plot
+# Plot drift
 p <- bplot + 
-  geom_path(data = track_BB_OL_10, 
+  geom_path(data = drift, 
             aes(x=long,y=lat,group=spotID), 
-            alpha = 0.5,
-            linewidth = 3) + # line width parameter is ignored for some reason
-  geom_point(data = track_BB_OL_10, 
+            alpha = 0.5) +
+  geom_point(data = drift, 
              aes(x=long,y=lat,group=spotID, size = OL_125, color = OL_125),
-             shape=19, stroke = 4) +
+             shape=19, stroke = 1) +
+
   #formatting  
@@
   theme(panel.grid = element_blank())
+
+# Plot AIS
+q <- bplot +
+  #geom_path(data = AIS_sf, aes(x=LON,y=LAT, group = "VesselName")) + 
+  geom_sf(data = AIS_sf, aes(x=LON, y=LAT, color = `VesselName`, size = 0.1), show.legend = FALSE) +
+  scale_x_continuous(limits = c(min(drift$long-0.7), max(drift$long + 0.7))) + 
+  scale_y_continuous(limits = c(min(drift$lat-0.7), max(drift$lat + 0.7))) +
+  labs(x=NULL, y=NULL, 
+       fill = 'Depth') +
+  theme(panel.grid = element_blank())
 ```
@@
 ```{r}
-# Animate
+# Animate drift
 anim = p +
@@
 animate(anim, nframes = 70, fps = 3)
+anim_save("drift10.gif")
+
+# Animate AIS
+anim = q +
+  transition_reveal(along = BaseDateTime)+
+  ease_aes('linear') +
+  ggtitle("Date: {frame_along}")
+
+# Animate with control over frames per second and number of frames
+animate(anim, nframes = 400, fps = 3)
 

### 2023-02-15 Geog_thesis_proposal.qmd
added slides

---
format: revealjs
title: "Assessing noise exposure to beaked and sperm whales in the California Current"
subtitle: "MS GIScience Thesis Proposal"
author: "Marina Bozinovic"
date: April 5, 2023
title-slide-attributes:
  data-background-image: images/IMG_1531.JPG
  data-background-opacity: "0.5"
---

## Introduction and Background

::: incremental
1.  Introduction to the nature of the study

    -   Soundscapes/acoustic habitats + human-generate noise

    -   Threatened and little-known whale species. Visually cryptic, long dive times make study difficult.

    -   Using drifting sensors (buoys) to track this relationship in space and time

    -   PAM
:::

::: notes
Speaker notes go here
:::

## Introduction and Background

2.  Theoretical framework and literature review

    -   Framework: Exploratory framework

    -   Lit review:

        -   More noise/increased shipping is expected.

        -   Beaked/sperm whale detections are difficult to collect, presence-only

        -   Soundscape research is on the rise, acoustic habitat is an parameter to consider in conservation/protected habitat management applications

## Introduction and Background

3.  Specific research objectives.

    -   Explore 2018 dataset from CCES survey with soundscape metrics and beaked whale detections.

        -   spatial proximity of vessels with high intensity sound events

    -   Create an open-source, reproducible tool for visualizing noise exposure (broad). This tool can lay groundwork for future hypotheses, better defining conservation priorities (outside the scope of my study) and visualizing events requiring further protection.

## Introduction and Background

4.  Rationale

    -   Researchers and managers need a tool to continue to investigate the relationship between human activity and threatened animals as the ocean becomes noisier. Patterns of noise within Sanctuaries is important. Effective for planning "noise-generating activities" (Hooker et al. 2019) around sensitive species.

## Method and Materials

1.  Description of study area - California Current Ecosystem (S. Oregon to N. Mexico)

    ![](images/allTracks.png)

    -   productive and well-managed/studied ecosystem

## Method and Materials

2.  Detailed description of methods to resolve research objectives

    -   Use RStudio:

        -   GOverall, generate map animations
        -   Packages: *gganimate,* etc.
        -   LOTS of data from CCES_2018, AIS to wrangle

    -   Use Github:

        -   reproducible code

## Projected Results

1.  What kind of info will this design acquire?
2.  How will you use this info to reach objectives?
3.  How will proposed research contribute to knowledge of field?

### 2023-03-08 readin.R
Create readin.R
Created R script to load, read in, wrangle raw data

@@ -1,0 +1,178 @@
+
+# Read in, load, and wrangle raw data
+
+###########################
+# Required packages
+###########################
+
+library(tidyverse)
+library(lubridate)
+library(sf)
+library(rerddap)
+library(terra)
+library(marmap)
+library(rnaturalearth)
+library(mapproj)
+library(raster)
+library("rerddapXtracto")
+library(PAMpal)
+library(maptiles)
+library(tmap)
+library(spData)
+library(tmaptools)
+library(grid)
+library(ggspatial)
+
+
+##########################
+# Read in soundscape data, NEED TO MAKE THIS A FUNCTION
+##########################
+
+# Round to nearest 20 minutes, create dateTime data type
+# Drift 8: Broadband
+BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
+BB_08 <- BB_08_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+BB_08 <- BB_08 %>%
+  mutate(dateTime = round_date(ymd_hms(BB_08$dateTime),"20 minutes"))
+
+# Drift 8: Octave Level
+OL_08_csv <- read_csv("Data/CCES_08_1Hz_1s_OL_2min.csv",show_col_types = FALSE)
+OL_08 <- OL_08_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+OL_08 <- OL_08 %>%
+  mutate(dateTime = round_date(ymd_hms(OL_08$dateTime),"20 minutes"))
+
+# Drift 8: TOL
+TOL_08_csv<- read_csv("data/CCES_08_1Hz_1s_TOL_2min.csv", show_col_types = FALSE)
+TOL_08 <- TOL_08_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+TOL_08 <- TOL_08 %>%
+  mutate(dateTime = round_date(ymd_hms(TOL_08$dateTime),"20 minutes"))
+
+# Drift 10: Broadband
+BB_10_csv <- read_csv("Data/CCES_10_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
+BB_10 <- BB_10_csv %>% 
[…]

### 2023-03-08 test.Rmd
Update test.Rmd
Removed reading in chunks. This will be code only

--- 2022-11-15 test.Rmd
@@
 library(lubridate)
+library(sf)
+library(rerddap)
+library(terra)
+library(marmap)
+library(rnaturalearth)
+library(mapproj)
+library(raster)
+library("rerddapXtracto")
+library(PAMpal)
+library(maptiles)
+library(tmap)
+library(spData)
+library(tmaptools)
+library(grid)
+library(ggspatial)
 ```
 
-## Read in data
+## Read in soundscape data, NEED TO MAKE THIS A FUNCTION
 
 ```{r}
-# Read in soundscape data, round to nearest 20 min
-BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv")
+# Round to nearest 20 minutes, create dateTime data type
+# Drift 8: Broadband
+BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
 BB_08 <- BB_08_csv %>% 
@@
   mutate(dateTime = round_date(ymd_hms(BB_08$dateTime),"20 minutes"))
-```
 
-### Read in Tracks
-```{r}
-# Read in tracks, filter out lost buoys
-tracks_csv <- read_csv("Data/AllTracks.csv")
-tracks <- tracks_csv %>%
-  dplyr::filter(station == "4"| station == "7"| station =="8" |station == "10"| 
-                station == "12"|station == "13"| station == "14"| station == "16"|
-                station == "17"|station == "18"| station == "19"| station == "20"| 
-                station == "21"|station == "22"| station == "23")
-
-# Filter by CCES_08 buoy
-tracks_08 <- tracks %>% dplyr::filter(station == "8")
-
-# Round CCES_08 to nearest 20 min
-tracks_08 <- tracks_08 %>% 
-  mutate(dateTime = ceiling_date(ymd_hms(tracks_08$dateTime), "20 minutes"))
-
-```
-
-### Read in OL
-```{r}
-# Read in OL data, round to nearest 20 min
-OL_08_csv <- read_csv("Data/CCES_08_1Hz_1s_OL_2min.csv")
+# Drift 8: Octave Level
+OL_08_csv <- read_csv("Data/CCES_08_1Hz_1s_OL_2min.csv",show_col_types = FALSE)
 OL_08 <- OL_08_csv %>% 
@@
   mutate(dateTime = round_date(ymd_hms(OL_08$dateTime),"20 minutes"))
-OL_08
+
+# Drift 8: TOL
+TOL_08_csv<- read_csv("data/CCES_08_1Hz_1s_TOL_2min.csv", show_col_types = FALSE)
+TOL_08 <- TOL_08_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+TOL_08 <- TOL_08 %>%
+  mutate(dateTime = round_date(ymd_hms(TOL_08$dateTime),"20 minutes"))
+
+# Drift 10: Broadband
+BB_10_csv <- read_csv("Data/CCES_10_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
+BB_10 <- BB_10_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+BB_10 <- BB_10 %>%
+  mutate(dateTime = round_date(ymd_hms(BB_10$dateTime),"20 minutes"))
+
+# Drift 10: TOL
+TOL_10_csv<- read_csv("data/CCES_10_1Hz_1s_TOL_2min.csv", show_col_types = FALSE)
+TOL_10 <- TOL_10_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+TOL_10 <- TOL_10 %>%
+  mutate(dateTime = round_date(ymd_hms(TOL_10$dateTime),"20 minutes"))
+
+# Drift 10: Octave level
+OL_10_csv <- read_csv("Data/CCES_10_1Hz_1s_OL_2min.csv", show_col_types = FALSE)
+OL_10 <- OL_10_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+OL_10 <- OL_10 %>%
+  mutate(dateTime = round_date(ymd_hms(OL_10$dateTime),"20 minutes"))
 
@@
 
-Joining tables
+### Read in Tracks
+
 ```{r}
-# Join tables
-track_BB_08 <- left_join(tracks_08_rnd, BB_08, by = "dateTime")
-track_BB_OL_08 <- left_join(track_BB_08, OL_08, by="dateTime")
+# Read in tracks, filter out lost buoys
+tracks <- read_csv("Data/AllTracks.csv", show_col_types = FALSE) %>%
+  dplyr::filter(station == "4"| station == "7"| station =="8" |station == "10"|          # filter out lost buoys
+                station == "12"|station == "13"| station == "14"| station == "16"|
+                station == "17"|station == "18"| station == "19"| station == "20"| 
+                station == "21"|station == "22"| station == "23") %>%
+  mutate(dateTime = round_date(ymd_hms(tracks$dateTime), "20 minutes"))                 
+  # round to nearest 20 minutes
 
-write_csv(track_BB_OL_08, "Data/track_bb_OL_08.csv")
+# Filter by buoy 8 and 10
+tracks_08 <- tracks %>% dplyr::filter(station == "8")
+tracks_10 <- tracks %>% dplyr::filter(station == "10")
+
+# Tracks on a map
+tracks_sf <- tracks %>% # make tracks an sf object
+  st_as_sf(coords= c("long", "lat"), crs=4326, remove=F)
+
 ```
+
+### Read in whale detections
+
+```{r}
+# All whale detections
+whales <- read_csv("Data/EventInfo.csv", show_col_types = FALSE) %>%
+  mutate(lat = as.numeric(Latitude)) %>%
+  mutate(long = as.numeric(Longitude)) %>%
+  mutate(bestNumber = as.numeric(bestNumber)) %>%   #convert nClicks from character to numeric
+  mutate(nClicks = as.numeric(nClicks)) %>%      # convert nClicks from character to numeric
+  mutate(StartTime = mdy_hm(StartTime)) %>%      #convert time from character to datetime
+  mutate(EndTime = mdy_hm(EndTime)) %>%
+  drop_na(...1)%>%                                 # remove NAs in column labeled "...1"
+  dplyr::select(-Project)
+
+BWsf <- whales %>% st_as_sf(coords = c("long","lat"), crs=4326)   # whales as sf
+
+# Drift 10 whales as sf
+whales10_sf <- BWsf %>%                            
+  dplyr::filter(Deployment == "10")
+
+whales08_sf <- BWsf %>%                            
+  dplyr::filter(Deployment == "8")
+
+# Drift 10 whales
+whales10 <- whales %>%
+  dplyr::filter(Deployment == "10")
+
+# Drift 8 whales
+whales8 <- whales %>%
+  dplyr::filter(Deployment == "8")
+```
+
+### Joining tracks with soundscape metrics with whale detections
+
+```{r}
+# Drift 08
+tracks_SS_08 <- left_join(tracks_08, BB_08, by = "dateTime") %>%   # join tracks with BB 08
+  left_join(., TOL_08, by = "dateTime") %>%                        # join TOL
+  left_join(., whales8, by = "long") %>%                           # join whales 
+  filter(spotID == "0-2571609") %>%                                # filter by one SPOT
+  mutate(BWpresence = if_else(is.na(nClicks), 0, 1)) %>%           # Add column to specify BW presence (1) or absence (0)
+  dplyr::select(dateTime, lat.x, long, station, `BB_20-24000`, TOL_63, 
+                TOL_125, TOL_2000, TOL_5000, TOL_20000, species, BWpresence)  # keep necessary columns
+
+#Drift 10
+tracks_SS_10 <- left_join(tracks_10, BB_10, by = "dateTime") %>%    # join tracks with BB 10
+  left_join(., TOL_10, by = "dateTime") %>%                         # join tracks with TOL 10 
+  left_join(., whales10, by = "long") %>%                           # join whales 
+  filter(spotID == "0-2572490") %>%                                 # filter by one SPOT
+  mutate(BWpresence = if_else(is.na(nClicks), 0, 1)) %>%           # Add column to specify BW presence (1) or absence (0)
+  dplyr::select(dateTime, lat.x, long, station, `BB_20-24000`, TOL_63, 
+                TOL_125, TOL_2000, TOL_5000, TOL_20000, species, BWpresence)  # keep necessary columns
+```
+
+## AIS
+
+For 10-01-18 to 10-10-18
+
+```{r}
+rawAIS <- read_csv("Data/AIS_100118_101018.csv", show_col_types = FALSE) # Vessel data from 10-01-18 to 10-10-18
+drift <- track_BB_OL_10
+
+# Filter by vessels over 100m in length, remove unnecessary columns
+AIS <- rawAIS %>%
+  filter(Length >= 100) %>%
+  dplyr::select(-SOG,-COG, -CallSign, -IMO, -Cargo, -Draft, -Status)
+
+# Currently unused
+bbox <- st_bbox(c(xmin = min(drift$lat-0.7), 
+            xmax = max(drift$lat + 0.7), 
+            ymax = max(drift$long + 0.7), 
+            ymin = min(drift$long-0.7)), crs = st_crs(4326))
+
+AIS_sf <- AIS %>%
+  st_as_sf(coords = c("LON", "LAT"), crs=4326, remove=F) #%>%
+  #st_crop(y = bbox)
+
+AIS_sf <- AIS_sf %>% 
+  filter(BaseDateTime > "2018-10-05 00:00:00")  # Shorten time range to 10/05/2018 to 10/10/2018
+```
+
+#### AIS
+
+For 08-10-18 to 08-17-18
+
+```{r}
+rawAIS <- read_csv("Data/AIS_081018_081718.csv", show_col_types = FALSE) # Vessel data from 08-10-18 to 08-17-18
+
+augAIS <- rawAIS %>% filter(Length >= 100) %>%
+  dplyr::select(-SOG,-COG, -CallSign, -IMO, -Cargo, -Draft, -Status)
+
+```
+
+### Exploratory analysis
+
+#### Not important: sound by latitude and dateTime
+
+```{r}
+# Does sound vary with latitude?
+lm(formula = OL_63 ~ lat, data = track_BB_OL_08)
+fit10 <- lm(formula = lat ~ OL_63, data = track_BB_OL_10)
+plot(fit10)
+ggplot(track_BB_OL_08, aes(x = lat, y = `BB_20-24000`)) + scale_y_continuous() +
+   geom_point() + geom_smooth()
+
+# Does sound vary with date/time?
+summary(lm(formula = `BB_20-24000`~ dateTime, data = track_BB_OL_10))
+
+ggplot(track_BB_OL_08, aes(x = dateTime, y = `BB_20-24000`)) +
+  geom_point() +
+  geom_smooth() +
+  ggtitle("Buoy 08 - Sound levels by date")
+
+ggplot(track_BB_OL_10, aes(x = dateTime, y = `BB_20-24000`)) +
+  geom_point() +
+  geom_smooth() +
+  ggtitle("Buoy 10 - Sound levels by date")
+
+plot(lm(formula = `BB_20-24000`~ dateTime, data = track_BB_OL_08))
+
+ggplot() +
+  geom_point(data = track_BB_OL_08, aes(x = dateTime, y = `BB_20-24000`, color = "red")) +
+  geom_point(data = track_BB_OL_10, aes(x = dateTime, y = `BB_20-24000`, color = "blue"))
+```
+
+#### Not important: tracks as sf and basemaps
+
+```{r}
+library(tmap); library(maptiles); library(sf)
+tmap_mode("plot")
+trackssf <- st_as_sf(tracks, coords = c("long", "lat"), crs=4326)
+tracks08sf <- st_as_sf(track_BB_OL_08, coords = c("long", "lat"), crs=4326)
+tracks10sf <- st_as_sf(track_BB_OL_10, coords = c("long", "lat"), crs=4326)
+trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("long", "lat.x"), crs=4326)
+trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("long", "lat.x"), crs=4326)
+trackBase <- get_tiles(whales_sf, provider = "Esri.WorldShadedRelief")
+allTrackbase <- get_tiles(trackssf, provider = "Stamen.TonerLite", zoom = 5)
+```
+
+### Sea Surface Temp with rerddap, Drift 8 only
+```{r}
+SSTInfo <- info('jplMURSST41')  # Multi-scale Ultra-high Resolution (MUR) SST Analysis fv04.1, Global, 0.01°, 2002-present, Daily 
+murSST <- griddap(sstInfo, latitude = c(35, 39), longitude = c(-122, -130), time = c("2018-08-16", "2018-10-01"), fields = 'analysed_sst') # drift8 dates only
+SST <- murSST$data
+
+# Raster of SST
+ggplot() + 
+  geom_tile(data = SST, aes(x = longitude, y = latitude, fill = analysed_sst)) +
+  scale_fill_gradient(low = "darkblue", high = "yellow") +
+  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(), 
+        axis.title = element_blank(),
+        panel.background = element_blank(),
+        panel.border = element_rect(fill = NA),
+        legend.position = "none")
+```
+
+### SST from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
+```{r}
+xcoord <- tracks_08$long
+ycoord <- tracks_08$lat
+tcoord <- tracks_08$dateTime
+xlen <- 0.2
+ylen <- 0.2
+parameter <- "analysed_sst"
+dataset <- 'jplMURSST41'
+#dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")
+
+#swchl <- rxtracto(dataInfo, 
+#                  parameter=parameter, 
+#                  xcoord=xcoord, ycoord=ycoord, 
+#                  tcoord=tcoord, xlen=xlen, ylen=ylen)
+
+#plotTrack(swchl, xcoord, ycoord, tcoord, plotColor = 'thermal')
+```
+
+### SST from PAMPal
+```{r}
+# Downloads data from servers
+# tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "analysedSST")
+
+# from file 
+tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'analysedSST.nc')
+tracks_SS_08
+```
+
+### Distance to continental slope
+
+```{r}
+# Increase cell size by 30x to be visible
+bathy <- rast("data/gebco_bathy.tif")
+agg <- terra::aggregate(bathy, 20)
+plot(agg)
+
+# Rasterize track with agg as the template
+#track_rast <- rasterize(tracks08sf, agg, field = "BB_20-24000")
+#TOL_rast <- rasterize(trk_TOL_08, agg, field = "TOL_125")
+sf08 <- st_as_sf(tracks_SS_08, coords = c("long", "lat.x"), crs=4326, remove=F)
+sf10 <- st_as_sf(tracks_SS_10, coords = c("long", "lat.x"), crs=4326, remove=F)
+trkraster08 <- rasterize(vect(sf08), agg)
+trkraster10 <- rasterize(vect(sf10), agg)
+
+bathy_df <- as.data.frame(bathy, xy = TRUE)
+
+# Map the coastline and cont. slope
+cont <- st_as_sf(as.contour(bathy))
+coast <- cont %>% filter(level == 0)
+cont_slope <- cont %>% filter(level == -2000)
+cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
+plot(cont_slope_rast, col = "red")
+cont_slope_rast <- rasterize(cont_slope, agg)
+plot(cont_slope_rast, col = "red")
+
+ggplot() + 
+  geom_raster(data = bathy_df, 
+              mapping = aes(x=x, y=y, 
+                            alpha = 0.8,
+                            fill = `gebco_bathy`))+
+  geom_sf(data = cont, col="black") + 
+  geom_sf(data = coast, col = "red") +
+  geom_sf(data = cont_slope, col = "purple") +
+  geom_sf(data = tracks08sf, mapping = aes(col= `BB_20-24000`)) +
+  geom_sf(data = tracks10sf, mapping = aes(col= `BB_20-24000`)) +
+  scale_color_gradient(low="green", high="red") 
+```
+
+### Distance to continental slope
+
+```{r}
+# Drift 8
+slopedist <- terra::distance(trkraster08, vect(cont_slope))     # measure distance from track to slope
+extdist <- terra::extract(slopedist, vect(sf08))       # extract this distance into a df
+tracks_SS_08 <- cbind(sf08,extdist) %>%                     # bind to main df
+  dplyr::select(-ID) %>%
+  rename("dist_slope" = gebco_bathy)
+
+#Drift 10
+slopedist <- terra::distance(trkraster10, vect(cont_slope))     # measure distance from track to slope
+extdist <- terra::extract(slopedist, vect(sf10))       # extract this distance into a df
+tracks_SS_10 <- cbind(sf10,extdist) %>%                     # bind to main df
+  dplyr::select(-ID) %>%
+  rename("dist_slope" = gebco_bathy)
+```
+
+#### Plot distance to continental slope
+
+```{r}
+# By color
+ggplot() +
+  geom_sf(data = pts08, 
+          mapping = aes(col = `Distance to Continental Slope`)) +
+  geom_sf(data = pts10, 
+          mapping = aes(col = `Distance to Continental Slope`)) +
+  geom_sf(data = cont_slope)
+
+# Plot distance by OL/TOL 
+ggplot(data = pts08, aes(x = `Distance to Continental Slope`, y = `OL_125`)) + geom_point()
+ggplot(data = TOL08, aes(x = `Distance to Continental Slope`, y = `TOL_125`)) + geom_point() + geom_smooth()
+   # what are the units for distance? Meters
+
+ggplot() +
+  geom_raster(data = bathy_df, 
+              mapping = aes(x=x, y=y,
+                            fill = `gebco_bathy`)) +
+  geom_sf(data = TOL08, 
+          mapping = aes(col = `Distance to Continental Slope`)) +
+  geom_sf(data = cont_slope) + scale_fill_continuous(low = "lightblue", high = "darkblue")
+
+summary(lm(formula = BB_20.24000 ~ `Distance to Continental Slope`, data = pts08))
+
+# PLOT distance by TOL with whales
+ggplot(data = TOL_whale_08, aes(x = `Distance to Continental Slope`, y = `TOL_125`)) + 
+  geom_point() + 
+  geom_smooth(color = "green") +
+  theme_classic()
+
+# SAME PLOT with size as whale detections
+ggplot(data = TOL_whale_08, aes(x = `Distance to Continental Slope`, y = `TOL_125`)) + 
+  geom_point(aes(size = minNumber)) + 
+  geom_smooth() +
+  theme_classic()
+```
+
+### Third octave levels histograms
+
+```{r}
+# Drift 8 - all TOL bands, facet wrap
+pivot <- tracks_SS_08 %>%
+  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+ggplot(data = pivot, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 1, color = "black") +
+  theme_classic() +
+  ggtitle("Distribution of third octave level intensities: Drift 8") +
+  labs(x = "Sound Intensity (dB)") +
+  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
+  scale_fill_manual(values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.85,0.2)) +
+  facet_wrap(~TOL)
+
+# Drift 10 - all TOL bands, facet grid
+pivot2 <- tracks_SS_10 %>%
+  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 5, color = "black") +
+  theme_classic() +
+  ggtitle("Distribution of third octave level intensities: Drift 10") +
+  scale_x_continuous(name = "Sound Intensity (dB)", breaks=seq(50,110,20)) +
+  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
+  scale_fill_manual(values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.6,0.76)) +
+  facet_grid(~TOL)
+
+# Drift 8 - TOL bands 125 and 2000 only
+pivot3 <- tracks_SS_08 %>%
+  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
+  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
+              `TOL_2000` = "TOL at 2000 Hz")
+
+ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 1, color = "black") +
+  theme_classic() +
+  ggtitle("DASBR 8: Distribution of Third-Octave Level (TOL) Intensities") +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
+  scale_fill_manual(values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.6,0.7)) +
+  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
+
+# Drift 10 - TOL bands 125 and 2000 only
+pivot4 <- tracks_SS_10 %>%
+  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
+  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+ggplot(data = pivot4, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 1, color = "black") +
+  theme_classic() +
+  #ggtitle("DASBR 10: Distribution of Third-Octave Level (TOL) Intensities") +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
+  scale_fill_manual(values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.6,0.7)) +
+  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
+
+# TOL linear model plots
+#plot(lm(formula = `Distance to Continental Slope` ~ bestNumber, data = TOL_whale_08))
+
+#ggplot(TOL_whale_08) +
+#  geom_point(aes(x = TOL_125, y = nClicks), color = "blue") #+
+#  geom_point(aes(x = TOL_63, y =  nClicks), color = "hotpink") +
+#  geom_point(aes(x = TOL_5000, y =  nClicks), color = "limegreen") +
+#  geom_point(aes(x = TOL_2000, y =  nClicks), color = "yellow")
+
+```
+
+### Does hour of day affect sound levels?
+
+```{r}
+# CCES_08
+hr_08 <- track_BB_OL_08 %>%
+  mutate(Hr = hour(track_BB_OL_08$dateTime)) %>%
+  group_by(Hr) %>%
+  summarize(meanBB = mean(`BB_20-24000`)) %>%
+  rename(`Buoy 08 Broadband` = `meanBB`)
+ggplot(data = hr_08, mapping = aes(x = Hr, y = `Buoy 08 Broadband`)) + 
+  geom_point() + geom_smooth()
+
+#CCES_10
+hr_10 <- track_BB_OL_10 %>%
+  mutate(Hr = hour(track_BB_OL_10$dateTime)) %>%
+  group_by(Hr) %>%
+  summarize(meanBB = mean(`BB_20-24000`)) %>%
+  rename(`Buoy 10 Broadband` = `meanBB`)
+ggplot(data = hr_10, mapping = aes(x = Hr, y = `Buoy 10 Broadband`)) + 
+  geom_point() + geom_smooth()
+
+# Join and graph
+hr_join <- left_join(hr_08, hr_10, by = "Hr")
+hr_piv <- hr_join %>%
+  pivot_longer(col=c(`Buoy 08 Broadband`,`Buoy 10 Broadband`), names_to = "buoys", values_to = "BB")
+
+ggplot(data = hr_piv, mapping = aes(Hr, BB, col=buoys)) +
+  geom_point() + geom_smooth()
+```
+
+### Are sound levels affected by species composition or number of clicks?
+
+```{r}
+# Plot
+library(tmap); library(maptiles)
+tmap_mode("plot")
+oceanBase <- get_tiles(whales_sf, provider="Esri.NatGeoWorldMap", crop = TRUE)
+tm_shape(oceanBase) + tm_rgb() +
+  tm_shape(whales_sf) + 
+  tm_symbols(col="species", size = "nClicks") +
+  tm_layout(legend.width = 0.5, 
+            legend.position = c("right","top"))
+
+# CCES 08
+BB_whales08 <- st_join(pts08, whales_sf) %>%
+  filter(!is.na(`species`))
+
+summary(glm(`BB_20.24000` ~ species + `Distance to Continental Slope` + nClicks, data = BB_whales08))
+
+# CCES 10
+BB_whales10 <- st_join(pts10, whales_sf) %>%
+  filter(!is.na(`species`))
+
+summary(glm(`BB_20.24000` ~ species + `Distance to Continental Slope` + nClicks, data = BB_whales10))
+
+#Raster base map
+tmap_mode("plot")
+oceanBase <- get_tiles(BB_whales10, provider="Esri.NatGeoWorldMap")
+tm_shape(bathy) + tm_raster() +
+  tm_shape(BB_whales10) + 
+  tm_symbols(col="species", size = "nClicks", style = "jenks") +
+  tm_layout(legend.width = 0.5)
+
+```
+
+### Explore environmental variables
+
+#### Bathymetry
+
+```{r}
+library(raster)
+bgo <- getNOAA.bathy(lon1 = -121, lon2 = -131,
+                     lat1 = 34, lat2= 40, resolution = 1)  # plot bounding box around drift 8 and 10
+# As raster
+bgo_rast <- marmap::as.raster(bgo)
+
+# Get depth of whale detections in drift 8
+depth08 <- get.depth(bgo, TOL_whale_08[,4:3], locator = FALSE)
+
+# Get depth of tracks
+trkdepth08 <- get.depth(bgo, TOL_whale_08[,4:3], locator = FALSE)
+trkjoin <- left_join(TOL_whale_08, trkdepth08, by = c("long"= "lon")) %>%
+  mutate(depth = depth*-1)%>%     # remove the negative sign
+  mutate(nClicks = coalesce(nClicks, 0))   # replace NAs with zeros
+
+#Plot
+bplot + geom_point(data = depth08, aes(x = lon, y = lat, color = depth, size = 2)) +
+  scale_color_gradient(low = "red", high = "beige") +
+  ggtitle("Depth of Drfit 8 whale detections")
+
+# Join depth08 with TOL_whale_08 on long
+join <- left_join(TOL_whale_08, depth08, by = c("long" = "lon"))
+
+ggplot(trkjoin) + geom_point(aes(x = depth, y = nClicks))
+
+summary(glm(formula = nClicks ~ TOL_125 + `Distance to Continental Slope` + depth + TOL_63 + TOL_2000 + TOL_5000, data = trkjoin, family = poisson))
+```
+
+#### Bathymetric slope
+
+```{r}
+# Slope
+bgo_slope <- terrain(bgo_rast, v = "slope")
+plot(bgo_slope, axes = FALSE)
+
+# Extract slope values at points along drift 8
+slope <- raster::extract(bgo_slope, tracks_SS_08[,4:3])
+tracks_SS_08 <- cbind(tracks_SS_08, slope)
+
+#ggplot() +
+#  geom_raster(data = bgo_slope, mapping = aes(x=x, y=y, fill=slope)) + 
+#  geom_point(data = depth08, aes(x = lon, y = lat, color = depth, size = 2))
+
+# Roughness (different from rugosity?)
+plot(terrain(bgo_rast, opt = "roughness"))    # what are the units on this?
+```
+
+#### Distance from coast - PAMPal
+
+```{r}
+# PAMpal function
+tracks_SS_08 <- as.data.frame(tracks_SS_08) %>% 
+  mutate(Latitude = lat.x, 
+         Longitude = long, 
+         UTC = as.POSIXct(dateTime, format='%Y-%m-%d %H:%M:%S', tz='UTC'))
+
+dist2shore <- erddapToEdinfo(dataset='dist2coast_1deg',
+                             baseurl='https://pae-paha.pacioos.hawaii.edu/erddap/')
+# Downloads nc file
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=dist2shore, fileName='Dist2Shore.nc')
+
+tracks_SS_08 <- matchEnvData(tracks_SS_08, nc='Dist2Shore.nc')  
+```
+
+#### Chlorophyll-a: NOAA Coastwatch
+
+```{r}
+# Code chunk taken from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
+xcoord <- tracks_08_rnd$long
+ycoord <- tracks_08_rnd$lat
+tcoord <- tracks_08_rnd$dateTime
+xlen <- 0.2
+ylen <- 0.2
+parameter <- "chlorophyll"
+dataset <- 'erdMH1chlamday'
+dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")
+
+swchl <- rxtracto(dataInfo, 
+                  parameter=parameter, 
+                  xcoord=xcoord, ycoord=ycoord,
+                  tcoord=tcoord, xlen=xlen, ylen=ylen)
+
+plotTrack(swchl, xcoord, ycoord, tcoord, plotColor = 'algae')
+
+```
+
+### Chlorophyll from PAMPal
+```{r}
+# Downloads from server
+tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdMH1chlamday', fileName='chlorophyll.nc')
+
+# Pulls from downloaded file
+tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'chlorophyll.nc')
+tracks_SS_08
+```
+
+#### Sea surface height anomalies - NOAA Coastwatch
+```{r}
+# Code chunk taken from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
+xcoord <- tracks_08_rnd$long
+ycoord <- tracks_08_rnd$lat
+tcoord <- tracks_08_rnd$dateTime
+xlen <- 0.2
+ylen <- 0.2
+parameter <- 'sla'
+dataset <- 'nesdisSSH1day'
+dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")
+
+swssh <- rxtracto(dataInfo, 
+                  parameter=parameter, 
+                  xcoord=xcoord, ycoord=ycoord, 
+                  tcoord=tcoord, xlen=xlen, ylen=ylen)
+
+plotTrack(swssh, xcoord, ycoord, tcoord, plotColor = 'deep')
+```
+
+#### Sea Surface Height Anomalies from PAMPal
+```{r}
+# Downloaded from server
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'nesdisSSH1day', fileName='sshanomalies.nc')
+
+# Pulled from downloaded nc file
+tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'sshanomalies.nc')
+tracks_SS_08
+```
+
+#### HyCom Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
+```{r}
+hycomData <- matchEnvData(tracks_SS_08, nc=PAMmisc::hycomList, raw=TRUE, depth = 0:1000, fileName = 'hycomData')
+
+```
+
+### Plot maps for Powerpoint images
+
+Map of whole study area
+```{r}
+# Establish bounding box
+bbox <- st_bbox(c(xmin = -110, 
+            xmax = -128, 
+            ymax = 44, 
+            ymin = 30), crs = st_crs(4326))
+
+#Establish basemap
+allTrackbase <- get_tiles(bbox, provider = "Esri.WorldGrayCanvas", zoom = 5)
+
+# Establish inset map
+us_region <- st_bbox(c(xmin = -90,
+            xmax = -120, 
+            ymax = 48, 
+            ymin = 25), crs = st_crs(4326))
+usa <- get_tiles(us_region, provider = "Esri.WorldGrayCanvas", zoom = 4)
+
+sg <- bb_poly(bb(tracks_sf))
+inset <- tm_shape(usa) + tm_rgb() + tm_shape(sg) + tm_borders(col = "red")
+
+#Map all tracks
+main <- tm_shape(allTrackbase) + tm_rgb() +
+  tm_shape(trackssf) +
+  tm_grid(alpha = 0.6, col = "grey") +
+  tm_dots(col = "darkblue", size = 0.001, legend.show = TRUE) +
+  tm_scale_bar(width = 0.3, text.size = 3, position = c(0.2, 0.03)) +
+  tm_compass(type = "arrow", position = c(0.05, 0.05), color.light = "white") +
+  tm_add_legend('symbol',
+                type = "line",
+                col = "darkblue",
+                labels = "Drifting buoy tracks",
+                size = 1,
+                lwd = 2)
+
+vp <- viewport(x=0.66, y=0.8, width = 0.26, height=0.4)
+main
+print(inset,vp=vp)
+
+# Map all tracks - ggplot
+basemap2 <- as.data.frame(allTrackbase, xy = TRUE)
+#ggplot() +
+#  geom_tile(data = basemap, aes(x=x, y=y, fill = "basemap")) +
+#  geom_sf(data = trackssf, mapping = aes(), color = "red")
+
+# Map tracks 08 and 10 only
+tm_shape(trackBase) + tm_rgb() + 
+  tm_shape(tracks08sf) + tm_dots(col = "red", 
+                                 legend.show = TRUE) +
+  tm_shape(tracks10sf) + tm_dots(col = "green")
+
+# Bounding box of drift tracks on map
+lon2 <- -129.04666
+lon1 <- -115.78247
+lat1 <- 28.21790
+lat2 <- 45.09226
+
+basemap3 <- ne_countries(scale = "medium", type = "map_units", returnclass = "sf")
+
+# Drift tracks
+ggplot(data = basemap3) +
+  geom_sf(fill = "antiquewhite") +
+  coord_sf(xlim = c(lon2 - 1, lon1 + 1), ylim = c(lat1 - 1, lat2 + 1), expand = FALSE) +
+  theme(panel.grid.major = element_blank()) +
+  geom_path(data = tracks_sf, mapping = aes(x=long,y=lat, group = factor(station)), color = "deeppink") +
+  theme(panel.background = element_rect(fill = "aliceblue"), 
+        axis.text = element_blank(), axis.ticks = element_blank(), 
+        axis.title = element_blank())
+```
+
+Map of soundscapes
+```{r}
+# CCES_08 + CCES_10 Broadband
+tm_shape(trackBase) + tm_rgb() + 
+  tm_shape(tracks08sf) + 
+  tm_symbols(col = "BB_20-24000", 
+             border.lwd = NA, 
+             size = 0.4, 
+             sizes.legend = 0.7, 
+             n = 10) +
+  tm_shape(tracks10sf) + 
+  tm_symbols(col = "BB_20-24000",
+             border.lwd = NA, 
+             size = 0.4,
+             n = 10) +
+  tm_layout(legend.width = 0.5, 
+            legend.position = c("right","top"),
+            legend.bg.color = "lightblue")
+
+#CCES_08 Octave Level
+tm_shape(trackBase) + tm_rgb() + 
+  tm_shape(tracks08sf) + 
+  tm_symbols(col = "OL_63", 
+             border.lwd = NA, 
+             size = 0.5, 
+             n = 10) + 
+  tm_layout(legend.width = 0.5, 
+            legend.position = c("right","top")) + 
+  tm_legend()
+
+ggplot() + 
+  geom_sf(data = trk_TOL_08, aes(color = TOL_125, size = TOL_125), show.legend = FALSE) +
+  geom_sf(data = trk_TOL_10, aes(color = TOL_125, size = TOL_125)) +
+  scale_x_continuous(limits = c(-130,-122)) + 
+  scale_y_continuous(limits = c(35,39)) +
+  scale_color_gradient(low = "navy", high = "yellow")+
+  scale_size(range = c(1,12)) +
+  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
+  theme(panel.border = element_rect(fill = NA),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(), 
+        axis.title = element_blank(),
+        panel.background = element_blank(),
+        legend.position = "none")
+```
+
+Map of whale detections
+```{r}
+# Whale detections
+ggplot() +
+  geom_point(data = whales8, mapping = aes(x=long, y=lat), shape = 21, color = "black", fill = "green4", size = 4) +
+  geom_point(data = whales10, mapping = aes(x=long, y=lat), shape = 21, color = "black", fill = "green4", size = 4) +
+  scale_x_continuous(limits = c(-130,-122)) + 
+  scale_y_continuous(limits = c(35,39)) +
+  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
+  theme(panel.border = element_rect(fill = NA),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(), 
+        axis.title = element_blank(),
+        panel.background = element_blank()
+        )
+#ggsave("BWdetect.png")
+
+```
+
+Map of bathymetry
+```{r}
+# Bathymetry
+bathym <- ggplot() +
+    geom_raster(data = bathy_df, 
+              mapping = aes(x=x, y=y,
+                            fill = `gebco_bathy`)) +
+  annotation_scale(location = "bl", width_hint = 0.4) +
+  theme(axis.text = element_blank(), 
+        axis.ticks = element_blank(), 
+        axis.title = element_blank(),
+        panel.background = element_blank(),
+        panel.border = element_rect(fill = NA))
+
+# Alternative Bathymetry (code from Ocean Tracking Network #https://www.youtube.com/watch?v=xT2F-Y9bKLk)
+bgo <- getNOAA.bathy(lon1 = -122, lon2 = -130,
+                     lat1 = 35, lat2= 39, resolution = 1)  # plot bounding box around drift 8 and 10
+autoplot(bgo, geom = c("raster", "contour"), coast = TRUE, show.legend = FALSE) + 
+  scale_fill_etopo() + #creates ggplot object
+  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
+  theme(axis.text = element_blank(), 
+        axis.ticks = element_blank(), 
+        axis.title = element_blank(),
+        panel.background = element_blank())
+
+# Soundscape + bathymetry
+ggplot() +
+  geom_raster(data = bathy_df, 
+              mapping = aes(x=x, y=y,
+                            fill = `gebco_bathy`), show.legend = FALSE) +
+  #geom_contour(data = bathy_df, mapping = aes(x=x, y=y, z=gebco_bathy, linewidth = 0.1), color = "blue", linewidth = 0.01,) +
+  geom_sf(data = trk_TOL_08, aes(color = TOL_125, size = TOL_125), show.legend = FALSE) +
+  geom_sf(data = tracks10sf, aes(color = TOL_125, size = TOL_125)) +
+  geom_sf(data = whales08_sf, aes(), color = "black") +
+  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
+  scale_color_gradient(low = "beige", high = "hotpink") +
+  scale_fill_etopo() +
+  theme(axis.text = element_blank(), 
+        axis.ticks = element_blank(), 
+        axis.title = element_blank(),
+        panel.background = element_blank(),
+        panel.border = element_rect(fill = NA))
+#ggsave("soundscape.png")
+```
+
+Map of AIS
+```{r}
+# Plot + crop to drift area
+basemap <- ne_countries(scale = "medium", returnclass = "sf")
+ggplot() + 
+  geom_sf(aes(), basemap) +
+  geom_path(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName), show.legend = F) + 
+  #geom_point(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName)) +
+  scale_x_continuous(limits = c(min(drift$long-0.7), max(drift$long + 0.7))) + 
+  scale_y_continuous(limits = c(min(drift$lat-0.7), max(drift$lat + 0.7)))
+
+# AIS data
+ggplot() + 
+  geom_path(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName, linewidth = 1), show.legend = F) +
+  scale_x_continuous(limits = c(-131,-121)) + 
+  scale_y_continuous(limits = c(34,40)) +
+  theme(axis.text = element_blank(), 
+        axis.ticks = element_blank(), 
+        axis.title = element_blank(),
+        panel.background = element_blank(),
+        panel.border = element_rect(fill = NA))
+#ggsave("AIS.png")
+```

### 2023-03-08 outline.qmd
updated with progress

---
title: "Thesis outline"
format: docx
editor: visual
---

## Goal 1: Characterize the soundscape

#### Explore the relationship between soundscape metrics and environmental variables, and beaked whale (BW) detections

-   Create histograms, scatterplots, time series, facet graphs of:
    -   A soundscape metric/BW presence + absence/env variable
        -   Comparing TOL intensities between drift 8 and 10
        -   Broadband levels, Third-octave levels at 63, 125, 2000, 5000, 20000 Hz
        -   Probability of detecting BW is highest when soundscape is \_\_\_.
        -   Which frequencies had the most loudness?
    -   Enviromental variables (source from PAMpal for events or ERDAPP)
        -   Bottom depth (**depth for whale detections in drift 8**)

        -   Bottom slope (**done for drift 8**)

        -   Distance from shore **(done for drift 8)**

        -   Distance from continental shelf (**done for drift 8**)

        -   Sea surface temperature (**in progress)**

        -   Sea surface height (need help sourcing, can only find SSH anomalies for 2018)

        -   Sea surface slope

        -   Chlorophyll-a (**mapped for drift 8)**

        -   Mixed layer depth (MLD), thermocline temperature (TTemp), temperature at 400 m, salinity (**Taiki's code)**
    -   Other variables: seasonality, hour of the day (create time series for avg daily env.variable)
    -   Questions: Does soundscape change when conditions change? Is sound seasonal?
-   Plug into model (GAM, zero-inflated, GLM)

## Goal 2: Influence from vessels

#### Explore how vessel presence influences the soundscape

-   Find instances of vessel in vicinity of drift vs. drift alone and compare

    -   Number of ship passages

        -   Create buffer around whale detections (unsure of buffer size)
        -   Find AIS vessels by lat/long that lie within buffer
        -   Quantify number of vessels that pass for each drift

    -   Closest ship passage (When a ship has a close encounter with a drift, how close is it on average? How is soundscape affected?)

-   Is vessel presence = loud soundscape?

    -   Compare TOL125 intensities vs. ship presence

-   Chronic vs. acute sound events

    -   Instances where drifts and vessels overlap, plot time on x-axis and TOL125 dB on y

### 2023-03-08 hycomThermocline[2998].rmd
Create hycomThermocline[2998].rmd

---
title: "HYCOM Data Access and Thermocline Variables"
author: "Taiki Sakai"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Prep some fake data. Environmental download functions require columns
named exatly `Longitude`, `Latitude`, and `UTC`. `UTC` must be converted
to POSIXct format.

```{r}
data <- data.frame(Latitude = c(38, 22.2, 15),
                   Longitude = c(-123.5, -152.9, 144),
                   UTC = c('2022-09-27 19:00:48',
                           '2022-09-27 17:57:48',
                           '2018-07-20 20:49:00'))
data$UTC <- as.POSIXct(data$UTC, format='%Y-%m-%d %H:%M:%S', tz='UTC')
```

Install latest PAMmisc from GitHub

```{r, eval=FALSE}
# only run this chunk once
devtools::install_github('TaikiSan21/PAMmisc')
```

## HYCOM Download

HYCOM has different datasets depending on the date, by passing this
`hycomList` object `PAMmisc` will sort out which one you need based on
your date ranges in your data, even figuring out if your download needs
to span multiple datasets.

The Hycom servers can be a bit slow, so this may take some time

```{r, cache=TRUE}
library(PAMmisc)
hycomData <- matchEnvData(data, nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
```

Normally the `matchEnvData` function will just add the environmental
variables to your dataframe, averaging them over any depth values. By
setting `raw=TRUE`, we instead have it return all values at the depths
so that we can work with them. The output is a list with length equal to
the number of rows in our dataframe.

```{r}
print(hycomData[[1]])
```

The matchXXX data show the coordinate values in the NetCDF files that
your data were matched to. The function will match your data to the
closest value in the NetCDF file.

## Thermocline Varibales

Now lets use the temperature at depth data to estimate thermocline
variables. This function returns estimates of mixed layer depth
(`mldDepth`), mixed layer temperature (`mldTemp`), thermocline depth
(`ttDepth`), and thermocline temperature (`ttTemp`). You don't need to
worry about the xxxMode parameters, they are place holders in case we
want to add different estimation methods in the future.

```{r}
calcThermoVars <- function(depth, temp, mldMode='SST8', thermMode='variso', plot=FALSE) {
    # depthInterp <- approxfun(x=temp, y=depth)
    switch(match.arg(mldMode),
           'SST8' = {
               mldTemp <- temp[1] - 0.8
               # first few temps are likely to have repeat values that trigger
               # warnings for approxfun, so cut off to only values after mldTemp
               useMin <- min(which(temp <= mldTemp)) - 1
               useIx <- useMin:length(temp)
               depthInterp <- approxfun(x=temp[useIx], y=depth[useIx])
               mld <- depthInterp(mldTemp)
           }
    )
    switch(match.arg(thermMode),
           'variso' = {
               t400 <- approx(x=depth, y=temp, xout=400)$y
               # therm end is half way to t400 temp
               tEnd <- mldTemp + (t400 - mldTemp)/2
               # therm temp is midpt from mld to end
               tt <- (mldTemp + tEnd)/2
               # tt <- mldTemp - .25 * (mldTemp - t400)
               ttDepth <- depthInterp(tt)
               tEndDepth <- depthInterp(tEnd)
           }
    )
    if(plot) {
        to300 <- depth <= 300
        plot(x=temp[to300], y=-depth[to300],
             main = paste0('MixLayer: ', round(mld,0), 'm ', round(mldTemp, 1), 'C',
                           '\nTherm: ', round(ttDepth, 0), 'm ', round(tt, 1), 'C'),
             xlab='Temp (C)', ylab='Depth (m)')

        lines(x=temp[to300], y=-depth[to300])
        lines(x=c(mldTemp, tEnd), y=c(-mld, -tEndDepth), col='darkgray', lwd=4)
        points(x=c(mldTemp, tt), y=c(-mld, -ttDepth), col='blue', cex=1.5, pch=15)
    }
    list(mldDepth = mld, mldTemp = mldTemp,
         ttDepth = ttDepth, ttTemp = tt)
}
```

We can calculate these and plot the results for our first data point.
The points are the temperature values at the depths in the HYCOM data.
The light gray line is the estimated thermocline based on the
calculations in [Fiedler
2010](https://aslopubs.onlinelibrary.wiley.com/doi/epdf/10.4319/lom.2010.8.313).
The blue squares mark the calculated mixed layer depth and thermocline
depth.

```{r}
calcThermoVars(hycomData[[1]]$matchDepth, hycomData[[1]]$water_temp, plot=TRUE)
```

And here's how we would connect these to our original dataframe, adding
in the temp & salinity at 400m. The list `hycomData` is in the same
order as the original rows, so we can just `cbind` them.

```{r}
library(dplyr)
thermData <- bind_rows(lapply(hycomData, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))
data <- cbind(data, thermData)
print(data)
```

### 2023-04-05 readin.R
Delete readin.R

@@ -1,178 +1,0 @@
-
-# Read in, load, and wrangle raw data
-
-###########################
-# Required packages
-###########################
-
-library(tidyverse)
-library(lubridate)
-library(sf)
-library(rerddap)
-library(terra)
-library(marmap)
-library(rnaturalearth)
-library(mapproj)
-library(raster)
-library("rerddapXtracto")
-library(PAMpal)
-library(maptiles)
-library(tmap)
-library(spData)
-library(tmaptools)
-library(grid)
-library(ggspatial)
-
-
-##########################
-# Read in soundscape data, NEED TO MAKE THIS A FUNCTION
-##########################
-
-# Round to nearest 20 minutes, create dateTime data type
-# Drift 8: Broadband
-BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
-BB_08 <- BB_08_csv %>% 
-  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
-BB_08 <- BB_08 %>%
-  mutate(dateTime = round_date(ymd_hms(BB_08$dateTime),"20 minutes"))
-
-# Drift 8: Octave Level
-OL_08_csv <- read_csv("Data/CCES_08_1Hz_1s_OL_2min.csv",show_col_types = FALSE)
-OL_08 <- OL_08_csv %>% 
-  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
-OL_08 <- OL_08 %>%
-  mutate(dateTime = round_date(ymd_hms(OL_08$dateTime),"20 minutes"))
-
-# Drift 8: TOL
-TOL_08_csv<- read_csv("data/CCES_08_1Hz_1s_TOL_2min.csv", show_col_types = FALSE)
-TOL_08 <- TOL_08_csv %>% 
-  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
-TOL_08 <- TOL_08 %>%
-  mutate(dateTime = round_date(ymd_hms(TOL_08$dateTime),"20 minutes"))
-
-# Drift 10: Broadband
-BB_10_csv <- read_csv("Data/CCES_10_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
-BB_10 <- BB_10_csv %>% 
[…]

### 2023-04-05 outline.qmd
Update outline.qmd

@@
 title: "Thesis outline"
+author: Marina Bozinovic
 format: docx
@@
 
-## Goal 1: Characterize the soundscape
+## Goal 1: Characterize the soundscape/acoustic habitat
 
-#### Explore the relationship between soundscape metrics and environmental variables, and beaked whale (BW) detections
+#### Explore the relationship (spatial/temporal variation?) between soundscape metrics and environmental variables, and beaked whale (BW) detections.
+(Alternative: Describe the acoustic environment where deep-diving species are present/absent.)
 
--   Create histograms, scatterplots, time series, facet graphs of:
+-   Exploratory: Create histograms, scatterplots, time series, facet graphs of:
     -   A soundscape metric/BW presence + absence/env variable
-        -   Comparing TOL intensities between drift 8 and 10
+        -   Comparing TOL intensities between drift 8 and 10 (**only two drifts for now**)
         -   Broadband levels, Third-octave levels at 63, 125, 2000, 5000, 20000 Hz
-        -   Probability of detecting BW is highest when soundscape is \_\_\_.
-        -   Which frequencies had the most loudness?
-    -   Enviromental variables (source from PAMpal for events or ERDAPP)
-        -   Bottom depth (**depth for whale detections in drift 8**)
+            -   Are we interested in high frequencies (20 kHz)?
+        -   Which frequencies had the most loudness? What does the soundscape look like when there are high BW detections?
+    -   Oceanographic characteristics
+        -   Bottom depth (**done for drift 8**)
 
@@
 
-        -   Sea surface temperature (**in progress)**
+        -   Sea surface temperature **(done for drift 8**)
 
-        -   Sea surface height (need help sourcing, can only find SSH anomalies for 2018)
+        -   Sea surface height anomalies **(done for drift 8)**
 
-        -   Sea surface slope
+        -   Sea surface height slope (in progress)
 
-        -   Chlorophyll-a (**mapped for drift 8)**
+        -   Curl of wind stress (in progress)
 
-        -   Mixed layer depth (MLD), thermocline temperature (TTemp), temperature at 400 m, salinity (**Taiki's code)**
-    -   Other variables: seasonality, hour of the day (create time series for avg daily env.variable)
+        -   Mixed layer depth (MLD) **(done for drift 8**)
+
+        -   Mixed layer depth Temperature **(done for drift 8**)
+
+        -   Thermocline temperature (TTemp) **(done for drift 8**)
+
+        -   Thermocline depth **(done for drift 8**)
+
+        -   Temperature at 400 m **(done for drift 8**)
+
+        -   Salinity at 400 m **(done for drift 8**)
+
+        -   Chlorophyll-a **(done for drift 8**)
+    -   Other potential variables: seasonality, hour of the day (create time series for avg daily env.variable)
     -   Questions: Does soundscape change when conditions change? Is sound seasonal?
--   Plug into model (GAM, zero-inflated, GLM)
+-   Fit a model (GAM, possibly zero-inflated, GLM, random forest, AIC)
 
-## Goal 2: Influence from vessels
+## Goal 2: Influence from vessels (connect sound to a source)
 
-#### Explore how vessel presence influences the soundscape
+#### Explore how vessel presence influences and contributes to the soundscape
 
@@
 
-    -   Number of ship passages
+    -   Number of ship passages, types of ships (**IN PROGRESS**)
 
-        -   Create buffer around whale detections (unsure of buffer size)
+        -   Create buffer around whale detections (using 50nmi)
+
         -   Find AIS vessels by lat/long that lie within buffer
-        -   Quantify number of vessels that pass for each drift
+
+        -   Identify number of vessels that pass each drift
 
@@
 
--   Is vessel presence = loud soundscape?
+-   Does vessel presence create a louder soundscape?
 
-    -   Compare TOL125 intensities vs. ship presence
+    -   Compare TOL 63 and125 intensities vs. ship presence
 
@@
 
-    -   Instances where drifts and vessels overlap, plot time on x-axis and TOL125 dB on y
+    -   Instances where drifts and vessels overlap, plot time on x-axis and TOL125 on y axis

### 2023-04-05 02_test.Rmd
Create 02_test.Rmd

---
title: "exploratory analysis"
output: html_document
date: "2022-11-06"
---

Required Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(sf)
library(rerddap)
library(terra)
library(marmap)
library(rnaturalearth)
library(mapproj)
library(raster)
library("rerddapXtracto")
library(PAMpal)
library(PAMmisc)
library(maptiles)
library(tmap)
library(spData)
library(tmaptools)
library(grid)
library(ggspatial)
```

Set Parameters
```{r}
theme_set(theme_bw())
```

### Exploratory analysis

#### Not important: tracks as sf and basemaps

```{r}
library(tmap); library(maptiles); library(sf)
tmap_mode("plot")
trackssf <- st_as_sf(tracks, coords = c("long", "lat"), crs=4326)
#tracks08sf <- st_as_sf(track_BB_OL_08, coords = c("long", "lat"), crs=4326)  # outdated
#tracks10sf <- st_as_sf(track_BB_OL_10, coords = c("long", "lat"), crs=4326)  # outdated
trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("long", "lat.x"), crs=4326)
trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("long", "lat.x"), crs=4326)
trackBase <- get_tiles(whales_8_10sf, provider = "Esri.WorldShadedRelief")
allTrackbase <- get_tiles(trackssf, provider = "Stamen.TonerLite", zoom = 5)
```

### Sea Surface Temp with rerddap, Drift 8 only
```{r}
# Multi-scale Ultra-high Resolution (MUR) SST Analysis fv04.1, Global, 0.01°, 2002-present, Daily
SSTInfo <- info('jplMURSST41') 
murSST <- griddap(SSTInfo, latitude = c(35, 39), longitude = c(-122, -130), 
                  time = c("2018-08-16", "2018-10-01"), 
                  fields = 'analysed_sst') # drift8 dates only
SST <- murSST$data

# Raster of SST
ggplot() + 
  geom_tile(data = SST, aes(x = longitude, y = latitude, fill = analysed_sst)) +
  scale_fill_gradient(low = "darkblue", high = "yellow") +
  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(), 
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(fill = NA),
        legend.position = "none")
```

### SST from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
```{r}
xcoord <- tracks_08$long
ycoord <- tracks_08$lat
tcoord <- tracks_08$dateTime
xlen <- 0.2
ylen <- 0.2
parameter <- "analysed_sst"
dataset <- 'jplMURSST41'
#dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")

#swchl <- rxtracto(dataInfo, 
#                  parameter=parameter, 
#                  xcoord=xcoord, ycoord=ycoord, 
#                  tcoord=tcoord, xlen=xlen, ylen=ylen)

#plotTrack(swchl, xcoord, ycoord, tcoord, plotColor = 'thermal')
```

### SST from PAMPal
```{r}
# Downloads data from servers
#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "analysedSST") %>%

# from file 
tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/analysedSST.nc') %>%
  rename(., sst_mean = analysed_sst_mean) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)   # rename new column to "sst_mean" and remove unnecessary columns
tracks_SS_08
```

### Distance to continental slope

```{r}
# Increase cell size by 30x to be visible
bathy <- rast("data/gebco_bathy.tif")
agg <- terra::aggregate(bathy, 20)
plot(agg)

# Rasterize track with agg as the template
#track_rast <- rasterize(tracks08sf, agg, field = "BB_20-24000")
#TOL_rast <- rasterize(trk_TOL_08, agg, field = "TOL_125")
sf08 <- st_as_sf(tracks_SS_08, coords = c("long", "lat.x"), crs=4326, remove=F)
sf10 <- st_as_sf(tracks_SS_10, coords = c("long", "lat.x"), crs=4326, remove=F)
trkraster08 <- rasterize(vect(sf08), agg)
trkraster10 <- rasterize(vect(sf10), agg)

bathy_df <- as.data.frame(bathy, xy = TRUE)

# Map the coastline and cont. slope
cont <- st_as_sf(as.contour(bathy))
coast <- cont %>% filter(level == 0)
cont_slope <- cont %>% filter(level == -2000)
cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
plot(cont_slope_rast, col = "red")
cont_slope_rast <- rasterize(cont_slope, agg)
plot(cont_slope_rast, col = "red")

ggplot() + 
  geom_raster(data = bathy_df, 
              mapping = aes(x=x, y=y, 
                            alpha = 0.8,
                            fill = `gebco_bathy`))+
  geom_sf(data = cont, col="black") + 
  geom_sf(data = coast, col = "red") +
  geom_sf(data = cont_slope, col = "purple") +
  geom_sf(data = whales_8_10sf, mapping = aes()) +
  scale_color_gradient(low="green", high="red") 
```

### Extract slope_dist and bind to drift tracks
```{r}
# Drift 8
slopedist <- terra::distance(trkraster08, vect(cont_slope))     # measure distance from track to slope
extdist <- terra::extract(slopedist, vect(sf08))       # extract this distance into a df
tracks_SS_08 <- cbind(sf08,extdist) %>%                     # bind to main df
  dplyr::select(-ID) %>%
  rename("dist_slope" = gebco_bathy)

#Drift 10
slopedist <- terra::distance(trkraster10, vect(cont_slope))     # measure distance from track to slope
extdist <- terra::extract(slopedist, vect(sf10))       # extract this distance into a df
tracks_SS_10 <- cbind(sf10,extdist) %>%                     # bind to main df
  dplyr::select(-ID) %>%
  rename("dist_slope" = gebco_bathy)
```

#### Plot distance to continental slope

```{r}
# By color
#ggplot() +
#  geom_sf(data = pts08, 
#          mapping = aes(col = `Distance to Continental Slope`)) +
#  geom_sf(data = pts10, 
#          mapping = aes(col = `Distance to Continental Slope`)) +
#  geom_sf(data = cont_slope)

# Plot distance by OL/TOL 
#ggplot(data = pts08, aes(x = `Distance to Continental Slope`, y = `OL_125`)) + geom_point()
#ggplot(data = TOL08, aes(x = `Distance to Continental Slope`, y = `TOL_125`)) + geom_point() + geom_smooth()
   # what are the units for distance? Meters

ggplot() +
  geom_raster(data = bathy_df, 
              mapping = aes(x=x, y=y,
                            fill = `gebco_bathy`)) +
  geom_sf(data = tracks_SS_08, 
          mapping = aes(col = dist_slope)) +
  geom_sf(data = cont_slope) + scale_fill_continuous(high = "lightgreen", low = "blue")

# PLOT distance by TOL with whales
ggplot(data = tracks_SS_08, aes(x = dist_slope, y = `TOL_125`)) + 
  geom_point() + 
  geom_smooth(color = "green") +
  theme_classic()
```

### Third octave levels histograms

```{r}
# Drift 8 - all TOL bands, facet wrap
pivot <- tracks_SS_08 %>%
  pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid

ggplot(data = pivot, aes(Hz, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 3, color = "black") +
  theme_classic() +
  ggtitle("Distribution of third octave level intensities: Drift 8") +
  labs(x = "Sound Intensity (dB)") +
  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
  scale_fill_manual(values = c("0" = "bisque1",
                               "1" = "brown4")) +
  theme(legend.position = c(0.7,0.8)) +
  facet_grid(~SL)

# Drift 10 - all TOL bands, facet grid
pivot2 <- tracks_SS_10 %>%
  pivot_longer(cols = starts_with("TOL")| starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid

ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 3, color = "black") +
  theme_classic() +
  ggtitle("Distribution of third octave level intensities: Drift 10") +
  scale_x_continuous(name = "Sound Intensity (dB)", breaks=seq(50,110,20)) +
  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
  scale_fill_manual(values = c("0" = "bisque1",
                               "1" = "brown4")) +
  theme(legend.position = c(0.6,0.76)) +
  facet_grid(~SL)

# Drift 8 - TOL bands 125 and 2000 only
pivot3 <- tracks_SS_08 %>%
  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid

TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
              `TOL_2000` = "TOL at 2000 Hz")
```

```{r drift-8-histogram}
ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 1, color = "black") +
  theme_classic() +
  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 8") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_manual(name = "Beaked Whale Presence", 
                    labels = c("0" = "Absent", "1" = "Present"),
                    values = c("0" = "bisque1",
                               "1" = "brown4")) +
  theme(legend.position = c(0.6,0.7)) +
  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
```

```{r drift-8-histogram}
```

```{r}
# Drift 10 - TOL bands 125 and 2000 only
pivot4 <- tracks_SS_10 %>%
  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid

ggplot(data = pivot4, aes(Hz, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 1, color = "black") +
  theme_classic() +
  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 10") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_manual(name = "Beaked Whale Presence", 
                    labels = c("0" = "Absent", "1" = "Present"),
                    values = c("0" = "bisque1",
                               "1" = "brown4")) +
  theme(legend.position = c(0.6,0.7)) +
  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))

TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
              `TOL_2000` = "TOL at 2000 Hz")
# TOL linear model plots
#plot(lm(formula = `Distance to Continental Slope` ~ bestNumber, data = TOL_whale_08))

#ggplot(TOL_whale_08) +
#  geom_point(aes(x = TOL_125, y = nClicks), color = "blue") #+
#  geom_point(aes(x = TOL_63, y =  nClicks), color = "hotpink") +
#  geom_point(aes(x = TOL_5000, y =  nClicks), color = "limegreen") +
#  geom_point(aes(x = TOL_2000, y =  nClicks), color = "yellow")

```

### Does hour of day affect sound levels?

```{r}
# CCES_08
#hr_08 <- track_BB_OL_08 %>%
#  mutate(Hr = hour(track_BB_OL_08$dateTime)) %>%
#  group_by(Hr) %>%
#  summarize(meanBB = mean(`BB_20-24000`)) %>%
#  rename(`Buoy 08 Broadband` = `meanBB`)
#ggplot(data = hr_08, mapping = aes(x = Hr, y = `Buoy 08 Broadband`)) + 
#  geom_point() + geom_smooth()

#CCES_10
#hr_10 <- track_BB_OL_10 %>%
#  mutate(Hr = hour(track_BB_OL_10$dateTime)) %>%
#  group_by(Hr) %>%
#  summarize(meanBB = mean(`BB_20-24000`)) %>%
#  rename(`Buoy 10 Broadband` = `meanBB`)
#ggplot(data = hr_10, mapping = aes(x = Hr, y = `Buoy 10 Broadband`)) + 
#  geom_point() + geom_smooth()

# Join and graph
#hr_join <- left_join(hr_08, hr_10, by = "Hr")
#hr_piv <- hr_join %>%
#  pivot_longer(col=c(`Buoy 08 Broadband`,`Buoy 10 Broadband`), names_to = "buoys", values_to = "BB")

#ggplot(data = hr_piv, mapping = aes(Hr, BB, col=buoys)) +
#  geom_point() + geom_smooth()
```

### Are sound levels affected by species composition or number of clicks?

```{r}
# Plot
library(tmap); library(maptiles)
tmap_mode("plot")
oceanBase <- get_tiles(tracks_SS_08, provider="Esri.NatGeoWorldMap", crop = TRUE)
tm_shape(oceanBase) + tm_rgb() +
  tm_shape(tracks_SS_08) +  tm_symbols(col="species", size = 1) +
  tm_layout(legend.width = 0.5, 
            legend.position = c("right","top"))

# CCES 08
#BB_whales08 <- st_join(pts08, whales_sf) %>%
#  filter(!is.na(`species`))

#summary(glm(`BB_20.24000` ~ species + `Distance to Continental Slope` + nClicks, data = BB_whales08))

# CCES 10
#BB_whales10 <- st_join(pts10, whales_sf) %>%
#  filter(!is.na(`species`))

#summary(glm(`BB_20.24000` ~ species + `Distance to Continental Slope` + nClicks, data = BB_whales10))

#Raster base map
tmap_mode("plot")
oceanBase <- get_tiles(tracks_SS_10, provider="Esri.NatGeoWorldMap")
tm_shape(bathy) + tm_raster() +
  tm_shape(tracks_SS_10) + 
  tm_symbols(style = "jenks") +
  tm_layout(legend.width = 0.5)

```

#### Bathymetry

```{r}
bgo <- getNOAA.bathy(lon1 = -121, lon2 = -131,
                     lat1 = 34, lat2= 40, resolution = 1)  # plot bounding box around drift 8 and 10
# As raster
bgo_rast <- marmap::as.raster(bgo)

# Convert sf to dataframe for this operation
tracks_SS_08 <- st_set_geometry(tracks_SS_08, value = NULL)

# Get depth of whale detections in drift 8
depth08 <- get.depth(bgo, tracks_SS_08[,3:2], locator = FALSE)

# Get depth of tracks
trkdepth08 <- get.depth(bgo, tracks_SS_08[,3:2], locator = FALSE)
trkjoin <- left_join(tracks_SS_08, trkdepth08, by = c("long"= "lon")) %>%
  mutate(depth = depth*-1)         # remove the negative sign

#Plot
ggplot() + geom_point(data = tracks_SS_08, aes(x = long, y = lat.x, size = 2)) +
  scale_color_gradient(low = "red", high = "beige") +
  ggtitle("Depth of Drift 8 whale detections")
```
#### Bathymetry - PAMpal
```{r}
# Downloads from server
#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'srtm15plus', fileName='depth.nc')

# Pulls from downloaded file
tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/depth.nc') %>%
  rename(., depth = z_mean) %>%        # rename new column to "depth"
  mutate(depth = depth*-1) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
tracks_SS_08
```

#### Bathymetric slope

```{r}
# Slope
bgo_slope <- terrain(bgo_rast, v = "slope")
plot(bgo_slope, axes = FALSE)

# Extract slope values at points along drift 8
slope <- raster::extract(bgo_slope, tracks_SS_08[,3:2])
tracks_SS_08 <- cbind(tracks_SS_08, slope)

#ggplot() +
#  geom_raster(data = bgo_slope, mapping = aes(x=x, y=y, fill=slope)) + 
#  geom_point(data = depth08, aes(x = lon, y = lat, color = depth, size = 2))

# Roughness (different from rugosity?)
plot(terrain(bgo_rast, opt = "roughness"))    # what are the units on this?
```

#### Distance from coast - PAMPal

```{r}
dist2shore <- erddapToEdinfo(dataset='dist2coast_1deg',
                             baseurl='https://pae-paha.pacioos.hawaii.edu/erddap/')
# Downloads nc file
#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=dist2shore, fileName='Dist2Shore.nc')

tracks_SS_08 <- matchEnvData(tracks_SS_08, nc='data/Dist2Shore.nc') %>%
  rename(., dist_shore = dist_mean) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
tracks_SS_08
```

#### Chlorophyll-a: NOAA Coastwatch

```{r}
# Code chunk taken from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
xcoord <- tracks_SS_08$long
ycoord <- tracks_SS_08$lat
tcoord <- tracks_SS_08$dateTime
xlen <- 0.2
ylen <- 0.2
parameter <- "chlorophyll"
dataset <- 'erdMH1chlamday'
dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")

swchl <- rxtracto(dataInfo, 
                  parameter=parameter, 
                  xcoord=xcoord, ycoord=ycoord,
                  tcoord=tcoord, xlen=xlen, ylen=ylen)

plotTrack(swchl, xcoord, ycoord, tcoord, plotColor = 'algae')

```

### Chlorophyll from PAMPal
```{r}
# Downloads from server
#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdMH1chlamday', fileName='chlorophyll.nc')

# Pulls from downloaded file
tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/chlorophyll.nc')
tracks_SS_08
```

#### Sea surface height anomalies - NOAA Coastwatch
```{r}
# Code chunk taken from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
xcoord <- tracks_SS_08$long
ycoord <- tracks_SS_08$lat
tcoord <- tracks_SS_08$dateTime
xlen <- 0.2
ylen <- 0.2
parameter <- 'sla'
dataset <- 'nesdisSSH1day'
dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")

swssh <- rxtracto(dataInfo, 
                  parameter=parameter, 
                  xcoord=xcoord, ycoord=ycoord, 
                  tcoord=tcoord, xlen=xlen, ylen=ylen)

plotTrack(swssh, xcoord, ycoord, tcoord, plotColor = 'deep')
```

#### Sea Surface Height Anomalies from PAMPal
```{r}
# Downloaded from server
#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'nesdisSSH1day', fileName='sshanomalies.nc')

# Pulled from downloaded nc file
tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/sshanomalies.nc') %>%
  rename(., ssh_mean = sla_mean) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
tracks_SS_08
```

### Wind Stress Curl from PAMPal *** NEEDS FIXING ****
```{r}
# Downloads from server
#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQSstress1day_LonPM180', #fileName='erdQSstress1day_LonPM180.nc', var = 'curl', depth = 0)

# Pulls from downloaded file
#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQSstress1day_LonPM180.nc')
#tracks_SS_08

#windcurl <- erddapToEdinfo('erdQSstress1day_LonPM180', baseurl = 'https://coastwatch.pfeg.noaa.gov/erddap/')

#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=windcurl, fileName='windcurl.nc')  
```

#### HyCom Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
```{r}
# These refer to nc files that have already been downloaded from the server

#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = #"hycomm0.nc")

#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = #"hycomm1.nc")

#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = #"hycomm2.nc")

#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName #= "hycomm3.nc")

#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName #= "hycomm4.nc")

#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName #= "hycomm5.nc")

#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName #= "hycomm6.nc")

# Run the 7 nc lists that make up tracks_SS_08
hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm0.nc")
hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm1.nc")
hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm2.nc")
hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm3.nc")
hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm4.nc")
hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm5.nc")
hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2140,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm6.nc")
```

Use function to estimate thermocline variables using the temperature at depth data
```{r}
calcThermoVars <- function(depth, temp, mldMode='SST8', thermMode='variso', plot=FALSE) {
    # depthInterp <- approxfun(x=temp, y=depth)
    switch(match.arg(mldMode),
           'SST8' = {
               mldTemp <- temp[1] - 0.8
               # first few temps are likely to have repeat values that trigger
               # warnings for approxfun, so cut off to only values after mldTemp
               useMin <- min(which(temp <= mldTemp)) - 1
               useIx <- useMin:length(temp)
               depthInterp <- approxfun(x=temp[useIx], y=depth[useIx])
               mld <- depthInterp(mldTemp)
           }
    )
    switch(match.arg(thermMode),
           'variso' = {
               t400 <- approx(x=depth, y=temp, xout=400)$y
               # therm end is half way to t400 temp
               tEnd <- mldTemp + (t400 - mldTemp)/2
               # therm temp is midpt from mld to end
               tt <- (mldTemp + tEnd)/2
               # tt <- mldTemp - .25 * (mldTemp - t400)
               ttDepth <- depthInterp(tt)
               tEndDepth <- depthInterp(tEnd)
           }
    )
    if(plot) {
        to300 <- depth <= 300
        plot(x=temp[to300], y=-depth[to300],
             main = paste0('MixLayer: ', round(mld,0), 'm ', round(mldTemp, 1), 'C',
                           '\nTherm: ', round(ttDepth, 0), 'm ', round(tt, 1), 'C'),
             xlab='Temp (C)', ylab='Depth (m)')

        lines(x=temp[to300], y=-depth[to300])
        lines(x=c(mldTemp, tEnd), y=c(-mld, -tEndDepth), col='darkgray', lwd=4)
        points(x=c(mldTemp, tt), y=c(-mld, -ttDepth), col='blue', cex=1.5, pch=15)
    }
    list(mldDepth = mld, mldTemp = mldTemp,
         ttDepth = ttDepth, ttTemp = tt)
}
```

```{r}
#calcThermoVars(hycomData[[1]]$matchDepth, hycomData[[1]]$water_temp, plot=TRUE)

thermData0 <- bind_rows(lapply(hycomData0, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))

thermData1 <- bind_rows(lapply(hycomData1, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))

thermData2 <- bind_rows(lapply(hycomData2, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))

thermData3 <- bind_rows(lapply(hycomData3, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))

thermData4 <- bind_rows(lapply(hycomData4, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))

thermData5 <- bind_rows(lapply(hycomData5, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))

thermData6 <- bind_rows(lapply(hycomData6, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))

thermData <- rbind(thermData0, thermData1, thermData2, thermData3, thermData4, thermData5, thermData6)

tracks_SS_08 <- cbind(tracks_SS_08, thermData)

# New tracks_SS_08 variable contains Hycom variables now
View(tracks_SS_08)
```

Get avg daily environmental variables
```{r}
means_08 <- tracks_SS_08 %>%
  group_by(as.Date(dateTime))%>%
  summarize(sst_mean = mean(sst_mean),
            dist_slope = mean(dist_slope),
            depth = mean(depth),
            slope = mean(slope),
            dist_shore = mean(dist_shore),
            chlorophyll_mean = mean(chlorophyll_mean),
            ssh_mean = mean(ssh_mean),
            mldDepth = mean(mldDepth),
            mldTemp = mean(mldTemp),
            ttDepth = mean(ttDepth),
            ttTemp = mean(ttTemp),
            temp400 = mean(temp400),
            sal400 = mean(sal400),
            BWpresence = sum(BWpresence))
```

Scatterplots

```{r}
# SST vs TOL 2000
ggplot(data = tracks_SS_08, aes(x = sst_mean, y = `TOL_2000`)) +
  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)

# SST over time
ggplot(tracks_SS_08, aes(x = dateTime, y = sst_mean)) +
  geom_line() +
  geom_point(aes(color = species)) +
  scale_color_discrete(na.translate=FALSE)
  #geom_smooth(method = "loess")

# Depth vs TOL 125
ggplot(data = tracks_SS_08, aes(x = depth, y = `TOL_125`)) +
  geom_line() +
  geom_point(aes(color = species)) +
  scale_color_discrete(na.translate=FALSE)

#Mean depth over time
ggplot(means_08, aes(x = `as.Date(dateTime)`, y = depth*(-1))) +
  geom_line()

# Chlorophyll vs TOL 125
ggplot(data = tracks_SS_08, aes(x = chlorophyll_mean, y = `TOL_125`)) +
  geom_point(aes(color = species))

# CHl over time
ggplot(tracks_SS_08, aes(x = dateTime, y = chlorophyll_mean)) +
  geom_line() +
  geom_point(aes(color = species)) +
  scale_color_discrete(na.translate=FALSE)
  #geom_smooth(method = "loess")

# Mixed Layer Depth vs TOL125 with species
ggplot(data = tracks_SS_08, aes(x = mldDepth, y = `TOL_125`)) +
  geom_point(aes(color = species))

# Mixed Layer Depth over time
ggplot(tracks_SS_08, aes(x = dateTime, y = mldDepth*(-1))) +
  geom_line() +
  geom_point(aes(color = species)) +
  scale_color_discrete(na.translate=FALSE)

ggplot(data = tracks_SS_08, aes(x = slope, y = `TOL_5000`)) +
  geom_point(aes(color = species))
```
Facet Grids for Env. Variables
```{r}
#For surface/depth variables USING DAILY MEAN VARIABLES
# See above for means_08 derivation
track08Long <- means_08 %>%
  pivot_longer(cols = sst_mean:ssh_mean,
               names_to="parameter",
               values_to="value") %>%
  filter(parameter %in% c("sst_mean", "dist_shore", "chlorophyll_mean", "ssh_mean")) %>%
  rename(Date = `as.Date(dateTime)`)

p <- ggplot(data = track08Long, aes(x = Date, y = value)) +
  geom_line()
 # geom_point(aes(size = BWpresence))   ### How best to represent number of BW detections per day?

p + facet_grid(parameter ~ ., scales = "free_y")

# For MLD/Thermocline variables 
track08Long2 <- tracks_SS_08 %>%
  pivot_longer(cols = mldDepth:sal400,
               names_to="parameter",
               values_to="value") %>%
  filter(parameter %in% c("mldDepth", "mldTemp", "ttDepth", "ttTemp", "temp400", "sal400")) %>%
    rename(Date = dateTime) %>%
    subset(species!="?BW")

q <- ggplot(data = track08Long2, aes(x = Date, y = value)) +
  geom_line() + geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
q + facet_grid(parameter ~ ., scales = "free_y")

# For temp400 and salinity400 variables
env.labs <- c(`sal400` = "Salinity at 400m", 
              `temp400` = "Temperature at 400m")

track08Long3 <- tracks_SS_08 %>%
  pivot_longer(cols = temp400:sal400,
               names_to="parameter",
               values_to="value") %>%
  filter(parameter %in% c("temp400", "sal400")) %>%
  rename(`Date (2018)` = dateTime)

q <- ggplot(data = track08Long3, aes(x = `Date (2018)`, y = value)) +
  geom_line() + 
  geom_point(aes(color = species)) + 
  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
q + facet_grid(parameter ~ ., scales = "free_y", labeller = as_labeller(env.labs))
```

Co-occurrence plots of drift 8 and vessels 
```{r}
#### 09/04/18 co-occurrence of vessel and drift within 10 km
track08_09.04 <- tracks_SS_08 %>%
  filter(dateTime >= "2018-09-04 00:00:00" & dateTime <= "2018-09-05 00:00:00")

ggplot(data = track08_09.04, aes(x = dateTime, y = `TOL_125`)) +
  geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
  geom_text(aes(as.POSIXct("2018-09-04 14:13:00"), 88, label = "Enter vessel", hjust = 1.0)) +
  geom_vline(xintercept = as.POSIXct("2018-09-04 15:17:00"), color = "yellow", linewidth = 2) +
  geom_text(aes(as.POSIXct("2018-09-04 15:13:00"), 87, label = "Exit vessel", hjust = -0.1)) +
  geom_line() + 
  geom_point(data = track08_09.04, aes(color = species, size = 2)) +
  scale_color_discrete(na.translate=FALSE) +
  scale_size_continuous(guide = "none") +
  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-04-18\n(Within 7 km)") +
  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

#### Zooms in further into listening area when vessel transits through.
track08_09.04a <- tracks_SS_08 %>%
  filter(dateTime >= "2018-09-04 12:00:00" & dateTime <= "2018-09-04 16:00:00")

ggplot(data = track08_09.04a, aes(x = dateTime, y = `TOL_125`)) +
  geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
  geom_text(aes(as.POSIXct("2018-09-04 14:13:00"), 86, label = "Enter vessel", hjust = 0.9)) +
  geom_vline(xintercept = as.POSIXct("2018-09-04 15:17:00"), color = "yellow", linewidth = 2) +
  geom_text(aes(as.POSIXct("2018-09-04 15:13:00"), 85, label = "Exit vessel", hjust = 0.3)) +
  geom_line() +
  geom_point(data = track08_09.04a, aes(color = species, size = 2)) +
  scale_color_discrete(na.translate=FALSE) +
  scale_size_continuous(guide = "none") +
  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-04-18\n(Within 7 km)") +
  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

#### 09/05/18 co-occurrence of vessel and drift within 10 km. NO BW detections
track08_09.05 <- tracks_SS_08 %>%
  filter(dateTime >= "2018-09-04 21:00:00" & dateTime <= "2018-09-06 00:00:00")

ggplot(data = track08_09.05, aes(x = dateTime, y = `TOL_125`)) +
  geom_vline(xintercept = as.POSIXct("2018-09-05 01:33:00"), color = "yellow", linewidth = 1) +
  geom_text(aes(as.POSIXct("2018-09-05 01:33:00"), 86, label = "Enter vessel", hjust = 1)) +
  geom_vline(xintercept = as.POSIXct("2018-09-05 02:08:00"), color = "yellow", linewidth = 1) +
  geom_text(aes(as.POSIXct("2018-09-05 02:08:00"), 85, label = "Exit vessel", hjust = -0.1)) +
  geom_line() + 
  scale_color_discrete(na.translate=FALSE) +
  scale_size_continuous(guide = "none") +
  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-05-18\n(Within 5 km)") +
  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

#### 09/10/18 co-occurrence of vessel and drift within 10 km.
track08_09.10 <- tracks_SS_08 %>%
  filter(dateTime >= "2018-09-10 00:00:00" & dateTime <= "2018-09-11 00:00:00")

ggplot(data = track08_09.10, aes(x = dateTime, y = `TOL_125`)) +
  geom_vline(xintercept = as.POSIXct("2018-09-10 12:58:00"), color = "yellow", linewidth = 1) +
  geom_text(aes(as.POSIXct("2018-09-10 12:58:00"), 86, label = "Enter vessel", hjust = 1)) +
  geom_vline(xintercept = as.POSIXct("2018-09-10 14:01:00"), color = "yellow", linewidth = 1) +
  geom_text(aes(as.POSIXct("2018-09-10 14:01:00"), 85, label = "Exit vessel", hjust = -0.1)) +
  geom_line() +
    geom_point(data = track08_09.10, aes(color = species, size = 2)) +
  scale_color_discrete(na.translate=FALSE) +
  scale_size_continuous(guide = "none") +
  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-10-18\n(Within 2 km)") +
  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

#### 09/22/18 co-occurrence of vessel and drift within 10 km
track08_09.22 <- tracks_SS_08 %>%
  filter(dateTime >= "2018-09-22 00:00:00" & dateTime <= "2018-09-23 00:00:00")

ggplot(data = track08_09.22, aes(x = dateTime, y = `TOL_125`)) +
  geom_vline(xintercept = as.POSIXct("2018-09-22 19:13:00"), color = "yellow", linewidth = 1) +
  geom_text(aes(as.POSIXct("2018-09-22 19:13:00"), 86, label = "Enter vessel", hjust = 1)) +
  geom_vline(xintercept = as.POSIXct("2018-09-22 19:34:00"), color = "yellow", linewidth = 1) +
  geom_text(aes(as.POSIXct("2018-09-22 19:34:00"), 85, label = "Exit vessel", hjust = -0.1)) +
  geom_line() + 
  geom_point(data = track08_09.22, aes(color = species, size = 2)) +
  scale_color_discrete(na.translate=FALSE) +
  scale_size_continuous(guide = "none") +
  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-22-18\n(Within 9 km)") +
  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

#### 09/23/18 co-occurrence of vessel and drift within 10 km. NO whale detections in this time period.
track08_09.23 <- tracks_SS_08 %>%
  filter(dateTime >= "2018-09-23 00:00:00" & dateTime <= "2018-09-24 00:00:00")

ggplot(data = track08_09.23, aes(x = dateTime, y = `TOL_125`)) +
  geom_vline(xintercept = as.POSIXct("2018-09-23 03:55:00"), color = "yellow", linewidth = 1) +
  geom_text(aes(as.POSIXct("2018-09-23 03:55:00"), 86, label = "Enter vessel", hjust = 1)) +
  geom_vline(xintercept = as.POSIXct("2018-09-23 04:23:00"), color = "yellow", linewidth = 1) +
  geom_text(aes(as.POSIXct("2018-09-23 04:23:00"), 85, label = "Exit vessel", hjust = -0.1)) +
  geom_line() +
  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-23-18\n(Within 6 km)") +
  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
```

BW Occurrences
```{r}
# Filter Drift 8 by BW presence and absence
BW1 <- tracks_SS_08 %>%
  filter(BWpresence == 1)
BW0 <- tracks_SS_08 %>%
  filter(BWpresence == 0)

ggplot(data = tracks_SS_08, mapping = aes(TOL_125)) + geom_histogram(binwidth = 0.1, color = "black")

ggplot() + 
  geom_histogram(data = BW0, aes(TOL_5000), binwidth = 0.5, fill = "black", color = "white") +
  geom_histogram(data = BW1, aes(TOL_5000), binwidth = 0.5, fill = "yellow", color = "black")

# Percent of time BWs are detected
print(paste(nrow(BW1) / nrow(tracks_SS_08) * 100,"%"))

# Most abundant BW species when whale is detected (remove NA values)
tracks_SS_08 %>%
  filter(!is.na(species)) %>%
  ggplot() + geom_bar(aes(species))

# Sound levels BW are exposed to by species
ZC <- tracks_SS_08 %>%
  filter(species == "ZC")

BW43 <- tracks_SS_08 %>%
  filter(species == "BW43")

print(max(ZC$TOL_125))
print(max(qBW$TOL_125))
print(max(BW43$TOL_125))

ggplot() + geom_histogram(data = ZC, mapping = aes(TOL_125), binwidth = 0.2, fill = "brown", color = "white") +
  geom_histogram(data = BW43, mapping = aes(TOL_125), binwidth = 0.2, fill = "yellow", color = "white") +
  geom_histogram(data = qBW, mapping = aes(TOL_125), binwidth = 0.2, fill = "blue", color = "white") +
  theme_classic()
```

Density Plots - Species at TOL sound levels. REMOVED ?BW HERE
```{r}
# Broadband
tracks_SS_08 %>%
  filter(!is.na(species)) %>%
  subset(species!="?BW") %>%
  ggplot(aes(`BB_20.24000`, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle("Broadband") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

# TOL 63
tracks_SS_08 %>%
  filter(!is.na(species)) %>%
  subset(species!="?BW") %>%
  ggplot(aes(TOL_63, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle("63 Hz") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

# TOL 125
tracks_SS_08 %>%
  filter(!is.na(species)) %>%
  subset(species!="?BW") %>%
  ggplot(aes(TOL_125, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle("125 Hz") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

# TOL 2000
tracks_SS_08 %>%
  filter(!is.na(species)) %>%
  subset(species!="?BW") %>%
  ggplot(aes(TOL_2000, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle("2000 Hz") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

# TOL 5000
tracks_SS_08 %>%
  filter(!is.na(species)) %>%
  subset(species!="?BW") %>%
  ggplot(aes(TOL_5000, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle("5000 Hz") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

# TOL 20,000
tracks_SS_08 %>%
  filter(!is.na(species)) %>%
  subset(species!="?BW") %>%
  ggplot(aes(TOL_20000, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle("20,000 Hz") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
```

Histograms with BW presence/absence and environmental variables
```{r}
# SST
ggplot(data = tracks_SS_08, aes(sst_mean, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.1, color = "black")

# Depth
ggplot(data = tracks_SS_08, aes(depth, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 10, color = "black")

# Bottom slope
ggplot(data = tracks_SS_08, aes(slope, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.005, color = "black")

# SSH
ggplot(data = tracks_SS_08, aes(ssh_mean, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.005, color = "black")

# Chlorophyll
ggplot(data = tracks_SS_08, aes(chlorophyll_mean, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.005, color = "black")

# Mixed layer depth Depth
ggplot(data = tracks_SS_08, aes(mldDepth, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.25, color = "black")

# Mixed layer depth Temperature
ggplot(data = tracks_SS_08, aes(mldTemp, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.1, color = "black")

# Thermocline Depth
ggplot(data = tracks_SS_08, aes(ttDepth, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.25, color = "black")

#Thermocline temperature
ggplot(data = tracks_SS_08, aes(ttTemp, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.05, color = "black")

# Temperature at 400m
ggplot(data = tracks_SS_08, aes(temp400, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.01, color = "black")

# Salinity at 400 m
ggplot(data = tracks_SS_08, aes(sal400, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.001, color = "black")
```

### Plot maps for Powerpoint images

Map of whole study area
```{r}
# Establish bounding box
bbox <- st_bbox(c(xmin = -110, 
            xmax = -128, 
            ymax = 44, 
            ymin = 30), crs = st_crs(4326))

#Establish basemap
allTrackbase <- get_tiles(bbox, provider = "Esri.WorldGrayCanvas", zoom = 5)

# Establish inset map
us_region <- st_bbox(c(xmin = -90,
            xmax = -120, 
            ymax = 48, 
            ymin = 25), crs = st_crs(4326))
usa <- get_tiles(us_region, provider = "Esri.WorldGrayCanvas", zoom = 4)

sg <- bb_poly(bb(tracks_sf))
inset <- tm_shape(usa) + tm_rgb() + tm_shape(sg) + tm_borders(col = "red")

#Map all tracks
main <- tm_shape(allTrackbase) + tm_rgb() +
  tm_shape(trackssf) +
  tm_grid(alpha = 0.6, col = "grey") +
  tm_dots(col = "darkblue", size = 0.001, legend.show = TRUE) +
  tm_scale_bar(width = 0.3, text.size = 3, position = c(0.2, 0.03)) +
  tm_compass(type = "arrow", position = c(0.05, 0.05), color.light = "white") +
  tm_add_legend('symbol',
                type = "line",
                col = "darkblue",
                labels = "Drifting buoy tracks",
                size = 1,
                lwd = 2)

vp <- viewport(x=0.66, y=0.8, width = 0.26, height=0.4)
main
print(inset,vp=vp)

# Map all tracks - ggplot
basemap2 <- as.data.frame(allTrackbase, xy = TRUE)
#ggplot() +
#  geom_tile(data = basemap, aes(x=x, y=y, fill = "basemap")) +
#  geom_sf(data = trackssf, mapping = aes(), color = "red")

# Map tracks 08 and 10 only
tm_shape(trackBase) + tm_rgb() + 
  tm_shape(trk_TOL_08) + tm_dots(col = "red", 
                                 legend.show = TRUE) +
  tm_shape(trk_TOL_10) + tm_dots(col = "green")

# Bounding box of drift tracks on map
lon2 <- -129.04666
lon1 <- -115.78247
lat1 <- 28.21790
lat2 <- 45.09226

basemap3 <- ne_countries(scale = "medium", type = "map_units", returnclass = "sf")

# Drift tracks
ggplot(data = basemap3) +
  geom_sf(fill = "antiquewhite") +
  coord_sf(xlim = c(lon2 - 1, lon1 + 1), ylim = c(lat1 - 1, lat2 + 1), expand = FALSE) +
  theme(panel.grid.major = element_blank()) +
  geom_path(data = tracks_sf, mapping = aes(x=long,y=lat, group = factor(station)), color = "deeppink") +
  theme(panel.background = element_rect(fill = "aliceblue"), 
        axis.text = element_blank(), axis.ticks = element_blank(), 
        axis.title = element_blank())
```

Map of soundscapes
```{r}
# CCES_08 + CCES_10 Broadband
tm_shape(trackBase) + tm_rgb() + 
  tm_shape(trk_TOL_08) + 
  tm_symbols(col = "BB_20-24000", 
             border.lwd = NA, 
             size = 0.4, 
             sizes.legend = 0.7, 
             n = 10) +
  tm_shape(trk_TOL_10) + 
  tm_symbols(col = "BB_20-24000",
             border.lwd = NA, 
             size = 0.4,
             n = 10) +
  tm_layout(legend.width = 0.5, 
            legend.position = c("right","top"),
            legend.bg.color = "lightblue")

#CCES_08 Third Octave Level
tm_shape(trackBase) + tm_rgb() + 
  tm_shape(trk_TOL_08) + 
  tm_symbols(col = "TOL_63", 
             border.lwd = NA, 
             size = 0.5, 
             n = 10) + 
  tm_layout(legend.width = 0.5, 
            legend.position = c("right","top")) + 
  tm_legend()

ggplot() + 
  geom_sf(data = trk_TOL_08, aes(color = TOL_125, size = TOL_125), show.legend = FALSE) +
  geom_sf(data = trk_TOL_10, aes(color = TOL_125, size = TOL_125)) +
  scale_x_continuous(limits = c(-130,-122)) + 
  scale_y_continuous(limits = c(35,39)) +
  scale_color_gradient(low = "navy", high = "yellow")+
  scale_size(range = c(1,12)) +
  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
  theme(panel.border = element_rect(fill = NA),
        axis.text = element_blank(),
        axis.ticks = element_blank(), 
        axis.title = element_blank(),
        panel.background = element_blank(),
        legend.position = "none")
```

Map of whale detections
```{r}
# Whale detections
ggplot() +
  geom_point(data = whales8, mapping = aes(x=long, y=lat), shape = 21, color = "black", fill = "green4", size = 4) +
  geom_point(data = whales10, mapping = aes(x=long, y=lat), shape = 21, color = "black", fill = "green4", size = 4) +
  scale_x_continuous(limits = c(-130,-122)) + 
  scale_y_continuous(limits = c(35,39)) +
  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
  theme(panel.border = element_rect(fill = NA),
        axis.text = element_blank(),
        axis.ticks = element_blank(), 
        axis.title = element_blank(),
        panel.background = element_blank()
        )
#ggsave("BWdetect.png")

```

Map of bathymetry
```{r}
# Bathymetry
bathym <- ggplot() +
    geom_raster(data = bathy_df, 
              mapping = aes(x=x, y=y,
                            fill = `gebco_bathy`)) +
  annotation_scale(location = "bl", width_hint = 0.4) +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(fill = NA))

# Alternative Bathymetry (code from Ocean Tracking Network #https://www.youtube.com/watch?v=xT2F-Y9bKLk)
bgo <- getNOAA.bathy(lon1 = -122, lon2 = -130,
                     lat1 = 35, lat2= 39, resolution = 1)  # plot bounding box around drift 8 and 10
autoplot(bgo, geom = c("raster", "contour"), coast = TRUE, show.legend = FALSE) + 
  scale_fill_etopo() + #creates ggplot object
  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank(),
        panel.background = element_blank())

# Soundscape + bathymetry
ggplot() +
  geom_raster(data = bathy_df, 
              mapping = aes(x=x, y=y,
                            fill = `gebco_bathy`), show.legend = FALSE) +
  #geom_contour(data = bathy_df, mapping = aes(x=x, y=y, z=gebco_bathy, linewidth = 0.1), color = "blue", linewidth = 0.01,) +
  geom_sf(data = trk_TOL_08, aes(color = TOL_125, size = TOL_125), show.legend = FALSE) +
  geom_sf(data = whales08_sf, aes(), color = "black") +
  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
  scale_color_gradient(low = "beige", high = "hotpink") +
  scale_fill_etopo() +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(fill = NA))
#ggsave("soundscape.png")
```

Map of AIS
```{r}
# Plot + crop to drift area
#basemap <- ne_countries(scale = "medium", returnclass = "sf")
#ggplot() + 
#  geom_sf(aes(), basemap) +
#  geom_path(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName), #show.legend = F) + 
  #geom_point(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName)) +
#  scale_x_continuous(limits = c(min(drift$long-0.7), max(drift$long + 0.7))) + 
#  scale_y_continuous(limits = c(min(drift$lat-0.7), max(drift$lat + 0.7)))

# AIS data
ggplot() + 
  geom_path(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName, linewidth = 1), show.legend = F) +
  scale_x_continuous(limits = c(-131,-121)) + 
  scale_y_continuous(limits = c(34,40)) +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(fill = NA))
#ggsave("AIS.png")
```

### 2023-04-05 01_readin.R
Create 01_readin.R

@@ -1,0 +1,246 @@
+
+# Read in, load, clean up, and wrangle raw data
+
+# Look into https://bookdown.org/yihui/rmarkdown-cookbook/managing-projects.html for info on 
+# managing projects with multiple Rmds and scripts
+
+###########################
+# Required packages
+###########################
+
+library(tidyverse)
+library(lubridate)
+library(sf)
+library(rerddap)
+library(terra)
+library(marmap)
+library(rnaturalearth)
+library(mapproj)
+library(raster)
+library("rerddapXtracto")
+library(PAMpal)
+library(maptiles)
+library(tmap)
+library(spData)
+library(tmaptools)
+library(grid)
+library(ggspatial)
+library(here)
+library(leaflet)
+
+##########################
+# Read in soundscape data, NEED TO MAKE THIS A FUNCTION
+# look at http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops for help on automation.
+##########################
+here()
+
+# Set system time zone to UTC
+Sys.setenv(TZ='UTC')
+
+worldmap <- ne_countries(scale = 'medium', type = 'map_units',
+                         returnclass = 'sf')
+calif <- st_crop(worldmap, 
+                 xmin = 33.7,
+                 xmax = 40,
+                 ymin = -131,
+                 ymax = -122)
+calif <- st_transform(calif, 4326)
+
+tmap_options(basemaps=c(Terrain = "Esri.WorldTerrain",
+                        Imagery = "Esri.WorldImagery", 
+                        OceanBasemap = "Esri.OceanBasemap", 
+                        Topo="OpenTopoMap",
+                        Ortho="GeoportailFrance.orthos"))
+
+
+# Round to nearest 20 minutes, create dateTime data type
+# Drift 8: Broadband
+BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
+BB_08 <- BB_08_csv %>% 
+  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
+BB_08 <- BB_08 %>%
+  mutate(dateTime = round_date(ymd_hms(BB_08$dateTime),"20 minutes"))
+
+# Drift 8: Octave Level
[…]

### 2023-04-19 01_readin.R
Update 01_readin.R
Added for loops, added code to process each drift

@@ -4,35 +4,20 @@
 # Look into https://bookdown.org/yihui/rmarkdown-cookbook/managing-projects.html for info on 
 # managing projects with multiple Rmds and scripts
 
-###########################
-# Required packages
-###########################
+# libraries ##########
 
 library(tidyverse)
 library(lubridate)
 library(sf)
-library(rerddap)
-library(terra)
-library(marmap)
 library(rnaturalearth)
-library(mapproj)
-library(raster)
-library("rerddapXtracto")
-library(PAMpal)
-library(maptiles)
 library(tmap)
-library(spData)
-library(tmaptools)
-library(grid)
-library(ggspatial)
 library(here)
-library(leaflet)
 
-##########################
-# Read in soundscape data, NEED TO MAKE THIS A FUNCTION
-# look at http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops for help on automation.
-##########################
+# set up #####
+
 here()
+#Remove global environment if needed
+#rm(list=ls()) 
 
 # Set system time zone to UTC
 Sys.setenv(TZ='UTC')
@@ -53,191 +38,247 @@ tmap_options(basemaps=c(Terrain = "Esri.WorldTerrain",
                         Ortho="GeoportailFrance.orthos"))
 
 
-# Round to nearest 20 minutes, create dateTime data type
-# Drift 8: Broadband
-BB_08_csv <- read_csv("Data/CCES_08_1Hz_1s_BB_2min.csv", show_col_types = FALSE)
-BB_08 <- BB_08_csv %>% 
-  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
-BB_08 <- BB_08 %>%
-  mutate(dateTime = round_date(ymd_hms(BB_08$dateTime),"20 minutes"))
+##########################
+# Read in soundscape data, NEED TO MAKE THIS A FUNCTION
+# look at http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops for help on automation.
+##########################
 
-# Drift 8: Octave Level
-OL_08_csv <- read_csv("Data/CCES_08_1Hz_1s_OL_2min.csv",show_col_types = FALSE)
-OL_08 <- OL_08_csv %>% 
-  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`)
-OL_08 <- OL_08 %>%
[…]

### 2023-04-19 02_AIS.Rmd
Added for loops to process all AIS files

---
title: "AIS"
author: "Marina Bozinovic"
date: "2023-04-11"
output: html_document
---
```{r}
#Required Packages

library(tidyverse)
library(sf)
```

```{r}
#########
# AIS
#########

# Need to add more AIS dates in data folder
AISlist <- list.files(path = paste0("data/"), pattern = "AIS_", recursive = TRUE)

for (v in AISlist) {
  Adata <- read_csv(paste0("data/", v), show_col_types = FALSE) %>%
    filter(Length >= 100) %>%                  # Filter out vessels over 100m in length
    dplyr::select(-SOG,-COG, -CallSign, -IMO, -Cargo, -Draft, -Status) %>%  #Remove unnecessary columns
    mutate(variable = paste0(substr(v, 16, 17), substr(v, 6, 7)))

  assign(paste0("AIS_", substr(v, 5, 10)), Adata)
}

AIS_sf <- AIS %>%
  st_as_sf(coords = c("LON", "LAT"), crs=4326, remove=F)
```

```{r}
##################################
# Create 10 km buffer around buoys (taken from M. McKenna's code)
# Should be malleable for each buoy's number to easily plug in.
##################################
trk8 <- tracks %>% 
  dplyr::filter(station == "8") %>%
  st_as_sf(coords= c("Longitude", "Latitude"), crs=4326, remove=F)

pois <- st_transform(trk8, crs = 4326)    # Specify a crs
pts_buff <- st_buffer(pois, dist = 10000)  # Buffer all points in trk to 10 km (10000m)
buff <- st_transform(pts_buff, crs = 4326)  # Specify crs for new buffer

### Crop full AIS to 10km buffer around drift 8 ###
AIS_crop <- st_crop(x = AIS_sf, y = buff)

# Intersect buffer with AIS. Starting this on first 10 rows to see how slowly it goes. 
## FIX ME
#inters <- st_intersection(buff, AIS_crop[1:140,])

#Another way to view with TRUE/FALSE if AIS points fall within buffer
intersect <- st_intersects(buff, AIS_crop)
lgth <- lengths(st_intersects(buff, AIS_crop)) > 0

# Speeding up st_intersection (from https://github.com/r-spatial/sf/issues/801)

st_intersection_faster <- function(x,y,...){
#faster replacement for st_intersection(x, y,...)

  y_subset <-
    st_intersects(x, y) %>%
    unlist() %>%
    unique() %>%
    sort() %>%
    {y[.,]}

  st_intersection(x, y_subset,...)
}
inters <- st_intersection_faster(buff, AIS_crop[1:140,])
inters2 <- st_intersection_faster(buff, AIS_crop[141:200,])
inters3 <- st_intersection_faster(buff, AIS_crop[201:300,])

```
```{r}
# Plot showing 10km buffer around drift 8
#tmap_mode("plot")
#tm_shape(buff) + tm_symbols(col = "blue", border.col = NA) +
#  tm_shape(trk) + tm_dots(col = "red")
```

```{r}
# Plot
ggplot(tracks_SS_08, aes(x = long, y = lat.x)) + geom_point() 
 + geom_sf(data = bbox8, aes(col = "red"))

# Plot showing AIS bounding box around drift 8 tracks
ggplot() + geom_sf(data = AIS_crop,color = "red")

ggplot() +
  geom_sf(data = calif) + coord_sf(default_crs = sf::st_crs(4326)) + theme_bw()
 # geom_sf(data = AIS_sf)

### 2023-04-19 03_exploratory.Rmd
Create 03_exploratory.Rmd

--- 2023-04-05 02_test.Rmd
@@
 library(raster)
-library("rerddapXtracto")
+#library("rerddapXtracto")
 library(PAMpal)
@@
 theme_set(theme_bw())
-```
-
-### Exploratory analysis
-
-#### Not important: tracks as sf and basemaps
-
-```{r}
-library(tmap); library(maptiles); library(sf)
 tmap_mode("plot")
-trackssf <- st_as_sf(tracks, coords = c("long", "lat"), crs=4326)
-#tracks08sf <- st_as_sf(track_BB_OL_08, coords = c("long", "lat"), crs=4326)  # outdated
-#tracks10sf <- st_as_sf(track_BB_OL_10, coords = c("long", "lat"), crs=4326)  # outdated
-trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("long", "lat.x"), crs=4326)
-trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("long", "lat.x"), crs=4326)
+```
+
+Create sf objects and basemaps for maps
+```{r}
+whales_8_10sf <- BWsf %>%                            
+dplyr::filter(station == "8" | station == "10")
+
+trackssf <- st_as_sf(tracks, coords = c("Longitude", "Latitude"), crs=4326)
+trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326)
+trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326)
 trackBase <- get_tiles(whales_8_10sf, provider = "Esri.WorldShadedRelief")
@@
 ```
-
-### Sea Surface Temp with rerddap, Drift 8 only
-```{r}
-# Multi-scale Ultra-high Resolution (MUR) SST Analysis fv04.1, Global, 0.01°, 2002-present, Daily
-SSTInfo <- info('jplMURSST41') 
-murSST <- griddap(SSTInfo, latitude = c(35, 39), longitude = c(-122, -130), 
-                  time = c("2018-08-16", "2018-10-01"), 
-                  fields = 'analysed_sst') # drift8 dates only
-SST <- murSST$data
-
-# Raster of SST
-ggplot() + 
-  geom_tile(data = SST, aes(x = longitude, y = latitude, fill = analysed_sst)) +
-  scale_fill_gradient(low = "darkblue", high = "yellow") +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(axis.text = element_blank(),
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA),
-        legend.position = "none")
-```
-
-### SST from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
-```{r}
-xcoord <- tracks_08$long
-ycoord <- tracks_08$lat
-tcoord <- tracks_08$dateTime
-xlen <- 0.2
-ylen <- 0.2
-parameter <- "analysed_sst"
-dataset <- 'jplMURSST41'
-#dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")
-
-#swchl <- rxtracto(dataInfo, 
-#                  parameter=parameter, 
-#                  xcoord=xcoord, ycoord=ycoord, 
-#                  tcoord=tcoord, xlen=xlen, ylen=ylen)
-
-#plotTrack(swchl, xcoord, ycoord, tcoord, plotColor = 'thermal')
-```
-
-### SST from PAMPal
-```{r}
-# Downloads data from servers
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "analysedSST") %>%
+### Exploratory analysis
+
+SST from PAMPal
+```{r}
+# Downloads data from servers, uncomment and only run once
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "analysedSST")
 
@@
   dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)   # rename new column to "sst_mean" and remove unnecessary columns
-tracks_SS_08
-```
-
-### Distance to continental slope
-
+```
+
+Distance to continental slope
 ```{r}
@@
 bathy <- rast("data/gebco_bathy.tif")
-agg <- terra::aggregate(bathy, 20)
+agg <- terra::aggregate(bathy, 5)
 plot(agg)
@@
 # Rasterize track with agg as the template
-#track_rast <- rasterize(tracks08sf, agg, field = "BB_20-24000")
+
 #TOL_rast <- rasterize(trk_TOL_08, agg, field = "TOL_125")
-sf08 <- st_as_sf(tracks_SS_08, coords = c("long", "lat.x"), crs=4326, remove=F)
-sf10 <- st_as_sf(tracks_SS_10, coords = c("long", "lat.x"), crs=4326, remove=F)
+sf08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
+sf10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
 trkraster08 <- rasterize(vect(sf08), agg)
@@
 
-ggplot() + 
-  geom_raster(data = bathy_df, 
-              mapping = aes(x=x, y=y, 
-                            alpha = 0.8,
-                            fill = `gebco_bathy`))+
-  geom_sf(data = cont, col="black") + 
-  geom_sf(data = coast, col = "red") +
-  geom_sf(data = cont_slope, col = "purple") +
-  geom_sf(data = whales_8_10sf, mapping = aes()) +
-  scale_color_gradient(low="green", high="red") 
+#ggplot() + 
+#  geom_raster(data = bathy_df, 
+#              mapping = aes(x=x, y=y, 
+#                            alpha = 0.8,
+#                            fill = `gebco_bathy`))+
+#  geom_sf(data = cont, col="black") + 
+#  geom_sf(data = coast, col = "red") +
+#  geom_sf(data = cont_slope, col = "purple") +
+#  geom_sf(data = whales_8_10sf, mapping = aes()) +
+#  scale_color_gradient(low="green", high="red") 
 ```
@@
   dplyr::select(-ID) %>%
-  rename("dist_slope" = gebco_bathy)
+  rename("dist2slope" = gebco_bathy) %>%
+  st_set_geometry(NULL)                         # Drop geometry to make df for future cbinds
 
@@
   dplyr::select(-ID) %>%
-  rename("dist_slope" = gebco_bathy)
-```
-
-#### Plot distance to continental slope
-
-```{r}
-# By color
-#ggplot() +
-#  geom_sf(data = pts08, 
-#          mapping = aes(col = `Distance to Continental Slope`)) +
-#  geom_sf(data = pts10, 
-#          mapping = aes(col = `Distance to Continental Slope`)) +
-#  geom_sf(data = cont_slope)
-
-# Plot distance by OL/TOL 
-#ggplot(data = pts08, aes(x = `Distance to Continental Slope`, y = `OL_125`)) + geom_point()
-#ggplot(data = TOL08, aes(x = `Distance to Continental Slope`, y = `TOL_125`)) + geom_point() + geom_smooth()
-   # what are the units for distance? Meters
-
-ggplot() +
-  geom_raster(data = bathy_df, 
-              mapping = aes(x=x, y=y,
-                            fill = `gebco_bathy`)) +
-  geom_sf(data = tracks_SS_08, 
-          mapping = aes(col = dist_slope)) +
-  geom_sf(data = cont_slope) + scale_fill_continuous(high = "lightgreen", low = "blue")
-
-# PLOT distance by TOL with whales
-ggplot(data = tracks_SS_08, aes(x = dist_slope, y = `TOL_125`)) + 
-  geom_point() + 
-  geom_smooth(color = "green") +
-  theme_classic()
-```
-
-### Third octave levels histograms
-
-```{r}
-# Drift 8 - all TOL bands, facet wrap
-pivot <- tracks_SS_08 %>%
-  pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of third octave level intensities: Drift 8") +
-  labs(x = "Sound Intensity (dB)") +
-  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
-  scale_fill_manual(values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.7,0.8)) +
-  facet_grid(~SL)
-
-# Drift 10 - all TOL bands, facet grid
-pivot2 <- tracks_SS_10 %>%
-  pivot_longer(cols = starts_with("TOL")| starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of third octave level intensities: Drift 10") +
-  scale_x_continuous(name = "Sound Intensity (dB)", breaks=seq(50,110,20)) +
-  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
-  scale_fill_manual(values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.76)) +
-  facet_grid(~SL)
-
-# Drift 8 - TOL bands 125 and 2000 only
-pivot3 <- tracks_SS_08 %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz")
-```
-
-```{r drift-8-histogram}
-ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 8") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Beaked Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
-```
-
-```{r drift-8-histogram}
-```
-
-```{r}
-# Drift 10 - TOL bands 125 and 2000 only
-pivot4 <- tracks_SS_10 %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot4, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 10") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Beaked Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
-
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz")
-# TOL linear model plots
-#plot(lm(formula = `Distance to Continental Slope` ~ bestNumber, data = TOL_whale_08))
-
-#ggplot(TOL_whale_08) +
-#  geom_point(aes(x = TOL_125, y = nClicks), color = "blue") #+
-#  geom_point(aes(x = TOL_63, y =  nClicks), color = "hotpink") +
-#  geom_point(aes(x = TOL_5000, y =  nClicks), color = "limegreen") +
-#  geom_point(aes(x = TOL_2000, y =  nClicks), color = "yellow")
-
-```
-
-### Does hour of day affect sound levels?
-
-```{r}
-# CCES_08
-#hr_08 <- track_BB_OL_08 %>%
-#  mutate(Hr = hour(track_BB_OL_08$dateTime)) %>%
-#  group_by(Hr) %>%
-#  summarize(meanBB = mean(`BB_20-24000`)) %>%
-#  rename(`Buoy 08 Broadband` = `meanBB`)
-#ggplot(data = hr_08, mapping = aes(x = Hr, y = `Buoy 08 Broadband`)) + 
-#  geom_point() + geom_smooth()
-
-#CCES_10
-#hr_10 <- track_BB_OL_10 %>%
-#  mutate(Hr = hour(track_BB_OL_10$dateTime)) %>%
-#  group_by(Hr) %>%
-#  summarize(meanBB = mean(`BB_20-24000`)) %>%
-#  rename(`Buoy 10 Broadband` = `meanBB`)
-#ggplot(data = hr_10, mapping = aes(x = Hr, y = `Buoy 10 Broadband`)) + 
-#  geom_point() + geom_smooth()
-
-# Join and graph
-#hr_join <- left_join(hr_08, hr_10, by = "Hr")
-#hr_piv <- hr_join %>%
-#  pivot_longer(col=c(`Buoy 08 Broadband`,`Buoy 10 Broadband`), names_to = "buoys", values_to = "BB")
-
-#ggplot(data = hr_piv, mapping = aes(Hr, BB, col=buoys)) +
-#  geom_point() + geom_smooth()
-```
-
-### Are sound levels affected by species composition or number of clicks?
-
-```{r}
-# Plot
-library(tmap); library(maptiles)
-tmap_mode("plot")
-oceanBase <- get_tiles(tracks_SS_08, provider="Esri.NatGeoWorldMap", crop = TRUE)
-tm_shape(oceanBase) + tm_rgb() +
-  tm_shape(tracks_SS_08) +  tm_symbols(col="species", size = 1) +
-  tm_layout(legend.width = 0.5, 
-            legend.position = c("right","top"))
-
-# CCES 08
-#BB_whales08 <- st_join(pts08, whales_sf) %>%
-#  filter(!is.na(`species`))
-
-#summary(glm(`BB_20.24000` ~ species + `Distance to Continental Slope` + nClicks, data = BB_whales08))
-
-# CCES 10
-#BB_whales10 <- st_join(pts10, whales_sf) %>%
-#  filter(!is.na(`species`))
-
-#summary(glm(`BB_20.24000` ~ species + `Distance to Continental Slope` + nClicks, data = BB_whales10))
-
-#Raster base map
-tmap_mode("plot")
-oceanBase <- get_tiles(tracks_SS_10, provider="Esri.NatGeoWorldMap")
-tm_shape(bathy) + tm_raster() +
-  tm_shape(tracks_SS_10) + 
-  tm_symbols(style = "jenks") +
-  tm_layout(legend.width = 0.5)
-
-```
-
-#### Bathymetry
-
-```{r}
-bgo <- getNOAA.bathy(lon1 = -121, lon2 = -131,
-                     lat1 = 34, lat2= 40, resolution = 1)  # plot bounding box around drift 8 and 10
-# As raster
-bgo_rast <- marmap::as.raster(bgo)
-
-# Convert sf to dataframe for this operation
-tracks_SS_08 <- st_set_geometry(tracks_SS_08, value = NULL)
-
-# Get depth of whale detections in drift 8
-depth08 <- get.depth(bgo, tracks_SS_08[,3:2], locator = FALSE)
-
-# Get depth of tracks
-trkdepth08 <- get.depth(bgo, tracks_SS_08[,3:2], locator = FALSE)
-trkjoin <- left_join(tracks_SS_08, trkdepth08, by = c("long"= "lon")) %>%
-  mutate(depth = depth*-1)         # remove the negative sign
-
-#Plot
-ggplot() + geom_point(data = tracks_SS_08, aes(x = long, y = lat.x, size = 2)) +
-  scale_color_gradient(low = "red", high = "beige") +
-  ggtitle("Depth of Drift 8 whale detections")
-```
+  #rename("dist2slope" = gebco_bathy) %>%
+  st_set_geometry(NULL)                        # Drop geometry to make df for future cbinds
+```
+
 #### Bathymetry - PAMpal
@@
   dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-tracks_SS_08
 ```
@@
 ```{r}
-# Slope
+# Slope (Hard-coded - can fix this)
+bgo <- getNOAA.bathy(lon1 = -121, lon2 = -131,
+                     lat1 = 34, lat2= 40, resolution = 1)  # plot bounding box around drift 8 and 10
+# As raster
+bgo_rast <- marmap::as.raster(bgo)
 bgo_slope <- terrain(bgo_rast, v = "slope")
-plot(bgo_slope, axes = FALSE)
 
 # Extract slope values at points along drift 8
-slope <- raster::extract(bgo_slope, tracks_SS_08[,3:2])
+slope <- raster::extract(bgo_slope, tracks_SS_08[,4:3])
 tracks_SS_08 <- cbind(tracks_SS_08, slope)
-
-#ggplot() +
-#  geom_raster(data = bgo_slope, mapping = aes(x=x, y=y, fill=slope)) + 
-#  geom_point(data = depth08, aes(x = lon, y = lat, color = depth, size = 2))
-
-# Roughness (different from rugosity?)
-plot(terrain(bgo_rast, opt = "roughness"))    # what are the units on this?
 ```
@@
 #### Distance from coast - PAMPal
-
-```{r}
-dist2shore <- erddapToEdinfo(dataset='dist2coast_1deg',
-                             baseurl='https://pae-paha.pacioos.hawaii.edu/erddap/')
+Need to have variable choose itself!! Also, how do I get it to not download each time?
+```{r}
+#dist2shore <- erddapToEdinfo(dataset='dist2coast_1deg',
+#                             baseurl='https://pae-paha.pacioos.hawaii.edu/erddap/',
+#                             chooseVars = FALSE)
+
+tracks_SS_08 <- matchEnvData(tracks_SS_08, nc='data/Dist2Shore.nc') %>%
+  rename(., dist2shore = dist_mean) %>%
+  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
+
+#May not need this part
+# Downloads from server
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'dist2coast_1deg', fileName='dist2coast.nc', var = 'dist')
+
+# Pulls from downloaded file
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/dist2coast.nc')
+
 # Downloads nc file
 #tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=dist2shore, fileName='Dist2Shore.nc')  
-
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc='data/Dist2Shore.nc') %>%
-  rename(., dist_shore = dist_mean) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-tracks_SS_08
-```
-
-#### Chlorophyll-a: NOAA Coastwatch
-
-```{r}
-# Code chunk taken from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
-xcoord <- tracks_SS_08$long
-ycoord <- tracks_SS_08$lat
-tcoord <- tracks_SS_08$dateTime
-xlen <- 0.2
-ylen <- 0.2
-parameter <- "chlorophyll"
-dataset <- 'erdMH1chlamday'
-dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")
-
-swchl <- rxtracto(dataInfo, 
-                  parameter=parameter, 
-                  xcoord=xcoord, ycoord=ycoord,
-                  tcoord=tcoord, xlen=xlen, ylen=ylen)
-
-plotTrack(swchl, xcoord, ycoord, tcoord, plotColor = 'algae')
-
 ```
@@
 tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/chlorophyll.nc')
-tracks_SS_08
-```
-
-#### Sea surface height anomalies - NOAA Coastwatch
-```{r}
-# Code chunk taken from NOAA CoastWatch : https://coastwatch.pfeg.noaa.gov/projects/r/xyt-matchup.html
-xcoord <- tracks_SS_08$long
-ycoord <- tracks_SS_08$lat
-tcoord <- tracks_SS_08$dateTime
-xlen <- 0.2
-ylen <- 0.2
-parameter <- 'sla'
-dataset <- 'nesdisSSH1day'
-dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")
-
-swssh <- rxtracto(dataInfo, 
-                  parameter=parameter, 
-                  xcoord=xcoord, ycoord=ycoord, 
-                  tcoord=tcoord, xlen=xlen, ylen=ylen)
-
-plotTrack(swssh, xcoord, ycoord, tcoord, plotColor = 'deep')
 ```
@@
 # Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQSstress1day_LonPM180', #fileName='erdQSstress1day_LonPM180.nc', var = 'curl', depth = 0)
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQMstress1day_LonPM180', fileName='erdQMstress1day_LonPM180.nc', var = 'curl')
 
 # Pulls from downloaded file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQSstress1day_LonPM180.nc')
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQMstress1day_LonPM180.nc')
 #tracks_SS_08
 
-#windcurl <- erddapToEdinfo('erdQSstress1day_LonPM180', baseurl = 'https://coastwatch.pfeg.noaa.gov/erddap/')
+#windcurl <- erddapToEdinfo('erdQMstress1day', baseurl = 'https://coastwatch.pfeg.noaa.gov/erddap/')
 
@@
 
-#### HyCom Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
-```{r}
-# These refer to nc files that have already been downloaded from the server
-
-#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = #"hycomm0.nc")
-
-#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = #"hycomm1.nc")
-
-#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = #"hycomm2.nc")
-
-#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName #= "hycomm3.nc")
-
-#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName #= "hycomm4.nc")
-
-#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName #= "hycomm5.nc")
-
-#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName #= "hycomm6.nc")
-
-# Run the 7 nc lists that make up tracks_SS_08
+#### HyCOM Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
+Doesn't bin itself into separate dataframe based on size!
+```{r}
+# These refer to nc files that have already been downloaded from the server.
+# Uncomment and only run once! Rows are hard-coded.
+
+#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm0.nc")
+
+#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm1.nc")
+
+#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm2.nc")
+
+#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm3.nc")
+
+#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm4.nc")
+
+#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm5.nc")
+
+#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm6.nc")
+
+#hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm7.nc")
+
+#hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm8.nc")
+
+# Run the 9 nc lists that make up tracks_SS_08
 hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm0.nc")
@@
 hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm5.nc")
-hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2140,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm6.nc")
-```
-
-Use function to estimate thermocline variables using the temperature at depth data
+hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm6.nc")
+hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm7.nc")
+hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm8.nc")
+```
+
+Create function to estimate thermocline variables using the temperature at depth data
 ```{r}
@@
 
+Use above function on the smaller hycom datasets (0-6)
 ```{r}
@@
 
-thermData <- rbind(thermData0, thermData1, thermData2, thermData3, thermData4, thermData5, thermData6)
-
+thermData7 <- bind_rows(lapply(hycomData7, function(x) {
+    # do thermo calcs
+    result <- calcThermoVars(x$matchDepth, x$water_temp)
+    # get specific values
+    result$temp400 <- x$water_temp[x$matchDepth == 400]
+    result$sal400 <- x$salinity[x$matchDepth == 400]
+    result
+}))
+
+thermData8 <- bind_rows(lapply(hycomData8, function(x) {
+    # do thermo calcs
+    result <- calcThermoVars(x$matchDepth, x$water_temp)
+    # get specific values
+    result$temp400 <- x$water_temp[x$matchDepth == 400]
+    result$sal400 <- x$salinity[x$matchDepth == 400]
+    result
+}))
+
+# Bind all thermocline datasets together
+thermData <- rbind(thermData0, thermData1, thermData2, thermData3, thermData4, 
+                   thermData5, thermData6, thermData7, thermData8)
+
+# Bind main thermocline dataset to main track/soundscape dataframe
 tracks_SS_08 <- cbind(tracks_SS_08, thermData)
@@
 # New tracks_SS_08 variable contains Hycom variables now
+##### FINAL tracks_ss_08 dataframe ######
 View(tracks_SS_08)
@@
 means_08 <- tracks_SS_08 %>%
-  group_by(as.Date(dateTime))%>%
+  group_by(as.Date(UTC))%>%
   summarize(sst_mean = mean(sst_mean),
-            dist_slope = mean(dist_slope),
+            dist_slope = mean(dist2slope),
             depth = mean(depth),
             slope = mean(slope),
-            dist_shore = mean(dist_shore),
+            dist_shore = mean(dist2shore),
             chlorophyll_mean = mean(chlorophyll_mean),
@@
 
+### Exploratory Analysis ######
+
+### Third octave levels histograms
+
+```{r}
+# Drift 8 - all TOL bands, facet wrap
+pivot <- tracks_SS_08 %>%
+  pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+ggplot(data = pivot, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 3, color = "black") +
+  theme_classic() +
+  ggtitle("Distribution of third octave level intensities: Drift 8") +
+  labs(x = "Sound Intensity (dB)") +
+  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
+  scale_fill_manual(values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.7,0.8)) +
+  facet_grid(~SL)
+
+# Drift 10 - all TOL bands, facet grid
+pivot2 <- tracks_SS_10 %>%
+  pivot_longer(cols = starts_with("TOL")| starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 3, color = "black") +
+  theme_classic() +
+  ggtitle("Distribution of third octave level intensities: Drift 10") +
+  scale_x_continuous(name = "Sound Intensity (dB)", breaks=seq(50,110,20)) +
+  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
+  scale_fill_manual(values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.6,0.76)) +
+  facet_grid(~SL)
+
+# Drift 8 - TOL bands 125 and 2000 only
+pivot3 <- tracks_SS_08 %>%
+  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
+  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
+              `TOL_2000` = "TOL at 2000 Hz")
+```
+
+```{r drift-8-histogram}
+ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 1, color = "black") +
+  theme_classic() +
+  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 8") +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_manual(name = "Beaked Whale Presence", 
+                    labels = c("0" = "Absent", "1" = "Present"),
+                    values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.6,0.7)) +
+  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
+```
+
+```{r}
+# Drift 10 - TOL bands 125 and 2000 only
+pivot4 <- tracks_SS_10 %>%
+  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
+  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+ggplot(data = pivot4, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 1, color = "black") +
+  theme_classic() +
+  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 10") +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_manual(name = "Beaked Whale Presence", 
+                    labels = c("0" = "Absent", "1" = "Present"),
+                    values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.6,0.7)) +
+  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
+
+TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
+              `TOL_2000` = "TOL at 2000 Hz")
+```
+
 Scatterplots
@@
 # SST over time
-ggplot(tracks_SS_08, aes(x = dateTime, y = sst_mean)) +
+ggplot(tracks_SS_08, aes(x = UTC, y = sst_mean)) +
   geom_line() +
@@
 #Mean depth over time
-ggplot(means_08, aes(x = `as.Date(dateTime)`, y = depth*(-1))) +
+ggplot(means_08, aes(x = `as.Date(UTC)`, y = depth*(-1))) +
   geom_line()
@@
 # CHl over time
-ggplot(tracks_SS_08, aes(x = dateTime, y = chlorophyll_mean)) +
+ggplot(tracks_SS_08, aes(x = UTC, y = chlorophyll_mean)) +
   geom_line() +
@@
 # Mixed Layer Depth over time
-ggplot(tracks_SS_08, aes(x = dateTime, y = mldDepth*(-1))) +
+ggplot(tracks_SS_08, aes(x = UTC, y = mldDepth*(-1))) +
   geom_line() +
@@
 ```
+
 Facet Grids for Env. Variables
@@
   filter(parameter %in% c("sst_mean", "dist_shore", "chlorophyll_mean", "ssh_mean")) %>%
-  rename(Date = `as.Date(dateTime)`)
+  rename(Date = `as.Date(UTC)`)
 
@@
   filter(parameter %in% c("mldDepth", "mldTemp", "ttDepth", "ttTemp", "temp400", "sal400")) %>%
-    rename(Date = dateTime) %>%
+    rename(Date = UTC) %>%
     subset(species!="?BW")
@@
   filter(parameter %in% c("temp400", "sal400")) %>%
-  rename(`Date (2018)` = dateTime)
+  rename(`Date (2018)` = UTC)
 
@@
   geom_point(aes(color = species)) + 
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
+  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43", "BW" = "BW"))
 q + facet_grid(parameter ~ ., scales = "free_y", labeller = as_labeller(env.labs))
@@
 track08_09.04 <- tracks_SS_08 %>%
-  filter(dateTime >= "2018-09-04 00:00:00" & dateTime <= "2018-09-05 00:00:00")
-
-ggplot(data = track08_09.04, aes(x = dateTime, y = `TOL_125`)) +
+  filter(UTC >= "2018-09-04 00:00:00" & UTC <= "2018-09-05 00:00:00")
+
+ggplot(data = track08_09.04, aes(x = UTC, y = `TOL_125`)) +
   geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
@@
 
-#### Zooms in further into listening area when vessel transits through.
-track08_09.04a <- tracks_SS_08 %>%
-  filter(dateTime >= "2018-09-04 12:00:00" & dateTime <= "2018-09-04 16:00:00")
-
-ggplot(data = track08_09.04a, aes(x = dateTime, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
-  geom_text(aes(as.POSIXct("2018-09-04 14:13:00"), 86, label = "Enter vessel", hjust = 0.9)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-04 15:17:00"), color = "yellow", linewidth = 2) +
-  geom_text(aes(as.POSIXct("2018-09-04 15:13:00"), 85, label = "Exit vessel", hjust = 0.3)) +
-  geom_line() +
-  geom_point(data = track08_09.04a, aes(color = species, size = 2)) +
-  scale_color_discrete(na.translate=FALSE) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-04-18\n(Within 7 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/05/18 co-occurrence of vessel and drift within 10 km. NO BW detections
+#### 09/05/18 co-occurrence of vessel and drift within 10 km.
 track08_09.05 <- tracks_SS_08 %>%
-  filter(dateTime >= "2018-09-04 21:00:00" & dateTime <= "2018-09-06 00:00:00")
-
-ggplot(data = track08_09.05, aes(x = dateTime, y = `TOL_125`)) +
+  filter(UTC >= "2018-09-04 21:00:00" & UTC <= "2018-09-06 00:00:00")
+
+ggplot(data = track08_09.05, aes(x = UTC, y = `TOL_125`)) +
   geom_vline(xintercept = as.POSIXct("2018-09-05 01:33:00"), color = "yellow", linewidth = 1) +
@@
   geom_line() + 
-  scale_color_discrete(na.translate=FALSE) +
+  geom_point(data = track08_09.05, aes(color = species, size = 2)) +
   scale_size_continuous(guide = "none") +
@@
 track08_09.10 <- tracks_SS_08 %>%
-  filter(dateTime >= "2018-09-10 00:00:00" & dateTime <= "2018-09-11 00:00:00")
-
-ggplot(data = track08_09.10, aes(x = dateTime, y = `TOL_125`)) +
+  filter(UTC >= "2018-09-10 00:00:00" & UTC <= "2018-09-11 00:00:00")
+
+ggplot(data = track08_09.10, aes(x = UTC, y = `TOL_125`)) +
   geom_vline(xintercept = as.POSIXct("2018-09-10 12:58:00"), color = "yellow", linewidth = 1) +
@@
   geom_point(data = track08_09.10, aes(color = species, size = 2)) +
-  scale_color_discrete(na.translate=FALSE) +
   scale_size_continuous(guide = "none") +
@@
 track08_09.22 <- tracks_SS_08 %>%
-  filter(dateTime >= "2018-09-22 00:00:00" & dateTime <= "2018-09-23 00:00:00")
-
-ggplot(data = track08_09.22, aes(x = dateTime, y = `TOL_125`)) +
+  filter(UTC >= "2018-09-22 00:00:00" & UTC <= "2018-09-23 00:00:00")
+
+ggplot(data = track08_09.22, aes(x = UTC, y = `TOL_125`)) +
   geom_vline(xintercept = as.POSIXct("2018-09-22 19:13:00"), color = "yellow", linewidth = 1) +
@@
   geom_point(data = track08_09.22, aes(color = species, size = 2)) +
-  scale_color_discrete(na.translate=FALSE) +
   scale_size_continuous(guide = "none") +
@@
 track08_09.23 <- tracks_SS_08 %>%
-  filter(dateTime >= "2018-09-23 00:00:00" & dateTime <= "2018-09-24 00:00:00")
-
-ggplot(data = track08_09.23, aes(x = dateTime, y = `TOL_125`)) +
+  filter(UTC >= "2018-09-23 00:00:00" & UTC <= "2018-09-24 00:00:00")
+
+ggplot(data = track08_09.23, aes(x = UTC, y = `TOL_125`)) +
   geom_vline(xintercept = as.POSIXct("2018-09-23 03:55:00"), color = "yellow", linewidth = 1) +
@@
 print(max(ZC$TOL_125))
-print(max(qBW$TOL_125))
 print(max(BW43$TOL_125))
@@
   geom_histogram(data = BW43, mapping = aes(TOL_125), binwidth = 0.2, fill = "yellow", color = "white") +
-  geom_histogram(data = qBW, mapping = aes(TOL_125), binwidth = 0.2, fill = "blue", color = "white") +
   theme_classic()
@@
 
-Density Plots - Species at TOL sound levels. REMOVED ?BW HERE
+Density Plots - Species at TOL sound levels
 ```{r}
@@
   filter(!is.na(species)) %>%
-  subset(species!="?BW") %>%
   ggplot(aes(`BB_20.24000`, fill = species)) + 
@@
   filter(!is.na(species)) %>%
-  subset(species!="?BW") %>%
   ggplot(aes(TOL_63, fill = species)) + 
@@
   filter(!is.na(species)) %>%
-  subset(species!="?BW") %>%
   ggplot(aes(TOL_125, fill = species)) + 
@@
   filter(!is.na(species)) %>%
-  subset(species!="?BW") %>%
   ggplot(aes(TOL_2000, fill = species)) + 
@@
   filter(!is.na(species)) %>%
-  subset(species!="?BW") %>%
   ggplot(aes(TOL_5000, fill = species)) + 
@@
   filter(!is.na(species)) %>%
-  subset(species!="?BW") %>%
   ggplot(aes(TOL_20000, fill = species)) + 
@@
 # Soundscape + bathymetry
+whales08_sf <- BWsf %>%                            
+  dplyr::filter(station == "8")
+
 ggplot() +
@@
 ```
+
+Map of Sea Surface Temp with rerddap, Drift 8 only
+```{r}
+# Multi-scale Ultra-high Resolution (MUR) SST Analysis fv04.1, Global, 0.01°, 2002-present, Daily
+SSTInfo <- info('jplMURSST41') 
+murSST <- griddap(SSTInfo, latitude = c(35, 39), longitude = c(-122, -130), 
+                  time = c("2018-08-16", "2018-10-01"), 
+                  fields = 'analysed_sst') # drift 8 dates only
+SST <- murSST$data
+
+# Raster of SST
+ggplot() + 
+  geom_tile(data = SST, aes(x = longitude, y = latitude, fill = analysed_sst)) +
+  scale_fill_gradient(low = "darkblue", high = "yellow") +
+  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(), 
+        axis.title = element_blank(),
+        panel.background = element_blank(),
+        panel.border = element_rect(fill = NA),
+        legend.position = "none")
+```
+
+Plot distance to continental slope
+```{r}
+ggplot() +
+  geom_raster(data = bathy_df, 
+              mapping = aes(x=x, y=y,
+                            fill = `gebco_bathy`)) +
+  geom_sf(data = tracks_SS_08, 
+          mapping = aes(col = dist_slope)) +
+  geom_sf(data = cont_slope) + scale_fill_continuous(high = "lightgreen", low = "blue")
+
+# PLOT distance by TOL with whales
+ggplot(data = tracks_SS_08, aes(x = dist_slope, y = `TOL_125`)) + 
+  geom_point() + 
+  geom_smooth(color = "green") +
+  theme_classic()
+```
+
+Another map - need to decide if important to keep
+```{r}
+# Plot
+oceanBase <- get_tiles(tracks_SS_08, provider="Esri.NatGeoWorldMap", crop = TRUE)
+tm_shape(oceanBase) + tm_rgb() +
+  tm_shape(tracks_SS_08) +  tm_symbols(col="species", size = 1) +
+  tm_layout(legend.width = 0.5, 
+            legend.position = c("right","top"))
+
+#Raster base map
+oceanBase <- get_tiles(tracks_SS_10, provider="Esri.NatGeoWorldMap")
+tm_shape(bathy) + tm_raster() +
+  tm_shape(tracks_SS_10) + 
+  tm_symbols(style = "jenks") +
+  tm_layout(legend.width = 0.5)
+```

### 2023-04-19 data/soundscape_metrics.R
Created separate scripts

@@ -1,0 +1,31 @@
+## code to prepare `soundscape` dataset goes here
+
+#NEED TO MAKE THIS A FUNCTION
+# look at http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops for help on automation.
+
+library(tidyverse)
+library(lubridate)
+
+
+#dlist <- c(4,7,8,10,12,13,14,16,17,18,19,20,21,22,23)
+sslist <- list.files(path = paste0("data-raw/"), pattern = "CCES_", recursive = TRUE)
+
+#code doesnt work, can delete
+#for (i in sslist) {
+#  read_csv(paste0("data-raw/",i), show_col_types = FALSE)%>%
+#  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
+#  mutate(dateTime = round_date(ymd_hms(.$dateTime),"20 minutes")) %>%
+#  assign(., value = paste0((substr(sslist[i], 16, 17)), substr(sslist[i], 6, 7)))
+#}
+
+#From ChatGPT, edited from above
+for (i in sslist) {
+  data <- read_csv(here(paste0("data-raw/", i)), show_col_types = FALSE) %>%
+    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
+    mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
+    mutate(variable = paste0(substr(i, 16, 17), substr(i, 6, 7)))
+  
+  assign(paste0(substr(i, 16, 17), "_", substr(i, 6, 7)), data)
+}
+
+# Need to save as .rda

### 2023-04-19 data/tracks.R

@@ -1,0 +1,23 @@
+## code to prepare in `tracks` dataset goes here
+
+library(tidyverse)
+
+# Load raw rda file
+load("data-raw/CCES2018_DriftTracks_Modified_03Nov2022.rda")
+
+# Read in tracks, filter out lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17)
+tracks <- AllTracks %>%
+  dplyr::select(-dist, -speed) %>%
+  mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>%      # round to nearest 20 minutes, will create duplicated in the data
+  filter(!station %in% c('1','2','3','4','5','6','9','11', '17')) %>%
+  rename(Longitude = long,                                 # Need these columns to specifically say this for ERDDAP matching
+         Latitude = lat,
+         UTC = dateTime)
+tracks <- tracks[!duplicated(tracks[c('UTC', 'station')]),]    # Remove duplicates
+
+# Save as RDS
+saveRDS(tracks, 'tracks.rda')
+
+# Filter by buoy
+#tracks_08 <- tracks %>% dplyr::filter(station == "8") 
+track4 <- AllTracks %>% dplyr::filter(station == "4")

### 2023-04-19 data/whales.R

@@ -1,0 +1,25 @@
+## code to prepare `whale` dataset here
+
+library(tidyverse)
+
+# Load raw RDA file
+#load("data-raw/CCES2018_BW_Detections.rda")             # Beaked whales only
+load("data-raw/CCES2018_BW_and_PM_Detections.rda")       # Beaked  + sperm whales
+
+# Change EventInfo object name to "whales" and edit columns
+whales <- EventInfo %>%
+  dplyr::select(-Project, -UID, -Id) %>%            # remove unnecessary columns
+  subset(species!= "?BW") %>%                    # remove rows containing ?BW and BW
+  subset(species!= "BW") %>%
+  mutate(dateTimeRound = round_date(.$StartTime, "20 minutes")) %>%   # Round to nearest 20 min
+  rename(station = Deployment)                    # Rename for common join column name next
+
+## NEED TO CHANGE STATION 15 TO STATION 14 HERE. Drift 15 and 14 are the same.
+whales$station[whales$station == 15] <- 14
+
+# Create common UTC field and remove unnecessary fields
+whales <- left_join(whales, tracks, by = join_by(station, closest("dateTimeRound" <= UTC))) %>%
+  dplyr::select(-Latitude.y, -Longitude.y, -spotID, -minNumber, -maxNumber, -bestNumber)
+
+# Whales data as sf
+BWsf <- whales %>% st_as_sf(coords = c("Longitude.x","Latitude.x"), crs=4326)   # whales as sf

### 2023-04-19 data-raw/soundscape_metrics.R
Added script for raw data

@@ -1,0 +1,33 @@
+## code to prepare soundscape dataset goes here
+
+##########################
+#NEED TO MAKE THIS A FUNCTION
+# look at http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops for help on automation.
+##########################
+
+library(tidyverse)
+library(lubridate)
+
+
+#dlist <- c(4,7,8,10,12,13,14,16,17,18,19,20,21,22,23)
+sslist <- list.files(path = paste0("data/"), pattern = "CCES_", recursive = TRUE)
+
+#code doesnt work, can delete
+#for (i in sslist) {
+#  read_csv(paste0("data/",i), show_col_types = FALSE)%>%
+#  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
+#  mutate(dateTime = round_date(ymd_hms(.$dateTime),"20 minutes")) %>%
+#  assign(., value = paste0((substr(sslist[i], 16, 17)), substr(sslist[i], 6, 7)))
+#}
+
+#From ChatGPT, edited from above
+for (i in sslist) {
+  data <- read_csv(here(paste0("data/", i)), show_col_types = FALSE) %>%
+    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
+    mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
+    mutate(variable = paste0(substr(i, 16, 17), substr(i, 6, 7)))
+  
+  assign(paste0(substr(i, 16, 17), "_", substr(i, 6, 7)), data)
+}
+
+# Need to save as .rda

### 2023-04-19 data-raw/tracks.R

@@ -1,0 +1,23 @@
+## code to prepare in AllTracks dataset goes here
+
+library(tidyverse)
+
+# Load raw rda file
+load("data/CCES2018_DriftTracks_Modified_03Nov2022.rda")
+
+# Read in tracks, filter out lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17)
+tracks <- AllTracks %>%
+  dplyr::select(-dist, -speed) %>%
+  mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>%      # round to nearest 20 minutes, will create duplicated in the data
+  filter(!station %in% c('1','2','3','4','5','6','9','11', '17')) %>%
+  rename(Longitude = long,                                 # Need these columns to specifically say this for ERDDAP matching
+         Latitude = lat,
+         UTC = dateTime)
+tracks <- tracks[!duplicated(tracks[c('UTC', 'station')]),]    # Remove duplicates
+
+# Save as RDS
+saveRDS(tracks, 'tracks.rda')
+
+# Filter by buoy
+#tracks_08 <- tracks %>% dplyr::filter(station == "8") 
+track4 <- AllTracks %>% dplyr::filter(station == "4")

### 2023-04-19 data-raw/whales.R

@@ -1,0 +1,25 @@
+## code to prepare whale detections dataset here
+
+library(tidyverse)
+
+# Load raw RDA file
+#load("data/CCES2018_BW_Detections.rda")             # Beaked whales only
+load("data/CCES2018_BW_and_PM_Detections.rda")       # Beaked  + sperm whales
+
+# Change EventInfo object name to "whales" and edit columns
+whales <- EventInfo %>%
+  dplyr::select(-Project, -UID, -Id) %>%            # remove unnecessary columns
+  subset(species!= "?BW") %>%                    # remove rows containing ?BW and BW
+  subset(species!= "BW") %>%
+  mutate(dateTimeRound = round_date(.$StartTime, "20 minutes")) %>%   # Round to nearest 20 min
+  rename(station = Deployment)                    # Rename for common join column name next
+
+## NEED TO CHANGE STATION 15 TO STATION 14 HERE. Drift 15 and 14 are the same.
+whales$station[whales$station == 15] <- 14
+
+# Create common UTC field and remove unnecessary fields
+whales <- left_join(whales, tracks, by = join_by(station, closest("dateTimeRound" <= UTC))) %>%
+  dplyr::select(-Latitude.y, -Longitude.y, -spotID, -minNumber, -maxNumber, -bestNumber)
+
+# Whales data as sf
+BWsf <- whales %>% st_as_sf(coords = c("Longitude.x","Latitude.x"), crs=4326)   # whales as sf

### 2023-04-19 02_AIS.Rmd
Added code for AIS pts within 10km buffer

@@
 library(sf)
-```
 
-```{r}
-#########
-# AIS
-#########
-
-# Need to add more AIS dates in data folder
-AISlist <- list.files(path = paste0("data/"), pattern = "AIS_", recursive = TRUE)
-
-for (v in AISlist) {
-  Adata <- read_csv(paste0("data/", v), show_col_types = FALSE) %>%
-    filter(Length >= 100) %>%                  # Filter out vessels over 100m in length
-    dplyr::select(-SOG,-COG, -CallSign, -IMO, -Cargo, -Draft, -Status) %>%  #Remove unnecessary columns
-    mutate(variable = paste0(substr(v, 16, 17), substr(v, 6, 7)))
-
-  assign(paste0("AIS_", substr(v, 5, 10)), Adata)
-}
-
-AIS_sf <- AIS %>%
-  st_as_sf(coords = c("LON", "LAT"), crs=4326, remove=F)
-```
-
-```{r}
-##################################
-# Create 10 km buffer around buoys (taken from M. McKenna's code)
-# Should be malleable for each buoy's number to easily plug in.
-##################################
-trk8 <- tracks %>% 
-  dplyr::filter(station == "8") %>%
-  st_as_sf(coords= c("Longitude", "Latitude"), crs=4326, remove=F)
-
-pois <- st_transform(trk8, crs = 4326)    # Specify a crs
-pts_buff <- st_buffer(pois, dist = 10000)  # Buffer all points in trk to 10 km (10000m)
-buff <- st_transform(pts_buff, crs = 4326)  # Specify crs for new buffer
-
-### Crop full AIS to 10km buffer around drift 8 ###
-AIS_crop <- st_crop(x = AIS_sf, y = buff)
-
-# Intersect buffer with AIS. Starting this on first 10 rows to see how slowly it goes. 
-## FIX ME
-#inters <- st_intersection(buff, AIS_crop[1:140,])
-
-#Another way to view with TRUE/FALSE if AIS points fall within buffer
-intersect <- st_intersects(buff, AIS_crop)
-lgth <- lengths(st_intersects(buff, AIS_crop)) > 0
-
+# Function
 # Speeding up st_intersection (from https://github.com/r-spatial/sf/issues/801)
@@
 }
+```
+
+```{r}
+#########
+# AIS
+#########
+
+# Need to add more AIS dates in data folder
+AISlist <- list.files(path = paste0("data-raw/"), pattern = "AIS_", recursive = TRUE)
+
+for (v in AISlist) {
+  Adata <- read_csv(paste0("data-raw/", v), show_col_types = FALSE) %>%
+    filter(Length >= 100) %>%                  # Filter out vessels over 100m in length
+    dplyr::select(-SOG,-COG, -CallSign, -IMO, -Cargo, -Draft, -Status) %>%  #Remove unnecessary columns
+    mutate(variable = paste0(substr(v, 16, 17), substr(v, 6, 7)))
+
+  assign(paste0("AIS_", substr(v, 5, 10)), Adata)
+}
+
+AIS_sf <- AIS_091218 %>%
+  st_as_sf(coords = c("LON", "LAT"), crs=4326, remove=F)
+```
+
+```{r}
+##################################
+# Create 10 km buffer around buoys (taken from M. McKenna's code)
+# Should be malleable for each buoy's number to easily plug in.
+##################################
+trk8 <- tracks %>% 
+  dplyr::filter(station == "8") %>%
+  st_as_sf(coords= c("Longitude", "Latitude"), crs=4326, remove=F)
+
+pois <- st_transform(trk8, crs = 4326)    # Specify a crs
+pts_buff <- st_buffer(pois, dist = 10000)  # Creates 10 km buffer around buoy points 10 km (10000m)
+buff <- st_transform(pts_buff, crs = 4326)  # Specify crs for new buffer
+
+### Crop full AIS to 10km buffer around drift 8 ###
+AIS_crop <- st_crop(x = AIS_sf, y = buff)
+
+# Intersect buffer with AIS. Starting this on first 10 rows to see how slowly it goes. 
+## FIX ME
+#inters <- st_intersection(buff, AIS_crop[1:140,])
+
+#Another way to view AIS points that fall within buffer. DOES NOT consider time.
+AIS_within_buff <- AIS_crop[buff,]
+by <- join_by(closest(UTC >= BaseDateTime))
+trk_AIS <- left_join(as.data.frame(trk8), as.data.frame(AIS_within_buff), by)
+
 inters <- st_intersection_faster(buff, AIS_crop[1:140,])
@@
 # Plot showing AIS bounding box around drift 8 tracks
-ggplot() + geom_sf(data = AIS_crop,color = "red")
+ggplot() + geom_sf(data = trk_AIS, color = "red")
+ggplot() + geom_sf(data = buff[lgth,] ,color = "red")
 

### 2023-04-21 01_readin.R
Update 01_readin.R
Removed code that is in other scripts, added joinTable function, added code to read in RDS files

@@ -1,5 +1,5 @@
-
-# Read in, load, clean up, and wrangle raw data
+## code to assemble clean dataframe for each buoy.
+## Prerequisites include whales.R, tracks.R, soundscape_metrics.R.
 
 # Look into https://bookdown.org/yihui/rmarkdown-cookbook/managing-projects.html for info on 
 # managing projects with multiple Rmds and scripts
@@ -15,120 +15,21 @@ library(here)
 
 # set up #####
 
-here()
-#Remove global environment if needed
-#rm(list=ls()) 
-
 # Set system time zone to UTC
 Sys.setenv(TZ='UTC')
+here()
 
-worldmap <- ne_countries(scale = 'medium', type = 'map_units',
-                         returnclass = 'sf')
-calif <- st_crop(worldmap, 
-                 xmin = 33.7,
-                 xmax = 40,
-                 ymin = -131,
-                 ymax = -122)
-calif <- st_transform(calif, 4326)
+# Clear global environment if needed
+# rm(list=ls()) 
 
-tmap_options(basemaps=c(Terrain = "Esri.WorldTerrain",
-                        Imagery = "Esri.WorldImagery", 
-                        OceanBasemap = "Esri.OceanBasemap", 
-                        Topo="OpenTopoMap",
-                        Ortho="GeoportailFrance.orthos"))
-
-
-##########################
-# Read in soundscape data, NEED TO MAKE THIS A FUNCTION
-# look at http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops for help on automation.
-##########################
-
-#dlist <- c(4,7,8,10,12,13,14,16,17,18,19,20,21,22,23)
-sslist <- list.files(path = paste0("data-raw/"), pattern = "CCES_", recursive = TRUE)
-
-#code doesnt work, can delete
-#for (i in sslist) {
-#  read_csv(paste0("data-raw/",i), show_col_types = FALSE)%>%
-#  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
-#  mutate(dateTime = round_date(ymd_hms(.$dateTime),"20 minutes")) %>%
-#  assign(., value = paste0((substr(sslist[i], 16, 17)), substr(sslist[i], 6, 7)))
-#}
-
-#From ChatGPT, edited from above
-for (i in sslist) {
[…]

### 2023-04-21 02_AIS.Rmd
Added code for measuring distance

@@
 library(sf)
+library(gganimate)
+#install.packages("remotes")
+#remotes::install_github("skgrange/gissr")
+#devtools::install_github("skgrange/gissr")
+library(gissr)
 
@@
 }
+
 ```
@@
 ##################################
+tracks <- readRDS("data-raw/tracks.rda")
+
 trk8 <- tracks %>% 
@@
 AIS_crop <- st_crop(x = AIS_sf, y = buff)
+```
 
+```{r}
 # Intersect buffer with AIS. Starting this on first 10 rows to see how slowly it goes. 
@@
 AIS_within_buff <- AIS_crop[buff,]
+f <- buff[AIS_within_buff,]
 by <- join_by(closest(UTC >= BaseDateTime))
+
+d <- full_join(buff, as.data.frame(AIS_within_buff), by)
+
+e <- d[!duplicated(d[c('BaseDateTime')]),]
+
 trk_AIS <- left_join(as.data.frame(trk8), as.data.frame(AIS_within_buff), by)
 
-inters <- st_intersection_faster(buff, AIS_crop[1:140,])
-inters2 <- st_intersection_faster(buff, AIS_crop[141:200,])
-inters3 <- st_intersection_faster(buff, AIS_crop[201:300,])
+proj4string <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
+trk_AIS_sf <- trk_AIS %>% st_as_sf(coords = c("Longitude", "Latitude"), crs=projcrs, remove=F)# %>%
+   st_set_crs(4326)
+trk_AIS_sf <- st_transform(trk_AIS_sf, crs = 4326)
 
+#inters <- st_intersection_faster(buff, AIS_crop[1:140,])
+#inters2 <- st_intersection_faster(buff, AIS_crop[141:200,])
+#inters3 <- st_intersection_faster(buff, AIS_crop[201:300,])
+
+#Another way to view AIS points that fall within buffer. DOES NOT consider time.
+
+st_is_within_distance(AIS_crop, pois, dist = 10000)
+st_distance(AIS_crop, pois, dist_fun = 10000)
+
+st_set_geometry(AIS_crop, value = NULL)
+st_set_geometry(pois, value = NULL)
+
+AIS_crop <- as_Spatial(AIS_crop)
+pois <- as_Spatial(pois)
+AIS_within_buff <- as_Spatial(AIS_within_buff)
+z <- sp_distance(AIS_crop, pois, features = TRUE, unit = "km")
+
+data_places <- pois@data
+data_places$distance <- sp_distance(AIS_within_buff, pois, features = TRUE, unit = "km")
+View(data_places)
+
+rgeos::gWithinDistance(AIS_crop, pois, features = TRUE, dist = 10000)
 ```
+```{r}
+# For each row in AIS df,
+  # calculate distance to closest drift
+  # return smallest distance
+  # attach column to df
+
+# 
+(geosphere::distHaversine(AIS_within_buff[9:20,], pois))/1000
+```
+
 ```{r}
@@
 
+plots
 ```{r}
@@
 # Plot showing AIS bounding box around drift 8 tracks
-ggplot() + geom_sf(data = trk_AIS, color = "red")
+ggplot() + geom_sf(data = trk_AIS_sf, color = "red")
 ggplot() + geom_sf(data = buff[lgth,] ,color = "red")
@@
  # geom_sf(data = AIS_sf)
+
+p <- ggplot() + geom_sf(data = buff, mapping = aes(x=Longitude,y=Latitude)) +
+  geom_point(data = AIS_within_buff, mapping = aes(x=LON,y=LAT, group = VesselName), color = "red")
+
+anim <- p +
+  transition_reveal(along = BaseDateTime)+
+  ease_aes('linear')+
+  #shadow_wake(0.1) +
+  ggtitle("Date: {frame_along}")
+
+animate(anim, nframes = 180, fps = 3)
+```

### 2023-04-21 data-raw/soundscape_metrics.R
Moved to raw data folder
Added loop to read files through hard drive

@@ -1,28 +1,17 @@
-## code to prepare soundscape dataset goes here
-
-##########################
-#NEED TO MAKE THIS A FUNCTION
-# look at http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops for help on automation.
-##########################
+## code to prepare `soundscape` dataset goes here
 
 library(tidyverse)
 library(lubridate)
 
+# Path to folder in Project/on GitHub
+sslist <- list.files(path = paste0("data-raw/"), pattern = "CCES_", recursive = TRUE)
 
-#dlist <- c(4,7,8,10,12,13,14,16,17,18,19,20,21,22,23)
-sslist <- list.files(path = paste0("data/"), pattern = "CCES_", recursive = TRUE)
+# Path to folder on my external hard drive
+dirlist <- dir("D:/Soundscape_metrics", recursive=TRUE, full.names=TRUE, pattern="CCES_")
 
-#code doesnt work, can delete
-#for (i in sslist) {
-#  read_csv(paste0("data/",i), show_col_types = FALSE)%>%
-#  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
-#  mutate(dateTime = round_date(ymd_hms(.$dateTime),"20 minutes")) %>%
-#  assign(., value = paste0((substr(sslist[i], 16, 17)), substr(sslist[i], 6, 7)))
-#}
-
-#From ChatGPT, edited from above
+# Read through soundscape files, change date format
 for (i in sslist) {
-  data <- read_csv(here(paste0("data/", i)), show_col_types = FALSE) %>%
+  data <- read_csv(here(paste0("data-raw/", i)), show_col_types = FALSE) %>%
     rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
     mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
     mutate(variable = paste0(substr(i, 16, 17), substr(i, 6, 7)))
@@ -30,4 +19,14 @@ for (i in sslist) {
   assign(paste0(substr(i, 16, 17), "_", substr(i, 6, 7)), data)
 }
 
-# Need to save as .rda
+# Read though files from external hard drive
+for (i in dirlist) {
+  data <- read_csv(i, show_col_types = FALSE) %>%
+    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
[…]

### 2023-04-21 data-raw/tracks.R

@@ -1,9 +1,9 @@
-## code to prepare in AllTracks dataset goes here
+## code to prepare in `tracks` dataset goes here
 
 library(tidyverse)
 
 # Load raw rda file
-load("data/CCES2018_DriftTracks_Modified_03Nov2022.rda")
+load("data-raw/CCES2018_DriftTracks_Modified_03Nov2022.rda")
 
 # Read in tracks, filter out lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17)
 tracks <- AllTracks %>%
@@ -16,7 +16,7 @@ tracks <- AllTracks %>%
 tracks <- tracks[!duplicated(tracks[c('UTC', 'station')]),]    # Remove duplicates
 
 # Save as RDS
-saveRDS(tracks, 'tracks.rda')
+saveRDS(tracks, 'data/tracks.rda')
 
 # Filter by buoy
 #tracks_08 <- tracks %>% dplyr::filter(station == "8") 

### 2023-04-21 data-raw/whales.R

@@ -1,10 +1,11 @@
-## code to prepare whale detections dataset here
+## code to prepare `whales` dataset here
 
 library(tidyverse)
+library(sf)
 
 # Load raw RDA file
-#load("data/CCES2018_BW_Detections.rda")             # Beaked whales only
-load("data/CCES2018_BW_and_PM_Detections.rda")       # Beaked  + sperm whales
+#load("data-raw/CCES2018_BW_Detections.rda")             # Beaked whales only
+load("data-raw/CCES2018_BW_and_PM_Detections.rda")       # Beaked + sperm whales
 
 # Change EventInfo object name to "whales" and edit columns
 whales <- EventInfo %>%
@@ -23,3 +24,6 @@ whales <- left_join(whales, tracks, by = join_by(station, closest("dateTimeRound
 
 # Whales data as sf
 BWsf <- whales %>% st_as_sf(coords = c("Longitude.x","Latitude.x"), crs=4326)   # whales as sf
+
+# Save object
+saveRDS(whales, 'data/whales.rda')

### 2023-04-21 data/tracks.R
Removed

@@ -1,23 +1,0 @@
-## code to prepare in `tracks` dataset goes here
-
-library(tidyverse)
-
-# Load raw rda file
-load("data-raw/CCES2018_DriftTracks_Modified_03Nov2022.rda")
-
-# Read in tracks, filter out lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17)
-tracks <- AllTracks %>%
-  dplyr::select(-dist, -speed) %>%
-  mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>%      # round to nearest 20 minutes, will create duplicated in the data
-  filter(!station %in% c('1','2','3','4','5','6','9','11', '17')) %>%
-  rename(Longitude = long,                                 # Need these columns to specifically say this for ERDDAP matching
-         Latitude = lat,
-         UTC = dateTime)
-tracks <- tracks[!duplicated(tracks[c('UTC', 'station')]),]    # Remove duplicates
-
-# Save as RDS
-saveRDS(tracks, 'tracks.rda')
-
-# Filter by buoy
-#tracks_08 <- tracks %>% dplyr::filter(station == "8") 
-track4 <- AllTracks %>% dplyr::filter(station == "4")

### 2023-04-21 data/whales.R

@@ -1,25 +1,0 @@
-## code to prepare `whale` dataset here
-
-library(tidyverse)
-
-# Load raw RDA file
-#load("data-raw/CCES2018_BW_Detections.rda")             # Beaked whales only
-load("data-raw/CCES2018_BW_and_PM_Detections.rda")       # Beaked  + sperm whales
-
-# Change EventInfo object name to "whales" and edit columns
-whales <- EventInfo %>%
-  dplyr::select(-Project, -UID, -Id) %>%            # remove unnecessary columns
-  subset(species!= "?BW") %>%                    # remove rows containing ?BW and BW
-  subset(species!= "BW") %>%
-  mutate(dateTimeRound = round_date(.$StartTime, "20 minutes")) %>%   # Round to nearest 20 min
-  rename(station = Deployment)                    # Rename for common join column name next
-
-## NEED TO CHANGE STATION 15 TO STATION 14 HERE. Drift 15 and 14 are the same.
-whales$station[whales$station == 15] <- 14
-
-# Create common UTC field and remove unnecessary fields
-whales <- left_join(whales, tracks, by = join_by(station, closest("dateTimeRound" <= UTC))) %>%
-  dplyr::select(-Latitude.y, -Longitude.y, -spotID, -minNumber, -maxNumber, -bestNumber)
-
-# Whales data as sf
-BWsf <- whales %>% st_as_sf(coords = c("Longitude.x","Latitude.x"), crs=4326)   # whales as sf

### 2023-05-11 manuscript.rmd
Added to manuscript

---
title: "manuscript"
format: html
editor: visual
---

# Manuscript Draft

## Introduction

\[Introduce sound, soundscapes, noise in the ocean, noise pollution\]\
Sound comes from a variety of sources in the ocean - biotic, abiotic, human. The ocean is naturally a noisy place, but recent increases in human activities are increasing the ocean sound levels. It's likely this noise pollution will interfere with soniferous animals, but it's unclear how, where, and what species will be affected.

\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]\
Among the protected species in the California Current Ecosystem are beaked whale and sperm whales. These are toothed whales known as some of the deepest divers in the ocean, reaching depths of nearly 10,000 ft, or 3,000 meters. They can dive for over 3 hours, followed by only a few minutes on the surface. This diving behavior, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them. Their reliance on sound can make them particularly sensitive to human noise. Sperm whales may have more acoustic tolerance than beaked whales (Miller et al. 2022). The acoustic interference from vessels, for example, can affect behavior, disrupt communication, or lead to serious harm, even at far distances (See Hooker et al. 2019 for more info). There are several documented cases of mid-frequency naval sonar activities coinciding with mass stranding events by beaked whales (see Anne's paper). Under the Marine Mammal Protection Act, all these species are federally protected, and several of these species have threatened or endangered conservation statuses. Despite their visually cryptic nature, their sound signals are very well-described. With automated classification methods, the clicks and pulses they make during foraging can be identified down to the species level. \[Can cite specific examples of noise as stressor to whales.\]

\[Introduce the importance of passive acoustic monitoring on this type of study and gaps in visual survey methods\]\
Passive acoustic monitoring is a tool that allows us to record and listen to the soundscape. Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen. One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters; it can drift freely with the currents for days or weeks at a time. The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only. Its basic components consist of a satellite transmitter above water to monitor GPS information and an acoustic recorder attached by line at \~100m of depth. In fact, most of what is known about these deep divers is because of acoustic monitoring.

\[Main research points\]\
Shipping and offshore activities are expected to increase in the future, causing elevated oceanic sound levels. This is a form of habitat degradation, so there is a need to understand how human activities change and contribute to the overall soundscape in space and time. It's a challenge to acquire information on these federally protected and acoustically-sensitive species but becoming increasingly important to understand where habitats overlap with human activity. Acoustics can serve as important indicators of ecosystem health, but also of how humans are changing the environment. Quantifying soundscapes in a standardized way and monitoring these changes over time are priorities for marine managers and researchers right now.

\[Goal\]\
Characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics and vessel tracking

\[How this work will contribute to gaps in literature\]\
This work will contribute to establishing baseline conditions of noise exposure to inform wind farm development and shipping lane changes. It'll also add to the knowledge base of beaked and sperm whale distribution through acoustic detections. Both of these outcomes are of high importance to managers and agencies especially NOAA.

## Methods

\[Overall data collection efforts\]\
Acoustic detections were collected with the use of free-drifting Drifting Acoustic Spar Buoy Recorders (DASBRs). These were deployed as part of the California Current Ecosystem Survey (CCES) spanning the offshore waters from southern Canada to northern Mexico in summer of 2018. DASBRs offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species. These DASBRs contained two satellite geo-location trackers, two hydrophones, etc and collected a total of 1910 hours of data. SWFSC ran acoustic detections in PAMGuard. Eight drifting recorders were lost at sea and therefore removed from the analysis. These contain 1, 5, 10, 25, 75, 95, and 99 percentile distribution from each metric.

\[Study area\]\
The California Current Ecosystem extends along the US' Economic Exclusion Zone along the west coast. It's a highly productive and complex ecosystem that contains six national marine sanctuaries, supports several wild fisheries, and is home to many protected species. Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area.

\[Add details about acoustic data collection information.\]\
Soundscape metrics were duty-cycled (most 18min off, 2 minutes on, but some variation in "off" times) and later processed by the NOAA team with calls and clicks identified to the species level (see Simonis et al 2020 report for details on this survey and Rand et al. 2022 for duty-cycle study), and soundscape metrics were binned into low, medium, high frequency bands.

\[Post-processing details, data cleaning\]\
Soundscape metrics were rounded to the nearest 20 minutes. These include third-octave level bands and broadband metrics. One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands. The band at 125 Hertz is often associated with shipping vessel noise and the 2000 Hertz or 2 kHz with mid-frequency naval sonar. Beaked whales are known to be sensitive to sonar in this latter range. Several TOL levels were chosen as proxies to vessel or sonar frequencies. Whale detections were rounded to nearest 20 minutes. Two types of BW species (?BW and ?BWC) were removed from analysis as the species is unknown. Whale detections reduced to "presence/absence" from number of clicks, estimated individuals, etc.

Drifter tracks were rounded to nearest 20 minutes, lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17) were removed. Drift 14 and 15 are the same. These three major datasets were joined together based on a common buoy number and time field. Environmetal and oceanographic variables thought to be predictors of BW/SW habitat were extracted from NOAA's Environmental Research Division Data Access Program (ERDDAP) and Hybrid Coordinate Ocean Model (HYCOM) using matchEnvData function in package PAMmisc. Variables at depth were calculated using function from Taiki and formula from McCullough et al 2021. Distance to slope and bathymetric slope were calculated from a bathymetry TIF taken from GEBCO. Can show table of variables and justification by papers (Virgili et al. 2022).

\[AIS\]\
Automatic Identification System vessel tracking data were collected from Marine Cadastre in order to evaluate the effect of spatial proximity to the buoys. This step is important in understanding how ship characteristics contribute to various frequency bands. I filtered out any vessels under 100 meters in length in order to focus only on large vessels. By exploring buoy/vessel co-occurrence with a 10-km "listening space" around the buoy (McKenna) can allow us to investigate patterns of when whales are calling and the vessel activity during co-occurrence.

\[Big analysis: statistics, modeling details\]

Can use non-parametric Kruskal-Wallis one-way analysis of variance test to examine if number of whale detections per hour differed depending on presence/absence of vessels (see Trickey et al. 2022).

## Results

\[Add here.\] \[Weekly/daily plot of acoustic presence of whales and ships over time from each drift? See Trickey et al. 2022\]

## Discussion

\[Known limitations to this study\] Duty cycle whale detection and soundscape method knowingly omits 18 of 20 minutes of data. A sperm whale's true location relative to buoy is unknown so preferred habitat may not reflect the environmental variables that are associated with buoy that detects it (Barkley et al. 2021) Rounding tracks and soundscape metrics date/times to nearest 20 minutes is not perfect. Often, times did not line up with :00, :20, :40 and were either omitted or duplicated.

This study only captures half of the year. If BW/SW have seasonal preferences outside of the study period, it was not captured here.

\[Add here.\]

## Acknowledgements

Taiki - help with code!\
Megan McKenna - guidance on AIS questions. \[Add here.\]

### 2023-05-11 01_readin.R
loop for read in rda files

@@ -33,11 +33,9 @@ whales <- readRDS('data/whales.rda')
 
 trk_whale <- left_join(tracks, whales, by = c("station","UTC")) #%>%
 #  filter(!station %in% c('4','17'))  ## Can filter out buoys 4 and 17 here
-View(trk_whale)
 
-#Show duplicates (none??)
+# Remove duplicates
 trk_whale <- trk_whale[!duplicated(trk_whale[c('UTC', 'station')]),]
-View(trk_whale)
 
 ###############
 # Filter whales by drift
@@ -54,18 +52,20 @@ for (drift in d) {
 
 ##############################################################
 # Joining tracks with soundscape metrics with whale detections
-# Make loop if possible
 ##############################################################
 
-### Objects for forloop if I can figure it out.
-#bblist <- c(BB08, BB10)
-#tolist <- c(TO08, TO10)
-#whlist <- c(whales8, whales10)
+## Make list of rda files to read in
+iilist <- c("BB_07", "BB_08", "BB_10", "BB_12", "BB_13", "BB_14", "BB_16", 
+            "BB_18", "BB_19", "BB_20", "BB_21", "BB_22", "BB_23", "TO_07",
+            "TO_08", "TO_10", "TO_12", "TO_13", "TO_14", "TO_16", "TO_18", 
+            "TO_19", "TO_20", "TO_21", "TO_22", "TO_23")
 
-# to include when I have more BB metrics in my environment       
-#BB12, BB13, BB14, BB15, BB16, BB18, BB19, BB20, BB21, BB22, BB23)
-#TO12, TO13, TO14, TO15, TO16, TO18, TO19, TO20, TO21, TO22, TO23)
-
+## Read in rda files and assign to object
+for (ii in iilist) {
+  dta <- readRDS(paste0('data/', ii, '.rda'))
+  
+  assign(ii, dta)
+}
 
 # Make function to join whales + tracks, broadband, and TOL soundscape metrics and edit columns
 joinTable <- function(w, s, t) {
@@ -87,7 +87,6 @@ joinTable <- function(w, s, t) {
 tracks_SS_07 <- joinTable(trk_whales7, BB_07, TO_07)
 tracks_SS_08 <- joinTable(trk_whales8, BB_08, TO_08)
 tracks_SS_10 <- joinTable(trk_whales10, BB_10, TO_10)
-tracks_SS_11 <- joinTable(trk_whales11, BB_11, TO_11)
 tracks_SS_12 <- joinTable(trk_whales12, BB_12, TO_12)
[…]

### 2023-05-11 02_AIS.Rmd
distance function added

@@
 library(gissr)
+```
 
+```{r}
 # Function
@@
 # 
-(geosphere::distHaversine(AIS_within_buff[9:20,], pois))/1000
+ship_dist <- (geosphere::distHaversine(AIS_within_buff@coords[10:13,], pois))/1000
+cbind(AIS_within_buff@data[10:13,], ship_dist)
 ```

### 2023-05-11 03_exploratory.Rmd
added functions
bathymetry slope and depth extraction from TIF

@@
 ---
-title: "exploratory analysis"
+title: "environmenal/exploratory analysis"
 output: html_document
@@
 library(tidyverse)
-library(lubridate)
 library(sf)
-library(rerddap)
+#library(raster)
 library(terra)
-library(marmap)
-library(rnaturalearth)
-library(mapproj)
-library(raster)
-#library("rerddapXtracto")
 library(PAMpal)
 library(PAMmisc)
-library(maptiles)
-library(tmap)
-library(spData)
-library(tmaptools)
-library(grid)
-library(ggspatial)
-```
-
-Set Parameters
-```{r}
-theme_set(theme_bw())
-tmap_mode("plot")
-```
-
-Create sf objects and basemaps for maps
-```{r}
-whales_8_10sf <- BWsf %>%                            
-dplyr::filter(station == "8" | station == "10")
-
-trackssf <- st_as_sf(tracks, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326)
-trackBase <- get_tiles(whales_8_10sf, provider = "Esri.WorldShadedRelief")
-allTrackbase <- get_tiles(trackssf, provider = "Stamen.TonerLite", zoom = 5)
-```
-### Exploratory analysis
-
-SST from PAMPal
-```{r}
-# Downloads data from servers, uncomment and only run once
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "analysedSST")
-
-# from file 
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/analysedSST.nc') %>%
+#library(marmap)
+library(rerddap)
+```
+
+### Get Environmental Variables
+
+For each buoy, the value of each environmental variable will be added to each track point based on Lat, Long, and UTC.
+Data only need to be downloaded once. After that, can refer to the file in the external hard drive ("D:").
+
+Environmental variables: Sea surface temp, Wind stress curl, Distance to continental slope, Depth, bathymetric slope,
+distance to shore, chlorophyll, sea surface height anomalies, + Hycom variables.
+
+Sea Surface Temperature from ERDDAP using PAMPal function
+```{r}
+# Downloads data from servers onto external hard drive. Uncomment and only run once.
+#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST07", var = c('analysed_sst'))
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST08", var = c('analysed_sst'))
+#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST10", var = c('analysed_sst'))
+#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST12", var = c('analysed_sst'))
+#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST13", var = c('analysed_sst'))
+#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST14", var = c('analysed_sst'))
+#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST16", var = c('analysed_sst'))
+#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST18", var = c('analysed_sst'))
+#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST19", var = c('analysed_sst'))
+#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST20", var = c('analysed_sst'))
+#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST21", var = c('analysed_sst'))
+#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST22", var = c('analysed_sst'))
+#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST23", var = c('analysed_sst'))
+```
+
+Once downloaded, matches SST from file
+```{r}
+SSTlist <- dir("D:/Environ_variables", recursive=TRUE, full.names=TRUE, pattern="analysedSST")
+
+for (ss in SSTlist) {
+  data <- matchEnvData(get(paste0("tracks_SS_", substr(ss, 33, 34))), nc = ss) %>%
   rename(., sst_mean = analysed_sst_mean) %>%
   dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)   # rename new column to "sst_mean" and remove unnecessary columns
-```
-
-Distance to continental slope
-```{r}
-# Increase cell size by 30x to be visible
-bathy <- rast("data/gebco_bathy.tif")
-agg <- terra::aggregate(bathy, 5)
-plot(agg)
-
-# Rasterize track with agg as the template
-
-#TOL_rast <- rasterize(trk_TOL_08, agg, field = "TOL_125")
-sf08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-sf10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-trkraster08 <- rasterize(vect(sf08), agg)
-trkraster10 <- rasterize(vect(sf10), agg)
-
-bathy_df <- as.data.frame(bathy, xy = TRUE)
-
-# Map the coastline and cont. slope
-cont <- st_as_sf(as.contour(bathy))
-coast <- cont %>% filter(level == 0)
-cont_slope <- cont %>% filter(level == -2000)
-cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
-plot(cont_slope_rast, col = "red")
-cont_slope_rast <- rasterize(cont_slope, agg)
-plot(cont_slope_rast, col = "red")
-
-#ggplot() + 
-#  geom_raster(data = bathy_df, 
-#              mapping = aes(x=x, y=y, 
-#                            alpha = 0.8,
-#                            fill = `gebco_bathy`))+
-#  geom_sf(data = cont, col="black") + 
-#  geom_sf(data = coast, col = "red") +
-#  geom_sf(data = cont_slope, col = "purple") +
-#  geom_sf(data = whales_8_10sf, mapping = aes()) +
-#  scale_color_gradient(low="green", high="red") 
-```
-
-### Extract slope_dist and bind to drift tracks
-```{r}
-# Drift 8
-slopedist <- terra::distance(trkraster08, vect(cont_slope))     # measure distance from track to slope
-extdist <- terra::extract(slopedist, vect(sf08))       # extract this distance into a df
-tracks_SS_08 <- cbind(sf08,extdist) %>%                     # bind to main df
+
+  assign(paste0("tracks_SS_", substr(ss, 33, 34)), data)
+}
+```
+### Wind Stress Curl from ERDDAP using PAMPal function
+```{r}
+# Downloads data from servers onto external hard drive. Uncomment and only run once.
+#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress07.nc", var = c('curl'))
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress08.nc", var = c('curl'))
+#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress10.nc", var = c('curl'))
+#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress12.nc", var = c('curl'))
+#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress13.nc", var = c('curl'))
+#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress14.nc", var = c('curl'))
+#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress16.nc", var = c('curl'))
+#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress18.nc", var = c('curl'))
+#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress19.nc", var = c('curl'))
+#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress20.nc", var = c('curl'))
+#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress21.nc", var = c('curl'))
+#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress22.nc", var = c('curl'))
+#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress23.nc", var = c('curl'))
+
+# Pulls from downloaded file
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data-raw/erdQMstress1day_LonPM180.nc')
+#tracks_SS_08
+
+#windcurl <- erddapToEdinfo('erdQMstress1day', baseurl = 'https://coastwatch.pfeg.noaa.gov/erddap/')
+
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=windcurl, fileName='windcurl.nc')
+
+```
+Once downloaded, matches wind stress curl from file
+```{r}
+windlist <- dir("D:/Environ_variables", recursive=TRUE, full.names=TRUE, pattern="windstress")
+
+for (w in windlist) {
+  data <- matchEnvData(get(paste0("tracks_SS_", substr(w, 32, 33))), nc = w) %>%
+     dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean, -matchDepth_mean)   # remove unnecessary columns
+
+  assign(paste0("tracks_SS_", substr(w, 32, 33)), data)
+}
+```
+
+### Distance to continental slope (uses terra package)
+```{r}
+bathy <- rast("D:/CCES_gebco_bathy.tif") # Bathymetry for all of CCE from https://download.gebco.net/
+agg <- terra::aggregate(bathy, 5)  # Increase cell size by factor of 5 to be visible
+
+cont <- st_as_sf(as.contour(bathy, )) # Create contour lines from bathymetry 
+cont_slope <- cont %>% filter(level == -2000) # Establish continental shelf. Is 2000m the correct depth for continental shelf?
+
+tracklist <- grep(pattern = "tracks_SS_", names(.GlobalEnv), value = T)
+
+for (t in tracklist) {
+  sf <- st_as_sf(get(t), coords = c("Longitude", "Latitude"), crs=4326, remove=F)
+  trk <- terra::rasterize(vect(sf), agg)
+  slop <- terra::distance(trk, vect(cont_slope))
+  ext <- terra::extract(slop, vect(sf))
+
+  dbind <- cbind(sf, ext) %>%
   dplyr::select(-ID) %>%
-  rename("dist2slope" = gebco_bathy) %>%
-  st_set_geometry(NULL)                         # Drop geometry to make df for future cbinds
-
-#Drift 10
-slopedist <- terra::distance(trkraster10, vect(cont_slope))     # measure distance from track to slope
-extdist <- terra::extract(slopedist, vect(sf10))       # extract this distance into a df
-tracks_SS_10 <- cbind(sf10,extdist) %>%                     # bind to main df
-  dplyr::select(-ID) %>%
-  #rename("dist2slope" = gebco_bathy) %>%
-  st_set_geometry(NULL)                        # Drop geometry to make df for future cbinds
-```
-
-#### Bathymetry - PAMpal
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'srtm15plus', fileName='depth.nc')
-
-# Pulls from downloaded file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/depth.nc') %>%
+  rename("dist2slope" = last) %>%
+  st_set_geometry(NULL) # remove geometry to prevent buoy dataframe from becoming sf object for future
+
+  assign(paste0("tracks_SS_", substr(t, 11, 12)), dbind)
+}
+```
+
+Depth from ERDDAP using PAMPal function
+```{r}
+# Downloads data from servers onto external hard drive. Uncomment and only run once.
+#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth07.nc", var = c('z'))
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth08.nc", var = c('z'))
+#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth10.nc", var = c('z'))
+#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth12.nc", var = c('z'))
+#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth13.nc", var = c('z'))
+#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth14.nc", var = c('z'))
+#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth16.nc", var = c('z'))
+#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth18.nc", var = c('z'))
+#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth19.nc", var = c('z'))
+#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth20.nc", var = c('z'))
+#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth21.nc", var = c('z'))
+#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth22.nc", var = c('z'))
+#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth23.nc", var = c('z'))
+```
+
+Once downloaded, matches depth from file
+```{r}
+depthlist <- dir("D:/Environ_variables", recursive=TRUE, full.names=TRUE, pattern="depth")
+
+for (dd in depthlist) {
+  data <- matchEnvData(get(paste0("tracks_SS_", substr(dd, 27, 28))), nc = dd) %>%
   rename(., depth = z_mean) %>%        # rename new column to "depth"
@@
   dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-```
-
-#### Bathymetric slope
-
-```{r}
-# Slope (Hard-coded - can fix this)
-bgo <- getNOAA.bathy(lon1 = -121, lon2 = -131,
-                     lat1 = 34, lat2= 40, resolution = 1)  # plot bounding box around drift 8 and 10
-# As raster
-bgo_rast <- marmap::as.raster(bgo)
-bgo_slope <- terrain(bgo_rast, v = "slope")
-
-# Extract slope values at points along drift 8
-slope <- raster::extract(bgo_slope, tracks_SS_08[,4:3])
-tracks_SS_08 <- cbind(tracks_SS_08, slope)
-```
-
-#### Distance from coast - PAMPal
-Need to have variable choose itself!! Also, how do I get it to not download each time?
-```{r}
-#dist2shore <- erddapToEdinfo(dataset='dist2coast_1deg',
-#                             baseurl='https://pae-paha.pacioos.hawaii.edu/erddap/',
-#                             chooseVars = FALSE)
-
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc='data/Dist2Shore.nc') %>%
+
+  assign(paste0("tracks_SS_", substr(dd, 27, 28)), data)
+}
+```
+
+#### Bathymetric slope (skipping loop for now)
+```{r}
+bathy_slope <- terra::terrain(bathy, v = "slope")
+
+for (t in tracklist) {
+  slope <- terra::extract(bathy_slope, get(t)[,4:3]) %>%  # Reads latitude and longitude of buoy dataframe
+  dplyr::select(-ID) %>%
+  rename("bathy_slope" = slope)
+
+  sbind <- cbind(get(t), slope)
+
+  assign(paste0("tracks_SS_", substr(t, 11, 12)), sbind)
+}
+```
+
+Distance to Shore from ERDDAP using PAMPal function
+```{r}
+# Downloads data from servers onto external hard drive. Uncomment and only run once.
+#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore07.nc", var = c('dist'))
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore08.nc", var = c('dist'))
+#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore10.nc", var = c('dist'))
+#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore12.nc", var = c('dist'))
+#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore13.nc", var = c('dist'))
+#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore14.nc", var = c('dist'))
+#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore16.nc", var = c('dist'))
+#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore18.nc", var = c('dist'))
+#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore19.nc", var = c('dist'))
+#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore20.nc", var = c('dist'))
+#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore21.nc", var = c('dist'))
+#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore22.nc", var = c('dist'))
+#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore23.nc", var = c('dist'))
+```
+
+Once downloaded, matches Distance to Shore from file
+```{r}
+distlist <- dir("D:/Environ_variables", recursive=TRUE, full.names=TRUE, pattern="dist2shore")
+
+for (d2 in distlist) {
+  data <- matchEnvData(get(paste0("tracks_SS_", substr(d2, 32, 33))), nc = d2) %>%
   rename(., dist2shore = dist_mean) %>%
@@
 
-#May not need this part
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'dist2coast_1deg', fileName='dist2coast.nc', var = 'dist')
-
-# Pulls from downloaded file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/dist2coast.nc')
-
-# Downloads nc file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=dist2shore, fileName='Dist2Shore.nc')  
-```
-
-### Chlorophyll from PAMPal
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdMH1chlamday', fileName='chlorophyll.nc')
-
-# Pulls from downloaded file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/chlorophyll.nc')
-```
-
-#### Sea Surface Height Anomalies from PAMPal
-```{r}
-# Downloaded from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'nesdisSSH1day', fileName='sshanomalies.nc')
-
-# Pulled from downloaded nc file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/sshanomalies.nc') %>%
+  assign(paste0("tracks_SS_", substr(d2, 32, 33)), data)
+}
+```
+
+### Chlorophyll from ERDDAP using PAMPal function
+```{r}
+# Downloads data from servers onto external hard drive. Uncomment and only run once.
+#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll07.nc", var = c('chlorophyll'))
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll08.nc", var = c('chlorophyll'))
+#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll10.nc", var = c('chlorophyll'))
+#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll12.nc", var = c('chlorophyll'))
+#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll13.nc", var = c('chlorophyll'))
+#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll14.nc", var = c('chlorophyll'))
+#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll16.nc", var = c('chlorophyll'))
+#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll18.nc", var = c('chlorophyll'))
+#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll19.nc", var = c('chlorophyll'))
+#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll20.nc", var = c('chlorophyll'))
+#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll21.nc", var = c('chlorophyll'))
+#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll22.nc", var = c('chlorophyll'))
+#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll23.nc", var = c('chlorophyll'))
+```
+
+Once downloaded, matches chlorophyll from file
+```{r}
+chlalist <- dir("D:/Environ_variables", recursive=TRUE, full.names=TRUE, pattern="chlorophyll")
+
+for (ch in chlalist) {
+  data <- matchEnvData(get(paste0("tracks_SS_", substr(ch, 33, 34))), nc = ch) %>%
+  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
+
+  assign(paste0("tracks_SS_", substr(ch, 33, 34)), data)
+}
+```
+
+Sea Surface Height Anomalies from ERDDAP using PAMPal function
+```{r}
+# Downloads data from servers onto external hard drive. Uncomment and only run once.
+#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh07.nc", var = c('sla'))
+#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh08.nc", var = c('sla'))
+#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh10.nc", var = c('sla'))
+#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh12.nc", var = c('sla'))
+#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh13.nc", var = c('sla'))
+#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh14.nc", var = c('sla'))
+#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh16.nc", var = c('sla'))
+#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh18.nc", var = c('sla'))
+#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh19.nc", var = c('sla'))
+#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh20.nc", var = c('sla'))
+#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh21.nc", var = c('sla'))
+#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh22.nc", var = c('sla'))
+#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh23.nc", var = c('sla'))
+```
+
+Once downloaded, matches sea surface height anomalies from file
+```{r}
+sshlist <- dir("D:/Environ_variables", recursive=TRUE, full.names=TRUE, pattern="ssh")
+
+for (sh in sshlist) {
+  data <- matchEnvData(get(paste0("tracks_SS_", substr(sh, 25, 26))), nc = sh) %>%
   rename(., ssh_mean = sla_mean) %>%
   dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-```
-
-### Wind Stress Curl from PAMPal *** NEEDS FIXING ****
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQMstress1day_LonPM180', fileName='erdQMstress1day_LonPM180.nc', var = 'curl')
-
-# Pulls from downloaded file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQMstress1day_LonPM180.nc')
-#tracks_SS_08
-
-#windcurl <- erddapToEdinfo('erdQMstress1day', baseurl = 'https://coastwatch.pfeg.noaa.gov/erddap/')
-
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=windcurl, fileName='windcurl.nc')  
+
+  assign(paste0("tracks_SS_", substr(sh, 25, 26)), data)
+}
 ```
@@
 #### HyCOM Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
-Doesn't bin itself into separate dataframe based on size!
-```{r}
-# These refer to nc files that have already been downloaded from the server.
-# Uncomment and only run once! Rows are hard-coded.
-
-#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm0.nc")
-
-#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm1.nc")
-
-#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm2.nc")
-
-#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm3.nc")
-
-#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm4.nc")
-
-#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm5.nc")
-
-#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm6.nc")
-
-#hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm7.nc")
-
-#hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm8.nc")
-
-# Run the 9 nc lists that make up tracks_SS_08
-hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm0.nc")
-hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm1.nc")
-hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm2.nc")
-hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm3.nc")
-hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm4.nc")
-hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm5.nc")
-hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm6.nc")
-hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm7.nc")
-hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm8.nc")
-```
 
@@
 
+```{r}
+# Create loop for each buoy to download HYCOM data
+
+##FIX THIS - try testing on ONE buoy
+for (t in tracklist) {
+  Hdata <- matchEnvData(as.data.frame(get(paste0("tracks_SS_", substr(t, 11, 12)))), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
+
+  thermData <- bind_rows(lapply(Hdata, function(x) {
+    # do thermo calcs
+    result <- calcThermoVars(x$matchDepth, x$water_temp)
+    # get specific values
+    result$temp400 <- x$water_temp[x$matchDepth == 400]
+    result$sal400 <- x$salinity[x$matchDepth == 400]
+    result
+}))
+  bindr <- cbind(get(paste0("tracks_SS_", substr(t, 11, 12))), thermData)
+  assign(paste0("tracks_SS_", substr(t, 11, 12)), bindr) #fix this part?
+}
+```
+
+```{r}
+test08 <- matchEnvData(as.data.frame(tracks_SS_08), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
+
+thermData08 <- bind_rows(lapply(test08, function(x) {
+    # do thermo calcs
+    result <- calcThermoVars(x$matchDepth, x$water_temp)
+    # get specific values
+    result$temp400 <- x$water_temp[x$matchDepth == 400]
+    result$sal400 <- x$salinity[x$matchDepth == 400]
+    result
+}))
+
+tracks_SS_08 <- cbind(tracks_SS_08, thermData08)
+```
+
+```{r}
+#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm0.nc")
+
+#test <- matchEnvData(as.data.frame(tracks_SS_07), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
+
+#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm1.nc")
+
+#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm2.nc")
+
+#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm3.nc")
+
+#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm4.nc")
+
+#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm5.nc")
+
+#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm6.nc")
+
+#hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm7.nc")
+
+#hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm8.nc")
+
+# Run the 9 nc lists that make up tracks_SS_08
+#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm0.nc")
+#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm1.nc")
+#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm2.nc")
+#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm3.nc")
+#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm4.nc")
+#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm5.nc")
+#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm6.nc")
+#hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm7.nc")
+#hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm8.nc")
+```
+
 Use above function on the smaller hycom datasets (0-6)
 ```{r}
-#calcThermoVars(hycomData[[1]]$matchDepth, hycomData[[1]]$water_temp, plot=TRUE)
+#Optional plot:
+calcThermoVars(test[[1]]$matchDepth, test[[1]]$water_temp, plot=TRUE)
 
@@
 View(tracks_SS_08)
+```
+```{r}
+thermData <- bind_rows(lapply(test, function(x) {
+    # do thermo calcs
+    result <- calcThermoVars(x$matchDepth, x$water_temp)
+    # get specific values
+    result$temp400 <- x$water_temp[x$matchDepth == 400]
+    result$sal400 <- x$salinity[x$matchDepth == 400]
+    result
+}))
+
+test_fin <- cbind(tracks_SS_07, thermData)
+print(data)
+```
+testing hycom bind on ONE drift first
+```{r}
+test22 <- matchEnvData(as.data.frame(tracks_SS_22), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
+
+thermData22 <- bind_rows(lapply(test22, function(x) {
+    # do thermo calcs
+    result <- calcThermoVars(x$matchDepth, x$water_temp)
+    # get specific values
+    result$temp400 <- x$water_temp[x$matchDepth == 400]
+    result$sal400 <- x$salinity[x$matchDepth == 400]
+    result
+}))
+
+test_fin <- cbind(tracks_SS_22, thermData22)
+View(test_fin)
+```
+
+```{r}
+saveRDS(tracks_SS_10, 'data/fulltrack10.rda')
+saveRDS(tracks_SS_12, 'data/fulltrack12.rda')
+saveRDS(tracks_SS_13, 'data/fulltrack13.rda')
+saveRDS(tracks_SS_14, 'data/fulltrack14.rda')
+saveRDS(tracks_SS_16, 'data/fulltrack16.rda')
+saveRDS(tracks_SS_18, 'data/fulltrack18.rda')
+saveRDS(tracks_SS_19, 'data/fulltrack19.rda')
+saveRDS(tracks_SS_20, 'data/fulltrack20.rda')
+saveRDS(tracks_SS_21, 'data/fulltrack21.rda')
+saveRDS(tracks_SS_22, 'data/fulltrack22.rda')
+saveRDS(tracks_SS_23, 'data/fulltrack23.rda')
+saveRDS(tracks_SS_07, 'data/fulltrack07.rda')
+saveRDS(tracks_SS_08, 'data/fulltrack08.rda')
 ```
@@
 
+Set Parameters
+```{r}
+theme_set(theme_bw())
+tmap_mode("plot")
+```
+
+Create sf objects and basemaps for maps (not needed for environmental variables)
+```{r}
+whales_8_10sf <- BWsf %>%                            
+dplyr::filter(station == "8" | station == "10")
+
+trackssf <- st_as_sf(tracks, coords = c("Longitude", "Latitude"), crs=4326)
+trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326)
+trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326)
+trackBase <- get_tiles(whales_8_10sf, provider = "Esri.WorldShadedRelief")
+allTrackbase <- get_tiles(trackssf, provider = "Stamen.TonerLite", zoom = 5)
+```
+
 Map of whole study area
@@
 ```
+
+```{r}
+# Plots for coastline and continental slope. Do I keep?
+sf08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
+sf10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
+trkraster08 <- rasterize(vect(sf08), agg)
+trkraster10 <- rasterize(vect(sf10), agg)
+
+bathy_df <- as.data.frame(bathy, xy = TRUE)
+
+# Map the coastline and cont. slope
+cont <- st_as_sf(as.contour(bathy, ))
+coast <- cont %>% filter(level == 0)
+cont_slope <- cont %>% filter(level == -2000) #Is 2000m the correct depth for continental shelf?
+cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
+plot(agg)
+plot(cont_slope_rast, col = "red")
+cont_slope_rast <- rasterize(cont_slope, agg)
+plot(cont_slope_rast, col = "red")
+
+ggplot() + 
+  geom_sf(data = cont, col="black") + 
+  geom_sf(data = coast, col = "red") +
+  geom_sf(data = cont_slope, col = "purple") +
+  #geom_sf(data = whales_8_10sf, mapping = aes()) +
+  scale_color_gradient(low="green", high="red")
+```

### 2023-05-11 data/tracks.R
edited track R file

@@ -1,0 +1,23 @@
+## code to prepare in `tracks` dataset goes here
+
+library(tidyverse)
+
+# Load raw rda file
+load("data-raw/CCES2018_DriftTracks_Modified_03Nov2022.rda")
+
+# Read in tracks, filter out lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17)
+tracks <- AllTracks %>%
+  dplyr::select(-dist, -speed) %>%
+  mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>%      # round to nearest 20 minutes, will create duplicated in the data
+  filter(!station %in% c('1','2','3','4','5','6','9','11', '17')) %>%
+  rename(Longitude = long,                                 # Need these columns to specifically say this for ERDDAP matching
+         Latitude = lat,
+         UTC = dateTime)
+tracks <- tracks[!duplicated(tracks[c('UTC', 'station')]),]    # Remove duplicates
+
+# Save as RDS
+saveRDS(tracks, 'data/tracks.rda')
+
+# Filter by buoy
+#tracks_08 <- tracks %>% dplyr::filter(station == "8") 
+track4 <- AllTracks %>% dplyr::filter(station == "4")

### 2023-05-11 data/soundscape_metrics.R
cleaned up file

@@ -1,31 +1,33 @@
 ## code to prepare `soundscape` dataset goes here
 
-#NEED TO MAKE THIS A FUNCTION
-# look at http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops for help on automation.
-
 library(tidyverse)
 library(lubridate)
+library(here)
 
-
-#dlist <- c(4,7,8,10,12,13,14,16,17,18,19,20,21,22,23)
+# Path to folder in Project/on GitHub
 sslist <- list.files(path = paste0("data-raw/"), pattern = "CCES_", recursive = TRUE)
 
-#code doesnt work, can delete
+# Path to folder on my external hard drive
+dirlist <- dir("D:/Soundscape_metrics", recursive=TRUE, full.names=TRUE, pattern="CCES_")
+
+# DO I NEED THIS SECTION??? Read through soundscape csv files, change date format
 #for (i in sslist) {
-#  read_csv(paste0("data-raw/",i), show_col_types = FALSE)%>%
-#  rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
-#  mutate(dateTime = round_date(ymd_hms(.$dateTime),"20 minutes")) %>%
-#  assign(., value = paste0((substr(sslist[i], 16, 17)), substr(sslist[i], 6, 7)))
+#  data <- read_csv(here(paste0("data-raw/", i)), show_col_types = FALSE) %>%
+#    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
+#    mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
+#    mutate(variable = paste0(substr(i, 16, 17), substr(i, 6, 7)))
+  
+#  assign(paste0(substr(i, 16, 17), "_", substr(i, 6, 7)), data)
 #}
 
-#From ChatGPT, edited from above
-for (i in sslist) {
-  data <- read_csv(here(paste0("data-raw/", i)), show_col_types = FALSE) %>%
+# Read though files from external hard drive
+for (i in dirlist) {
+  data <- read_csv(i, show_col_types = FALSE) %>%
     rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
     mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
-    mutate(variable = paste0(substr(i, 16, 17), substr(i, 6, 7)))
[…]

### 2023-05-11 03_environ_var.Rmd
renamed to environ_var, removed code with plots

--- 2023-05-11 03_exploratory.Rmd
@@
 ---
-title: "environmenal/exploratory analysis"
+title: " extracting environmenal and oceanographic variables"
 output: html_document
@@
 
-##FIX THIS - try testing on ONE buoy
+## This only downloads 11 of 13 successfully - FIX
 for (t in tracklist) {
@@
   bindr <- cbind(get(paste0("tracks_SS_", substr(t, 11, 12))), thermData)
-  assign(paste0("tracks_SS_", substr(t, 11, 12)), bindr) #fix this part?
+  assign(paste0("tracks_SS_", substr(t, 11, 12)), bindr)
 }
@@
 ```{r}
+## HYCOM function above stops at buoys 07 and 08, need to manually run it here. FIX!!
 test08 <- matchEnvData(as.data.frame(tracks_SS_08), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
@@
 
-```{r}
-#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm0.nc")
-
-#test <- matchEnvData(as.data.frame(tracks_SS_07), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
-
-#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm1.nc")
-
-#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm2.nc")
-
-#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm3.nc")
-
-#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm4.nc")
-
-#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm5.nc")
-
-#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm6.nc")
-
-#hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm7.nc")
-
-#hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "data-raw/hycomm8.nc")
-
-# Run the 9 nc lists that make up tracks_SS_08
-#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm0.nc")
-#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm1.nc")
-#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm2.nc")
-#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm3.nc")
-#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm4.nc")
-#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm5.nc")
-#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm6.nc")
-#hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm7.nc")
-#hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "data-raw/hycomm8.nc")
-```
-
-Use above function on the smaller hycom datasets (0-6)
-```{r}
-#Optional plot:
-calcThermoVars(test[[1]]$matchDepth, test[[1]]$water_temp, plot=TRUE)
-
-thermData0 <- bind_rows(lapply(hycomData0, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData1 <- bind_rows(lapply(hycomData1, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData2 <- bind_rows(lapply(hycomData2, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData3 <- bind_rows(lapply(hycomData3, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData4 <- bind_rows(lapply(hycomData4, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData5 <- bind_rows(lapply(hycomData5, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData6 <- bind_rows(lapply(hycomData6, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData7 <- bind_rows(lapply(hycomData7, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData8 <- bind_rows(lapply(hycomData8, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-# Bind all thermocline datasets together
-thermData <- rbind(thermData0, thermData1, thermData2, thermData3, thermData4, 
-                   thermData5, thermData6, thermData7, thermData8)
-
-# Bind main thermocline dataset to main track/soundscape dataframe
-tracks_SS_08 <- cbind(tracks_SS_08, thermData)
-
-# New tracks_SS_08 variable contains Hycom variables now
-##### FINAL tracks_ss_08 dataframe ######
-View(tracks_SS_08)
-```
-```{r}
-thermData <- bind_rows(lapply(test, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-test_fin <- cbind(tracks_SS_07, thermData)
-print(data)
-```
-testing hycom bind on ONE drift first
-```{r}
-test22 <- matchEnvData(as.data.frame(tracks_SS_22), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
-
-thermData22 <- bind_rows(lapply(test22, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-test_fin <- cbind(tracks_SS_22, thermData22)
-View(test_fin)
-```
-
+Save final buoy df for analysis use next.
 ```{r}
@@
 ```
-
-Get avg daily environmental variables
-```{r}
-means_08 <- tracks_SS_08 %>%
-  group_by(as.Date(UTC))%>%
-  summarize(sst_mean = mean(sst_mean),
-            dist_slope = mean(dist2slope),
-            depth = mean(depth),
-            slope = mean(slope),
-            dist_shore = mean(dist2shore),
-            chlorophyll_mean = mean(chlorophyll_mean),
-            ssh_mean = mean(ssh_mean),
-            mldDepth = mean(mldDepth),
-            mldTemp = mean(mldTemp),
-            ttDepth = mean(ttDepth),
-            ttTemp = mean(ttTemp),
-            temp400 = mean(temp400),
-            sal400 = mean(sal400),
-            BWpresence = sum(BWpresence))
-```
-
-### Exploratory Analysis ######
-
-### Third octave levels histograms
-
-```{r}
-# Drift 8 - all TOL bands, facet wrap
-pivot <- tracks_SS_08 %>%
-  pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of third octave level intensities: Drift 8") +
-  labs(x = "Sound Intensity (dB)") +
-  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
-  scale_fill_manual(values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.7,0.8)) +
-  facet_grid(~SL)
-
-# Drift 10 - all TOL bands, facet grid
-pivot2 <- tracks_SS_10 %>%
-  pivot_longer(cols = starts_with("TOL")| starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of third octave level intensities: Drift 10") +
-  scale_x_continuous(name = "Sound Intensity (dB)", breaks=seq(50,110,20)) +
-  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
-  scale_fill_manual(values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.76)) +
-  facet_grid(~SL)
-
-# Drift 8 - TOL bands 125 and 2000 only
-pivot3 <- tracks_SS_08 %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz")
-```
-
-```{r drift-8-histogram}
-ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 8") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Beaked Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
-```
-
-```{r}
-# Drift 10 - TOL bands 125 and 2000 only
-pivot4 <- tracks_SS_10 %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot4, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 10") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Beaked Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
-
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz")
-```
-
-Scatterplots
-
-```{r}
-# SST vs TOL 2000
-ggplot(data = tracks_SS_08, aes(x = sst_mean, y = `TOL_2000`)) +
-  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
-
-# SST over time
-ggplot(tracks_SS_08, aes(x = UTC, y = sst_mean)) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-  #geom_smooth(method = "loess")
-
-# Depth vs TOL 125
-ggplot(data = tracks_SS_08, aes(x = depth, y = `TOL_125`)) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-
-#Mean depth over time
-ggplot(means_08, aes(x = `as.Date(UTC)`, y = depth*(-1))) +
-  geom_line()
-
-# Chlorophyll vs TOL 125
-ggplot(data = tracks_SS_08, aes(x = chlorophyll_mean, y = `TOL_125`)) +
-  geom_point(aes(color = species))
-
-# CHl over time
-ggplot(tracks_SS_08, aes(x = UTC, y = chlorophyll_mean)) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-  #geom_smooth(method = "loess")
-
-# Mixed Layer Depth vs TOL125 with species
-ggplot(data = tracks_SS_08, aes(x = mldDepth, y = `TOL_125`)) +
-  geom_point(aes(color = species))
-
-# Mixed Layer Depth over time
-ggplot(tracks_SS_08, aes(x = UTC, y = mldDepth*(-1))) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-
-ggplot(data = tracks_SS_08, aes(x = slope, y = `TOL_5000`)) +
-  geom_point(aes(color = species))
-```
-
-Facet Grids for Env. Variables
-```{r}
-#For surface/depth variables USING DAILY MEAN VARIABLES
-# See above for means_08 derivation
-track08Long <- means_08 %>%
-  pivot_longer(cols = sst_mean:ssh_mean,
-               names_to="parameter",
-               values_to="value") %>%
-  filter(parameter %in% c("sst_mean", "dist_shore", "chlorophyll_mean", "ssh_mean")) %>%
-  rename(Date = `as.Date(UTC)`)
-
-p <- ggplot(data = track08Long, aes(x = Date, y = value)) +
-  geom_line()
- # geom_point(aes(size = BWpresence))   ### How best to represent number of BW detections per day?
-
-p + facet_grid(parameter ~ ., scales = "free_y")
-
-# For MLD/Thermocline variables 
-track08Long2 <- tracks_SS_08 %>%
-  pivot_longer(cols = mldDepth:sal400,
-               names_to="parameter",
-               values_to="value") %>%
-  filter(parameter %in% c("mldDepth", "mldTemp", "ttDepth", "ttTemp", "temp400", "sal400")) %>%
-    rename(Date = UTC) %>%
-    subset(species!="?BW")
-
-q <- ggplot(data = track08Long2, aes(x = Date, y = value)) +
-  geom_line() + geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
-q + facet_grid(parameter ~ ., scales = "free_y")
-
-# For temp400 and salinity400 variables
-env.labs <- c(`sal400` = "Salinity at 400m", 
-              `temp400` = "Temperature at 400m")
-
-track08Long3 <- tracks_SS_08 %>%
-  pivot_longer(cols = temp400:sal400,
-               names_to="parameter",
-               values_to="value") %>%
-  filter(parameter %in% c("temp400", "sal400")) %>%
-  rename(`Date (2018)` = UTC)
-
-q <- ggplot(data = track08Long3, aes(x = `Date (2018)`, y = value)) +
-  geom_line() + 
-  geom_point(aes(color = species)) + 
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43", "BW" = "BW"))
-q + facet_grid(parameter ~ ., scales = "free_y", labeller = as_labeller(env.labs))
-```
-
-Co-occurrence plots of drift 8 and vessels 
-```{r}
-#### 09/04/18 co-occurrence of vessel and drift within 10 km
-track08_09.04 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-04 00:00:00" & UTC <= "2018-09-05 00:00:00")
-
-ggplot(data = track08_09.04, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
-  geom_text(aes(as.POSIXct("2018-09-04 14:13:00"), 88, label = "Enter vessel", hjust = 1.0)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-04 15:17:00"), color = "yellow", linewidth = 2) +
-  geom_text(aes(as.POSIXct("2018-09-04 15:13:00"), 87, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.04, aes(color = species, size = 2)) +
-  scale_color_discrete(na.translate=FALSE) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-04-18\n(Within 7 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/05/18 co-occurrence of vessel and drift within 10 km.
-track08_09.05 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-04 21:00:00" & UTC <= "2018-09-06 00:00:00")
-
-ggplot(data = track08_09.05, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-05 01:33:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-05 01:33:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-05 02:08:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-05 02:08:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.05, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-05-18\n(Within 5 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/10/18 co-occurrence of vessel and drift within 10 km.
-track08_09.10 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-10 00:00:00" & UTC <= "2018-09-11 00:00:00")
-
-ggplot(data = track08_09.10, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-10 12:58:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-10 12:58:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-10 14:01:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-10 14:01:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() +
-  geom_point(data = track08_09.10, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-10-18\n(Within 2 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/22/18 co-occurrence of vessel and drift within 10 km
-track08_09.22 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-22 00:00:00" & UTC <= "2018-09-23 00:00:00")
-
-ggplot(data = track08_09.22, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-22 19:13:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-22 19:13:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-22 19:34:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-22 19:34:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.22, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-22-18\n(Within 9 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/23/18 co-occurrence of vessel and drift within 10 km. NO whale detections in this time period.
-track08_09.23 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-23 00:00:00" & UTC <= "2018-09-24 00:00:00")
-
-ggplot(data = track08_09.23, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-23 03:55:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-23 03:55:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-23 04:23:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-23 04:23:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-23-18\n(Within 6 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-```
-
-BW Occurrences
-```{r}
-# Filter Drift 8 by BW presence and absence
-BW1 <- tracks_SS_08 %>%
-  filter(BWpresence == 1)
-BW0 <- tracks_SS_08 %>%
-  filter(BWpresence == 0)
-
-ggplot(data = tracks_SS_08, mapping = aes(TOL_125)) + geom_histogram(binwidth = 0.1, color = "black")
-
-ggplot() + 
-  geom_histogram(data = BW0, aes(TOL_5000), binwidth = 0.5, fill = "black", color = "white") +
-  geom_histogram(data = BW1, aes(TOL_5000), binwidth = 0.5, fill = "yellow", color = "black")
-
-# Percent of time BWs are detected
-print(paste(nrow(BW1) / nrow(tracks_SS_08) * 100,"%"))
-
-# Most abundant BW species when whale is detected (remove NA values)
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot() + geom_bar(aes(species))
-
-# Sound levels BW are exposed to by species
-ZC <- tracks_SS_08 %>%
-  filter(species == "ZC")
-
-BW43 <- tracks_SS_08 %>%
-  filter(species == "BW43")
-
-print(max(ZC$TOL_125))
-print(max(BW43$TOL_125))
-
-ggplot() + geom_histogram(data = ZC, mapping = aes(TOL_125), binwidth = 0.2, fill = "brown", color = "white") +
-  geom_histogram(data = BW43, mapping = aes(TOL_125), binwidth = 0.2, fill = "yellow", color = "white") +
-  theme_classic()
-```
-
-Density Plots - Species at TOL sound levels
-```{r}
-# Broadband
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(`BB_20.24000`, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("Broadband") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 63
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_63, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("63 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 125
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_125, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("125 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 2000
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_2000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("2000 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 5000
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_5000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("5000 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 20,000
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_20000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("20,000 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-```
-
-Histograms with BW presence/absence and environmental variables
-```{r}
-# SST
-ggplot(data = tracks_SS_08, aes(sst_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.1, color = "black")
-
-# Depth
-ggplot(data = tracks_SS_08, aes(depth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 10, color = "black")
-
-# Bottom slope
-ggplot(data = tracks_SS_08, aes(slope, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
-
-# SSH
-ggplot(data = tracks_SS_08, aes(ssh_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
-
-# Chlorophyll
-ggplot(data = tracks_SS_08, aes(chlorophyll_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
-
-# Mixed layer depth Depth
-ggplot(data = tracks_SS_08, aes(mldDepth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.25, color = "black")
-
-# Mixed layer depth Temperature
-ggplot(data = tracks_SS_08, aes(mldTemp, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.1, color = "black")
-
-# Thermocline Depth
-ggplot(data = tracks_SS_08, aes(ttDepth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.25, color = "black")
-
-#Thermocline temperature
-ggplot(data = tracks_SS_08, aes(ttTemp, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.05, color = "black")
-
-# Temperature at 400m
-ggplot(data = tracks_SS_08, aes(temp400, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.01, color = "black")
-
-# Salinity at 400 m
-ggplot(data = tracks_SS_08, aes(sal400, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.001, color = "black")
-```
-
-### Plot maps for Powerpoint images
-
-Set Parameters
-```{r}
-theme_set(theme_bw())
-tmap_mode("plot")
-```
-
-Create sf objects and basemaps for maps (not needed for environmental variables)
-```{r}
-whales_8_10sf <- BWsf %>%                            
-dplyr::filter(station == "8" | station == "10")
-
-trackssf <- st_as_sf(tracks, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326)
-trackBase <- get_tiles(whales_8_10sf, provider = "Esri.WorldShadedRelief")
-allTrackbase <- get_tiles(trackssf, provider = "Stamen.TonerLite", zoom = 5)
-```
-
-Map of whole study area
-```{r}
-# Establish bounding box
-bbox <- st_bbox(c(xmin = -110, 
-            xmax = -128, 
-            ymax = 44, 
-            ymin = 30), crs = st_crs(4326))
-
-#Establish basemap
-allTrackbase <- get_tiles(bbox, provider = "Esri.WorldGrayCanvas", zoom = 5)
-
-# Establish inset map
-us_region <- st_bbox(c(xmin = -90,
-            xmax = -120, 
-            ymax = 48, 
-            ymin = 25), crs = st_crs(4326))
-usa <- get_tiles(us_region, provider = "Esri.WorldGrayCanvas", zoom = 4)
-
-sg <- bb_poly(bb(tracks_sf))
-inset <- tm_shape(usa) + tm_rgb() + tm_shape(sg) + tm_borders(col = "red")
-
-#Map all tracks
-main <- tm_shape(allTrackbase) + tm_rgb() +
-  tm_shape(trackssf) +
-  tm_grid(alpha = 0.6, col = "grey") +
-  tm_dots(col = "darkblue", size = 0.001, legend.show = TRUE) +
-  tm_scale_bar(width = 0.3, text.size = 3, position = c(0.2, 0.03)) +
-  tm_compass(type = "arrow", position = c(0.05, 0.05), color.light = "white") +
-  tm_add_legend('symbol',
-                type = "line",
-                col = "darkblue",
-                labels = "Drifting buoy tracks",
-                size = 1,
-                lwd = 2)
-
-vp <- viewport(x=0.66, y=0.8, width = 0.26, height=0.4)
-main
-print(inset,vp=vp)
-
-# Map all tracks - ggplot
-basemap2 <- as.data.frame(allTrackbase, xy = TRUE)
-#ggplot() +
-#  geom_tile(data = basemap, aes(x=x, y=y, fill = "basemap")) +
-#  geom_sf(data = trackssf, mapping = aes(), color = "red")
-
-# Map tracks 08 and 10 only
-tm_shape(trackBase) + tm_rgb() + 
-  tm_shape(trk_TOL_08) + tm_dots(col = "red", 
-                                 legend.show = TRUE) +
-  tm_shape(trk_TOL_10) + tm_dots(col = "green")
-
-# Bounding box of drift tracks on map
-lon2 <- -129.04666
-lon1 <- -115.78247
-lat1 <- 28.21790
-lat2 <- 45.09226
-
-basemap3 <- ne_countries(scale = "medium", type = "map_units", returnclass = "sf")
-
-# Drift tracks
-ggplot(data = basemap3) +
-  geom_sf(fill = "antiquewhite") +
-  coord_sf(xlim = c(lon2 - 1, lon1 + 1), ylim = c(lat1 - 1, lat2 + 1), expand = FALSE) +
-  theme(panel.grid.major = element_blank()) +
-  geom_path(data = tracks_sf, mapping = aes(x=long,y=lat, group = factor(station)), color = "deeppink") +
-  theme(panel.background = element_rect(fill = "aliceblue"), 
-        axis.text = element_blank(), axis.ticks = element_blank(), 
-        axis.title = element_blank())
-```
-
-Map of soundscapes
-```{r}
-# CCES_08 + CCES_10 Broadband
-tm_shape(trackBase) + tm_rgb() + 
-  tm_shape(trk_TOL_08) + 
-  tm_symbols(col = "BB_20-24000", 
-             border.lwd = NA, 
-             size = 0.4, 
-             sizes.legend = 0.7, 
-             n = 10) +
-  tm_shape(trk_TOL_10) + 
-  tm_symbols(col = "BB_20-24000",
-             border.lwd = NA, 
-             size = 0.4,
-             n = 10) +
-  tm_layout(legend.width = 0.5, 
-            legend.position = c("right","top"),
-            legend.bg.color = "lightblue")
-
-#CCES_08 Third Octave Level
-tm_shape(trackBase) + tm_rgb() + 
-  tm_shape(trk_TOL_08) + 
-  tm_symbols(col = "TOL_63", 
-             border.lwd = NA, 
-             size = 0.5, 
-             n = 10) + 
-  tm_layout(legend.width = 0.5, 
-            legend.position = c("right","top")) + 
-  tm_legend()
-
-ggplot() + 
-  geom_sf(data = trk_TOL_08, aes(color = TOL_125, size = TOL_125), show.legend = FALSE) +
-  geom_sf(data = trk_TOL_10, aes(color = TOL_125, size = TOL_125)) +
-  scale_x_continuous(limits = c(-130,-122)) + 
-  scale_y_continuous(limits = c(35,39)) +
-  scale_color_gradient(low = "navy", high = "yellow")+
-  scale_size(range = c(1,12)) +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(panel.border = element_rect(fill = NA),
-        axis.text = element_blank(),
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        legend.position = "none")
-```
-
-Map of whale detections
-```{r}
-# Whale detections
-ggplot() +
-  geom_point(data = whales8, mapping = aes(x=long, y=lat), shape = 21, color = "black", fill = "green4", size = 4) +
-  geom_point(data = whales10, mapping = aes(x=long, y=lat), shape = 21, color = "black", fill = "green4", size = 4) +
-  scale_x_continuous(limits = c(-130,-122)) + 
-  scale_y_continuous(limits = c(35,39)) +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(panel.border = element_rect(fill = NA),
-        axis.text = element_blank(),
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank()
-        )
-#ggsave("BWdetect.png")
-
-```
-
-Map of bathymetry
-```{r}
-# Bathymetry
-bathym <- ggplot() +
-    geom_raster(data = bathy_df, 
-              mapping = aes(x=x, y=y,
-                            fill = `gebco_bathy`)) +
-  annotation_scale(location = "bl", width_hint = 0.4) +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA))
-
-# Alternative Bathymetry (code from Ocean Tracking Network #https://www.youtube.com/watch?v=xT2F-Y9bKLk)
-bgo <- getNOAA.bathy(lon1 = -122, lon2 = -130,
-                     lat1 = 35, lat2= 39, resolution = 1)  # plot bounding box around drift 8 and 10
-autoplot(bgo, geom = c("raster", "contour"), coast = TRUE, show.legend = FALSE) + 
-  scale_fill_etopo() + #creates ggplot object
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank())
-
-# Soundscape + bathymetry
-whales08_sf <- BWsf %>%                            
-  dplyr::filter(station == "8")
-
-ggplot() +
-  geom_raster(data = bathy_df, 
-              mapping = aes(x=x, y=y,
-                            fill = `gebco_bathy`), show.legend = FALSE) +
-  #geom_contour(data = bathy_df, mapping = aes(x=x, y=y, z=gebco_bathy, linewidth = 0.1), color = "blue", linewidth = 0.01,) +
-  geom_sf(data = trk_TOL_08, aes(color = TOL_125, size = TOL_125), show.legend = FALSE) +
-  geom_sf(data = whales08_sf, aes(), color = "black") +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  scale_color_gradient(low = "beige", high = "hotpink") +
-  scale_fill_etopo() +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA))
-#ggsave("soundscape.png")
-```
-
-Map of AIS
-```{r}
-# Plot + crop to drift area
-#basemap <- ne_countries(scale = "medium", returnclass = "sf")
-#ggplot() + 
-#  geom_sf(aes(), basemap) +
-#  geom_path(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName), #show.legend = F) + 
-  #geom_point(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName)) +
-#  scale_x_continuous(limits = c(min(drift$long-0.7), max(drift$long + 0.7))) + 
-#  scale_y_continuous(limits = c(min(drift$lat-0.7), max(drift$lat + 0.7)))
-
-# AIS data
-ggplot() + 
-  geom_path(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName, linewidth = 1), show.legend = F) +
-  scale_x_continuous(limits = c(-131,-121)) + 
-  scale_y_continuous(limits = c(34,40)) +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA))
-#ggsave("AIS.png")
-```
-
-Map of Sea Surface Temp with rerddap, Drift 8 only
-```{r}
-# Multi-scale Ultra-high Resolution (MUR) SST Analysis fv04.1, Global, 0.01°, 2002-present, Daily
-SSTInfo <- info('jplMURSST41') 
-murSST <- griddap(SSTInfo, latitude = c(35, 39), longitude = c(-122, -130), 
-                  time = c("2018-08-16", "2018-10-01"), 
-                  fields = 'analysed_sst') # drift 8 dates only
-SST <- murSST$data
-
-# Raster of SST
-ggplot() + 
-  geom_tile(data = SST, aes(x = longitude, y = latitude, fill = analysed_sst)) +
-  scale_fill_gradient(low = "darkblue", high = "yellow") +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(axis.text = element_blank(),
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA),
-        legend.position = "none")
-```
-
-Plot distance to continental slope
-```{r}
-ggplot() +
-  geom_raster(data = bathy_df, 
-              mapping = aes(x=x, y=y,
-                            fill = `gebco_bathy`)) +
-  geom_sf(data = tracks_SS_08, 
-          mapping = aes(col = dist_slope)) +
-  geom_sf(data = cont_slope) + scale_fill_continuous(high = "lightgreen", low = "blue")
-
-# PLOT distance by TOL with whales
-ggplot(data = tracks_SS_08, aes(x = dist_slope, y = `TOL_125`)) + 
-  geom_point() + 
-  geom_smooth(color = "green") +
-  theme_classic()
-```
-
-Another map - need to decide if important to keep
-```{r}
-# Plot
-oceanBase <- get_tiles(tracks_SS_08, provider="Esri.NatGeoWorldMap", crop = TRUE)
-tm_shape(oceanBase) + tm_rgb() +
-  tm_shape(tracks_SS_08) +  tm_symbols(col="species", size = 1) +
-  tm_layout(legend.width = 0.5, 
-            legend.position = c("right","top"))
-
-#Raster base map
-oceanBase <- get_tiles(tracks_SS_10, provider="Esri.NatGeoWorldMap")
-tm_shape(bathy) + tm_raster() +
-  tm_shape(tracks_SS_10) + 
-  tm_symbols(style = "jenks") +
-  tm_layout(legend.width = 0.5)
-```
-
-```{r}
-# Plots for coastline and continental slope. Do I keep?
-sf08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-sf10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-trkraster08 <- rasterize(vect(sf08), agg)
-trkraster10 <- rasterize(vect(sf10), agg)
-
-bathy_df <- as.data.frame(bathy, xy = TRUE)
-
-# Map the coastline and cont. slope
-cont <- st_as_sf(as.contour(bathy, ))
-coast <- cont %>% filter(level == 0)
-cont_slope <- cont %>% filter(level == -2000) #Is 2000m the correct depth for continental shelf?
-cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
-plot(agg)
-plot(cont_slope_rast, col = "red")
-cont_slope_rast <- rasterize(cont_slope, agg)
-plot(cont_slope_rast, col = "red")
-
-ggplot() + 
-  geom_sf(data = cont, col="black") + 
-  geom_sf(data = coast, col = "red") +
-  geom_sf(data = cont_slope, col = "purple") +
-  #geom_sf(data = whales_8_10sf, mapping = aes()) +
-  scale_color_gradient(low="green", high="red")
-```

### 2023-05-11 01_readin.R
removed commented code (unnecessary)

@@ -1,16 +1,11 @@
 ## code to assemble clean dataframe for each buoy.
 ## Prerequisites include whales.R, tracks.R, soundscape_metrics.R.
 
-# Look into https://bookdown.org/yihui/rmarkdown-cookbook/managing-projects.html for info on 
-# managing projects with multiple Rmds and scripts
 
 # libraries ##########
 
 library(tidyverse)
 library(lubridate)
-library(sf)
-library(rnaturalearth)
-library(tmap)
 library(here)
 
 # set up #####
@@ -19,20 +14,14 @@ library(here)
 Sys.setenv(TZ='UTC')
 here()
 
-# Clear global environment if needed
-# rm(list=ls()) 
-
-
 ######################
 # Join TRACKS with WHALES (Start Rmd here)
 ######################
 tracks <- readRDS('data/tracks.rda')
 whales <- readRDS('data/whales.rda')
-# Read in EACH BB and TO rda file??
-# Should I add code to remove file if it exists?
 
-trk_whale <- left_join(tracks, whales, by = c("station","UTC")) #%>%
-#  filter(!station %in% c('4','17'))  ## Can filter out buoys 4 and 17 here
+# Join tracks and whales
+trk_whale <- left_join(tracks, whales, by = c("station","UTC"))
 
 # Remove duplicates
 trk_whale <- trk_whale[!duplicated(trk_whale[c('UTC', 'station')]),]
@@ -79,10 +68,8 @@ joinTable <- function(w, s, t) {
                 TOL_5000, TOL_20000, species, BWpresence)
 }
  
-#m <- left_join(trk_whales7, BB_07, by = c("UTC" = "dateTime")) %>%
-#  .[!duplicated(.['UTC']),]
 
-# join whales detections, soundscape metrics to each buoy track
+# join whales detections + soundscape metrics (BB and TO) to each buoy track
 # How can I loop this?
 tracks_SS_07 <- joinTable(trk_whales7, BB_07, TO_07)
 tracks_SS_08 <- joinTable(trk_whales8, BB_08, TO_08)
@@ -98,25 +85,3 @@ tracks_SS_21 <- joinTable(trk_whales21, BB_21, TO_21)
 tracks_SS_22 <- joinTable(trk_whales22, BB_22, TO_22)
 tracks_SS_23 <- joinTable(trk_whales23, BB_23, TO_23)
 
-
-
-#Plots
-worldmap <- ne_countries(scale = 'medium', type = 'map_units',
[…]

### 2023-05-11 04_analysis.Rmd
added code to read in rds files

--- 2023-04-19 03_exploratory.Rmd
@@
 ---
-title: "exploratory analysis"
+title: "04_analysis"
+author: "Marina Bozinovic"
+date: "2023-05-11"
 output: html_document
-date: "2022-11-06"
 ---
@@
 Required Packages
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(tidyverse)
-library(lubridate)
-library(sf)
-library(rerddap)
-library(terra)
-library(marmap)
-library(rnaturalearth)
-library(mapproj)
-library(raster)
-#library("rerddapXtracto")
-library(PAMpal)
-library(PAMmisc)
-library(maptiles)
-library(tmap)
-library(spData)
-library(tmaptools)
-library(grid)
-library(ggspatial)
+```{r}
+# add packages here!!
 ```
 
-Set Parameters
+Read in RDS files of each buoy + all environ variables + AIS (pending)
 ```{r}
-theme_set(theme_bw())
-tmap_mode("plot")
-```
+ftlist <- dir("data/", recursive=TRUE, full.names=TRUE, pattern="fulltrack")
 
-Create sf objects and basemaps for maps
-```{r}
-whales_8_10sf <- BWsf %>%                            
-dplyr::filter(station == "8" | station == "10")
+for (ft in ftlist) {
+  df <- readRDS(ft)
+  assign(paste0("fulltrack", substr(ft, 15, 16)), df)
+}
 
-trackssf <- st_as_sf(tracks, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326)
-trackBase <- get_tiles(whales_8_10sf, provider = "Esri.WorldShadedRelief")
-allTrackbase <- get_tiles(trackssf, provider = "Stamen.TonerLite", zoom = 5)
-```
-### Exploratory analysis
-
-SST from PAMPal
-```{r}
-# Downloads data from servers, uncomment and only run once
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "analysedSST")
-
-# from file 
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/analysedSST.nc') %>%
-  rename(., sst_mean = analysed_sst_mean) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)   # rename new column to "sst_mean" and remove unnecessary columns
-```
-
-Distance to continental slope
-```{r}
-# Increase cell size by 30x to be visible
-bathy <- rast("data/gebco_bathy.tif")
-agg <- terra::aggregate(bathy, 5)
-plot(agg)
-
-# Rasterize track with agg as the template
-
-#TOL_rast <- rasterize(trk_TOL_08, agg, field = "TOL_125")
-sf08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-sf10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-trkraster08 <- rasterize(vect(sf08), agg)
-trkraster10 <- rasterize(vect(sf10), agg)
-
-bathy_df <- as.data.frame(bathy, xy = TRUE)
-
-# Map the coastline and cont. slope
-cont <- st_as_sf(as.contour(bathy))
-coast <- cont %>% filter(level == 0)
-cont_slope <- cont %>% filter(level == -2000)
-cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
-plot(cont_slope_rast, col = "red")
-cont_slope_rast <- rasterize(cont_slope, agg)
-plot(cont_slope_rast, col = "red")
-
-#ggplot() + 
-#  geom_raster(data = bathy_df, 
-#              mapping = aes(x=x, y=y, 
-#                            alpha = 0.8,
-#                            fill = `gebco_bathy`))+
-#  geom_sf(data = cont, col="black") + 
-#  geom_sf(data = coast, col = "red") +
-#  geom_sf(data = cont_slope, col = "purple") +
-#  geom_sf(data = whales_8_10sf, mapping = aes()) +
-#  scale_color_gradient(low="green", high="red") 
-```
-
-### Extract slope_dist and bind to drift tracks
-```{r}
-# Drift 8
-slopedist <- terra::distance(trkraster08, vect(cont_slope))     # measure distance from track to slope
-extdist <- terra::extract(slopedist, vect(sf08))       # extract this distance into a df
-tracks_SS_08 <- cbind(sf08,extdist) %>%                     # bind to main df
-  dplyr::select(-ID) %>%
-  rename("dist2slope" = gebco_bathy) %>%
-  st_set_geometry(NULL)                         # Drop geometry to make df for future cbinds
-
-#Drift 10
-slopedist <- terra::distance(trkraster10, vect(cont_slope))     # measure distance from track to slope
-extdist <- terra::extract(slopedist, vect(sf10))       # extract this distance into a df
-tracks_SS_10 <- cbind(sf10,extdist) %>%                     # bind to main df
-  dplyr::select(-ID) %>%
-  #rename("dist2slope" = gebco_bathy) %>%
-  st_set_geometry(NULL)                        # Drop geometry to make df for future cbinds
-```
-
-#### Bathymetry - PAMpal
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'srtm15plus', fileName='depth.nc')
-
-# Pulls from downloaded file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/depth.nc') %>%
-  rename(., depth = z_mean) %>%        # rename new column to "depth"
-  mutate(depth = depth*-1) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-```
-
-#### Bathymetric slope
-
-```{r}
-# Slope (Hard-coded - can fix this)
-bgo <- getNOAA.bathy(lon1 = -121, lon2 = -131,
-                     lat1 = 34, lat2= 40, resolution = 1)  # plot bounding box around drift 8 and 10
-# As raster
-bgo_rast <- marmap::as.raster(bgo)
-bgo_slope <- terrain(bgo_rast, v = "slope")
-
-# Extract slope values at points along drift 8
-slope <- raster::extract(bgo_slope, tracks_SS_08[,4:3])
-tracks_SS_08 <- cbind(tracks_SS_08, slope)
-```
-
-#### Distance from coast - PAMPal
-Need to have variable choose itself!! Also, how do I get it to not download each time?
-```{r}
-#dist2shore <- erddapToEdinfo(dataset='dist2coast_1deg',
-#                             baseurl='https://pae-paha.pacioos.hawaii.edu/erddap/',
-#                             chooseVars = FALSE)
-
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc='data/Dist2Shore.nc') %>%
-  rename(., dist2shore = dist_mean) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-
-#May not need this part
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'dist2coast_1deg', fileName='dist2coast.nc', var = 'dist')
-
-# Pulls from downloaded file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/dist2coast.nc')
-
-# Downloads nc file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=dist2shore, fileName='Dist2Shore.nc')  
-```
-
-### Chlorophyll from PAMPal
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdMH1chlamday', fileName='chlorophyll.nc')
-
-# Pulls from downloaded file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/chlorophyll.nc')
-```
-
-#### Sea Surface Height Anomalies from PAMPal
-```{r}
-# Downloaded from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'nesdisSSH1day', fileName='sshanomalies.nc')
-
-# Pulled from downloaded nc file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/sshanomalies.nc') %>%
-  rename(., ssh_mean = sla_mean) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-```
-
-### Wind Stress Curl from PAMPal *** NEEDS FIXING ****
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQMstress1day_LonPM180', fileName='erdQMstress1day_LonPM180.nc', var = 'curl')
-
-# Pulls from downloaded file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQMstress1day_LonPM180.nc')
-#tracks_SS_08
-
-#windcurl <- erddapToEdinfo('erdQMstress1day', baseurl = 'https://coastwatch.pfeg.noaa.gov/erddap/')
-
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=windcurl, fileName='windcurl.nc')  
-```
-
-#### HyCOM Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
-Doesn't bin itself into separate dataframe based on size!
-```{r}
-# These refer to nc files that have already been downloaded from the server.
-# Uncomment and only run once! Rows are hard-coded.
-
-#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm0.nc")
-
-#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm1.nc")
-
-#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm2.nc")
-
-#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm3.nc")
-
-#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm4.nc")
-
-#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm5.nc")
-
-#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm6.nc")
-
-#hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm7.nc")
-
-#hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm8.nc")
-
-# Run the 9 nc lists that make up tracks_SS_08
-hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm0.nc")
-hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm1.nc")
-hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm2.nc")
-hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm3.nc")
-hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm4.nc")
-hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm5.nc")
-hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm6.nc")
-hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm7.nc")
-hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm8.nc")
-```
-
-Create function to estimate thermocline variables using the temperature at depth data
-```{r}
-calcThermoVars <- function(depth, temp, mldMode='SST8', thermMode='variso', plot=FALSE) {
-    # depthInterp <- approxfun(x=temp, y=depth)
-    switch(match.arg(mldMode),
-           'SST8' = {
-               mldTemp <- temp[1] - 0.8
-               # first few temps are likely to have repeat values that trigger
-               # warnings for approxfun, so cut off to only values after mldTemp
-               useMin <- min(which(temp <= mldTemp)) - 1
-               useIx <- useMin:length(temp)
-               depthInterp <- approxfun(x=temp[useIx], y=depth[useIx])
-               mld <- depthInterp(mldTemp)
-           }
-    )
-    switch(match.arg(thermMode),
-           'variso' = {
-               t400 <- approx(x=depth, y=temp, xout=400)$y
-               # therm end is half way to t400 temp
-               tEnd <- mldTemp + (t400 - mldTemp)/2
-               # therm temp is midpt from mld to end
-               tt <- (mldTemp + tEnd)/2
-               # tt <- mldTemp - .25 * (mldTemp - t400)
-               ttDepth <- depthInterp(tt)
-               tEndDepth <- depthInterp(tEnd)
-           }
-    )
-    if(plot) {
-        to300 <- depth <= 300
-        plot(x=temp[to300], y=-depth[to300],
-             main = paste0('MixLayer: ', round(mld,0), 'm ', round(mldTemp, 1), 'C',
-                           '\nTherm: ', round(ttDepth, 0), 'm ', round(tt, 1), 'C'),
-             xlab='Temp (C)', ylab='Depth (m)')
-
-        lines(x=temp[to300], y=-depth[to300])
-        lines(x=c(mldTemp, tEnd), y=c(-mld, -tEndDepth), col='darkgray', lwd=4)
-        points(x=c(mldTemp, tt), y=c(-mld, -ttDepth), col='blue', cex=1.5, pch=15)
-    }
-    list(mldDepth = mld, mldTemp = mldTemp,
-         ttDepth = ttDepth, ttTemp = tt)
-}
-```
-
-Use above function on the smaller hycom datasets (0-6)
-```{r}
-#calcThermoVars(hycomData[[1]]$matchDepth, hycomData[[1]]$water_temp, plot=TRUE)
-
-thermData0 <- bind_rows(lapply(hycomData0, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData1 <- bind_rows(lapply(hycomData1, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData2 <- bind_rows(lapply(hycomData2, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData3 <- bind_rows(lapply(hycomData3, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData4 <- bind_rows(lapply(hycomData4, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData5 <- bind_rows(lapply(hycomData5, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData6 <- bind_rows(lapply(hycomData6, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData7 <- bind_rows(lapply(hycomData7, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData8 <- bind_rows(lapply(hycomData8, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-# Bind all thermocline datasets together
-thermData <- rbind(thermData0, thermData1, thermData2, thermData3, thermData4, 
-                   thermData5, thermData6, thermData7, thermData8)
-
-# Bind main thermocline dataset to main track/soundscape dataframe
-tracks_SS_08 <- cbind(tracks_SS_08, thermData)
-
-# New tracks_SS_08 variable contains Hycom variables now
-##### FINAL tracks_ss_08 dataframe ######
-View(tracks_SS_08)
 ```

### 2023-05-12 03_environ_var.Rmd
saves fulltracks to hard drive and data folder

@@
 saveRDS(tracks_SS_08, 'data/fulltrack08.rda')
+
+# also saves to external hard drive
+saveRDS(tracks_SS_10, 'D:/Buoy_dfs/fulltrack10.rda')
+saveRDS(tracks_SS_12, 'D:/Buoy_dfs/fulltrack12.rda')
+saveRDS(tracks_SS_13, 'D:/Buoy_dfs/fulltrack13.rda')
+saveRDS(tracks_SS_14, 'D:/Buoy_dfs/fulltrack14.rda')
+saveRDS(tracks_SS_16, 'D:/Buoy_dfs/fulltrack16.rda')
+saveRDS(tracks_SS_18, 'D:/Buoy_dfs/fulltrack18.rda')
+saveRDS(tracks_SS_19, 'D:/Buoy_dfs/fulltrack19.rda')
+saveRDS(tracks_SS_20, 'D:/Buoy_dfs/fulltrack20.rda')
+saveRDS(tracks_SS_21, 'D:/Buoy_dfs/fulltrack21.rda')
+saveRDS(tracks_SS_22, 'D:/Buoy_dfs/fulltrack22.rda')
+saveRDS(tracks_SS_23, 'D:/Buoy_dfs/fulltrack23.rda')
+saveRDS(tracks_SS_07, 'D:/Buoy_dfs/fulltrack07.rda')
+saveRDS(tracks_SS_08, 'D:/Buoy_dfs/fulltrack08.rda')
 ```

### 2023-05-15 manuscript.rmd
added Manocci reference

@@
 These three major datasets were joined together based on a common buoy number and time field.
+
+\[Environmental Data\]\
 Environmetal and oceanographic variables thought to be predictors of BW/SW habitat were extracted from NOAA's Environmental Research Division Data Access Program (ERDDAP) and Hybrid Coordinate Ocean Model (HYCOM) using matchEnvData function in package PAMmisc.
@@
 2022).
+
+[On wind curl stress]{.underline}: "Positive wind stress curl lifts the thermocline, which results in enhanced primary production" (Manocci et al 2015).
 
@@
 
+[Example of GAM]{.underline}: Manocci et al. 2015.
+
 ## Results
 
-\[Add here.\] \[Weekly/daily plot of acoustic presence of whales and ships over time from each drift? See Trickey et al. 2022\]
+\[Add here.\]
+
+\[Weekly/daily plot of acoustic presence of whales and ships over time from each drift? See JS Trickey et al. 2022, Fig 3\]
 

### 2023-05-15 data-raw/whales.R
added row counter

@@ -23,7 +23,10 @@ whales <- left_join(whales, tracks, by = join_by(station, closest("dateTimeRound
   dplyr::select(-Latitude.y, -Longitude.y, -spotID, -minNumber, -maxNumber, -bestNumber)
 
 # Whales data as sf
-BWsf <- whales %>% st_as_sf(coords = c("Longitude.x","Latitude.x"), crs=4326)   # whales as sf
+Wsf <- whales %>% st_as_sf(coords = c("Longitude.x","Latitude.x"), crs=4326)   # whales as sf
 
 # Save object
 saveRDS(whales, 'data/whales.rda')
+
+# Number of rows per station
+nrow(whales[whales$station==12,])

### 2023-05-15 data-raw/soundscape_metrics.R
added here package

@@ -2,6 +2,7 @@
 
 library(tidyverse)
 library(lubridate)
+library(here)
 
 # Path to folder in Project/on GitHub
 sslist <- list.files(path = paste0("data-raw/"), pattern = "CCES_", recursive = TRUE)

### 2023-05-15 01_readin.R
added join_by code in joinTable fx

@@ -58,14 +58,14 @@ for (ii in iilist) {
 
 # Make function to join whales + tracks, broadband, and TOL soundscape metrics and edit columns
 joinTable <- function(w, s, t) {
-  left_join(w, s, by = c("UTC" = "dateTime")) %>%   # join tracks with broadband metrics
-    .[!duplicated(.['UTC']),] %>%                   # removes duplicates
-  left_join(., t, by = c("UTC"= "dateTime")) %>%
-    .[!duplicated(.['UTC']),] %>%                   # removes duplicated again
-  mutate(BWpresence = if_else(is.na(nClicks), 0, 1)) %>%    # Add column to specify BW presence (1) or absence (0)
+  left_join(w, s, by = join_by(closest("UTC" <= "dateTime"))) %>%     # join tracks with broadband metrics
+    .[!duplicated(.['UTC']),] %>%                                     # removes duplicates
+  left_join(., t, by = join_by(closest("UTC" <= "dateTime"))) %>%    # joins tracks/BB with TOL metrics
+    .[!duplicated(.['UTC']),] %>%                                     # removes duplicates formed from new join
+  mutate(Wpresence = if_else(is.na(nClicks), 0, 1)) %>%    # Add column to specify whale presence (1) or absence (0)
   dplyr::select(UTC, spotID, Latitude, Longitude, station, 
                 `BB_20-24000`,TOL_63, TOL_125, TOL_2000, 
-                TOL_5000, TOL_20000, species, BWpresence)
+                TOL_5000, TOL_20000, species, Wpresence)
 }
  
 

### 2023-05-15 03_environ_var.Rmd
removed code
deleted unused packages and test HYCOM code

@@
 ---
+For each buoy, the value of each environmental variable will be added to each track point based on Lat, Long, and UTC.
+Data only need to be downloaded once. After that, can refer to the file in the external hard drive ("D:").
+
+Environmental variables: Sea surface temp, Wind stress curl, Distance to continental slope, Depth, bathymetric slope,
+distance to shore, chlorophyll, sea surface height anomalies, + Hycom variables.
+
+This takes about 2.5 hours to execute completely!
 
@@
 library(sf)
-#library(raster)
 library(terra)
@@
 library(PAMmisc)
-#library(marmap)
 library(rerddap)
@@
 ### Get Environmental Variables
-
-For each buoy, the value of each environmental variable will be added to each track point based on Lat, Long, and UTC.
-Data only need to be downloaded once. After that, can refer to the file in the external hard drive ("D:").
-
-Environmental variables: Sea surface temp, Wind stress curl, Distance to continental slope, Depth, bathymetric slope,
-distance to shore, chlorophyll, sea surface height anomalies, + Hycom variables.
 
@@
 
+# runs hycom variable download and bind for one drift. can ignore if above works.
 ```{r}
 ## HYCOM function above stops at buoys 07 and 08, need to manually run it here. FIX!!
-test08 <- matchEnvData(as.data.frame(tracks_SS_08), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
+#test08 <- matchEnvData(as.data.frame(tracks_SS_08), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
 
-thermData08 <- bind_rows(lapply(test08, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
+#thermData08 <- bind_rows(lapply(test08, function(x) {
+#    # do thermo calcs
+#    result <- calcThermoVars(x$matchDepth, x$water_temp)
+#    # get specific values
+#    result$temp400 <- x$water_temp[x$matchDepth == 400]
+#    result$sal400 <- x$salinity[x$matchDepth == 400]
+#    result
+#}))
 
-tracks_SS_08 <- cbind(tracks_SS_08, thermData08)
+#tracks_SS_08 <- cbind(tracks_SS_08, thermData08)
 ```

### 2023-05-15 04_analysis.Rmd
updated with ALL buoys

@@
 ---
+# Exploratory Analysis and Modeling for all 13 buoys
 
@@
 ```{r}
-# add packages here!!
+library(tidyverse)
 ```
@@
 
-Get avg daily environmental variables
+Bind ALL DRIFTS into one dataframe
 ```{r}
-means_08 <- tracks_SS_08 %>%
-  group_by(as.Date(UTC))%>%
-  summarize(sst_mean = mean(sst_mean),
-            dist_slope = mean(dist2slope),
-            depth = mean(depth),
-            slope = mean(slope),
-            dist_shore = mean(dist2shore),
-            chlorophyll_mean = mean(chlorophyll_mean),
-            ssh_mean = mean(ssh_mean),
-            mldDepth = mean(mldDepth),
-            mldTemp = mean(mldTemp),
-            ttDepth = mean(ttDepth),
-            ttTemp = mean(ttTemp),
-            temp400 = mean(temp400),
-            sal400 = mean(sal400),
-            BWpresence = sum(BWpresence))
+allDrifts <- rbind(fulltrack07, fulltrack08, fulltrack10, fulltrack12, fulltrack13, fulltrack14, 
+                   fulltrack16, fulltrack18, fulltrack19, fulltrack20, fulltrack21, fulltrack22, fulltrack23)
 ```
-
-### Exploratory Analysis ######
 
@@
 
-```{r}
-# Drift 8 - all TOL bands, facet wrap
-pivot <- tracks_SS_08 %>%
+```{r all-drifts-all-freq-histogram}
+# create list of fulltrack dfs from global environment to loop through
+dlist <- grep(pattern = "fulltrack", names(.GlobalEnv), value = T)
+
+# all Drifts - all TOL bands, facet wrap
+for (ft in dlist) {
+pivot1 <- get(ft) %>%
   pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
 
-ggplot(data = pivot, aes(Hz, fill = factor(BWpresence))) +
+print(ggplot(data = pivot1, aes(Hz, fill = factor(BWpresence))) +
   geom_histogram(binwidth = 3, color = "black") +
   theme_classic() +
-  ggtitle("Distribution of third octave level intensities: Drift 8") +
+  ggtitle(paste0("Distribution of third octave level intensities: Drift ", substr(ft,10,11))) +
   labs(x = "Sound Intensity (dB)") +
-  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
+  guides(fill = guide_legend(title = "Whale Presence")) +
+  scale_fill_manual(values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.7,0.8)) +
+  facet_grid(~SL))
+}
+
+# Single plot for ALL DRIFTS
+pivot3 <- allDrifts %>%
+  pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 3, color = "black") +
+  theme_classic() +
+  ggtitle("Distribution of third octave level intensities: All Drifts Combined") +
+  labs(x = "Sound Intensity (dB)") +
+  guides(fill = guide_legend(title = "Whale Presence")) +
   scale_fill_manual(values = c("0" = "bisque1",
@@
 
-# Drift 10 - all TOL bands, facet grid
-pivot2 <- tracks_SS_10 %>%
-  pivot_longer(cols = starts_with("TOL")| starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of third octave level intensities: Drift 10") +
-  scale_x_continuous(name = "Sound Intensity (dB)", breaks=seq(50,110,20)) +
-  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
-  scale_fill_manual(values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.76)) +
-  facet_grid(~SL)
-
-# Drift 8 - TOL bands 125 and 2000 only
-pivot3 <- tracks_SS_08 %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz")
 ```
 
+## ignore??
 ```{r drift-8-histogram}
@@
 
-```{r}
-# Drift 10 - TOL bands 125 and 2000 only
-pivot4 <- tracks_SS_10 %>%
+```{r all-drifts-two-freq-histogram}
+# All Drifts - TOL bands 125 and 2000 only
+
+TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
+              `TOL_2000` = "TOL at 2000 Hz") #creates label names for plots below
+
+for (ft in dlist) {
+pivot2 <- get(ft) %>%
+  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
+  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+print(ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 1, color = "black") +
+  theme_classic() +
+  ggtitle(paste0("Distribution of Third-Octave Level (TOL) Intensities: Drift ", substr(ft,10,11))) +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_manual(name = "Whale Presence", 
+                    labels = c("0" = "Absent", "1" = "Present"),
+                    values = c("0" = "bisque1",
+                               "1" = "brown4")) +
+  theme(legend.position = c(0.6,0.7)) +
+  facet_wrap(~TOL, labeller = as_labeller(TOL.labs)))
+}
+
+# Single plot for ALL DRIFTS combined
+pivot4 <- allDrifts %>%
   dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
@@
   theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 10") +
+  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: All Drifts Combined") +
   scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Beaked Whale Presence", 
+  scale_fill_manual(name = "Whale Presence", 
                     labels = c("0" = "Absent", "1" = "Present"),
@@
   facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
+```
 
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz")
+Timeline of drifts in ocean
+```{r}
+ggplot(data = allDrifts, aes(x = UTC, y = station*(-1))) + geom_point(aes(color = factor(station)))
+
+#Note the break in drift 20 - I think it wasn't transmitting for several days (11/15-11/19)
 ```
@@
 # SST vs TOL 2000
-ggplot(data = tracks_SS_08, aes(x = sst_mean, y = `TOL_2000`)) +
+ggplot(data = allDrifts, aes(x = sst_mean, y = `TOL_2000`)) +
   geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
 
-# SST over time
-ggplot(tracks_SS_08, aes(x = UTC, y = sst_mean)) +
+# SST over time (strange plot)
+ggplot(allDrifts, aes(x = UTC, y = sst_mean)) +
   geom_line() +
@@
 # Depth vs TOL 125
-ggplot(data = tracks_SS_08, aes(x = depth, y = `TOL_125`)) +
+ggplot(data = allDrifts, aes(x = depth, y = `TOL_125`)) +
   geom_line() +
@@
 #Mean depth over time
-ggplot(means_08, aes(x = `as.Date(UTC)`, y = depth*(-1))) +
-  geom_line()
+#ggplot(means_08, aes(x = `as.Date(UTC)`, y = depth*(-1))) +
+#  geom_line()
 
 # Chlorophyll vs TOL 125
-ggplot(data = tracks_SS_08, aes(x = chlorophyll_mean, y = `TOL_125`)) +
-  geom_point(aes(color = species))
+ggplot(data = allDrifts, aes(x = chlorophyll_mean, y = `TOL_125`)) +
+  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
 
 # CHl over time
-ggplot(tracks_SS_08, aes(x = UTC, y = chlorophyll_mean)) +
+ggplot(allDrifts, aes(x = UTC, y = chlorophyll_mean)) +
   geom_line() +
@@
 # Mixed Layer Depth vs TOL125 with species
-ggplot(data = tracks_SS_08, aes(x = mldDepth, y = `TOL_125`)) +
-  geom_point(aes(color = species))
+ggplot(data = allDrifts, aes(x = mldDepth, y = `TOL_125`)) +
+  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
 
 # Mixed Layer Depth over time
-ggplot(tracks_SS_08, aes(x = UTC, y = mldDepth*(-1))) +
+ggplot(allDrifts, aes(x = UTC, y = mldDepth*(-1))) +
   geom_line() +
@@
 
-ggplot(data = tracks_SS_08, aes(x = slope, y = `TOL_5000`)) +
-  geom_point(aes(color = species))
+ggplot(data = allDrifts, aes(x = `TOL_125`, y = bathy_slope)) +
+  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
+
 ```
@@
 ```{r}
+#Get avg daily environmental variables
+means <- allDrifts %>%
+  group_by(as.Date(UTC))%>%
+  summarize(sst_mean = mean(sst_mean),
+            dist_slope = mean(dist2slope),
+            depth = mean(depth),
+            slope = mean(bathy_slope),
+            dist_shore = mean(dist2shore),
+            chlorophyll_mean = mean(chlorophyll_mean),
+            ssh_mean = mean(ssh_mean),
+            mldDepth = mean(mldDepth),
+            mldTemp = mean(mldTemp),
+            ttDepth = mean(ttDepth),
+            ttTemp = mean(ttTemp),
+            temp400 = mean(temp400),
+            sal400 = mean(sal400),
+            curlmean = mean(curl_mean),
+            Wh_presence = sum(BWpresence))
+```
+
+```{r}
 #For surface/depth variables USING DAILY MEAN VARIABLES
-# See above for means_08 derivation
-track08Long <- means_08 %>%
+trackLong <- means %>%
   pivot_longer(cols = sst_mean:ssh_mean,
@@
 
-p <- ggplot(data = track08Long, aes(x = Date, y = value)) +
+p <- ggplot(data = trackLong, aes(x = Date, y = value)) +
   geom_line()
- # geom_point(aes(size = BWpresence))   ### How best to represent number of BW detections per day?
 
@@
 # For MLD/Thermocline variables 
-track08Long2 <- tracks_SS_08 %>%
+trackLong2 <- allDrifts %>%
   pivot_longer(cols = mldDepth:sal400,
@@
 
-q <- ggplot(data = track08Long2, aes(x = Date, y = value)) +
+q <- ggplot(data = trackLong2, aes(x = Date, y = value)) +
   geom_line() + geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
@@
 
-track08Long3 <- tracks_SS_08 %>%
+trackLong3 <- allDrifts %>%
   pivot_longer(cols = temp400:sal400,
@@
 
-q <- ggplot(data = track08Long3, aes(x = `Date (2018)`, y = value)) +
-  geom_line() + 
+q <- ggplot(data = trackLong3, aes(x = `Date (2018)`, y = value)) +
+  #geom_line() + 
   geom_point(aes(color = species)) + 
@@
 
+NEEDS FIXING/LOOP/AIS DATA FOR OTHER DRIFTS
 Co-occurrence plots of drift 8 and vessels 
@@
 
-BW Occurrences
+IS THIS HELPFUL??
+Whale Occurrences
 ```{r}
-# Filter Drift 8 by BW presence and absence
-BW1 <- tracks_SS_08 %>%
+# Filter Drift 8 by Whale presence and absence
+W1 <- tracks_SS_08 %>%
   filter(BWpresence == 1)
-BW0 <- tracks_SS_08 %>%
+W0 <- tracks_SS_08 %>%
   filter(BWpresence == 0)
@@
 ggplot() + 
-  geom_histogram(data = BW0, aes(TOL_5000), binwidth = 0.5, fill = "black", color = "white") +
-  geom_histogram(data = BW1, aes(TOL_5000), binwidth = 0.5, fill = "yellow", color = "black")
+  geom_histogram(data = W0, aes(TOL_5000), binwidth = 0.5, fill = "black", color = "white") +
+  geom_histogram(data = W1, aes(TOL_5000), binwidth = 0.5, fill = "yellow", color = "black")
 
 # Percent of time BWs are detected
-print(paste(nrow(BW1) / nrow(tracks_SS_08) * 100,"%"))
+print(paste(nrow(W1) / nrow(tracks_SS_08) * 100,"%"))
 
-# Most abundant BW species when whale is detected (remove NA values)
-tracks_SS_08 %>%
+# Most abundant whale species when whale is detected (remove NA values)
+allDrifts %>%
   filter(!is.na(species)) %>%
@@
 
-# Sound levels BW are exposed to by species
-ZC <- tracks_SS_08 %>%
+# Sound levels whales are exposed to by species
+ZC <- allDrifts %>%
   filter(species == "ZC")
 
-BW43 <- tracks_SS_08 %>%
+BW43 <- allDrifts %>%
   filter(species == "BW43")
+
+BB <- allDrifts %>%
+  filter(species == "BB")
+
+BWC <- allDrifts %>%
+  filter(species == "BWC")
+
+BW37V <- allDrifts %>%
+  filter(species == "BW37V")
+
+PM <- allDrifts %>%
+  filter(species == "PM")
 
@@
 print(max(BW43$TOL_125))
+print(max(BB$TOL_125))
+print(max(BWC$TOL_125))
+print(max(BW37V$TOL_125))
+print(max(PM$TOL_125))
 
@@
 Density Plots - Species at TOL sound levels
-```{r}
+```{r density-plots}
 # Broadband
-tracks_SS_08 %>%
+for (ft in dlist) {
+plt <- get(ft) %>%
+  filter(!is.na(species))
+
+  print(ggplot(data = plt, aes(`BB_20.24000`, fill = species)) + 
+  geom_density(alpha = 0.5) +
+  ggtitle(paste0("Broadband Drift ", substr(ft, 10, 11))) +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
+}
+
+# Broadband SINGLE PLOT ALL DRIFTS
+allDrifts %>%
   filter(!is.na(species)) %>%
@@
   geom_density(alpha = 0.5) +
-  ggtitle("Broadband") +
+  ggtitle("Broadband ALL Drifts Combined") +
   scale_x_continuous(name = "Sound Intensity (dB)") +
@@
 # TOL 63
-tracks_SS_08 %>%
+for (ft in dlist) {
+plt <- get(ft) %>%
+  filter(!is.na(species))
+
+  print(ggplot(data = plt, aes(TOL_63, fill = species)) + 
+  geom_density(alpha = 0.5) +
+  ggtitle(paste0("TOL 63, Drift ", substr(ft, 10, 11))) +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
+}
+
+# TOL 63 SINGLE PLOT ALL DRIFTS
+allDrifts %>%
   filter(!is.na(species)) %>%
@@
   geom_density(alpha = 0.5) +
-  ggtitle("63 Hz") +
+  ggtitle("TOL 63 ALL Drifts Combined") +
   scale_x_continuous(name = "Sound Intensity (dB)") +
@@
 # TOL 125
-tracks_SS_08 %>%
+for (ft in dlist) {
+plt <- get(ft) %>%
+  filter(!is.na(species))
+
+  print(ggplot(data = plt, aes(TOL_125, fill = species)) + 
+  geom_density(alpha = 0.5) +
+  ggtitle(paste0("TOL 125, Drift ", substr(ft, 10, 11))) +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
+}
+
+# TOL 125 SINGLE PLOT ALL DRIFTS
+allDrifts %>%
   filter(!is.na(species)) %>%
@@
   geom_density(alpha = 0.5) +
-  ggtitle("125 Hz") +
+  ggtitle("TOL 125 ALL Drifts Combined") +
   scale_x_continuous(name = "Sound Intensity (dB)") +
@@
 # TOL 2000
-tracks_SS_08 %>%
+for (ft in dlist) {
+plt <- get(ft) %>%
+  filter(!is.na(species))
+
+  print(ggplot(data = plt, aes(TOL_2000, fill = species)) + 
+  geom_density(alpha = 0.5) +
+  ggtitle(paste0("TOL 2000, Drift ", substr(ft, 10, 11))) +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
+}
+
+# TOL 2000 SINGLE PLOT ALL DRIFTS
+allDrifts %>%
   filter(!is.na(species)) %>%
@@
   geom_density(alpha = 0.5) +
-  ggtitle("2000 Hz") +
+  ggtitle("TOL 2000 ALL Drifts Combined") +
   scale_x_continuous(name = "Sound Intensity (dB)") +
@@
 # TOL 5000
-tracks_SS_08 %>%
+for (ft in dlist) {
+plt <- get(ft) %>%
+  filter(!is.na(species))
+
+  print(ggplot(data = plt, aes(TOL_5000, fill = species)) + 
+  geom_density(alpha = 0.5) +
+  ggtitle(paste0("TOL 5000, Drift ", substr(ft, 10, 11))) +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
+}
+
+# TOL 5000 SINGLE PLOT ALL DRIFTS
+allDrifts %>%
   filter(!is.na(species)) %>%
@@
   geom_density(alpha = 0.5) +
-  ggtitle("5000 Hz") +
+  ggtitle("TOL 5000 ALL Drifts Combined") +
   scale_x_continuous(name = "Sound Intensity (dB)") +
@@
 # TOL 20,000
-tracks_SS_08 %>%
+for (ft in dlist) {
+plt <- get(ft) %>%
+  filter(!is.na(species))
+
+  print(ggplot(data = plt, aes(TOL_20000, fill = species)) + 
+  geom_density(alpha = 0.5) +
+  ggtitle(paste0("TOL 20000, Drift ", substr(ft, 10, 11))) +
+  scale_x_continuous(name = "Sound Intensity (dB)") +
+  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
+}
+
+# TOL 20,000 SINGLE PLOT ALL DRIFTS
+allDrifts %>%
   filter(!is.na(species)) %>%
@@
   geom_density(alpha = 0.5) +
-  ggtitle("20,000 Hz") +
+  ggtitle("TOL 20,000 ALL Drifts Combined") +
   scale_x_continuous(name = "Sound Intensity (dB)") +
@@
 
-Histograms with BW presence/absence and environmental variables
+ARE THESE HELPFUL??
+Histograms with whale presence/absence and environmental variables. Doesnt differentiate by species though
 ```{r}
 # SST
-ggplot(data = tracks_SS_08, aes(sst_mean, fill = factor(BWpresence))) +
+ggplot(data = allDrifts, aes(sst_mean, fill = factor(BWpresence))) +
   geom_histogram(binwidth = 0.1, color = "black")
@@
 # Depth
-ggplot(data = tracks_SS_08, aes(depth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 10, color = "black")
+ggplot(data = allDrifts, aes(depth, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 100, color = "black")
 
 # Bottom slope
-ggplot(data = tracks_SS_08, aes(slope, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
+ggplot(data = allDrifts, aes(bathy_slope, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 1, color = "black")
 
 # SSH
-ggplot(data = tracks_SS_08, aes(ssh_mean, fill = factor(BWpresence))) +
+ggplot(data = allDrifts, aes(ssh_mean, fill = factor(BWpresence))) +
   geom_histogram(binwidth = 0.005, color = "black")
@@
 # Chlorophyll
-ggplot(data = tracks_SS_08, aes(chlorophyll_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
+ggplot(data = allDrifts, aes(chlorophyll_mean, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 0.05, color = "black")
+
+#Wind Curl
+ggplot(data = allDrifts, aes(curl_mean, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 0.0000005, color = "black")
 
 # Mixed layer depth Depth
-ggplot(data = tracks_SS_08, aes(mldDepth, fill = factor(BWpresence))) +
+ggplot(data = allDrifts, aes(mldDepth, fill = factor(BWpresence))) +
   geom_histogram(binwidth = 0.25, color = "black")
@@
 # Mixed layer depth Temperature
-ggplot(data = tracks_SS_08, aes(mldTemp, fill = factor(BWpresence))) +
+ggplot(data = allDrifts, aes(mldTemp, fill = factor(BWpresence))) +
   geom_histogram(binwidth = 0.1, color = "black")
@@
 # Thermocline Depth
-ggplot(data = tracks_SS_08, aes(ttDepth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.25, color = "black")
+ggplot(data = allDrifts, aes(ttDepth, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 0.75, color = "black")
 
 #Thermocline temperature
-ggplot(data = tracks_SS_08, aes(ttTemp, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.05, color = "black")
+ggplot(data = allDrifts, aes(ttTemp, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 0.1, color = "black")
 
 # Temperature at 400m
-ggplot(data = tracks_SS_08, aes(temp400, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.01, color = "black")
+ggplot(data = allDrifts, aes(temp400, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 0.03, color = "black")
 
 # Salinity at 400 m
-ggplot(data = tracks_SS_08, aes(sal400, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.001, color = "black")
+ggplot(data = allDrifts, aes(sal400, fill = factor(BWpresence))) +
+  geom_histogram(binwidth = 0.005, color = "black")
 ```
@@
 ```{r}
-whales_8_10sf <- BWsf %>%                            
+whales_8_10sf <- Wsf %>%                            
 dplyr::filter(station == "8" | station == "10")
@@
         )
-#ggsave("BWdetect.png")
+#ggsave("Wdetect.png")
 
@@
 # Soundscape + bathymetry
-whales08_sf <- BWsf %>%                            
+whales08_sf <- Wsf %>%                            
   dplyr::filter(station == "8")

### 2023-05-25 manuscript.rmd
added text and references

@@
 editor: visual
+bibliography: references.bib
 ---
@@
 Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen.
-One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters; it can drift freely with the currents for days or weeks at a time.
+One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
+Studying beaked whales can be challenging due to how offshore they live (@robbins2022).
+It can drift freely with the currents for days or weeks at a time.
 The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only.
+Historically, it has been challenging to record foraging calls at several hundred meters of depth (\@barlow).
+
 Its basic components consist of a satellite transmitter above water to monitor GPS information and an acoustic recorder attached by line at \~100m of depth.
@@
 Distance to slope and bathymetric slope were calculated from a bathymetry TIF taken from GEBCO.
-Can show table of variables and justification by papers (Virgili et al.
-2022).
+
+\[Can show table of variables and justification by papers (Virgili et al. 2022).\]
+
+| Covariate | Unit | Source                                                   | Justification |
+|-----------|------|----------------------------------------------------------|---------------|
+|           |      | \[can use A, B, C to avoid repetition, cite in caption\] |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
+|           |      |                                                          |               |
 
@@
 
-[Example of GAM]{.underline}: Manocci et al. 2015.
+[Example of GAM]{.underline}: Manocci et al. 2015. @mannocci2015
+
+AIC explanation @symonds2011
 
@@
 
+Are BW or SW known to have seasonal movements? @robbins2022 suggest they do, though there's conflicting evidence.
+
 \[Add here.\]

### 2023-05-25 04_analysis.Rmd
more gam and glm varieties

@@
 library(tidyverse)
+library(mgcv)
 ```
@@
 ```{r drift-8-histogram}
-ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 8") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Beaked Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
+#ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
+#  geom_histogram(binwidth = 1, color = "black") +
+#  theme_classic() +
+#  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 8") +
+#  scale_x_continuous(name = "Sound Intensity (dB)") +
+#  scale_fill_manual(name = "Beaked Whale Presence", 
+#                    labels = c("0" = "Absent", "1" = "Present"),
+#                    values = c("0" = "bisque1",
+#                               "1" = "brown4")) +
+#  theme(legend.position = c(0.6,0.7)) +
+#  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
 ```
@@
 W1 <- tracks_SS_08 %>%
-  filter(BWpresence == 1)
+  filter(Wpresence == 1)
 W0 <- tracks_SS_08 %>%
-  filter(BWpresence == 0)
+  filter(Wpresence == 0)
 
@@
   geom_histogram(binwidth = 0.005, color = "black")
+```
+### GAMs
+
+```{r}
+model3 <- gam(BWpresence ~ depth + TOL_125 + dist2shore + mldTemp, 
+              data = allDrifts, family = binomial)
+plot(model3)
+
+#try fitting linear model
+
+lm_depth <- lm(BWpresence ~ depth, data = allDrifts)
+ggplot(allDrifts, aes(sst_mean, chlorophyll_mean)) +
+  geom_point() +
+  geom_smooth()
+```
+Categorical term
+```{r}
+# ALL variables, GAM
+model1 <- gam(BWpresence>0 ~ s(BB_20.24000) + s(TOL_63) + s(TOL_125) + s(TOL_2000)
+              + s(TOL_5000) + s(TOL_20000) + s(depth) + s(mldDepth) + s(Latitude)
+              + s(Longitude) + s(sst_mean) + s(curl_mean) + s(dist2slope) 
+              + s(bathy_slope) + s(dist2shore) + s(chlorophyll_mean) + s(ssh_mean) 
+              + s(mldTemp) + s(ttDepth) + s(ttTemp) + s(temp400) + s(sal400),
+              data = allDrifts, family = binomial, method = "REML")
+
+# ALL variables, GLM
+model1GLM <- glm(BWpresence ~ BB_20.24000 + TOL_63 + TOL_125 + TOL_2000
+              + TOL_5000 + TOL_20000 + depth + mldDepth + Latitude
+              + Longitude + sst_mean + curl_mean + dist2slope
+              + bathy_slope + dist2shore + chlorophyll_mean + ssh_mean 
+              + mldTemp + ttDepth + ttTemp + temp400 + sal400,
+              data = allDrifts, family = binomial)
+
+# By species? doesnt work, may need to create dummy variables
+model2 <- gam(BWpresence ~ s(sst_mean, by = as.factor(species)), 
+              data = allDrifts, family = binomial, method = "REML")
+
+# Soundscape metrics only, GAM
+model4 <- gam(BWpresence ~ s(BB_20.24000, k = 10) + 
+                s(TOL_63, k = 10) + s(TOL_125, k = 10) + 
+                s(TOL_2000, k = 10) + s(TOL_5000, k = 10) 
+              + s(TOL_20000, k = 10),
+              data = allDrifts, family = binomial, method = "REML")
+
+# Soundscape metrics only, GLM
+model4GLM <- glm(BWpresence ~ BB_20.24000 + TOL_63 + TOL_125 + TOL_2000
+              + TOL_5000 + TOL_20000,
+              data = allDrifts, family = binomial)
+
+model125 <- gam(BWpresence ~ s(TOL_125, k = 10),
+              data = allDrifts, family = binomial, method = "REML")
+plot(model125, trans = plogis)
+summary(model125)
+gam.check(model125)
+
+# Environmental var only
+model5 <- gam(BWpresence ~ s(depth) + s(mldDepth) + s(Latitude)
+              + s(Longitude) + s(sst_mean) + s(curl_mean) + s(dist2slope) 
+              + s(bathy_slope) + s(dist2shore) + s(chlorophyll_mean) + s(ssh_mean) 
+              + s(mldTemp) + s(ttDepth) + s(ttTemp) + s(temp400) + s(sal400),
+              data = allDrifts, family = binomial, method = "REML")
+
+plogis(coef(model1[1]))
+summary(model1)
+summary(model4)
+summary(model5)
+summary(model4GLM)
+plot(model4GLM)
+gam.check(model1)
+gam.check(model4)
+
+#Log odds scale (when used with family = binomial)
+plot(model1, pages = 6)
+plot(model4, pages = 4, trans = plogis)
+
+#Probability scale + centered on the intercept + adding intercept-related uncertainty with seWithMean arg
+plot(model1, pages = 14, trans = plogis,
+     shift = coef(model1)[1])#,
+     #seWithMean = TRUE)
+
+plot(model2, pages = 1, trans = plogis,
+     shift = coef(model2)[1])
+
+plot(model4, pages = 6, trans = plogis,
+     shift = coef(model4)[1])
 ```

### 2023-05-25 01_readin.R
added more TOL band columns to large dataframe
for next run

@@ -64,8 +64,9 @@ joinTable <- function(w, s, t) {
     .[!duplicated(.['UTC']),] %>%                                     # removes duplicates formed from new join
   mutate(Wpresence = if_else(is.na(nClicks), 0, 1)) %>%    # Add column to specify whale presence (1) or absence (0)
   dplyr::select(UTC, spotID, Latitude, Longitude, station, 
-                `BB_20-24000`,TOL_63, TOL_125, TOL_2000, 
-                TOL_5000, TOL_20000, species, Wpresence)
+                `BB_20-24000`,TOL_63, TOL_125, TOL_1600, 
+                TOL_2000, TOL_3150, TOL_5000, TOL_8000, 
+                TOL_10000, TOL_12500, TOL_20000, species, Wpresence)
 }
  
 

### 2023-05-25 manuscript.rmd
added to all sections

@@
 
-\[Introduce sound, soundscapes, noise in the ocean, noise pollution\]\
-Sound comes from a variety of sources in the ocean - biotic, abiotic, human.
-The ocean is naturally a noisy place, but recent increases in human activities are increasing the ocean sound levels.
+**\[Introduce sound, soundscapes, noise in the ocean, noise pollution\]**\
+Soundscapes in the ocean are made up of three broad categories of sound sources: biophonies, which come from marine mammals, fishes, and invertebrates; geophonies, which consist of natural sounds like rain, wind, earthquakes; and anthropophonies, which consist of human-made sounds from vessels, offshore development, active sonar pingers, and seismic surveys (@miksis-olds, @duarte2021).
+
+The ocean is naturally a noisy place, but recent increases in human activities are increasing the ocean sound levels (cite).
+"Ambient noise off the coast of California has increased many-fold over the past several decades...
+(@moore2013 + other sources w/in that).
 It's likely this noise pollution will interfere with soniferous animals, but it's unclear how, where, and what species will be affected.
+\[add language about sensitivity and vulnerability @duarte2021??
+i think).
 
-\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]\
+**\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
 Among the protected species in the California Current Ecosystem are beaked whale and sperm whales.
 These are toothed whales known as some of the deepest divers in the ocean, reaching depths of nearly 10,000 ft, or 3,000 meters.
-They can dive for over 3 hours, followed by only a few minutes on the surface.
+They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993).
 This diving behavior, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them.
+
 Their reliance on sound can make them particularly sensitive to human noise.
@@
 2022).
-The acoustic interference from vessels, for example, can affect behavior, disrupt communication, or lead to serious harm, even at far distances (See Hooker et al.
-2019 for more info).
-There are several documented cases of mid-frequency naval sonar activities coinciding with mass stranding events by beaked whales (see Anne's paper).
-Under the Marine Mammal Protection Act, all these species are federally protected, and several of these species have threatened or endangered conservation statuses.
+The acoustic interference from vessels, for example, can affect behavior, disrupt communication, or lead to serious harm, even at far distances (See @hooker2019 for more info).
+~~disruption of group cohesion when locating prey (Joy et al.
+2019)~~ killer whales only, spending time and effort fleeing instead of foraging (Joy et al.
+2019, @deruiter2013), masking signals for communication (@erbe) or environmental cues (@duarte2021), reduced mating calls (Weilgart, 2007), and hearing loss (cite).
+
+There are several documented cases of mid-frequency naval sonar activities coinciding with mass stranding events by beaked whales (see @simonis2020).
+Under the Marine Mammal Protection Act, all these species are federally protected, several of these species have threatened or endangered conservation statuses, but most remain listed as "data deficient" (cite IUCN).
+Arguments have been made that the term data-deficient does not illicit the same conservation urgency and therefore gets lowered on the priority list (@parsons2016).
+
 Despite their visually cryptic nature, their sound signals are very well-described.
-With automated classification methods, the clicks and pulses they make during foraging can be identified down to the species level.
-\[Can cite specific examples of noise as stressor to whales.\]
+Beaked whale produce clicks, pulses, and buzzes during foraging starting at the 400(?) m range @tyack.
+The production of sound at this depth may be an adaptation to avoid predation from killer whales, who spend most of its time shallow (from @tyack Baird et al 1998, follow up on this).
+**\[add info about sperm whale echolocation characteristics.\]** Sperm whales also emit series of clicks when socializing (@poupard2022 ).
 
-\[Introduce the importance of passive acoustic monitoring on this type of study and gaps in visual survey methods\]\
+With automated classification methods, these vocalizations can be identified down to the species level. \[Can cite specific examples of noise as stressor to whales.\]
+
+**\[Introduce the importance of passive acoustic monitoring on this type of study, gaps in visual survey methods\]**\
 Passive acoustic monitoring is a tool that allows us to record and listen to the soundscape.
 Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen.
-One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
+**\[List pros and cons of all types of PAM: acoustic tags, towed hydrophones, bottom-mounted\].** One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
 Studying beaked whales can be challenging due to how offshore they live (@robbins2022).
@@
 The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only.
-Historically, it has been challenging to record foraging calls at several hundred meters of depth (\@barlow).
+Historically, it has been challenging to record foraging calls at several hundred meters of depth @barlow2005.
 
@@
 In fact, most of what is known about these deep divers is because of acoustic monitoring.
+Similar designs of drifting buoys have been used in estuarine and coral reef soundscape studies @lillis2018.
 
-\[Main research points\]\
+**\[Main research points\]**\
 Shipping and offshore activities are expected to increase in the future, causing elevated oceanic sound levels.
@@
 Acoustics can serve as important indicators of ecosystem health, but also of how humans are changing the environment.
-Quantifying soundscapes in a standardized way and monitoring these changes over time are priorities for marine managers and researchers right now.
 
-\[Goal\]\
+**\[Goal\]**\
 Characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics and vessel tracking
 
-\[How this work will contribute to gaps in literature\]\
+**\[How this work will contribute to gaps in literature\]**\
 This work will contribute to establishing baseline conditions of noise exposure to inform wind farm development and shipping lane changes.
@@
 Both of these outcomes are of high importance to managers and agencies especially NOAA.
+Quantifying soundscapes in a standardized way and monitoring these changes over time are priorities for marine managers and researchers right now.
+Connecting human noise sources to whale habitat in space and time is needed for efficient management @haver2019.
+Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity.
+With AIS, we can calculate distance from buoy to vessel and sound levels in those moments.
+
+\"Increased understanding of temporal and spatial distribution of beaked whales will enable more effective planning for noise-generating activities to occur where and when fewer sensitive animals are present.\" @hooker2019
+
+"Very large-scale studies, conducted on the spatial, temporal and spectral scales at which these highly mobile predators live their lives, may be needed to evaluate the influence of changes in a marine mammal's acoustic environment." @erbe2014
 
@@
 
-\[Overall data collection efforts\]\
-Acoustic detections were collected with the use of free-drifting Drifting Acoustic Spar Buoy Recorders (DASBRs).
+**\[Overall data collection efforts\]**\
+Acoustic detections were collected with the use of free-drifting Drifting Acoustic Spar Buoy Recorders (hereafter buoys).
 These were deployed as part of the California Current Ecosystem Survey (CCES) spanning the offshore waters from southern Canada to northern Mexico in summer of 2018.
-DASBRs offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species.
-These DASBRs contained two satellite geo-location trackers, two hydrophones, etc and collected a total of 1910 hours of data.
+Buoys offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species.
+These buoys contained two satellite geo-location trackers, two hydrophones, etc and collected a total of 1910 hours of data.
 SWFSC ran acoustic detections in PAMGuard.
-Eight drifting recorders were lost at sea and therefore removed from the analysis.
+Eight buoys were lost at sea and therefore removed from the analysis.
 These contain 1, 5, 10, 25, 75, 95, and 99 percentile distribution from each metric.
 
-\[Study area\]\
+**\[Study area\]**\
 The California Current Ecosystem extends along the US' Economic Exclusion Zone along the west coast.
+It is characterized by...
 It's a highly productive and complex ecosystem that contains six national marine sanctuaries, supports several wild fisheries, and is home to many protected species.
@@
 
-\[Add details about acoustic data collection information.\]\
+\[Can discuss the warm and cool phases that occurred during study period.\]
+
+**\[Add details about acoustic data collection information.\]**\
 Soundscape metrics were duty-cycled (most 18min off, 2 minutes on, but some variation in "off" times) and later processed by the NOAA team with calls and clicks identified to the species level (see Simonis et al 2020 report for details on this survey and Rand et al.
@@
 
-\[Post-processing details, data cleaning\]\
-Soundscape metrics were rounded to the nearest 20 minutes.
-These include third-octave level bands and broadband metrics.
+**\[Post-processing details, data cleaning\]**\
+Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to UTC.
+These include third-octave level bands and broadband (20-24,000 Hz) metrics.
 One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.
 The band at 125 Hertz is often associated with shipping vessel noise and the 2000 Hertz or 2 kHz with mid-frequency naval sonar.
-Beaked whales are known to be sensitive to sonar in this latter range.
-Several TOL levels were chosen as proxies to vessel or sonar frequencies.
+Beaked whales are known to be sensitive to sonar in this latter range (cite).
+Several TOL levels were chosen as proxies to vessel or sonar frequencies (TOL 63 to 20,000).
+
 Whale detections were rounded to nearest 20 minutes.
+A separate column was created as a common join field that contained nearest timestamp in UTC to that of drifter tracks.
 Two types of BW species (?BW and ?BWC) were removed from analysis as the species is unknown.
-Whale detections reduced to "presence/absence" from number of clicks, estimated individuals, etc.
+Whale detections were reduced to "presence/absence" from number of clicks, estimated individuals, etc.
 
@@
 Drift 14 and 15 are the same.
-These three major datasets were joined together based on a common buoy number and time field.
+These three major datasets were joined together based on a common buoy number (column "station") and time field column "UTC").
 
-\[Environmental Data\]\
+**\[Environmental Data\]**\
 Environmetal and oceanographic variables thought to be predictors of BW/SW habitat were extracted from NOAA's Environmental Research Division Data Access Program (ERDDAP) and Hybrid Coordinate Ocean Model (HYCOM) using matchEnvData function in package PAMmisc.
-Variables at depth were calculated using function from Taiki and formula from McCullough et al 2021.
+Variables at depth were calculated using function from Taiki and formula from @mccullough2021.
 Distance to slope and bathymetric slope were calculated from a bathymetry TIF taken from GEBCO.
@@
 | Covariate                     | Unit | Source                                                   | Justification |
-|-----------|------|----------------------------------------------------------|---------------|
+|------------|------------|--------------------------------------|------------|
 |                               |      | \[can use A, B, C to avoid repetition, cite in caption\] |               |
-|           |      |                                                          |               |
-|           |      |                                                          |               |
+| Distance to continental slope |      | GEBCO tif                                                |               |
+| Bathmetric slope              |      | GEBCO tif                                                |               |
 |                               |      |                                                          |               |
@@
 
-\[AIS\]\
+**\[AIS\]**\
 Automatic Identification System vessel tracking data were collected from Marine Cadastre in order to evaluate the effect of spatial proximity to the buoys.
@@
 
-\[Big analysis: statistics, modeling details\]
+**\[Big analysis: statistics, modeling details\]**
 
@@
 
-\[Add here.\]
+\[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day)\]
 
@@
 
-\[Known limitations to this study\] Duty cycle whale detection and soundscape method knowingly omits 18 of 20 minutes of data.
+**\[Known limitations to this study\]\
+**Duty cycle whale detection and soundscape method knowingly omits 18 of 20 minutes of data.
 A sperm whale's true location relative to buoy is unknown so preferred habitat may not reflect the environmental variables that are associated with buoy that detects it (Barkley et al.
@@
 Taiki - help with code!\
-Megan McKenna - guidance on AIS questions. \[Add here.\]
+Megan McKenna - guidance on AIS questions.
+
+\[Add here.\]

### 2023-05-31 manuscript.rmd
added info to all sections

@@
 **\[Introduce sound, soundscapes, noise in the ocean, noise pollution\]**\
-Soundscapes in the ocean are made up of three broad categories of sound sources: biophonies, which come from marine mammals, fishes, and invertebrates; geophonies, which consist of natural sounds like rain, wind, earthquakes; and anthropophonies, which consist of human-made sounds from vessels, offshore development, active sonar pingers, and seismic surveys (@miksis-olds, @duarte2021).
+@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources: \
+- biophonies, which come from marine mammals, fishes, and invertebrates; \
+- geophonies, which consist of natural sounds like rain, wind, earthquakes; \
+- anthropophonies, which consist of human-made sounds from vessels, offshore development, active sonar pingers, and seismic surveys (@miksis-olds, @duarte2021).
 
-The ocean is naturally a noisy place, but recent increases in human activities are increasing the ocean sound levels (cite).
+The ocean is naturally a noisy place, but recent increases in human activities are increasing the ocean sound levels (@tyack2008 ).
+Frequencies of these activities often overlap in frequency range, space, and time.
+
+\[Chronic/residual/ambient/transient/continuous/impulsive sound\] @mckenna2021 and @richardson2013 \
+Transient sounds have a short duration, while continuous sounds keep going. @richardson2013.
+
+Lower frequencies travel further, and higher frequencies tend to attentuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ).
+Ambient noise in shallow water has more variability than deeper water (@richardson2013)
+
+\[Where and to what extend is noise pollution regulated?\] Marine Strategy Framework Directive by European Union.
+
 "Ambient noise off the coast of California has increased many-fold over the past several decades...
@@
 **\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
-Among the protected species in the California Current Ecosystem are beaked whale and sperm whales.
-These are toothed whales known as some of the deepest divers in the ocean, reaching depths of nearly 10,000 ft, or 3,000 meters.
+[Beaked whales]{.underline} (family) routinely dive for hours at a time to depths greater than 1000 ft.
 They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993).
 This diving behavior, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them.
+This may be an anti-predation adaptation to avoid killer whales (cite) as well as a synchronized foraging dives following a silent descent (@aguilardesoto2020 ).
+Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite).
+It is thought that this anti-predator response is exactly what makes them sensitive to human noise.
 
-Their reliance on sound can make them particularly sensitive to human noise.
-Sperm whales may have more acoustic tolerance than beaked whales (Miller et al.
-2022).
-The acoustic interference from vessels, for example, can affect behavior, disrupt communication, or lead to serious harm, even at far distances (See @hooker2019 for more info).
-~~disruption of group cohesion when locating prey (Joy et al.
-2019)~~ killer whales only, spending time and effort fleeing instead of foraging (Joy et al.
-2019, @deruiter2013), masking signals for communication (@erbe) or environmental cues (@duarte2021), reduced mating calls (Weilgart, 2007), and hearing loss (cite).
+[Sperm whales]{.underline} (*Physeter macrocephalus*) may have more acoustic tolerance than beaked whales (Miller et al. 2022).
 
-There are several documented cases of mid-frequency naval sonar activities coinciding with mass stranding events by beaked whales (see @simonis2020).
+The acoustic interference from vessels, for example, can :\
+- affect behavior (displacement from habitat, spending time and effort fleeing instead of foraging (Joy et al. 2019, @deruiter2013, @aguilarsoto)\
+- disrupt communication and echolocation\
+- lead to physical harm, even at far distances (See @hooker2019 for more info) like hearing loss and elevated stress levels\
+- masking signals for communication (\@erbe, or environmental cues (@duarte2021)\
+- reduced mating calls (Weilgart, 2007)
+
+There are several documented cases of naval mid-frequency active sonar (MFAS) activities coinciding with mass stranding events by beaked whales (see @simonis2020, table 1 in @weilgart, @deruiter2013 , @stanistreet2022, @bernaldodequiros).
+[For sperm whales]{.underline}: "The potential impact of full-scale military exercises on sperm whales are not known.
+@stanistreet2022, though there is evidence that male sperm whales stop feeding and change behaviors ( @curé2013).
+
 Under the Marine Mammal Protection Act, all these species are federally protected, several of these species have threatened or endangered conservation statuses, but most remain listed as "data deficient" (cite IUCN).
@@
 
-Despite their visually cryptic nature, their sound signals are very well-described.
+Despite their visually cryptic nature, their sound signals are very well-described, down to the species level @baumann-pickering2013, @baumann-pickering2014.
 Beaked whale produce clicks, pulses, and buzzes during foraging starting at the 400(?) m range @tyack.
 The production of sound at this depth may be an adaptation to avoid predation from killer whales, who spend most of its time shallow (from @tyack Baird et al 1998, follow up on this).
+Some are still being discovered and identified - see @manzano-roth2023.
+
 **\[add info about sperm whale echolocation characteristics.\]** Sperm whales also emit series of clicks when socializing (@poupard2022 ).
-
-With automated classification methods, these vocalizations can be identified down to the species level. \[Can cite specific examples of noise as stressor to whales.\]
 
@@
 Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen.
-**\[List pros and cons of all types of PAM: acoustic tags, towed hydrophones, bottom-mounted\].** One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
+**\[List pros and cons of all types of PAM: acoustic tags, towed hydrophones, bottom-mounted, static (Mellinger et al)\].** One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
 Studying beaked whales can be challenging due to how offshore they live (@robbins2022).
@@
 
+**\[AIS\]**
+
+Automatic Identification System. This was collected to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.
+
+@aguilarsoto2006 -- elevated sound levels duing close encounters between ships and beaked whales can be detrimental.
+
 **\[Main research points\]**\
@@
 **\[Goal\]**\
-Characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics and vessel tracking
+Characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics and vessel tracking.
+Characterizing the acoustic environment and identifying co-occurrence of whales and anthropogenic sound has been done in a similar study by @weiss2021.
 
-**\[How this work will contribute to gaps in literature\]**\
+**\[How this work will contribute to gaps in literature, "To date, studies have focused on this and this, but not on this"\]**\
 This work will contribute to establishing baseline conditions of noise exposure to inform wind farm development and shipping lane changes.
+
 It'll also add to the knowledge base of beaked and sperm whale distribution through acoustic detections.
+
 Both of these outcomes are of high importance to managers and agencies especially NOAA.
+Within their Ocean Noise Strategy Roadmap, NOAA is seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic soudn sources contribute to the soundscape.
+Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity.
+With AIS, we can calculate distance from buoy to vessel and sound levels in those moments.
+And 3) NOAA is seeking more information on vocally active species' distributions.
+
+\[add here\]
+
 Quantifying soundscapes in a standardized way and monitoring these changes over time are priorities for marine managers and researchers right now.
 Connecting human noise sources to whale habitat in space and time is needed for efficient management @haver2019.
-Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity.
-With AIS, we can calculate distance from buoy to vessel and sound levels in those moments.
 
-\"Increased understanding of temporal and spatial distribution of beaked whales will enable more effective planning for noise-generating activities to occur where and when fewer sensitive animals are present.\" @hooker2019
+"Increased understanding of temporal and spatial distribution of beaked whales will enable more effective planning for noise-generating activities to occur where and when fewer sensitive animals are present." @hooker2019
 
 "Very large-scale studies, conducted on the spatial, temporal and spectral scales at which these highly mobile predators live their lives, may be needed to evaluate the influence of changes in a marine mammal's acoustic environment." @erbe2014
+
+"Research should be conducted on how the level of anthropogenic activity (such as the types and numbers of vessels) is related to resulting noise levels." @hildebrand2009
 
@@
 
+see @harvey for report on conditions
+
 **\[Add details about acoustic data collection information.\]**\
@@
 2022 for duty-cycle study), and soundscape metrics were binned into low, medium, high frequency bands.
+Can include software used, sampling rate?
 
@@
 These include third-octave level bands and broadband (20-24,000 Hz) metrics.
-One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.
-The band at 125 Hertz is often associated with shipping vessel noise and the 2000 Hertz or 2 kHz with mid-frequency naval sonar.
+~~One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.~~ The band at 125 Hertz is often associated with shipping vessel noise and the 2000 Hertz or 2 kHz with mid-frequency naval sonar (cite).
 Beaked whales are known to be sensitive to sonar in this latter range (cite).
@@
 
-\[Can show table of variables and justification by papers (Virgili et al. 2022).\]
+\[Can show table of variables and justification by papers (Virgili et al. 2022)\]
+
+use @torreblanca2022 for deep habitat info\
+see @johnston2008 and @manzano-roth2023 for seamount habitat importance
 
 | Covariate                     | Unit | Source                                                          | Justification |
-|------------|------------|--------------------------------------|------------|
+|--------------|--------------|-------------------------------|--------------|
 |                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |               |
-| Distance to continental slope |      | GEBCO tif                                                |               |
-| Bathmetric slope              |      | GEBCO tif                                                |               |
+| Distance to continental slope |      | General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015 |               |
+| Bathymetric slope             |      | GEBCO tif                                                       |               |
 |                               |      |                                                                 |               |
@@
 
-[Example of GAM]{.underline}: Manocci et al. 2015. @mannocci2015
+[Example of GAM]{.underline}: @mannocci2015
 
@@
 
-\[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day)\]
+\[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day, drift with most detections/day)\]
 
@@
 
+Non-AIS vessels or small vessels (fishing, recreational) may contribute to the soundscape in relevant frequency bands (@hermannsen2019 , @hildebrand2009 ) and this does not get captured here.
+
 \[Add here.\]

### 2023-05-31 04_analysis.Rmd
added violin and box plots

@@
 #Note the break in drift 20 - I think it wasn't transmitting for several days (11/15-11/19)
+```
+
+Avg sound levels by band (scatterplot)
+```{r}
+avgdB <- TO_12 %>%
+  summarize(TOL_25 = mean(TOL_25),
+            TOL_31.5 = mean(TOL_31.5),
+            TOL_40 = mean(TOL_40),
+            TOL_50 = mean(TOL_50),
+            TOL_63 = mean(TOL_63),
+            TOL_80 = mean(TOL_80),
+            TOL_100 = mean(TOL_100),
+            TOL_125 = mean(TOL_125),
+            TOL_160 = mean(TOL_160),
+            TOL_200 = mean(TOL_200),
+            TOL_250 = mean(TOL_250),
+            TOL_315 = mean(TOL_315),
+            TOL_400 = mean(TOL_400),
+            TOL_500 = mean(TOL_500),
+            TOL_630 = mean(TOL_630),
+            TOL_800 = mean(TOL_800),
+            TOL_1000 = mean(TOL_1000),
+            TOL_1250 = mean(TOL_1250),
+            TOL_1600 = mean(TOL_1600),
+            TOL_2000 = mean(TOL_2000),
+            TOL_2500 = mean(TOL_2500),
+            TOL_3150 = mean(TOL_3150),
+            TOL_4000 = mean(TOL_4000),
+            TOL_5000 = mean(TOL_5000),
+            TOL_6300 = mean(TOL_6300),
+            TOL_8000 = mean(TOL_8000),
+            TOL_10000 = mean(TOL_10000),
+            TOL_12500 = mean(TOL_12500),
+            TOL_16000 = mean(TOL_16000),
+            TOL_20000 = mean(TOL_20000)) %>%
+  pivot_longer(cols = TOL_25:TOL_20000,
+               names_to = "frequency",
+               values_to="dB")
+
+ggplot(avgdB, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
+  geom_point() +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```
+
+Box plots
+```{r}
+TO_longer <- pivot_longer(TO_14, cols = TOL_25:TOL_20000,
+               names_to = "frequency",
+               values_to="dB")
+ggplot(TO_longer, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
+  geom_boxplot() +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+# Violin plot
+ggplot(TO_longer, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
+  geom_violin() +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 ```

### 2023-06-01 manuscript.rmd
added to all sections

@@
 
-**\[Introduce sound, soundscapes, noise in the ocean, noise pollution\]**\
-@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources: \
-- biophonies, which come from marine mammals, fishes, and invertebrates; \
-- geophonies, which consist of natural sounds like rain, wind, earthquakes; \
-- anthropophonies, which consist of human-made sounds from vessels, offshore development, active sonar pingers, and seismic surveys (@miksis-olds, @duarte2021).
+**\[Sound, soundscapes, noise in the ocean, noise pollution\]**\
+@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources:\
+- biophonies, which come from marine mammals, fishes, and invertebrates;\
+- geophonies, which consist of natural sounds like rain, wind, earthquakes;\
+- anthropophonies, which consist of human-made sounds from vessels, offshore development, active sonar pingers/echosounders, and seismic surveys (@miksis-olds, @duarte2021).
 
@@
 
-\[Chronic/residual/ambient/transient/continuous/impulsive sound\] @mckenna2021 and @richardson2013 \
+\[Chronic/residual/ambient/transient/continuous/impulsive sound\] @mckenna2021 and @richardson2013\
 Transient sounds have a short duration, while continuous sounds keep going. @richardson2013.
 
-Lower frequencies travel further, and higher frequencies tend to attentuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ).
+Lower frequencies travel further, and higher frequencies tend to attenuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ).
 Ambient noise in shallow water has more variability than deeper water (@richardson2013)
@@
 **\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
-[Beaked whales]{.underline} (family) routinely dive for hours at a time to depths greater than 1000 ft.
+[Beaked whales]{.underline} (*Ziphiidae*) routinely dive for hours at a time to depths greater than 1000 ft.
 They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993).
@@
 Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite).
+Why do they have both adaptations?
 It is thought that this anti-predator response is exactly what makes them sensitive to human noise.
 
-[Sperm whales]{.underline} (*Physeter macrocephalus*) may have more acoustic tolerance than beaked whales (Miller et al. 2022).
+Species studied:\
+- ZC Cuvier's beaked whale *Ziphius cavirostris\
+-* BW43, possibly Perrin's beaked whale *M*. *perrini*\
+- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi\
+-* BB, --???\
+- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens\
+-* PM, sperm whale
 
-The acoustic interference from vessels, for example, can :\
-- affect behavior (displacement from habitat, spending time and effort fleeing instead of foraging (Joy et al. 2019, @deruiter2013, @aguilarsoto)\
-- disrupt communication and echolocation\
-- lead to physical harm, even at far distances (See @hooker2019 for more info) like hearing loss and elevated stress levels\
-- masking signals for communication (\@erbe, or environmental cues (@duarte2021)\
-- reduced mating calls (Weilgart, 2007)
+sp[erm whales]{.underline} (*Physeter macrocephalus*) may have more acoustic tolerance than beaked whales (Miller et al. 2022).
+
+The acoustic interference from human activity (IN ODONTOCETES OR ALL MARINE LIFE?), for example, can :\
+- affect **behavior** (displacement from habitat, spending time and effort fleeing instead of foraging (Joy et al.
+2019, @deruiter2013, @aguilarsoto2006 , @stanistreet2022, @joyce2020, @falcone2017, @cholewiak2017 )\
+- disrupt **communication** and echolocation (same as above?)\
+- lead to physical **harm**, even at far distances (up to 100 km away from @falcone2017) like hearing loss, tissue damage, and elevated stress levels. (See @hooker2019 for more info)\
+- **masking** signals for communication (@erbe2014?, ) or environmental cues (@duarte2021)\
+- reduced mating calls (Weilgart, 2007)\
+- see Cox et al 2006 for impacts of sound on BW
 
@@
 The production of sound at this depth may be an adaptation to avoid predation from killer whales, who spend most of its time shallow (from @tyack Baird et al 1998, follow up on this).
-Some are still being discovered and identified - see @manzano-roth2023.
+Some are still being discovered and identified globally- see @manzano-roth2023, @yamada2019.
+
+Do beaked whales show temporal variation in their calls? See @ziegenhorn2023
 
@@
 Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen.
-**\[List pros and cons of all types of PAM: acoustic tags, towed hydrophones, bottom-mounted, static (Mellinger et al)\].** One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
+
+**\[List pros and cons of all types of PAM: acoustic tags, towed hydrophones, bottom-mounted, static (Mellinger et al)\].** One type of passive acoustic recorder called a drifting buoy that drifts freely with the currents for days or weeks at a time.
+It allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
+Buoys offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species.
 Studying beaked whales can be challenging due to how offshore they live (@robbins2022).
-It can drift freely with the currents for days or weeks at a time.
 The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only.
@@
 
-Automatic Identification System. This was collected to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.
+Automatic Identification System.
+Ships are abundant throughout the study area due to major US ports.
+This was collected to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.
 
 @aguilarsoto2006 -- elevated sound levels duing close encounters between ships and beaked whales can be detrimental.
+
+How have other studies separated vessel noise (contribution to soundscape) from ambient soundscape? See @weiss2021 and @haver2020. Power spectral density ambient sound levels.
 
@@
 Characterizing the acoustic environment and identifying co-occurrence of whales and anthropogenic sound has been done in a similar study by @weiss2021.
+This serves as baseline monitoring of soundscape for offshore areas.
 
@@
 Both of these outcomes are of high importance to managers and agencies especially NOAA.
-Within their Ocean Noise Strategy Roadmap, NOAA is seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic soudn sources contribute to the soundscape.
+Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic soudn sources contribute to the soundscape.
 Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity.
@@
 These were deployed as part of the California Current Ecosystem Survey (CCES) spanning the offshore waters from southern Canada to northern Mexico in summer of 2018.
-Buoys offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species.
 These buoys contained two satellite geo-location trackers, two hydrophones, etc and collected a total of 1910 hours of data.
@@
 These include third-octave level bands and broadband (20-24,000 Hz) metrics.
-~~One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.~~ The band at 125 Hertz is often associated with shipping vessel noise and the 2000 Hertz or 2 kHz with mid-frequency naval sonar (cite).
+~~One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.~~ The band at 125 Hertz is often associated with shipping vessel noise and the 2-5 kHz range with mid-frequency naval sonar ( @garrett2016 ).
 Beaked whales are known to be sensitive to sonar in this latter range (cite).
@@
 A separate column was created as a common join field that contained nearest timestamp in UTC to that of drifter tracks.
-Two types of BW species (?BW and ?BWC) were removed from analysis as the species is unknown.
+Two types of BW species (?BW and ?BWC) were removed from analysis as the species is unknown and unsure.
 Whale detections were reduced to "presence/absence" from number of clicks, estimated individuals, etc.
@@
 | Covariate                     | Unit | Source                                                          | Justification |
-|--------------|--------------|-------------------------------|--------------|
+|-----------------|-----------------|----------------------|-----------------|
 |                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |               |
@@
 
-\[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day, drift with most detections/day)\]
+[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day ( @ziegenhorn2023 ) , drift with most detections/day)]
 
 \[Weekly/daily plot of acoustic presence of whales and ships over time from each drift? See JS Trickey et al. 2022, Fig 3\]
+
+Look at power spectral density? @haver2020
 

### 2023-06-29 RF_ZC_presence.R
Create RF_ZC_presence.R

@@ -1,0 +1,66 @@
+#Use a Random Forest to identify most important variables
+#in beaked and sperm whale presence
+#Modified from Anne Simonis (6/23/2023)
+
+#Required Packages
+library(randomForest)
+library(dplyr)
+library(rfPermute)
+
+#Define response variable (presence/absence as 0/1)
+#I suggest creating a variable for presence of each species, and then create a
+#random forest model for each species 
+
+# For ZC only
+allDrifts$BWpresence<-as.factor(allDrifts$BWpresence) %>%
+  droplevels(c("BB", "BW43", "BW37V", "PM", "BWC"))
+
+
+#List all variables you want to evaluate (I only included a few here, 
+#but I know you have more environmental and AIS parameters to add)
+covariate.list<-list(c('TOL_125','TOL_2000','TOL_5000',
+                       'sst_mean','dist2slope','depth',
+                       'chlorophyll_mean','mldDepth'))
+
+include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
+
+#make a text string of all covariates considered for a file name 
+string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
+
+#Create a dataframe for the RF model
+DF.modelZC <- na.omit(allDrifts)
+
+# create balanced sample sizes of response for tree construction to 
+# avoid biases associated with imbalanced data
+# For ZC only
+sampsizeZC <- balancedSampsize(DF.modelZC$BWpresence)
+
+# set seed for reproducibility
+set.seed(123)
+
+# create balanced sample size random forest model
+
+#By ZC, doesn't work
+RF.model.ZC <- rfPermute(DF.modelZC$BWpresence ~ ., DF.modelZC[,include.covars], ntree=1000)
+RF.model.ZC
+
+#Evaluate RF models
+#Create a PDF of summary plots
+pdf(paste(Dep,"_",string.covars.used,".pdf"), width=14, height=10) ## Not working?
+round(importance(RF.model.ZC),3) #ranking of importance for each variable
+plotImportance(RF.model.ZC) #visualize importance
+plotTrace(RF.model.ZC)   #model stability w/number of trees (this should be flat!)
+plotImpPreds(RF.model.ZC, DF.model, "Bwpresence")  #distribution of predictors 
+plotPredictedProbs(RF.model.ZC)
+plotProximity(RF.model.ZC) ## Not working?? 
+dev.off()
+
[…]

### 2023-06-29 RF_example_bwpresence.R
Create RF_example_bwpresence.R

@@ -1,0 +1,72 @@
+#Use a Random Forest to identify most important variables
+#in beaked and sperm whale presence
+#Anne Simonis 6/23/2023
+
+#Required Packages
+library(randomForest)
+library(dplyr)
+library(rfPermute)
+
+#Define response variable (presence/absence as 0/1)
+#I suggest creating a variable for presence of each species, and then create a
+#random forest model for each species 
+allDrifts$Bwpresence<-as.factor(allDrifts$BWpresence)
+
+# For ZC only
+allDrifts$BWpresence<-as.factor(allDrifts$BWpresence) %>%
+  droplevels(c("BB", "BW43", "BW37V", "PM", "BWC"))
+
+
+#List all variables you want to evaluate (I only included a few here, 
+#but I know you have more environmental and AIS parameters to add)
+covariate.list<-list(c('TOL_125','TOL_2000','TOL_5000',
+                       'sst_mean','dist2slope','depth',
+                       'chlorophyll_mean','mldDepth'))
+
+include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
+
+#make a text string of all covariates considered for a file name 
+string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
+
+#Create a dataframe for the RF model
+DF.model <- na.omit(allDrifts)
+DF.modelZC <- na.omit(allDrifts)
+
+# create balanced sample sizes of response for tree construction to 
+# avoid biases associated with imbalanced data
+sampsize <- balancedSampsize(DF.model$Bwpresence)
+
+# For ZC only
+sampsizeZC <- balancedSampsize(DF.modelZC$BWpresence)
+
+# set seed for reproducibility
+set.seed(123)
+
+# create balanced sample size random forest model
+RF.model <- rfPermute(DF.model$Bwpresence ~ ., DF.model[,include.covars], ntree=1000)
+RF.model
+
+#By ZC, doesn't work
+RF.model.ZC <- rfPermute(DF.modelZC$BWpresence ~ ., DF.modelZC[,include.covars], ntree=1000)
+RF.model.ZC
+
+#Evaluate RF models
+#Create a PDF of summary plots
+pdf(paste(Dep,"_",string.covars.used,".pdf"), width=14, height=10)
+round(importance(RF.model),3) #ranking of importance for each variable
+plotImportance(RF.model.ZC) #visualize importance
[…]

### 2023-06-29 manuscript.rmd
Update manuscript.rmd

@@
 
-**\[Sound, soundscapes, noise in the ocean, noise pollution\]**\
-@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources:\
-- biophonies, which come from marine mammals, fishes, and invertebrates;\
-- geophonies, which consist of natural sounds like rain, wind, earthquakes;\
-- anthropophonies, which consist of human-made sounds from vessels, offshore development, active sonar pingers/echosounders, and seismic surveys (@miksis-olds, @duarte2021).
+### Soundscapes
+
+**\[Noise pollution is a growing concern and increasing in the ocean.\]**
 
@@
 Frequencies of these activities often overlap in frequency range, space, and time.
+Climate change is also expected to alter the temperature and salinity of the ocean, causing an increase in global sound speed @affatati2022 .
 
-\[Chronic/residual/ambient/transient/continuous/impulsive sound\] @mckenna2021 and @richardson2013\
-Transient sounds have a short duration, while continuous sounds keep going. @richardson2013.
+Ships account for over 80% of the global trade, and an estimated 2.1% annual increase in shipping in expected (@unitedstatesarmycorpsofengineersusace2018) .
+Shipping accounts for the most prominent anthropogenic noise in the ocean, so sound levels are expected to increase.
+\[Discuss post-covid elevated levels.
+\] The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades @hildebrand2009.
+
+Within their Marine Strategy Framework Directive, the European Union has recognized noise as a form of pollution @directiv2008.
+
+**\[Generally, how are cetaceans/fishes/invertebrates affected?\]**
+
+Sound is a critical form of communication for many species in the ocean.
+
+**\[Sound sources\]**\
+\@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources:\
+- biophonies, which come from marine mammals, fishes, and invertebrates;\
+- geophonies, which consist of natural sounds like rain, wind, earthquakes;\
+- anthropophonies, which consist of human-made sounds from vessels (@miksis-olds, @duarte2021) .
+These can be incidental or deliberate @chou2021, @chahouri2022 .
+Incidental = shipping, offshore development, drilling.
+Deliberate = active sonar pingers/echosounders, and seismic surveys.
+
+**\[Chronic (long-term)/residual/ambient/transient/continuous/impulsive/acute sound\]** @mckenna2021 and @richardson2013\
+Transient sounds have a short duration, while continuous sounds keep going.
+@richardson2013.
+How are transient and continuous sounds known to affect marine mammals differently?
+Close approach of a vessel, for example, appears to disrupt beaked whale foraging, as opposed to some medium levels of chronic ship noise ( @aguilarsoto2006 ).
+It is unknown if short-term effects translate into long-term effects
 
@@
 Ambient noise in shallow water has more variability than deeper water (@richardson2013)
-
-\[Where and to what extend is noise pollution regulated?\] Marine Strategy Framework Directive by European Union.
 
@@
 
+### Beaked Whales/Sperm Whales
+
 **\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
@@
 This may be an anti-predation adaptation to avoid killer whales (cite) as well as a synchronized foraging dives following a silent descent (@aguilardesoto2020 ).
-Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite).
+Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite)).
 Why do they have both adaptations?
@@
 
+"Sperm whales and Cuvier's beaked whales are deep-diving species that generally live in pelagic areas characterized by deep basins and trenches or steep slopes.
+As such, their distribution is closely related to bathymetry, with distinctive preference for depths of 1000m and 500-1500m for sperm whales and Cuvier's beaked whales, respectively." ([**Awbery et al 2021**](https://www.frontiersin.org/articles/10.3389/fmars.2022.860242/full))
+
 Species studied:\
-- ZC Cuvier's beaked whale *Ziphius cavirostris\
--* BW43, possibly Perrin's beaked whale *M*. *perrini*\
-- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi\
+- ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
+-* BW43, possibly Perrin's beaked whale *M*. *perrini (IUCN status: endangered)*\
+- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)\
 -* BB, --???\
-- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens\
--* PM, sperm whale
+- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)\
+-* PM, sperm whale *(IUCN status: vulnerable)*
 
-sp[erm whales]{.underline} (*Physeter macrocephalus*) may have more acoustic tolerance than beaked whales (Miller et al. 2022).
+sp[erm whales]{.underline} (*Physeter macrocephalus*) may have more acoustic tolerance than beaked whales @miller2009, @miller2022 .
+Look at 3D3 project of behavioral responses of sperm whales to sonar
 
-The acoustic interference from human activity (IN ODONTOCETES OR ALL MARINE LIFE?), for example, can :\
-- affect **behavior** (displacement from habitat, spending time and effort fleeing instead of foraging (Joy et al.
+**Effects of Acoustic Disturbance**
+
+The acoustic interference from human activity in BW/SW, for example, can :\
+- affect **behavior** (displacement from habitat, spending time and effort fleeing instead of foraging, increasing call intensity (cite?) (Joy et al.
 2019, @deruiter2013, @aguilarsoto2006 , @stanistreet2022, @joyce2020, @falcone2017, @cholewiak2017 )\
@@
 
-There are several documented cases of naval mid-frequency active sonar (MFAS) activities coinciding with mass stranding events by beaked whales (see @simonis2020, table 1 in @weilgart, @deruiter2013 , @stanistreet2022, @bernaldodequiros).
+There are several documented cases of naval mid-frequency active sonar (MFAS) activities coinciding with mass stranding events by beaked whales (see @simonis2020, table 1 in @weilgart, @deruiter2013 , @stanistreet2022, @bernaldodequiros, @fernández2005 ).
 [For sperm whales]{.underline}: "The potential impact of full-scale military exercises on sperm whales are not known.
@@
 
-Under the Marine Mammal Protection Act, all these species are federally protected, several of these species have threatened or endangered conservation statuses, but most remain listed as "data deficient" (cite IUCN).
+Under the Marine Mammal Protection Act, all these species are federally protected, several of these species have threatened or endangered conservation statuses (Endangered Species Act), but most remain listed as "data deficient" (cite IUCN).
+\[Also include National Marine Sanctuary Act\].
 Arguments have been made that the term data-deficient does not illicit the same conservation urgency and therefore gets lowered on the priority list (@parsons2016).
@@
 
-**\[add info about sperm whale echolocation characteristics.\]** Sperm whales also emit series of clicks when socializing (@poupard2022 ).
+**\[add info about sperm whale echolocation characteristics.\]** Sperm whales also emit series of clicks when socializing (@poupard2022 ).\
 
-**\[Introduce the importance of passive acoustic monitoring on this type of study, gaps in visual survey methods\]**\
+### Passive Acoustic Monitoring
+
 Passive acoustic monitoring is a tool that allows us to record and listen to the soundscape.
-Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen.
+It is not fit for all applications or ecological questions, but beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen ( @fleishman ).
 
-**\[List pros and cons of all types of PAM: acoustic tags, towed hydrophones, bottom-mounted, static (Mellinger et al)\].** One type of passive acoustic recorder called a drifting buoy that drifts freely with the currents for days or weeks at a time.
+Bottom-mounted/static: high temporal resolution, low spatial resolution @mellinger
+
+Acoustic tags:
+
+Towed arrays:
+
+Tradeoffs include costly equipment or deployment, data/battery storage requirements, chance of return, spatial coverage, temporal coverage, capturing fine-scale movements, number of individuals, parsing out individuals, through inclement weather or night-time @johnson2009 @fleishman
+
+One type of passive acoustic recorder called a drifting buoy that drifts freely with the currents for days or weeks at a time.
 It allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
@@
 
-**\[AIS\]**
+### **AIS**
 
@@
 
-@aguilarsoto2006 -- elevated sound levels duing close encounters between ships and beaked whales can be detrimental.
+@aguilarsoto2006 -- elevated sound levels during close encounters between ships and beaked whales can be detrimental.
 
@@
 
-**\[Main research points\]**\
+**Laws and Regulations** \[Where and to what extent is noise pollution regulated?\]
+
+see @chou2021 for a list of documents, reports, programs, efforts in ocean noise reduction
+
+Marine Strategy Framework Directive by European Union establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise.
+@directiv2008, @borsanij.f.2023
+
+For mid-frequency cetaceans, auditory impairment threshold levels are as follows:\
+peak sound pressure level (Impulsive) - 230 dB, recommended accumulation period of 24 hrs (impulsive) - 185 dB, recommended accumulation period of 24 hrs (impulsive) - 198 dB (@noaa2018) .
+
+IMO guidelines from 2014 are being updated by July 2023. @internationalmaritimeorganization
+
+Canada is developing an Ocean Noise Strategy
+
+In the US, it's difficult to achieve compliance from Navy.
+
+### Research Objective
+
+\
 Shipping and offshore activities are expected to increase in the future, causing elevated oceanic sound levels.
@@
 It's a challenge to acquire information on these federally protected and acoustically-sensitive species but becoming increasingly important to understand where habitats overlap with human activity.
-Acoustics can serve as important indicators of ecosystem health, but also of how humans are changing the environment.
-
-**\[Goal\]**\
+Acoustics can serve as important indicators of ecosystem health, but also of how humans are changing the environment.\
+\
 Characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics and vessel tracking.
@@
 
-**\[How this work will contribute to gaps in literature, "To date, studies have focused on this and this, but not on this"\]**\
-This work will contribute to establishing baseline conditions of noise exposure to inform wind farm development and shipping lane changes.
+**\[Gaps in literature, rationale for research\]**
+
+"To date, studies have focused on this and this, but not on this".This work will contribute to establishing baseline conditions of noise exposure to inform wind farm development and shipping lane changes.
+\[add here\]
 
@@
 
-Both of these outcomes are of high importance to managers and agencies especially NOAA.
+Both of these outcomes are of high importance to managers and agencies especially NOAA and BOEM (see 2023-2024 report).
 Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic soudn sources contribute to the soundscape.
@@
 And 3) NOAA is seeking more information on vocally active species' distributions.
+This is part of a bill introduced by Congress to "assess underwater sound in high-priority" environments (https://www.congress.gov/bill/117th-congress/house-bill/6987).
 
@@
 
+### **Study Area**
+
+\
+The California Current Ecosystem extends along the US' Economic Exclusion Zone along the west coast.
+It is characterized by...
+It's a highly productive and complex ecosystem that contains five current and one proposed national marine sanctuaries, supports several wild fisheries, and is home to many protected species.
+Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area.
+
+In 2021, California, Washington, and Oregon ranked 3rd, 5th, and 23rd in most trafficked U.S.
+waterways, respectively (@unitedstatesarmycorpsofengineersusace2018.
+Lots of overlap between vessel activity from busy ports and important habitat for protected species.
+
+\[Can discuss the warm and cool phases that occurred during study period.\]
+
+see @harvey for report on conditions
+
+### Data Collection and Processing
+
 **\[Overall data collection efforts\]**\
@@
 
-**\[Study area\]**\
-The California Current Ecosystem extends along the US' Economic Exclusion Zone along the west coast.
-It is characterized by...
-It's a highly productive and complex ecosystem that contains six national marine sanctuaries, supports several wild fisheries, and is home to many protected species.
-Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area.
-
-\[Can discuss the warm and cool phases that occurred during study period.\]
-
-see @harvey for report on conditions
-
 **\[Add details about acoustic data collection information.\]**\
-Soundscape metrics were duty-cycled (most 18min off, 2 minutes on, but some variation in "off" times) and later processed by the NOAA team with calls and clicks identified to the species level (see Simonis et al 2020 report for details on this survey and Rand et al.
-2022 for duty-cycle study), and soundscape metrics were binned into low, medium, high frequency bands.
+Soundscape metrics were duty-cycled (most 18min off, 2 minutes on, but some variation in "off" times) and later processed by the NOAA team with calls and clicks identified to the species level (see @simonis2020 for details on this survey and @rand2022 for duty-cycle study).
+Soundscape metrics were binned into low, medium, high frequency bands.
 Can include software used, sampling rate?
@@
 Beaked whales are known to be sensitive to sonar in this latter range (cite).
-Several TOL levels were chosen as proxies to vessel or sonar frequencies (TOL 63 to 20,000).
+Several TOL levels were chosen as proxies to vessel or sonar frequencies (from TOL 63 to 20,000).
 
@@
 | Covariate                     | Unit | Source                                                          | Justification                                                                                                         |
-|-----------------|-----------------|----------------------|-----------------|
+|-------------------------------|------|-----------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------|
 |                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |                                                                                                                       |
@@
 |                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |               |
+| Wind curl stress              |      | ERDDAP                                                          | "Positive wind stress curl lifts the thermocline, which results in enhanced primary production" (Manocci et al 2015).
+|
 |                               |      |                                                                 |                                                                                                                       |
@@
 
-[On wind curl stress]{.underline}: "Positive wind stress curl lifts the thermocline, which results in enhanced primary production" (Manocci et al 2015).
-
 **\[AIS\]**\
@@
 
-Can use non-parametric Kruskal-Wallis one-way analysis of variance test to examine if number of whale detections per hour differed depending on presence/absence of vessels (see Trickey et al. 2022).
+Use Random Forest model to identify most important predictors.
+
+Can use non-parametric Kruskal-Wallis one-way analysis of variance test to examine *if number of whale detections per hour differed depending on presence/absence of vessels* (see Trickey et al. 2022).
 

### 2023-06-29 data-raw/soundscape_metrics.R
Update soundscape_metrics.R
added code to read in percentiles

@@ -5,22 +5,23 @@ library(lubridate)
 library(here)
 
 # Path to folder in Project/on GitHub
-sslist <- list.files(path = paste0("data-raw/"), pattern = "CCES_", recursive = TRUE)
+#sslist <- list.files(path = paste0("data-raw/"), pattern = "CCES_", recursive = TRUE)
+#### Might not need this - it links to a few raw data files in GitHub ###############
+
+# Read through a few soundscape files on GitHub, change date format
+#for (i in sslist) {
+#  data <- read_csv(here(paste0("data-raw/", i)), show_col_types = FALSE) %>%
+#    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
+#    mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
+#    mutate(variable = paste0(substr(i, 16, 17), substr(i, 6, 7)))
+  
+#  assign(paste0(substr(i, 16, 17), "_", substr(i, 6, 7)), data)
+#}
 
 # Path to folder on my external hard drive
 dirlist <- dir("D:/Soundscape_metrics", recursive=TRUE, full.names=TRUE, pattern="CCES_")
 
-# Read through soundscape files, change date format
-for (i in sslist) {
-  data <- read_csv(here(paste0("data-raw/", i)), show_col_types = FALSE) %>%
-    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
-    mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
-    mutate(variable = paste0(substr(i, 16, 17), substr(i, 6, 7)))
-  
-  assign(paste0(substr(i, 16, 17), "_", substr(i, 6, 7)), data)
-}
-
-# Read though files from external hard drive
+# Read though files from external hard drive, change date format
 for (i in dirlist) {
   data <- read_csv(i, show_col_types = FALSE) %>%
     rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
@@ -31,3 +32,20 @@ for (i in dirlist) {
 
 }
 # Creates BB_07, TO_07, BB_08, TO_08, etc. RDA files for each buoy
+
+
+####### Read in Percentiles ###############
+
+# Path to folder on my external hard drive
+pclist <- dir("D:/Soundscape_metrics/percentiles", 
[…]

### 2023-06-29 07_summary_stats.Rmd
Create 07_summary_stats.Rmd

---
title: "07_summary_stats"
author: "Marina Bozinovic"
date: "2023-06-28"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

Read in RDS files of each buoy + all environ variables + AIS (pending)
```{r}
ftlist <- dir("data/", recursive=TRUE, full.names=TRUE, pattern="fulltrack")

for (ft in ftlist) {
  df <- readRDS(ft)
  assign(paste0("fulltrack", substr(ft, 15, 16)), df)
}
```

Read in RDA files for percentiles
```{r}
## Make list of rda files to read in
pclist <- c("BB_pct01_07", "BB_pct01_08", "BB_pct01_10", "BB_pct01_12", "BB_pct01_13", "BB_pct01_14", "BB_pct01_16", 
            "BB_pct01_18", "BB_pct01_19", "BB_pct01_20", "BB_pct01_21", "BB_pct01_22", "BB_pct01_23", "TO_pct01_07",
            "TO_pct01_08", "TO_pct01_10", "TO_pct01_12", "TO_pct01_13", "TO_pct01_14", "TO_pct01_16", "TO_pct01_18", 
            "TO_pct01_19", "TO_pct01_20", "TO_pct01_21", "TO_pct01_22", "TO_pct01_23", "BB_pct99_07", "BB_pct99_08", 
            "BB_pct99_10", "BB_pct99_12", "BB_pct99_13", "BB_pct99_14", "BB_pct99_16", "BB_pct99_18", "BB_pct99_19", 
            "BB_pct99_20", "BB_pct99_21", "BB_pct99_22", "BB_pct99_23", "TO_pct99_07", "TO_pct99_08", "TO_pct99_10", 
            "TO_pct99_12", "TO_pct99_13", "TO_pct99_14", "TO_pct99_16", "TO_pct99_18", "TO_pct99_19", "TO_pct99_20", 
            "TO_pct99_21", "TO_pct99_22", "TO_pct99_23")

## Read in rda files and assign to object
for (ii in pclist) {
  dta <- readRDS(paste0('data/', ii, '.rda'))

  assign(ii, dta)
}

pct01_07 <- cbind(BB_pct01_07, TO_pct01_07) %>%
  select(., -3)
pct01_08 <- cbind(BB_pct01_08, TO_pct01_08) %>%
  select(., -3)
pct01_10 <- cbind(BB_pct01_10, TO_pct01_10) %>%
  select(., -3)
pct01_12 <- cbind(BB_pct01_12, TO_pct01_12) %>%
  select(., -3)
pct01_13 <- cbind(BB_pct01_13, TO_pct01_13) %>%
  select(., -3)
pct01_14 <- cbind(BB_pct01_14, TO_pct01_14) %>%
  select(., -3)
pct01_16 <- cbind(BB_pct01_16, TO_pct01_16) %>%
  select(., -3)
pct01_18 <- cbind(BB_pct01_18, TO_pct01_18) %>%
  select(., -3)
pct01_19 <- cbind(BB_pct01_19, TO_pct01_19) %>%
  select(., -3)
pct01_20 <- cbind(BB_pct01_20, TO_pct01_20) %>%
  select(., -3)
pct01_21 <- cbind(BB_pct01_21, TO_pct01_21) %>%
  select(., -3)
pct01_22 <- cbind(BB_pct01_22, TO_pct01_22) %>%
  select(., -3)
pct01_23 <- cbind(BB_pct01_23, TO_pct01_23) %>%
  select(., -3)
```

Bind ALL DRIFTS into one dataframe
```{r}
allDrifts <- rbind(fulltrack07, fulltrack08, fulltrack10, 
                   fulltrack12, fulltrack13, fulltrack14, 
                   fulltrack16, fulltrack18, fulltrack19, 
                   fulltrack20, fulltrack21, fulltrack22, fulltrack23)
```

Where are loudest sound levels?
```{r}
pct01_07 %>%
  summarise_if(is.numeric, list(~ min(., na.rm=TRUE)))

```

Where are the quietest sound levels by band?
```{r}
pctlist <- c("07", "08", "10", "12", "14", "16", "18", "19",
             "20", "21", "22", "23")
# Doesn't show time stamp!
for (pc in pctlist) {
  data <- get(paste0("pct01_", pc)) %>%
    summarise_if(is.numeric, list(~ min(., na.rm=TRUE)))

  assign(paste0("min_", pc), data)
}

```

Where do we hear the most whale detections?
```{r}

```

Summary of each species: highest/lowest sound levels, number of detections
```{r}
allDrifts %>%

```

### 2023-06-29 04_analysis.Rmd
Update 04_analysis.Rmd

--- 2023-05-25 04_analysis.Rmd
@@
   theme(legend.position = c(0.7,0.8)) +
-  facet_grid(~SL))
+  (facet_grid(~factor(SL, levels = c('BB_20.24000', 'TOL_63', 'TOL_125', 'TOL_2000', 'TOL_5000', 'TOL_20000')))))
 }
@@
   pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
+
+#SL$size_f <- factor(SL$ , levels = c('BB_20.2400', 'TOL_63', 'TOL_125', 'TOL_2000', 'TOL_5000', 'TOL_20000'))))
 
@@
   theme(legend.position = c(0.7,0.8)) +
-  facet_grid(~SL)
+  facet_grid(~factor(SL, levels = c('BB_20.24000', 'TOL_63', 'TOL_125', 'TOL_2000', 'TOL_5000', 'TOL_20000')))
 
@@
 #Note the break in drift 20 - I think it wasn't transmitting for several days (11/15-11/19)
+```
+
+Avg sound levels by band (scatterplot)
+```{r}
+avgdB <- TO_23 %>%
+  summarize(TOL_25 = mean(TOL_25),
+            TOL_31.5 = mean(TOL_31.5),
+            TOL_40 = mean(TOL_40),
+            TOL_50 = mean(TOL_50),
+            TOL_63 = mean(TOL_63),
+            TOL_80 = mean(TOL_80),
+            TOL_100 = mean(TOL_100),
+            TOL_125 = mean(TOL_125),
+            TOL_160 = mean(TOL_160),
+            TOL_200 = mean(TOL_200),
+            TOL_250 = mean(TOL_250),
+            TOL_315 = mean(TOL_315),
+            TOL_400 = mean(TOL_400),
+            TOL_500 = mean(TOL_500),
+            TOL_630 = mean(TOL_630),
+            TOL_800 = mean(TOL_800),
+            TOL_1000 = mean(TOL_1000),
+            TOL_1250 = mean(TOL_1250),
+            TOL_1600 = mean(TOL_1600),
+            TOL_2000 = mean(TOL_2000),
+            TOL_2500 = mean(TOL_2500),
+            TOL_3150 = mean(TOL_3150),
+            TOL_4000 = mean(TOL_4000),
+            TOL_5000 = mean(TOL_5000),
+            TOL_6300 = mean(TOL_6300),
+            TOL_8000 = mean(TOL_8000),
+            TOL_10000 = mean(TOL_10000),
+            TOL_12500 = mean(TOL_12500),
+            TOL_16000 = mean(TOL_16000),
+            TOL_20000 = mean(TOL_20000)) %>%
+  pivot_longer(cols = TOL_25:TOL_20000,
+               names_to = "frequency",
+               values_to="dB")
+
+ggplot(avgdB, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
+  geom_point() +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```
+
+Box plots
+```{r}
+TO_longer <- pivot_longer(TO_10, cols = TOL_25:TOL_20000,
+               names_to = "frequency",
+               values_to="dB")
+ggplot(TO_longer, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
+  geom_boxplot() +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+# Violin plot
+ggplot(TO_longer, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
+  geom_violin() +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 ```
@@
               data = allDrifts, family = binomial)
-plot(model3)
+#plot(model3)
 
@@
 ```
-
-### Plot maps for Powerpoint images
-
-Set Parameters
-```{r}
-theme_set(theme_bw())
-tmap_mode("plot")
-```
-
-Create sf objects and basemaps for maps (not needed for environmental variables)
-```{r}
-whales_8_10sf <- Wsf %>%                            
-dplyr::filter(station == "8" | station == "10")
-
-trackssf <- st_as_sf(tracks, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326)
-trackBase <- get_tiles(whales_8_10sf, provider = "Esri.WorldShadedRelief")
-allTrackbase <- get_tiles(trackssf, provider = "Stamen.TonerLite", zoom = 5)
-```
-
-Map of whole study area
-```{r}
-# Establish bounding box
-bbox <- st_bbox(c(xmin = -110, 
-            xmax = -128, 
-            ymax = 44, 
-            ymin = 30), crs = st_crs(4326))
-
-#Establish basemap
-allTrackbase <- get_tiles(bbox, provider = "Esri.WorldGrayCanvas", zoom = 5)
-
-# Establish inset map
-us_region <- st_bbox(c(xmin = -90,
-            xmax = -120, 
-            ymax = 48, 
-            ymin = 25), crs = st_crs(4326))
-usa <- get_tiles(us_region, provider = "Esri.WorldGrayCanvas", zoom = 4)
-
-sg <- bb_poly(bb(tracks_sf))
-inset <- tm_shape(usa) + tm_rgb() + tm_shape(sg) + tm_borders(col = "red")
-
-#Map all tracks
-main <- tm_shape(allTrackbase) + tm_rgb() +
-  tm_shape(trackssf) +
-  tm_grid(alpha = 0.6, col = "grey") +
-  tm_dots(col = "darkblue", size = 0.001, legend.show = TRUE) +
-  tm_scale_bar(width = 0.3, text.size = 3, position = c(0.2, 0.03)) +
-  tm_compass(type = "arrow", position = c(0.05, 0.05), color.light = "white") +
-  tm_add_legend('symbol',
-                type = "line",
-                col = "darkblue",
-                labels = "Drifting buoy tracks",
-                size = 1,
-                lwd = 2)
-
-vp <- viewport(x=0.66, y=0.8, width = 0.26, height=0.4)
-main
-print(inset,vp=vp)
-
-# Map all tracks - ggplot
-basemap2 <- as.data.frame(allTrackbase, xy = TRUE)
-#ggplot() +
-#  geom_tile(data = basemap, aes(x=x, y=y, fill = "basemap")) +
-#  geom_sf(data = trackssf, mapping = aes(), color = "red")
-
-# Map tracks 08 and 10 only
-tm_shape(trackBase) + tm_rgb() + 
-  tm_shape(trk_TOL_08) + tm_dots(col = "red", 
-                                 legend.show = TRUE) +
-  tm_shape(trk_TOL_10) + tm_dots(col = "green")
-
-# Bounding box of drift tracks on map
-lon2 <- -129.04666
-lon1 <- -115.78247
-lat1 <- 28.21790
-lat2 <- 45.09226
-
-basemap3 <- ne_countries(scale = "medium", type = "map_units", returnclass = "sf")
-
-# Drift tracks
-ggplot(data = basemap3) +
-  geom_sf(fill = "antiquewhite") +
-  coord_sf(xlim = c(lon2 - 1, lon1 + 1), ylim = c(lat1 - 1, lat2 + 1), expand = FALSE) +
-  theme(panel.grid.major = element_blank()) +
-  geom_path(data = tracks_sf, mapping = aes(x=long,y=lat, group = factor(station)), color = "deeppink") +
-  theme(panel.background = element_rect(fill = "aliceblue"), 
-        axis.text = element_blank(), axis.ticks = element_blank(), 
-        axis.title = element_blank())
-```
-
-Map of soundscapes
-```{r}
-# CCES_08 + CCES_10 Broadband
-tm_shape(trackBase) + tm_rgb() + 
-  tm_shape(trk_TOL_08) + 
-  tm_symbols(col = "BB_20-24000", 
-             border.lwd = NA, 
-             size = 0.4, 
-             sizes.legend = 0.7, 
-             n = 10) +
-  tm_shape(trk_TOL_10) + 
-  tm_symbols(col = "BB_20-24000",
-             border.lwd = NA, 
-             size = 0.4,
-             n = 10) +
-  tm_layout(legend.width = 0.5, 
-            legend.position = c("right","top"),
-            legend.bg.color = "lightblue")
-
-#CCES_08 Third Octave Level
-tm_shape(trackBase) + tm_rgb() + 
-  tm_shape(trk_TOL_08) + 
-  tm_symbols(col = "TOL_63", 
-             border.lwd = NA, 
-             size = 0.5, 
-             n = 10) + 
-  tm_layout(legend.width = 0.5, 
-            legend.position = c("right","top")) + 
-  tm_legend()
-
-ggplot() + 
-  geom_sf(data = trk_TOL_08, aes(color = TOL_125, size = TOL_125), show.legend = FALSE) +
-  geom_sf(data = trk_TOL_10, aes(color = TOL_125, size = TOL_125)) +
-  scale_x_continuous(limits = c(-130,-122)) + 
-  scale_y_continuous(limits = c(35,39)) +
-  scale_color_gradient(low = "navy", high = "yellow")+
-  scale_size(range = c(1,12)) +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(panel.border = element_rect(fill = NA),
-        axis.text = element_blank(),
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        legend.position = "none")
-```
-
-Map of whale detections
-```{r}
-# Whale detections
-ggplot() +
-  geom_point(data = whales8, mapping = aes(x=long, y=lat), shape = 21, color = "black", fill = "green4", size = 4) +
-  geom_point(data = whales10, mapping = aes(x=long, y=lat), shape = 21, color = "black", fill = "green4", size = 4) +
-  scale_x_continuous(limits = c(-130,-122)) + 
-  scale_y_continuous(limits = c(35,39)) +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(panel.border = element_rect(fill = NA),
-        axis.text = element_blank(),
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank()
-        )
-#ggsave("Wdetect.png")
-
-```
-
-Map of bathymetry
-```{r}
-# Bathymetry
-bathym <- ggplot() +
-    geom_raster(data = bathy_df, 
-              mapping = aes(x=x, y=y,
-                            fill = `gebco_bathy`)) +
-  annotation_scale(location = "bl", width_hint = 0.4) +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA))
-
-# Alternative Bathymetry (code from Ocean Tracking Network #https://www.youtube.com/watch?v=xT2F-Y9bKLk)
-bgo <- getNOAA.bathy(lon1 = -122, lon2 = -130,
-                     lat1 = 35, lat2= 39, resolution = 1)  # plot bounding box around drift 8 and 10
-autoplot(bgo, geom = c("raster", "contour"), coast = TRUE, show.legend = FALSE) + 
-  scale_fill_etopo() + #creates ggplot object
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank())
-
-# Soundscape + bathymetry
-whales08_sf <- Wsf %>%                            
-  dplyr::filter(station == "8")
-
-ggplot() +
-  geom_raster(data = bathy_df, 
-              mapping = aes(x=x, y=y,
-                            fill = `gebco_bathy`), show.legend = FALSE) +
-  #geom_contour(data = bathy_df, mapping = aes(x=x, y=y, z=gebco_bathy, linewidth = 0.1), color = "blue", linewidth = 0.01,) +
-  geom_sf(data = trk_TOL_08, aes(color = TOL_125, size = TOL_125), show.legend = FALSE) +
-  geom_sf(data = whales08_sf, aes(), color = "black") +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  scale_color_gradient(low = "beige", high = "hotpink") +
-  scale_fill_etopo() +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA))
-#ggsave("soundscape.png")
-```
-
-Map of AIS
-```{r}
-# Plot + crop to drift area
-#basemap <- ne_countries(scale = "medium", returnclass = "sf")
-#ggplot() + 
-#  geom_sf(aes(), basemap) +
-#  geom_path(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName), #show.legend = F) + 
-  #geom_point(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName)) +
-#  scale_x_continuous(limits = c(min(drift$long-0.7), max(drift$long + 0.7))) + 
-#  scale_y_continuous(limits = c(min(drift$lat-0.7), max(drift$lat + 0.7)))
-
-# AIS data
-ggplot() + 
-  geom_path(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName, linewidth = 1), show.legend = F) +
-  scale_x_continuous(limits = c(-131,-121)) + 
-  scale_y_continuous(limits = c(34,40)) +
-  theme(axis.text = element_blank(), 
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA))
-#ggsave("AIS.png")
-```
-
-Map of Sea Surface Temp with rerddap, Drift 8 only
-```{r}
-# Multi-scale Ultra-high Resolution (MUR) SST Analysis fv04.1, Global, 0.01°, 2002-present, Daily
-SSTInfo <- info('jplMURSST41') 
-murSST <- griddap(SSTInfo, latitude = c(35, 39), longitude = c(-122, -130), 
-                  time = c("2018-08-16", "2018-10-01"), 
-                  fields = 'analysed_sst') # drift 8 dates only
-SST <- murSST$data
-
-# Raster of SST
-ggplot() + 
-  geom_tile(data = SST, aes(x = longitude, y = latitude, fill = analysed_sst)) +
-  scale_fill_gradient(low = "darkblue", high = "yellow") +
-  annotation_scale(location = "br", width_hint = 0.2, plot_unit = "km") +
-  theme(axis.text = element_blank(),
-        axis.ticks = element_blank(), 
-        axis.title = element_blank(),
-        panel.background = element_blank(),
-        panel.border = element_rect(fill = NA),
-        legend.position = "none")
-```
-
-Plot distance to continental slope
-```{r}
-ggplot() +
-  geom_raster(data = bathy_df, 
-              mapping = aes(x=x, y=y,
-                            fill = `gebco_bathy`)) +
-  geom_sf(data = tracks_SS_08, 
-          mapping = aes(col = dist_slope)) +
-  geom_sf(data = cont_slope) + scale_fill_continuous(high = "lightgreen", low = "blue")
-
-# PLOT distance by TOL with whales
-ggplot(data = tracks_SS_08, aes(x = dist_slope, y = `TOL_125`)) + 
-  geom_point() + 
-  geom_smooth(color = "green") +
-  theme_classic()
-```
-
-Another map - need to decide if important to keep
-```{r}
-# Plot
-oceanBase <- get_tiles(tracks_SS_08, provider="Esri.NatGeoWorldMap", crop = TRUE)
-tm_shape(oceanBase) + tm_rgb() +
-  tm_shape(tracks_SS_08) +  tm_symbols(col="species", size = 1) +
-  tm_layout(legend.width = 0.5, 
-            legend.position = c("right","top"))
-
-#Raster base map
-oceanBase <- get_tiles(tracks_SS_10, provider="Esri.NatGeoWorldMap")
-tm_shape(bathy) + tm_raster() +
-  tm_shape(tracks_SS_10) + 
-  tm_symbols(style = "jenks") +
-  tm_layout(legend.width = 0.5)
-```
-
-```{r}
-# Plots for coastline and continental slope. Do I keep?
-sf08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-sf10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-trkraster08 <- rasterize(vect(sf08), agg)
-trkraster10 <- rasterize(vect(sf10), agg)
-
-bathy_df <- as.data.frame(bathy, xy = TRUE)
-
-# Map the coastline and cont. slope
-cont <- st_as_sf(as.contour(bathy, ))
-coast <- cont %>% filter(level == 0)
-cont_slope <- cont %>% filter(level == -2000) #Is 2000m the correct depth for continental shelf?
-cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
-plot(agg)
-plot(cont_slope_rast, col = "red")
-cont_slope_rast <- rasterize(cont_slope, agg)
-plot(cont_slope_rast, col = "red")
-
-ggplot() + 
-  geom_sf(data = cont, col="black") + 
-  geom_sf(data = coast, col = "red") +
-  geom_sf(data = cont_slope, col = "purple") +
-  #geom_sf(data = whales_8_10sf, mapping = aes()) +
-  scale_color_gradient(low="green", high="red")
-```

### 2023-08-08 RF_ZC_presence_7-11_edits.R
Create RF_ZC_presence_7-11_edits.R
added RF model with no balanced sample size for reference

@@ -1,0 +1,80 @@
+#Use a Random Forest to identify most important variables
+#in beaked and sperm whale presence
+#Modified from Anne Simonis (7/11/2023)
+
+#Required Packages
+library(randomForest)
+library(dplyr)
+library(rfPermute)
+
+#Define response variable (presence/absence as 0/1)
+#I suggest creating a variable for presence of each species, and then create a
+#random forest model for each species 
+
+#Descriptive label for analysis
+Dep<-'CCES_Zc'
+
+# For ZC only
+allDrifts$species<-as.factor(allDrifts$species) %>%
+  droplevels(c("BB", "BW43", "BW37V", "PM", "BWC"))
+
+allDrifts$ZcPresence<-factor(ifelse(allDrifts$species=='ZC',1,0))
+
+#List all variables you want to evaluate (I only included a few here, 
+#but I know you have more environmental and AIS parameters to add)
+#covariate.list<-list(c('ZcPresence','TOL_125','TOL_2000','TOL_5000',
+ #                      'sst_mean','dist2slope','depth',
+  #                     'chlorophyll_mean','mldDepth'))
+#more parameters
+covariate.list<-list(c('ZcPresence','BB_20.24000','TOL_63','TOL_125','TOL_2000','TOL_5000',
+                       'TOL_20000', 'curl_mean', 'bathy_slope', 'ssh_mean',  
+                       'sst_mean','dist2slope','depth',
+                       'chlorophyll_mean','mldDepth', 'mldTemp', 'ttDepth',
+                       'ttTemp', 'temp400', 'sal400'))
+
+include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
+
+#make a text string of all covariates considered for a file name 
+string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
+
+#Create a dataframe for the RF model
+DF.modelZC <- na.omit(allDrifts)
+
+# create balanced sample sizes of response for tree construction to 
+# avoid biases associated with imbalanced data
+# For ZC only
+sampsizeZC <- balancedSampsize(DF.modelZC$ZcPresence)
+
+# set seed for reproducibility
+set.seed(123)
+
+# create balanced sample size random forest model
+#By ZC
+RF.model.ZC <- rfPermute(ZcPresence ~ ., data=DF.modelZC[,include.covars],
[…]

### 2023-08-08 RF_ZC_presence.R
Update RF_ZC_presence.R
old version of ZC rf model before edits

@@ -28,7 +28,10 @@ include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
 string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
 
 #Create a dataframe for the RF model
-DF.modelZC <- na.omit(allDrifts)
+#DF.modelZC <- na.omit(allDrifts)
+
+DF.modelZC <- allDrifts %>%
+  filter(!species %in% c("BB", "BW43", "BW37V", "PM", "BWC"))
 
 # create balanced sample sizes of response for tree construction to 
 # avoid biases associated with imbalanced data
@@ -40,9 +43,11 @@ set.seed(123)
 
 # create balanced sample size random forest model
 
-#By ZC, doesn't work
-RF.model.ZC <- rfPermute(DF.modelZC$BWpresence ~ ., DF.modelZC[,include.covars], ntree=1000)
+#Model for ZC
+RF.model.ZC <- rfPermute(DF.modelZC$BWpresence ~ ., DF.modelZC[,include.covars], ntree=1000) ##doesnt work!!
 RF.model.ZC
+
+Dep <- "CCES_ZC_"
 
 #Evaluate RF models
 #Create a PDF of summary plots

### 2023-08-08 RF_example_bwpresence.R
Update RF_example_bwpresence.R
from Anne's code

@@ -32,8 +32,10 @@ string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse
 DF.model <- na.omit(allDrifts)
 DF.modelZC <- na.omit(allDrifts)
 
+
 # create balanced sample sizes of response for tree construction to 
 # avoid biases associated with imbalanced data
+sampsize <- balancedSampsize(DF.model$Bwpresence)
 sampsize <- balancedSampsize(DF.model$Bwpresence)
 
 # For ZC only

### 2023-08-08 manuscript.rmd
Update manuscript.rmd
added to intro

@@
 
+Sound is a critical form of communication for many species in the ocean.
+Since sound travels over four times faster in water than in air and can carry information over far distances @urick1983 , the use of sound has evolved as a critical component of communication, navigation, and foraging for many invertebrates, fishes and cetaceans @duarte2021 .
+Under proper condition, sounds have the ability to travel great distances across ocean basins @miksis-olds.
+
+The total collection of sounds in an environment is called a soundscape, and these are grouped into three broad categories: biophonies, sound produced by cetaceans, fishes, and invertebrates; geophonies, which consist of natural sounds like rain, wind, and earthquakes; anthropophonies, which consist of human-made sounds from shipping vessels, sonar, and survey activities (@miksis-olds, @duarte2021).
+Ocean sound levels exhibit high spatio-temporal variation based on traits such as properties of water, organism diversity, human activities, and bathymetry of a region @andré2011.
+Within soundscape ecology, ocean soundscapes have unique enough sound signatures to be characterized, measured, and monitored as indicators of ecosystem health, species diversity @radford2010 @haver2019.
+
+**\[what do they have to lose in a noisy environment?\]**
+
+In an exceptionally noisy environment, organisms can be negatively impacted.
+Invertebrates and fishes may have stunted development and elevated stress responses (cite), while cetaceans' behavior, calling patterns, physiology and stress levels, ability to locate prey or conspecifics are affected.
+Ecosystem-level effects?
+Each species is affected differently, some more sensitive to certain types of noise.
+While terrestrial ecosystem-level effects have been documented, (Buxton 2017, Shannon 2016, Francis 2012), those within aquatic ecosystems are less studied @kunc2016.
+
 **\[Noise pollution is a growing concern and increasing in the ocean.\]**
 
-The ocean is naturally a noisy place, but recent increases in human activities are increasing the ocean sound levels (@tyack2008 ).
-Frequencies of these activities often overlap in frequency range, space, and time.
-Climate change is also expected to alter the temperature and salinity of the ocean, causing an increase in global sound speed @affatati2022 .
+Ocean noise pollution is among the anthropogenic changes that have recently grown to concerning levels @tyack2008, @hildebrand.
+The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades and has a close correlation to global economic trends @frisk2012.
+with commercial shipping accounting for most of the low-frequency anthropogenic noise in the ocean @hildebrand2009.
+Ships have been increasing in terms of gross tonnage and fleet size @hildebrand2009 (and age?), and now account for over 80% of the global trade, with an estimated 1.4% expected annual increase in shipping through 2027 @unctad.
 
-Ships account for over 80% of the global trade, and an estimated 2.1% annual increase in shipping in expected (@unitedstatesarmycorpsofengineersusace2018) .
-Shipping accounts for the most prominent anthropogenic noise in the ocean, so sound levels are expected to increase.
-\[Discuss post-covid elevated levels.
-\] The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades @hildebrand2009.
+These noise-generating activities are categorized as either incidental or deliberate @chou2021, @chahouri2022.
+Deliberate activities such as active sonar pingers, resource extraction from the seabed, seismic surveys, and echosounders are intentionally produce sound to measure an area and can utilize both high and low frequencies.
+Incidental activities like dredging, offshore development, and shipping produce sound as a by-product @hawkins2017.
+Anthropogenic noise can further be characterized by the frequency and duration of the sound.
+For example, shipping has a low-frequency (\<200 Hz), but continuous sound that is widely distributed across the world's oceans @hildebrand2009.
+Because this type of noise can travel great distances (add number), they can have effects further away from the source.
+Sonar for military surveillance or research such as seafloor mapping occupies the mid-frequency bands, but attenuates quicker and occupies a smaller acoustic space than shipping (shorter distance).
+Close approach of a vessel, for example, appears to disrupt beaked whale foraging, as opposed to some medium levels of chronic ship noise ( @aguilarsoto2006 ).
 
-Within their Marine Strategy Framework Directive, the European Union has recognized noise as a form of pollution @directiv2008.
-
-**\[Generally, how are cetaceans/fishes/invertebrates affected?\]**
-
-Sound is a critical form of communication for many species in the ocean.
-
-**\[Sound sources\]**\
-\@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources:\
-- biophonies, which come from marine mammals, fishes, and invertebrates;\
-- geophonies, which consist of natural sounds like rain, wind, earthquakes;\
-- anthropophonies, which consist of human-made sounds from vessels (@miksis-olds, @duarte2021) .
-These can be incidental or deliberate @chou2021, @chahouri2022 .
-Incidental = shipping, offshore development, drilling.
-Deliberate = active sonar pingers/echosounders, and seismic surveys.
-
-**\[Chronic (long-term)/residual/ambient/transient/continuous/impulsive/acute sound\]** @mckenna2021 and @richardson2013\
-Transient sounds have a short duration, while continuous sounds keep going.
-@richardson2013.
-How are transient and continuous sounds known to affect marine mammals differently?
-Close approach of a vessel, for example, appears to disrupt beaked whale foraging, as opposed to some medium levels of chronic ship noise ( @aguilarsoto2006 ).
-It is unknown if short-term effects translate into long-term effects
+It remains unknown if chronic effects from continuous sound or acute effects from impulsive sounds can alter the normal behavior or physiology of cetaceans enough to affect reproduction and population levels.
 
 Lower frequencies travel further, and higher frequencies tend to attenuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ).
-Ambient noise in shallow water has more variability than deeper water (@richardson2013)
+Ambient noise in shallow water has more variability than deeper water (@richardson2013).
+Climate change is also expected to alter the temperature, salinity, and pH of the ocean, causing an additional increase in global sound speed due to changing absorption qualities @affatati2022 @kunc2016 .
 
-"Ambient noise off the coast of California has increased many-fold over the past several decades...
-(@moore2013 + other sources w/in that).
-It's likely this noise pollution will interfere with soniferous animals, but it's unclear how, where, and what species will be affected.
-\[add language about sensitivity and vulnerability @duarte2021??
-i think).
+**\[This is an issue, but the extent of damage remains unclear, so it's difficult to produce legislation to that effect\]**
+
+As soniferous animals, cetaceans have varying levels of sensitivity and vulnerability to certain types of noise.
+Mysticetes are more affected by low-frequency noise produced by large shipping vessels.
+Odontocetes produce and hear mid- to high-frequency sounds, so human activities that produce sounds at these frequencies interfere with these signals.
+While many studies document negative responses of cetaceans to noise pollution with general frequency bands (cite), the circumstances surrounding these acoustic disturbances (chronic or impulsive, low, mid, or high frequency, shallow or deep water, overlap with anthropogenic activity) and their long-term effects are unclear.
 
@@
 **\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
-[Beaked whales]{.underline} (*Ziphiidae*) routinely dive for hours at a time to depths greater than 1000 ft.
+[Beaked whales]{.underline} (*Ziphiidae*) are among the least studied and most sensitive animals to anthropogenic noise.
+They routinely dive for hours at a time to depths greater than 1000 ft.
 They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993).
@@
 - ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
--* BW43, possibly Perrin's beaked whale *M*. *perrini (IUCN status: endangered)*\
+-* BW43, possibly Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed??\
 - BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)\
@@
 Marine Strategy Framework Directive by European Union establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise.
-@directiv2008, @borsanij.f.2023
+@directiv2008, @borsanij.f.2023 Within their Marine Strategy Framework Directive, the European Union is the only entity to date that has recognized noise as a form of pollution and restricts noise from reaching levels that may cause negative effects to the environment @directiv2008.
+[What about UK Marine Strategy?] See @chou2021 for larger list of efforts.
 
@@
 Both of these outcomes are of high importance to managers and agencies especially NOAA and BOEM (see 2023-2024 report).
-Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic soudn sources contribute to the soundscape.
+Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic sound sources contribute to the soundscape.
 Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity.
@@
 Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area.
+"Ambient noise off the coast of California has increased many-fold over the past several decades..." (@moore2013 + other sources w/in that).
 
@@
 Beaked whales are known to be sensitive to sonar in this latter range (cite).
-Several TOL levels were chosen as proxies to vessel or sonar frequencies (from TOL 63 to 20,000).
+Several TOL levels were chosen as proxies to vessel or sonar frequencies, ranging from TOL 63 to 20,000.
 
@@
 | Covariate                     | Unit | Source                                                          | Justification                                                                                                         |
-|-------------------------------|------|-----------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------|
+|-----------------|-----------------|-----------------|----------------------|
 |                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |                                                                                                                       |
@@
 By exploring buoy/vessel co-occurrence with a 10-km "listening space" around the buoy (McKenna) can allow us to investigate patterns of when whales are calling and the vessel activity during co-occurrence.
+
+AIS was downloaded, processed (converted to UTC), and matched to each drifter using Taiki's PAMscapes package and functions.
 

### 2023-08-08 data-raw/tracks.R
Update tracks.R
added track4 to see alltracks without rounding

@@ -8,9 +8,9 @@ load("data-raw/CCES2018_DriftTracks_Modified_03Nov2022.rda")
 # Read in tracks, filter out lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17)
 tracks <- AllTracks %>%
   dplyr::select(-dist, -speed) %>%
-  mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>%      # round to nearest 20 minutes, will create duplicated in the data
+  mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>% # round to nearest 20 minutes, will create duplicated in the data
   filter(!station %in% c('1','2','3','4','5','6','9','11', '17')) %>%
-  rename(Longitude = long,                                 # Need these columns to specifically say this for ERDDAP matching
+  rename(Longitude = long,              # Need these columns to specifically say this for ERDDAP matching
          Latitude = lat,
          UTC = dateTime)
 tracks <- tracks[!duplicated(tracks[c('UTC', 'station')]),]    # Remove duplicates
@@ -21,3 +21,16 @@ saveRDS(tracks, 'data/tracks.rda')
 # Filter by buoy
 #tracks_08 <- tracks %>% dplyr::filter(station == "8") 
 track4 <- AllTracks %>% dplyr::filter(station == "4")
+
+
+
+### Option to create AllTracks before rounding to nearest 20 min, for AIS
+tracks2 <- AllTracks %>%
+  dplyr::select(-dist, -speed) %>%
+  filter(!station %in% c('1','2','3','4','5','6','9','11', '17')) %>%
+  rename(Longitude = long,            # Need these columns to specifically say this for ERDDAP matching
+         Latitude = lat,
+         UTC = dateTime)
+
+tracks7 <- tracks2 %>%
+  filter(station== 7)

### 2023-08-08 07_summary_stats.Rmd
Update 07_summary_stats.Rmd
added

@@
 
+# 1 Percentile
 pct01_07 <- cbind(BB_pct01_07, TO_pct01_07) %>%
@@
   select(., -3)
+
+pct01_all <- rbind(pct01_07, pct01_08, pct01_10, pct01_12, pct01_13,
+                   pct01_14, pct01_16, pct01_18, pct01_19, pct01_20,
+                   pct01_21, pct01_22, pct01_23)
+# need to add column with buoy name
+
+# 99 Percentile
+pct99_07 <- cbind(BB_pct99_07, TO_pct99_07) %>%
+  select(., -3)
+pct99_08 <- cbind(BB_pct99_08, TO_pct99_08) %>%
+  select(., -3)
+pct99_10 <- cbind(BB_pct99_10, TO_pct99_10) %>%
+  select(., -3)
+pct99_12 <- cbind(BB_pct99_12, TO_pct99_12) %>%
+  select(., -3)
+pct99_13 <- cbind(BB_pct99_13, TO_pct99_13) %>%
+  select(., -3)
+pct99_14 <- cbind(BB_pct99_14, TO_pct99_14) %>%
+  select(., -3)
+pct99_16 <- cbind(BB_pct99_16, TO_pct99_16) %>%
+  select(., -3)
+pct99_18 <- cbind(BB_pct99_18, TO_pct99_18) %>%
+  select(., -3)
+pct99_19 <- cbind(BB_pct99_19, TO_pct99_19) %>%
+  select(., -3)
+pct99_20 <- cbind(BB_pct99_20, TO_pct99_20) %>%
+  select(., -3)
+pct99_21 <- cbind(BB_pct99_21, TO_pct99_21) %>%
+  select(., -3)
+pct99_22 <- cbind(BB_pct99_22, TO_pct99_22) %>%
+  select(., -3)
+pct99_23 <- cbind(BB_pct99_23, TO_pct99_23) %>%
+  select(., -3)
 ```
@@
 ```{r}
+# For buoy 7
 pct01_07 %>%
@@
 
+pct01_07 %>% filter(.$`BB_20-24000` > 120)
+
+########################
+
+pctlist <- c("07", "08", "10", "12", "14", "16", "18", "19",
+             "20", "21", "22", "23")
+# Doesn't show time stamp!
+for (pc in pctlist) {
+  data <- get(paste0("pct99_", pc)) %>%
+    summarise_if(is.numeric, list(~ max(., na.rm=TRUE)))
+
+  assign(paste0("max_", pc), data)
+}
+
+print(max_07)
+
+## Add multi. Sum up rows and extract min/max
+ggplot(data = TO_pct01_07, aes()) + 
+  geom_histogram()
+
+summary(TO_pct01_07)
 ```
@@
 
+print(min_07)
 ```
 
-Where do we hear the most whale detections?
+Where do we hear the most whale detections? Which buoy had the most detections?
 ```{r}
+# Summarize number of detections by species
+allDrifts %>%
+  group_by(species) %>%
+  summarize(n())
+## ZC had the most
+
+#Which buoy had the most detections?
+allDrifts %>%
+  count(station, BWpresence) %>%
+  filter(!BWpresence == 0) %>%
+  arrange(desc(n))
+## Buoy 18 has the most
+
+#Plot buoys
+ggplot() +
+     geom_path(data=allDrifts, aes(x=Longitude, y=Latitude, group=factor(station), col=factor(station))) 
+```
+
+Summary of each species: highest/lowest sound levels, avg sound levels they're exposed to
+```{r}
+# Sort sperm whale detection by decreasing BB sound levels
+allDrifts %>%
+  dplyr::filter(species == "PM") %>%
+  arrange(desc(BB_20.24000))
+
+PM <- allDrifts %>%
+  dplyr::filter(species == "PM")
+
+print(paste("PM Max BB sound level: ", max(PM$BB_20.24000, na.rm = TRUE)))
+print(paste("PM Min BB sound level: ", min(PM$BB_20.24000, na.rm = TRUE)))
+print(paste("PM Max TOL125 sound level: ", max(PM$TOL_125, na.rm = TRUE)))
+print(paste("PM Min TOL125 sound level: ", min(PM$TOL_125, na.rm = TRUE)))
+print(paste("PM Max TOL20000 sound level: ", max(PM$TOL_20000, na.rm = TRUE)))
+print(paste("PM Min TOL20000 sound level: ", min(PM$TOL_20000, na.rm = TRUE)))
+
+# One detection value for BWC
+BWC <- allDrifts %>%
+  dplyr::filter(species == "BWC")
+print(paste("BWC Max BB sound level: ", max(BWC$BB_20.24000, na.rm = TRUE)))
+print(paste("BWC Min BB sound level: ", min(BWC$BB_20.24000, na.rm = TRUE)))
+print(paste("BWC Max TOL125 sound level: ", max(BWC$TOL_125, na.rm = TRUE)))
+print(paste("BWC Min TOL125 sound level: ", min(BWC$TOL_125, na.rm = TRUE)))
+print(paste("BWC Max TOL20000 sound level: ", max(BWC$TOL_20000, na.rm = TRUE)))
+print(paste("BWC Min TOL20000 sound level: ", min(BWC$TOL_20000, na.rm = TRUE)))
+
+BB <- allDrifts %>%
+  dplyr::filter(species == "BB")
+print(paste("BB Max BB sound level: ", max(BB$BB_20.24000, na.rm = TRUE)))
+print(paste("BB Min BB sound level: ", min(BB$BB_20.24000, na.rm = TRUE)))
+print(paste("BB Max TOL125 sound level: ", max(BB$TOL_125, na.rm = TRUE)))
+print(paste("BB Min TOL125 sound level: ", min(BB$TOL_125, na.rm = TRUE)))
+print(paste("BB Max TOL20000 sound level: ", max(BB$TOL_20000, na.rm = TRUE)))
+print(paste("BB Min TOL20000 sound level: ", min(BB$TOL_20000, na.rm = TRUE)))
+
+BW37V <- allDrifts %>%
+  dplyr::filter(species == "BW37V")
+print(paste("BW37V Max BB sound level: ", max(BW37V$BB_20.24000, na.rm = TRUE)))
+print(paste("BW37V Min BB sound level: ", min(BW37V$BB_20.24000, na.rm = TRUE)))
+print(paste("BW37V Max TOL125 sound level: ", max(BW37V$TOL_125, na.rm = TRUE)))
+print(paste("BW37V Min TOL125 sound level: ", min(BW37V$TOL_125, na.rm = TRUE)))
+print(paste("BW37V Max TOL20000 sound level: ", max(BW37V$TOL_20000, na.rm = TRUE)))
+print(paste("BW37V Min TOL20000 sound level: ", min(BW37V$TOL_20000, na.rm = TRUE)))
+
+BW43 <- allDrifts %>%
+  dplyr::filter(species == "BW43")
+print(paste("BW43 Max BB sound level: ", max(BW43$BB_20.24000, na.rm = TRUE)))
+print(paste("BW43 Min BB sound level: ", min(BW43$BB_20.24000, na.rm = TRUE)))
+print(paste("BW43 Max TOL125 sound level: ", max(BW43$TOL_125, na.rm = TRUE)))
+print(paste("BW43 Min TOL125 sound level: ", min(BW43$TOL_125, na.rm = TRUE)))
+print(paste("BW43 Max TOL20000 sound level: ", max(BW43$TOL_20000, na.rm = TRUE)))
+print(paste("BW43 Min TOL20000 sound level: ", min(BW43$TOL_20000, na.rm = TRUE)))
+
+ZC <- allDrifts %>%
+  dplyr::filter(species == "ZC")
+print(paste("ZC Max BB sound level: ", max(ZC$BB_20.24000, na.rm = TRUE)))
+print(paste("ZC Min BB sound level: ", min(ZC$BB_20.24000, na.rm = TRUE)))
+print(paste("ZC Max TOL125 sound level: ", max(ZC$TOL_125, na.rm = TRUE)))
+print(paste("ZC Min TOL125 sound level: ", min(ZC$TOL_125, na.rm = TRUE)))
+print(paste("ZC Max TOL20000 sound level: ", max(ZC$TOL_20000, na.rm = TRUE)))
+print(paste("ZC Min TOL20000 sound level: ", min(ZC$TOL_20000, na.rm = TRUE)))
 
 ```
+Number of detections
+```{r}
+print(paste("Number of PM detections: ", nrow(PM)))
+print(paste("Number of ZC detections: ", nrow(ZC)))
+print(paste("Number of BW43 detections: ", nrow(BW43)))
+print(paste("Number of BWC detections: ", nrow(BWC)))
+print(paste("Number of BB detections: ", nrow(BB)))
+print(paste("Number of BW37V detections: ", nrow(BW37V)))
 
-Summary of each species: highest/lowest sound levels, number of detections
+#Plot
+sp <- allDrifts %>% filter(!species == 'NA')
+ggplot(sp, aes(species)) +
+  geom_bar()
+```
+How do I find WHERE the highest/lowest detections are happening?
 ```{r}
+ZC <- allDrifts %>%
+  dplyr::filter(species == "ZC")
+
+ZC[which.max(ZC$BB_20.24000),]
+ZCmax2000 <- ZC[which.max(ZC$TOL_2000),]
+
+#Point of ZC max TOL2000 sound level. Need to put in context of map.
+ggplot(data = ZCmax2000, aes(x = Longitude, y = Latitude)) +
+  geom_point()
+
+#not working
 allDrifts %>%
+  group_by(species) %>%
+  filter(max(allDrifts$BB_20.24000))
+
+allDrifts %>%
+  summarise_if(is.numeric, list(~ min(., na.rm=TRUE)))
 

### 2023-08-08 02_AIS.Rmd
added note to update AIS directory
Update 02_AIS.Rmd
added Taiki's workflow to this

@@
 ---
+
+#Introduction
+AIS download from Marine Cadastre and match to buoy datasets.
+
+Required Packages
 ```{r}
-#Required Packages
-
+library(PAMscapes)
 library(tidyverse)
-library(sf)
-library(gganimate)
-#install.packages("remotes")
-#remotes::install_github("skgrange/gissr")
-#devtools::install_github("skgrange/gissr")
-library(gissr)
 ```
@@
 
-# Need to add more AIS dates in data folder
+##### Need to add more AIS dates to D: DRIVE!!!
 AISlist <- list.files(path = paste0("data-raw/"), pattern = "AIS_", recursive = TRUE)
@@
 
+-----------------------
+From Taiki's AIS Workflow
+
 ```{r}
-# Plot showing 10km buffer around drift 8
-#tmap_mode("plot")
-#tm_shape(buff) + tm_symbols(col = "blue", border.col = NA) +
-#  tm_shape(trk) + tm_dots(col = "red")
+marcadFiles07 <- downloadMarCadAIS(tracks_SS_07, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
 ```
 
-plots
+Subset to US West Coast
 ```{r}
-# Plot
-ggplot(tracks_SS_08, aes(x = long, y = lat.x)) + geom_point() 
- + geom_sf(data = bbox8, aes(col = "red"))
+subsetMarCadAIS(inDir="D:/AIS", outDir="D:/AIS/AIS_West", name='West_', 
+                latRange=c(20,50), lonRange=c(-140,-110),
+                overwrite=FALSE, progress=TRUE)
+```
+Now read in only ais points that come in contact with buoys.
+```{r}
+ais_near_07 <- readLocalAIS(tracks_SS_07, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+View(ais_near_07)
 
-# Plot showing AIS bounding box around drift 8 tracks
-ggplot() + geom_sf(data = trk_AIS_sf, color = "red")
-ggplot() + geom_sf(data = buff[lgth,] ,color = "red")
+gpsClose <- addAIS(tracks_SS_07, ais_near_07, interpType = 'close', interpTime = 120)
+gpsNone <- addAIS(tracks_SS_07, ais_near_07, interpType = 'none')
+gpsAll <- addAIS(tracks_SS_07, ais_near_07, interpType = 'all', interpTime = 120)
 
+## Using PRE-ROUNDED dataset to try out
+ais_near_raw7 <- readLocalAIS(tracks7, aisDir = "D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+gpsClose_raw7 <- addAIS(tracks7, ais_near_raw7, interpType = 'close', interpTime = 120)
+```
+
+Plot
+```{r}
 ggplot() +
-  geom_sf(data = calif) + coord_sf(default_crs = sf::st_crs(4326)) + theme_bw()
- # geom_sf(data = AIS_sf)
+    geom_path(data=tracks_SS_07, aes(x=Longitude, y=Latitude), col='black') +
+    geom_path(data=ais_near_07, aes(x=Longitude, y=Latitude, group=factor(MMSI), col=inDist)) +
+    xlim(min(tracks_SS_07$Longitude), max(tracks_SS_07$Longitude)) +
+    ylim(min(tracks_SS_07$Latitude), max(tracks_SS_07$Latitude))
 
-p <- ggplot() + geom_sf(data = buff, mapping = aes(x=Longitude,y=Latitude)) +
-  geom_point(data = AIS_within_buff, mapping = aes(x=LON,y=LAT, group = VesselName), color = "red")
+baseGps <- ggplot(data=tracks7, aes(x=Longitude, y=Latitude), col='black') +
+    geom_path() +
+    xlim(min(tracks7$Longitude), max(tracks7$Longitude)) +
+    ylim(min(tracks7$Latitude), max(tracks7$Latitude))
 
-anim <- p +
-  transition_reveal(along = BaseDateTime)+
-  ease_aes('linear')+
-  #shadow_wake(0.1) +
-  ggtitle("Date: {frame_along}")
+baseGps + geom_path(data=gpsClose_raw7, aes(x=shipLong, y=shipLat, group=factor(MMSI), color=shipDist < 10e3)) +
+     geom_point(data=gpsClose_raw7, aes(x=shipLong, y=shipLat, color=shipDist < 10e3), size = 0.1) +
+    ggtitle('CloseInterp')
 
-animate(anim, nframes = 180, fps = 3)
+# Viewing only ships within 10 km of buoys.
+close2ships7 <- gpsClose %>% filter(shipDist < 10e3)
 ```

### 2023-08-08 manuscript.rmd
Update manuscript.rmd
added to BW and SW section of intro

@@
 Since sound travels over four times faster in water than in air and can carry information over far distances @urick1983 , the use of sound has evolved as a critical component of communication, navigation, and foraging for many invertebrates, fishes and cetaceans @duarte2021 .
-Under proper condition, sounds have the ability to travel great distances across ocean basins @miksis-olds.
+Under the right conditions, sounds have the ability to travel across ocean basins @miksis-olds.
 
-The total collection of sounds in an environment is called a soundscape, and these are grouped into three broad categories: biophonies, sound produced by cetaceans, fishes, and invertebrates; geophonies, which consist of natural sounds like rain, wind, and earthquakes; anthropophonies, which consist of human-made sounds from shipping vessels, sonar, and survey activities (@miksis-olds, @duarte2021).
-Ocean sound levels exhibit high spatio-temporal variation based on traits such as properties of water, organism diversity, human activities, and bathymetry of a region @andré2011.
-Within soundscape ecology, ocean soundscapes have unique enough sound signatures to be characterized, measured, and monitored as indicators of ecosystem health, species diversity @radford2010 @haver2019.
+The total collection of sounds in an environment is called a soundscape, and these sources of sound are grouped into three broad categories: biophonies, sound produced by animals like cetaceans, fishes, and invertebrates; geophonies, natural sounds like rain, wind, and earthquakes; and anthropophonies, human-made sounds from shipping vessels, sonar, and survey activities (@miksis-olds, @duarte2021).
+Every region of the ocean experiences high spatio-temporal variation in sound levels and sources based on traits such as properties of water, organism diversity, human activities, and bathymetry @andré2011.
+Because of their unique sound signatures, ocean soundscapes are dynamic and can be characterized, measured, and monitored as indicators of ecosystem health or species diversity @radford2010 @haver2019 @weiss2021 @pijanowski2011 .
 
@@
 
-In an exceptionally noisy environment, organisms can be negatively impacted.
-Invertebrates and fishes may have stunted development and elevated stress responses (cite), while cetaceans' behavior, calling patterns, physiology and stress levels, ability to locate prey or conspecifics are affected.
-Ecosystem-level effects?
+When noise levels reach above normal levels, organisms can be negatively impacted.
+Invertebrates and fishes may have stunted development and elevated stress responses (cite), fishes \[add negative behavior here\], and cetaceans' behavior, calling patterns, physiology and stress levels, ability to locate prey or conspecifics are impacted (cite).
 Each species is affected differently, some more sensitive to certain types of noise.
-While terrestrial ecosystem-level effects have been documented, (Buxton 2017, Shannon 2016, Francis 2012), those within aquatic ecosystems are less studied @kunc2016.
+While terrestrial ecosystem-level effects have been documented, (Buxton 2017, Shannon 2016, Francis 2012), those within aquatic ecosystems lack studies @kunc2016.
 
@@
 Ocean noise pollution is among the anthropogenic changes that have recently grown to concerning levels @tyack2008, @hildebrand.
-The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades and has a close correlation to global economic trends @frisk2012.
-with commercial shipping accounting for most of the low-frequency anthropogenic noise in the ocean @hildebrand2009.
+The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades and has a close correlation to global economic trends @frisk2012, with commercial shipping accounting for most of the low-frequency anthropogenic noise in the ocean @hildebrand2009.
 Ships have been increasing in terms of gross tonnage and fleet size @hildebrand2009 (and age?), and now account for over 80% of the global trade, with an estimated 1.4% expected annual increase in shipping through 2027 @unctad.
@@
 Anthropogenic noise can further be characterized by the frequency and duration of the sound.
-For example, shipping has a low-frequency (\<200 Hz), but continuous sound that is widely distributed across the world's oceans @hildebrand2009.
-Because this type of noise can travel great distances (add number), they can have effects further away from the source.
-Sonar for military surveillance or research such as seafloor mapping occupies the mid-frequency bands, but attenuates quicker and occupies a smaller acoustic space than shipping (shorter distance).
-Close approach of a vessel, for example, appears to disrupt beaked whale foraging, as opposed to some medium levels of chronic ship noise ( @aguilarsoto2006 ).
+For example, shipping produces a continuous, low-frequency (\<200 Hz) sound that is widely distributed across the world's oceans @hildebrand2009.
+~~Because this type of noise can travel great distances, they can have effects further away from the source~~.
+Sonar for military surveillance or research such as seafloor mapping occupies the mid-frequency bands, but attenuates quicker, therefore impacting a smaller acoustic space.
 
 It remains unknown if chronic effects from continuous sound or acute effects from impulsive sounds can alter the normal behavior or physiology of cetaceans enough to affect reproduction and population levels.
+(though see @new2013 for habitat displacement affecting reproduction or survival)
 
@@
 Odontocetes produce and hear mid- to high-frequency sounds, so human activities that produce sounds at these frequencies interfere with these signals.
+The lines are not always so clear cut as odontocetes may be affected by the frequencies of motorized shipping typically known to affect baleen whales @aguilarsoto2006.
+
 While many studies document negative responses of cetaceans to noise pollution with general frequency bands (cite), the circumstances surrounding these acoustic disturbances (chronic or impulsive, low, mid, or high frequency, shallow or deep water, overlap with anthropogenic activity) and their long-term effects are unclear.
+
+habitat degradation
 
@@
 
-**\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
-[Beaked whales]{.underline} (*Ziphiidae*) are among the least studied and most sensitive animals to anthropogenic noise.
-They routinely dive for hours at a time to depths greater than 1000 ft.
-They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993).
-This diving behavior, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them.
-This may be an anti-predation adaptation to avoid killer whales (cite) as well as a synchronized foraging dives following a silent descent (@aguilardesoto2020 ).
-Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite)).
-Why do they have both adaptations?
-It is thought that this anti-predator response is exactly what makes them sensitive to human noise.
+[Beaked whales]{.underline} consist of 24 species within the family *Ziphiidae* and are among the least understood cetaceans and deepest divers in the ocean.
+They regularly dive for over an hour to depths greater than 800 m, and have been known to dive over three hours at a time to depths greater than 2000 m, followed by only a few minutes on the surface @tyack @baird2006.
+Their preference for deep water, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them with visual observations @awbery2022 .
+Much of what is known comes from strandings and stomach content analysis @west2017.
+Even when sighting events occur, they risk misidentification due to the paucity of data regarding morphological variations in some species @aguilardesoto2017.
+Tagging studies prove useful, but difficult to scale/acquire sizable dataset @baird2006 @baird2008 @tyack .
 
-"Sperm whales and Cuvier's beaked whales are deep-diving species that generally live in pelagic areas characterized by deep basins and trenches or steep slopes.
-As such, their distribution is closely related to bathymetry, with distinctive preference for depths of 1000m and 500-1500m for sperm whales and Cuvier's beaked whales, respectively." ([**Awbery et al 2021**](https://www.frontiersin.org/articles/10.3389/fmars.2022.860242/full))
+Beaked whales are also known to be unusually sensitive to noise.
+There are several documented cases of naval mid-frequency active sonar (MFAS) activities coinciding with mass stranding events by beaked whales (see @simonis2020, table 1 in @weilgart, @deruiter2013 , @stanistreet2022, @bernaldodequiros, @fernández2005 ).
+This occurs for two reasons: 1) their anti-predation adaptation to avoid killer whales (*Orcinus orca*)(synchronized foraging, followed by silent, scattered vertical ascent) mimics that of anthropogenic noise and leads to strandings @aguilardesoto2020; 2) the same physiological adaptations that allow for extreme diving are observed to be deadly when forced to alter behavior and flee the noise source more so than other odontocetes @tyack @cox2006 @joyce2020 .
+Furthermore, these responses vary based on the individuals' age, sex, past experience with noise, or current activity @weilgart.
 
-Species studied:\
+Despite their visually cryptic nature, their sound signals are very well-described, down to the species level @baumann-pickering2013, @baumann-pickering2014.
+Beaked whale produce clicks, pulses, and buzzes during foraging starting at the 400(?) m range @tyack.
+With increasing acoustic surveys in the literature, new vocalization patterns are being detected globally @manzano-roth2023 and notable temporal patterns are emerging @yamada2019.
+
+Sperm whales (*Physeter macrocephalus*) also occupy deep ocean habitats, but their dive depths and times are reduced in comparison.
+Sperm whales have been known to dive regularly between 400-600 m and often over 1000 m @watkins1993 @awbery2022 @amano2003.
+They also appear to have more acoustic tolerance than beaked whales @miller2009, @miller2022.
+While avoidance responses in sperm whales have been recorded following naval sonar events @curé2013 @stanistreet2022 @sivle2012 , the full extent is unknown.
+
+Sperm whales also emit series of targeted clicks when socializing (@marcoux2006 @fais2015) and when foraging @fais2016, but it's unlikely they debilitate their prey with their vocalizations as previously hypothesized @norris1972 .
+More is known about sperm whales than beaked whales as they are less visually cryptic and have a
+
+Species studied: \
 - ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
--* BW43, possibly Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed??\
-- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)\
--* BB, --???\
-- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)\
--* PM, sperm whale *(IUCN status: vulnerable)*
-
-sp[erm whales]{.underline} (*Physeter macrocephalus*) may have more acoustic tolerance than beaked whales @miller2009, @miller2022 .
-Look at 3D3 project of behavioral responses of sperm whales to sonar
+-* BW43, Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed in 2021 @baumann-pickering2013 @barlow2021\
+- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)* @simonis2020*\
+**-* BB, --???\
+- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)* @simonis2020*\
+**-* PM, sperm whale *(IUCN status: vulnerable)*
 
@@
 
-The acoustic interference from human activity in BW/SW, for example, can :\
-- affect **behavior** (displacement from habitat, spending time and effort fleeing instead of foraging, increasing call intensity (cite?) (Joy et al.
-2019, @deruiter2013, @aguilarsoto2006 , @stanistreet2022, @joyce2020, @falcone2017, @cholewiak2017 )\
-- disrupt **communication** and echolocation (same as above?)\
-- lead to physical **harm**, even at far distances (up to 100 km away from @falcone2017) like hearing loss, tissue damage, and elevated stress levels. (See @hooker2019 for more info)\
-- **masking** signals for communication (@erbe2014?, ) or environmental cues (@duarte2021)\
-- reduced mating calls (Weilgart, 2007)\
+The specific effects of acoustic interference from human activity in beaked and sperm whales are well-documented.
+Intense noise events can affect behavior through habitat displacement, spending time and effort fleeing instead of foraging, communication and echolocation click disruption, increasing call intensity (cite?) (@deruiter2013, @aguilarsoto2006 , @stanistreet2022, @joyce2020, @falcone2017, @cholewiak2017 ), cause physiological harm like hearing loss, tissue damage, and elevated stress levels (see @hooker2019) even at far distances (up to 100 km away @falcone2017), masksignals for communication (@erbe2014), environmental cues (@duarte2021), or mating @weilgart.\
 - see Cox et al 2006 for impacts of sound on BW
 
-There are several documented cases of naval mid-frequency active sonar (MFAS) activities coinciding with mass stranding events by beaked whales (see @simonis2020, table 1 in @weilgart, @deruiter2013 , @stanistreet2022, @bernaldodequiros, @fernández2005 ).
-[For sperm whales]{.underline}: "The potential impact of full-scale military exercises on sperm whales are not known.
-@stanistreet2022, though there is evidence that male sperm whales stop feeding and change behaviors ( @curé2013).
-
-Under the Marine Mammal Protection Act, all these species are federally protected, several of these species have threatened or endangered conservation statuses (Endangered Species Act), but most remain listed as "data deficient" (cite IUCN).
-\[Also include National Marine Sanctuary Act\].
-Arguments have been made that the term data-deficient does not illicit the same conservation urgency and therefore gets lowered on the priority list (@parsons2016).
-
-Despite their visually cryptic nature, their sound signals are very well-described, down to the species level @baumann-pickering2013, @baumann-pickering2014.
-Beaked whale produce clicks, pulses, and buzzes during foraging starting at the 400(?) m range @tyack.
-The production of sound at this depth may be an adaptation to avoid predation from killer whales, who spend most of its time shallow (from @tyack Baird et al 1998, follow up on this).
-Some are still being discovered and identified globally- see @manzano-roth2023, @yamada2019.
-
-Do beaked whales show temporal variation in their calls? See @ziegenhorn2023
-
-**\[add info about sperm whale echolocation characteristics.\]** Sperm whales also emit series of clicks when socializing (@poupard2022 ).\
+Under the Marine Mammal Protection Act, all these species are federally protected, several receive addition protection under the Endangered Species Act due to a vulnerable or endangered conservation status @taylor @perrin , but some remain listed as "data deficient" ( @pitman @ginkgodens).
+The latter designation has been scrutinized for not illiciting the same conservation urgency as species whose populations are well-documented therefore gets lowered on the priority list (@parsons2016).
 
@@
 It is not fit for all applications or ecological questions, but beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen ( @fleishman ).
+This method provides reliable estimates @arranz2023.
 
@@
 It is characterized by...
-It's a highly productive and complex ecosystem that contains five current and one proposed national marine sanctuaries, supports several wild fisheries, and is home to many protected species.
+It's a highly productive and complex ecosystem that contains five current and one proposed national marine sanctuaries \[add National Marine Sanctuary Act support\], supports several wild fisheries, and is home to many protected species.
 Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area.
@@
 | Covariate                     | Unit | Source                                                          | Justification                                                                                                         |
-|-----------------|-----------------|-----------------|----------------------|
+|-----------------|-----------------|-----------------|---------------------|
 |                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |                                                                                                                       |
@@
 
+**\[Recommendations?\]**
+
 ## Acknowledgements

### 2023-08-16 manuscript.rmd
Update manuscript.rmd
Added to research objective, methods

@@
 There are several documented cases of naval mid-frequency active sonar (MFAS) activities coinciding with mass stranding events by beaked whales (see @simonis2020, table 1 in @weilgart, @deruiter2013 , @stanistreet2022, @bernaldodequiros, @fernández2005 ).
-This occurs for two reasons: 1) their anti-predation adaptation to avoid killer whales (*Orcinus orca*)(synchronized foraging, followed by silent, scattered vertical ascent) mimics that of anthropogenic noise and leads to strandings @aguilardesoto2020; 2) the same physiological adaptations that allow for extreme diving are observed to be deadly when forced to alter behavior and flee the noise source more so than other odontocetes @tyack @cox2006 @joyce2020 .
+This occurs for two reasons: 1) their anti-predation adaptation to avoid killer whales (*Orcinus orca*)(synchronized foraging, followed by silent, scattered vertical ascent) mimics that of anthropogenic noise and leads to strandings @aguilardesoto2020; 2) the same physiological adaptations that allow for extreme diving are observed to be deadly when forced to alter behavior and flee the noise source more so than other odontocetes @tyack @cox2006 @joyce2020 / can increase decompression sickness risk (Kvadsheim et al. [**2012**](https://esajournals-onlinelibrary-wiley-com.jpllnet.sfsu.edu/doi/full/10.1890/15-0040#eap1216-bib-0036), Fahlman et al. [**2014**](https://esajournals-onlinelibrary-wiley-com.jpllnet.sfsu.edu/doi/full/10.1890/15-0040#eap1216-bib-0024) ).
 Furthermore, these responses vary based on the individuals' age, sex, past experience with noise, or current activity @weilgart.
@@
 Sperm whales have been known to dive regularly between 400-600 m and often over 1000 m @watkins1993 @awbery2022 @amano2003.
-They also appear to have more acoustic tolerance than beaked whales @miller2009, @miller2022.
-While avoidance responses in sperm whales have been recorded following naval sonar events @curé2013 @stanistreet2022 @sivle2012 , the full extent is unknown.
+They also appear to have more acoustic tolerance than beaked whales @miller2009, @miller2022 @madsen2002 @miller2009 @winsor2017 .
+While avoidance responses in sperm whales have been recorded following naval sonar events @curé2013 @stanistreet2022 @sivle2012 and seismic survey vessels @bowels 1994 range from minor @isojunno2022 @madsen2002 to moderate @isojunno2016 , the circumstances surrounding the noise exposure is not well understood.
 
 Sperm whales also emit series of targeted clicks when socializing (@marcoux2006 @fais2015) and when foraging @fais2016, but it's unlikely they debilitate their prey with their vocalizations as previously hypothesized @norris1972 .
-More is known about sperm whales than beaked whales as they are less visually cryptic and have a
+More is known about sperm whales than beaked whales as they are less visually cryptic and more success being seen on the surface (visual surveys).
 
-Species studied: \
+Species studied:\
 - ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
 -* BW43, Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed in 2021 @baumann-pickering2013 @barlow2021\
-- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)* @simonis2020*\
-**-* BB, --???\
-- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)* @simonis2020*\
-**-* PM, sperm whale *(IUCN status: vulnerable)*
+- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)* \@simonis\
+- BB, -- Baird's beaked whale *B. bairidii* *(IUCN status: least concern)*\
+- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)* @simonis\
+- PM, sperm whale *(IUCN status: vulnerable)*
 
@@
 The specific effects of acoustic interference from human activity in beaked and sperm whales are well-documented.
-Intense noise events can affect behavior through habitat displacement, spending time and effort fleeing instead of foraging, communication and echolocation click disruption, increasing call intensity (cite?) (@deruiter2013, @aguilarsoto2006 , @stanistreet2022, @joyce2020, @falcone2017, @cholewiak2017 ), cause physiological harm like hearing loss, tissue damage, and elevated stress levels (see @hooker2019) even at far distances (up to 100 km away @falcone2017), masksignals for communication (@erbe2014), environmental cues (@duarte2021), or mating @weilgart.\
-- see Cox et al 2006 for impacts of sound on BW
+Intense noise events can affect behavior through habitat displacement, spending time and effort fleeing instead of foraging, communication and echolocation click disruption, increasing call intensity (cite?) (@deruiter2013, @aguilarsoto2006 , @stanistreet2022, @joyce2020, @falcone2017, @cholewiak2017 ), cause physiological harm like hearing loss, tissue damage, and elevated stress levels (see @hooker2019 and @cox2006) even at far distances (up to 100 km away @falcone2017), mask signals for communication (@erbe2014), environmental cues (@duarte2021), or mating @weilgart.
 
-Under the Marine Mammal Protection Act, all these species are federally protected, several receive addition protection under the Endangered Species Act due to a vulnerable or endangered conservation status @taylor @perrin , but some remain listed as "data deficient" ( @pitman @ginkgodens).
+Under the Marine Mammal Protection Act, all these species are federally protected, several receive addition protection under the Endangered Species Act due to a vulnerable or endangered conservation status @taylor @perrin , but some remain listed as "data deficient" by the International Union for the Conservation of Nature (IUCN) Red List of Threatened Species @pitman @ginkgodens.
 The latter designation has been scrutinized for not illiciting the same conservation urgency as species whose populations are well-documented therefore gets lowered on the priority list (@parsons2016).
@@
 
-Passive acoustic monitoring is a tool that allows us to record and listen to the soundscape.
-It is not fit for all applications or ecological questions, but beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen ( @fleishman ).
-This method provides reliable estimates @arranz2023.
+Passive acoustic monitoring (PAM) is a tool that allows researchers to record and listen to the soundscape.
+It is also an ideal method for assessing habitat use and improving species distribution models of visually cryptic cetaceans like beaked whales and sperm whales that are more likely to be heard than seen ( @fleishman ), live relatively far offshore ( @robbins2022), and have low-density populations ( @hooker2019, @arranz2023 @hildebrand2015).
 
-Bottom-mounted/static: high temporal resolution, low spatial resolution @mellinger
+Passive acoustic recorders come in two forms: fixed or mobile, each better suited to a certain ecological application.
+A mobile (towed or free-drifting) buoy moves either behind a vessel or autonomously with the currents for days or weeks at a time.
+It captures the dynamic nature of sound over a large spatial area, especially in deep and less studied parts of the ocean, making it ideal for detecting offshore or mobile species.
+Bottom-mounted recorders allow for greater temporal coverage and are often used when studying specific regions or known habitats of study species ( @nosal2007, @marques2012 @jarvis2022).
 
-Acoustic tags:
-
-Towed arrays:
-
-Tradeoffs include costly equipment or deployment, data/battery storage requirements, chance of return, spatial coverage, temporal coverage, capturing fine-scale movements, number of individuals, parsing out individuals, through inclement weather or night-time @johnson2009 @fleishman
-
-One type of passive acoustic recorder called a drifting buoy that drifts freely with the currents for days or weeks at a time.
-It allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters.
-Buoys offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species.
-Studying beaked whales can be challenging due to how offshore they live (@robbins2022).
-The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only.
-Historically, it has been challenging to record foraging calls at several hundred meters of depth @barlow2005.
-
-Its basic components consist of a satellite transmitter above water to monitor GPS information and an acoustic recorder attached by line at \~100m of depth.
-In fact, most of what is known about these deep divers is because of acoustic monitoring.
-Similar designs of drifting buoys have been used in estuarine and coral reef soundscape studies @lillis2018.
+The benefits of using PAM over visual surveys include use during night hours or through inclement weather, cost to charter and @johnson2009 @fleishman.
+The trade-offs for using PAM include costly equipment or deployment, data/battery storage requirements, uncertain chance of return, limited spatial coverage for fixed recorders, and limited temporal coverage for mobile recorders.
+Even with experienced acousticians, noisy detections can obfuscate number of individuals, species detected, or distance or direction from recorder for species with large call ranges such as sperm whales @barlow2005 @mellinger2007.
 
@@
 
-Automatic Identification System.
-Ships are abundant throughout the study area due to major US ports.
-This was collected to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.
+Automatic Identification System (AIS) is a mandatory vessel tracking system for vessels over 300 gross tons to aid in position communication and safe navigation @uscg .
+While not originally intended for use in research, it has become a valuable resource for vessel traffic data, especially in studying effects of vessel strikes @greig2020 @reimer2016 and noise pollution @erbe2014 @aguilarsoto2006 on cetaceans.
+Because ships are abundant throughout the study area due to major US ports, AIS was included in this study to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.
 
-@aguilarsoto2006 -- elevated sound levels during close encounters between ships and beaked whales can be detrimental.
+Previous studies have separated vessel and anthropogenic noise from ambient soundscape based on known spectral signatures, frequency bands, and percentiles @weiss2021 @haver2020.
+The inclusion of AIS to soundscape metrics in this study provides fine-scale evaluation of sound and source.
 
-How have other studies separated vessel noise (contribution to soundscape) from ambient soundscape? See @weiss2021 and @haver2020. Power spectral density ambient sound levels.
+Power spectral density ambient sound levels (see @haver2020 for details on what percentiles mean)
 
-**Laws and Regulations** \[Where and to what extent is noise pollution regulated?\]
+**Laws and Regulations**
 
-see @chou2021 for a list of documents, reports, programs, efforts in ocean noise reduction
+While there is general consensus that anthropogenic noise in the ocean is increasing, policies that aim to mitigate chronic noise are generally lacking.
+This is due to knowledge gaps regarding the long-term and cumulative effects of noise on certain species @markus2018.
 
-Marine Strategy Framework Directive by European Union establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise.
-@directiv2008, @borsanij.f.2023 Within their Marine Strategy Framework Directive, the European Union is the only entity to date that has recognized noise as a form of pollution and restricts noise from reaching levels that may cause negative effects to the environment @directiv2008.
-[What about UK Marine Strategy?] See @chou2021 for larger list of efforts.
-
-For mid-frequency cetaceans, auditory impairment threshold levels are as follows:\
-peak sound pressure level (Impulsive) - 230 dB, recommended accumulation period of 24 hrs (impulsive) - 185 dB, recommended accumulation period of 24 hrs (impulsive) - 198 dB (@noaa2018) .
-
-IMO guidelines from 2014 are being updated by July 2023. @internationalmaritimeorganization
-
-Canada is developing an Ocean Noise Strategy
-
-In the US, it's difficult to achieve compliance from Navy.
+International and regional efforts are underway (see @chou2021) , but still have a ways to go.
+Within the European Union, the Marine Strategy Framework Directive establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise @directiv2008, @borsanij.f.2023.
+Some studies have independently tested @haver2021 @garrett2016 @sebastianutto2015 or used @mustonen2019 these metrics.
+The National Oceanic and Atmospheric Administration (NOAA) has developed an Ocean Noise Strategy Roadmap, an agency-wise approach to mitigating both chronic and acute effects of noise exposure to animals @noaa2018.
+Within it, they agree upon impairment thresholds for mid-frequency cetaceans (@noaa2018) .
+Because chronic effects are not fully understood, they remain to be regulated.
+The International Maritime Organization (IMO) has also been addressing noise pollution from commercial vessels through quieter ship design @internationalmaritimeorganization.
+Updates to their 2014 guidelines are set to be finalized by January 2024.
+Canada is also developing an Ocean Noise Strategy @breeze2022.
 
@@
 
+With anthropogenic noise-generating activities on the rise, global oceanic sound levels are expected to increase.
+Noise is a form of habitat degradation, so there is a need for research on a large spatial scale to understand how these highly mobile animals overlap in space and time with human activities and how they're affected @haver2019 @hooker2019 @erbe2014.
+Passive acoustic monitoring Acquiring information on these federally protected and acoustically-sensitive species is challenging\
 \
-Shipping and offshore activities are expected to increase in the future, causing elevated oceanic sound levels.
-This is a form of habitat degradation, so there is a need to understand how human activities change and contribute to the overall soundscape in space and time.
-It's a challenge to acquire information on these federally protected and acoustically-sensitive species but becoming increasingly important to understand where habitats overlap with human activity.
-Acoustics can serve as important indicators of ecosystem health, but also of how humans are changing the environment.\
-\
-Characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics and vessel tracking.
-Characterizing the acoustic environment and identifying co-occurrence of whales and anthropogenic sound has been done in a similar study by @weiss2021.
-This serves as baseline monitoring of soundscape for offshore areas.
+Logistical difficulties associated with offshore research and low-density populations of visually cryptic deep diving whales have created a paucity of data in the California Current Ecosystem.
+This research provides a step towards understanding soundscape variation in the area and beaked and sperm whale distributions and their acoustic environment.
 
-**\[Gaps in literature, rationale for research\]**
+The purpose of this study is to characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics, environmental variables, and AIS.
+We determine significant predictors of whale presence for each species and provide explicit sources for noise-generating human activities through overlay of large vessel activity.
 
-"To date, studies have focused on this and this, but not on this".This work will contribute to establishing baseline conditions of noise exposure to inform wind farm development and shipping lane changes.
-\[add here\]
-
-It'll also add to the knowledge base of beaked and sperm whale distribution through acoustic detections.
-
-Both of these outcomes are of high importance to managers and agencies especially NOAA and BOEM (see 2023-2024 report).
-Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic sound sources contribute to the soundscape.
-Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity.
-With AIS, we can calculate distance from buoy to vessel and sound levels in those moments.
-And 3) NOAA is seeking more information on vocally active species' distributions.
-This is part of a bill introduced by Congress to "assess underwater sound in high-priority" environments (https://www.congress.gov/bill/117th-congress/house-bill/6987).
-
-\[add here\]
-
-Quantifying soundscapes in a standardized way and monitoring these changes over time are priorities for marine managers and researchers right now.
-Connecting human noise sources to whale habitat in space and time is needed for efficient management @haver2019.
-
-"Increased understanding of temporal and spatial distribution of beaked whales will enable more effective planning for noise-generating activities to occur where and when fewer sensitive animals are present." @hooker2019
-
-"Very large-scale studies, conducted on the spatial, temporal and spectral scales at which these highly mobile predators live their lives, may be needed to evaluate the influence of changes in a marine mammal's acoustic environment." @erbe2014
-
-"Research should be conducted on how the level of anthropogenic activity (such as the types and numbers of vessels) is related to resulting noise levels." @hildebrand2009
+This study highlights spatial and temporal acoustic variability within the study region. It will also contribute to noise exposure levels through acoustic detections.
 
@@
 
+The California Current Ecosystem (CCE) is a highly productive and complex ecosystem that contains one proposed and five current national marine sanctuaries @sanctuaries , supports several wild fisheries @field2006 , and is home to nineteen protected whale species @fisheries.
+The 1,141,800 km2 area extends along the US' Economic Exclusion Zone from south of Vancouver to north of Baja California and is a characterized by deep canyons, continental shelf @barlow2009 @barlow2007.
+Characteristic of the area are upwelling events that carry nutrients to the surface often drive primary productivity, creating an ideal habitat where whales' prey are found and driving them there @ryther1969.
+The year 2018 was characterized in the CCE with slightly warmer than average sea surface temperatures as the recent El-Nino and marine heatwave events were ending @harvey.
+Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area @peña2007 @coleman2008 @oldach2022.
 \
-The California Current Ecosystem extends along the US' Economic Exclusion Zone along the west coast.
-It is characterized by...
-It's a highly productive and complex ecosystem that contains five current and one proposed national marine sanctuaries \[add National Marine Sanctuary Act support\], supports several wild fisheries, and is home to many protected species.
-Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area.
-"Ambient noise off the coast of California has increased many-fold over the past several decades..." (@moore2013 + other sources w/in that).
-
+\
 In 2021, California, Washington, and Oregon ranked 3rd, 5th, and 23rd in most trafficked U.S.
 waterways, respectively (@unitedstatesarmycorpsofengineersusace2018.
-Lots of overlap between vessel activity from busy ports and important habitat for protected species.
-
-\[Can discuss the warm and cool phases that occurred during study period.\]
-
-see @harvey for report on conditions
+Between vessel activity like fishing, shipping and naval exercises in California alone, there is notable overlap of anthropogenic presence with important habitat for protected species.
+Furthermore, the future development of offshore wind turbines off of Morro Bay and Humboldt, California are expected to be noise-producing activities during construction and operation, increasing sound levels in those regions @cooperman @madsen2006.
 
@@
 
-**\[Overall data collection efforts\]**\
-Acoustic detections were collected with the use of free-drifting Drifting Acoustic Spar Buoy Recorders (hereafter buoys).
-These were deployed as part of the California Current Ecosystem Survey (CCES) spanning the offshore waters from southern Canada to northern Mexico in summer of 2018.
-These buoys contained two satellite geo-location trackers, two hydrophones, etc and collected a total of 1910 hours of data.
-SWFSC ran acoustic detections in PAMGuard.
-Eight buoys were lost at sea and therefore removed from the analysis.
-These contain 1, 5, 10, 25, 75, 95, and 99 percentile distribution from each metric.
+**Overall data collection efforts**\
+Acoustic detections were collected as part of the 2018 California Current Ecosystem Survey (CCES) by the Southwest Fisheries Science Center (SWFSC).
+Twenty-three free-drifting buoys (numbered 1-23) were deployed at varying times between July and November in the offshore waters of the study area.
+These buoys, numbered 1-23, contained two satellite geo-locators above water to monitor GPS information and two hydrophones at depth attached by a 150m line.
+They collected a total of 1910 hours of data.
+SWFSC detected and classified acoustic detections in PAMGuard to species level as per methods mentioned in @simonis2020 .
 
-**\[Add details about acoustic data collection information.\]**\
-Soundscape metrics were duty-cycled (most 18min off, 2 minutes on, but some variation in "off" times) and later processed by the NOAA team with calls and clicks identified to the species level (see @simonis2020 for details on this survey and @rand2022 for duty-cycle study).
-Soundscape metrics were binned into low, medium, high frequency bands.
-Can include software used, sampling rate?
+Acoustic data were duty-cycled, with most buoys sampling at a rate of 18 minutes off and 2 minutes on.
+The SWFSC team analyzed the data and detected echolocation pulses produced by beaked and sperm whales.
+These were classified by peak frequency and spectral signature and identified to the species level (see @simonis2020 for details on this survey and @rand2022 for duty-cycle study).
+Soundscape metrics were binned by peak frequency into low, medium, high frequency bands, along with the 1, 5, 10, 25, 75, 95, and 99 percentile distribution from each metric.
 
-**\[Post-processing details, data cleaning\]**\
-Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to UTC.
-These include third-octave level bands and broadband (20-24,000 Hz) metrics.
-~~One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.~~ The band at 125 Hertz is often associated with shipping vessel noise and the 2-5 kHz range with mid-frequency naval sonar ( @garrett2016 ).
-Beaked whales are known to be sensitive to sonar in this latter range (cite).
-Several TOL levels were chosen as proxies to vessel or sonar frequencies, ranging from TOL 63 to 20,000.
+Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to the UTC time zone.
+These metrics include third-octave level (TOL) bands from 63 to 20,000 Hz and broadband (20-24,000 Hz) metrics.
+Several TOL levels were chosen as proxies to vessel or sonar frequencies.
 
-Whale detections were rounded to nearest 20 minutes.
-A separate column was created as a common join field that contained nearest timestamp in UTC to that of drifter tracks.
-Two types of BW species (?BW and ?BWC) were removed from analysis as the species is unknown and unsure.
-Whale detections were reduced to "presence/absence" from number of clicks, estimated individuals, etc.
+Whale detection timestamps were rounded to nearest 20 minutes based on start time to align with the soundscape metrics.
+A separate column was created as a common join field that contained nearest timestamp in UTC to that of buoy tracks (see below).
+Two types of beaked whale detections (species code ?BW and ?BWC) were removed from analysis as the species was unknown.
 
-Drifter tracks were rounded to nearest 20 minutes, lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17) were removed.
-Drift 14 and 15 are the same.
-These three major datasets were joined together based on a common buoy number (column "station") and time field column "UTC").
+Buoy tracks were similarly rounded to nearest 20 minutes.
+Seven buoys (1,2,3,5,6,9,11) were lost at sea and two buoys had corrupted data (4,17) so these were removed from the analysis.
+Buoy 14 and 15 were the same and therefore consolidated to 14.
+These three preliminary datasets were joined together based on a common buoy number (column "station") and time field column "UTC").
+Whale detections (column Wpresence) were reduced to binary data for absence/presence and all other fields of that dataset were eliminated.
 
-**\[Environmental Data\]**\
-Environmetal and oceanographic variables thought to be predictors of BW/SW habitat were extracted from NOAA's Environmental Research Division Data Access Program (ERDDAP) and Hybrid Coordinate Ocean Model (HYCOM) using matchEnvData function in package PAMmisc.
-Variables at depth were calculated using function from Taiki and formula from @mccullough2021.
-Distance to slope and bathymetric slope were calculated from a bathymetry TIF taken from GEBCO.
+**Environmental Data**\
+Possible environmental and oceanographic predictors of cetacean habitat were extracted from the NOAA Environmental Research Division Data Access Program (ERDDAP) server ((https://coastwatch.pfeg.noaa.gov/erddap/) and Hybrid Coordinate Ocean Model (HYCOM) using the *matchEnvData* function in R package *PAMmisc*.
+Six modeled variables at depth (mixed layer depth, mixed layer temperature, temperature at 400m, thermocline temperature, thermocline depth, salinity at 400m) were calculated using a formula from @mccullough2021.
+Such variables are especially important to incorporate as it better reflects the deep environment that beaked and sperm whales are known to inhabit.
+Distance to slope and bathymetric slope were calculated from a bathymetry TIF acquired from the General Bathymetric Chart of the Oceans (GEBCO).
 
@@
 
-use @torreblanca2022 for deep habitat info\
-see @johnston2008 and @manzano-roth2023 for seamount habitat importance
+@forney2012 shows how BW model had a lot of uncertainty regarding distribution predictors. So few detections that some spp had to be lumped together (also @barlow2009).
 
 | Covariate                     | Unit | Source                                                          | Justification                |
-|-----------------|-----------------|-----------------|---------------------|
-|                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |                                                                                                                       |
-| Distance to continental slope |      | General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015 |                                                                                                                       |
-| Bathymetric slope             |      | GEBCO tif                                                       |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
-| Wind curl stress              |      | ERDDAP                                                          | "Positive wind stress curl lifts the thermocline, which results in enhanced primary production" (Manocci et al 2015).
-|
-|                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
-|                               |      |                                                                 |                                                                                                                       |
+|----------------|----------------|----------------|------------------------|
+| Salinity at 400m              |      |                                                                 |                              |
+| Distance to continental slope |      | General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015 | @barlow2009                  |
+| Bathymetric slope             |      | GEBCO tif                                                       | @becker2010                  |
+| Depth                         |      |                                                                 | @forney2012 @becker2010      |
+| Wind curl stress              |      | ERDDAP                                                          | @mannocci2015                |
+| MLD depth                     |      |                                                                 | @mccullough2021, @forney2012 |
+| Thermocline temp              |      |                                                                 | @mccullough2021              |
+| Thermocline depth             |      |                                                                 |                              |
+| MLD Temp                      |      |                                                                 | @mccullough2021              |
+| Temperature at 400m           |      |                                                                 |                              |
+| SST mean                      |      |                                                                 | Becker 2007, @forney2012     |
+| CHL                           |      |                                                                 | @forney2012                  |
+| SSH                           |      |                                                                 |                              |
 
 **\[AIS\]**\
-Automatic Identification System vessel tracking data were collected from Marine Cadastre in order to evaluate the effect of spatial proximity to the buoys.
+Automatic Identification System data were collected in order to evaluate the effect of spatial proximity to the buoys.
 This step is important in understanding how ship characteristics contribute to various frequency bands.
@@
 
-AIS was downloaded, processed (converted to UTC), and matched to each drifter using Taiki's PAMscapes package and functions.
+AIS was downloaded from Marine Cadastre, processed, converted to UTC timezone, and matched to each buoy's timestamp using the package *PAMscapes* and function \_\_.
 
@@
 
-[Example of GAM]{.underline}: @mannocci2015
+[Example of GAM]{.underline}: @mannocci2015, Becket et al 2020, Becker 2007
 
@@
 
-[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day ( @ziegenhorn2023 ) , drift with most detections/day)]
+[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day ( @ziegenhorn2023 ) , drift with most detections/day), where/when were most detections heard, number of species]
 
-\[Weekly/daily plot of acoustic presence of whales and ships over time from each drift? See JS Trickey et al. 2022, Fig 3\]
+\[Weekly/daily plot of acoustic presence of whales and ships over time from each drift? See JS Trickey et al. 2022, Fig 3\] Which buoy had the most AIS overlap and when?
 
 Look at power spectral density? @haver2020
+
+The band at 125 Hertz is often associated with shipping vessel noise and the 2-5 kHz range with mid-frequency naval sonar ( @garrett2016 ).
+Beaked whales are known to be sensitive to sonar in this latter range (cite)
 
@@
 
+Does not account for the differences in relative noise contribution by vessel type (as @hatch2008 does)
+
 \[Add here.\]
+
+This work will contribute to establishing baseline conditions of noise exposure to inform wind farm development and shipping lane changes.
+In better understanding this, human activities can be altered to avoid overlap with sensitive species @hooker2019.
+
+Both of these outcomes are of high importance to managers and agencies especially NOAA and BOEM (see 2023-2024 report).
+Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic sound sources contribute to the soundscape.
+Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity.
+With AIS, we can calculate distance from buoy to vessel and sound levels in those moments.
+And 3) NOAA is seeking more information on vocally active species' distributions.
+This is part of a bill introduced by Congress to "assess underwater sound in high-priority" environments (https://www.congress.gov/bill/117th-congress/house-bill/6987).
+
+Outside of cetacean research, similar designs of drifting buoys have been used in estuarine and coral reef soundscape studies @lillis2018.
 

### 2023-08-16 07_summary_stats.Rmd
made note about prelim results

@@
 
-Where are loudest sound levels?
+Where are loudest sound levels? See @haver2020 for details on percentiles
+5th percentile = exceeded most of the time, so refers to background sound
+90th percentile = exceeded 10% of the time, so refers to PEAKS in sound level
+Also see @haver2021 (fig 2) for daily TOL levels by site (can do by buoy)
 ```{r}

### 2023-08-16 04_analysis.Rmd
added code to save allDrifts as rda

@@
                    fulltrack16, fulltrack18, fulltrack19, fulltrack20, fulltrack21, fulltrack22, fulltrack23)
+#save(allDrifts, file = "allDrifts.rda")
 ```

### 2023-08-16 02_AIS.Rmd
overwrote old code, added Taiki's code

--- 2023-04-21 02_AIS.Rmd
@@
 ---
-```{r}
-#Required Packages
 
+#Introduction
+AIS download from Marine Cadastre and match to buoy datasets.
+
+Required Packages
+```{r, }
+library(PAMscapes)
 library(tidyverse)
-library(sf)
-library(gganimate)
-#install.packages("remotes")
-#remotes::install_github("skgrange/gissr")
-#devtools::install_github("skgrange/gissr")
-library(gissr)
-
-# Function
-# Speeding up st_intersection (from https://github.com/r-spatial/sf/issues/801)
-
-st_intersection_faster <- function(x,y,...){
-#faster replacement for st_intersection(x, y,...)
-
-  y_subset <-
-    st_intersects(x, y) %>%
-    unlist() %>%
-    unique() %>%
-    sort() %>%
-    {y[.,]}
-
-  st_intersection(x, y_subset,...)
-}
-
 ```
 
+-----------------------
+From Taiki's AIS Workflow
+
 ```{r}
-#########
-# AIS
-#########
-
-# Need to add more AIS dates in data folder
-AISlist <- list.files(path = paste0("data-raw/"), pattern = "AIS_", recursive = TRUE)
-
-for (v in AISlist) {
-  Adata <- read_csv(paste0("data-raw/", v), show_col_types = FALSE) %>%
-    filter(Length >= 100) %>%                  # Filter out vessels over 100m in length
-    dplyr::select(-SOG,-COG, -CallSign, -IMO, -Cargo, -Draft, -Status) %>%  #Remove unnecessary columns
-    mutate(variable = paste0(substr(v, 16, 17), substr(v, 6, 7)))
-
-  assign(paste0("AIS_", substr(v, 5, 10)), Adata)
-}
-
-AIS_sf <- AIS_091218 %>%
-  st_as_sf(coords = c("LON", "LAT"), crs=4326, remove=F)
+#marcadFiles07 <- downloadMarCadAIS(tracks_SS_07, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles08 <- downloadMarCadAIS(tracks_SS_08, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles10 <- downloadMarCadAIS(tracks_SS_10, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles12 <- downloadMarCadAIS(tracks_SS_12, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles13 <- downloadMarCadAIS(tracks_SS_13, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles14 <- downloadMarCadAIS(tracks_SS_14, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles16 <- downloadMarCadAIS(tracks_SS_16, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles18 <- downloadMarCadAIS(tracks_SS_18, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles19 <- downloadMarCadAIS(tracks_SS_19, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles20 <- downloadMarCadAIS(tracks_SS_20, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles21 <- downloadMarCadAIS(tracks_SS_21, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles22 <- downloadMarCadAIS(tracks_SS_22, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles23 <- downloadMarCadAIS(tracks_SS_23, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
 ```
 
+Subset to US West Coast
 ```{r}
-##################################
-# Create 10 km buffer around buoys (taken from M. McKenna's code)
-# Should be malleable for each buoy's number to easily plug in.
-##################################
-tracks <- readRDS("data-raw/tracks.rda")
+subsetMarCadAIS(inDir="D:/AIS", outDir="D:/AIS/AIS_West", name='West_', 
+                latRange=c(20,50), lonRange=c(-140,-110),
+                overwrite=FALSE, progress=TRUE)
+```
+Now read into R only ais points that come in contact with buoys
+```{r}
+ais_near_07 <- readLocalAIS(tracks_SS_07, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
 
-trk8 <- tracks %>% 
-  dplyr::filter(station == "8") %>%
-  st_as_sf(coords= c("Longitude", "Latitude"), crs=4326, remove=F)
+ais_near_08 <- readLocalAIS(tracks_SS_08, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
 
-pois <- st_transform(trk8, crs = 4326)    # Specify a crs
-pts_buff <- st_buffer(pois, dist = 10000)  # Creates 10 km buffer around buoy points 10 km (10000m)
-buff <- st_transform(pts_buff, crs = 4326)  # Specify crs for new buffer
+ais_near_10 <- readLocalAIS(tracks_SS_10, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
 
-### Crop full AIS to 10km buffer around drift 8 ###
-AIS_crop <- st_crop(x = AIS_sf, y = buff)
+ais_near_12 <- readLocalAIS(tracks_SS_12, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+ais_near_13 <- readLocalAIS(tracks_SS_13, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+ais_near_14 <- readLocalAIS(tracks_SS_14, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+ais_near_16 <- readLocalAIS(tracks_SS_16, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+ais_near_18 <- readLocalAIS(tracks_SS_18, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+ais_near_19 <- readLocalAIS(tracks_SS_19, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+ais_near_20 <- readLocalAIS(tracks_SS_20, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+ais_near_21 <- readLocalAIS(tracks_SS_21, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+#empty table because no large vessels interacted with this buoy
+
+ais_near_22 <- readLocalAIS(tracks_SS_22, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+ais_near_23 <- readLocalAIS(tracks_SS_23, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+   filter(vesselLength >= 100)  #remove vessels under 100m in length
 ```
 
+Match AIS points with buoy points
 ```{r}
-# Intersect buffer with AIS. Starting this on first 10 rows to see how slowly it goes. 
-## FIX ME
-#inters <- st_intersection(buff, AIS_crop[1:140,])
-
-#Another way to view AIS points that fall within buffer. DOES NOT consider time.
-AIS_within_buff <- AIS_crop[buff,]
-f <- buff[AIS_within_buff,]
-by <- join_by(closest(UTC >= BaseDateTime))
-
-d <- full_join(buff, as.data.frame(AIS_within_buff), by)
-
-e <- d[!duplicated(d[c('BaseDateTime')]),]
-
-trk_AIS <- left_join(as.data.frame(trk8), as.data.frame(AIS_within_buff), by)
-
-proj4string <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
-trk_AIS_sf <- trk_AIS %>% st_as_sf(coords = c("Longitude", "Latitude"), crs=projcrs, remove=F)# %>%
-   st_set_crs(4326)
-trk_AIS_sf <- st_transform(trk_AIS_sf, crs = 4326)
-
-#inters <- st_intersection_faster(buff, AIS_crop[1:140,])
-#inters2 <- st_intersection_faster(buff, AIS_crop[141:200,])
-#inters3 <- st_intersection_faster(buff, AIS_crop[201:300,])
-
-#Another way to view AIS points that fall within buffer. DOES NOT consider time.
-
-st_is_within_distance(AIS_crop, pois, dist = 10000)
-st_distance(AIS_crop, pois, dist_fun = 10000)
-
-st_set_geometry(AIS_crop, value = NULL)
-st_set_geometry(pois, value = NULL)
-
-AIS_crop <- as_Spatial(AIS_crop)
-pois <- as_Spatial(pois)
-AIS_within_buff <- as_Spatial(AIS_within_buff)
-z <- sp_distance(AIS_crop, pois, features = TRUE, unit = "km")
-
-data_places <- pois@data
-data_places$distance <- sp_distance(AIS_within_buff, pois, features = TRUE, unit = "km")
-View(data_places)
-
-rgeos::gWithinDistance(AIS_crop, pois, features = TRUE, dist = 10000)
-```
-```{r}
-# For each row in AIS df,
-  # calculate distance to closest drift
-  # return smallest distance
-  # attach column to df
-
-# 
-(geosphere::distHaversine(AIS_within_buff[9:20,], pois))/1000
+gpsClose07 <- addAIS(tracks_SS_07, ais_near_07, interpType = 'close', interpTime = 120)
+gpsClose08 <- addAIS(tracks_SS_08, ais_near_08, interpType = 'close', interpTime = 120)
+gpsClose10 <- addAIS(tracks_SS_10, ais_near_10, interpType = 'close', interpTime = 120)
+gpsClose12 <- addAIS(tracks_SS_12, ais_near_12, interpType = 'close', interpTime = 120)
+gpsClose13 <- addAIS(tracks_SS_13, ais_near_13, interpType = 'close', interpTime = 120)
+gpsClose14 <- addAIS(tracks_SS_14, ais_near_14, interpType = 'close', interpTime = 120)
+gpsClose16 <- addAIS(tracks_SS_16, ais_near_16, interpType = 'close', interpTime = 120)
+gpsClose18 <- addAIS(tracks_SS_18, ais_near_18, interpType = 'close', interpTime = 120)
+gpsClose19 <- addAIS(tracks_SS_19, ais_near_19, interpType = 'close', interpTime = 120)
+gpsClose20 <- addAIS(tracks_SS_20, ais_near_20, interpType = 'close', interpTime = 120)
+gpsClose21 <- addAIS(tracks_SS_21, ais_near_21, interpType = 'close', interpTime = 120) #empty table
+gpsClose22 <- addAIS(tracks_SS_22, ais_near_22, interpType = 'close', interpTime = 120)
+gpsClose23 <- addAIS(tracks_SS_23, ais_near_23, interpType = 'close', interpTime = 120)
 ```
 
+Create column 'shipsin10km' that displays number of ships associated with timestamp
+Need to make a loop!
 ```{r}
-# Plot showing 10km buffer around drift 8
-#tmap_mode("plot")
-#tm_shape(buff) + tm_symbols(col = "blue", border.col = NA) +
-#  tm_shape(trk) + tm_dots(col = "red")
+# Viewing only ships within 10 km of buoys.
+close2ships7 <- gpsClose07 %>% filter(shipDist < 10e3)
+
+#Add column of ROUNDED UTC values to the lowest minute. Allows common field to bind
+z <- as.POSIXct(close2ships7$UTC)
+close2ships7$UTCrnd <- lubridate::floor_date(z, "minute")
+
+#Bind rows with ship nearby
+#Summarize unique ships and display number of nearby vessels
+nearbyShp07 <- inner_join(tracks_ss_07, close2ships7, by = join_by("UTC" == "UTCrnd")) %>%
+  select(UTC, UTC2, MMSI, vesselLength, vesselType, shipLat, shipLong, shipDist) %>%
+  group_by(UTC2) %>%
+  summarize(shipin10km = n_distinct(MMSI))
+
+# ready to join to main df and add zeros to Null values
 ```
-
-plots
-```{r}
-# Plot
-ggplot(tracks_SS_08, aes(x = long, y = lat.x)) + geom_point() 
- + geom_sf(data = bbox8, aes(col = "red"))
-
-# Plot showing AIS bounding box around drift 8 tracks
-ggplot() + geom_sf(data = trk_AIS_sf, color = "red")
-ggplot() + geom_sf(data = buff[lgth,] ,color = "red")
-
-ggplot() +
-  geom_sf(data = calif) + coord_sf(default_crs = sf::st_crs(4326)) + theme_bw()
- # geom_sf(data = AIS_sf)
-
-p <- ggplot() + geom_sf(data = buff, mapping = aes(x=Longitude,y=Latitude)) +
-  geom_point(data = AIS_within_buff, mapping = aes(x=LON,y=LAT, group = VesselName), color = "red")
-
-anim <- p +
-  transition_reveal(along = BaseDateTime)+
-  ease_aes('linear')+
-  #shadow_wake(0.1) +
-  ggtitle("Date: {frame_along}")
-
-animate(anim, nframes = 180, fps = 3)
-```

### 2023-08-16 RF_ZC_presence_7-11_edits.R
Update RF_ZC_presence_7-11_edits.R
ran randomforest instead of rfpermute

@@ -7,6 +7,9 @@ library(randomForest)
 library(dplyr)
 library(rfPermute)
 
+# Load in main dataframe
+load(file = "allDrifts.rda")
+
 #Define response variable (presence/absence as 0/1)
 #I suggest creating a variable for presence of each species, and then create a
 #random forest model for each species 
@@ -51,12 +54,25 @@ set.seed(123)
 # create balanced sample size random forest model
 #By ZC
 RF.model.ZC <- rfPermute(ZcPresence ~ ., data=DF.modelZC[,include.covars],
-                            replace=FALSE, ntree=10000, sampsize=sampsizeZC, proximity=FALSE)
+                            replace=FALSE, ntree=15000, sampsize=sampsizeZC, proximity=FALSE)
 RF.model.ZC
+#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# Use random forest, not rfPermute
+# This code associated with PDF from 8/15/23 at 2:53 am
+RanFor.model.ZC <- randomForest(ZcPresence ~ ., data=DF.modelZC[,include.covars],
+                         replace=FALSE, ntree=70000, sampsize=sampsizeZC, 
+                         proximity=FALSE, importance = TRUE)
+RanFor.model.ZC
 
-#Playing around with no balanced sample size
-RF.model.ZC.nosamp <- rfPermute(ZcPresence ~ ., data=DF.modelZC[,include.covars],
-                         replace=FALSE, ntree=10000, proximity=FALSE)
+pdf(paste(Dep,"_",string.covars.used,".pdf"), width=14, height=10)
+round(importance(RanFor.model.ZC),3) #ranking of importance for each variable
+plotImportance(RanFor.model.ZC) #visualize importance
+plotTrace(RanFor.model.ZC)   #model stability w/number of trees (this should be flat!)
+plotImpPreds(RanFor.model.ZC, DF.modelZC, "ZcPresence")  #distribution of predictors 
+plotPredictedProbs(RanFor.model.ZC)
+dev.off()
+save.image(paste(Dep,"_", string.covars.used, ".RData", sep=""))
+#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
 #Evaluate RF models
 #Create a PDF of summary plots
[…]

### 2023-09-13 data-raw/whales.R
removed rounding, prepped for join

@@ -1,32 +1,23 @@
 ## code to prepare `whales` dataset here
 
+# Required library
 library(tidyverse)
-library(sf)
 
 # Load raw RDA file
-#load("data-raw/CCES2018_BW_Detections.rda")             # Beaked whales only
-load("data-raw/CCES2018_BW_and_PM_Detections.rda")       # Beaked + sperm whales
+load("data-raw/CCES2018_BW_and_PM_Detections.rda")
 
 # Change EventInfo object name to "whales" and edit columns
 whales <- EventInfo %>%
-  dplyr::select(-Project, -UID, -Id) %>%            # remove unnecessary columns
-  subset(species!= "?BW") %>%                    # remove rows containing ?BW and BW
-  subset(species!= "BW") %>%
-  mutate(dateTimeRound = round_date(.$StartTime, "20 minutes")) %>%   # Round to nearest 20 min
-  rename(station = Deployment)                    # Rename for common join column name next
+  dplyr::select(-Project, -UID, -Id, -minNumber, -maxNumber, -bestNumber, -nClicks) %>%   # remove unnecessary columns
+  subset(species!="?BW" & species!="BW") %>%        # remove rows containing ?BW and BW
+  mutate(UTC = StartTime) %>%
+  filter(!Deployment %in% c('1','2','3','4','5','6','9','11', '17'))
 
-## NEED TO CHANGE STATION 15 TO STATION 14 HERE. Drift 15 and 14 are the same.
-whales$station[whales$station == 15] <- 14
-
-# Create common UTC field and remove unnecessary fields
-whales <- left_join(whales, tracks, by = join_by(station, closest("dateTimeRound" <= UTC))) %>%
-  dplyr::select(-Latitude.y, -Longitude.y, -spotID, -minNumber, -maxNumber, -bestNumber)
-
-# Whales data as sf
-Wsf <- whales %>% st_as_sf(coords = c("Longitude.x","Latitude.x"), crs=4326)   # whales as sf
+## Change Station 15 to Station 14 here. Drift 15 and 14 are the same (mix-up during fieldwork)
+whales$Deployment[whales$Deployment == 15] <- 14
 
 # Save object
 saveRDS(whales, 'data/whales.rda')
 
-# Number of rows per station
-nrow(whales[whales$station==12,])
+
+

### 2023-09-13 data-raw/tracks.R
Updated
removed rounding, updated code with NEW csv

@@ -2,35 +2,17 @@
 
 library(tidyverse)
 
-# Load raw rda file
-load("data-raw/CCES2018_DriftTracks_Modified_03Nov2022.rda")
+# Read in csv with GPS points
+AllTracks_raw <- read_csv("data-raw/AllCCES_GPS.csv")
 
 # Read in tracks, filter out lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17)
-tracks <- AllTracks %>%
-  dplyr::select(-dist, -speed) %>%
-  mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>% # round to nearest 20 minutes, will create duplicated in the data
-  filter(!station %in% c('1','2','3','4','5','6','9','11', '17')) %>%
-  rename(Longitude = long,              # Need these columns to specifically say this for ERDDAP matching
-         Latitude = lat,
-         UTC = dateTime)
-tracks <- tracks[!duplicated(tracks[c('UTC', 'station')]),]    # Remove duplicates
+AllTracks <- AllTracks_raw %>%
+  dplyr::select(Latitude, Longitude, UTC, DriftName) %>%
+  #mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>% # round to nearest 20 minutes, will create duplicated in the data
+  filter(!DriftName %in% c('CCES_004','CCES_017'))
+
+AllTracks <- AllTracks[!duplicated(AllTracks[c('Longitude', 'Latitude','UTC', 'DriftName')]),]    # Remove duplicate timestamps
 
 # Save as RDS
-saveRDS(tracks, 'data/tracks.rda')
-
-# Filter by buoy
-#tracks_08 <- tracks %>% dplyr::filter(station == "8") 
-track4 <- AllTracks %>% dplyr::filter(station == "4")
-
-
-
-### Option to create AllTracks before rounding to nearest 20 min, for AIS
-tracks2 <- AllTracks %>%
-  dplyr::select(-dist, -speed) %>%
-  filter(!station %in% c('1','2','3','4','5','6','9','11', '17')) %>%
-  rename(Longitude = long,            # Need these columns to specifically say this for ERDDAP matching
-         Latitude = lat,
-         UTC = dateTime)
-
-tracks7 <- tracks2 %>%
-  filter(station== 7)
+saveRDS(AllTracks, 'data/AllTracks.rda')
+ 

### 2023-09-13 data-raw/soundscape_metrics.R
Update soundscape_metrics.R
removed rounding

@@ -37,7 +37,7 @@ for (i in dirlist) {
 ####### Read in Percentiles ###############
 
 # Path to folder on my external hard drive
-pclist <- dir("D:/Soundscape_metrics/percentiles", 
+pclist <- dir("D:/Soundscape_metrics/10_90_percentiles", 
                recursive=TRUE, full.names=TRUE, pattern="CCES_")
 
 
@@ -45,7 +45,8 @@ pclist <- dir("D:/Soundscape_metrics/percentiles",
 for (i in pclist) {
   data <- read_csv(i, show_col_types = FALSE) %>%
     rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
-    mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes"))
+    mutate(UTC = as.POSIXct(dateTime))
   
-  saveRDS(data, file = paste0("data/", substr(i, 50, 57), "_", substr(i, 40, 41), ".rda"))
+  saveRDS(data, file = paste0("data/", substr(i, 60, 64), "_", substr(i, 46, 47), ".rda"))
 }
+

### 2023-09-13 07_summary_stats.Rmd
Update 07_summary_stats.Rmd
removed broadband metrics

--- 2023-08-08 07_summary_stats.Rmd
@@
 library(tidyverse)
+library(lubridate)
+library(PAMpal)
+library(PAMmisc)
+library(PAMscapes)
+library(patchwork)
+library(knitr)
+library(kableExtra)
 ```
@@
 ## Make list of rda files to read in
-pclist <- c("BB_pct01_07", "BB_pct01_08", "BB_pct01_10", "BB_pct01_12", "BB_pct01_13", "BB_pct01_14", "BB_pct01_16", 
-            "BB_pct01_18", "BB_pct01_19", "BB_pct01_20", "BB_pct01_21", "BB_pct01_22", "BB_pct01_23", "TO_pct01_07",
-            "TO_pct01_08", "TO_pct01_10", "TO_pct01_12", "TO_pct01_13", "TO_pct01_14", "TO_pct01_16", "TO_pct01_18", 
-            "TO_pct01_19", "TO_pct01_20", "TO_pct01_21", "TO_pct01_22", "TO_pct01_23", "BB_pct99_07", "BB_pct99_08", 
-            "BB_pct99_10", "BB_pct99_12", "BB_pct99_13", "BB_pct99_14", "BB_pct99_16", "BB_pct99_18", "BB_pct99_19", 
-            "BB_pct99_20", "BB_pct99_21", "BB_pct99_22", "BB_pct99_23", "TO_pct99_07", "TO_pct99_08", "TO_pct99_10", 
-            "TO_pct99_12", "TO_pct99_13", "TO_pct99_14", "TO_pct99_16", "TO_pct99_18", "TO_pct99_19", "TO_pct99_20", 
-            "TO_pct99_21", "TO_pct99_22", "TO_pct99_23")
+pclist <- c("pct10_07", "pct10_08", "pct10_10", "pct10_12", "pct10_13",
+            "pct10_14", "pct10_16", "pct10_18", "pct10_19", "pct10_20",
+            "pct10_21", "pct10_22", "pct10_23", "pct90_07", "pct90_08",
+            "pct90_10", "pct90_12", "pct90_13", "pct90_14", "pct90_16",
+            "pct90_18", "pct90_19", "pct90_20", "pct90_21", "pct90_22",
+            "pct90_23", "pct50_07", "pct50_08", "pct50_10", "pct50_12",
+            "pct50_13", "pct50_14", "pct50_16", "pct50_18", "pct50_19", 
+            "pct50_20", "pct50_21", "pct50_22", "pct50_23")
 
@@
 
+```
+
+Create df from 1 percentile
+```{r}
 # 1 Percentile
@@
 
+pct01_07$buoy <- ("07")
+pct01_08$buoy <- ("08")
+pct01_10$buoy <- ("10")
+pct01_12$buoy <- ("12")
+pct01_13$buoy <- ("13")
+pct01_14$buoy <- ("14")
+pct01_16$buoy <- ("16")
+pct01_18$buoy <- ("18")
+pct01_19$buoy <- ("19")
+pct01_20$buoy <- ("20")
+pct01_21$buoy <- ("21")
+pct01_22$buoy <- ("22")
+pct01_23$buoy <- ("23")
+
 pct01_all <- rbind(pct01_07, pct01_08, pct01_10, pct01_12, pct01_13,
@@
                    pct01_21, pct01_22, pct01_23)
-# need to add column with buoy name
 
+# Group by and summarize
+pct01_all %>%
+  group_by(buoy) %>%
+  summarize(max(`TOL_125`))
+
+```
+
+Create df from 99 percentile
+```{r}
 # 99 Percentile
@@
   select(., -3)
+
+# Add column with buoy number
+pct99_07$buoy <- ("07")
+pct99_08$buoy <- ("08")
+pct99_10$buoy <- ("10")
+pct99_12$buoy <- ("12")
+pct99_13$buoy <- ("13")
+pct99_14$buoy <- ("14")
+pct99_16$buoy <- ("16")
+pct99_18$buoy <- ("18")
+pct99_19$buoy <- ("19")
+pct99_20$buoy <- ("20")
+pct99_21$buoy <- ("21")
+pct99_22$buoy <- ("22")
+pct99_23$buoy <- ("23")
+
+pct99_all <- rbind(pct99_07, pct99_08, pct99_10, pct99_12, pct99_13,
+                   pct99_14, pct99_16, pct99_18, pct99_19, pct99_20,
+                   pct99_21, pct99_22, pct99_23)
+```
+
+Plot 99pct and 1pct by buoy
+```{r fig.height=5, fig.width=13}
+# Group by and summarize
+pct99_all %>%
+  group_by(buoy) %>%
+  summarize(max(`TOL_125`)) %>%
+  ggplot(aes(x = buoy, y = `max(TOL_125)`)) + geom_bar(stat = "identity")
+
+#TOL timeline by buoy
+pct99_all %>%
+  ggplot(aes(x = dateTime, y = TOL_2000, color = buoy)) + geom_line()
+
+# Average dB by hour for each frequency
+pct99_16 %>%
+  group_by(time = floor_date(dateTime, "1 hour")) %>%
+  summarize(mean_BB = mean(`BB_20-24000`),
+            meanTOL_125 = mean(TOL_125),
+            meanTOL_250 = mean(TOL_250),
+            meanTOL_500 = mean(TOL_500),
+            meanTOL_2000 = mean(TOL_2000),
+            meanTOL_4000 = mean(TOL_4000),
+            meanTOL_20000 = mean(TOL_20000)) %>%
+  pivot_longer(cols = mean_BB:meanTOL_20000,
+               names_to = "frequency",
+               values_to="dB") %>%
+  ggplot(aes(x = time, y = dB)) + geom_line(aes(color = frequency)) #+
+  #geom_point(data = allDrifts, aes(color = species))
+# Show more dates/times on x axis
+# 99pct, buoy 16 has a strong peak later in journey at 4kHz
+
+ggplot() +
+  geom_point(data = pct01_23, aes(x = lubridate::round_date(dateTime, "2 hours"), y = TOL_125), color = 'steelblue') +
+  #geom_line(data = pct99_23, aes(x = dateTime, y = TOL_125), color = 'coral') +
+  facet_grid()
+
+```
+
+Bind percentile df (pctx_x) with GPS track df (GPSwhalex)
+
+Buoy 7
+```{r}
+# Create columns with percentiles and BIND
+pct50_07_select <- pct50_07 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_07_select <- pct10_07 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_07_select <- pct90_07 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct07 <- cbind(pct10_07_select, pct50_07_select, pct90_07_select)
+
+# Join GPS points with the three percentiles
+pct07_whl <- timeJoin(GPSwhale7, pct07, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy7pctSummary <- pct07_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 8
+```{r}
+# Create columns with percentiles and BIND
+pct50_08_select <- pct50_08 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_08_select <- pct10_08 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_08_select <- pct90_08 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct08 <- cbind(pct10_08_select, pct50_08_select, pct90_08_select)
+
+# Join GPS points with the three percentiles
+pct08_whl <- timeJoin(GPSwhale8, pct08, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy8pctSummary <- pct08_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 10
+```{r}
+# Create columns with percentiles and BIND
+pct50_10_select <- pct50_10 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_10_select <- pct10_10 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_10_select <- pct90_10 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct10 <- cbind(pct10_10_select, pct50_10_select, pct90_10_select)
+
+# Join GPS points with the three percentiles
+pct10_whl <- timeJoin(GPSwhale10, pct10, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy10pctSummary <- pct10_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 12
+```{r}
+# Create columns with percentiles and BIND
+pct50_12_select <- pct50_12 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_12_select <- pct10_12 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_12_select <- pct90_12 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct12 <- cbind(pct10_12_select, pct50_12_select, pct90_12_select)
+
+# Join GPS points with the three percentiles
+pct12_whl <- timeJoin(GPSwhale12, pct12, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy12pctSummary <- pct12_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 13
+```{r}
+# Create columns with percentiles and BIND
+pct50_13_select <- pct50_13 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_13_select <- pct10_13 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_13_select <- pct90_13 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct13 <- cbind(pct10_13_select, pct50_13_select, pct90_13_select)
+
+# Join GPS points with the three percentiles
+pct13_whl <- timeJoin(GPSwhale13, pct13, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy13pctSummary <- pct13_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 14 # Issue binding pct50 with others due to differing number of rows
+```{r}
+# Create columns with percentiles and BIND
+pct50_14_select <- pct50_14 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_14_select <- pct10_14 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_14_select <- pct90_14 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+#pct14 <- cbind(pct10_14_select, pct50_14_select, pct90_14_select)
+
+# Join GPS points with the three percentiles
+pct14_whl <- timeJoin(GPSwhale14, pct14, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy14pctSummary <- pct14_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+Buoy 16
+```{r}
+# Create columns with percentiles and BIND
+pct50_16_select <- pct50_16 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_16_select <- pct10_16 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_16_select <- pct90_16 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct16 <- cbind(pct10_16_select, pct50_16_select, pct90_16_select)
+
+# Join GPS points with the three percentiles
+pct16_whl <- timeJoin(GPSwhale16, pct16, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy16pctSummary <- pct16_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 18
+```{r}
+# Create columns with percentiles and BIND
+pct50_18_select <- pct50_18 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_18_select <- pct10_18 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_18_select <- pct90_18 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct18 <- cbind(pct10_18_select, pct50_18_select, pct90_18_select)
+
+# Join GPS points with the three percentiles
+pct18_whl <- timeJoin(GPSwhale18, pct18, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy18pctSummary <- pct18_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 19
+```{r}
+# Create columns with percentiles and BIND
+pct50_19_select <- pct50_19 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_19_select <- pct10_19 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_19_select <- pct90_19 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct19 <- cbind(pct10_19_select, pct50_19_select, pct90_19_select)
+
+# Join GPS points with the three percentiles
+pct19_whl <- timeJoin(GPSwhale19, pct19, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy19pctSummary <- pct19_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 20
+```{r}
+# Create columns with percentiles and BIND
+pct50_20_select <- pct50_20 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_20_select <- pct10_20 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_20_select <- pct90_20 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct20 <- cbind(pct10_20_select, pct50_20_select, pct90_20_select)
+
+# Join GPS points with the three percentiles
+pct20_whl <- timeJoin(GPSwhale20, pct20, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy20pctSummary <- pct20_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 21
+```{r}
+# Create columns with percentiles and BIND
+pct50_21_select <- pct50_21 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_21_select <- pct10_21 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_21_select <- pct90_21 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct21 <- cbind(pct10_21_select, pct50_21_select, pct90_21_select)
+
+# Join GPS points with the three percentiles
+pct21_whl <- timeJoin(GPSwhale21, pct21, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy21pctSummary <- pct21_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 22
+```{r}
+# Create columns with percentiles and BIND
+pct50_22_select <- pct50_22 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_22_select <- pct10_22 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_22_select <- pct90_22 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct22 <- cbind(pct10_22_select, pct50_22_select, pct90_22_select)
+
+# Join GPS points with the three percentiles
+pct22_whl <- timeJoin(GPSwhale22, pct22, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy22pctSummary <- pct22_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Buoy 23
+```{r}
+# Create columns with percentiles and BIND
+pct50_23_select <- pct50_23 %>% select(UTC, TOL_125, TOL_2000) %>% 
+  rename(pct50_TOL125 = TOL_125,
+         pct50_TOL2000 = TOL_2000)
+pct10_23_select <- pct10_23 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct10_TOL125 = TOL_125,
+         pct10_TOL2000 = TOL_2000)
+pct90_23_select <- pct90_23 %>% select(UTC, TOL_125, TOL_2000) %>%
+   rename(pct90_TOL125 = TOL_125,
+         pct90_TOL2000 = TOL_2000)
+
+# Bind
+pct23 <- cbind(pct10_23_select, pct50_23_select, pct90_23_select)
+
+# Join GPS points with the three percentiles
+pct23_whl <- timeJoin(GPSwhale23, pct23, thresh = 3600, interpolate = FALSE)
+
+#Summarize
+buoy23pctSummary <- pct23_whl %>%
+  group_by(species) %>%
+  summarize(pct10_125_mean = mean(pct10_TOL125),
+            pct50_125_mean = mean(pct50_TOL125),
+            pct90_125_mean = mean(pct90_TOL125),
+            pct10_2000_mean = mean(pct10_TOL2000),
+            pct50_2000_mean = mean(pct50_TOL2000),
+            pct90_2000_mean = mean(pct90_TOL2000))
+```
+
+Knit all summary percentiles into a table
+```{r}
+
+## Buoy 7 ##
+sumPct7 <- kable(buoy7pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 7") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 8 ##
+sumPct8 <- kable(buoy8pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 8") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 10 ##
+sumPct10 <- kable(buoy10pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 10") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 12 ##
+sumPct12 <- kable(buoy12pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 12") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 13 ##
+sumPct13 <- kable(buoy13pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 13") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 14 ##
+#kable(buoy14pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+#                                     caption = "Summary Percentiles for Buoy 14") %>%
+#  kableExtra::kable_classic() %>%
+#  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 16 ##
+sumPct16 <- kable(buoy16pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 16") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 18 ##
+sumPct18 <- kable(buoy18pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 18") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 19 ##
+sumPct19 <- kable(buoy19pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 19") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 20 ##
+sumPct20 <- kable(buoy20pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 20") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 21 ##
+sumPct21 <- kable(buoy21pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 21") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 22 ##
+sumPct22 <- kable(buoy22pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 22") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+## Buoy 23 ##
+sumPct23 <- kable(buoy23pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 23") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+
+```
+
+Time Series Heat Map
+```{r fig.width=12}
+
+pdf(paste("timeSeries8-31",".pdf"), width=14, height=10)
+
+### Buoy 7 ####
+W.07 <- fulltrack07 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.07plot <- plotPresGrid(W.07, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack07, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 7: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.07, bin = 'hour', type = 'presence', start = min(fulltrack07$UTC),
+                         end = max(fulltrack07$UTC), fill = NA, color = "white")
+
+# Create plot with TOL125
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+
+t1 <- plotTimeseries(fulltrack07, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 7: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.07, bin = 'hour', type = 'presence', start = min(fulltrack07$UTC),
+                         end = max(fulltrack07$UTC), fill = NA, color = "white")
+
+# Create plot with TOL2000
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+###### Buoy 8 ####
+W.08 <- fulltrack08 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.08plot <- plotPresGrid(W.08, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack08, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.08, bin = 'hour', type = 'presence', start = min(fulltrack08$UTC),
+                         end = max(fulltrack08$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack08, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.08, bin = 'hour', type = 'presence', start = min(fulltrack08$UTC),
+                         end = max(fulltrack08$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+#### Buoy 10  ######
+
+W.10 <- fulltrack10 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.10plot <- plotPresGrid(W.10, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack10, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
+                         end = max(fulltrack10$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
+                         end = max(fulltrack10$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 12
+W.12 <- fulltrack12 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.12plot <- plotPresGrid(W.12, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack12, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 12: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.12, bin = 'hour', type = 'presence', start = min(fulltrack12$UTC),
+                         end = max(fulltrack12$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack12, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 12: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.12, bin = 'hour', type = 'presence', start = min(fulltrack12$UTC),
+                         end = max(fulltrack12$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 13
+W.13 <- fulltrack13 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.13plot <- plotPresGrid(W.13, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack13, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(fulltrack13$UTC),
+                         end = max(fulltrack13$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack13, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(fulltrack13$UTC),
+                         end = max(fulltrack13$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 14
+W.14 <- fulltrack14 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.14plot <- plotPresGrid(W.14, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack14, column='TOL_125', bin='hour', style = 'heatmap', title='Buoy 14: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.14, bin = 'hour', type = 'presence', start = min(fulltrack14$UTC),
+                         end = max(fulltrack14$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack14, column='TOL_2000', bin='hour', style = 'heatmap', title='Buoy 14: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.14, bin = 'hour', type = 'presence', start = min(fulltrack14$UTC),
+                         end = max(fulltrack14$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 16
+W.16 <- fulltrack16 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.16plot <- plotPresGrid(W.16, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack16, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 16: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.16, bin = 'hour', type = 'presence', start = min(fulltrack16$UTC),
+                         end = max(fulltrack16$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack16, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 16: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.16, bin = 'hour', type = 'presence', start = min(fulltrack16$UTC),
+                         end = max(fulltrack16$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 18
+W.18 <- fulltrack18 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.18plot <- plotPresGrid(W.18, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack18, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 18: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.18, bin = 'hour', type = 'presence', start = min(fulltrack18$UTC),
+                         end = max(fulltrack18$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack18, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 18: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.18, bin = 'hour', type = 'presence', start = min(fulltrack18$UTC),
+                         end = max(fulltrack18$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 19
+W.19 <- fulltrack19 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.19plot <- plotPresGrid(W.19, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack19, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 19: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.19, bin = 'hour', type = 'presence', start = min(fulltrack19$UTC),
+                         end = max(fulltrack19$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack19, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 19: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.19, bin = 'hour', type = 'presence', start = min(fulltrack19$UTC),
+                         end = max(fulltrack19$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 20
+W.20 <- fulltrack20 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.20plot <- plotPresGrid(W.20, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack20, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(fulltrack20$UTC),
+                         end = max(fulltrack20$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack20, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(fulltrack20$UTC),
+                         end = max(fulltrack20$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 21
+W.21 <- fulltrack21 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.21plot <- plotPresGrid(W.21, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack21, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 21: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.21, bin = 'hour', type = 'presence', start = min(fulltrack21$UTC),
+                         end = max(fulltrack21$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack21, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 21: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.21, bin = 'hour', type = 'presence', start = min(fulltrack21$UTC),
+                         end = max(fulltrack21$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 12
+W.22 <- fulltrack22 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.22plot <- plotPresGrid(W.22, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack22, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 22: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.22, bin = 'hour', type = 'presence', start = min(fulltrack22$UTC),
+                         end = max(fulltrack22$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack22, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 22: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.22, bin = 'hour', type = 'presence', start = min(fulltrack22$UTC),
+                         end = max(fulltrack22$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 23
+W.23 <- fulltrack23 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.23plot <- plotPresGrid(W.23, bin = 'hour', type = 'presence')
+
+g1 <- (plotTimeseries(fulltrack23, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 23: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.23, bin = 'hour', type = 'presence', start = min(fulltrack23$UTC),
+                         end = max(fulltrack23$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(fulltrack23, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 23: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.23, bin = 'hour', type = 'presence', start = min(fulltrack23$UTC),
+                         end = max(fulltrack23$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r}
+save.image(file = ".RData")
+```
+
+plot whale presence and SS side by side. Stacked?
+```{r}
+W.08 <- fulltrack08 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.08plot <- plotPresGrid(W.08, bin = 'hour', type = 'presence')
+
+plotTimeseries(fulltrack08, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 2000 Plot') +  W.08plot
+
+#############
+
+W.10 <- fulltrack10 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.10plot <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
+                         end = max(fulltrack10$UTC), by = 'BWpresence', alpha = 0.01, color = "red")
+
+plotTimeseries(fulltrack10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot') +  W.10plot
+plotTimeseries(fulltrack10, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 125 Plot') +  W.10plot
+#############
+
+W.13 <- fulltrack13 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.13plot <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(fulltrack13$UTC),
+                         end = max(fulltrack13$UTC), by = 'BWpresence', alpha = 0.5)
+
+plotTimeseries(fulltrack13, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 2000 Plot') +  W.13plot
+plotTimeseries(fulltrack13, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 125 Plot') +  W.13plot
+##############
+
+W.20 <- fulltrack20 %>%
+  select(UTC, BWpresence, species) %>%
+  filter(species == "ZC")
+
+W.20plot <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(fulltrack20$UTC),
+                         end = max(fulltrack20$UTC),  by = 'BWpresence', alpha = 0.5)
+
+plotTimeseries(fulltrack20, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot') +  W.20plot
+plotTimeseries(fulltrack20, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot') +  W.20plot
+
+##########
+
+g1 <- plotTimeseries(fulltrack10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot')
+g2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
+                         end = max(fulltrack10$UTC), fill = NA, color = "white")
+S.10 <- fulltrack10 %>% select(UTC, shipin10km) %>% filter(shipin10km == "1")
+g3 <- plotPresGrid(S.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
+                         end = max(fulltrack10$UTC), fill = NA, color = "red")
+
+combined <- g1
+combined$layers <- c(combined$layers, g2$layers, g3$layers)
+combined
 ```
@@
 
-Where are loudest sound levels?
+Where are loudest sound levels? See @haver2020 for details on percentiles
+5th percentile = exceeded most of the time, so refers to background sound
+90th percentile = exceeded 10% of the time, so refers to PEAKS in sound level
+Also see @haver2021 (fig 2) for daily TOL levels by site (can do by buoy)
 ```{r}
-# For buoy 7
-pct01_07 %>%
-  summarise_if(is.numeric, list(~ min(., na.rm=TRUE)))
-
-pct01_07 %>% filter(.$`BB_20-24000` > 120)
-
-########################
 
@@
   dplyr::filter(species == "PM")
+
+PM %>%
+  group_by(station) %>%
+  summarize(mean = mean(TOL_2000), n = n())
 

### 2023-09-13 04_analysis.Rmd
Update 04_analysis.Rmd
added some GAM models

--- 2023-06-29 04_analysis.Rmd
@@
                    fulltrack16, fulltrack18, fulltrack19, fulltrack20, fulltrack21, fulltrack22, fulltrack23)
+save(allDrifts, file = "allDrifts.rda")
 ```
@@
 ```
+Looking at TOl2000 sound levels for each buoy and where Whale Detections happen
+```{r fig.width=10}
+ggplot(fulltrack23, aes(x = UTC, y = TOL_2000)) + geom_line() + 
+  geom_smooth() + 
+  geom_point(aes(color = species, size = 2)) +
+  geom_point(color = shipin10km) +
+  scale_color_discrete(na.translate=FALSE)
+```
 
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```
+
+```{r}
+ggplot(allDrifts, aes(factor(BWpresence), TOL_2000)) + geom_boxplot()
 ```
@@
 ```
+
 ### GAMs
@@
 ```
+
+Top 5 environmental variables (by RF) with one drift
+```{r}
+# Environ only
+model1 <- gam(BWpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400,
+              data = fulltrack22, family = binomial, method = "REML")
+
+plot(model1, trans = plogis, all.terms = TRUE)
+k.check(model1)
+summary(model1)
+gam.check(model1)
+# 16% deviance explained (buoy 7)
+# 0.68% deviance explained (buoy 8)
+# 8.08% deviance explained (buoy 10)
+# 0.64% deviance explained (buoy 12)
+# 6.94 % deviance explained (buoy 13)
+# 19.5% deviance explained (buoy 14)
+# 12.2% deviance explained (buoy 16)
+# 8.15% deviance explained (buoy 18)
+# 7.7% deviance explained (buoy 19)
+# 5.5% deviance explained (buoy 20)
+# 33.5% deviance explained (buoy 21)
+# 13.5% deviance explained (buoy 22)
+# 12.1% deviance explained (buoy 23)
+
+# Add soundscape metrics to model 1
+model2 <- gam(BWpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400
+              + TOL_125 + TOL_2000,
+              data = fulltrack22, family = binomial, method = "REML")
+
+#plot(model2, trans = plogis) ## doesn't work
+summary(model2)
+gam.check(model2)
+# 18.3% deviance explained (buoy 7)
+# 1.25% deviance explained (buoy 8)
+# 8.56% deviance explained (buoy 10)
+# 1.39% deviance explained (buoy 12)
+# 11.1% deviance explained (buoy 13)
+# 28% deviance explained (buoy 14)
+# 15% deviance explained (buoy 16)
+# 9.92% deviance explained (buoy 18)
+# 8.57% deviance explained (buoy 19)
+# 7.3% deviance explained (buoy 20)
+# 38.3% deviance explained (buoy 21)
+# 15.2 % deviance explained (buoy 22)
+# 20.4% deviance explained (buoy 23)
+```
+GAM with FULL DATASET (not great)
+```{r}
+alldriftGAM <- gam(BWpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400
+              + TOL_125 + TOL_2000,
+              data = allDrifts, family = binomial, method = "REML")
+summary(alldriftGAM)
+gam.check(alldriftGAM)
+```
+
+GAMs with s()
+```{r}
+# Environ only, with s()
+model3 <- gam(BWpresence>0 ~ s(depth) + s(sst_mean) + s(mldTemp) + 
+                s(ttDepth) + s(sal400),
+              data = fulltrack21, family = binomial, method = "REML")
+
+plot(model3, trans = plogis)
+summary(model3)
+gam.check(model3)
+# 61.9 % deviance explained (buoy 7)
+# 1.42% deviance explained (buoy 8)
+# 12.5% deviance explained (buoy 10)
+# 1.66% deviance explained (buoy 12)
+# 12.8% deviance explained (buoy 13)
+# 41.4% deviance explained (buoy 14)
+# 39% deviance explained (buoy 16)
+# 16.6% deviance explained (buoy 18)
+# 15.5% deviance explained (buoy 19)
+# 5.8% deviance explained (buoy 20)
+## Iteration limit reached without full convergence -- 71.7% deviance explained (buoy 21) ???
+# 18.6% deviance explained (buoy 22)
+# 49.6% deviance explained (buoy 23)
+
+#Add soundscape metrics to model 3
+model4 <- gam(BWpresence>0 ~ s(depth) + s(sst_mean) + s(mldTemp) + 
+                s(ttDepth) + s(sal400) + s(TOL_125) + s(TOL_2000),
+              data = fulltrack10, family = binomial, method = "REML")
+
+plot(model4, trans = plogis)
+summary(model4)
+gam.check(model4)
+# 14.5% deviance explained
+
+# mix and match?
+model5 <- gam(BWpresence>0 ~ depth + sst_mean + s(mldTemp) + 
+                s(ttDepth) + sal400,
+              data = fulltrack21, family = binomial, method = "REML")
+
+summary(model5)
+gam.check(model5)
+# 57.4 % deviance explained
+plot(model5)
+
+```
+Higher edfs describe more wiggly curves. Doesn't mean significant

### 2023-09-13 03_environ_var.Rmd
Update 03_environ_var.Rmd
created env# variable for each buoy, reran with better GPS pts

@@
 
+Read in GPSwhale rdas
+```{r}
+env07 <- readRDS(file = "data/GPSwhaleAIS7.rda")
+env08 <- readRDS(file = "data/GPSwhaleAIS8.rda")
+env10 <- readRDS(file = "data/GPSwhaleAIS10.rda")
+env12 <- readRDS(file = "data/GPSwhaleAIS12.rda")
+env13 <- readRDS(file = "data/GPSwhaleAIS13.rda")
+env14 <- readRDS(file = "data/GPSwhaleAIS14.rda")
+env16 <- readRDS(file = "data/GPSwhaleAIS16.rda")
+env18 <- readRDS(file = "data/GPSwhaleAIS18.rda")
+env19 <- readRDS(file = "data/GPSwhaleAIS19.rda")
+env20 <- readRDS(file = "data/GPSwhaleAIS20.rda")
+env21 <- readRDS(file = "data/GPSwhaleAIS21.rda")
+env22 <- readRDS(file = "data/GPSwhaleAIS22.rda")
+env23 <- readRDS(file = "data/GPSwhaleAIS23.rda")
+```
+
 ### Get Environmental Variables
@@
 Sea Surface Temperature from ERDDAP using PAMPal function
+Daily Multi-scale Ultra-high Resolution (MUR) SST Analysis fv04.1, Global, 0.01°
 ```{r}
 # Downloads data from servers onto external hard drive. Uncomment and only run once.
-#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST07", var = c('analysed_sst'))
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST08", var = c('analysed_sst'))
-#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST10", var = c('analysed_sst'))
-#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST12", var = c('analysed_sst'))
-#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST13", var = c('analysed_sst'))
-#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST14", var = c('analysed_sst'))
-#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST16", var = c('analysed_sst'))
-#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST18", var = c('analysed_sst'))
-#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST19", var = c('analysed_sst'))
-#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST20", var = c('analysed_sst'))
-#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST21", var = c('analysed_sst'))
-#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST22", var = c('analysed_sst'))
-#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST23", var = c('analysed_sst'))
+#env07 <- matchEnvData(GPSwhaleAIS7, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST07", var = c('analysed_sst'))
+#env08 <- matchEnvData(GPSwhaleAIS8, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST08", var = c('analysed_sst'))
+#env10 <- matchEnvData(GPSwhaleAIS10, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST10", var = c('analysed_sst'))
+#env12 <- matchEnvData(GPSwhaleAIS12, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST12", var = c('analysed_sst'))
+#env13 <- matchEnvData(GPSwhaleAIS13, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST13", var = c('analysed_sst'))
+#env14 <- matchEnvData(GPSwhaleAIS14, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST14", var = c('analysed_sst'))
+#env16 <- matchEnvData(GPSwhaleAIS16, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST16", var = c('analysed_sst'))
+#env18 <- matchEnvData(GPSwhaleAIS18, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST18", var = c('analysed_sst'))
+#env19 <- matchEnvData(GPSwhaleAIS19, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST19", var = c('analysed_sst'))
+#env20 <- matchEnvData(GPSwhaleAIS20, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST20", var = c('analysed_sst'))
+#env21 <- matchEnvData(GPSwhaleAIS21, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST21", var = c('analysed_sst'))
+#env22 <- matchEnvData(GPSwhaleAIS22, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST22", var = c('analysed_sst'))
+#env23 <- matchEnvData(GPSwhaleAIS23, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST23", var = c('analysed_sst'))
 ```
@@
 for (ss in SSTlist) {
-  data <- matchEnvData(get(paste0("tracks_SS_", substr(ss, 33, 34))), nc = ss) %>%
+  data <- matchEnvData(get(paste0("env", substr(ss, 33, 34))), nc = ss) %>%
   rename(., sst_mean = analysed_sst_mean) %>%
@@
 
-  assign(paste0("tracks_SS_", substr(ss, 33, 34)), data)
+  assign(paste0("env", substr(ss, 33, 34)), data)
 }
@@
 ### Wind Stress Curl from ERDDAP using PAMPal function
+Wind Stress, Metop-A ASCAT, 0.25°, Global, Near Real Time, 2009-present (1 Day) 
 ```{r}
 # Downloads data from servers onto external hard drive. Uncomment and only run once.
-#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress07.nc", var = c('curl'))
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress08.nc", var = c('curl'))
-#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress10.nc", var = c('curl'))
-#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress12.nc", var = c('curl'))
-#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress13.nc", var = c('curl'))
-#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress14.nc", var = c('curl'))
-#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress16.nc", var = c('curl'))
-#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress18.nc", var = c('curl'))
-#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress19.nc", var = c('curl'))
-#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress20.nc", var = c('curl'))
-#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress21.nc", var = c('curl'))
-#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress22.nc", var = c('curl'))
-#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress23.nc", var = c('curl'))
-
-# Pulls from downloaded file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data-raw/erdQMstress1day_LonPM180.nc')
-#tracks_SS_08
-
-#windcurl <- erddapToEdinfo('erdQMstress1day', baseurl = 'https://coastwatch.pfeg.noaa.gov/erddap/')
-
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=windcurl, fileName='windcurl.nc')
-
+#env07 <- matchEnvData(env07, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress07.nc", var = c('curl'))
+#env08 <- matchEnvData(env08, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress08.nc", var = c('curl'))
+#env10 <- matchEnvData(env10, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress10.nc", var = c('curl'))
+#env12 <- matchEnvData(env12, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress12.nc", var = c('curl'))
+#env13 <- matchEnvData(env13, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress13.nc", var = c('curl'))
+#env14 <- matchEnvData(env14, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress14.nc", var = c('curl'))
+#env16 <- matchEnvData(env16, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress16.nc", var = c('curl'))
+#env18 <- matchEnvData(env18, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress18.nc", var = c('curl'))
+#env19 <- matchEnvData(env19, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress19.nc", var = c('curl'))
+#env20 <- matchEnvData(env20, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress20.nc", var = c('curl'))
+#env21 <- matchEnvData(env21, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress21.nc", var = c('curl'))
+#env22 <- matchEnvData(env22, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress22.nc", var = c('curl'))
+#env23 <- matchEnvData(env23, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress23.nc", var = c('curl'))
 ```
@@
 for (w in windlist) {
-  data <- matchEnvData(get(paste0("tracks_SS_", substr(w, 32, 33))), nc = w) %>%
+  data <- matchEnvData(get(paste0("env", substr(w, 32, 33))), nc = w) %>%
      dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean, -matchDepth_mean)   # remove unnecessary columns
 
-  assign(paste0("tracks_SS_", substr(w, 32, 33)), data)
+  assign(paste0("env", substr(w, 32, 33)), data)
 }
@@
 
-### Distance to continental slope (uses terra package)
+### Distance to continental slope (uses terra and sf package)
 ```{r}
@@
 
-tracklist <- grep(pattern = "tracks_SS_", names(.GlobalEnv), value = T)
+tracklist <- grep(pattern = "env", names(.GlobalEnv), value = T)
 
@@
 
-  assign(paste0("tracks_SS_", substr(t, 11, 12)), dbind)
+  assign(paste0("env", substr(t, 4, 5)), dbind)
 }
@@
 Depth from ERDDAP using PAMPal function
+SRTM15_PLUS Estimated Topography, 15 seconds, Global, v1 
 ```{r}
 # Downloads data from servers onto external hard drive. Uncomment and only run once.
-#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth07.nc", var = c('z'))
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth08.nc", var = c('z'))
-#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth10.nc", var = c('z'))
-#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth12.nc", var = c('z'))
-#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth13.nc", var = c('z'))
-#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth14.nc", var = c('z'))
-#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth16.nc", var = c('z'))
-#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth18.nc", var = c('z'))
-#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth19.nc", var = c('z'))
-#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth20.nc", var = c('z'))
-#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth21.nc", var = c('z'))
-#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth22.nc", var = c('z'))
-#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth23.nc", var = c('z'))
+#env07 <- matchEnvData(env07, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth07.nc", var = c('z'))
+#env08 <- matchEnvData(env08, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth08.nc", var = c('z'))
+#env10 <- matchEnvData(env10, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth10.nc", var = c('z'))
+#env12 <- matchEnvData(env12, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth12.nc", var = c('z'))
+#env13 <- matchEnvData(env13, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth13.nc", var = c('z'))
+#env14 <- matchEnvData(env14, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth14.nc", var = c('z'))
+#env16 <- matchEnvData(env16, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth16.nc", var = c('z'))
+#env18 <- matchEnvData(env18, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth18.nc", var = c('z'))
+#env19 <- matchEnvData(env19, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth19.nc", var = c('z'))
+#env20 <- matchEnvData(env20, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth20.nc", var = c('z'))
+#env21 <- matchEnvData(env21, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth21.nc", var = c('z'))
+#env22 <- matchEnvData(env22, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth22.nc", var = c('z'))
+#env23 <- matchEnvData(env23, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth23.nc", var = c('z'))
 ```
@@
 for (dd in depthlist) {
-  data <- matchEnvData(get(paste0("tracks_SS_", substr(dd, 27, 28))), nc = dd) %>%
+  data <- matchEnvData(get(paste0("env", substr(dd, 27, 28))), nc = dd) %>%
   rename(., depth = z_mean) %>%        # rename new column to "depth"
@@
 
-  assign(paste0("tracks_SS_", substr(dd, 27, 28)), data)
+  assign(paste0("env", substr(dd, 27, 28)), data)
 }
@@
 
-#### Bathymetric slope (skipping loop for now)
+#### Bathymetric slope. Units in degrees
 ```{r}
@@
 for (t in tracklist) {
-  slope <- terra::extract(bathy_slope, get(t)[,4:3]) %>%  # Reads latitude and longitude of buoy dataframe
+  slope <- terra::extract(bathy_slope, get(t)[,3:2]) %>%  # Reads latitude and longitude of buoy dataframe
   dplyr::select(-ID) %>%
@@
 
-  assign(paste0("tracks_SS_", substr(t, 11, 12)), sbind)
+  assign(paste0("env", substr(t, 4, 5)), sbind)
 }
@@
 # Downloads data from servers onto external hard drive. Uncomment and only run once.
-#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore07.nc", var = c('dist'))
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore08.nc", var = c('dist'))
-#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore10.nc", var = c('dist'))
-#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore12.nc", var = c('dist'))
-#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore13.nc", var = c('dist'))
-#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore14.nc", var = c('dist'))
-#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore16.nc", var = c('dist'))
-#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore18.nc", var = c('dist'))
-#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore19.nc", var = c('dist'))
-#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore20.nc", var = c('dist'))
-#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore21.nc", var = c('dist'))
-#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore22.nc", var = c('dist'))
-#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore23.nc", var = c('dist'))
+#env07 <- matchEnvData(env07, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore07.nc", var = c('dist'))
+#env08 <- matchEnvData(env08, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore08.nc", var = c('dist'))
+#env10 <- matchEnvData(env10, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore10.nc", var = c('dist'))
+#env12 <- matchEnvData(env12, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore12.nc", var = c('dist'))
+#env13 <- matchEnvData(env13, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore13.nc", var = c('dist'))
+#env14 <- matchEnvData(env14, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore14.nc", var = c('dist'))
+#env16 <- matchEnvData(env16, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore16.nc", var = c('dist'))
+#env18 <- matchEnvData(env18, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore18.nc", var = c('dist'))
+#env19 <- matchEnvData(env19, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore19.nc", var = c('dist'))
+#env20 <- matchEnvData(env20, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore20.nc", var = c('dist'))
+#env21 <- matchEnvData(env21, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore21.nc", var = c('dist'))
+#env22 <- matchEnvData(env22, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore22.nc", var = c('dist'))
+#env23 <- matchEnvData(env23, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore23.nc", var = c('dist'))
 ```
@@
 for (d2 in distlist) {
-  data <- matchEnvData(get(paste0("tracks_SS_", substr(d2, 32, 33))), nc = d2) %>%
+  data <- matchEnvData(get(paste0("env", substr(d2, 32, 33))), nc = d2) %>%
   rename(., dist2shore = dist_mean) %>%
@@
 
-  assign(paste0("tracks_SS_", substr(d2, 32, 33)), data)
+  assign(paste0("env", substr(d2, 32, 33)), data)
 }
@@
 ### Chlorophyll from ERDDAP using PAMPal function
+Chlorophyll-a, Aqua MODIS, NPP, L3SMI, Global, 4km, Science Quality, 2003-
+present (Monthly Composite) 
 ```{r}
 # Downloads data from servers onto external hard drive. Uncomment and only run once.
-#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll07.nc", var = c('chlorophyll'))
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll08.nc", var = c('chlorophyll'))
-#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll10.nc", var = c('chlorophyll'))
-#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll12.nc", var = c('chlorophyll'))
-#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll13.nc", var = c('chlorophyll'))
-#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll14.nc", var = c('chlorophyll'))
-#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll16.nc", var = c('chlorophyll'))
-#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll18.nc", var = c('chlorophyll'))
-#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll19.nc", var = c('chlorophyll'))
-#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll20.nc", var = c('chlorophyll'))
-#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll21.nc", var = c('chlorophyll'))
-#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll22.nc", var = c('chlorophyll'))
-#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll23.nc", var = c('chlorophyll'))
+#env07 <- matchEnvData(env07, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll07.nc", var = c('chlorophyll'))
+#env08 <- matchEnvData(env08, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll08.nc", var = c('chlorophyll'))
+#env10 <- matchEnvData(env10, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll10.nc", var = c('chlorophyll'))
+#env12 <- matchEnvData(env12, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll12.nc", var = c('chlorophyll'))
+#env13 <- matchEnvData(env13, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll13.nc", var = c('chlorophyll'))
+#env14 <- matchEnvData(env14, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll14.nc", var = c('chlorophyll'))
+#env16 <- matchEnvData(env16, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll16.nc", var = c('chlorophyll'))
+#env18 <- matchEnvData(env18, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll18.nc", var = c('chlorophyll'))
+#env19 <- matchEnvData(env19, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll19.nc", var = c('chlorophyll'))
+#env20 <- matchEnvData(env20, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll20.nc", var = c('chlorophyll'))
+#env21 <- matchEnvData(env21, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll21.nc", var = c('chlorophyll'))
+#env22 <- matchEnvData(env22, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll22.nc", var = c('chlorophyll'))
+#env23 <- matchEnvData(env23, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll23.nc", var = c('chlorophyll'))
 ```
@@
 for (ch in chlalist) {
-  data <- matchEnvData(get(paste0("tracks_SS_", substr(ch, 33, 34))), nc = ch) %>%
+  data <- matchEnvData(get(paste0("env", substr(ch, 33, 34))), nc = ch) %>%
   dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
 
-  assign(paste0("tracks_SS_", substr(ch, 33, 34)), data)
+  assign(paste0("env", substr(ch, 33, 34)), data)
 }
@@
 Sea Surface Height Anomalies from ERDDAP using PAMPal function
+Sea Surface Height Anomalies from Altimetry, Global, 2017-
+present (meters)
 ```{r}
 # Downloads data from servers onto external hard drive. Uncomment and only run once.
-#tracks_SS_07 <- matchEnvData(tracks_SS_07, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh07.nc", var = c('sla'))
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh08.nc", var = c('sla'))
-#tracks_SS_10 <- matchEnvData(tracks_SS_10, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh10.nc", var = c('sla'))
-#tracks_SS_12 <- matchEnvData(tracks_SS_12, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh12.nc", var = c('sla'))
-#tracks_SS_13 <- matchEnvData(tracks_SS_13, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh13.nc", var = c('sla'))
-#tracks_SS_14 <- matchEnvData(tracks_SS_14, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh14.nc", var = c('sla'))
-#tracks_SS_16 <- matchEnvData(tracks_SS_16, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh16.nc", var = c('sla'))
-#tracks_SS_18 <- matchEnvData(tracks_SS_18, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh18.nc", var = c('sla'))
-#tracks_SS_19 <- matchEnvData(tracks_SS_19, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh19.nc", var = c('sla'))
-#tracks_SS_20 <- matchEnvData(tracks_SS_20, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh20.nc", var = c('sla'))
-#tracks_SS_21 <- matchEnvData(tracks_SS_21, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh21.nc", var = c('sla'))
-#tracks_SS_22 <- matchEnvData(tracks_SS_22, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh22.nc", var = c('sla'))
-#tracks_SS_23 <- matchEnvData(tracks_SS_23, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh23.nc", var = c('sla'))
+#env07 <- matchEnvData(env07, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh07.nc", var = c('sla'))
+#env08 <- matchEnvData(env08, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh08.nc", var = c('sla'))
+#env10 <- matchEnvData(env10, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh10.nc", var = c('sla'))
+#env12 <- matchEnvData(env12, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh12.nc", var = c('sla'))
+#env13 <- matchEnvData(env13, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh13.nc", var = c('sla'))
+#env14 <- matchEnvData(env14, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh14.nc", var = c('sla'))
+#env16 <- matchEnvData(env16, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh16.nc", var = c('sla'))
+#env18 <- matchEnvData(env18, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh18.nc", var = c('sla'))
+#env19 <- matchEnvData(env19, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh19.nc", var = c('sla'))
+#env20 <- matchEnvData(env20, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh20.nc", var = c('sla'))
+#env21 <- matchEnvData(env21, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh21.nc", var = c('sla'))
+#env22 <- matchEnvData(env22, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh22.nc", var = c('sla'))
+#env23 <- matchEnvData(env23, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh23.nc", var = c('sla'))
 ```
@@
 for (sh in sshlist) {
-  data <- matchEnvData(get(paste0("tracks_SS_", substr(sh, 25, 26))), nc = sh) %>%
+  data <- matchEnvData(get(paste0("env", substr(sh, 25, 26))), nc = sh) %>%
   rename(., ssh_mean = sla_mean) %>%
@@
 
-  assign(paste0("tracks_SS_", substr(sh, 25, 26)), data)
+  assign(paste0("env", substr(sh, 25, 26)), data)
 }
@@
 for (t in tracklist) {
-  Hdata <- matchEnvData(as.data.frame(get(paste0("tracks_SS_", substr(t, 11, 12)))), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
+  Hdata <- matchEnvData(as.data.frame(get(paste0("env", substr(t, 4, 5)))), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
 
@@
 }))
-  bindr <- cbind(get(paste0("tracks_SS_", substr(t, 11, 12))), thermData)
-  assign(paste0("tracks_SS_", substr(t, 11, 12)), bindr)
+  bindr <- cbind(get(paste0("env", substr(t, 4, 5))), thermData)
+  assign(paste0("env", substr(t, 4, 5)), bindr)
 }
@@
 
-Save final buoy df for analysis use next.
+Save final df 'GPSwhaleAISenv' to add soundscape metrics next.
 ```{r}
-saveRDS(tracks_SS_10, 'data/fulltrack10.rda')
-saveRDS(tracks_SS_12, 'data/fulltrack12.rda')
-saveRDS(tracks_SS_13, 'data/fulltrack13.rda')
-saveRDS(tracks_SS_14, 'data/fulltrack14.rda')
-saveRDS(tracks_SS_16, 'data/fulltrack16.rda')
-saveRDS(tracks_SS_18, 'data/fulltrack18.rda')
-saveRDS(tracks_SS_19, 'data/fulltrack19.rda')
-saveRDS(tracks_SS_20, 'data/fulltrack20.rda')
-saveRDS(tracks_SS_21, 'data/fulltrack21.rda')
-saveRDS(tracks_SS_22, 'data/fulltrack22.rda')
-saveRDS(tracks_SS_23, 'data/fulltrack23.rda')
-saveRDS(tracks_SS_07, 'data/fulltrack07.rda')
-saveRDS(tracks_SS_08, 'data/fulltrack08.rda')
+saveRDS(env10, 'data/GPSwhaleAISenv10.rda')
+saveRDS(env12, 'data/GPSwhaleAISenv12.rda')
+saveRDS(env13, 'data/GPSwhaleAISenv13.rda')
+saveRDS(env14, 'data/GPSwhaleAISenv14.rda')
+saveRDS(env16, 'data/GPSwhaleAISenv16.rda')
+saveRDS(env18, 'data/GPSwhaleAISenv18.rda')
+saveRDS(env19, 'data/GPSwhaleAISenv19.rda')
+saveRDS(env20, 'data/GPSwhaleAISenv20.rda')
+saveRDS(env21, 'data/GPSwhaleAISenv21.rda')
+saveRDS(env22, 'data/GPSwhaleAISenv22.rda')
+saveRDS(env23, 'data/GPSwhaleAISenv23.rda')
+saveRDS(env07, 'data/GPSwhaleAISenv07.rda')
+saveRDS(env08, 'data/GPSwhaleAISenv08.rda')
 
 # also saves to external hard drive
-saveRDS(tracks_SS_10, 'D:/Buoy_dfs/fulltrack10.rda')
-saveRDS(tracks_SS_12, 'D:/Buoy_dfs/fulltrack12.rda')
-saveRDS(tracks_SS_13, 'D:/Buoy_dfs/fulltrack13.rda')
-saveRDS(tracks_SS_14, 'D:/Buoy_dfs/fulltrack14.rda')
-saveRDS(tracks_SS_16, 'D:/Buoy_dfs/fulltrack16.rda')
-saveRDS(tracks_SS_18, 'D:/Buoy_dfs/fulltrack18.rda')
-saveRDS(tracks_SS_19, 'D:/Buoy_dfs/fulltrack19.rda')
-saveRDS(tracks_SS_20, 'D:/Buoy_dfs/fulltrack20.rda')
-saveRDS(tracks_SS_21, 'D:/Buoy_dfs/fulltrack21.rda')
-saveRDS(tracks_SS_22, 'D:/Buoy_dfs/fulltrack22.rda')
-saveRDS(tracks_SS_23, 'D:/Buoy_dfs/fulltrack23.rda')
-saveRDS(tracks_SS_07, 'D:/Buoy_dfs/fulltrack07.rda')
-saveRDS(tracks_SS_08, 'D:/Buoy_dfs/fulltrack08.rda')
+saveRDS(env10, 'D:/Buoy_dfs/GPSwhaleAISenv10.rda')
+saveRDS(env12, 'D:/Buoy_dfs/GPSwhaleAISenv12.rda')
+saveRDS(env13, 'D:/Buoy_dfs/GPSwhaleAISenv13.rda')
+saveRDS(env14, 'D:/Buoy_dfs/GPSwhaleAISenv14.rda')
+saveRDS(env16, 'D:/Buoy_dfs/GPSwhaleAISenv16.rda')
+saveRDS(env18, 'D:/Buoy_dfs/GPSwhaleAISenv18.rda')
+saveRDS(env19, 'D:/Buoy_dfs/GPSwhaleAISenv19.rda')
+saveRDS(env20, 'D:/Buoy_dfs/GPSwhaleAISenv20.rda')
+saveRDS(env21, 'D:/Buoy_dfs/GPSwhaleAISenv21.rda')
+saveRDS(env22, 'D:/Buoy_dfs/GPSwhaleAISenv22.rda')
+saveRDS(env23, 'D:/Buoy_dfs/GPSwhaleAISenv23.rda')
+saveRDS(env07, 'D:/Buoy_dfs/GPSwhaleAISenv07.rda')
+saveRDS(env08, 'D:/Buoy_dfs/GPSwhaleAISenv08.rda')
 ```

### 2023-09-13 02_AIS.Rmd
Update 02_AIS.Rmd
reran with updated GPS pts

--- 2023-05-11 02_AIS.Rmd
@@
 ---
-```{r}
-#Required Packages
 
+#Introduction
+AIS download from Marine Cadastre and match to buoy datasets.
+
+Required Packages
+```{r, }
+library(PAMscapes)
 library(tidyverse)
-library(sf)
-library(gganimate)
-#install.packages("remotes")
-#remotes::install_github("skgrange/gissr")
-#devtools::install_github("skgrange/gissr")
-library(gissr)
 ```
 
+-----------------------
+From Taiki's AIS Workflow
 ```{r}
-# Function
-# Speeding up st_intersection (from https://github.com/r-spatial/sf/issues/801)
+#marcadFiles07 <- downloadMarCadAIS(GPSwhale7, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles08 <- downloadMarCadAIS(GPSwhale8, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles10 <- downloadMarCadAIS(GPSwhale10, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles12 <- downloadMarCadAIS(GPSwhale12, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles13 <- downloadMarCadAIS(GPSwhale13, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles14 <- downloadMarCadAIS(GPSwhale14, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles16 <- downloadMarCadAIS(GPSwhale16, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles18 <- downloadMarCadAIS(GPSwhale18, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles19 <- downloadMarCadAIS(GPSwhale19, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles20 <- downloadMarCadAIS(GPSwhale20, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles21 <- downloadMarCadAIS(GPSwhale21, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles22 <- downloadMarCadAIS(GPSwhale22, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles23 <- downloadMarCadAIS(GPSwhale23, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+```
 
-st_intersection_faster <- function(x,y,...){
-#faster replacement for st_intersection(x, y,...)
+Subset to US West Coast
+```{r}
+subsetMarCadAIS(inDir="D:/AIS", outDir="D:/AIS/AIS_West", name='West_', 
+                latRange=c(20,50), lonRange=c(-140,-110),
+                overwrite=FALSE, progress=TRUE)
+```
+Now read into R only ais points that come in contact with buoys
+```{r}
+#ais_near_07 <- readLocalAIS(GPSwhale7, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
 
-  y_subset <-
-    st_intersects(x, y) %>%
-    unlist() %>%
-    unique() %>%
-    sort() %>%
-    {y[.,]}
+#ais_near_08 <- readLocalAIS(GPSwhale8, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
 
-  st_intersection(x, y_subset,...)
+#ais_near_10 <- readLocalAIS(GPSwhale10, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_12 <- readLocalAIS(GPSwhale12, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_13 <- readLocalAIS(GPSwhale13, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_14 <- readLocalAIS(GPSwhale14, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_16 <- readLocalAIS(GPSwhale16, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_18 <- readLocalAIS(GPSwhale18, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_19 <- readLocalAIS(GPSwhale19, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_20 <- readLocalAIS(GPSwhale20, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_21 <- readLocalAIS(GPSwhale21, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+#empty table because no large vessels interacted with this buoy
+
+#ais_near_22 <- readLocalAIS(GPSwhale22, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#     filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_23 <- readLocalAIS(GPSwhale23, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+```
+
+Save `ais_near_` for all buoys
+```{r}
+#saveRDS(ais_near_07, file = "D:AIS/ais_near_07.rda")
+#saveRDS(ais_near_08, file = "D:AIS/ais_near_08.rda")
+#saveRDS(ais_near_10, file = "D:AIS/ais_near_10.rda")
+#saveRDS(ais_near_12, file = "D:AIS/ais_near_12.rda")
+#saveRDS(ais_near_13, file = "D:AIS/ais_near_13.rda")
+#saveRDS(ais_near_14, file = "D:AIS/ais_near_14.rda")
+#saveRDS(ais_near_16, file = "D:AIS/ais_near_16.rda")
+#saveRDS(ais_near_18, file = "D:AIS/ais_near_18.rda")
+#saveRDS(ais_near_19, file = "D:AIS/ais_near_19.rda")
+#saveRDS(ais_near_20, file = "D:AIS/ais_near_20.rda")
+#saveRDS(ais_near_21, file = "D:AIS/ais_near_21.rda")
+#saveRDS(ais_near_22, file = "D:AIS/ais_near_22.rda")
+#saveRDS(ais_near_23, file = "D:AIS/ais_near_23.rda")
+```
+
+Load rds files into R with AIS points near buoys
+```{r}
+ais_near_07 <- readRDS("D:/AIS/ais_near_07.rda")
+ais_near_08 <- readRDS('D:/AIS/ais_near_08.rda')
+ais_near_10 <- readRDS('D:/AIS/ais_near_10.rda')
+ais_near_12 <- readRDS('D:/AIS/ais_near_12.rda')
+ais_near_13 <- readRDS('D:/AIS/ais_near_13.rda')
+ais_near_14 <- readRDS('D:/AIS/ais_near_14.rda')
+ais_near_16 <- readRDS('D:/AIS/ais_near_16.rda')
+ais_near_18 <- readRDS('D:/AIS/ais_near_18.rda')
+ais_near_19 <- readRDS('D:/AIS/ais_near_19.rda')
+ais_near_20 <- readRDS('D:/AIS/ais_near_20.rda')
+ais_near_21 <- readRDS('D:/AIS/ais_near_21.rda')
+ais_near_22 <- readRDS('D:/AIS/ais_near_22.rda')
+ais_near_23 <- readRDS('D:/AIS/ais_near_23.rda')
+```
+
+Match AIS points with buoy points
+```{r}
+gpsClose07 <- addAIS(GPSwhale7, ais_near_07, interpType = 'close', interpTime = 120)
+gpsClose08 <- addAIS(GPSwhale8, ais_near_08, interpType = 'close', interpTime = 120)
+gpsClose10 <- addAIS(GPSwhale10, ais_near_10, interpType = 'close', interpTime = 120)
+gpsClose12 <- addAIS(GPSwhale12, ais_near_12, interpType = 'close', interpTime = 120)
+gpsClose13 <- addAIS(GPSwhale13, ais_near_13, interpType = 'close', interpTime = 120)
+gpsClose14 <- addAIS(GPSwhale14, ais_near_14, interpType = 'close', interpTime = 120)
+gpsClose16 <- addAIS(GPSwhale16, ais_near_16, interpType = 'close', interpTime = 120)
+gpsClose18 <- addAIS(GPSwhale18, ais_near_18, interpType = 'close', interpTime = 120)
+gpsClose19 <- addAIS(GPSwhale19, ais_near_19, interpType = 'close', interpTime = 120)
+gpsClose20 <- addAIS(GPSwhale20, ais_near_20, interpType = 'close', interpTime = 120)
+gpsClose21 <- addAIS(GPSwhale21, ais_near_21, interpType = 'close', interpTime = 120) #empty table
+gpsClose22 <- addAIS(GPSwhale22, ais_near_22, interpType = 'close', interpTime = 120)
+gpsClose23 <- addAIS(GPSwhale23, ais_near_23, interpType = 'close', interpTime = 120)
+```
+
+Create column 'shipsin10km' that displays number of ships associated with timestamp
+```{r}
+# Viewing only ships within 10 km of buoys.
+close2ships07 <- gpsClose07 %>% filter(shipDist < 10e3)
+close2ships08 <- gpsClose08 %>% filter(shipDist < 10e3)
+close2ships10 <- gpsClose10 %>% filter(shipDist < 10e3)
+close2ships12 <- gpsClose12 %>% filter(shipDist < 10e3)
+close2ships13 <- gpsClose13 %>% filter(shipDist < 10e3)
+close2ships14 <- gpsClose14 %>% filter(shipDist < 10e3)
+close2ships16 <- gpsClose16 %>% filter(shipDist < 10e3)
+close2ships18 <- gpsClose18 %>% filter(shipDist < 10e3)
+close2ships19 <- gpsClose19 %>% filter(shipDist < 10e3)
+close2ships20 <- gpsClose20 %>% filter(shipDist < 10e3)
+#close2ships21 <- gpsClose21 %>% filter(shipDist < 10e3) #not long enough of a buoy to detect
+close2ships22 <- gpsClose22 %>% filter(shipDist < 10e3)
+close2ships23 <- gpsClose23 %>% filter(shipDist < 10e3)
+
+#Add column of ROUNDED UTC values to the lowest minute. Allows common field to bind
+# Close2ships21 removed because no ships nearby
+sslist <- c("close2ships07", "close2ships08", "close2ships10", "close2ships12", "close2ships13",
+        "close2ships14", "close2ships16", "close2ships18", "close2ships19", "close2ships20",
+        "close2ships22", "close2ships23")
+
+for (sx in sslist) {
+  dta <- get(sx)
+  dta$UTCrnd <- lubridate::floor_date(as.POSIXct(dta$UTC), "minute")
+  assign(sx, dta)
+}
+
+# Function that creates a column with number of ships within 10 km (shipin10km)
+#Bind rows with ship nearby
+#Summarize unique ships and display number of nearby vessels
+ShpRows <- function(w, s) {
+  z <- inner_join(w, s, by = join_by("UTC" == "UTCrnd")) %>%
+    select(UTC, UTC.y, MMSI, vesselLength, vesselType, shipLat, shipLong, shipDist)%>%
+    group_by(UTC) %>%
+    summarize(shipin10km = n_distinct(MMSI))
 }
@@
 
+Join to main df, add zeros to null values
 ```{r}
-#########
-# AIS
-#########
+nearbyShp07 <- ShpRows(GPSwhale7, close2ships07)
+GPSwhaleAIS7 <- left_join(GPSwhale7, nearbyShp07, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
 
-# Need to add more AIS dates in data folder
-AISlist <- list.files(path = paste0("data-raw/"), pattern = "AIS_", recursive = TRUE)
+nearbyShp08 <- ShpRows(GPSwhale8, close2ships08)
+GPSwhaleAIS8 <- left_join(GPSwhale8, nearbyShp08, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
 
-for (v in AISlist) {
-  Adata <- read_csv(paste0("data-raw/", v), show_col_types = FALSE) %>%
-    filter(Length >= 100) %>%                  # Filter out vessels over 100m in length
-    dplyr::select(-SOG,-COG, -CallSign, -IMO, -Cargo, -Draft, -Status) %>%  #Remove unnecessary columns
-    mutate(variable = paste0(substr(v, 16, 17), substr(v, 6, 7)))
+nearbyShp10 <- ShpRows(GPSwhale10, close2ships10)
+GPSwhaleAIS10 <- left_join(GPSwhale10, nearbyShp10, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
 
-  assign(paste0("AIS_", substr(v, 5, 10)), Adata)
-}
+nearbyShp12 <- ShpRows(GPSwhale12, close2ships12)
+GPSwhaleAIS12 <- left_join(GPSwhale12, nearbyShp12, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
 
-AIS_sf <- AIS_091218 %>%
-  st_as_sf(coords = c("LON", "LAT"), crs=4326, remove=F)
+nearbyShp13 <- ShpRows(GPSwhale13, close2ships13)
+GPSwhaleAIS13 <- left_join(GPSwhale13, nearbyShp13, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
+
+nearbyShp14 <- ShpRows(GPSwhale14, close2ships14)
+GPSwhaleAIS14 <- left_join(GPSwhale14, nearbyShp14, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
+
+nearbyShp16 <- ShpRows(GPSwhale16, close2ships16)
+GPSwhaleAIS16 <- left_join(GPSwhale16, nearbyShp16, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
+
+nearbyShp18 <- ShpRows(GPSwhale18, close2ships18)
+GPSwhaleAIS18 <- left_join(GPSwhale18, nearbyShp18, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
+
+nearbyShp19 <- ShpRows(GPSwhale19, close2ships19)
+GPSwhaleAIS19 <- left_join(GPSwhale19, nearbyShp19, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
+
+nearbyShp20 <- ShpRows(GPSwhale20, close2ships20)
+GPSwhaleAIS20 <- left_join(GPSwhale20, nearbyShp20, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
+
+# Manually create column for buoys with no vessel activity
+GPSwhaleAIS21 <- GPSwhale21 %>% add_column(shipin10km = 0)
+
+nearbyShp22 <- ShpRows(GPSwhale22, close2ships22)
+GPSwhaleAIS22 <- left_join(GPSwhale22, nearbyShp22, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
+
+nearbyShp23 <- ShpRows(GPSwhale23, close2ships23)
+GPSwhaleAIS23 <- left_join(GPSwhale23, nearbyShp23, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
 ```
 
+Save new fulltrack rdas to D drive
 ```{r}
-##################################
-# Create 10 km buffer around buoys (taken from M. McKenna's code)
-# Should be malleable for each buoy's number to easily plug in.
-##################################
-tracks <- readRDS("data-raw/tracks.rda")
-
-trk8 <- tracks %>% 
-  dplyr::filter(station == "8") %>%
-  st_as_sf(coords= c("Longitude", "Latitude"), crs=4326, remove=F)
-
-pois <- st_transform(trk8, crs = 4326)    # Specify a crs
-pts_buff <- st_buffer(pois, dist = 10000)  # Creates 10 km buffer around buoy points 10 km (10000m)
-buff <- st_transform(pts_buff, crs = 4326)  # Specify crs for new buffer
-
-### Crop full AIS to 10km buffer around drift 8 ###
-AIS_crop <- st_crop(x = AIS_sf, y = buff)
+saveRDS(GPSwhaleAIS7, file = "data/GPSwhaleAIS7.rda")
+saveRDS(GPSwhaleAIS8, file = "data/GPSwhaleAIS8.rda")
+saveRDS(GPSwhaleAIS10, file = "data/GPSwhaleAIS10.rda")
+saveRDS(GPSwhaleAIS12, file = "data/GPSwhaleAIS12.rda")
+saveRDS(GPSwhaleAIS13, file = "data/GPSwhaleAIS13.rda")
+saveRDS(GPSwhaleAIS14, file = "data/GPSwhaleAIS14.rda")
+saveRDS(GPSwhaleAIS16, file = "data/GPSwhaleAIS16.rda")
+saveRDS(GPSwhaleAIS18, file = "data/GPSwhaleAIS18.rda")
+saveRDS(GPSwhaleAIS19, file = "data/GPSwhaleAIS19.rda")
+saveRDS(GPSwhaleAIS20, file = "data/GPSwhaleAIS20.rda")
+saveRDS(GPSwhaleAIS21, file = "data/GPSwhaleAIS21.rda")
+saveRDS(GPSwhaleAIS22, file = "data/GPSwhaleAIS22.rda")
+saveRDS(GPSwhaleAIS23, file = "data/GPSwhaleAIS23.rda")
 ```
-
-```{r}
-# Intersect buffer with AIS. Starting this on first 10 rows to see how slowly it goes. 
-## FIX ME
-#inters <- st_intersection(buff, AIS_crop[1:140,])
-
-#Another way to view AIS points that fall within buffer. DOES NOT consider time.
-AIS_within_buff <- AIS_crop[buff,]
-f <- buff[AIS_within_buff,]
-by <- join_by(closest(UTC >= BaseDateTime))
-
-d <- full_join(buff, as.data.frame(AIS_within_buff), by)
-
-e <- d[!duplicated(d[c('BaseDateTime')]),]
-
-trk_AIS <- left_join(as.data.frame(trk8), as.data.frame(AIS_within_buff), by)
-
-proj4string <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
-trk_AIS_sf <- trk_AIS %>% st_as_sf(coords = c("Longitude", "Latitude"), crs=projcrs, remove=F)# %>%
-   st_set_crs(4326)
-trk_AIS_sf <- st_transform(trk_AIS_sf, crs = 4326)
-
-#inters <- st_intersection_faster(buff, AIS_crop[1:140,])
-#inters2 <- st_intersection_faster(buff, AIS_crop[141:200,])
-#inters3 <- st_intersection_faster(buff, AIS_crop[201:300,])
-
-#Another way to view AIS points that fall within buffer. DOES NOT consider time.
-
-st_is_within_distance(AIS_crop, pois, dist = 10000)
-st_distance(AIS_crop, pois, dist_fun = 10000)
-
-st_set_geometry(AIS_crop, value = NULL)
-st_set_geometry(pois, value = NULL)
-
-AIS_crop <- as_Spatial(AIS_crop)
-pois <- as_Spatial(pois)
-AIS_within_buff <- as_Spatial(AIS_within_buff)
-z <- sp_distance(AIS_crop, pois, features = TRUE, unit = "km")
-
-data_places <- pois@data
-data_places$distance <- sp_distance(AIS_within_buff, pois, features = TRUE, unit = "km")
-View(data_places)
-
-rgeos::gWithinDistance(AIS_crop, pois, features = TRUE, dist = 10000)
-```
-```{r}
-# For each row in AIS df,
-  # calculate distance to closest drift
-  # return smallest distance
-  # attach column to df
-
-# 
-ship_dist <- (geosphere::distHaversine(AIS_within_buff@coords[10:13,], pois))/1000
-cbind(AIS_within_buff@data[10:13,], ship_dist)
-```
-
-```{r}
-# Plot showing 10km buffer around drift 8
-#tmap_mode("plot")
-#tm_shape(buff) + tm_symbols(col = "blue", border.col = NA) +
-#  tm_shape(trk) + tm_dots(col = "red")
-```
-
-plots
-```{r}
-# Plot
-ggplot(tracks_SS_08, aes(x = long, y = lat.x)) + geom_point() 
- + geom_sf(data = bbox8, aes(col = "red"))
-
-# Plot showing AIS bounding box around drift 8 tracks
-ggplot() + geom_sf(data = trk_AIS_sf, color = "red")
-ggplot() + geom_sf(data = buff[lgth,] ,color = "red")
-
-ggplot() +
-  geom_sf(data = calif) + coord_sf(default_crs = sf::st_crs(4326)) + theme_bw()
- # geom_sf(data = AIS_sf)
-
-p <- ggplot() + geom_sf(data = buff, mapping = aes(x=Longitude,y=Latitude)) +
-  geom_point(data = AIS_within_buff, mapping = aes(x=LON,y=LAT, group = VesselName), color = "red")
-
-anim <- p +
-  transition_reveal(along = BaseDateTime)+
-  ease_aes('linear')+
-  #shadow_wake(0.1) +
-  ggtitle("Date: {frame_along}")
-
-animate(anim, nframes = 180, fps = 3)
-```

### 2023-09-13 01_readin.R
Update 01_readin.R
updated with better GPS pts

@@ -1,11 +1,12 @@
 ## code to assemble clean dataframe for each buoy.
-## Prerequisites include whales.R, tracks.R, soundscape_metrics.R.
+## Prerequisites include whales.R, tracks.R
+## Results in dataframe for each buoy containing whale detections and buoy GPS points.
 
 
-# libraries ##########
+# libraries #######
 
 library(tidyverse)
-library(lubridate)
+library(PAMpal)
 library(here)
 
 # set up #####
@@ -14,29 +15,21 @@ library(here)
 Sys.setenv(TZ='UTC')
 here()
 
-######################
-# Join TRACKS with WHALES (Start Rmd here)
-######################
-tracks <- readRDS('data/tracks.rda')
+# Read in GPS tracks and whale detections
+AllTracks <- readRDS('data/AllTracks.rda')
 whales <- readRDS('data/whales.rda')
-
-# Join tracks and whales
-trk_whale <- left_join(tracks, whales, by = c("station","UTC"))
-
-# Remove duplicates
-trk_whale <- trk_whale[!duplicated(trk_whale[c('UTC', 'station')]),]
 
 ###############
 # Filter whales by drift
 ##############
 
 # Create list of drifts
-d <- unique(trk_whale$station)
+#d <- unique(trk_whale$station)
 
 # Filter by drifts and assign to new variable
-for (drift in d) {
-  assign(paste0("trk_whales", drift), trk_whale %>% dplyr::filter(station == drift))
-}
+#for (drift in d) {
+#  assign(paste0("trk_whales", drift), trk_whale %>% dplyr::filter(station == drift))
+#}
 # Results in trk_whales7, trk_whales8... objects
 
 ##############################################################
@@ -44,45 +37,395 @@ for (drift in d) {
 ##############################################################
 
 ## Make list of rda files to read in
-iilist <- c("BB_07", "BB_08", "BB_10", "BB_12", "BB_13", "BB_14", "BB_16", 
-            "BB_18", "BB_19", "BB_20", "BB_21", "BB_22", "BB_23", "TO_07",
-            "TO_08", "TO_10", "TO_12", "TO_13", "TO_14", "TO_16", "TO_18", 
-            "TO_19", "TO_20", "TO_21", "TO_22", "TO_23")
+#iilist <- c("BB_07", "BB_08", "BB_10", "BB_12", "BB_13", "BB_14", "BB_16", 
[…]

### 2023-10-16 RF_ZC_presence_7-11_edits.R
Update RF_ZC_presence_7-11_edits.R
edited RF models with randomForest instead of rfPermute

@@ -29,8 +29,7 @@ allDrifts$ZcPresence<-factor(ifelse(allDrifts$species=='ZC',1,0))
  #                      'sst_mean','dist2slope','depth',
   #                     'chlorophyll_mean','mldDepth'))
 #more parameters
-covariate.list<-list(c('ZcPresence','BB_20.24000','TOL_63','TOL_125','TOL_2000','TOL_5000',
-                       'TOL_20000', 'curl_mean', 'bathy_slope', 'ssh_mean',  
+covariate.list<-list(c('ZcPresence', 'curl_mean', 'bathy_slope', 'ssh_mean',  
                        'sst_mean','dist2slope','depth',
                        'chlorophyll_mean','mldDepth', 'mldTemp', 'ttDepth',
                        'ttTemp', 'temp400', 'sal400'))
@@ -53,14 +52,34 @@ set.seed(123)
 
 # create balanced sample size random forest model
 #By ZC
-RF.model.ZC <- rfPermute(ZcPresence ~ ., data=DF.modelZC[,include.covars],
-                            replace=FALSE, ntree=15000, sampsize=sampsizeZC, proximity=FALSE)
-RF.model.ZC
+#RF.model.ZC <- rfPermute(ZcPresence ~ ., data=DF.modelZC[,include.covars],
+#                            replace=FALSE, ntree=15000, sampsize=sampsizeZC, proximity=FALSE)
+#RF.model.ZC
+# Error message - "Error in randomForest.default(x = list(BB_20.24000 = c(101.991578, 98.486802,  : 
+#'R_Calloc' could not allocate memory (9976 of 8 bytes)"
+
+#Evaluate RF models
+#Create a PDF of summary plots
+#pdf(paste(Dep,"_",string.covars.used,".pdf"), width=14, height=10)
+#round(importance(RF.model.ZC),3) #ranking of importance for each variable
+#plotImportance(RF.model.ZC) #visualize importance
+#plotTrace(RF.model.ZC)   #model stability w/number of trees (this should be flat!)
+#plotImpPreds(RF.model.ZC, DF.modelZC, "ZcPresence")  #distribution of predictors 
+#plotPredictedProbs(RF.model.ZC)
+#plotProximity(RF.model.ZC) ## Not working??
+#dev.off()
+
+#Confusion matrix
+#confusionMatrix(RF.model.ZC)
+
[…]

### 2023-10-16 maps.Rmd
Create maps.Rmd
moved to new rmd

--- 2023-04-19 03_exploratory.Rmd
@@
 ---
-title: "exploratory analysis"
+title: "maps"
+author: "Marina Bozinovic"
+date: "2023-06-19"
 output: html_document
-date: "2022-11-06"
 ---
 
-Required Packages
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(tidyverse)
-library(lubridate)
-library(sf)
-library(rerddap)
-library(terra)
-library(marmap)
-library(rnaturalearth)
-library(mapproj)
-library(raster)
-#library("rerddapXtracto")
-library(PAMpal)
-library(PAMmisc)
-library(maptiles)
-library(tmap)
-library(spData)
-library(tmaptools)
-library(grid)
-library(ggspatial)
-```
+### Plot maps for Powerpoint images
 
@@
 
-Create sf objects and basemaps for maps
+Create sf objects and basemaps for maps (not needed for environmental variables)
 ```{r}
-whales_8_10sf <- BWsf %>%                            
+whales_8_10sf <- Wsf %>%                            
 dplyr::filter(station == "8" | station == "10")
@@
 ```
-### Exploratory analysis
-
-SST from PAMPal
-```{r}
-# Downloads data from servers, uncomment and only run once
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'jplMURSST41', fileName = "analysedSST")
-
-# from file 
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/analysedSST.nc') %>%
-  rename(., sst_mean = analysed_sst_mean) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)   # rename new column to "sst_mean" and remove unnecessary columns
-```
-
-Distance to continental slope
-```{r}
-# Increase cell size by 30x to be visible
-bathy <- rast("data/gebco_bathy.tif")
-agg <- terra::aggregate(bathy, 5)
-plot(agg)
-
-# Rasterize track with agg as the template
-
-#TOL_rast <- rasterize(trk_TOL_08, agg, field = "TOL_125")
-sf08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-sf10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
-trkraster08 <- rasterize(vect(sf08), agg)
-trkraster10 <- rasterize(vect(sf10), agg)
-
-bathy_df <- as.data.frame(bathy, xy = TRUE)
-
-# Map the coastline and cont. slope
-cont <- st_as_sf(as.contour(bathy))
-coast <- cont %>% filter(level == 0)
-cont_slope <- cont %>% filter(level == -2000)
-cont_slope_rast <- rasterize(cont_slope, agg, field = "level")
-plot(cont_slope_rast, col = "red")
-cont_slope_rast <- rasterize(cont_slope, agg)
-plot(cont_slope_rast, col = "red")
-
-#ggplot() + 
-#  geom_raster(data = bathy_df, 
-#              mapping = aes(x=x, y=y, 
-#                            alpha = 0.8,
-#                            fill = `gebco_bathy`))+
-#  geom_sf(data = cont, col="black") + 
-#  geom_sf(data = coast, col = "red") +
-#  geom_sf(data = cont_slope, col = "purple") +
-#  geom_sf(data = whales_8_10sf, mapping = aes()) +
-#  scale_color_gradient(low="green", high="red") 
-```
-
-### Extract slope_dist and bind to drift tracks
-```{r}
-# Drift 8
-slopedist <- terra::distance(trkraster08, vect(cont_slope))     # measure distance from track to slope
-extdist <- terra::extract(slopedist, vect(sf08))       # extract this distance into a df
-tracks_SS_08 <- cbind(sf08,extdist) %>%                     # bind to main df
-  dplyr::select(-ID) %>%
-  rename("dist2slope" = gebco_bathy) %>%
-  st_set_geometry(NULL)                         # Drop geometry to make df for future cbinds
-
-#Drift 10
-slopedist <- terra::distance(trkraster10, vect(cont_slope))     # measure distance from track to slope
-extdist <- terra::extract(slopedist, vect(sf10))       # extract this distance into a df
-tracks_SS_10 <- cbind(sf10,extdist) %>%                     # bind to main df
-  dplyr::select(-ID) %>%
-  #rename("dist2slope" = gebco_bathy) %>%
-  st_set_geometry(NULL)                        # Drop geometry to make df for future cbinds
-```
-
-#### Bathymetry - PAMpal
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'srtm15plus', fileName='depth.nc')
-
-# Pulls from downloaded file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/depth.nc') %>%
-  rename(., depth = z_mean) %>%        # rename new column to "depth"
-  mutate(depth = depth*-1) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-```
-
-#### Bathymetric slope
-
-```{r}
-# Slope (Hard-coded - can fix this)
-bgo <- getNOAA.bathy(lon1 = -121, lon2 = -131,
-                     lat1 = 34, lat2= 40, resolution = 1)  # plot bounding box around drift 8 and 10
-# As raster
-bgo_rast <- marmap::as.raster(bgo)
-bgo_slope <- terrain(bgo_rast, v = "slope")
-
-# Extract slope values at points along drift 8
-slope <- raster::extract(bgo_slope, tracks_SS_08[,4:3])
-tracks_SS_08 <- cbind(tracks_SS_08, slope)
-```
-
-#### Distance from coast - PAMPal
-Need to have variable choose itself!! Also, how do I get it to not download each time?
-```{r}
-#dist2shore <- erddapToEdinfo(dataset='dist2coast_1deg',
-#                             baseurl='https://pae-paha.pacioos.hawaii.edu/erddap/',
-#                             chooseVars = FALSE)
-
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc='data/Dist2Shore.nc') %>%
-  rename(., dist2shore = dist_mean) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-
-#May not need this part
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'dist2coast_1deg', fileName='dist2coast.nc', var = 'dist')
-
-# Pulls from downloaded file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/dist2coast.nc')
-
-# Downloads nc file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=dist2shore, fileName='Dist2Shore.nc')  
-```
-
-### Chlorophyll from PAMPal
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdMH1chlamday', fileName='chlorophyll.nc')
-
-# Pulls from downloaded file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/chlorophyll.nc')
-```
-
-#### Sea Surface Height Anomalies from PAMPal
-```{r}
-# Downloaded from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'nesdisSSH1day', fileName='sshanomalies.nc')
-
-# Pulled from downloaded nc file
-tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'data/sshanomalies.nc') %>%
-  rename(., ssh_mean = sla_mean) %>%
-  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)
-```
-
-### Wind Stress Curl from PAMPal *** NEEDS FIXING ****
-```{r}
-# Downloads from server
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQMstress1day_LonPM180', fileName='erdQMstress1day_LonPM180.nc', var = 'curl')
-
-# Pulls from downloaded file
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc = 'erdQMstress1day_LonPM180.nc')
-#tracks_SS_08
-
-#windcurl <- erddapToEdinfo('erdQMstress1day', baseurl = 'https://coastwatch.pfeg.noaa.gov/erddap/')
-
-#tracks_SS_08 <- matchEnvData(tracks_SS_08, nc=windcurl, fileName='windcurl.nc')  
-```
-
-#### HyCOM Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
-Doesn't bin itself into separate dataframe based on size!
-```{r}
-# These refer to nc files that have already been downloaded from the server.
-# Uncomment and only run once! Rows are hard-coded.
-
-#hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm0.nc")
-
-#hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm1.nc")
-
-#hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm2.nc")
-
-#hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm3.nc")
-
-#hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm4.nc")
-
-#hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm5.nc")
-
-#hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm6.nc")
-
-#hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm7.nc")
-
-#hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, fileName = "hycomm8.nc")
-
-# Run the 9 nc lists that make up tracks_SS_08
-hycomData0 <- matchEnvData(as.data.frame(tracks_SS_08[1:250,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm0.nc")
-hycomData1 <- matchEnvData(as.data.frame(tracks_SS_08[251:500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm1.nc")
-hycomData2 <- matchEnvData(as.data.frame(tracks_SS_08[501:750,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm2.nc")
-hycomData3 <- matchEnvData(as.data.frame(tracks_SS_08[751:1200,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm3.nc")
-hycomData4 <- matchEnvData(as.data.frame(tracks_SS_08[1201:1500,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm4.nc")
-hycomData5 <- matchEnvData(as.data.frame(tracks_SS_08[1501:1800,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm5.nc")
-hycomData6 <- matchEnvData(as.data.frame(tracks_SS_08[1801:2145,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm6.nc")
-hycomData7 <- matchEnvData(as.data.frame(tracks_SS_08[2146:2469,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm7.nc")
-hycomData8 <- matchEnvData(as.data.frame(tracks_SS_08[2470:2769,]), var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000, nc = "hycomm8.nc")
-```
-
-Create function to estimate thermocline variables using the temperature at depth data
-```{r}
-calcThermoVars <- function(depth, temp, mldMode='SST8', thermMode='variso', plot=FALSE) {
-    # depthInterp <- approxfun(x=temp, y=depth)
-    switch(match.arg(mldMode),
-           'SST8' = {
-               mldTemp <- temp[1] - 0.8
-               # first few temps are likely to have repeat values that trigger
-               # warnings for approxfun, so cut off to only values after mldTemp
-               useMin <- min(which(temp <= mldTemp)) - 1
-               useIx <- useMin:length(temp)
-               depthInterp <- approxfun(x=temp[useIx], y=depth[useIx])
-               mld <- depthInterp(mldTemp)
-           }
-    )
-    switch(match.arg(thermMode),
-           'variso' = {
-               t400 <- approx(x=depth, y=temp, xout=400)$y
-               # therm end is half way to t400 temp
-               tEnd <- mldTemp + (t400 - mldTemp)/2
-               # therm temp is midpt from mld to end
-               tt <- (mldTemp + tEnd)/2
-               # tt <- mldTemp - .25 * (mldTemp - t400)
-               ttDepth <- depthInterp(tt)
-               tEndDepth <- depthInterp(tEnd)
-           }
-    )
-    if(plot) {
-        to300 <- depth <= 300
-        plot(x=temp[to300], y=-depth[to300],
-             main = paste0('MixLayer: ', round(mld,0), 'm ', round(mldTemp, 1), 'C',
-                           '\nTherm: ', round(ttDepth, 0), 'm ', round(tt, 1), 'C'),
-             xlab='Temp (C)', ylab='Depth (m)')
-
-        lines(x=temp[to300], y=-depth[to300])
-        lines(x=c(mldTemp, tEnd), y=c(-mld, -tEndDepth), col='darkgray', lwd=4)
-        points(x=c(mldTemp, tt), y=c(-mld, -ttDepth), col='blue', cex=1.5, pch=15)
-    }
-    list(mldDepth = mld, mldTemp = mldTemp,
-         ttDepth = ttDepth, ttTemp = tt)
-}
-```
-
-Use above function on the smaller hycom datasets (0-6)
-```{r}
-#calcThermoVars(hycomData[[1]]$matchDepth, hycomData[[1]]$water_temp, plot=TRUE)
-
-thermData0 <- bind_rows(lapply(hycomData0, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData1 <- bind_rows(lapply(hycomData1, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData2 <- bind_rows(lapply(hycomData2, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData3 <- bind_rows(lapply(hycomData3, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData4 <- bind_rows(lapply(hycomData4, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData5 <- bind_rows(lapply(hycomData5, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData6 <- bind_rows(lapply(hycomData6, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData7 <- bind_rows(lapply(hycomData7, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-thermData8 <- bind_rows(lapply(hycomData8, function(x) {
-    # do thermo calcs
-    result <- calcThermoVars(x$matchDepth, x$water_temp)
-    # get specific values
-    result$temp400 <- x$water_temp[x$matchDepth == 400]
-    result$sal400 <- x$salinity[x$matchDepth == 400]
-    result
-}))
-
-# Bind all thermocline datasets together
-thermData <- rbind(thermData0, thermData1, thermData2, thermData3, thermData4, 
-                   thermData5, thermData6, thermData7, thermData8)
-
-# Bind main thermocline dataset to main track/soundscape dataframe
-tracks_SS_08 <- cbind(tracks_SS_08, thermData)
-
-# New tracks_SS_08 variable contains Hycom variables now
-##### FINAL tracks_ss_08 dataframe ######
-View(tracks_SS_08)
-```
-
-Get avg daily environmental variables
-```{r}
-means_08 <- tracks_SS_08 %>%
-  group_by(as.Date(UTC))%>%
-  summarize(sst_mean = mean(sst_mean),
-            dist_slope = mean(dist2slope),
-            depth = mean(depth),
-            slope = mean(slope),
-            dist_shore = mean(dist2shore),
-            chlorophyll_mean = mean(chlorophyll_mean),
-            ssh_mean = mean(ssh_mean),
-            mldDepth = mean(mldDepth),
-            mldTemp = mean(mldTemp),
-            ttDepth = mean(ttDepth),
-            ttTemp = mean(ttTemp),
-            temp400 = mean(temp400),
-            sal400 = mean(sal400),
-            BWpresence = sum(BWpresence))
-```
-
-### Exploratory Analysis ######
-
-### Third octave levels histograms
-
-```{r}
-# Drift 8 - all TOL bands, facet wrap
-pivot <- tracks_SS_08 %>%
-  pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of third octave level intensities: Drift 8") +
-  labs(x = "Sound Intensity (dB)") +
-  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
-  scale_fill_manual(values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.7,0.8)) +
-  facet_grid(~SL)
-
-# Drift 10 - all TOL bands, facet grid
-pivot2 <- tracks_SS_10 %>%
-  pivot_longer(cols = starts_with("TOL")| starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of third octave level intensities: Drift 10") +
-  scale_x_continuous(name = "Sound Intensity (dB)", breaks=seq(50,110,20)) +
-  guides(fill = guide_legend(title = "Beaked Whale Presence")) +
-  scale_fill_manual(values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.76)) +
-  facet_grid(~SL)
-
-# Drift 8 - TOL bands 125 and 2000 only
-pivot3 <- tracks_SS_08 %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz")
-```
-
-```{r drift-8-histogram}
-ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 8") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Beaked Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
-```
-
-```{r}
-# Drift 10 - TOL bands 125 and 2000 only
-pivot4 <- tracks_SS_10 %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot4, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 10") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Beaked Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
-
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz")
-```
-
-Scatterplots
-
-```{r}
-# SST vs TOL 2000
-ggplot(data = tracks_SS_08, aes(x = sst_mean, y = `TOL_2000`)) +
-  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
-
-# SST over time
-ggplot(tracks_SS_08, aes(x = UTC, y = sst_mean)) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-  #geom_smooth(method = "loess")
-
-# Depth vs TOL 125
-ggplot(data = tracks_SS_08, aes(x = depth, y = `TOL_125`)) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-
-#Mean depth over time
-ggplot(means_08, aes(x = `as.Date(UTC)`, y = depth*(-1))) +
-  geom_line()
-
-# Chlorophyll vs TOL 125
-ggplot(data = tracks_SS_08, aes(x = chlorophyll_mean, y = `TOL_125`)) +
-  geom_point(aes(color = species))
-
-# CHl over time
-ggplot(tracks_SS_08, aes(x = UTC, y = chlorophyll_mean)) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-  #geom_smooth(method = "loess")
-
-# Mixed Layer Depth vs TOL125 with species
-ggplot(data = tracks_SS_08, aes(x = mldDepth, y = `TOL_125`)) +
-  geom_point(aes(color = species))
-
-# Mixed Layer Depth over time
-ggplot(tracks_SS_08, aes(x = UTC, y = mldDepth*(-1))) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-
-ggplot(data = tracks_SS_08, aes(x = slope, y = `TOL_5000`)) +
-  geom_point(aes(color = species))
-```
-
-Facet Grids for Env. Variables
-```{r}
-#For surface/depth variables USING DAILY MEAN VARIABLES
-# See above for means_08 derivation
-track08Long <- means_08 %>%
-  pivot_longer(cols = sst_mean:ssh_mean,
-               names_to="parameter",
-               values_to="value") %>%
-  filter(parameter %in% c("sst_mean", "dist_shore", "chlorophyll_mean", "ssh_mean")) %>%
-  rename(Date = `as.Date(UTC)`)
-
-p <- ggplot(data = track08Long, aes(x = Date, y = value)) +
-  geom_line()
- # geom_point(aes(size = BWpresence))   ### How best to represent number of BW detections per day?
-
-p + facet_grid(parameter ~ ., scales = "free_y")
-
-# For MLD/Thermocline variables 
-track08Long2 <- tracks_SS_08 %>%
-  pivot_longer(cols = mldDepth:sal400,
-               names_to="parameter",
-               values_to="value") %>%
-  filter(parameter %in% c("mldDepth", "mldTemp", "ttDepth", "ttTemp", "temp400", "sal400")) %>%
-    rename(Date = UTC) %>%
-    subset(species!="?BW")
-
-q <- ggplot(data = track08Long2, aes(x = Date, y = value)) +
-  geom_line() + geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
-q + facet_grid(parameter ~ ., scales = "free_y")
-
-# For temp400 and salinity400 variables
-env.labs <- c(`sal400` = "Salinity at 400m", 
-              `temp400` = "Temperature at 400m")
-
-track08Long3 <- tracks_SS_08 %>%
-  pivot_longer(cols = temp400:sal400,
-               names_to="parameter",
-               values_to="value") %>%
-  filter(parameter %in% c("temp400", "sal400")) %>%
-  rename(`Date (2018)` = UTC)
-
-q <- ggplot(data = track08Long3, aes(x = `Date (2018)`, y = value)) +
-  geom_line() + 
-  geom_point(aes(color = species)) + 
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43", "BW" = "BW"))
-q + facet_grid(parameter ~ ., scales = "free_y", labeller = as_labeller(env.labs))
-```
-
-Co-occurrence plots of drift 8 and vessels 
-```{r}
-#### 09/04/18 co-occurrence of vessel and drift within 10 km
-track08_09.04 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-04 00:00:00" & UTC <= "2018-09-05 00:00:00")
-
-ggplot(data = track08_09.04, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
-  geom_text(aes(as.POSIXct("2018-09-04 14:13:00"), 88, label = "Enter vessel", hjust = 1.0)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-04 15:17:00"), color = "yellow", linewidth = 2) +
-  geom_text(aes(as.POSIXct("2018-09-04 15:13:00"), 87, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.04, aes(color = species, size = 2)) +
-  scale_color_discrete(na.translate=FALSE) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-04-18\n(Within 7 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/05/18 co-occurrence of vessel and drift within 10 km.
-track08_09.05 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-04 21:00:00" & UTC <= "2018-09-06 00:00:00")
-
-ggplot(data = track08_09.05, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-05 01:33:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-05 01:33:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-05 02:08:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-05 02:08:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.05, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-05-18\n(Within 5 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/10/18 co-occurrence of vessel and drift within 10 km.
-track08_09.10 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-10 00:00:00" & UTC <= "2018-09-11 00:00:00")
-
-ggplot(data = track08_09.10, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-10 12:58:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-10 12:58:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-10 14:01:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-10 14:01:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() +
-  geom_point(data = track08_09.10, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-10-18\n(Within 2 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/22/18 co-occurrence of vessel and drift within 10 km
-track08_09.22 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-22 00:00:00" & UTC <= "2018-09-23 00:00:00")
-
-ggplot(data = track08_09.22, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-22 19:13:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-22 19:13:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-22 19:34:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-22 19:34:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.22, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-22-18\n(Within 9 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/23/18 co-occurrence of vessel and drift within 10 km. NO whale detections in this time period.
-track08_09.23 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-23 00:00:00" & UTC <= "2018-09-24 00:00:00")
-
-ggplot(data = track08_09.23, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-23 03:55:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-23 03:55:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-23 04:23:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-23 04:23:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-23-18\n(Within 6 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-```
-
-BW Occurrences
-```{r}
-# Filter Drift 8 by BW presence and absence
-BW1 <- tracks_SS_08 %>%
-  filter(BWpresence == 1)
-BW0 <- tracks_SS_08 %>%
-  filter(BWpresence == 0)
-
-ggplot(data = tracks_SS_08, mapping = aes(TOL_125)) + geom_histogram(binwidth = 0.1, color = "black")
-
-ggplot() + 
-  geom_histogram(data = BW0, aes(TOL_5000), binwidth = 0.5, fill = "black", color = "white") +
-  geom_histogram(data = BW1, aes(TOL_5000), binwidth = 0.5, fill = "yellow", color = "black")
-
-# Percent of time BWs are detected
-print(paste(nrow(BW1) / nrow(tracks_SS_08) * 100,"%"))
-
-# Most abundant BW species when whale is detected (remove NA values)
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot() + geom_bar(aes(species))
-
-# Sound levels BW are exposed to by species
-ZC <- tracks_SS_08 %>%
-  filter(species == "ZC")
-
-BW43 <- tracks_SS_08 %>%
-  filter(species == "BW43")
-
-print(max(ZC$TOL_125))
-print(max(BW43$TOL_125))
-
-ggplot() + geom_histogram(data = ZC, mapping = aes(TOL_125), binwidth = 0.2, fill = "brown", color = "white") +
-  geom_histogram(data = BW43, mapping = aes(TOL_125), binwidth = 0.2, fill = "yellow", color = "white") +
-  theme_classic()
-```
-
-Density Plots - Species at TOL sound levels
-```{r}
-# Broadband
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(`BB_20.24000`, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("Broadband") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 63
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_63, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("63 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 125
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_125, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("125 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 2000
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_2000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("2000 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 5000
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_5000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("5000 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 20,000
-tracks_SS_08 %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_20000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("20,000 Hz") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-```
-
-Histograms with BW presence/absence and environmental variables
-```{r}
-# SST
-ggplot(data = tracks_SS_08, aes(sst_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.1, color = "black")
-
-# Depth
-ggplot(data = tracks_SS_08, aes(depth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 10, color = "black")
-
-# Bottom slope
-ggplot(data = tracks_SS_08, aes(slope, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
-
-# SSH
-ggplot(data = tracks_SS_08, aes(ssh_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
-
-# Chlorophyll
-ggplot(data = tracks_SS_08, aes(chlorophyll_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
-
-# Mixed layer depth Depth
-ggplot(data = tracks_SS_08, aes(mldDepth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.25, color = "black")
-
-# Mixed layer depth Temperature
-ggplot(data = tracks_SS_08, aes(mldTemp, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.1, color = "black")
-
-# Thermocline Depth
-ggplot(data = tracks_SS_08, aes(ttDepth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.25, color = "black")
-
-#Thermocline temperature
-ggplot(data = tracks_SS_08, aes(ttTemp, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.05, color = "black")
-
-# Temperature at 400m
-ggplot(data = tracks_SS_08, aes(temp400, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.01, color = "black")
-
-# Salinity at 400 m
-ggplot(data = tracks_SS_08, aes(sal400, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.001, color = "black")
-```
-
-### Plot maps for Powerpoint images
 

### 2023-10-16 manuscript.rmd
Update manuscript.rmd
intro changes

@@
 title: "manuscript"
-format: html
+format: docx
 editor: visual
 bibliography: references.bib
+fontsize: 12pt
 ---
@@
 
-Sound is a critical form of communication for many species in the ocean.
-Since sound travels over four times faster in water than in air and can carry information over far distances @urick1983 , the use of sound has evolved as a critical component of communication, navigation, and foraging for many invertebrates, fishes and cetaceans @duarte2021 .
-Under the right conditions, sounds have the ability to travel across ocean basins @miksis-olds.
+Sound serves as the main sensory input for many species in the ocean.
+Since sound travels over four times faster in water than in air and can carry information over far distances [@urick1983], the use of sound has evolved as a critical component of communication, navigation, and foraging for many invertebrates, fishes, and cetaceans [@duarte2021].
+Under the right conditions, sounds can travel across ocean basins [@miksis-olds2018].
 
-The total collection of sounds in an environment is called a soundscape, and these sources of sound are grouped into three broad categories: biophonies, sound produced by animals like cetaceans, fishes, and invertebrates; geophonies, natural sounds like rain, wind, and earthquakes; and anthropophonies, human-made sounds from shipping vessels, sonar, and survey activities (@miksis-olds, @duarte2021).
-Every region of the ocean experiences high spatio-temporal variation in sound levels and sources based on traits such as properties of water, organism diversity, human activities, and bathymetry @andré2011.
-Because of their unique sound signatures, ocean soundscapes are dynamic and can be characterized, measured, and monitored as indicators of ecosystem health or species diversity @radford2010 @haver2019 @weiss2021 @pijanowski2011 .
+The total collection of sounds in an environment is called a soundscape, and these sources of sound are grouped into three broad categories: biophonies, sound produced by animals like cetaceans, fishes, and invertebrates; geophonies, natural sounds like rain, wind, and earthquakes; and anthropophonies, human-made sounds from shipping vessels, sonar, and oceanographic survey activities [@miksis-olds2018; @duarte2021].
+Every region of the ocean experiences high spatio-temporal variation in sound levels and sources based on traits such as properties of water, organism diversity, human activities, and bathymetry [@andré2011].
+Because of their unique sound signatures, ocean soundscapes can be characterized, measured, and monitored as indicators of ecosystem health or species diversity [@radford2010; @pijanowski2011; @haver2019; @weiss2021].
 
-**\[what do they have to lose in a noisy environment?\]**
+When noise levels reach higher than normal levels, organisms can be negatively impacted.
+Invertebrates and fishes may experience stunted development, elevated stress responses, and physical injuries [@weilgart2018], while cetaceans' behavior, calling patterns, physiology and stress levels, and ability to locate prey or conspecifics are impacted [@weilgart2007].
+Sensitivities and responses to noise can vary among species and even individuals [@kunc2016].
+While terrestrial ecosystem-level effects have been well-documented [@francis2012; @shannon2016; @buxton2017], effects within aquatic ecosystems require further studies [@kunc2016].
 
-When noise levels reach above normal levels, organisms can be negatively impacted.
-Invertebrates and fishes may have stunted development and elevated stress responses (cite), fishes \[add negative behavior here\], and cetaceans' behavior, calling patterns, physiology and stress levels, ability to locate prey or conspecifics are impacted (cite).
-Each species is affected differently, some more sensitive to certain types of noise.
-While terrestrial ecosystem-level effects have been documented, (Buxton 2017, Shannon 2016, Francis 2012), those within aquatic ecosystems lack studies @kunc2016.
+Ocean noise pollution is among the anthropogenic changes that have recently grown to concerning levels [@tyack2008; @hildebrand2009].
+The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades and has a close correlation to global economic trends [@frisk2012].
+Ships have been increasing in terms of gross tonnage and fleet size [@hildebrand2009] and now account for over 80% of the global trade, with an estimated 1.4% expected annual increase in shipping through 2027 [@unctad2022].
+Commercial shipping now accounts for most of the low-frequency anthropogenic noise in the ocean [@hildebrand2009].
 
-**\[Noise pollution is a growing concern and increasing in the ocean.\]**
-
-Ocean noise pollution is among the anthropogenic changes that have recently grown to concerning levels @tyack2008, @hildebrand.
-The global ambient noise in the ocean is estimated to have increased by 12 decibels in the last few decades and has a close correlation to global economic trends @frisk2012, with commercial shipping accounting for most of the low-frequency anthropogenic noise in the ocean @hildebrand2009.
-Ships have been increasing in terms of gross tonnage and fleet size @hildebrand2009 (and age?), and now account for over 80% of the global trade, with an estimated 1.4% expected annual increase in shipping through 2027 @unctad.
-
-These noise-generating activities are categorized as either incidental or deliberate @chou2021, @chahouri2022.
-Deliberate activities such as active sonar pingers, resource extraction from the seabed, seismic surveys, and echosounders are intentionally produce sound to measure an area and can utilize both high and low frequencies.
-Incidental activities like dredging, offshore development, and shipping produce sound as a by-product @hawkins2017.
+These noise-generating activities are categorized as either incidental or deliberate [@chou2021; @chahouri2022].
+Deliberate activities such as active sonar pingers, resource extraction from the seabed, seismic surveys, and echosounders intentionally produce sound to measure an area and can utilize both high and low frequencies [@duarte2021].
+Incidental activities like dredging, offshore development, and shipping produce sound as a by-product [@hawkins2017; @southall2017].
 Anthropogenic noise can further be characterized by the frequency and duration of the sound.
-For example, shipping produces a continuous, low-frequency (\<200 Hz) sound that is widely distributed across the world's oceans @hildebrand2009.
-~~Because this type of noise can travel great distances, they can have effects further away from the source~~.
-Sonar for military surveillance or research such as seafloor mapping occupies the mid-frequency bands, but attenuates quicker, therefore impacting a smaller acoustic space.
-
-It remains unknown if chronic effects from continuous sound or acute effects from impulsive sounds can alter the normal behavior or physiology of cetaceans enough to affect reproduction and population levels.
-(though see @new2013 for habitat displacement affecting reproduction or survival)
-
-Lower frequencies travel further, and higher frequencies tend to attenuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ).
-Ambient noise in shallow water has more variability than deeper water (@richardson2013).
-Climate change is also expected to alter the temperature, salinity, and pH of the ocean, causing an additional increase in global sound speed due to changing absorption qualities @affatati2022 @kunc2016 .
-
-**\[This is an issue, but the extent of damage remains unclear, so it's difficult to produce legislation to that effect\]**
+For example, shipping occurs in all oceans and produces a continuous, low-frequency (\<200 Hz) sound that travels far distances [@hildebrand2009].
+Sonar for military surveillance or research such as seafloor mapping produces mid-frequency sounds, but attenuates quicker, therefore impacting a smaller acoustic space [@hildebrand2009; @richardson2013].
+Noise has been shown to propagate differently based on static features like bathymetry and depth [@richardson2013; @vagle2021], as well as dynamic variables like temperature, salinity, and pH from a changing climate [@kunc2016; @affatati2022].
 
 As soniferous animals, cetaceans have varying levels of sensitivity and vulnerability to certain types of noise.
-Mysticetes are more affected by low-frequency noise produced by large shipping vessels.
-Odontocetes produce and hear mid- to high-frequency sounds, so human activities that produce sounds at these frequencies interfere with these signals.
-The lines are not always so clear cut as odontocetes may be affected by the frequencies of motorized shipping typically known to affect baleen whales @aguilarsoto2006.
+Generally, mysticetes are more affected by low-frequency noise produced by large shipping vessels [@southall2017].
+Odontocetes produce and detect mid- to high-frequency sounds and are more likely to interfere with human activities at these frequencies.
+These generalizations appear to have exceptions, as odontocete species like beaked whales have shown avoidance behaviors in response to broadband and low frequency vessel noise [@aguilarsoto2006; @pirotta2012].
+It remains uncertain how short-term effects of chronic or acute noise pollution for cetaceans translates into long-term or population level changes.
+@new2013 suggest enough exposure can cause beaked whales to cease foraging and eventually lower reproduction rates.
 
-While many studies document negative responses of cetaceans to noise pollution with general frequency bands (cite), the circumstances surrounding these acoustic disturbances (chronic or impulsive, low, mid, or high frequency, shallow or deep water, overlap with anthropogenic activity) and their long-term effects are unclear.
+### Beaked Whales and Sperm Whales
 
-habitat degradation
+Beaked whales (family *Ziphiidae)* are a group of 24 species and among the least understood cetaceans and deepest divers in the ocean.
+They regularly dive for over an hour to depths greater than 800 meters, sometimes exceeding three hours and 2000 meters, followed by only a few minutes on the surface [@tyack2006; @baird2006].
+Their preference for deep water, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them with visual observations [@awbery2022].
+Much of what is known comes from strandings and stomach content analysis [@west2017].
+Even when sighting events occur, they risk misidentification due to the paucity of data of morphological variations in some species [@aguilardesoto2017].
+Tagging studies prove useful, but it is often difficult to acquire a sizable dataset [@baird2006; @tyack2006; @baird2008].
 
-### Beaked Whales/Sperm Whales
+Beaked whales are known to be exceptionally sensitive to noise [@hooker2019].
+There are several documented cases of naval mid-frequency active sonar (MFAS) activities coinciding with mass stranding events by beaked whales [@fernández2005; @weilgart2007; @deruiter2013; @bernaldodequiros2019; @simonis2020; @stanistreet2022].
+The same physiological adaptations that allow for extreme diving are observed to increase their decompression risk when they alter their behavior in response to noise [@kvadsheim2012; @fahlman2014].
+Furthermore, these responses vary based on the individuals' age, sex, past experience with noise, or current activity [@weilgart2007].
 
-[Beaked whales]{.underline} consist of 24 species within the family *Ziphiidae* and are among the least understood cetaceans and deepest divers in the ocean.
-They regularly dive for over an hour to depths greater than 800 m, and have been known to dive over three hours at a time to depths greater than 2000 m, followed by only a few minutes on the surface @tyack @baird2006.
-Their preference for deep water, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them with visual observations @awbery2022 .
-Much of what is known comes from strandings and stomach content analysis @west2017.
-Even when sighting events occur, they risk misidentification due to the paucity of data regarding morphological variations in some species @aguilardesoto2017.
-Tagging studies prove useful, but difficult to scale/acquire sizable dataset @baird2006 @baird2008 @tyack .
-
-Beaked whales are also known to be unusually sensitive to noise.
-There are several documented cases of naval mid-frequency active sonar (MFAS) activities coinciding with mass stranding events by beaked whales (see @simonis2020, table 1 in @weilgart, @deruiter2013 , @stanistreet2022, @bernaldodequiros, @fernández2005 ).
-This occurs for two reasons: 1) their anti-predation adaptation to avoid killer whales (*Orcinus orca*)(synchronized foraging, followed by silent, scattered vertical ascent) mimics that of anthropogenic noise and leads to strandings @aguilardesoto2020; 2) the same physiological adaptations that allow for extreme diving are observed to be deadly when forced to alter behavior and flee the noise source more so than other odontocetes @tyack @cox2006 @joyce2020 / can increase decompression sickness risk (Kvadsheim et al. [**2012**](https://esajournals-onlinelibrary-wiley-com.jpllnet.sfsu.edu/doi/full/10.1890/15-0040#eap1216-bib-0036), Fahlman et al. [**2014**](https://esajournals-onlinelibrary-wiley-com.jpllnet.sfsu.edu/doi/full/10.1890/15-0040#eap1216-bib-0024) ).
-Furthermore, these responses vary based on the individuals' age, sex, past experience with noise, or current activity @weilgart.
-
-Despite their visually cryptic nature, their sound signals are very well-described, down to the species level @baumann-pickering2013, @baumann-pickering2014.
-Beaked whale produce clicks, pulses, and buzzes during foraging starting at the 400(?) m range @tyack.
-With increasing acoustic surveys in the literature, new vocalization patterns are being detected globally @manzano-roth2023 and notable temporal patterns are emerging @yamada2019.
+Despite their visually cryptic nature, their sound signals are very well-described, down to the species level [@baumann-pickering2013; @baumann-pickering2014].
+Beaked whales produce clicks, pulses, and buzzes during foraging below 500 meters of depth [@tyack2006].
+Increased acoustic surveys reported in the literature have detected new vocalization patterns globally [@manzano-roth2023], with notable temporal patterns emerging [@yamada2019].
 
 Sperm whales (*Physeter macrocephalus*) also occupy deep ocean habitats, but their dive depths and times are reduced in comparison.
-Sperm whales have been known to dive regularly between 400-600 m and often over 1000 m @watkins1993 @awbery2022 @amano2003.
-They also appear to have more acoustic tolerance than beaked whales @miller2009, @miller2022 @madsen2002 @miller2009 @winsor2017 .
-While avoidance responses in sperm whales have been recorded following naval sonar events @curé2013 @stanistreet2022 @sivle2012 and seismic survey vessels @bowels 1994 range from minor @isojunno2022 @madsen2002 to moderate @isojunno2016 , the circumstances surrounding the noise exposure is not well understood.
+Sperm whales have been known to dive regularly between 400-600 meters and often over 1000 meters [@watkins1993; @amano2003; @awbery2022].
+They also appear to have more acoustic tolerance than beaked whales [@madsen2002; @miller2009; @winsor2017; @miller2022].
+While avoidance responses in sperm whales have been recorded following MFAS events [@sivle2012; @curé2013; @stanistreet2022] and seismic survey pulses [@madsen2002] , and can range from minor [@isojunno2022] to moderate [@isojunno2016] in severity, the circumstances surrounding the noise exposure is not well understood.
 
-Sperm whales also emit series of targeted clicks when socializing (@marcoux2006 @fais2015) and when foraging @fais2016, but it's unlikely they debilitate their prey with their vocalizations as previously hypothesized @norris1972 .
-More is known about sperm whales than beaked whales as they are less visually cryptic and more success being seen on the surface (visual surveys).
+Sperm whales also emit series of targeted clicks when socializing [@marcoux2006; @fais2015] and foraging [@fais2016], but it's unlikely they debilitate their prey with their vocalizations as previously hypothesized [@norris1972].
+More is known about sperm whales than beaked whales as they spend more time on the surface, lending themselves to better detection during visual surveys.
 
-Species studied:\
-- ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
--* BW43, Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed in 2021 @baumann-pickering2013 @barlow2021\
-- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)* \@simonis\
-- BB, -- Baird's beaked whale *B. bairidii* *(IUCN status: least concern)*\
-- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)* @simonis\
-- PM, sperm whale *(IUCN status: vulnerable)*
+Several studies to date have documented specific effects of acoustic interference from human activity in beaked and sperm whales.
+Intense noise events can affect behavior through habitat displacement, time and effort fleeing spent instead of foraging, and communication and echolocation click disruption [@aguilarsoto2006; @deruiter2013; @cholewiak2017; @falcone2017; @joyce2020; @stanistreet2022]; cause physiological harm like hearing loss, tissue damage, and elevated stress levels [@cox2006; @hooker2019] even up to 100 km away [@falcone2017]; and mask signals for communication [@erbe2014], environmental cues [@duarte2021], or mating [@weilgart2007].
 
-**Effects of Acoustic Disturbance**
-
-The specific effects of acoustic interference from human activity in beaked and sperm whales are well-documented.
-Intense noise events can affect behavior through habitat displacement, spending time and effort fleeing instead of foraging, communication and echolocation click disruption, increasing call intensity (cite?) (@deruiter2013, @aguilarsoto2006 , @stanistreet2022, @joyce2020, @falcone2017, @cholewiak2017 ), cause physiological harm like hearing loss, tissue damage, and elevated stress levels (see @hooker2019 and @cox2006) even at far distances (up to 100 km away @falcone2017), mask signals for communication (@erbe2014), environmental cues (@duarte2021), or mating @weilgart.
-
-Under the Marine Mammal Protection Act, all these species are federally protected, several receive addition protection under the Endangered Species Act due to a vulnerable or endangered conservation status @taylor @perrin , but some remain listed as "data deficient" by the International Union for the Conservation of Nature (IUCN) Red List of Threatened Species @pitman @ginkgodens.
-The latter designation has been scrutinized for not illiciting the same conservation urgency as species whose populations are well-documented therefore gets lowered on the priority list (@parsons2016).
+Under the Marine Mammal Protection Act, these species are federally protected, and several receive addition protection under the Endangered Species Act due to a vulnerable or endangered conservation status [@taylor], but some remain listed as "data deficient" by the International Union for the Conservation of Nature (IUCN) Red List of Threatened Species [@pitman; @pitman2020].
+The latter designation has been scrutinized for not eliciting the same conservation urgency as species whose populations are well-documented [@parsons2016], therefore emphasizing the need for population and abundance studies of these species.
 
@@
 Passive acoustic monitoring (PAM) is a tool that allows researchers to record and listen to the soundscape.
-It is also an ideal method for assessing habitat use and improving species distribution models of visually cryptic cetaceans like beaked whales and sperm whales that are more likely to be heard than seen ( @fleishman ), live relatively far offshore ( @robbins2022), and have low-density populations ( @hooker2019, @arranz2023 @hildebrand2015).
+It is also an ideal method for assessing habitat use and improving species distribution models of visually cryptic cetaceans like beaked whales and sperm whales that are more likely to be heard than seen [@fleishman2023], live relatively far offshore [@robbins2022], and have low-density populations [@hildebrand2015; @hooker2019; @arranz2023].
 
 Passive acoustic recorders come in two forms: fixed or mobile, each better suited to a certain ecological application.
-A mobile (towed or free-drifting) buoy moves either behind a vessel or autonomously with the currents for days or weeks at a time.
+A mobile (towed or free drifting) buoy is either towed behind a vessel or moves autonomously with the currents for days or weeks at a time.
 It captures the dynamic nature of sound over a large spatial area, especially in deep and less studied parts of the ocean, making it ideal for detecting offshore or mobile species.
-Bottom-mounted recorders allow for greater temporal coverage and are often used when studying specific regions or known habitats of study species ( @nosal2007, @marques2012 @jarvis2022).
+Bottom-mounted recorders allow for greater temporal coverage and are often used when studying specific regions or known habitats of study species [@nosal2007; @marques2012; @jarvis2022].
 
-The benefits of using PAM over visual surveys include use during night hours or through inclement weather, cost to charter and @johnson2009 @fleishman.
-The trade-offs for using PAM include costly equipment or deployment, data/battery storage requirements, uncertain chance of return, limited spatial coverage for fixed recorders, and limited temporal coverage for mobile recorders.
-Even with experienced acousticians, noisy detections can obfuscate number of individuals, species detected, or distance or direction from recorder for species with large call ranges such as sperm whales @barlow2005 @mellinger2007.
+The benefits of using PAM over visual surveys include use during night hours or through inclement weather, lower cost to charter and little experience needed to conduct a survey [@johnson2009; @fleishman2023].
+The trade-offs include costly equipment or deployment, high data/battery storage requirements, uncertain chance of return, limited spatial coverage for fixed recorders, and limited temporal coverage for mobile recorders.
+Even with experienced acousticians, noisy detections can obfuscate number of individuals or species detected and distance or direction from recorder for species with large call ranges such as sperm whales [@barlow2005; @mellinger2007].
 
-### **AIS**
+### Automatic Identification System
 
-Automatic Identification System (AIS) is a mandatory vessel tracking system for vessels over 300 gross tons to aid in position communication and safe navigation @uscg .
-While not originally intended for use in research, it has become a valuable resource for vessel traffic data, especially in studying effects of vessel strikes @greig2020 @reimer2016 and noise pollution @erbe2014 @aguilarsoto2006 on cetaceans.
-Because ships are abundant throughout the study area due to major US ports, AIS was included in this study to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.
-
-Previous studies have separated vessel and anthropogenic noise from ambient soundscape based on known spectral signatures, frequency bands, and percentiles @weiss2021 @haver2020.
+Automatic Identification System (AIS) is a mandatory vessel tracking system for vessels over 300 gross tons to aid in position communication and safe navigation [@federalregister2003].
+While not originally intended for use in research, it has become a valuable resource for vessel traffic data, especially in studying effects of vessel strikes [@reimer2016; @greig2020] and noise pollution [@aguilarsoto2006; @erbe2014] on cetaceans.
+Because ships are abundant throughout the study area due to major U.S.
+ports, AIS was included in this study to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.
+Previous studies have separated vessel and anthropogenic noise from ambient soundscape based on known spectral signatures, frequency bands, and percentiles [@haver2020; @weiss2021].
 The inclusion of AIS to soundscape metrics in this study provides fine-scale evaluation of sound and source.
 
-Power spectral density ambient sound levels (see @haver2020 for details on what percentiles mean)
+### **Laws and Regulations**
 
-**Laws and Regulations**
-
-While there is general consensus that anthropogenic noise in the ocean is increasing, policies that aim to mitigate chronic noise are generally lacking.
-This is due to knowledge gaps regarding the long-term and cumulative effects of noise on certain species @markus2018.
-
-International and regional efforts are underway (see @chou2021) , but still have a ways to go.
-Within the European Union, the Marine Strategy Framework Directive establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise @directiv2008, @borsanij.f.2023.
-Some studies have independently tested @haver2021 @garrett2016 @sebastianutto2015 or used @mustonen2019 these metrics.
-The National Oceanic and Atmospheric Administration (NOAA) has developed an Ocean Noise Strategy Roadmap, an agency-wise approach to mitigating both chronic and acute effects of noise exposure to animals @noaa2018.
-Within it, they agree upon impairment thresholds for mid-frequency cetaceans (@noaa2018) .
+While there is consensus that anthropogenic noise in the ocean is increasing, policies that aim to mitigate chronic noise are generally lacking.
+This is due to knowledge gaps regarding the long-term and cumulative effects of noise on certain species [@markus2018].
+International and regional efforts are underway (see @chou2021) but are not fully complete.
+Within the European Union, the Marine Strategy Framework Directive establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise [@directiv2008; @borsanij.f.2023].
+Some studies have independently tested these threshold values [@sebastianutto2015; @garrett2016; @haver2021] or used them as guidelines in their study [@mustonen2019].
+The National Oceanic and Atmospheric Administration (NOAA) has developed an Ocean Noise Strategy Roadmap, an agency-wise approach to mitigating both chronic and acute effects of noise exposure to animals [@noaa2018].
+Within it, the authors agree upon impairment thresholds for mid-frequency cetaceans [@noaa2018].
 Because chronic effects are not fully understood, they remain to be regulated.
-The International Maritime Organization (IMO) has also been addressing noise pollution from commercial vessels through quieter ship design @internationalmaritimeorganization.
+The International Maritime Organization (IMO) has also been addressing noise pollution from commercial vessels through quieter ship design [@internationalmaritimeorganization2014].
 Updates to their 2014 guidelines are set to be finalized by January 2024.
-Canada is also developing an Ocean Noise Strategy @breeze2022.
+Finally, Canada has launched a Quiet Vessel Initiative and is expected to publish a Ocean Noise Strategy for underwater noise management in 2023 [@breeze2022].
 
@@
 With anthropogenic noise-generating activities on the rise, global oceanic sound levels are expected to increase.
-Noise is a form of habitat degradation, so there is a need for research on a large spatial scale to understand how these highly mobile animals overlap in space and time with human activities and how they're affected @haver2019 @hooker2019 @erbe2014.
-Passive acoustic monitoring Acquiring information on these federally protected and acoustically-sensitive species is challenging\
-\
+Noise is a form of habitat degradation, so there is a need for research on a large spatial scale to understand how these highly mobile animals overlap in space and time with human activities and how they're affected [@erbe2014; @haver2019; @hooker2019].
+Acquiring information on these federally protected and acoustically sensitive species is challenging, but certain PAM techniques can be well-suited for this type of research.
 Logistical difficulties associated with offshore research and low-density populations of visually cryptic deep diving whales have created a paucity of data in the California Current Ecosystem.
@@
 
-The purpose of this study is to characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics, environmental variables, and AIS.
-We determine significant predictors of whale presence for each species and provide explicit sources for noise-generating human activities through overlay of large vessel activity.
-
-This study highlights spatial and temporal acoustic variability within the study region. It will also contribute to noise exposure levels through acoustic detections.
+The purpose of this study is to characterize the spatial and temporal acoustic variability of beaked and sperm whale habitat in the California Current Ecosystem using soundscape metrics, environmental variables, and AIS.
+We determine significant predictors of whale presence for each species with the use of Random Forest models and generalized additive models and provide explicit sources for noise-generating human activities through overlay of large vessel activity.
 
@@
 
-The California Current Ecosystem (CCE) is a highly productive and complex ecosystem that contains one proposed and five current national marine sanctuaries @sanctuaries , supports several wild fisheries @field2006 , and is home to nineteen protected whale species @fisheries.
-The 1,141,800 km2 area extends along the US' Economic Exclusion Zone from south of Vancouver to north of Baja California and is a characterized by deep canyons, continental shelf @barlow2009 @barlow2007.
-Characteristic of the area are upwelling events that carry nutrients to the surface often drive primary productivity, creating an ideal habitat where whales' prey are found and driving them there @ryther1969.
-The year 2018 was characterized in the CCE with slightly warmer than average sea surface temperatures as the recent El-Nino and marine heatwave events were ending @harvey.
-Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area @peña2007 @coleman2008 @oldach2022.
-\
+The California Current Ecosystem (CCE) is a highly productive and complex ecosystem that contains one proposed and five current national marine sanctuaries [@officeofnationalmarinesanctuaries], supports several wild fisheries [@field2006], and is home to nineteen protected whale species [@noaafisheries].
+The 1,141,800 km2 area extends along the US' Economic Exclusion Zone from southern Canada to northern Mexico and is a characterized by deep canyons and a continental shelf [@barlow2007; @barlow2009].
+Characteristic of the area are upwelling events that carry nutrients to the surface and drive primary productivity, creating an ideal habitat where whales' prey are found [@ryther1969].
+The year 2018 was characterized in the CCE with slightly warmer than average sea surface temperatures as the recent El-Nino and marine heatwave events were ending [@harvey].
+Because this area has high economic value and natural ecological activity, it's a well-managed and well-researched area [@peña2007; @coleman2008; @oldach2022].\
 \
 In 2021, California, Washington, and Oregon ranked 3rd, 5th, and 23rd in most trafficked U.S.
-waterways, respectively (@unitedstatesarmycorpsofengineersusace2018.
+waterways, respectively [@unitedstatesarmycorpsofengineersusace2018].
 Between vessel activity like fishing, shipping and naval exercises in California alone, there is notable overlap of anthropogenic presence with important habitat for protected species.
-Furthermore, the future development of offshore wind turbines off of Morro Bay and Humboldt, California are expected to be noise-producing activities during construction and operation, increasing sound levels in those regions @cooperman @madsen2006.
+Furthermore, the future development of offshore wind turbines off of Morro Bay and Humboldt, California are expected to be noise-producing activities during construction and operation, increasing sound levels in those regions [@madsen2006; @cooperman2022].
 
@@
 
-**Overall data collection efforts**\
 Acoustic detections were collected as part of the 2018 California Current Ecosystem Survey (CCES) by the Southwest Fisheries Science Center (SWFSC).
-Twenty-three free-drifting buoys (numbered 1-23) were deployed at varying times between July and November in the offshore waters of the study area.
-These buoys, numbered 1-23, contained two satellite geo-locators above water to monitor GPS information and two hydrophones at depth attached by a 150m line.
+Twenty-three free drifting buoys (numbered 1-23) were deployed at varying times between July and November in the offshore waters of the study area.
+These buoys contained two satellite geo-locators above water to monitor GPS information and two hydrophones at depth attached by a 150-meter line.
 They collected a total of 1910 hours of data.
-SWFSC detected and classified acoustic detections in PAMGuard to species level as per methods mentioned in @simonis2020 .
+SWFSC detected and classified acoustic detections in PAMGuard to species level as per methods mentioned in @simonis2020a.
 
@@
 The SWFSC team analyzed the data and detected echolocation pulses produced by beaked and sperm whales.
-These were classified by peak frequency and spectral signature and identified to the species level (see @simonis2020 for details on this survey and @rand2022 for duty-cycle study).
+These were classified by peak frequency and spectral signature and identified to the species level (see @simonis2020a for details on this survey and @rand2022 for duty-cycle study).
 Soundscape metrics were binned by peak frequency into low, medium, high frequency bands, along with the 1, 5, 10, 25, 75, 95, and 99 percentile distribution from each metric.
 
-Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to the UTC time zone.
-These metrics include third-octave level (TOL) bands from 63 to 20,000 Hz and broadband (20-24,000 Hz) metrics.
-Several TOL levels were chosen as proxies to vessel or sonar frequencies.
+~~Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to the UTC time zone.
+These metrics include third-octave level (TOL) bands from 63 to 20,000 Hz and broadband (20-24,000 Hz) metrics.~~ Several TOL levels were chosen as proxies to vessel or sonar frequencies.
 
-Whale detection timestamps were rounded to nearest 20 minutes based on start time to align with the soundscape metrics.
-A separate column was created as a common join field that contained nearest timestamp in UTC to that of buoy tracks (see below).
-Two types of beaked whale detections (species code ?BW and ?BWC) were removed from analysis as the species was unknown.
+~~Whale detection timestamps were rounded to nearest 20 minutes based on start time to align with the soundscape metrics.
+A separate column was created as a common join field that contained nearest timestamp in UTC to that of buoy tracks (see below).~~ Two types of beaked whale detections (species code *?BW* and *?BWC*) were removed from analysis as the species was unknown.
 
-Buoy tracks were similarly rounded to nearest 20 minutes.
-Seven buoys (1,2,3,5,6,9,11) were lost at sea and two buoys had corrupted data (4,17) so these were removed from the analysis.
-Buoy 14 and 15 were the same and therefore consolidated to 14.
+**\[New addition as of 08/25/23\] \[\[**Two SPOT buoys sit on every buoy, so some GPS points were recorded twice.
+These duplicate tracks were eliminated.\]\] ~~Buoy tracks were similarly rounded to nearest 20 minutes.~~ Seven buoys (1,2,3,5,6,9,11) were lost at sea and two buoys had corrupted data (4,17) so these were removed from the analysis.
+Buoys 14 and 15 were the same and therefore consolidated to 14.
 These three preliminary datasets were joined together based on a common buoy number (column "station") and time field column "UTC").
-Whale detections (column Wpresence) were reduced to binary data for absence/presence and all other fields of that dataset were eliminated.
+Whale detections (column "Wpresence") were reduced to binary data for absence/presence and all other fields of that dataset were eliminated.
 
-**Environmental Data**\
+### **Environmental Data**
+
 Possible environmental and oceanographic predictors of cetacean habitat were extracted from the NOAA Environmental Research Division Data Access Program (ERDDAP) server ((https://coastwatch.pfeg.noaa.gov/erddap/) and Hybrid Coordinate Ocean Model (HYCOM) using the *matchEnvData* function in R package *PAMmisc*.
@@
 
-\[Can show table of variables and justification by papers (Virgili et al. 2022)\]
+\[See Virgili et al 2019 for extracting slope from GEBCO Data and dynamic variables!!\]
 
-@forney2012 shows how BW model had a lot of uncertainty regarding distribution predictors. So few detections that some spp had to be lumped together (also @barlow2009).
+### **Modeling**
 
-| Covariate                     | Unit | Source                                                          | Justification                |
-|----------------|----------------|----------------|------------------------|
-| Salinity at 400m              |      |                                                                 |                              |
-| Distance to continental slope |      | General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015 | @barlow2009                  |
-| Bathymetric slope             |      | GEBCO tif                                                       | @becker2010                  |
-| Depth                         |      |                                                                 | @forney2012 @becker2010      |
-| Wind curl stress              |      | ERDDAP                                                          | @mannocci2015                |
-| MLD depth                     |      |                                                                 | @mccullough2021, @forney2012 |
-| Thermocline temp              |      |                                                                 | @mccullough2021              |
-| Thermocline depth             |      |                                                                 |                              |
-| MLD Temp                      |      |                                                                 | @mccullough2021              |
-| Temperature at 400m           |      |                                                                 |                              |
-| SST mean                      |      |                                                                 | Becker 2007, @forney2012     |
-| CHL                           |      |                                                                 | @forney2012                  |
-| SSH                           |      |                                                                 |                              |
+We used a Random Forest (RF) model in R (R Core Team, 2023) using the package () to identify most important environmental or acoustic predictors for beaked and sperm whales.
+A separate model was run for each species.
 
-**\[AIS\]**\
-Automatic Identification System data were collected in order to evaluate the effect of spatial proximity to the buoys.
-This step is important in understanding how ship characteristics contribute to various frequency bands.
-I filtered out any vessels under 100 meters in length in order to focus only on large vessels.
-By exploring buoy/vessel co-occurrence with a 10-km "listening space" around the buoy (McKenna) can allow us to investigate patterns of when whales are calling and the vessel activity during co-occurrence.
-
-AIS was downloaded from Marine Cadastre, processed, converted to UTC timezone, and matched to each buoy's timestamp using the package *PAMscapes* and function \_\_.
-
-**\[Big analysis: statistics, modeling details\]**
-
-Use Random Forest model to identify most important predictors.
-
-Can use non-parametric Kruskal-Wallis one-way analysis of variance test to examine *if number of whale detections per hour differed depending on presence/absence of vessels* (see Trickey et al. 2022).
-
-[Example of GAM]{.underline}: @mannocci2015, Becket et al 2020, Becker 2007
-
-AIC explanation @symonds2011
+We built generalized additive models in R using the package *mcgv* to relate beaked and sperm whale detections to an environmental or acoustic variable.
 
@@
 [Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day ( @ziegenhorn2023 ) , drift with most detections/day), where/when were most detections heard, number of species]
+
+-   Buoy 18 had the most whale detections
+
+-   ZC was the most frequently detected whale
+
+-   The loudest sound levels were only occupied by PM
+
+-   Density of ship encounters
+
+-   Report on environmental conditions for ZC/PM?
+
+-   Report on analysis of env variables WITH and WITHOUT soundscape metrics (see @fiedler2023, results 3.1)
 
@@
 
+AIC explanation @symonds2011
+
 ## Discussion
+
+\[Laws and Regulations\]
+
+While there is consensus that anthropogenic noise in the ocean is increasing, policies that aim to mitigate chronic noise are generally lacking.
+This is due to knowledge gaps regarding the long-term and cumulative effects of noise on certain species (Markus and Sánchez 2018).
+International and regional efforts are underway (see Chou et al.
+(2021)) but are not fully complete.
+Within the European Union, the Marine Strategy Framework Directive establishes a directive about environmental regulation on marine noise, specifically requiring member states to adopt strategies in compliance with threshold values for impulsive and continuous noise ("Directive 2008/56/EC of the European Parliament and of the Council of 17 June 2008 Establishing a Framework for Community Action in the Field of Marine Environmental Policy (Marine Strategy Framework Directive) (Text with EEA Relevance)" 2008; Borsani, J.F.
+et al.
+2023).
+Some studies have independently tested these threshold values (Sebastianutto et al.
+2015; Garrett et al.
+2016; Haver et al.
+2021) or used them as guidelines in their study (Mustonen et al.
+2019).
+The National Oceanic and Atmospheric Administration (NOAA) has developed an Ocean Noise Strategy Roadmap, an agency-wise approach to mitigating both chronic and acute effects of noise exposure to animals (NOAA 2018).
+Within it, the authors agree upon impairment thresholds for mid-frequency cetaceans (NOAA 2018).
+Because chronic effects are not fully understood, they remain to be regulated.
+The International Maritime Organization (IMO) has also been addressing noise pollution from commercial vessels through quieter ship design (International Maritime Organization 2014).
+Updates to their 2014 guidelines are set to be finalized by January 2024.
+Finally, Canada has launched a Quiet Vessel Initiative and is expected to publish a Ocean Noise Strategy for underwater noise management in 2023 (Breeze et al.
+2022).
 
@@
 
+The year 2018 was characterized in the CCE with slightly warmer than average sea surface temperatures as the recent El-Nino and marine heatwave events were ending (Harvey et al. 2019).
+
+@croll2001 found that whale presence more closely related to prey abundance than sound levels
+
+Can use non-parametric Kruskal-Wallis one-way analysis of variance test to examine *if number of whale detections per hour differed depending on presence/absence of vessels* (see Trickey et al. 2022).
+
 **\[Recommendations?\]**
@@
 \[Add here.\]
+
+### Tables
+
+1\) Species studied:\
+- ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
+-* BW43, Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed in 2021 @baumann-pickering2013 @barlow2021\
+- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)* \@simonis\
+- BB, -- Baird's beaked whale *B. bairidii* *(IUCN status: least concern)*\
+- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)* @simonis\
+- PM, sperm whale *(IUCN status: vulnerable)*
+
+2\) Environmental covariates (group by physiographic, oceanographic, etc)
+
+| Covariate                          | Resolution       | Source                                                                                                                                            | Justification                           |
+|------------------------------------|------------------|---------------------------------------------------------------------------------------------------------------------------------------------------|-----------------------------------------|
+|                                    |                  |                                                                                                                                                   |                                         |
+| Distance to shore (m)              | 0.01-Degree Grid | ERDAPP                                                                                                                                            |                                         |
+| Distance to continental slope (m)  | 15 arc sec       | Derived from General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015                                                                      | @barlow2009                             |
+| Bathymetric slope (°)              | 15 arc sec       | Derived from GEBCO tif                                                                                                                            | @becker2010 @virgili2022                |
+| Depth (m)                          | 15 arc sec       | ERDAPP                                                                                                                                            | @forney2012 @becker2010 @virgili2022    |
+| Wind curl stress                   | 1°, 6-hourly     | ERDDAP                                                                                                                                            | @mannocci2015                           |
+| Mixed layer depth (m)              | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
+| @mccullough2021, @forney2012 @moore2021 |
+| Thermocline temperature (°C)       | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
+| @mccullough2021                         |
+| Thermocline depth (m)              | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
+|                                         |
+| Mixed layer depth Temperature (°C) | 0.08°, 3-hourly  | Calculated from HYCOM                                                                                                                             | @mccullough2021                         |
+| Salinity at 400m                   | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0                                                                                                                 |                                         |
+| Temperature at 400m (°C)           | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0                                                                                                                 |                                         |
+| SST mean                           | 0.01° , daily    | ERDAPP                                                                                                                                            | Becker 2007, @forney2012 @virgili2022   |
+| CHL                                | 4 km, monthly    | ERDAPP                                                                                                                                            | @forney2012 @virgili2019                |
+| Sea Surface Height Anomalies (m)   | daily            | ERDAPP                                                                                                                                            | @virgili2019                            |
+
+\[Can show table of variables and justification by papers (Virgili et al. 2022 and Virgili et al 2019)\]
+
+@forney2012 shows how BW model had a lot of uncertainty regarding distribution predictors. So few detections that some spp had to be lumped together (also @barlow2009).

### 2023-10-16 interpolation.Rmd
Create interpolation.Rmd
new rmd

---
title: "interpolation"
author: "Marina Bozinovic"
date: "2023-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mgcv)
```

Read in rda
```{r}
# Read in original buoy 20 to compare
# Change name so as not to alter it
env20.orig <- readRDS(file = "D:/Buoy_dfs/GPSwhaleAIS20.rda")

# Interpolated buoy 20
env20.interp <- readRDS(file = "data/GPSwhaleAIS20_alt.rda")
```

Download and attach environmental variables to interpolated pts for buoy 20
```{r}
# Sea Surface Temp
env20.interp <- matchEnvData(env20.interp, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST20_interp", var = c('analysed_sst')) # Download SST
env20.interp <- matchEnvData(env20.interp, nc = ss) %>%
  rename(., sst_mean = analysed_sst_mean) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)  # Match SST

# Wind curl
env20.interp <- matchEnvData(env20.interp, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress20_interp.nc", var = c('curl')) # Download wind curl
env20.interp <- matchEnvData(env20.interp, nc = w) %>%
     dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)   # Match wind curl

# Distance to continental slope
bathy <- rast("D:/CCES_gebco_bathy.tif") # Bathymetry for all of CCE from https://download.gebco.net/
agg <- terra::aggregate(bathy, 5)  # Increase cell size by factor of 5 to be visible
cont <- st_as_sf(as.contour(bathy, )) # Create contour lines from bathymetry 
cont_slope <- cont %>% filter(level == -2000) # Establish continental shelf. Is 2000m the correct depth for continental shelf?
sf <- st_as_sf(env20.interp, coords = c("Longitude", "Latitude"), crs=4326, remove=F)
trk <- terra::rasterize(vect(sf), agg)
slop <- terra::distance(trk, vect(cont_slope))
ext <- terra::extract(slop, vect(sf))

dbind <- cbind(sf, ext) %>%
  dplyr::select(-ID) %>%
  rename("dist2slope" = last) %>%
  st_set_geometry(NULL) # remove geometry to prevent buoy dataframe from becoming sf object for future

# Depth
env20.interp <- matchEnvData(env20.interp, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth20_alt.nc", var = c('z')) #Download depth
env20.interp <- matchEnvData(env20.interp, nc = dd) %>%
  rename(., depth = z_mean) %>%        # rename new column to "depth"
  mutate(depth = depth*-1) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean) #Match depth

# Bathymetric slope
bathy_slope <- terra::terrain(bathy, v = "slope")
slope <- terra::extract(bathy_slope, env20.interp[,3:2]) %>%  # Reads latitude and longitude of buoy dataframe
  dplyr::select(-ID) %>%
  rename("bathy_slope" = slope)
sbind <- cbind(env20.interp, slope)

# Distance to shore
env20.interp <- matchEnvData(env20.interp, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore20_interp.nc", var = c('dist')) #Download bathy slope
env20.interp <- matchEnvData(env20.interp, nc = d2) %>%
  rename(., dist2shore = dist_mean) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean) # Match bathy slope

# Chlorophyll
env20.interp <- matchEnvData(env20.interp, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll20_interp.nc", var = c('chlorophyll')) #Download chl
env20.interp <- matchEnvData(env20.interp, nc = ch) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean) #Match chl

# Sea Surface Height
env20.interp <- matchEnvData(env20.interp, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh20_interp.nc", var = c('sla')) #Download SSH
env20.interp <- matchEnvData(env20.interp, nc = sh) %>%
  rename(., ssh_mean = sla_mean) %>%
  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean) # Match SSH

# Deep ocean variables
# Assumes themData function has run and exists in global environment
Htest <- matchEnvData(env20.interp, nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)

thermData <- bind_rows(lapply(Htest, function(x) {
    # do thermo calcs
    result <- calcThermoVars(x$matchDepth, x$water_temp)
    # get specific values
    result$temp400 <- x$water_temp[x$matchDepth == 400]
    result$sal400 <- x$salinity[x$matchDepth == 400]
    result
}))

env20.interp <- cbind(env20.interp, thermData)

```

Save
```{r}
# To project
saveRDS(env20.interp, 'data/GPSwhaleAISenv20_interp.rda')

# To external hard drive
saveRDS(env20.interp, 'D:/Buoy_dfs/GPSwhaleAISenv20_interp.rda')
```

What is GAM of Buoy 20 without interpolation (original)
```{r}
trk20_noInterp <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + ttDepth + sal400
                      + TOL_125 + TOL_2000,
              data = env20.orig, family = binomial, method = "REML")

plot(trk20_noInterp, trans = plogis, all.terms = TRUE)
summary(trk20_noInterp)
# Env varb only: Deviance explained = 8.17%, R-sq = 0.051
# Env + SS variables: Deviance explained = 8.9%, R-sq = 0.0532
```

What is GAM of Buoy 20 with interpolation?
```{r}  
trk20_withInterp <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + ttDepth + sal400
                        + TOL_125 + TOL_2000,
              data = env20.interp, family = binomial, method = "REML")

plot(trk20_withInterp, trans = plogis, all.terms = TRUE)
summary(trk20_withInterp)
# Env varb only: Deviance explained = 7.91%, R-sq = 0.0634
# Env + SS variables: Deviance explained = 8.92%, R-sq = 0.0703
```

Summary Stats of Buoy 20 with/without interpolation
```{r}
# DF with removed questionable rows
#Before gap
#b4_mean <- sapply(env20.orig[830:849, 6:21], mean)
#b4_mean <- cbind(b4_mean, rownames(b4_mean))

#b4_sd <- sapply(env20.orig[830:849,6:21], sd)
#b4_sd <- cbind(b4_sd, rownames(b4_sd))

#b4_min <- sapply(env20.orig[830:849, 6:21], min)
#b4_min <- cbind(b4_min, rownames(b4_min))

#b4_max <- sapply(env20.orig[830:849, 6:21], max)
#b4_max <- cbind(b4_max, rownames(b4_max))
```

```{r}
#After gap
#aft_mean <- sapply(env20.orig[871:890, 6:21], mean)
#aft_mean <- cbind(aft_mean, rownames(aft_mean))

#aft_sd <- sapply(env20.orig[871:890, 6:21], sd)
#aft_sd <- cbind(aft_sd, rownames(aft_sd))

#aft_min <- sapply(env20.orig[871:890, 6:21], min)
#aft_min <- cbind(aft_min, rownames(aft_min))

#aft_max <- sapply(env20.orig[871:890, 6:21], max)
#aft_max <- cbind(aft_max, rownames(aft_max))
```

```{r}
b4af_mean <- sapply(env20.orig[840:860, 6:21], mean)
b4af_mean <- cbind(b4af_mean, rownames(b4af_mean))

b4af_sd <- sapply(env20.orig[840:860, 6:21], sd)
b4af_sd <- cbind(b4af_sd, rownames(b4af_sd))

b4af_min <- sapply(env20.orig[840:860, 6:21], min)
b4af_min <- cbind(b4af_min, rownames(b4af_min))

b4af_max <- sapply(env20.orig[840:860, 6:21], max)
b4af_max <- cbind(b4af_max, rownames(b4af_max))
```

```{r}
# DF of interpolated section
interp_mean <- sapply(env20.interp[850:870, 6:21], mean)
interp_mean <- cbind(interp_mean, rownames(interp_mean))

interp_sd <- sapply(env20.interp[850:870, 6:21], sd)
interp_sd <- cbind(interp_sd, rownames(interp_sd))

interp_min <- sapply(env20.interp[850:870, 6:21], min)
interp_min <- cbind(interp_min, rownames(interp_min))

interp_max <- sapply(env20.interp[850:870, 6:21], max)
interp_max <- cbind(interp_max, rownames(interp_max))
```

Create df for comparing before/after with interpolated points
```{r fig.width=10}
env20.orig <- env20.orig %>% mutate(dist2slopekm = dist2slope/1000,
                                  depthKm = depth/1000)

env20.interp <- env20.interp %>% mutate(dist2slopekm = dist2slope/1000,
                                  depthKm = depth/1000)

b4af_sample <- env20.orig[840:860, 6:24] %>%
  add_column(type = "beforeAfter")

interp_sample <- env20.interp[850:870, 6:24] %>%
  add_column(type = "interpolated")

compare_pts <- rbind(b4af_sample,interp_sample)

compare_longer1 <- pivot_longer(compare_pts, cols = c('sst_mean', 'mldTemp','mldTemp', 'ttTemp','temp400'),
             names_to = "variable",
             values_to = "values")

compare_longer2 <- pivot_longer(compare_pts, cols = c('TOL_125', 'TOL_2000', 'dist2slopekm'),
             names_to = "variable",
             values_to = "values")

compare_longer3 <- pivot_longer(compare_pts, cols = c('chlorophyll_mean', 'curl_mean', 'ssh_mean', 'depthKm'),
             names_to = "variable",
             values_to = "values")

compare_longer4 <- pivot_longer(compare_pts, cols = c('sal400', 'bathy_slope','ttDepth'),
             names_to = "variable",
             values_to = "values")

compare_longer5 <- pivot_longer(compare_pts, cols = c('mldDepth', 'ttDepth', 'depthKm'),
                                names_to = "variable",
                                values_to = "values")

depthKmplot <- pivot_longer(compare_pts, cols = 'depthKm',
                                names_to = "variable",
                                values_to = "values")

mldDepthplot <- pivot_longer(compare_pts, cols = 'mldDepth',
                                names_to = "variable",
                                values_to = "values")

ttDepthplot <- pivot_longer(compare_pts, cols = 'ttDepth',
                                names_to = "variable",
                                values_to = "values")
# Missing dist2shore because value does not change much
```

Plot
```{r fig.width=8}
ggplot(compare_longer1, aes(x = type, y = values, color = type)) + geom_boxplot() + 
  facet_wrap(~variable)
ggplot(compare_longer2, aes(x = type, y = values, color = type)) + geom_boxplot() + 
  facet_wrap(~variable)
ggplot(compare_longer3, aes(x = type, y = values, color = type)) + geom_boxplot() + 
  facet_wrap(~variable)
ggplot(compare_longer4, aes(x = type, y = values, color = type)) + geom_boxplot() + 
  facet_wrap(~variable)
#Depths only
ggplot(compare_longer5, aes(x = type, y = values, color = type)) + geom_boxplot() + 
  facet_wrap(~variable)

ggplot(depthKmplot, aes(x = type, y = values, color = type)) + geom_boxplot() +
  labs(title = "depth in Km")
ggplot(mldDepthplot, aes(x = type, y = values, color = type)) + geom_boxplot() +
  labs(title = "mldDepth")
ggplot(ttDepthplot, aes(x = type, y = values, color = type)) + geom_boxplot() +
  labs(title = "ttDepth")

```

Bind results & plot
```{r}
interp_results <- cbind(interp_mean, b4af_mean, interp_sd, b4af_sd, 
                        interp_min, b4af_min, interp_max, b4af_max) %>%
  format(scientific = F, digits = 1) %>%
  as.data.frame()

mean(as.numeric(interp_results$interp_sd))
mean(as.numeric(interp_results$b4af_sd))

#ggplot(interp_results) + geom_boxplot()

write.csv(interp_results, file = "interp_stats_drift20.csv")
```

### 2023-10-16 data-raw/whales.R
Update whales.R

@@ -16,8 +16,63 @@ whales <- EventInfo %>%
 ## Change Station 15 to Station 14 here. Drift 15 and 14 are the same (mix-up during fieldwork)
 whales$Deployment[whales$Deployment == 15] <- 14
 
-# Save object
-saveRDS(whales, 'data/whales.rda')
+
+# Filter by drifts and assign WHALE to new variable (creates whale7, whale8,...)
+g <- unique(whales$Deployment)
+
+for (whale in g) {
+  assign(paste0("whale", whale), whales %>% dplyr::filter(Deployment == whale))
+}
 
 
+############################################################################
+## For interpolation of whale detections to fill in incorrect coordinates ##
+############################################################################
 
+## Interpolation. Need to access track.R first
+approxLong <- approx(x = track_020$UTC, y=track_020$Longitude,
+                     xout = whale20$UTC)$y
+approxLat <- approx(x = track_020$UTC, y=track_020$Latitude,
+                    xout = whale20$UTC)$y
+
+# Create whale20.alt with new points
+whale20.alt <- whale20 %>% mutate(Longitude = approxLong, Latitude = approxLat)
+
+
+# Adding newly interpolated points to track_020:
+columnselect <- whale20.alt %>% select(Latitude, Longitude, UTC) %>% add_column(DriftName = "CCES_020")
+rowselect <- columnselect[51:69,]  #select whale detection rows that are missing from GPS
+
+
+# Create new track_020.alt for buoy 20 interpolation analysis
+# by pasting new points where GPS went dark
+track_020.alt <- rbind(track_020[1:800,],
+                       rowselect,
+                       track_020[801:957,])
+
+##############################################################################
+## Removal of incorrect coordinates where buoy 20 GPS was malfunctioning #####
+##############################################################################
+
+# Bind top part with bottom part of df, removing questionable rows
+whale20 <- rbind(whale20[1:50,], whale20[70:96,])
+# whale20 rows go from 96 to 77
+
+
+#############################################################
+# Save newly created objects
+saveRDS(whale7, 'data/whale7.rda')
+saveRDS(whale8, 'data/whale8.rda')
+saveRDS(whale10, 'data/whale10.rda')
+saveRDS(whale12, 'data/whale12.rda')
[…]

### 2023-10-16 data-raw/tracks.R
Update tracks.R

@@ -3,16 +3,41 @@
 library(tidyverse)
 
 # Read in csv with GPS points
-AllTracks_raw <- read_csv("data-raw/AllCCES_GPS.csv")
+AllTracks_raw <- read_csv("data-raw/AllCCES_GPS.csv",show_col_types = FALSE)
+
+## GPS tracks have 30 minute intervals, but note that there are TWO GPS transmitters that record on their own 30-min intervals.
+## This is why GPS tracks are usually ~2 or ~30 minutes apart.
 
 # Read in tracks, filter out lost buoys (1,2,3,5,6,9,11) and corrupted buoys (4,17)
 AllTracks <- AllTracks_raw %>%
   dplyr::select(Latitude, Longitude, UTC, DriftName) %>%
-  #mutate(dateTime = round_date(.$dateTime, "20 minutes")) %>% # round to nearest 20 minutes, will create duplicated in the data
   filter(!DriftName %in% c('CCES_004','CCES_017'))
 
 AllTracks <- AllTracks[!duplicated(AllTracks[c('Longitude', 'Latitude','UTC', 'DriftName')]),]    # Remove duplicate timestamps
 
+
+# Create new datafram for each buoy
+d <- unique(AllTracks$DriftName)
+
+for (drift in d) {
+  assign(paste0("track", substr(drift, 5,8)), AllTracks %>% 
+           dplyr::filter(DriftName == drift))
+}
+
+
 # Save as RDS
-saveRDS(AllTracks, 'data/AllTracks.rda')
- 
+saveRDS(track_007, 'data/track_007.rda')
+saveRDS(track_008, 'data/track_008.rda')
+saveRDS(track_010, 'data/track_010.rda')
+saveRDS(track_012, 'data/track_012.rda')
+saveRDS(track_013, 'data/track_013.rda')
+saveRDS(track_014, 'data/track_014.rda')
+saveRDS(track_016, 'data/track_016.rda')
+saveRDS(track_018, 'data/track_018.rda')
+saveRDS(track_019, 'data/track_019.rda')
+saveRDS(track_020, 'data/track_020.rda')
+saveRDS(track_021, 'data/track_021.rda')
+saveRDS(track_022, 'data/track_022.rda')
+saveRDS(track_023, 'data/track_023.rda')
+
+

### 2023-10-16 data-raw/soundscape_metrics.R
Delete soundscape_metrics.R

@@ -1,52 +1,0 @@
-## code to prepare `soundscape` dataset goes here
-
-library(tidyverse)
-library(lubridate)
-library(here)
-
-# Path to folder in Project/on GitHub
-#sslist <- list.files(path = paste0("data-raw/"), pattern = "CCES_", recursive = TRUE)
-#### Might not need this - it links to a few raw data files in GitHub ###############
-
-# Read through a few soundscape files on GitHub, change date format
-#for (i in sslist) {
-#  data <- read_csv(here(paste0("data-raw/", i)), show_col_types = FALSE) %>%
-#    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
-#    mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
-#    mutate(variable = paste0(substr(i, 16, 17), substr(i, 6, 7)))
-  
-#  assign(paste0(substr(i, 16, 17), "_", substr(i, 6, 7)), data)
-#}
-
-# Path to folder on my external hard drive
-dirlist <- dir("D:/Soundscape_metrics", recursive=TRUE, full.names=TRUE, pattern="CCES_")
-
-# Read though files from external hard drive, change date format
-for (i in dirlist) {
-  data <- read_csv(i, show_col_types = FALSE) %>%
-    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
-    mutate(dateTime = round_date(ymd_hms(dateTime), "20 minutes")) %>%
-    mutate(variable = paste0(substr(i, 38, 39), substr(i,28, 29)))
-  
-  saveRDS(data, file = paste0("data/", substr(i, 38, 39), "_", substr(i, 28, 29), ".rda"))
-
-}
-# Creates BB_07, TO_07, BB_08, TO_08, etc. RDA files for each buoy
-
-
-####### Read in Percentiles ###############
-
-# Path to folder on my external hard drive
-pclist <- dir("D:/Soundscape_metrics/10_90_percentiles", 
-               recursive=TRUE, full.names=TRUE, pattern="CCES_")
-
-
-#Read through percentiles in external hard drive, change date format
-for (i in pclist) {
-  data <- read_csv(i, show_col_types = FALSE) %>%
-    rename(dateTime = `yyyy-mm-ddTHH:MM:SSZ`) %>%
-    mutate(UTC = as.POSIXct(dateTime))
-  
[…]

### 2023-10-16 data-raw/04_soundscape_metrics.R
Create 04_soundscape_metrics.R

@@ -1,0 +1,150 @@
+## code to prepare `soundscape` dataset goes here
+## adding column for whale detections here
+## percentiles are at the bottom
+
+## Required Packages
+library(tidyverse)
+library(PAMpal)
+
+# Set system time zone to UTC
+Sys.setenv(TZ='UTC')
+
+#### Read in GPSwhaleAISenv dataframes for each buoy from external drive.
+#Create list of objects
+# Be sure to clear global environment of objects like GPSwhaleAISenvSSx
+buoylist <- dir("D:/Buoy_dfs", recursive=TRUE, full.names=TRUE, pattern="GPSwhaleAISenv")
+
+# Read though RDA files from external hard drive and assign to object
+## Won't work if GPS...SSx is already created in the D drive.
+for (i in buoylist) {
+  data <- readRDS(file = paste0("D:/Buoy_dfs/GPSwhaleAISenv", substr(i, 27,28), ".rda"))
+  assign(x = paste0("GPSwhaleAISenv", substr(i,27,28)), value = data)
+}
+
+# To read in one buoy file
+GPSwhaleAIS20.alt <- readRDS(file = "D:/Buoy_dfs/GPSwhaleAIS20_alt.rda")
+
+
+### Read in soundscape metrics (TOL only)
+# Path to folder on my external hard drive
+dirlist <- dir("D:/Soundscape_metrics/TOL_metrics", recursive=TRUE, full.names=TRUE, pattern="CCES_")
+
+# Read though files from external hard drive, change date format
+for (i in dirlist) {
+  data <- read_csv(i, show_col_types = FALSE) %>%
+    mutate(UTC = as.POSIXct(`yyyy-mm-ddTHH:MM:SSZ`, tz = 'UTC')) %>%
+    select(UTC, TOL_125, TOL_2000)
+  
+  assign(x = paste0("tol_", substr(i, 40, 41)), data)
+}
+
+#### To read in one soundscape file
[…]

### 2023-10-16 animations.Rmd
Create animations.Rmd

---
title: "animations"
author: "Marina Bozinovic"
date: "2023-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Map Animations

```{r}
# Install
#devtools::install_github("ropenscilabs/rnaturalearth")
#install.packages("rnaturalearth",
#                 repos = "http://packages.ropensci.org",
#                 type = "source")
#install.packages("gganimate")
#install.packages("gifski")
#install.packages("marmap")

# Libraries
library(sf)
library(tidyverse)
library(marmap)
library(gifski)
library(gganimate)
library(rnaturalearth)
```

### New ggplot wtih bplot as basemap

```{r}
library(mapproj)
drift <- track_BB_OL_10

# Get bathymetry (code from Ocean Tracking Network #https://www.youtube.com/watch?v=xT2F-Y9bKLk)
bgo <- getNOAA.bathy(lon1 = min(drift$long-0.7), lon2 = max(drift$long + 0.7),
                     lat1 = min(drift$lat-0.7), lat2=max(drift$lat + 0.7), resolution = 1)

# Create basemap
#bf <- fortify.bathy(bgo, geom=c("r", "c")) + scale_fill_etopo() # converts bathy object to data frame 
# ABOVE DOESNT WORK
bplot <- autoplot(bgo, geom = c("raster", "contour"), coast = TRUE) + 
  scale_fill_etopo() #creates ggplot object

# Plot drift
p <- bplot + 
  geom_path(data = drift, 
            aes(x=long,y=lat,group=spotID), 
            alpha = 0.5) +
  geom_point(data = drift, 
             aes(x=long,y=lat,group=spotID, size = OL_125, color = OL_125),
             shape=19, stroke = 1) +

  #formatting  
  scale_color_viridis_c(option = "inferno") +
  scale_size_continuous(range = c(0.1,10)) +
  labs(x=NULL, y=NULL, 
       fill = 'Depth',
       color = 'OL 125')+
  theme(panel.grid = element_blank())

# Plot AIS
b <- bplot + 
  geom_point(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName)) +
  scale_x_continuous(limits = c(min(drift$long-0.7), max(drift$long + 0.7))) + 
  scale_y_continuous(limits = c(min(drift$lat-0.7), max(drift$lat + 0.7))) +
  theme(panel.grid = element_blank())

# Plot AIS + drift
z <- bplot +
  geom_point(data = AIS_sf, mapping = aes(x=LON,y=LAT, group = VesselName)) +
  geom_path(data = drift, 
            aes(x=long,y=lat,group=spotID), 
            alpha = 0.5) +
  geom_point(data = drift, 
             aes(x=long,y=lat,group=spotID, size = OL_125, color = OL_125),
             shape=19, stroke = 1) +
  #formatting  
  scale_color_viridis_c(option = "inferno") +
  scale_size_continuous(range = c(0.1,10)) +
  labs(x=NULL, y=NULL, 
       fill = 'Depth',
       color = 'OL 125')+
  scale_x_continuous(limits = c(min(drift$long-0.7), max(drift$long + 0.7))) + 
  scale_y_continuous(limits = c(min(drift$lat-0.7), max(drift$lat + 0.7)))
```

```{r}
# Animate drift
anim <- p +
  transition_reveal(along = dateTime)+
  ease_aes('linear')+
  shadow_wake(0.1) +
  ggtitle("Date: {frame_along}") +

# Animate with control over frames per second and number of frames
animate(anim, nframes = 70, fps = 3)
anim_save("gifs/drift10.gif")

# Animate AIS. Need to continue working on this.
anim <-  b +
  transition_reveal(BaseDateTime) +
  ease_aes('linear') +
  shadow_wake(wake_length = 0.5) +
  ggtitle("Date: {frame_along}")

# Animate with control over frames per second and number of frames
animate(anim, nframes = 400, fps = 3)

```

### Plotting whale data

```{r}
# shorten time range
#whales10 <- whales10 %>% filter(StartTime > "2018-10-05 00:00:00")

# too smooth, doesnt work.
d <- bplot +
  geom_point(data = whales10, aes(x = as.numeric(Longitude), y = as.numeric(Latitude)), color = "cyan") +
  ggtitle("Drift 10: Whale detections")

anim <-  d +
  transition_reveal(along = StartTime) +
  exit_shrink() +
  ggtitle("Date: {frame_along}")

animate(anim, nframes = 400, fps = 3)

## The issue with showing drift path and whale detections is that gganimate only allows one type of transition for the whole plot. May need to join drift and whale detections and include null values?
```

### Join whale detections with drift 10

```{r}
w_d <- left_join(drift, whales10, by = "lat", keep = TRUE)

w_d_plt <- ggplot() +
  geom_path(data = w_d, aes(x=lon, y=lat.x)) +
  geom_point(data = w_d, aes(x=lon, y = lat.x, color = "species"))

w_d_plt + 
  transition_reveal(along = dateTime) +
  ease_aes() +
  shadow_trail(distance = 0.7) +
  ggtitle("Date:{frame_along}")

## Needs shorter time frame

```

### 2023-10-16 07_summary_stats.Rmd
Update 07_summary_stats.Rmd

@@
 ---
+
+# 10th, 50th (mean), and 90th Percentiles
 
@@
 
-Read in RDS files of each buoy + all environ variables + AIS (pending)
-```{r}
-ftlist <- dir("data/", recursive=TRUE, full.names=TRUE, pattern="fulltrack")
-
-for (ft in ftlist) {
-  df <- readRDS(ft)
-  assign(paste0("fulltrack", substr(ft, 15, 16)), df)
-}
-```
-
 Read in RDA files for percentiles
@@
 
-Create df from 1 percentile
-```{r}
-# 1 Percentile
-pct01_07 <- cbind(BB_pct01_07, TO_pct01_07) %>%
-  select(., -3)
-pct01_08 <- cbind(BB_pct01_08, TO_pct01_08) %>%
-  select(., -3)
-pct01_10 <- cbind(BB_pct01_10, TO_pct01_10) %>%
-  select(., -3)
-pct01_12 <- cbind(BB_pct01_12, TO_pct01_12) %>%
-  select(., -3)
-pct01_13 <- cbind(BB_pct01_13, TO_pct01_13) %>%
-  select(., -3)
-pct01_14 <- cbind(BB_pct01_14, TO_pct01_14) %>%
-  select(., -3)
-pct01_16 <- cbind(BB_pct01_16, TO_pct01_16) %>%
-  select(., -3)
-pct01_18 <- cbind(BB_pct01_18, TO_pct01_18) %>%
-  select(., -3)
-pct01_19 <- cbind(BB_pct01_19, TO_pct01_19) %>%
-  select(., -3)
-pct01_20 <- cbind(BB_pct01_20, TO_pct01_20) %>%
-  select(., -3)
-pct01_21 <- cbind(BB_pct01_21, TO_pct01_21) %>%
-  select(., -3)
-pct01_22 <- cbind(BB_pct01_22, TO_pct01_22) %>%
-  select(., -3)
-pct01_23 <- cbind(BB_pct01_23, TO_pct01_23) %>%
-  select(., -3)
-
-pct01_07$buoy <- ("07")
-pct01_08$buoy <- ("08")
-pct01_10$buoy <- ("10")
-pct01_12$buoy <- ("12")
-pct01_13$buoy <- ("13")
-pct01_14$buoy <- ("14")
-pct01_16$buoy <- ("16")
-pct01_18$buoy <- ("18")
-pct01_19$buoy <- ("19")
-pct01_20$buoy <- ("20")
-pct01_21$buoy <- ("21")
-pct01_22$buoy <- ("22")
-pct01_23$buoy <- ("23")
-
-pct01_all <- rbind(pct01_07, pct01_08, pct01_10, pct01_12, pct01_13,
-                   pct01_14, pct01_16, pct01_18, pct01_19, pct01_20,
-                   pct01_21, pct01_22, pct01_23)
-
-# Group by and summarize
-pct01_all %>%
-  group_by(buoy) %>%
-  summarize(max(`TOL_125`))
-
-```
-
-Create df from 99 percentile
-```{r}
-# 99 Percentile
-pct99_07 <- cbind(BB_pct99_07, TO_pct99_07) %>%
-  select(., -3)
-pct99_08 <- cbind(BB_pct99_08, TO_pct99_08) %>%
-  select(., -3)
-pct99_10 <- cbind(BB_pct99_10, TO_pct99_10) %>%
-  select(., -3)
-pct99_12 <- cbind(BB_pct99_12, TO_pct99_12) %>%
-  select(., -3)
-pct99_13 <- cbind(BB_pct99_13, TO_pct99_13) %>%
-  select(., -3)
-pct99_14 <- cbind(BB_pct99_14, TO_pct99_14) %>%
-  select(., -3)
-pct99_16 <- cbind(BB_pct99_16, TO_pct99_16) %>%
-  select(., -3)
-pct99_18 <- cbind(BB_pct99_18, TO_pct99_18) %>%
-  select(., -3)
-pct99_19 <- cbind(BB_pct99_19, TO_pct99_19) %>%
-  select(., -3)
-pct99_20 <- cbind(BB_pct99_20, TO_pct99_20) %>%
-  select(., -3)
-pct99_21 <- cbind(BB_pct99_21, TO_pct99_21) %>%
-  select(., -3)
-pct99_22 <- cbind(BB_pct99_22, TO_pct99_22) %>%
-  select(., -3)
-pct99_23 <- cbind(BB_pct99_23, TO_pct99_23) %>%
-  select(., -3)
-
-# Add column with buoy number
-pct99_07$buoy <- ("07")
-pct99_08$buoy <- ("08")
-pct99_10$buoy <- ("10")
-pct99_12$buoy <- ("12")
-pct99_13$buoy <- ("13")
-pct99_14$buoy <- ("14")
-pct99_16$buoy <- ("16")
-pct99_18$buoy <- ("18")
-pct99_19$buoy <- ("19")
-pct99_20$buoy <- ("20")
-pct99_21$buoy <- ("21")
-pct99_22$buoy <- ("22")
-pct99_23$buoy <- ("23")
-
-pct99_all <- rbind(pct99_07, pct99_08, pct99_10, pct99_12, pct99_13,
-                   pct99_14, pct99_16, pct99_18, pct99_19, pct99_20,
-                   pct99_21, pct99_22, pct99_23)
-```
-
-Plot 99pct and 1pct by buoy
-```{r fig.height=5, fig.width=13}
-# Group by and summarize
-pct99_all %>%
-  group_by(buoy) %>%
-  summarize(max(`TOL_125`)) %>%
-  ggplot(aes(x = buoy, y = `max(TOL_125)`)) + geom_bar(stat = "identity")
-
-#TOL timeline by buoy
-pct99_all %>%
-  ggplot(aes(x = dateTime, y = TOL_2000, color = buoy)) + geom_line()
-
-# Average dB by hour for each frequency
-pct99_16 %>%
-  group_by(time = floor_date(dateTime, "1 hour")) %>%
-  summarize(mean_BB = mean(`BB_20-24000`),
-            meanTOL_125 = mean(TOL_125),
-            meanTOL_250 = mean(TOL_250),
-            meanTOL_500 = mean(TOL_500),
-            meanTOL_2000 = mean(TOL_2000),
-            meanTOL_4000 = mean(TOL_4000),
-            meanTOL_20000 = mean(TOL_20000)) %>%
-  pivot_longer(cols = mean_BB:meanTOL_20000,
-               names_to = "frequency",
-               values_to="dB") %>%
-  ggplot(aes(x = time, y = dB)) + geom_line(aes(color = frequency)) #+
-  #geom_point(data = allDrifts, aes(color = species))
-# Show more dates/times on x axis
-# 99pct, buoy 16 has a strong peak later in journey at 4kHz
-
-ggplot() +
-  geom_point(data = pct01_23, aes(x = lubridate::round_date(dateTime, "2 hours"), y = TOL_125), color = 'steelblue') +
-  #geom_line(data = pct99_23, aes(x = dateTime, y = TOL_125), color = 'coral') +
-  facet_grid()
-
-```
-
 Bind percentile df (pctx_x) with GPS track df (GPSwhalex)
@@
 ```{r}
+# 50th Percentile for Buoy 14 was calculated later and somehow has a differing number of rows with slightly different UTCs
+# To allow for a cbind, I add two rows that are missing and use NAs for soundscape metrics.
+
 # Create columns with percentiles and BIND
@@
 
-# Bind
-#pct14 <- cbind(pct10_14_select, pct50_14_select, pct90_14_select)
+# Remove last two rows of pct50_14_select in order to bind.
+pct50_14_select <- pct50_14_select[1:1979,]
+
+# Bind
+pct14 <- cbind(pct10_14_select, pct50_14_select, pct90_14_select)
 
@@
             pct90_2000_mean = mean(pct90_TOL2000))
+
+ggplot(buoy23pctSummary) + geom_bar(aes(x = species))
 ```
@@
 ## Buoy 14 ##
-#kable(buoy14pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
-#                                     caption = "Summary Percentiles for Buoy 14") %>%
-#  kableExtra::kable_classic() %>%
-#  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
+sumPct14 <- kable(buoy14pctSummary, col.names = c("Animal", "10th", "Mean", "90th", "10th", "Mean", "90th"),
+                                     caption = "Summary Percentiles for Buoy 14") %>%
+  kableExtra::kable_classic() %>%
+  kableExtra::add_header_above(c(" " = 1, "TOL 125" = 3, "TOL 2000" = 3))
 
@@
 
-pdf(paste("timeSeries8-31",".pdf"), width=14, height=10)
+#pdf(paste("timeSeries8-31",".pdf"), width=14, height=10)
 
 ### Buoy 7 ####
-W.07 <- fulltrack07 %>%
-  select(UTC, BWpresence, species) %>%
+W.07 <- GPSwhaleAISenvSS07 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-g1 <- (plotTimeseries(fulltrack07, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 7: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.07, bin = 'hour', type = 'presence', start = min(fulltrack07$UTC),
-                         end = max(fulltrack07$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS07, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 7: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.07, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS07$UTC),
+                         end = max(GPSwhaleAISenvSS07$UTC), fill = NA, color = "white")
 
@@
 
-t1 <- plotTimeseries(fulltrack07, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 7: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.07, bin = 'hour', type = 'presence', start = min(fulltrack07$UTC),
-                         end = max(fulltrack07$UTC), fill = NA, color = "white")
+t1 <- plotTimeseries(GPSwhaleAISenvSS07, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 7: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.07, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS07$UTC),
+                         end = max(GPSwhaleAISenvSS07$UTC), fill = NA, color = "white")
 
@@
 ###### Buoy 8 ####
-W.08 <- fulltrack08 %>%
-  select(UTC, BWpresence, species) %>%
-  filter(species == "ZC")
+W.08 <- GPSwhaleAISenvSS08 %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "PM")
 
@@
 
-g1 <- (plotTimeseries(fulltrack08, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.08, bin = 'hour', type = 'presence', start = min(fulltrack08$UTC),
-                         end = max(fulltrack08$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack08, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.08, bin = 'hour', type = 'presence', start = min(fulltrack08$UTC),
-                         end = max(fulltrack08$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS08, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 125 Plot + PM'))
+g2 <- plotPresGrid(W.08, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS08$UTC),
+                         end = max(GPSwhaleAISenvSS08$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS08, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 2000 Plot + PM')
+t2 <- plotPresGrid(W.08, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS08$UTC),
+                         end = max(GPSwhaleAISenvSS08$UTC), fill = NA, color = "white")
 
@@
 
-W.10 <- fulltrack10 %>%
-  select(UTC, BWpresence, species) %>%
-  filter(species == "ZC")
+W.10 <- GPSwhaleAISenvSS10 %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "PM")
 
@@
 
-g1 <- (plotTimeseries(fulltrack10, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
-                         end = max(fulltrack10$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
-                         end = max(fulltrack10$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS10, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 125 Plot + PM'))
+g2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS10$UTC),
+                         end = max(GPSwhaleAISenvSS10$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot + PM')
+t2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS10$UTC),
+                         end = max(GPSwhaleAISenvSS10$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 12
-W.12 <- fulltrack12 %>%
-  select(UTC, BWpresence, species) %>%
-  filter(species == "ZC")
+W.12 <- GPSwhaleAISenvSS12 %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "PM")
 
@@
 
-g1 <- (plotTimeseries(fulltrack12, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 12: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.12, bin = 'hour', type = 'presence', start = min(fulltrack12$UTC),
-                         end = max(fulltrack12$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack12, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 12: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.12, bin = 'hour', type = 'presence', start = min(fulltrack12$UTC),
-                         end = max(fulltrack12$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS12, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 12: TOL 125 Plot + PM'))
+g2 <- plotPresGrid(W.12, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS12$UTC),
+                         end = max(GPSwhaleAISenvSS12$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS12, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 12: TOL 2000 Plot + PM')
+t2 <- plotPresGrid(W.12, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS12$UTC),
+                         end = max(GPSwhaleAISenvSS12$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 13
-W.13 <- fulltrack13 %>%
-  select(UTC, BWpresence, species) %>%
+W.13 <- GPSwhaleAISenvSS13 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-g1 <- (plotTimeseries(fulltrack13, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(fulltrack13$UTC),
-                         end = max(fulltrack13$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack13, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(fulltrack13$UTC),
-                         end = max(fulltrack13$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS13, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS13$UTC),
+                         end = max(GPSwhaleAISenvSS13$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS13, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS13$UTC),
+                         end = max(GPSwhaleAISenvSS13$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 14
-W.14 <- fulltrack14 %>%
-  select(UTC, BWpresence, species) %>%
-  filter(species == "ZC")
+W.14 <- GPSwhaleAISenvSS14 %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "PM")
 
@@
 
-g1 <- (plotTimeseries(fulltrack14, column='TOL_125', bin='hour', style = 'heatmap', title='Buoy 14: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.14, bin = 'hour', type = 'presence', start = min(fulltrack14$UTC),
-                         end = max(fulltrack14$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack14, column='TOL_2000', bin='hour', style = 'heatmap', title='Buoy 14: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.14, bin = 'hour', type = 'presence', start = min(fulltrack14$UTC),
-                         end = max(fulltrack14$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS14, column='TOL_125', bin='hour', style = 'heatmap', title='Buoy 14: TOL 125 Plot + PM'))
+g2 <- plotPresGrid(W.14, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS14$UTC),
+                         end = max(GPSwhaleAISenvSS14$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS14, column='TOL_2000', bin='hour', style = 'heatmap', title='Buoy 14: TOL 2000 Plot + PM')
+t2 <- plotPresGrid(W.14, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS14$UTC),
+                         end = max(GPSwhaleAISenvSS14$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 16
-W.16 <- fulltrack16 %>%
-  select(UTC, BWpresence, species) %>%
+W.16 <- GPSwhaleAISenvSS16 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-g1 <- (plotTimeseries(fulltrack16, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 16: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.16, bin = 'hour', type = 'presence', start = min(fulltrack16$UTC),
-                         end = max(fulltrack16$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack16, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 16: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.16, bin = 'hour', type = 'presence', start = min(fulltrack16$UTC),
-                         end = max(fulltrack16$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS16, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 16: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.16, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS16$UTC),
+                         end = max(GPSwhaleAISenvSS16$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS16, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 16: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.16, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS16$UTC),
+                         end = max(GPSwhaleAISenvSS16$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 18
-W.18 <- fulltrack18 %>%
-  select(UTC, BWpresence, species) %>%
+W.18 <- GPSwhaleAISenvSS18 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-g1 <- (plotTimeseries(fulltrack18, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 18: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.18, bin = 'hour', type = 'presence', start = min(fulltrack18$UTC),
-                         end = max(fulltrack18$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack18, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 18: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.18, bin = 'hour', type = 'presence', start = min(fulltrack18$UTC),
-                         end = max(fulltrack18$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS18, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 18: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.18, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS18$UTC),
+                         end = max(GPSwhaleAISenvSS18$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS18, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 18: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.18, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS18$UTC),
+                         end = max(GPSwhaleAISenvSS18$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 19
-W.19 <- fulltrack19 %>%
-  select(UTC, BWpresence, species) %>%
+W.19 <- GPSwhaleAISenvSS19 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-g1 <- (plotTimeseries(fulltrack19, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 19: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.19, bin = 'hour', type = 'presence', start = min(fulltrack19$UTC),
-                         end = max(fulltrack19$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack19, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 19: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.19, bin = 'hour', type = 'presence', start = min(fulltrack19$UTC),
-                         end = max(fulltrack19$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS19, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 19: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.19, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS19$UTC),
+                         end = max(GPSwhaleAISenvSS19$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS19, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 19: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.19, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS19$UTC),
+                         end = max(GPSwhaleAISenvSS19$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 20
-W.20 <- fulltrack20 %>%
-  select(UTC, BWpresence, species) %>%
+W.20 <- GPSwhaleAISenvSS20 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-g1 <- (plotTimeseries(fulltrack20, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(fulltrack20$UTC),
-                         end = max(fulltrack20$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack20, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(fulltrack20$UTC),
-                         end = max(fulltrack20$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS20, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS20$UTC),
+                         end = max(GPSwhaleAISenvSS20$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS20, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS20$UTC),
+                         end = max(GPSwhaleAISenvSS20$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 21
-W.21 <- fulltrack21 %>%
-  select(UTC, BWpresence, species) %>%
+W.21 <- GPSwhaleAISenvSS21 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-g1 <- (plotTimeseries(fulltrack21, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 21: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.21, bin = 'hour', type = 'presence', start = min(fulltrack21$UTC),
-                         end = max(fulltrack21$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack21, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 21: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.21, bin = 'hour', type = 'presence', start = min(fulltrack21$UTC),
-                         end = max(fulltrack21$UTC), fill = NA, color = "white")
-
-combined.t <- t1
-combined.t$layers <- c(combined.t$layers, t2$layers)
-
-combined.g + combined.t
-```
-
-```{r fig.width=12}
-## Buoy 12
-W.22 <- fulltrack22 %>%
-  select(UTC, BWpresence, species) %>%
-  filter(species == "ZC")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS21, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 21: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.21, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS21$UTC),
+                         end = max(GPSwhaleAISenvSS21$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS21, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 21: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.21, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS21$UTC),
+                         end = max(GPSwhaleAISenvSS21$UTC), fill = NA, color = "white")
+
+combined.t <- t1
+combined.t$layers <- c(combined.t$layers, t2$layers)
+
+combined.g + combined.t
+```
+
+```{r fig.width=12}
+## Buoy 22
+W.22 <- GPSwhaleAISenvSS22 %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "PM")
 
@@
 
-g1 <- (plotTimeseries(fulltrack22, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 22: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.22, bin = 'hour', type = 'presence', start = min(fulltrack22$UTC),
-                         end = max(fulltrack22$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack22, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 22: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.22, bin = 'hour', type = 'presence', start = min(fulltrack22$UTC),
-                         end = max(fulltrack22$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS22, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 22: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.22, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS22$UTC),
+                         end = max(GPSwhaleAISenvSS22$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS22, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 22: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.22, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS22$UTC),
+                         end = max(GPSwhaleAISenvSS22$UTC), fill = NA, color = "white")
 
@@
 ## Buoy 23
-W.23 <- fulltrack23 %>%
-  select(UTC, BWpresence, species) %>%
+W.23 <- GPSwhaleAISenvSS23 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-g1 <- (plotTimeseries(fulltrack23, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 23: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.23, bin = 'hour', type = 'presence', start = min(fulltrack23$UTC),
-                         end = max(fulltrack23$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(fulltrack23, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 23: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.23, bin = 'hour', type = 'presence', start = min(fulltrack23$UTC),
-                         end = max(fulltrack23$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISenvSS23, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 23: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.23, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS23$UTC),
+                         end = max(GPSwhaleAISenvSS23$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISenvSS23, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 23: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.23, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS23$UTC),
+                         end = max(GPSwhaleAISenvSS23$UTC), fill = NA, color = "white")
 
@@
 ```{r}
-W.08 <- fulltrack08 %>%
-  select(UTC, BWpresence, species) %>%
+W.08 <- GPSwhaleAISenvSS08 %>%
+  select(UTC, Wpresence, species) %>%
   filter(species == "ZC")
@@
 
-plotTimeseries(fulltrack08, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 2000 Plot') +  W.08plot
+plotTimeseries(GPSwhaleAISenvSS08, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 2000 Plot') +  W.08plot
 
@@
 
-W.10 <- fulltrack10 %>%
-  select(UTC, BWpresence, species) %>%
-  filter(species == "ZC")
-
-W.10plot <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
-                         end = max(fulltrack10$UTC), by = 'BWpresence', alpha = 0.01, color = "red")
-
-plotTimeseries(fulltrack10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot') +  W.10plot
-plotTimeseries(fulltrack10, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 125 Plot') +  W.10plot
+W.10 <-  GPSwhaleAISenvSS10 %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "ZC")
+
+W.10plot <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min( GPSwhaleAISenvSS10$UTC),
+                         end = max( GPSwhaleAISenvSS10$UTC), by = 'Wpresence', alpha = 0.01, color = "red")
+
+plotTimeseries( GPSwhaleAISenvSS10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot') +  W.10plot
+plotTimeseries( GPSwhaleAISenvSS10, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 125 Plot') +  W.10plot
 #############
 
-W.13 <- fulltrack13 %>%
-  select(UTC, BWpresence, species) %>%
-  filter(species == "ZC")
-
-W.13plot <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(fulltrack13$UTC),
-                         end = max(fulltrack13$UTC), by = 'BWpresence', alpha = 0.5)
-
-plotTimeseries(fulltrack13, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 2000 Plot') +  W.13plot
-plotTimeseries(fulltrack13, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 125 Plot') +  W.13plot
+W.13 <-  GPSwhaleAISenvSS13 %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "ZC")
+
+W.13plot <- plotPresGrid(W.13, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS13$UTC),
+                         end = max(GPSwhaleAISenvSS13$UTC), by = 'Wpresence', alpha = 0.5)
+
+plotTimeseries(GPSwhaleAISenvSS13, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 2000 Plot') +  W.13plot
+plotTimeseries(GPSwhaleAISenvSS13, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 13: TOL 125 Plot') +  W.13plot
 ##############
 
-W.20 <- fulltrack20 %>%
-  select(UTC, BWpresence, species) %>%
-  filter(species == "ZC")
-
-W.20plot <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(fulltrack20$UTC),
-                         end = max(fulltrack20$UTC),  by = 'BWpresence', alpha = 0.5)
-
-plotTimeseries(fulltrack20, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot') +  W.20plot
-plotTimeseries(fulltrack20, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot') +  W.20plot
+W.20 <- GPSwhaleAISenvSS20 %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "ZC")
+
+W.20plot <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS20$UTC),
+                         end = max(GPSwhaleAISenvSS20$UTC),  by = 'Wpresence', alpha = 0.5)
+
+plotTimeseries(GPSwhaleAISenvSS20, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot') +  W.20plot
+plotTimeseries(GPSwhaleAISenvSS20, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot') +  W.20plot
 
@@
 
-g1 <- plotTimeseries(fulltrack10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot')
-g2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
-                         end = max(fulltrack10$UTC), fill = NA, color = "white")
-S.10 <- fulltrack10 %>% select(UTC, shipin10km) %>% filter(shipin10km == "1")
-g3 <- plotPresGrid(S.10, bin = 'hour', type = 'presence', start = min(fulltrack10$UTC),
-                         end = max(fulltrack10$UTC), fill = NA, color = "red")
+g1 <- plotTimeseries(GPSwhaleAISenvSS10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot')
+g2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS10$UTC),
+                         end = max(GPSwhaleAISenvSS10$UTC), fill = NA, color = "white")
+S.10 <- GPSwhaleAISenvSS10 %>% select(UTC, shipin10km) %>% filter(shipin10km == "1")
+g3 <- plotPresGrid(S.10, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS10$UTC),
+                         end = max(GPSwhaleAISenvSS10$UTC), fill = NA, color = "red")
 
@@
 
-Bind ALL DRIFTS into one dataframe
-```{r}
-allDrifts <- rbind(fulltrack07, fulltrack08, fulltrack10, 
-                   fulltrack12, fulltrack13, fulltrack14, 
-                   fulltrack16, fulltrack18, fulltrack19, 
-                   fulltrack20, fulltrack21, fulltrack22, fulltrack23)
+Load ALL DRIFTS into one dataframe
+```{r}
+load('data/allDrifts.rda')
 ```
@@
 allDrifts %>%
-  count(station, BWpresence) %>%
-  filter(!BWpresence == 0) %>%
+  count(Deployment, Wpresence) %>%
+  filter(!Wpresence == 0) %>%
   arrange(desc(n))
@@
 ggplot() +
-     geom_path(data=allDrifts, aes(x=Longitude, y=Latitude, group=factor(station), col=factor(station))) 
+     geom_path(data=allDrifts, aes(x=Longitude, y=Latitude, group=factor(Deployment), col=factor(Deployment))) 
 ```
@@
 # Sort sperm whale detection by decreasing BB sound levels
-allDrifts %>%
-  dplyr::filter(species == "PM") %>%
-  arrange(desc(BB_20.24000))
-
 PM <- allDrifts %>%
@@
 PM %>%
-  group_by(station) %>%
+  group_by(Deployment) %>%
   summarize(mean = mean(TOL_2000), n = n())
 
-print(paste("PM Max BB sound level: ", max(PM$BB_20.24000, na.rm = TRUE)))
-print(paste("PM Min BB sound level: ", min(PM$BB_20.24000, na.rm = TRUE)))
 print(paste("PM Max TOL125 sound level: ", max(PM$TOL_125, na.rm = TRUE)))
 print(paste("PM Min TOL125 sound level: ", min(PM$TOL_125, na.rm = TRUE)))
-print(paste("PM Max TOL20000 sound level: ", max(PM$TOL_20000, na.rm = TRUE)))
-print(paste("PM Min TOL20000 sound level: ", min(PM$TOL_20000, na.rm = TRUE)))
+print(paste("PM Max TOL20000 sound level: ", max(PM$TOL_2000, na.rm = TRUE)))
+print(paste("PM Min TOL20000 sound level: ", min(PM$TOL_2000, na.rm = TRUE)))
 
@@
   dplyr::filter(species == "BWC")
-print(paste("BWC Max BB sound level: ", max(BWC$BB_20.24000, na.rm = TRUE)))
-print(paste("BWC Min BB sound level: ", min(BWC$BB_20.24000, na.rm = TRUE)))
 print(paste("BWC Max TOL125 sound level: ", max(BWC$TOL_125, na.rm = TRUE)))
 print(paste("BWC Min TOL125 sound level: ", min(BWC$TOL_125, na.rm = TRUE)))
-print(paste("BWC Max TOL20000 sound level: ", max(BWC$TOL_20000, na.rm = TRUE)))
-print(paste("BWC Min TOL20000 sound level: ", min(BWC$TOL_20000, na.rm = TRUE)))
+print(paste("BWC Max TOL20000 sound level: ", max(BWC$TOL_2000, na.rm = TRUE)))
+print(paste("BWC Min TOL20000 sound level: ", min(BWC$TOL_2000, na.rm = TRUE)))
 
@@
   dplyr::filter(species == "BB")
-print(paste("BB Max BB sound level: ", max(BB$BB_20.24000, na.rm = TRUE)))
-print(paste("BB Min BB sound level: ", min(BB$BB_20.24000, na.rm = TRUE)))
 print(paste("BB Max TOL125 sound level: ", max(BB$TOL_125, na.rm = TRUE)))
 print(paste("BB Min TOL125 sound level: ", min(BB$TOL_125, na.rm = TRUE)))
-print(paste("BB Max TOL20000 sound level: ", max(BB$TOL_20000, na.rm = TRUE)))
-print(paste("BB Min TOL20000 sound level: ", min(BB$TOL_20000, na.rm = TRUE)))
+print(paste("BB Max TOL20000 sound level: ", max(BB$TOL_2000, na.rm = TRUE)))
+print(paste("BB Min TOL20000 sound level: ", min(BB$TOL_2000, na.rm = TRUE)))
 
@@
   dplyr::filter(species == "BW37V")
-print(paste("BW37V Max BB sound level: ", max(BW37V$BB_20.24000, na.rm = TRUE)))
-print(paste("BW37V Min BB sound level: ", min(BW37V$BB_20.24000, na.rm = TRUE)))
 print(paste("BW37V Max TOL125 sound level: ", max(BW37V$TOL_125, na.rm = TRUE)))
 print(paste("BW37V Min TOL125 sound level: ", min(BW37V$TOL_125, na.rm = TRUE)))
-print(paste("BW37V Max TOL20000 sound level: ", max(BW37V$TOL_20000, na.rm = TRUE)))
-print(paste("BW37V Min TOL20000 sound level: ", min(BW37V$TOL_20000, na.rm = TRUE)))
+print(paste("BW37V Max TOL20000 sound level: ", max(BW37V$TOL_2000, na.rm = TRUE)))
+print(paste("BW37V Min TOL20000 sound level: ", min(BW37V$TOL_2000, na.rm = TRUE)))
 
@@
   dplyr::filter(species == "BW43")
-print(paste("BW43 Max BB sound level: ", max(BW43$BB_20.24000, na.rm = TRUE)))
-print(paste("BW43 Min BB sound level: ", min(BW43$BB_20.24000, na.rm = TRUE)))
 print(paste("BW43 Max TOL125 sound level: ", max(BW43$TOL_125, na.rm = TRUE)))
 print(paste("BW43 Min TOL125 sound level: ", min(BW43$TOL_125, na.rm = TRUE)))
-print(paste("BW43 Max TOL20000 sound level: ", max(BW43$TOL_20000, na.rm = TRUE)))
-print(paste("BW43 Min TOL20000 sound level: ", min(BW43$TOL_20000, na.rm = TRUE)))
+print(paste("BW43 Max TOL20000 sound level: ", max(BW43$TOL_2000, na.rm = TRUE)))
+print(paste("BW43 Min TOL20000 sound level: ", min(BW43$TOL_2000, na.rm = TRUE)))
 
@@
   dplyr::filter(species == "ZC")
-print(paste("ZC Max BB sound level: ", max(ZC$BB_20.24000, na.rm = TRUE)))
-print(paste("ZC Min BB sound level: ", min(ZC$BB_20.24000, na.rm = TRUE)))
 print(paste("ZC Max TOL125 sound level: ", max(ZC$TOL_125, na.rm = TRUE)))
 print(paste("ZC Min TOL125 sound level: ", min(ZC$TOL_125, na.rm = TRUE)))
-print(paste("ZC Max TOL20000 sound level: ", max(ZC$TOL_20000, na.rm = TRUE)))
-print(paste("ZC Min TOL20000 sound level: ", min(ZC$TOL_20000, na.rm = TRUE)))
-
-```
+print(paste("ZC Max TOL20000 sound level: ", max(ZC$TOL_2000, na.rm = TRUE)))
+print(paste("ZC Min TOL20000 sound level: ", min(ZC$TOL_2000, na.rm = TRUE)))
+```
+
 Number of detections
 ```{r}
+# Percent of time whales are detected
+print(paste(nrow(allDrifts[allDrifts$Wpresence == 1,]) / nrow(allDrifts) * 100,"%"))
+
 print(paste("Number of PM detections: ", nrow(PM)))
@@
 ```
-How do I find WHERE the highest/lowest detections are happening?
-```{r}
-ZC <- allDrifts %>%
-  dplyr::filter(species == "ZC")
-
-ZC[which.max(ZC$BB_20.24000),]
-ZCmax2000 <- ZC[which.max(ZC$TOL_2000),]
-
-#Point of ZC max TOL2000 sound level. Need to put in context of map.
-ggplot(data = ZCmax2000, aes(x = Longitude, y = Latitude)) +
-  geom_point()
-
-#not working
-allDrifts %>%
-  group_by(species) %>%
-  filter(max(allDrifts$BB_20.24000))
-
-allDrifts %>%
-  summarise_if(is.numeric, list(~ min(., na.rm=TRUE)))
-
-```

### 2023-10-16 06_gam.Rmd
Create 06_gam.Rmd

---
title: "06_gam"
author: "Marina Bozinovic"
date: "2023-10-09"
output: html_document
---

Compare results of GAM to Virgili et al. 2019, Fiedler et al. 2023

Required Packages
```{r setup}
library(tidyverse)
library(mgcv)
```

Load allDrifts object
```{r setup, include=FALSE}
load(allDrifts, file = "data/allDrifts.rda")
```

### GAMs

```{r}
model3 <- gam(Wpresence ~ depth + TOL_125 + dist2shore + mldTemp, 
              data = allDrifts, family = binomial)
#plot(model3)

#try fitting linear model

lm_depth <- lm(Wpresence ~ depth, data = GPSwhaleAISenvSS10)
ggplot(allDrifts, aes(sst_mean, chlorophyll_mean)) +
  geom_point() +
  geom_smooth()
```
Categorical term
```{r}
# ALL variables, GAM
model1 <- gam(Wpresence>0 ~ s(TOL_125) + s(TOL_2000) + s(depth) + s(mldDepth) + s(Latitude)
              + s(Longitude) + s(sst_mean) + s(curl_mean) + s(dist2slope) 
              + s(bathy_slope) + s(dist2shore) + s(chlorophyll_mean) + s(ssh_mean) 
              + s(mldTemp) + s(ttDepth) + s(ttTemp) + s(temp400) + s(sal400),
              data = allDrifts, family = binomial, method = "REML")

# ALL variables, GLM
model1GLM <- glm(Wpresence ~ TOL_125 + TOL_2000 + depth + mldDepth + Latitude
              + Longitude + sst_mean + curl_mean + dist2slope
              + bathy_slope + dist2shore + chlorophyll_mean + ssh_mean 
              + mldTemp + ttDepth + ttTemp + temp400 + sal400,
              data = allDrifts, family = binomial)

# By species? doesnt work, may need to create dummy variables
model2 <- gam(Wpresence ~ s(sst_mean, by = as.factor(species)), 
              data = allDrifts, family = binomial, method = "REML")

# Soundscape metrics only, GAM
model4 <- gam(Wpresence ~ s(TOL_125, k = 10) + 
                s(TOL_2000, k = 10),
              data = allDrifts, family = binomial, method = "REML")

# Soundscape metrics only, GLM
model4GLM <- glm(Wpresence ~ TOL_125 + TOL_2000,
              data = allDrifts, family = binomial)

model125 <- gam(Wpresence ~ s(TOL_125, k = 10),
              data = allDrifts, family = binomial, method = "REML")
plot(model125, trans = plogis)
summary(model125)
gam.check(model125)

# Environmental var only
model5 <- gam(Wpresence ~ s(depth) + s(mldDepth) + s(Latitude)
              + s(Longitude) + s(sst_mean) + s(curl_mean) + s(dist2slope) 
              + s(bathy_slope) + s(dist2shore) + s(chlorophyll_mean) + s(ssh_mean) 
              + s(mldTemp) + s(ttDepth) + s(ttTemp) + s(temp400) + s(sal400),
              data = allDrifts, family = binomial, method = "REML")

plogis(coef(model1[1]))
summary(model1)
summary(model4)
summary(model5)
summary(model4GLM)
plot(model4GLM)
gam.check(model1)
gam.check(model4)

#Log odds scale (when used with family = binomial)
plot(model1, pages = 6)
plot(model4, pages = 4, trans = plogis)

#Probability scale + centered on the intercept + adding intercept-related uncertainty with seWithMean arg
plot(model1, pages = 14, trans = plogis,
     shift = coef(model1)[1])#,
     #seWithMean = TRUE)

plot(model2, pages = 1, trans = plogis,
     shift = coef(model2)[1])

plot(model4, pages = 6, trans = plogis,
     shift = coef(model4)[1])
```

Top 5 environmental variables (by RF) with one drift
```{r}
# Environ only
model1 <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + ttDepth + sal400,
              data = GPSwhaleAISenvSS23, family = binomial, method = "REML")

plot(model1, trans = plogis, all.terms = TRUE)
k.check(model1)
summary(model1)
gam.check(model1)

# Add soundscape metrics to model 1
model2 <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + ttDepth + sal400
              + TOL_125 + TOL_2000,
              data = GPSwhaleAISenvSS10, family = binomial, method = "REML")

#plot(model2, trans = plogis) ## doesn't work
summary(model2)
gam.check(model2)
# 18.3% deviance explained (buoy 7)
# 1.25% deviance explained (buoy 8)
# 8.56% deviance explained (buoy 10)
# 1.39% deviance explained (buoy 12)
# 11.1% deviance explained (buoy 13)
# 28% deviance explained (buoy 14)
# 15% deviance explained (buoy 16)
# 9.92% deviance explained (buoy 18)
# 8.57% deviance explained (buoy 19)
# 7.3% deviance explained (buoy 20)
# 38.3% deviance explained (buoy 21)
# 15.2 % deviance explained (buoy 22)
# 20.4% deviance explained (buoy 23)

# Depth parameter only
model_depth_only <- gam(Wpresence>0 ~ depth,
                        data = GPSwhaleAISenvSS10, family = binomial, method = "REML")
summary(model_depth_only)
gam.check(model_depth_only)
```
GAM with FULL DATASET (not great)
```{r}
alldriftGAM <- gam(Wpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400
              + TOL_125 + TOL_2000,
              data = allDrifts, family = binomial, method = "REML")
summary(alldriftGAM)
gam.check(alldriftGAM)
```

GAMs with s()
```{r}
# Environ only, with s()
model3 <- gam(Wpresence>0 ~ s(depth) + s(sst_mean) + s(curl_mean) + 
                s(ttDepth) + s(sal400),
              data = GPSwhaleAISenvSS07, family = binomial, method = "REML")

plot(model3, trans = plogis)
summary(model3)
gam.check(model3)
# 61.9 % deviance explained (buoy 7)
# 1.42% deviance explained (buoy 8)
# 12.5% deviance explained (buoy 10)
# 1.66% deviance explained (buoy 12)
# 12.8% deviance explained (buoy 13)
# 41.4% deviance explained (buoy 14)
# 39% deviance explained (buoy 16)
# 16.6% deviance explained (buoy 18)
# 15.5% deviance explained (buoy 19)
# 5.8% deviance explained (buoy 20)
## Iteration limit reached without full convergence -- 71.7% deviance explained (buoy 21) ???
# 18.6% deviance explained (buoy 22)
# 49.6% deviance explained (buoy 23)

#Add soundscape metrics to model 3
model4 <- gam(Wpresence>0 ~ s(depth) + s(sst_mean) + s(curl_mean) + 
                s(ttDepth) + s(sal400) + s(TOL_125) + s(TOL_2000),
              data = GPSwhaleAISenvSS07, family = binomial, method = "REML")

plot(model4, trans = plogis)
summary(model4)
gam.check(model4)
# 14.5% deviance explained

# mix and match?
model5 <- gam(BWpresence>0 ~ depth + sst_mean + s(mldTemp) + 
                s(ttDepth) + sal400,
              data = fulltrack21, family = binomial, method = "REML")

summary(model5)
gam.check(model5)
# 57.4 % deviance explained
plot(model5)

```
Higher edfs describe more wiggly curves. Doesn't mean significant

#GAM by ZC
```{r}
ZC_filt <- allDrifts %>% 
  filter(species %in% c("ZC", NA))

ZC_GAM <- gam(Wpresence>0 ~ depth + mldTemp + ttDepth + sal400,
              data = ZC_filt, family = binomial, method = "REML")
summary(ZC_GAM)
gam.check(ZC_GAM)
```

# GAM by PM
```{r}
PM_filt <- allDrifts %>% 
  filter(species %in% c("PM", NA))

PM_GAM <- gam(Wpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400 +
                TOL_125 + TOL_2000,
              data = PM_filt, family = binomial, method = "REML")
summary(PM_GAM)
gam.check(PM_GAM)

# Using variables from Fiedler et al. 2023 for sperm whales
PM_GAM_fiedlerVar <- gam(Wpresence>0 ~ depth + mldDepth + curl_mean + ssh_mean + sal400,
              data = PM_filt, family = binomial, method = "REML")
summary(PM_GAM_fiedlerVar)
```

### 2023-10-16 05_analysis.Rmd
Create 05_analysis.Rmd

---
title: "04_analysis"
author: "Marina Bozinovic"
date: "2023-05-11"
output: html_document
---
# Exploratory Analysis and Modeling for all 13 buoys

Required Packages
```{r setup}
library(tidyverse)
```

Read in RDS files of each buoy + all environ variables + AIS
```{r}
ftlist <- dir("D:/Buoy_dfs/", recursive=TRUE, full.names=TRUE, pattern="GPSwhaleAISenvSS")

for (ft in ftlist) {
  df <- readRDS(ft)
  assign(paste0("GPSwhaleAISenvSS", substr(ft, 29, 30)), df)
}
```

Bind ALL DRIFTS into one dataframe
```{r}
dlist <- grep(pattern = "GPSwhaleAISenvSS", names(.GlobalEnv), value = T)

# Create a new column that specifies the drift number before the bind.
for (col in dlist){
  data <- get(col) %>% mutate(Deployment = substr(col,17,18))
  assign(paste0("GPSwhaleAISenvSS", substr(col,17,18)), data)
}

# Bind and save allDrifts dataframe
allDrifts <- rbind(GPSwhaleAISenvSS07, GPSwhaleAISenvSS08, GPSwhaleAISenvSS10, GPSwhaleAISenvSS12,
                   GPSwhaleAISenvSS13, GPSwhaleAISenvSS14, GPSwhaleAISenvSS16, GPSwhaleAISenvSS18,
                   GPSwhaleAISenvSS19, GPSwhaleAISenvSS20, GPSwhaleAISenvSS21, GPSwhaleAISenvSS22,
                   GPSwhaleAISenvSS23)
save(allDrifts, file = "data/allDrifts.rda")
```

### Third octave levels histograms
```{r each-drift-histogram}
# create list of fulltrack dfs from global environment to loop through
dlist <- grep(pattern = "GPSwhaleAISenvSS", names(.GlobalEnv), value = T)

# all Drifts - all TOL bands, facet wrap
for (ft in dlist) {
pivot1 <- get(ft) %>%
  pivot_longer(cols = starts_with("TOL"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid

print(ggplot(data = pivot1, aes(Hz, fill = factor(Wpresence))) +
  geom_histogram(binwidth = 1, color = "black") +
  theme_classic() +
  ggtitle(paste0("Distribution of TOL intensities: Drift", substr(ft,17,18))) +
  labs(x = "Sound Intensity (dB)") +
  guides(fill = guide_legend(title = "Whale Presence")) +
  scale_fill_manual(values = c("0" = "bisque1",
                               "1" = "brown4")) +
  theme(legend.position = c(0.5,0.8)) +
  (facet_grid(~factor(SL, levels = c('TOL_125', 'TOL_2000')))))
}
```

```{r all-drifts-histogram}
# Single plot for ALL DRIFTS
pivot3 <- allDrifts %>%
  # Pivot longer for histogram facet grid
  pivot_longer(cols = starts_with("TOL"), names_to = "SL", values_to = "Hz")

ggplot(data = pivot3, aes(Hz, fill = factor(Wpresence))) +
  geom_histogram(binwidth = 2, color = "black") +
  theme_classic() +
  ggtitle("Distribution of third octave level intensities: All Drifts Combined") +
  labs(x = "Sound Intensity (dB)") +
  guides(fill = guide_legend(title = "Whale Presence")) +
  scale_fill_manual(values = c("0" = "bisque1",
                               "1" = "brown4")) +
  theme(legend.position = c(0.7,0.8)) +
  facet_grid(~factor(SL, levels = c('TOL_125', 'TOL_2000')))
```

Timeline of drifts in ocean
```{r}
ggplot(data = allDrifts, aes(x = UTC, y = Deployment)) + geom_point(aes(color = factor(Deployment)))
```

Looking at TOl2000 sound levels for each buoy and where Whale Detections happen
```{r fig.width=10}
hourly <- GPSwhaleAISenvSS10 %>% group_by(time = floor_date(UTC, '4 hours')) %>%
  summarize(mean_TOL2000 = mean(TOL_2000),
            sum_AIS = sum(shipin10km),
            sum_Wh = sum(Wpresence))

ggplot(hourly, aes(x = time, y = mean_TOL2000)) + geom_line() + 
  geom_smooth() + 
  geom_point() +
  #geom_point(aes(color = species)) +
  #geom_vline(xintercept = hourly$UTC[hourly$shipin10km == 1], color = "yellow", alpha = 0.5) +
  scale_color_discrete(na.translate=FALSE) #+
 # scale_x_datetime(date_breaks = "1 day",limits = as.POSIXct(c("2018-09-24 00:00:00", "2018-10-03 00:00:00")))
```
# Sound levels across the whole range
```{r fig.width = 10}
ggplot(allDrifts, aes(UTC,TOL_2000)) + geom_point() 
```

# Using rollApply for ZC
```{r}
ZC_filt <- GPSwhaleAISenvSS23 %>% 
  filter(species %in% c("ZC", NA)) %>%
  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA))

ZC_pivot <- pivot_longer(ZC_filt, cols = rollAvg_Whl:rollAvg_125, names_to = "avg", values_to = "value")

#Plot facet grid
ggplot(ZC_pivot, aes(x = UTC, y = value, color = avg)) + geom_smooth() + 
  facet_grid(vars(avg), scales = "free") +
  scale_x_datetime(date_breaks = "3 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Averages of ZC Detections alongside Soundscape Metrics - Buoy 23")
```
# Using rollApply for PM
```{r}
PM_filt <- GPSwhaleAISenvSS23 %>% 
  filter(species %in% c("PM", NA)) %>%
  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA))

PM_pivot <- pivot_longer(PM_filt, cols = rollAvg_Whl:rollAvg_125, names_to = "avg", values_to = "value")

#Plot facet grid
ggplot(PM_pivot, aes(x = UTC, y = value, color = avg)) + geom_smooth() + 
  facet_grid(vars(avg), scales = "free") +
  scale_x_datetime(date_breaks = "3 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Averages of PM Detections alongside Soundscape Metrics - Buoy 23")
```

Compare overall TOL2000sound levels when whales are absent vs. present
```{r}
ggplot(allDrifts, aes(factor(Wpresence), TOL_125)) + geom_boxplot()
```

Scatterplots

```{r}
# SST vs TOL 2000
ggplot(data = allDrifts, aes(x = sst_mean, y = `TOL_2000`)) +
  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)

# SST over time (strange plot)
ggplot(allDrifts, aes(x = UTC, y = sst_mean)) +
  #geom_line() +
  geom_point(aes(color = species)) +
  scale_color_discrete(na.translate=FALSE)
  #geom_smooth(method = "loess")

# Depth vs TOL 125
ggplot(data = allDrifts, aes(x = `TOL_125`, y = depth *(-1))) +
  #geom_line() +
  geom_point(aes(color = species)) +
  scale_color_discrete(na.translate=FALSE)

# Chlorophyll vs TOL 125
ggplot(data = allDrifts, aes(x = chlorophyll_mean, y = `TOL_125`)) +
  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)

# CHl over time
ggplot(allDrifts, aes(x = UTC, y = chlorophyll_mean)) +
  geom_line() +
  geom_point(aes(color = species)) +
  scale_color_discrete(na.translate=FALSE)
  #geom_smooth(method = "loess")

# Mixed Layer Depth vs TOL125 with species
ggplot(data = allDrifts, aes(x = mldDepth, y = `TOL_125`)) +
  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)

# Mixed Layer Depth over time
ggplot(allDrifts, aes(x = UTC, y = mldDepth*(-1))) +
  geom_line() +
  geom_point(aes(color = species)) +
  scale_color_discrete(na.translate=FALSE)

ggplot(data = allDrifts, aes(x = `TOL_125`, y = bathy_slope)) +
  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)

```

Facet Grids for Env. Variables
```{r}
#Get avg daily environmental variables
means <- allDrifts %>%
  group_by(as.Date(UTC))%>%
  summarize(sst_mean = mean(sst_mean),
            dist_slope = mean(dist2slope),
            depth = mean(depth),
            slope = mean(bathy_slope),
            dist_shore = mean(dist2shore),
            chlorophyll_mean = mean(chlorophyll_mean),
            ssh_mean = mean(ssh_mean),
            mldDepth = mean(mldDepth),
            mldTemp = mean(mldTemp),
            ttDepth = mean(ttDepth),
            ttTemp = mean(ttTemp),
            temp400 = mean(temp400),
            sal400 = mean(sal400),
            curlmean = mean(curl_mean),
            Wh_presence = sum(Wpresence))
View(means)
```

```{r}
#For surface/depth variables USING DAILY MEAN VARIABLES
trackLong <- means %>%
  pivot_longer(cols = sst_mean:ssh_mean,
               names_to="parameter",
               values_to="value") %>%
  filter(parameter %in% c("sst_mean", "dist_shore", "chlorophyll_mean", "ssh_mean")) %>%
  rename(Date = `as.Date(UTC)`)

p <- ggplot(data = trackLong, aes(x = Date, y = value)) +
  geom_line()

p + facet_grid(parameter ~ ., scales = "free_y")

# For MLD/Thermocline variables 
trackLong2 <- allDrifts %>%
  pivot_longer(cols = mldDepth:sal400,
               names_to="parameter",
               values_to="value") %>%
  filter(parameter %in% c("mldDepth", "mldTemp", "ttDepth", "ttTemp", "temp400", "sal400")) %>%
    rename(Date = UTC) %>%
    subset(species!="?BW")

q <- ggplot(data = trackLong2, aes(x = Date, y = value)) +
  geom_line() + geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
q + facet_grid(parameter ~ ., scales = "free_y")

# For temp400 and salinity400 variables
env.labs <- c(`sal400` = "Salinity at 400m", 
              `temp400` = "Temperature at 400m")

trackLong3 <- allDrifts %>%
  pivot_longer(cols = temp400:sal400,
               names_to="parameter",
               values_to="value") %>%
  filter(parameter %in% c("temp400", "sal400")) %>%
  rename(`Date (2018)` = UTC)

q <- ggplot(data = trackLong3, aes(x = `Date (2018)`, y = value)) +
  #geom_line() + 
  geom_point(aes(color = species)) + 
  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43", "BW" = "BW"))
q + facet_grid(parameter ~ ., scales = "free_y", labeller = as_labeller(env.labs))
```

Use as reference for similar plot (potentially??)
Co-occurrence plots of drift 8 and vessels 
```{r}
#### 09/04/18 co-occurrence of vessel and drift within 10 km
#track08_09.04 <- tracks_SS_08 %>%
#  filter(UTC >= "2018-09-04 00:00:00" & UTC <= "2018-09-05 00:00:00")

#ggplot(data = track08_09.04, aes(x = UTC, y = `TOL_125`)) +
#  geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
#  geom_text(aes(as.POSIXct("2018-09-04 14:13:00"), 88, label = "Enter vessel", hjust = 1.0)) +
#  geom_vline(xintercept = as.POSIXct("2018-09-04 15:17:00"), color = "yellow", linewidth = 2) +
#  geom_text(aes(as.POSIXct("2018-09-04 15:13:00"), 87, label = "Exit vessel", hjust = -0.1)) +
#  geom_line() + 
#  geom_point(data = track08_09.04, aes(color = species, size = 2)) +
#  scale_color_discrete(na.translate=FALSE) +
#  scale_size_continuous(guide = "none") +
#  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-04-18\n(Within 7 km)") +
#  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
#  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
```

Density Plots - Species at TOL sound levels
```{r density-plots}
# TOL 125
for (ft in dlist) {
plt <- get(ft) %>%
  filter(!is.na(species))

  print(ggplot(data = plt, aes(TOL_125, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle(paste0("TOL 125, Drift ", substr(ft, 17, 18))) +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
}

# TOL 125 SINGLE PLOT ALL DRIFTS
allDrifts %>%
  filter(!is.na(species)) %>%
  ggplot(aes(TOL_125, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle("TOL 125 ALL Drifts Combined") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))

# TOL 2000
for (ft in dlist) {
plt <- get(ft) %>%
  filter(!is.na(species))

  print(ggplot(data = plt, aes(TOL_2000, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle(paste0("TOL 2000, Drift ", substr(ft, 17, 18))) +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
}

# TOL 2000 SINGLE PLOT ALL DRIFTS
allDrifts %>%
  filter(!is.na(species)) %>%
  ggplot(aes(TOL_2000, fill = species)) + 
  geom_density(alpha = 0.5) +
  ggtitle("TOL 2000 ALL Drifts Combined") +
  scale_x_continuous(name = "Sound Intensity (dB)") +
  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
```

Visualize SS for buoy 20 when GPS is malfunctioning
```{r}
# Visualize where detections are missing
GPSwhaleAISenvSS20$missingGPS <-ifelse(is.na(GPSwhaleAISenvSS20$Latitude),
                                       "yes", "no")
# Need to create rollnoise column
#ggplot(GPSwhaleAISenvSS20, aes(UTC, TOL_125)) + 
#  geom_point(aes(color = missingGPS)) + geom_line(aes(y=rollnoise))

```

Does time elapsed affect whale presence?

Linear model
```{r}
model6 <- lm(Wpresence>0 ~ timeAfterShip, data = allDrifts)
summary(model6)
plot(model6)
```

Avg. number of hours after ship encounter?
```{r}
mean(allDrifts$timeAfterShip) / 60 / 60
```

## Soundscape before, during, after ship passage?
Elevated TOL_125 sound levels with ship presence, only slight difference with TOL_2000
```{r}
# During
ship_event <- allDrifts %>% filter(shipin10km == 1)
no_ship <- allDrifts %>% filter(shipin10km == 0)

# Plot TOL 125
ggplot(allDrifts, aes(factor(shipin10km), TOL_125)) + geom_boxplot() + labs(x = "Ship Present?")
summary(ship_event$TOL_125)
summary(no_ship$TOL_125)

# Plot TOL_2000
ggplot(allDrifts, aes(factor(shipin10km), TOL_2000)) + geom_boxplot() + labs(x = "Ship Present?")
summary(ship_event$TOL_2000)
summary(no_ship$TOL_2000)

```

How often are there detections immediately following ship passage?
```{r}
# Filter by whale detections and mutate to hrs + days after ship event
detections <- allDrifts %>% 
  filter(Wpresence == 1) %>%
  mutate(hrsAfterShip = timeAfterShip/3600) %>%
  mutate(daysAfterShip = hrsAfterShip/24)

# Plot number of days after ship event by species
ggplot(detections, aes(species, daysAfterShip)) + geom_boxplot()
summary(detections$daysAfterShip) # Avg 5.4 days after ship
```

Visualize whale detections and ship events
```{r fig.width=10}
detsOnly <- GPSwhaleAISenvSS08 %>% filter(Wpresence == 1)
shipOnly <- GPSwhaleAISenvSS08 %>% filter(shipin10km == 1)

ggplot(GPSwhaleAISenvSS08) +
  geom_bar(shipOnly, mapping = aes(x = UTC, color = "shipin10km")) +
  geom_bar(detsOnly, mapping = aes(x = UTC, color = "Wpresence")) +
  scale_x_datetime(date_breaks = "2 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

sppcolor <- factor(detsOnly$species)

# Version 1 - too busy??
ggplot(GPSwhaleAISenvSS08) +
  geom_point(aes(x = UTC, y = TOL_125), alpha = 0.5) +
  geom_smooth(aes(x = UTC, y = TOL_125)) +
  scale_x_datetime(date_breaks = "2 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_vline(xintercept = shipOnly$UTC, color = "green", linewidth = 1) +
  geom_vline(xintercept = detsOnly$UTC, color = "purple", linewidth = 1)

# Version 2 -  Visualize sound levels when whales present by buoy
detsOnly2 <- GPSwhaleAISenvSS08 %>% filter(Wpresence == 1)
shipOnly2 <- GPSwhaleAISenvSS08 %>% filter(shipin10km == 1)
ggplot(detsOnly2) +
  geom_vline(xintercept = shipOnly2$UTC, color = "green") +  # add line for ship presence
  geom_point(aes(x = UTC, y = TOL_2000, color = species), size = 2) +
  scale_x_datetime(date_breaks = "2 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Version 3 - presence/absence sound levels
ggplot(GPSwhaleAISenvSS08) +
  geom_point(aes(x = UTC, y = TOL_125, color = factor(species)))

# All buoys
detsOnly3 <- allDrifts %>% filter(Wpresence == 1)
shipOnly3 <- allDrifts %>% filter(shipin10km == 1)
ggplot(detsOnly3) +
  geom_vline(xintercept = shipOnly3$UTC, color = "green") +  # add line for ship presence
  geom_point(aes(x = UTC, y = TOL_125, color = species), size = 2) +
  scale_x_datetime(date_breaks = "2 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#Sound levels for all species
```{r}
# Sound levels for all species
ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_125)) +
  labs(title = "TOL 125 Sound levels by Species - All Buoys")
ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_2000)) +
    labs(title = "TOL 2000 Sound levels by Species - All Buoys")
```

Visualize whale detections and ship events with plotPresBar
```{r}
detsOnly <- GPSwhaleAISenvSS07 %>% filter(Wpresence == 1)
shipOnly <- GPSwhaleAISenvSS07 %>% filter(shipin10km == 1)

PAMmisc::plotPresBar(detsOnly, bin = 'hour', by = "species")

PAMmisc::plotPresBar(shipOnly, bin='hour')

```

Bin time elapsed after ship event into categories and plot
```{r}
# Drift 20
example_drift20 <- GPSwhaleAISenvSS20 %>%
  .[!is.na(GPSwhaleAISenvSS20$timeAfterShip),] %>%
  mutate(timeAfterShipCat = cut(timeAfterShip, breaks = c(0, 43200, 129600, 216000, max(timeAfterShip)),
         labels = c("12 hr", "36 hr", "60 hr", "over 60 hr"))) %>%    # cuts 12, 36, 60 hrs
  filter(!is.na(species))
View(example_drift20)

# Plot
ggplot(example_drift20, aes(x=species, fill=timeAfterShipCat)) + geom_bar(position="dodge")

ggplot(example_drift20) + geom_histogram(aes(x = timeAfterShip)) + facet_grid(species ~.)
#ggplot(example_drift10) + geom_histogram(aes(x = timeAfterShip)) + facet_grid(species ~.)
```

#Hours After Ship Event by Species
```{r}
no_na <- allDrifts %>% filter(Wpresence == 1) %>%
  mutate(HrsafterShip = timeAfterShip /3600)
ggplot(no_na) + geom_histogram(aes(x = HrsafterShip), bins = 30) + facet_grid(species ~.) +
  labs(title = "Hours Elapsed from Ship Event by Species - All buoys")
```

Same as above with different drift
```{r}
# Drift 10

# Find index of first ship
whereShips <- GPSwhaleAISenvSS10$shipin10km > 0
firstship <- min(which(whereShips)) # Index of the first ship

# Replace with NA where index is less than index of min value
GPSwhaleAISenvSS10[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
GPSwhaleAISenvSS10[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA

example_drift10 <- GPSwhaleAISenvSS10 %>%
  .[!is.na(GPSwhaleAISenvSS10$timeAfterShip),] %>%
  mutate(timeAfterShipCat = cut(timeAfterShip, breaks = c(0, 43200, 129600, 216000, max(timeAfterShip)),
         labels = c("12 hr", "36 hr", "60 hr", "over 60 hr"))) %>%    # cuts 12, 36, 60 hrs
  filter(!is.na(species))
#View(example_drift10)

ggplot(example_drift10, aes(x=species, fill=timeAfterShipCat)) + geom_bar(position="dodge")
```

### 2023-10-16 04_analysis.Rmd
Update 04_analysis.Rmd

--- 2023-06-29 04_analysis.Rmd
@@
 Required Packages
-```{r}
+```{r setup}
 library(tidyverse)
@@
 
-Read in RDS files of each buoy + all environ variables + AIS (pending)
+Read in RDS files of each buoy + all environ variables + AIS
 ```{r}
-ftlist <- dir("data/", recursive=TRUE, full.names=TRUE, pattern="fulltrack")
+ftlist <- dir("D:/Buoy_dfs/", recursive=TRUE, full.names=TRUE, pattern="GPSwhaleAISenvSS")
 
@@
   df <- readRDS(ft)
-  assign(paste0("fulltrack", substr(ft, 15, 16)), df)
+  assign(paste0("GPSwhaleAISenvSS", substr(ft, 29, 30)), df)
 }
-
 ```
@@
 ```{r}
-allDrifts <- rbind(fulltrack07, fulltrack08, fulltrack10, fulltrack12, fulltrack13, fulltrack14, 
-                   fulltrack16, fulltrack18, fulltrack19, fulltrack20, fulltrack21, fulltrack22, fulltrack23)
+dlist <- grep(pattern = "GPSwhaleAISenvSS", names(.GlobalEnv), value = T)
+
+# Create a new column that specifies the drift number before the bind.
+for (col in dlist){
+  data <- get(col) %>% mutate(Deployment = substr(col,17,18))
+  assign(paste0("GPSwhaleAISenvSS", substr(col,17,18)), data)
+}
+
+# Bind and save allDrifts dataframe
+allDrifts <- rbind(GPSwhaleAISenvSS07, GPSwhaleAISenvSS08, GPSwhaleAISenvSS10, GPSwhaleAISenvSS12,
+                   GPSwhaleAISenvSS13, GPSwhaleAISenvSS14, GPSwhaleAISenvSS16, GPSwhaleAISenvSS18,
+                   GPSwhaleAISenvSS19, GPSwhaleAISenvSS20, GPSwhaleAISenvSS21, GPSwhaleAISenvSS22,
+                   GPSwhaleAISenvSS23)
+save(allDrifts, file = "data/allDrifts.rda")
 ```
@@
 ### Third octave levels histograms
-
-```{r all-drifts-all-freq-histogram}
+```{r each-drift-histogram}
 # create list of fulltrack dfs from global environment to loop through
-dlist <- grep(pattern = "fulltrack", names(.GlobalEnv), value = T)
+dlist <- grep(pattern = "GPSwhaleAISenvSS", names(.GlobalEnv), value = T)
 
@@
 pivot1 <- get(ft) %>%
-  pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
+  pivot_longer(cols = starts_with("TOL"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
 
-print(ggplot(data = pivot1, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
+print(ggplot(data = pivot1, aes(Hz, fill = factor(Wpresence))) +
+  geom_histogram(binwidth = 1, color = "black") +
   theme_classic() +
-  ggtitle(paste0("Distribution of third octave level intensities: Drift ", substr(ft,10,11))) +
+  ggtitle(paste0("Distribution of TOL intensities: Drift", substr(ft,17,18))) +
   labs(x = "Sound Intensity (dB)") +
@@
                                "1" = "brown4")) +
-  theme(legend.position = c(0.7,0.8)) +
-  (facet_grid(~factor(SL, levels = c('BB_20.24000', 'TOL_63', 'TOL_125', 'TOL_2000', 'TOL_5000', 'TOL_20000')))))
+  theme(legend.position = c(0.5,0.8)) +
+  (facet_grid(~factor(SL, levels = c('TOL_125', 'TOL_2000')))))
 }
+```
 
+```{r all-drifts-histogram}
 # Single plot for ALL DRIFTS
 pivot3 <- allDrifts %>%
-  pivot_longer(cols = starts_with("TOL") | starts_with("BB"), names_to = "SL", values_to = "Hz")    # Pivot longer for histogram facet grid
+  # Pivot longer for histogram facet grid
+  pivot_longer(cols = starts_with("TOL"), names_to = "SL", values_to = "Hz")
 
-#SL$size_f <- factor(SL$ , levels = c('BB_20.2400', 'TOL_63', 'TOL_125', 'TOL_2000', 'TOL_5000', 'TOL_20000'))))
-
-ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 3, color = "black") +
+ggplot(data = pivot3, aes(Hz, fill = factor(Wpresence))) +
+  geom_histogram(binwidth = 2, color = "black") +
   theme_classic() +
@@
   theme(legend.position = c(0.7,0.8)) +
-  facet_grid(~factor(SL, levels = c('BB_20.24000', 'TOL_63', 'TOL_125', 'TOL_2000', 'TOL_5000', 'TOL_20000')))
-
-```
-
-## ignore??
-```{r drift-8-histogram}
-#ggplot(data = pivot3, aes(Hz, fill = factor(BWpresence))) +
-#  geom_histogram(binwidth = 1, color = "black") +
-#  theme_classic() +
-#  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: Buoy 8") +
-#  scale_x_continuous(name = "Sound Intensity (dB)") +
-#  scale_fill_manual(name = "Beaked Whale Presence", 
-#                    labels = c("0" = "Absent", "1" = "Present"),
-#                    values = c("0" = "bisque1",
-#                               "1" = "brown4")) +
-#  theme(legend.position = c(0.6,0.7)) +
-#  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
-```
-
-```{r all-drifts-two-freq-histogram}
-# All Drifts - TOL bands 125 and 2000 only
-
-TOL.labs <- c(`TOL_125` = "TOL at 125 Hz", 
-              `TOL_2000` = "TOL at 2000 Hz") #creates label names for plots below
-
-for (ft in dlist) {
-pivot2 <- get(ft) %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-print(ggplot(data = pivot2, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle(paste0("Distribution of Third-Octave Level (TOL) Intensities: Drift ", substr(ft,10,11))) +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs)))
-}
-
-# Single plot for ALL DRIFTS combined
-pivot4 <- allDrifts %>%
-  dplyr::select(-TOL_63, -TOL_20000, -TOL_5000) %>%
-  pivot_longer(cols = starts_with("TOL"), names_to = "TOL", values_to = "Hz")    # Pivot longer for histogram facet grid
-
-ggplot(data = pivot4, aes(Hz, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black") +
-  theme_classic() +
-  ggtitle("Distribution of Third-Octave Level (TOL) Intensities: All Drifts Combined") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_manual(name = "Whale Presence", 
-                    labels = c("0" = "Absent", "1" = "Present"),
-                    values = c("0" = "bisque1",
-                               "1" = "brown4")) +
-  theme(legend.position = c(0.6,0.7)) +
-  facet_wrap(~TOL, labeller = as_labeller(TOL.labs))
+  facet_grid(~factor(SL, levels = c('TOL_125', 'TOL_2000')))
 ```
@@
 ```{r}
-ggplot(data = allDrifts, aes(x = UTC, y = station*(-1))) + geom_point(aes(color = factor(station)))
-
-#Note the break in drift 20 - I think it wasn't transmitting for several days (11/15-11/19)
+ggplot(data = allDrifts, aes(x = UTC, y = Deployment)) + geom_point(aes(color = factor(Deployment)))
 ```
 
-Avg sound levels by band (scatterplot)
-```{r}
-avgdB <- TO_23 %>%
-  summarize(TOL_25 = mean(TOL_25),
-            TOL_31.5 = mean(TOL_31.5),
-            TOL_40 = mean(TOL_40),
-            TOL_50 = mean(TOL_50),
-            TOL_63 = mean(TOL_63),
-            TOL_80 = mean(TOL_80),
-            TOL_100 = mean(TOL_100),
-            TOL_125 = mean(TOL_125),
-            TOL_160 = mean(TOL_160),
-            TOL_200 = mean(TOL_200),
-            TOL_250 = mean(TOL_250),
-            TOL_315 = mean(TOL_315),
-            TOL_400 = mean(TOL_400),
-            TOL_500 = mean(TOL_500),
-            TOL_630 = mean(TOL_630),
-            TOL_800 = mean(TOL_800),
-            TOL_1000 = mean(TOL_1000),
-            TOL_1250 = mean(TOL_1250),
-            TOL_1600 = mean(TOL_1600),
-            TOL_2000 = mean(TOL_2000),
-            TOL_2500 = mean(TOL_2500),
-            TOL_3150 = mean(TOL_3150),
-            TOL_4000 = mean(TOL_4000),
-            TOL_5000 = mean(TOL_5000),
-            TOL_6300 = mean(TOL_6300),
-            TOL_8000 = mean(TOL_8000),
-            TOL_10000 = mean(TOL_10000),
-            TOL_12500 = mean(TOL_12500),
-            TOL_16000 = mean(TOL_16000),
-            TOL_20000 = mean(TOL_20000)) %>%
-  pivot_longer(cols = TOL_25:TOL_20000,
-               names_to = "frequency",
-               values_to="dB")
+Looking at TOl2000 sound levels for each buoy and where Whale Detections happen
+```{r fig.width=10}
+hourly <- GPSwhaleAISenvSS10 %>% group_by(time = floor_date(UTC, '4 hours')) %>%
+  summarize(mean_TOL2000 = mean(TOL_2000),
+            sum_AIS = sum(shipin10km),
+            sum_Wh = sum(Wpresence))
 
-ggplot(avgdB, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
+ZC_filt <- GPSwhaleAISenvSS10 %>% 
+  filter(species %in% c("ZC", NA)) %>%
+  mutate(rollAvg = zoo::rollapply(Wpresence, 6, mean, fill = NA))
+
+ggplot(hourly, aes(x = time, y = mean_TOL2000)) + geom_line() + 
+  geom_smooth() + 
   geom_point() +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  #geom_point(aes(color = species)) +
+  #geom_vline(xintercept = hourly$UTC[hourly$shipin10km == 1], color = "yellow", alpha = 0.5) +
+  scale_color_discrete(na.translate=FALSE) #+
+ # scale_x_datetime(date_breaks = "1 day",limits = as.POSIXct(c("2018-09-24 00:00:00", "2018-10-03 00:00:00")))
+
 ```
 
-Box plots
+Compare overall TOL2000sound levels when whales are absent vs. present
 ```{r}
-TO_longer <- pivot_longer(TO_10, cols = TOL_25:TOL_20000,
-               names_to = "frequency",
-               values_to="dB")
-ggplot(TO_longer, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
-  geom_boxplot() +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-
-# Violin plot
-ggplot(TO_longer, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
-  geom_violin() +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+ggplot(allDrifts, aes(factor(Wpresence), TOL_2000)) + geom_boxplot()
 ```
@@
 ggplot(allDrifts, aes(x = UTC, y = sst_mean)) +
-  geom_line() +
+  #geom_line() +
   geom_point(aes(color = species)) +
@@
 # Depth vs TOL 125
-ggplot(data = allDrifts, aes(x = depth, y = `TOL_125`)) +
-  geom_line() +
+ggplot(data = allDrifts, aes(x = `TOL_125`, y = depth *(-1))) +
+  #geom_line() +
   geom_point(aes(color = species)) +
   scale_color_discrete(na.translate=FALSE)
-
-#Mean depth over time
-#ggplot(means_08, aes(x = `as.Date(UTC)`, y = depth*(-1))) +
-#  geom_line()
 
@@
             curlmean = mean(curl_mean),
-            Wh_presence = sum(BWpresence))
+            Wh_presence = sum(Wpresence))
+View(means)
 ```
@@
 
-NEEDS FIXING/LOOP/AIS DATA FOR OTHER DRIFTS
+Use as reference for similar plot (potentially??)
 Co-occurrence plots of drift 8 and vessels 
@@
 #### 09/04/18 co-occurrence of vessel and drift within 10 km
-track08_09.04 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-04 00:00:00" & UTC <= "2018-09-05 00:00:00")
+#track08_09.04 <- tracks_SS_08 %>%
+#  filter(UTC >= "2018-09-04 00:00:00" & UTC <= "2018-09-05 00:00:00")
 
-ggplot(data = track08_09.04, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
-  geom_text(aes(as.POSIXct("2018-09-04 14:13:00"), 88, label = "Enter vessel", hjust = 1.0)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-04 15:17:00"), color = "yellow", linewidth = 2) +
-  geom_text(aes(as.POSIXct("2018-09-04 15:13:00"), 87, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.04, aes(color = species, size = 2)) +
-  scale_color_discrete(na.translate=FALSE) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-04-18\n(Within 7 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/05/18 co-occurrence of vessel and drift within 10 km.
-track08_09.05 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-04 21:00:00" & UTC <= "2018-09-06 00:00:00")
-
-ggplot(data = track08_09.05, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-05 01:33:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-05 01:33:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-05 02:08:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-05 02:08:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.05, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-05-18\n(Within 5 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/10/18 co-occurrence of vessel and drift within 10 km.
-track08_09.10 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-10 00:00:00" & UTC <= "2018-09-11 00:00:00")
-
-ggplot(data = track08_09.10, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-10 12:58:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-10 12:58:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-10 14:01:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-10 14:01:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() +
-  geom_point(data = track08_09.10, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-10-18\n(Within 2 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/22/18 co-occurrence of vessel and drift within 10 km
-track08_09.22 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-22 00:00:00" & UTC <= "2018-09-23 00:00:00")
-
-ggplot(data = track08_09.22, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-22 19:13:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-22 19:13:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-22 19:34:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-22 19:34:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() + 
-  geom_point(data = track08_09.22, aes(color = species, size = 2)) +
-  scale_size_continuous(guide = "none") +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-22-18\n(Within 9 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-#### 09/23/18 co-occurrence of vessel and drift within 10 km. NO whale detections in this time period.
-track08_09.23 <- tracks_SS_08 %>%
-  filter(UTC >= "2018-09-23 00:00:00" & UTC <= "2018-09-24 00:00:00")
-
-ggplot(data = track08_09.23, aes(x = UTC, y = `TOL_125`)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-23 03:55:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-23 03:55:00"), 86, label = "Enter vessel", hjust = 1)) +
-  geom_vline(xintercept = as.POSIXct("2018-09-23 04:23:00"), color = "yellow", linewidth = 1) +
-  geom_text(aes(as.POSIXct("2018-09-23 04:23:00"), 85, label = "Exit vessel", hjust = -0.1)) +
-  geom_line() +
-  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-23-18\n(Within 6 km)") +
-  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
-  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-```
-
-IS THIS HELPFUL??
-Whale Occurrences
-```{r}
-# Filter Drift 8 by Whale presence and absence
-W1 <- tracks_SS_08 %>%
-  filter(Wpresence == 1)
-W0 <- tracks_SS_08 %>%
-  filter(Wpresence == 0)
-
-ggplot(data = tracks_SS_08, mapping = aes(TOL_125)) + geom_histogram(binwidth = 0.1, color = "black")
-
-ggplot() + 
-  geom_histogram(data = W0, aes(TOL_5000), binwidth = 0.5, fill = "black", color = "white") +
-  geom_histogram(data = W1, aes(TOL_5000), binwidth = 0.5, fill = "yellow", color = "black")
-
-# Percent of time BWs are detected
-print(paste(nrow(W1) / nrow(tracks_SS_08) * 100,"%"))
-
-# Most abundant whale species when whale is detected (remove NA values)
-allDrifts %>%
-  filter(!is.na(species)) %>%
-  ggplot() + geom_bar(aes(species))
-
-# Sound levels whales are exposed to by species
-ZC <- allDrifts %>%
-  filter(species == "ZC")
-
-BW43 <- allDrifts %>%
-  filter(species == "BW43")
-
-BB <- allDrifts %>%
-  filter(species == "BB")
-
-BWC <- allDrifts %>%
-  filter(species == "BWC")
-
-BW37V <- allDrifts %>%
-  filter(species == "BW37V")
-
-PM <- allDrifts %>%
-  filter(species == "PM")
-
-print(max(ZC$TOL_125))
-print(max(BW43$TOL_125))
-print(max(BB$TOL_125))
-print(max(BWC$TOL_125))
-print(max(BW37V$TOL_125))
-print(max(PM$TOL_125))
-
-ggplot() + geom_histogram(data = ZC, mapping = aes(TOL_125), binwidth = 0.2, fill = "brown", color = "white") +
-  geom_histogram(data = BW43, mapping = aes(TOL_125), binwidth = 0.2, fill = "yellow", color = "white") +
-  theme_classic()
+#ggplot(data = track08_09.04, aes(x = UTC, y = `TOL_125`)) +
+#  geom_vline(xintercept = as.POSIXct("2018-09-04 14:13:00"), color = "yellow", linewidth = 2) +
+#  geom_text(aes(as.POSIXct("2018-09-04 14:13:00"), 88, label = "Enter vessel", hjust = 1.0)) +
+#  geom_vline(xintercept = as.POSIXct("2018-09-04 15:17:00"), color = "yellow", linewidth = 2) +
+#  geom_text(aes(as.POSIXct("2018-09-04 15:13:00"), 87, label = "Exit vessel", hjust = -0.1)) +
+#  geom_line() + 
+#  geom_point(data = track08_09.04, aes(color = species, size = 2)) +
+#  scale_color_discrete(na.translate=FALSE) +
+#  scale_size_continuous(guide = "none") +
+#  ggtitle("Sound levels during co-occurrence of Drift 8 and a vessel on 09-04-18\n(Within 7 km)") +
+#  scale_y_continuous(name = "Sound Intensity (dB) at TOL 125 Hz") +
+#  scale_color_discrete(na.translate=FALSE, labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
 ```
@@
 ```{r density-plots}
-# Broadband
-for (ft in dlist) {
-plt <- get(ft) %>%
-  filter(!is.na(species))
-
-  print(ggplot(data = plt, aes(`BB_20.24000`, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle(paste0("Broadband Drift ", substr(ft, 10, 11))) +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
-}
-
-# Broadband SINGLE PLOT ALL DRIFTS
-allDrifts %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(`BB_20.24000`, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("Broadband ALL Drifts Combined") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 63
-for (ft in dlist) {
-plt <- get(ft) %>%
-  filter(!is.na(species))
-
-  print(ggplot(data = plt, aes(TOL_63, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle(paste0("TOL 63, Drift ", substr(ft, 10, 11))) +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
-}
-
-# TOL 63 SINGLE PLOT ALL DRIFTS
-allDrifts %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_63, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("TOL 63 ALL Drifts Combined") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
 # TOL 125
@@
   geom_density(alpha = 0.5) +
-  ggtitle(paste0("TOL 125, Drift ", substr(ft, 10, 11))) +
+  ggtitle(paste0("TOL 125, Drift ", substr(ft, 17, 18))) +
   scale_x_continuous(name = "Sound Intensity (dB)") +
@@
   geom_density(alpha = 0.5) +
-  ggtitle(paste0("TOL 2000, Drift ", substr(ft, 10, 11))) +
+  ggtitle(paste0("TOL 2000, Drift ", substr(ft, 17, 18))) +
   scale_x_continuous(name = "Sound Intensity (dB)") +
@@
   scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 5000
-for (ft in dlist) {
-plt <- get(ft) %>%
-  filter(!is.na(species))
-
-  print(ggplot(data = plt, aes(TOL_5000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle(paste0("TOL 5000, Drift ", substr(ft, 10, 11))) +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
-}
-
-# TOL 5000 SINGLE PLOT ALL DRIFTS
-allDrifts %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_5000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("TOL 5000 ALL Drifts Combined") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
-
-# TOL 20,000
-for (ft in dlist) {
-plt <- get(ft) %>%
-  filter(!is.na(species))
-
-  print(ggplot(data = plt, aes(TOL_20000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle(paste0("TOL 20000, Drift ", substr(ft, 10, 11))) +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43")))
-}
-
-# TOL 20,000 SINGLE PLOT ALL DRIFTS
-allDrifts %>%
-  filter(!is.na(species)) %>%
-  ggplot(aes(TOL_20000, fill = species)) + 
-  geom_density(alpha = 0.5) +
-  ggtitle("TOL 20,000 ALL Drifts Combined") +
-  scale_x_continuous(name = "Sound Intensity (dB)") +
-  scale_fill_discrete(labels = c("ZC" = "Cuvier's", "BW43" = "BW43"))
 ```
 
-ARE THESE HELPFUL??
-Histograms with whale presence/absence and environmental variables. Doesnt differentiate by species though
+Visualize SS for buoy 20 when GPS is malfunctioning
 ```{r}
-# SST
-ggplot(data = allDrifts, aes(sst_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.1, color = "black")
+# Visualize where detections are missing
+GPSwhaleAISenvSS20$missingGPS <-ifelse(is.na(GPSwhaleAISenvSS20$Latitude),
+                                       "yes", "no")
+# Need to create rollnoise column
+#ggplot(GPSwhaleAISenvSS20, aes(UTC, TOL_125)) + 
+#  geom_point(aes(color = missingGPS)) + geom_line(aes(y=rollnoise))
 
-# Depth
-ggplot(data = allDrifts, aes(depth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 100, color = "black")
+```
 
-# Bottom slope
-ggplot(data = allDrifts, aes(bathy_slope, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 1, color = "black")
+Does time elapsed affect whale presence?
 
-# SSH
-ggplot(data = allDrifts, aes(ssh_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
+Linear model
+```{r}
+model6 <- lm(Wpresence>0 ~ timeAfterShip, data = allDrifts)
+summary(model6)
+plot(model6)
+```
 
-# Chlorophyll
-ggplot(data = allDrifts, aes(chlorophyll_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.05, color = "black")
+Avg. number of hours after ship encounter?
+```{r}
+mean(allDrifts$timeAfterShip) / 60 / 60
+```
 
-#Wind Curl
-ggplot(data = allDrifts, aes(curl_mean, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.0000005, color = "black")
+## Soundscape before, during, after ship passage?
+Elevated TOL_125 sound levels with ship presence, only slight difference with TOL_2000
+```{r}
+# During
+ship_event <- allDrifts %>% filter(shipin10km == 1)
+no_ship <- allDrifts %>% filter(shipin10km == 0)
 
-# Mixed layer depth Depth
-ggplot(data = allDrifts, aes(mldDepth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.25, color = "black")
+# Plot TOL 125
+ggplot(allDrifts, aes(factor(shipin10km), TOL_125)) + geom_boxplot() + labs(x = "Ship Present?")
+summary(ship_event$TOL_125)
+summary(no_ship$TOL_125)
 
-# Mixed layer depth Temperature
-ggplot(data = allDrifts, aes(mldTemp, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.1, color = "black")
+# Plot TOL_2000
+ggplot(allDrifts, aes(factor(shipin10km), TOL_2000)) + geom_boxplot() + labs(x = "Ship Present?")
+summary(ship_event$TOL_2000)
+summary(no_ship$TOL_2000)
 
-# Thermocline Depth
-ggplot(data = allDrifts, aes(ttDepth, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.75, color = "black")
+## How to show BEOFRE AND AFTER???
+# create factors by days. discrete values like 10,000 - 20,000 seconds.
+```
 
-#Thermocline temperature
-ggplot(data = allDrifts, aes(ttTemp, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.1, color = "black")
+How often are there detections immediately following ship passage?
+```{r}
+# Filter by whale detections and mutate to hrs + days after ship event
+detections <- allDrifts %>% 
+  filter(Wpresence == 1) %>%
+  mutate(hrsAfterShip = timeAfterShip/3600) %>%
+  mutate(daysAfterShip = hrsAfterShip/24)
 
-# Temperature at 400m
-ggplot(data = allDrifts, aes(temp400, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.03, color = "black")
+# Plot number of days after ship event by species
+ggplot(detections, aes(daysAfterShip, species)) + geom_boxplot()
+summary(detections$daysAfterShip) # Avg 5.4 days after ship
+```
+Are whales present before a ship passage?
+```{r}
+detect_b4_ship <- allDrifts %>%
+  filter(Wpresence == 1) %>%
+  mutate(hrsBeforeShip = timeBeforeShip/3600) %>%
+  mutate(daysBeforeShip = hrsBeforeShip/24)
 
-# Salinity at 400 m
-ggplot(data = allDrifts, aes(sal400, fill = factor(BWpresence))) +
-  geom_histogram(binwidth = 0.005, color = "black")
+ggplot(detect_b4_ship, aes(daysBeforeShip)) + geom_histogram()
+summary(detect_b4_ship$daysBeforeShip) # Avg 5.9 days before ship
 ```
-### GAMs
+Visualize whale detections and ship events
+```{r fig.width=10}
+detsOnly <- GPSwhaleAISenvSS08 %>% filter(Wpresence == 1)
+shipOnly <- GPSwhaleAISenvSS08 %>% filter(shipin10km == 1)
 
+ggplot(GPSwhaleAISenvSS08) +
+  geom_bar(shipOnly, mapping = aes(x = UTC, color = "shipin10km")) +
+  geom_bar(detsOnly, mapping = aes(x = UTC, color = "Wpresence")) +
+  scale_x_datetime(date_breaks = "2 days") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+sppcolor <- factor(detsOnly$species)
+
+# Version 1 - too busy??
+ggplot(GPSwhaleAISenvSS08) +
+  geom_point(aes(x = UTC, y = TOL_125), alpha = 0.5) +
+  geom_smooth(aes(x = UTC, y = TOL_125)) +
+  scale_x_datetime(date_breaks = "2 days") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+  geom_vline(xintercept = shipOnly$UTC, color = "green", linewidth = 1) +
+  geom_vline(xintercept = detsOnly$UTC, color = "purple", linewidth = 1)
+
+# Version 2 -  Visualize sound levels when whales present by buoy
+detsOnly2 <- GPSwhaleAISenvSS18 %>% filter(Wpresence == 1)
+shipOnly2 <- GPSwhaleAISenvSS20 %>% filter(shipin10km == 1)
+ggplot(detsOnly2) +
+  geom_vline(xintercept = shipOnly2$UTC, color = "green") +  # add line for ship presence
+  geom_point(aes(x = UTC, y = TOL_2000, color = species), size = 2) +
+  scale_x_datetime(date_breaks = "2 days") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+# Version 3 - presence/absence sound levels
+ggplot(GPSwhaleAISenvSS12) +
+  geom_point(aes(x = UTC, y = TOL_125, color = factor(species)))
+
+# All buoys
+detsOnly3 <- allDrifts %>% filter(Wpresence == 1)
+shipOnly3 <- allDrifts %>% filter(shipin10km == 1)
+ggplot(detsOnly3) +
+  geom_vline(xintercept = shipOnly3$UTC, color = "green") +  # add line for ship presence
+  geom_point(aes(x = UTC, y = TOL_125, color = species), size = 2) +
+  scale_x_datetime(date_breaks = "2 days") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+ggplot(detsOnly2) + geom_boxplot(aes(x = species, y = TOL_2000))
+```
+
+Visualize whale detections and ship events with plotPresBar
 ```{r}
-model3 <- gam(BWpresence ~ depth + TOL_125 + dist2shore + mldTemp, 
-              data = allDrifts, family = binomial)
-#plot(model3)
+detsOnly <- GPSwhaleAISenvSS07 %>% filter(Wpresence == 1)
+shipOnly <- GPSwhaleAISenvSS07 %>% filter(shipin10km == 1)
 
-#try fitting linear model
+PAMmisc::plotPresBar(detsOnly, bin = 'hour', by = "species")
 
-lm_depth <- lm(BWpresence ~ depth, data = allDrifts)
-ggplot(allDrifts, aes(sst_mean, chlorophyll_mean)) +
-  geom_point() +
-  geom_smooth()
+PAMmisc::plotPresBar(shipOnly, bin='hour')
+
 ```
-Categorical term
+
+Bin time elapsed after ship event into categories and plot
 ```{r}
-# ALL variables, GAM
-model1 <- gam(BWpresence>0 ~ s(BB_20.24000) + s(TOL_63) + s(TOL_125) + s(TOL_2000)
-              + s(TOL_5000) + s(TOL_20000) + s(depth) + s(mldDepth) + s(Latitude)
-              + s(Longitude) + s(sst_mean) + s(curl_mean) + s(dist2slope) 
-              + s(bathy_slope) + s(dist2shore) + s(chlorophyll_mean) + s(ssh_mean) 
-              + s(mldTemp) + s(ttDepth) + s(ttTemp) + s(temp400) + s(sal400),
-              data = allDrifts, family = binomial, method = "REML")
+# Drift 20
+example_drift20 <- GPSwhaleAIS20 %>%
+  .[!is.na(GPSwhaleAIS20$timeAfterShip),] %>%
+  mutate(timeAfterShipCat = cut(timeAfterShip, breaks = c(0, 43200, 129600, 216000, max(timeAfterShip)),
+         labels = c("12 hr", "36 hr", "60 hr", "over 60 hr"))) %>%    # cuts 12, 36, 60 hrs
+  filter(!is.na(species))
+View(example_drift20)
 
-# ALL variables, GLM
-model1GLM <- glm(BWpresence ~ BB_20.24000 + TOL_63 + TOL_125 + TOL_2000
-              + TOL_5000 + TOL_20000 + depth + mldDepth + Latitude
-              + Longitude + sst_mean + curl_mean + dist2slope
-              + bathy_slope + dist2shore + chlorophyll_mean + ssh_mean 
-              + mldTemp + ttDepth + ttTemp + temp400 + sal400,
-              data = allDrifts, family = binomial)
+# Plot
+ggplot(example_drift20, aes(x=species, fill=timeAfterShipCat)) + geom_bar(position="dodge")
 
-# By species? doesnt work, may need to create dummy variables
-model2 <- gam(BWpresence ~ s(sst_mean, by = as.factor(species)), 
-              data = allDrifts, family = binomial, method = "REML")
+ggplot(example_drift20) + geom_histogram(aes(x = timeAfterShip)) + facet_grid(species ~.)
+ggplot(example_drift10) + geom_histogram(aes(x = timeAfterShip)) + facet_grid(species ~.)
 
-# Soundscape metrics only, GAM
-model4 <- gam(BWpresence ~ s(BB_20.24000, k = 10) + 
-                s(TOL_63, k = 10) + s(TOL_125, k = 10) + 
-                s(TOL_2000, k = 10) + s(TOL_5000, k = 10) 
-              + s(TOL_20000, k = 10),
-              data = allDrifts, family = binomial, method = "REML")
+no_na <- allDrifts %>% filter(Wpresence == 1) %>%
+  mutate(HrsafterShip = timeAfterShip /3600)
+ggplot(no_na) + geom_histogram(aes(x = HrsafterShip)) + facet_grid(species ~.)
+```
 
-# Soundscape metrics only, GLM
-model4GLM <- glm(BWpresence ~ BB_20.24000 + TOL_63 + TOL_125 + TOL_2000
-              + TOL_5000 + TOL_20000,
-              data = allDrifts, family = binomial)
+Same as above with different drift
+```{r}
+# Drift 10
 
-model125 <- gam(BWpresence ~ s(TOL_125, k = 10),
-              data = allDrifts, family = binomial, method = "REML")
-plot(model125, trans = plogis)
-summary(model125)
-gam.check(model125)
+# Find index of first ship
+whereShips <- GPSwhaleAIS10$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
 
-# Environmental var only
-model5 <- gam(BWpresence ~ s(depth) + s(mldDepth) + s(Latitude)
-              + s(Longitude) + s(sst_mean) + s(curl_mean) + s(dist2slope) 
-              + s(bathy_slope) + s(dist2shore) + s(chlorophyll_mean) + s(ssh_mean) 
-              + s(mldTemp) + s(ttDepth) + s(ttTemp) + s(temp400) + s(sal400),
-              data = allDrifts, family = binomial, method = "REML")
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS10[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS10[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
 
-plogis(coef(model1[1]))
-summary(model1)
-summary(model4)
-summary(model5)
-summary(model4GLM)
-plot(model4GLM)
-gam.check(model1)
-gam.check(model4)
+example_drift10 <- GPSwhaleAIS10 %>%
+  .[!is.na(GPSwhaleAIS10$timeAfterShip),] %>%
+  mutate(timeAfterShipCat = cut(timeAfterShip, breaks = c(0, 43200, 129600, 216000, max(timeAfterShip)),
+         labels = c("12 hr", "36 hr", "60 hr", "over 60 hr"))) %>%    # cuts 12, 36, 60 hrs
+  filter(!is.na(species))
+#View(example_drift10)
 
-#Log odds scale (when used with family = binomial)
-plot(model1, pages = 6)
-plot(model4, pages = 4, trans = plogis)
-
-#Probability scale + centered on the intercept + adding intercept-related uncertainty with seWithMean arg
-plot(model1, pages = 14, trans = plogis,
-     shift = coef(model1)[1])#,
-     #seWithMean = TRUE)
-
-plot(model2, pages = 1, trans = plogis,
-     shift = coef(model2)[1])
-
-plot(model4, pages = 6, trans = plogis,
-     shift = coef(model4)[1])
+ggplot(example_drift10, aes(x=species, fill=timeAfterShipCat)) + geom_bar(position="dodge")
 ```

### 2023-10-16 03_environ_var.Rmd
Update 03_environ_var.Rmd

@@
 library(PAMmisc)
-library(rerddap)
 ```
@@
 ```{r}
-env07 <- readRDS(file = "data/GPSwhaleAIS7.rda")
-env08 <- readRDS(file = "data/GPSwhaleAIS8.rda")
-env10 <- readRDS(file = "data/GPSwhaleAIS10.rda")
-env12 <- readRDS(file = "data/GPSwhaleAIS12.rda")
-env13 <- readRDS(file = "data/GPSwhaleAIS13.rda")
-env14 <- readRDS(file = "data/GPSwhaleAIS14.rda")
-env16 <- readRDS(file = "data/GPSwhaleAIS16.rda")
-env18 <- readRDS(file = "data/GPSwhaleAIS18.rda")
-env19 <- readRDS(file = "data/GPSwhaleAIS19.rda")
-env20 <- readRDS(file = "data/GPSwhaleAIS20.rda")
-env21 <- readRDS(file = "data/GPSwhaleAIS21.rda")
-env22 <- readRDS(file = "data/GPSwhaleAIS22.rda")
-env23 <- readRDS(file = "data/GPSwhaleAIS23.rda")
+GPSwhaleAIS7 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS07.rda')
+GPSwhaleAIS8 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS08.rda')
+GPSwhaleAIS10 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS10.rda')
+GPSwhaleAIS12 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS12.rda')
+GPSwhaleAIS13 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS13.rda')
+GPSwhaleAIS14 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS14.rda')
+GPSwhaleAIS16 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS16.rda')
+GPSwhaleAIS18 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS18.rda')
+GPSwhaleAIS19 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS19.rda')
+GPSwhaleAIS20 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS20.rda')
+GPSwhaleAIS21 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS21.rda')
+GPSwhaleAIS22 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS22.rda')
+GPSwhaleAIS23 <- readRDS(file = 'D:/Buoy_dfs/GPSwhaleAIS23.rda')
 ```
@@
 # Downloads data from servers onto external hard drive. Uncomment and only run once.
-#env07 <- matchEnvData(GPSwhaleAIS7, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST07", var = c('analysed_sst'))
-#env08 <- matchEnvData(GPSwhaleAIS8, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST08", var = c('analysed_sst'))
-#env10 <- matchEnvData(GPSwhaleAIS10, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST10", var = c('analysed_sst'))
-#env12 <- matchEnvData(GPSwhaleAIS12, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST12", var = c('analysed_sst'))
-#env13 <- matchEnvData(GPSwhaleAIS13, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST13", var = c('analysed_sst'))
-#env14 <- matchEnvData(GPSwhaleAIS14, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST14", var = c('analysed_sst'))
-#env16 <- matchEnvData(GPSwhaleAIS16, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST16", var = c('analysed_sst'))
-#env18 <- matchEnvData(GPSwhaleAIS18, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST18", var = c('analysed_sst'))
-#env19 <- matchEnvData(GPSwhaleAIS19, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST19", var = c('analysed_sst'))
-#env20 <- matchEnvData(GPSwhaleAIS20, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST20", var = c('analysed_sst'))
-#env21 <- matchEnvData(GPSwhaleAIS21, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST21", var = c('analysed_sst'))
-#env22 <- matchEnvData(GPSwhaleAIS22, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST22", var = c('analysed_sst'))
-#env23 <- matchEnvData(GPSwhaleAIS23, nc = 'jplMURSST41', fileName = "D:/Environ_variables/analysedSST23", var = c('analysed_sst'))
+env07 <- matchEnvData(GPSwhaleAIS7, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST07", var = c('analysed_sst'))
+env08 <- matchEnvData(GPSwhaleAIS8, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST08", var = c('analysed_sst'))
+env10 <- matchEnvData(GPSwhaleAIS10, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST10", var = c('analysed_sst'))
+env12 <- matchEnvData(GPSwhaleAIS12, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST12", var = c('analysed_sst'))
+env13 <- matchEnvData(GPSwhaleAIS13, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST13", var = c('analysed_sst'))
+env14 <- matchEnvData(GPSwhaleAIS14, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST14", var = c('analysed_sst'))
+env16 <- matchEnvData(GPSwhaleAIS16, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST16", var = c('analysed_sst'))
+env18 <- matchEnvData(GPSwhaleAIS18, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST18", var = c('analysed_sst'))
+env19 <- matchEnvData(GPSwhaleAIS19, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST19", var = c('analysed_sst'))
+env20 <- matchEnvData(GPSwhaleAIS20, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST20", var = c('analysed_sst'))
+env21 <- matchEnvData(GPSwhaleAIS21, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST21", var = c('analysed_sst'))
+env22 <- matchEnvData(GPSwhaleAIS22, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST22", var = c('analysed_sst'))
+env23 <- matchEnvData(GPSwhaleAIS23, nc = 'jplMURSST41', 
+                      fileName = "D:/Environ_variables/analysedSST23", var = c('analysed_sst'))
 ```
@@
 ```
-### Wind Stress Curl from ERDDAP using PAMPal function
-Wind Stress, Metop-A ASCAT, 0.25°, Global, Near Real Time, 2009-present (1 Day) 
+
+### Curl of Surface Wind Stress from ERDDAP using PAMPal function
+FNMOC Ekman Transports, 1°, 6-hourly, Lon+/-180, Units: M Pa m-1
 ```{r}
 # Downloads data from servers onto external hard drive. Uncomment and only run once.
-#env07 <- matchEnvData(env07, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress07.nc", var = c('curl'))
-#env08 <- matchEnvData(env08, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress08.nc", var = c('curl'))
-#env10 <- matchEnvData(env10, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress10.nc", var = c('curl'))
-#env12 <- matchEnvData(env12, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress12.nc", var = c('curl'))
-#env13 <- matchEnvData(env13, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress13.nc", var = c('curl'))
-#env14 <- matchEnvData(env14, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress14.nc", var = c('curl'))
-#env16 <- matchEnvData(env16, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress16.nc", var = c('curl'))
-#env18 <- matchEnvData(env18, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress18.nc", var = c('curl'))
-#env19 <- matchEnvData(env19, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress19.nc", var = c('curl'))
-#env20 <- matchEnvData(env20, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress20.nc", var = c('curl'))
-#env21 <- matchEnvData(env21, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress21.nc", var = c('curl'))
-#env22 <- matchEnvData(env22, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress22.nc", var = c('curl'))
-#env23 <- matchEnvData(env23, nc = 'erdQAstress1day', fileName = "D:/Environ_variables/windstress23.nc", var = c('curl'))
+#env07 <- matchEnvData(env07, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress07.nc", var = c('curl'))
+#env08 <- matchEnvData(env08, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress08.nc", var = c('curl'))
+#env10 <- matchEnvData(env10, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress10.nc", var = c('curl'))
+#env12 <- matchEnvData(env12, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress12.nc", var = c('curl'))
+#env13 <- matchEnvData(env13, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress13.nc", var = c('curl'))
+#env14 <- matchEnvData(env14, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress14.nc", var = c('curl'))
+#env16 <- matchEnvData(env16, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress16.nc", var = c('curl'))
+#env18 <- matchEnvData(env18, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress18.nc", var = c('curl'))
+#env19 <- matchEnvData(env19, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress19.nc", var = c('curl'))
+#env20 <- matchEnvData(env20, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress20.nc", var = c('curl'))
+#env21 <- matchEnvData(env21, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress21.nc", var = c('curl'))
+#env22 <- matchEnvData(env22, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress22.nc", var = c('curl'))
+#env23 <- matchEnvData(env23, nc = 'erdlasFnTran6_LonPM180', fileName = "D:/Environ_variables/windstress23.nc", var = c('curl'))
+
 ```
@@
   data <- matchEnvData(get(paste0("env", substr(w, 32, 33))), nc = w) %>%
-     dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean, -matchDepth_mean)   # remove unnecessary columns
+     dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean)   # remove unnecessary columns
 
@@
 ### Distance to continental slope (uses terra and sf package)
+## Takes about 20 minutes to run
 ```{r}
@@
 
+#Make sure to clear Global Environment of other objects with "env"
 tracklist <- grep(pattern = "env", names(.GlobalEnv), value = T)
@@
 }
+```
+```{r} 
+sf <- st_as_sf(get(t), coords = c("Longitude", "Latitude"), crs=4326, remove=F)
+  trk <- terra::rasterize(vect(sf), agg)
+  slop <- terra::distance(trk, vect(cont_slope))
+  ext <- terra::extract(slop, vect(sf))
+
+  dbind <- cbind(sf, ext) %>%
+  dplyr::select(-ID) %>%
+  rename("dist2slope" = last) %>%
+  st_set_geometry(NULL) # remove geometry to prevent buoy dataframe from becoming sf object for future 
 ```
@@
 #env23 <- matchEnvData(env23, nc = 'srtm15plus', fileName = "D:/Environ_variables/depth23.nc", var = c('z'))
+
 ```
@@
 Distance to Shore from ERDDAP using PAMPal function
+Distance to Nearest Coastline: 0.01-Degree Grid
 ```{r}
@@
 #env23 <- matchEnvData(env23, nc = 'dist2coast_1deg', fileName = "D:/Environ_variables/dist2shore23.nc", var = c('dist'))
+
 ```
@@
 present (Monthly Composite)
+Monthly is a better resolution because anything finer leaves NaN in columns
 ```{r}
@@
 #env23 <- matchEnvData(env23, nc = 'erdMH1chlamday', fileName = "D:/Environ_variables/chlorophyll23.nc", var = c('chlorophyll'))
+
 ```
@@
 #env23 <- matchEnvData(env23, nc = 'nesdisSSH1day', fileName = "D:/Environ_variables/ssh23.nc", var = c('sla'))
+
 ```
@@
 
-#### HyCOM Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
+#### Download HYCOM Data - MLDepth, MLTemp, TTemp, TDepth, Salinity
 
@@
 ```{r}
-# Create loop for each buoy to download HYCOM data
-
-## This only downloads 11 of 13 successfully - FIX
+# Create loop for each buoy to download HYCOM data. Takes a long time to run.
 for (t in tracklist) {
-  Hdata <- matchEnvData(as.data.frame(get(paste0("env", substr(t, 4, 5)))), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
+  Hdata <- matchEnvData(as.data.frame(get(paste0("env", substr(t, 4, 5)))), 
+                        nc=PAMmisc::hycomList, 
+                        var=c('water_temp', 'salinity'), 
+                        raw=TRUE, 
+                        depth = 0:1000)
 
@@
 
-# runs hycom variable download and bind for one drift. can ignore if above works.
+# runs hycom variable download and bind for one drift. Run if above does not work.
 ```{r}
-## HYCOM function above stops at buoys 07 and 08, need to manually run it here. FIX!!
-#test08 <- matchEnvData(as.data.frame(tracks_SS_08), nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
+Htest <- matchEnvData(env23, nc=PAMmisc::hycomList, var=c('water_temp', 'salinity'), raw=TRUE, depth = 0:1000)
 
-#thermData08 <- bind_rows(lapply(test08, function(x) {
-#    # do thermo calcs
-#    result <- calcThermoVars(x$matchDepth, x$water_temp)
-#    # get specific values
-#    result$temp400 <- x$water_temp[x$matchDepth == 400]
-#    result$sal400 <- x$salinity[x$matchDepth == 400]
-#    result
-#}))
+thermData <- bind_rows(lapply(Htest, function(x) {
+    # do thermo calcs
+    result <- calcThermoVars(x$matchDepth, x$water_temp)
+    # get specific values
+    result$temp400 <- x$water_temp[x$matchDepth == 400]
+    result$sal400 <- x$salinity[x$matchDepth == 400]
+    result
+}))
 
-#tracks_SS_08 <- cbind(tracks_SS_08, thermData08)
+env23 <- cbind(env23, thermData)
 ```

### 2023-10-16 02_AIS.Rmd
Update 02_AIS.Rmd

@@
 Required Packages
-```{r, }
+```{r setup}
 library(PAMscapes)
 library(tidyverse)
+```
+
+Read in GPS/whale detections
+```{r}
+GPSwhale7 <- readRDS(file = "D:Buoy_dfs/GPSwhale7.rda")
+GPSwhale8 <- readRDS(file = "D:Buoy_dfs/GPSwhale8.rda")
+GPSwhale10 <- readRDS(file = "D:Buoy_dfs/GPSwhale10.rda")
+GPSwhale12 <- readRDS(file = "D:Buoy_dfs/GPSwhale12.rda")
+GPSwhale13 <- readRDS(file = "D:Buoy_dfs/GPSwhale13.rda")
+GPSwhale14 <- readRDS(file = "D:Buoy_dfs/GPSwhale14.rda")
+GPSwhale16 <- readRDS(file = "D:Buoy_dfs/GPSwhale16.rda")
+GPSwhale18 <- readRDS(file = "D:Buoy_dfs/GPSwhale18.rda")
+GPSwhale19 <- readRDS(file = "D:Buoy_dfs/GPSwhale19.rda")
+GPSwhale20 <- readRDS(file = "D:Buoy_dfs/GPSwhale20.rda")
+GPSwhale21 <- readRDS(file = "D:Buoy_dfs/GPSwhale21.rda")
+GPSwhale22 <- readRDS(file = "D:Buoy_dfs/GPSwhale22.rda")
+GPSwhale23 <- readRDS(file = "D:Buoy_dfs/GPSwhale23.rda")
+GPSwhale20.alt <- readRDS(file = "D:Buoy_dfs/GPSwhale20_alt.rda")
 ```
@@
 #marcadFiles23 <- downloadMarCadAIS(GPSwhale23, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
+#marcadFiles20.alt <- downloadMarCadAIS(GPSwhale20.alt, outDir = "D:/AIS", overwrite = FALSE, unzip = TRUE)
 ```
@@
 Now read into R only ais points that come in contact with buoys
+Uncomment and run -- otherwise is saved in D: drive.
 ```{r}
@@
 #   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
+#ais_near_20.alt <- readLocalAIS(GPSwhale20.alt, aisDir="D:/AIS/AIS_West", distance=10e3) %>%
+#   filter(vesselLength >= 100)  #remove vessels under 100m in length
+
 ```
@@
 #saveRDS(ais_near_23, file = "D:AIS/ais_near_23.rda")
+#saveRDS(ais_near_20.alt, file = "D:AIS/ais_near_20_alt.rda")
 ```
@@
 ais_near_23 <- readRDS('D:/AIS/ais_near_23.rda')
+ais_near_20.alt <- readRDS('D:/AIS/ais_near_20_alt.rda')
 ```
@@
 Match AIS points with buoy points
+Warning of collpsing to unique x values.
 ```{r}
@@
 gpsClose23 <- addAIS(GPSwhale23, ais_near_23, interpType = 'close', interpTime = 120)
+gpsClose20.alt <- addAIS(GPSwhale20.alt, ais_near_20.alt, interpType = 'close', interpTime = 120)
 ```
@@
 close2ships23 <- gpsClose23 %>% filter(shipDist < 10e3)
+close2ships20.alt <- gpsClose20.alt %>% filter(shipDist < 10e3)
 
@@
         "close2ships14", "close2ships16", "close2ships18", "close2ships19", "close2ships20",
-        "close2ships22", "close2ships23")
+        "close2ships22", "close2ships23", "close2ships20.alt")
 
@@
 nearbyShp07 <- ShpRows(GPSwhale7, close2ships07)
-GPSwhaleAIS7 <- left_join(GPSwhale7, nearbyShp07, by = join_by("UTC")) %>%
+GPSwhaleAIS07 <- left_join(GPSwhale7, nearbyShp07, by = join_by("UTC")) %>%
   mutate(shipin10km = coalesce(shipin10km, 0))
@@
 nearbyShp08 <- ShpRows(GPSwhale8, close2ships08)
-GPSwhaleAIS8 <- left_join(GPSwhale8, nearbyShp08, by = join_by("UTC")) %>%
+GPSwhaleAIS08 <- left_join(GPSwhale8, nearbyShp08, by = join_by("UTC")) %>%
   mutate(shipin10km = coalesce(shipin10km, 0))
@@
   mutate(shipin10km = coalesce(shipin10km, 0))
+
+nearbyShp20.alt <- ShpRows(GPSwhale20.alt, close2ships20.alt)
+GPSwhaleAIS20.alt <- left_join(GPSwhale20.alt, nearbyShp20.alt, by = join_by("UTC")) %>%
+  mutate(shipin10km = coalesce(shipin10km, 0))
 ```
 
-Save new fulltrack rdas to D drive
+## Add time elapsed after last ship event
+
 ```{r}
-saveRDS(GPSwhaleAIS7, file = "data/GPSwhaleAIS7.rda")
-saveRDS(GPSwhaleAIS8, file = "data/GPSwhaleAIS8.rda")
+shiplist <- c("GPSwhaleAIS10", "GPSwhaleAIS12", "GPSwhaleAIS07", "GPSwhaleAIS08",
+              "GPSwhaleAIS13", "GPSwhaleAIS14", "GPSwhaleAIS16", "GPSwhaleAIS18",
+              "GPSwhaleAIS19", "GPSwhaleAIS20", "GPSwhaleAIS21", "GPSwhaleAIS22",
+              "GPSwhaleAIS23")
+
+for (ship in shiplist) {
+  data <- get(ship) %>%
+   arrange(UTC) %>%
+   mutate(tmpG = cumsum(c(FALSE, as.logical(diff(shipin10km))))) %>%
+   mutate(tmp_a = c(0, diff(UTC)) * !shipin10km) %>%
+   group_by(tmpG) %>%
+   mutate(timeAfterShip = cumsum(tmp_a)) %>%
+   ungroup() %>%
+   select(-c(tmp_a, tmpG))
+
+  assign(paste0("GPSwhaleAIS", substr(ship,12, 13)), data)
+
+}
+```
+
+For buoy 20 alt. only
+```{r}
+GPSwhaleAIS20.alt <- GPSwhaleAIS20.alt %>%
+   arrange(UTC) %>%
+   mutate(tmpG = cumsum(c(FALSE, as.logical(diff(shipin10km))))) %>%
+   mutate(tmp_a = c(0, diff(UTC)) * !shipin10km) %>%
+   group_by(tmpG) %>%
+   mutate(timeAfterShip = cumsum(tmp_a)) %>%
+   ungroup() %>%
+   select(-c(tmp_a, tmpG))
+```
+
+Change times before first ship event to NA
+```{r}
+# Buoy 7
+# Find index of first ship
+whereShips <- GPSwhaleAIS07$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS07[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS07[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 8
+whereShips <- GPSwhaleAIS08$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS08[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS08[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 10
+whereShips <- GPSwhaleAIS10$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS10[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS10[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 12
+whereShips <- GPSwhaleAIS12$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS12[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS12[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+#Buoy 13
+whereShips <- GPSwhaleAIS13$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS13[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS13[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 14
+whereShips <- GPSwhaleAIS14$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS14[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS14[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 16
+whereShips <- GPSwhaleAIS16$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS16[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS16[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 18
+whereShips <- GPSwhaleAIS18$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS18[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS18[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 19
+whereShips <- GPSwhaleAIS19$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS19[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS19[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 20
+whereShips <- GPSwhaleAIS20$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS20[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS20[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 21 -- no ships occur, make all NAs
+GPSwhaleAIS21[,"timeAfterShip"] = NA
+
+# Buoy 22
+whereShips <- GPSwhaleAIS22$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS22[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS22[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+
+# Buoy 23 - no ships occur, make all NAs
+GPSwhaleAIS23[,"timeAfterShip"] = NA
+
+#Buoy 20 with interpolated data
+whereShips <- GPSwhaleAIS20.alt$shipin10km > 0
+firstship <- min(which(whereShips)) # Index of the first ship
+
+# Replace with NA where index is less than index of min value
+GPSwhaleAIS20.alt[1,"timeAfterShip"] = NA  # Need to make first row with a zero into NA separately
+GPSwhaleAIS20.alt[2:firstship - 1, "timeAfterShip"] = NA  # then can make the rest of rows NA
+```
+
+Save new GPSwhaleAISx .rda
+```{r}
+saveRDS(GPSwhaleAIS07, file = "data/GPSwhaleAIS07.rda")
+saveRDS(GPSwhaleAIS08, file = "data/GPSwhaleAIS08.rda")
 saveRDS(GPSwhaleAIS10, file = "data/GPSwhaleAIS10.rda")
@@
 saveRDS(GPSwhaleAIS23, file = "data/GPSwhaleAIS23.rda")
+saveRDS(GPSwhaleAIS20.alt, file = "data/GPSwhaleAIS20_alt.rda")
 ```
+
+Also save to external D: drive
+```{r}
+saveRDS(GPSwhaleAIS10, 'D:/Buoy_dfs/GPSwhaleAIS10.rda')
+saveRDS(GPSwhaleAIS12, 'D:/Buoy_dfs/GPSwhaleAIS12.rda')
+saveRDS(GPSwhaleAIS13, 'D:/Buoy_dfs/GPSwhaleAIS13.rda')
+saveRDS(GPSwhaleAIS14, 'D:/Buoy_dfs/GPSwhaleAIS14.rda')
+saveRDS(GPSwhaleAIS16, 'D:/Buoy_dfs/GPSwhaleAIS16.rda')
+saveRDS(GPSwhaleAIS18, 'D:/Buoy_dfs/GPSwhaleAIS18.rda')
+saveRDS(GPSwhaleAIS19, 'D:/Buoy_dfs/GPSwhaleAIS19.rda')
+saveRDS(GPSwhaleAIS20, 'D:/Buoy_dfs/GPSwhaleAIS20.rda')
+saveRDS(GPSwhaleAIS21, 'D:/Buoy_dfs/GPSwhaleAIS21.rda')
+saveRDS(GPSwhaleAIS22, 'D:/Buoy_dfs/GPSwhaleAIS22.rda')
+saveRDS(GPSwhaleAIS23, 'D:/Buoy_dfs/GPSwhaleAIS23.rda')
+saveRDS(GPSwhaleAIS07, 'D:/Buoy_dfs/GPSwhaleAIS07.rda')
+saveRDS(GPSwhaleAIS08, 'D:/Buoy_dfs/GPSwhaleAIS08.rda')
+saveRDS(GPSwhaleAIS20.alt, 'D:/Buoy_dfs/GPSwhaleAIS20_alt.rda')
+```

### 2023-10-16 01_readin.R
Update 01_readin.R

@@ -2,101 +2,53 @@
 ## Prerequisites include whales.R, tracks.R
 ## Results in dataframe for each buoy containing whale detections and buoy GPS points.
 
-
-# libraries #######
+#### libraries #######
 
 library(tidyverse)
 library(PAMpal)
 library(here)
 
-# set up #####
+##### set up #####
 
 # Set system time zone to UTC
 Sys.setenv(TZ='UTC')
 here()
 
-# Read in GPS tracks and whale detections
-AllTracks <- readRDS('data/AllTracks.rda')
-whales <- readRDS('data/whales.rda')
+# Read in whale detections
+whale7 <- readRDS('data/whale7.rda')
+whale8 <- readRDS('data/whale8.rda')
+whale10 <- readRDS('data/whale10.rda')
+whale12 <- readRDS('data/whale12.rda')
+whale13 <- readRDS('data/whale13.rda')
+whale14 <- readRDS('data/whale14.rda')
+whale16 <- readRDS('data/whale16.rda')
+whale18 <- readRDS('data/whale18.rda')
+whale19 <- readRDS('data/whale19.rda')
+whale20 <- readRDS('data/whale20.rda')
+whale21 <- readRDS('data/whale21.rda')
+whale22 <- readRDS('data/whale22.rda')
+whale23 <- readRDS('data/whale23.rda')
+whale20.alt <- readRDS('data/whale20_alt.rda')
 
-###############
-# Filter whales by drift
-##############
-
-# Create list of drifts
-#d <- unique(trk_whale$station)
-
-# Filter by drifts and assign to new variable
-#for (drift in d) {
-#  assign(paste0("trk_whales", drift), trk_whale %>% dplyr::filter(station == drift))
-#}
-# Results in trk_whales7, trk_whales8... objects
-
-##############################################################
-# Joining tracks with soundscape metrics with whale detections
-##############################################################
-
-## Make list of rda files to read in
-#iilist <- c("BB_07", "BB_08", "BB_10", "BB_12", "BB_13", "BB_14", "BB_16", 
-#            "BB_18", "BB_19", "BB_20", "BB_21", "BB_22", "BB_23", "TO_07",
-#            "TO_08", "TO_10", "TO_12", "TO_13", "TO_14", "TO_16", "TO_18", 
-#            "TO_19", "TO_20", "TO_21", "TO_22", "TO_23")
-
-## Read in rda files and assign to object
-#for (ii in iilist) {
[…]

### 2023-10-20 RF_ZC_presence_7-11_edits.R
Delete RF_ZC_presence_7-11_edits.R

@@ -1,99 +1,0 @@
-#Use a Random Forest to identify most important variables
-#in beaked and sperm whale presence
-#Modified from Anne Simonis (7/11/2023)
-
-#Required Packages
-library(randomForest)
-library(dplyr)
-library(rfPermute)
-
-# Load in main dataframe
-load(file = "allDrifts.rda")
-
-#Define response variable (presence/absence as 0/1)
-#I suggest creating a variable for presence of each species, and then create a
-#random forest model for each species 
-
-#Descriptive label for analysis
-Dep<-'CCES_Zc'
-
-# For ZC only
-allDrifts$species<-as.factor(allDrifts$species) %>%
-  droplevels(c("BB", "BW43", "BW37V", "PM", "BWC"))
-
-allDrifts$ZcPresence<-factor(ifelse(allDrifts$species=='ZC',1,0))
-
-#List all variables you want to evaluate (I only included a few here, 
-#but I know you have more environmental and AIS parameters to add)
-#covariate.list<-list(c('ZcPresence','TOL_125','TOL_2000','TOL_5000',
- #                      'sst_mean','dist2slope','depth',
-  #                     'chlorophyll_mean','mldDepth'))
-#more parameters
-covariate.list<-list(c('ZcPresence', 'curl_mean', 'bathy_slope', 'ssh_mean',  
-                       'sst_mean','dist2slope','depth',
-                       'chlorophyll_mean','mldDepth', 'mldTemp', 'ttDepth',
-                       'ttTemp', 'temp400', 'sal400'))
-
-include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
-
-#make a text string of all covariates considered for a file name 
-string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
-
-#Create a dataframe for the RF model
-DF.modelZC <- na.omit(allDrifts)
-
-# create balanced sample sizes of response for tree construction to 
-# avoid biases associated with imbalanced data
-# For ZC only
-sampsizeZC <- balancedSampsize(DF.modelZC$ZcPresence)
-
-# set seed for reproducibility
-set.seed(123)
-
-# create balanced sample size random forest model
-#By ZC
[…]

### 2023-10-20 manuscript.tex
Create manuscript.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\usepackage{soul}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\setlength{\parindent}{4em}
\setlength{\parskip}{0em}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={manuscript},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{manuscript}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\hypertarget{manuscript-draft}{%
\section{Manuscript Draft}\label{manuscript-draft}}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\hypertarget{soundscapes}{%
\subsubsection{Soundscapes}\label{soundscapes}}

Sound serves as the main sensory input for many species in the ocean.
Since sound travels over four times faster in water than in air and can
carry information over far distances (Urick, 1983), the use of sound has
evolved as a critical component of communication, navigation, and
foraging for many invertebrates, fishes, and cetaceans (Duarte et al.,
2021). Under the right conditions, sounds can travel across ocean basins
(Miksis-Olds, Martin, and Tyack, 2018).\\

The total collection of sounds in an environment is called a soundscape,
and these sources of sound are grouped into three broad categories:
biophonies, sound produced by animals like cetaceans, fishes, and
invertebrates; geophonies, natural sounds like rain, wind, and
earthquakes; and anthropophonies, human-made sounds from shipping
vessels, sonar, and oceanographic survey activities (Duarte et al.,
2021; Miksis-Olds et al., 2018). Every region of the ocean experiences
high spatio-temporal variation in sound levels and sources based on
traits such as properties of water, organism diversity, human
activities, and bathymetry (André et al., 2011). Because of their unique
sound signatures, ocean soundscapes can be characterized, measured, and
monitored as indicators of ecosystem health or species diversity (Haver
et al., 2019; Pijanowski, Farina, Gage, Dumyahn, and Krause, 2011;
Radford, Stanley, Tindle, Montgomery, and Jeffs, 2010; Weiss et al.,
2021).\\

When noise levels reach higher than normal levels, organisms can be
negatively impacted. Invertebrates and fishes may experience stunted
development, elevated stress responses, and physical injuries (L.
Weilgart, 2018), while cetaceans' behavior, calling patterns, physiology
and stress levels, and ability to locate prey or conspecifics are
impacted (L. S. Weilgart, 2007). Sensitivities and responses to noise
can vary among species and even individuals (Kunc, McLaughlin, and
Schmidt, 2016). While terrestrial ecosystem-level effects have been
well-documented (Buxton et al., 2017; Francis, Kleist, Ortega, and Cruz,
2012; Shannon et al., 2016), effects within aquatic ecosystems require
further studies (Kunc et al., 2016).\\

Ocean noise pollution is among the anthropogenic changes that have
recently grown to concerning levels (Hildebrand, 2009; Tyack, 2008). The
global ambient noise in the ocean is estimated to have increased by 12
decibels in the last few decades and has a close correlation to global
economic trends (Frisk, 2012). Ships have been increasing in terms of
gross tonnage and fleet size (Hildebrand, 2009) and now account for over
80\% of the global trade, with an estimated 1.4\% expected annual
increase in shipping through 2027 (UNCTAD, 2022). Commercial shipping
now accounts for most of the low-frequency anthropogenic noise in the
ocean (Hildebrand, 2009).\\

These noise-generating activities are categorized as either incidental
or deliberate (Chahouri, Elouahmani, and Ouchene, 2022; Chou, Southall,
Robards, and Rosenbaum, 2021). Deliberate activities such as active
sonar pingers, resource extraction from the seabed, seismic surveys, and
echosounders intentionally produce sound to measure an area and can
utilize both high and low frequencies (Duarte et al., 2021). Incidental
activities like dredging, offshore development, and shipping produce
sound as a by-product (Hawkins and Popper, 2017; Southall et al., 2017).
Anthropogenic noise can further be characterized by the frequency and
duration of the sound. For example, shipping occurs in all oceans and
produces a continuous, low-frequency (\textless200 Hz) sound that
travels far distances (Hildebrand, 2009). Sonar for military
surveillance or research such as seafloor mapping produces mid-frequency
sounds, but attenuates quicker, therefore impacting a smaller acoustic
space (Hildebrand, 2009; Richardson, Jr, Malme, and Thomson, 2013).
Noise has been shown to propagate differently based on static features
like bathymetry and depth (Richardson et al., 2013; Vagle, Burnham,
O'Neill, and Yurk, 2021), as well as dynamic variables like temperature,
salinity, and pH from a changing climate (Affatati, Scaini, and Salon,
2022; Kunc et al., 2016).\\

As soniferous animals, cetaceans have varying levels of sensitivity and
vulnerability to certain types of noise. Generally, mysticetes are more
affected by low-frequency noise produced by large shipping vessels
(Southall et al., 2017). Odontocetes produce and detect mid- to
high-frequency sounds and are more likely to interfere with human
activities at these frequencies. These generalizations appear to have
exceptions, as odontocete species like beaked whales have shown
avoidance behaviors in response to broadband and low frequency vessel
noise (Aguilar Soto et al., 2006; Pirotta et al., 2012). It remains
uncertain how short-term effects of chronic or acute noise pollution for
cetaceans translates into long-term or population level changes. New,
Moretti, Hooker, Costa, and Simmons (2013) suggest enough exposure can
cause beaked whales to cease foraging and eventually lower reproduction
rates.\\

\hypertarget{beaked-whales-and-sperm-whales}{%
\subsubsection{Beaked Whales and Sperm
Whales}\label{beaked-whales-and-sperm-whales}}

Beaked whales (family \emph{Ziphiidae)} are a group of 24 species and
among the least understood cetaceans and deepest divers in the ocean.
They regularly dive for over an hour to depths greater than 800 meters,
sometimes exceeding three hours and 2000 meters, followed by only a few
minutes on the surface (Baird et al., 2006; Tyack, Aguilar Soto,
Johnson, Sturlese, and Madsen, 2006). Their preference for deep water,
coupled with a low-profile body makes them nearly invisible on the
surface, so very little is known about them with visual observations
(Awbery, 2022). Much of what is known comes from strandings and stomach
content analysis (West, Walker, Baird, Mead, and Collins, 2017). Even
when sighting events occur, they risk misidentification due to the
paucity of data of morphological variations in some species (Aguilar de
Soto et al., 2017). Tagging studies prove useful, but it is often
difficult to acquire a sizable dataset (Baird, 2008; Baird et al., 2006;
Tyack et al., 2006).

Beaked whales are known to be exceptionally sensitive to noise (Hooker
et al., 2019). There are several documented cases of naval mid-frequency
active sonar (MFAS) activities coinciding with mass stranding events by
beaked whales (Bernaldo de Quiros et al., 2019; DeRuiter et al., 2013;
Fernández et al., 2005; Simonis, Brownell, et al., 2020; Stanistreet et
al., 2022; L. S. Weilgart, 2007). The same physiological adaptations
that allow for extreme diving are observed to increase their
decompression risk when they alter their behavior in response to noise
(Fahlman, Tyack, Miller, and Kvadsheim, 2014; Kvadsheim et al., 2012).
Furthermore, these responses vary based on the individuals' age, sex,
past experience with noise, or current activity (L. S. Weilgart, 2007).

Despite their visually cryptic nature, their sound signals are very
well-described, down to the species level (Baumann-Pickering et al.,
2013; Baumann-Pickering et al., 2014). Beaked whales produce clicks,
pulses, and buzzes during foraging below 500 meters of depth (Tyack et
al., 2006). Increased acoustic surveys reported in the literature have
detected new vocalization patterns globally (Manzano-Roth et al., 2023),
with notable temporal patterns emerging (Yamada et al., 2019).

Sperm whales (\emph{Physeter macrocephalus}) also occupy deep ocean
habitats, but their dive depths and times are reduced in comparison.
Sperm whales have been known to dive regularly between 400-600 meters
and often over 1000 meters (Amano and Yoshioka, 2003; Awbery, 2022;
Watkins, Daher, Fristrup, Howald, and Di Sciara, 1993). They also appear
to have more acoustic tolerance than beaked whales (Madsen, Mohl,
Nielsen, and Wahlberg, 2002; Patrick J. O. Miller et al., 2022; P. J. O.
Miller et al., 2009; Winsor, Irvine, and Mate, 2017). While avoidance
responses in sperm whales have been recorded following MFAS events (Curé
et al., 2013; Sivle et al., 2012; Stanistreet et al., 2022) and seismic
survey pulses (Madsen et al., 2002) , and can range from minor (Isojunno
et al., 2022) to moderate (Isojunno et al., 2016) in severity, the
circumstances surrounding the noise exposure is not well understood.

Sperm whales also emit series of targeted clicks when socializing (Fais
et al., 2015; Marcoux, Whitehead, and Rendell, 2006) and foraging (Fais,
Johnson, Wilson, Aguilar Soto, and Madsen, 2016), but it's unlikely they
debilitate their prey with their vocalizations as previously
hypothesized (Norris and Harvey, 1972). More is known about sperm whales
than beaked whales as they spend more time on the surface, lending
themselves to better detection during visual surveys.

Several studies to date have documented specific effects of acoustic
interference from human activity in beaked and sperm whales. Intense
noise events can affect behavior through habitat displacement, time and
effort fleeing spent instead of foraging, and communication and
echolocation click disruption (Aguilar Soto et al., 2006; Cholewiak,
DeAngelis, Palka, Corkeron, and Van Parijs, 2017; DeRuiter et al., 2013;
Falcone et al., 2017; Joyce et al., 2020; Stanistreet et al., 2022);
cause physiological harm like hearing loss, tissue damage, and elevated
stress levels (Cox et al., 2006; Hooker et al., 2019) even up to 100 km
away (Falcone et al., 2017); and mask signals for communication (Erbe,
Williams, Sandilands, and Ashe, 2014), environmental cues (Duarte et
al., 2021), or mating (L. S. Weilgart, 2007).

Under the Marine Mammal Protection Act, these species are federally
protected, and several receive addition protection under the Endangered
Species Act due to a vulnerable or endangered conservation status
(Taylor et al., n.d.), but some remain listed as ``data deficient'' by
the International Union for the Conservation of Nature (IUCN) Red List
of Threatened Species (Pitman and Brownell, n.d., 2020). The latter
designation has been scrutinized for not eliciting the same conservation
urgency as species whose populations are well-documented (Parsons,
2016), therefore emphasizing the need for population and abundance
studies of these species.

\hypertarget{passive-acoustic-monitoring}{%
\subsubsection{Passive Acoustic
Monitoring}\label{passive-acoustic-monitoring}}

Passive acoustic monitoring (PAM) is a tool that allows researchers to
record and listen to the soundscape. It is also an ideal method for
assessing habitat use and improving species distribution models of
visually cryptic cetaceans like beaked whales and sperm whales that are
more likely to be heard than seen (Fleishman et al., 2023), live
relatively far offshore (Robbins, Bell, Potts, Babey, and Marley, 2022),
and have low-density populations (Arranz et al., 2023; Hildebrand et
al., 2015; Hooker et al., 2019).

Passive acoustic recorders come in two forms: fixed or mobile, each
better suited to a certain ecological application. A mobile (towed or
free drifting) buoy is either towed behind a vessel or moves
autonomously with the currents for days or weeks at a time. It captures
the dynamic nature of sound over a large spatial area, especially in
deep and less studied parts of the ocean, making it ideal for detecting
offshore or mobile species. Bottom-mounted recorders allow for greater
temporal coverage and are often used when studying specific regions or
known habitats of study species (Jarvis, DiMarzio, Watwood, Dolan, and
Morrissey, 2022; Marques et al., 2012; Nosal and Frazer, 2007).

The benefits of using PAM over visual surveys include use during night
hours or through inclement weather, lower cost to charter and little
experience needed to conduct a survey (Fleishman et al., 2023; Johnson,
Soto, and Madsen, 2009). The trade-offs include costly equipment or
deployment, high data/battery storage requirements, uncertain chance of
return, limited spatial coverage for fixed recorders, and limited
temporal coverage for mobile recorders. Even with experienced
acousticians, noisy detections can obfuscate number of individuals or
species detected and distance or direction from recorder for species
with large call ranges such as sperm whales (Barlow and Gisiner, 2005;
Mellinger, Stafford, Moore, Dziak, and Matsumoto, 2007).

\hypertarget{automatic-identification-system}{%
\subsubsection{Automatic Identification
System}\label{automatic-identification-system}}

Automatic Identification System (AIS) is a mandatory vessel tracking
system for vessels over 300 gross tons to aid in position communication
and safe navigation (Federal Register, 2003). While not originally
intended for use in research, it has become a valuable resource for
vessel traffic data, especially in studying effects of vessel strikes
(Greig, Hines, Cope, and Liu, 2020; Reimer, Gravel, Brown, and Taggart,
2016) and noise pollution (Aguilar Soto et al., 2006; Erbe et al., 2014)
on cetaceans. Because ships are abundant throughout the study area due
to major U.S. ports, AIS was included in this study to investigate
patterns of co-occurrence between vessels and whales and sound levels in
those moments. Previous studies have separated vessel and anthropogenic
noise from ambient soundscape based on known spectral signatures,
frequency bands, and percentiles (Haver et al., 2020; Weiss et al.,
2021). The inclusion of AIS to soundscape metrics in this study provides
fine-scale evaluation of sound and source.

\hypertarget{laws-and-regulations}{%
\subsubsection{\texorpdfstring{\textbf{Laws and
Regulations}}{Laws and Regulations}}\label{laws-and-regulations}}

While there is consensus that anthropogenic noise in the ocean is
increasing, policies that aim to mitigate chronic noise are generally
lacking. This is due to knowledge gaps regarding the long-term and
cumulative effects of noise on certain species (Markus and Sánchez,
2018). International and regional efforts are underway (see Chou et al.
(2021)) but are not fully complete. Within the European Union, the
Marine Strategy Framework Directive establishes a directive about
environmental regulation on marine noise, specifically requiring member
states to adopt strategies in compliance with threshold values for
impulsive and continuous noise (Borsani, J.F., Juretzek C., Klauson A.,
Leaper R., Le Courtois F., Liebschner A., Maglio A., Mueller A. , Norro
A., Novellino A., Outinen O., Popit A., Prospathopoulos A., Sigray P.,
Thomsen F., Tougaard J., Vukadin P., and Weilgart L., Borsani, J.F.,
Andersson M., André M., Azzellino A., Bou M., Castellote M., Ceyrac L.,
Dellong D., Folegot T., Hedgeland D., and Juretzek C., Klauson A.,
Leaper R., Le Courtois F., Liebschner A., Maglio A., Mueller A. , Norro
A., Novellino A., Outinen O., 2023; \emph{Directive 2008/56/EC of the
European Parliament and of the Council of 17 June 2008 establishing a
framework for community action in the field of marine environmental
policy (Marine Strategy Framework Directive) (Text with EEA relevance)},
2008). Some studies have independently tested these threshold values
(Garrett et al., 2016; Haver et al., 2021; Sebastianutto, Fortuna,
Mackelworth, Holcer, and Rako Gospić, 2015) or used them as guidelines
in their study (Mustonen et al., 2019). The National Oceanic and
Atmospheric Administration (NOAA) has developed an Ocean Noise Strategy
Roadmap, an agency-wise approach to mitigating both chronic and acute
effects of noise exposure to animals (NOAA, 2018). Within it, the
authors agree upon impairment thresholds for mid-frequency cetaceans
(NOAA, 2018). Because chronic effects are not fully understood, they
remain to be regulated. The International Maritime Organization (IMO)
has also been addressing noise pollution from commercial vessels through
quieter ship design (International Maritime Organization, 2014). Updates
to their 2014 guidelines are set to be finalized by January 2024.
Finally, Canada has launched a Quiet Vessel Initiative and is expected
to publish a Ocean Noise Strategy for underwater noise management in
2023 (Breeze et al., 2022).

\hypertarget{research-objective}{%
\subsubsection{Research Objective}\label{research-objective}}

With anthropogenic noise-generating activities on the rise, global
oceanic sound levels are expected to increase. Noise is a form of
habitat degradation, so there is a need for research on a large spatial
scale to understand how these highly mobile animals overlap in space and
time with human activities and how they're affected (Erbe et al., 2014;
Haver et al., 2019; Hooker et al., 2019). Acquiring information on these
federally protected and acoustically sensitive species is challenging,
but certain PAM techniques can be well-suited for this type of research.
Logistical difficulties associated with offshore research and
low-density populations of visually cryptic deep diving whales have
created a paucity of data in the California Current Ecosystem. This
research provides a step towards understanding soundscape variation in
the area and beaked and sperm whale distributions and their acoustic
environment.

The purpose of this study is to characterize the spatial and temporal
acoustic variability of beaked and sperm whale habitat in the California
Current Ecosystem using soundscape metrics, environmental variables, and
AIS. We determine significant predictors of whale presence for each
species with the use of Random Forest models and generalized additive
models and provide explicit sources for noise-generating human
activities through overlay of large vessel activity.

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\hypertarget{study-area}{%
\subsubsection{\texorpdfstring{\textbf{Study
Area}}{Study Area}}\label{study-area}}

The California Current Ecosystem (CCE) is a highly productive and
complex ecosystem that contains one proposed and five current national
marine sanctuaries (Office of National Marine Sanctuaries, n.d.),
supports several wild fisheries (Field and Francis, 2006), and is home
to nineteen protected whale species (NOAA Fisheries, n.d.). The
1,141,800 km2 area extends along the US' Economic Exclusion Zone from
southern Canada to northern Mexico and is a characterized by deep
canyons and a continental shelf (Barlow et al., 2009; Barlow and Forney,
2007). Characteristic of the area are upwelling events that carry
nutrients to the surface and drive primary productivity, creating an
ideal habitat where whales' prey are found (Ryther, 1969). The year 2018
was characterized in the CCE with slightly warmer than average sea
surface temperatures as the recent El-Nino and marine heatwave events
were ending (Harvey et al., n.d.). Because this area has high economic
value and natural ecological activity, it's a well-managed and
well-researched area (Coleman, 2008; Oldach et al., 2022; Peña and
Bograd, 2007).\\
\strut \\
In 2021, California, Washington, and Oregon ranked 3rd, 5th, and 23rd in
most trafficked U.S. waterways, respectively (United States Army Corps
of Engineers {[}USACE{]}, 2018). Between vessel activity like fishing,
shipping and naval exercises in California alone, there is notable
overlap of anthropogenic presence with important habitat for protected
species. Furthermore, the future development of offshore wind turbines
off of Morro Bay and Humboldt, California are expected to be
noise-producing activities during construction and operation, increasing
sound levels in those regions (Cooperman et al., 2022; Madsen, Wahlberg,
Tougaard, Lucke, and Tyack, 2006).

\hypertarget{data-collection-and-processing}{%
\subsubsection{Data Collection and
Processing}\label{data-collection-and-processing}}

Acoustic detections were collected as part of the 2018 California
Current Ecosystem Survey (CCES) by the Southwest Fisheries Science
Center (SWFSC). Twenty-three free drifting buoys (numbered 1-23) were
deployed at varying times between July and November in the offshore
waters of the study area. These buoys contained two satellite
geo-locators above water to monitor GPS information and two hydrophones
at depth attached by a 150-meter line. They collected a total of 1910
hours of data. SWFSC detected and classified acoustic detections in
PAMGuard to species level as per methods mentioned in Simonis, Trickey,
et al. (2020).

Acoustic data were duty-cycled, with most buoys sampling at a rate of 18
minutes off and 2 minutes on. The SWFSC team analyzed the data and
detected echolocation pulses produced by beaked and sperm whales. These
were classified by peak frequency and spectral signature and identified
to the species level (see Simonis, Trickey, et al. (2020) for details on
this survey and Rand, Wood, and Oswald (2022) for duty-cycle study).
Soundscape metrics were binned by peak frequency into low, medium, high
frequency bands, along with the 1, 5, 10, 25, 75, 95, and 99 percentile
distribution from each metric.

\st{Soundscape metrics were rounded to the nearest 20 minutes (:00, :20,
:40) and converted to the UTC time zone. These metrics include
third-octave level (TOL) bands from 63 to 20,000 Hz and broadband
(20-24,000 Hz) metrics.} Several TOL levels were chosen as proxies to
vessel or sonar frequencies.

\st{Whale detection timestamps were rounded to nearest 20 minutes based
on start time to align with the soundscape metrics. A separate column
was created as a common join field that contained nearest timestamp in
UTC to that of buoy tracks (see below).} Two types of beaked whale
detections (species code \emph{?BW} and \emph{?BWC}) were removed from
analysis as the species was unknown.

**\[New addition as of 08/25/23 -** Two SPOT buoys sit on every buoy, so some GPS points were recorded twice. These duplicate tracks were eliminated.\]
\st{Buoy tracks were similarly rounded to nearest 20 minutes.} Seven
buoys (1,2,3,5,6,9,11) were lost at sea and two buoys had corrupted data
(4,17) so these were removed from the analysis. Buoys 14 and 15 were the
same and therefore consolidated to 14. These three preliminary datasets
were joined together based on a common buoy number (column ``station'')
and time field column ``UTC''). Whale detections (column ``Wpresence'')
were reduced to binary data for absence/presence and all other fields of
that dataset were eliminated.

\hypertarget{environmental-data}{%
\subsubsection{\texorpdfstring{\textbf{Environmental
Data}}{Environmental Data}}\label{environmental-data}}

Possible environmental and oceanographic predictors of cetacean habitat
were extracted from the NOAA Environmental Research Division Data Access
Program (ERDDAP) server
((\url{https://coastwatch.pfeg.noaa.gov/erddap/}) and Hybrid Coordinate
Ocean Model (HYCOM) using the \emph{matchEnvData} function in R package
\emph{PAMmisc}. Six modeled variables at depth (mixed layer depth, mixed
layer temperature, temperature at 400m, thermocline temperature,
thermocline depth, salinity at 400m) were calculated using a formula
from McCullough et al. (2021). Such variables are especially important
to incorporate as it better reflects the deep environment that beaked
and sperm whales are known to inhabit. Distance to slope and bathymetric
slope were calculated from a bathymetry TIF acquired from the General
Bathymetric Chart of the Oceans (GEBCO).

\[See Virgili et al 2019 for extracting slope from GEBCO Data and dynamic variables!!\]

\hypertarget{modeling}{%
\subsubsection{\texorpdfstring{\textbf{Modeling}}{Modeling}}\label{modeling}}

We used a Random Forest (RF) model in R (R Core Team, 2023) using the
package () to identify most important environmental or acoustic
predictors for beaked and sperm whales. A separate model was run for
each species.

We built generalized additive models in R using the package \emph{mcgv}
to relate beaked and sperm whale detections to an environmental or
acoustic variable.

\hypertarget{results}{%
\subsection{Results}\label{results}}

(Describe simple statistical relationships, correlations, exploratory
data findings (highest, lowest, etc., time of day ( Ziegenhorn et al.,
2023 ) , drift with most detections/day), where/when were most
detections heard, number of species)

\begin{itemize}
\item
  Buoy 18 had the most whale detections
\item
  ZC was the most frequently detected whale
\item
  The loudest sound levels were only occupied by PM
\item
  Density of ship encounters
\item
  Report on environmental conditions for ZC/PM?
\item
  Report on analysis of env variables WITH and WITHOUT soundscape
  metrics (see Fiedler, Becker, Forney, Barlow, and Moore (2023),
  results 3.1)
\end{itemize}

Weekly/daily plot of acoustic presence of whales and ships over time
from each drift? See JS Trickey et al.~2022, Fig 3 Which buoy had the
most AIS overlap and when?

Look at power spectral density? Haver et al. (2020)

The band at 125 Hertz is often associated with shipping vessel noise and
the 2-5 kHz range with mid-frequency naval sonar ( Garrett et al. (2016)
). Beaked whales are known to be sensitive to sonar in this latter range
(cite)

AIC explanation Symonds and Moussalli (2011)

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\[Laws and Regulations\]

While there is consensus that anthropogenic noise in the ocean is
increasing, policies that aim to mitigate chronic noise are generally
lacking. This is due to knowledge gaps regarding the long-term and
cumulative effects of noise on certain species (Markus and Sánchez
2018). International and regional efforts are underway (see Chou et
al.~(2021)) but are not fully complete. Within the European Union, the
Marine Strategy Framework Directive establishes a directive about
environmental regulation on marine noise, specifically requiring member
states to adopt strategies in compliance with threshold values for
impulsive and continuous noise (``Directive 2008/56/EC of the European
Parliament and of the Council of 17 June 2008 Establishing a Framework
for Community Action in the Field of Marine Environmental Policy (Marine
Strategy Framework Directive) (Text with EEA Relevance)'' 2008; Borsani,
J.F. et al.~2023). Some studies have independently tested these
threshold values (Sebastianutto et al.~2015; Garrett et al.~2016; Haver
et al.~2021) or used them as guidelines in their study (Mustonen et
al.~2019). The National Oceanic and Atmospheric Administration (NOAA)
has developed an Ocean Noise Strategy Roadmap, an agency-wise approach
to mitigating both chronic and acute effects of noise exposure to
animals (NOAA 2018). Within it, the authors agree upon impairment
thresholds for mid-frequency cetaceans (NOAA 2018). Because chronic
effects are not fully understood, they remain to be regulated. The
International Maritime Organization (IMO) has also been addressing noise
pollution from commercial vessels through quieter ship design
(International Maritime Organization 2014). Updates to their 2014
guidelines are set to be finalized by January 2024. Finally, Canada has
launched a Quiet Vessel Initiative and is expected to publish a Ocean
Noise Strategy for underwater noise management in 2023 (Breeze et
al.~2022).

\[Known limitations to this study\]\\
Duty cycle whale detection and soundscape method knowingly omits 18 of
20 minutes of data. A sperm whale's true location relative to buoy is
unknown so preferred habitat may not reflect the environmental variables
that are associated with buoy that detects it (Barkley et al.~2021)
Rounding tracks and soundscape metrics date/times to nearest 20 minutes
is not perfect. Often, times did not line up with :00, :20, :40 and were
either omitted or duplicated.

This study only captures half of the year. If BW/SW have seasonal
preferences outside of the study period, it was not captured here.

Are BW or SW known to have seasonal movements? Robbins et al. (2022)
suggest they do, though there's conflicting evidence.

Non-AIS vessels or small vessels (fishing, recreational) may contribute
to the soundscape in relevant frequency bands (Hermannsen et al. (2019)
, Hildebrand (2009) ) and this does not get captured here.

Does not account for the differences in relative noise contribution by
vessel type (as Hatch et al. (2008) does)

\[Add here.\]

This work will contribute to establishing baseline conditions of noise
exposure to inform wind farm development and shipping lane changes. In
better understanding this, human activities can be altered to avoid
overlap with sensitive species Hooker et al. (2019).

Both of these outcomes are of high importance to managers and agencies
especially NOAA and BOEM (see 2023-2024 report). Within their Ocean
Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification
of the spatial and temporal variability of ambient noise conditions, and
2) understanding of how anthropogenic sound sources contribute to the
soundscape. Co-occurrence of buoys and vessels will provide fine-scale
snapshots of sound levels surrounding vessels and concurrent whale
foraging activity. With AIS, we can calculate distance from buoy to
vessel and sound levels in those moments. And 3) NOAA is seeking more
information on vocally active species' distributions. This is part of a
bill introduced by Congress to ``assess underwater sound in
high-priority'' environments
(\url{https://www.congress.gov/bill/117th-congress/house-bill/6987}).

Outside of cetacean research, similar designs of drifting buoys have
been used in estuarine and coral reef soundscape studies Lillis et al.
(2018).

The year 2018 was characterized in the CCE with slightly warmer than
average sea surface temperatures as the recent El-Nino and marine
heatwave events were ending (Harvey et al.~2019).

Croll, Clark, Calambokidis, Ellison, and Tershy (2001) found that whale
presence more closely related to prey abundance than sound levels

Can use non-parametric Kruskal-Wallis one-way analysis of variance test
to examine \emph{if number of whale detections per hour differed
depending on presence/absence of vessels} (see Trickey et al.~2022).

\[Recommendations?\]

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

Taiki - help with code!\\
Megan McKenna - guidance on AIS questions.

Add here.

\hypertarget{tables}{%
\subsubsection{Tables}\label{tables}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  Species studied:\\
\end{enumerate}

\begin{itemize}
\tightlist
\item
  ZC Cuvier's beaked whale~\emph{Ziphius cavirostris (IUCN status: least
  concern)\\
  -} BW43, Perrin's beaked whale~\emph{M}.~\emph{perrini (IUCN status:
  endangered)} confirmed in 2021 Baumann-Pickering et al. (2013) Barlow
  et al. (2021)\\
\item
  BW37V, possibly Hubbs' beaked whale \emph{M. carlhubbsi (IUCN status:
  data deficient)} Simonis, Trickey, et al. (2020)\\
\item
  BB, -- Baird's beaked whale \emph{B. bairidii} \emph{(IUCN status:
  least concern)}\\
\item
  BWC, possibly ginkgo-toothed beaked whale \emph{M. ginkgodens (IUCN
  status: data deficient)} Simonis, Trickey, et al. (2020)\\
\item
  PM, sperm whale \emph{(IUCN status: vulnerable)}
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  Environmental covariates (group by physiographic, oceanographic, etc)
\end{enumerate}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2361}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2361}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2917}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2361}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Covariate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Resolution
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Justification
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
& & & \\
Distance to shore (m) & 0.01-Degree Grid & ERDAPP & \\
Distance to continental slope (m) & 15 arc sec & Derived from General
Bathymetric Chart of the Oceans (GEBCO) Weatherall et al. (2015) &
Barlow et al. (2009) \\
Bathymetric slope (°) & 15 arc sec & Derived from GEBCO tif & Becker et
al. (2010) Virgili et al. (2022) \\
Depth (m) & 15 arc sec & ERDAPP & Forney et al. (2012) Becker et al.
(2010) Virgili et al. (2022) \\
Wind curl stress & 1°, 6-hourly & ERDDAP & Mannocci, Monestiez, Spitz,
and Ridoux (2015) \\
Mixed layer depth (m) & 0.08°, 3-hourly & HYCOM + NCODA, GLBy0.08/expt
93.0, calculated with variable representative isotherm method (Fiedler
2010), as described in McCullough et al 2021. & McCullough et al.
(2021), Forney et al. (2012) Moore (2021) \\
Thermocline temperature (°C) & 0.08°, 3-hourly & HYCOM + NCODA,
GLBy0.08/expt 93.0, calculated with variable representative isotherm
method (Fiedler 2010), as described in McCullough et al 2021. &
McCullough et al. (2021) \\
Thermocline depth (m) & 0.08°, 3-hourly & HYCOM + NCODA, GLBy0.08/expt
93.0, calculated with variable representative isotherm method (Fiedler
2010), as described in McCullough et al 2021. & \\
Mixed layer depth Temperature (°C) & 0.08°, 3-hourly & Calculated from
HYCOM & McCullough et al. (2021) \\
Salinity at 400m & 0.08°, 3-hourly & HYCOM + NCODA, GLBy0.08/expt 93.0
& \\
Temperature at 400m (°C) & 0.08°, 3-hourly & HYCOM + NCODA,
GLBy0.08/expt 93.0 & \\
SST mean & 0.01° , daily & ERDAPP & Becker 2007, Forney et al. (2012)
Virgili et al. (2022) \\
CHL & 4 km, monthly & ERDAPP & Forney et al. (2012) Virgili et al.
(2019) \\
Sea Surface Height Anomalies (m) & daily & ERDAPP & Virgili et al.
(2019) \\
\end{longtable}

Can show table of variables and justification by papers (Virgili et
al.~2022 and Virgili et al 2019)

Forney et al. (2012) shows how BW model had a lot of uncertainty
regarding distribution predictors. So few detections that some spp had
to be lumped together (also Barlow et al. (2009)).

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-affatati2022}{}}%
Affatati, A., Scaini, C., \& Salon, S. (2022). Ocean Sound Propagation
in a Changing Climate: Global Sound Speed Changes and Identification of
Acoustic Hotspots. \emph{Earth's Future}, \emph{10}(3), e2021EF002099.
\url{https://doi.org/10.1029/2021EF002099}

\leavevmode\vadjust pre{\hypertarget{ref-aguilardesoto2017}{}}%
Aguilar de Soto, N. A. de, Martín, V., Silva, M., Edler, R., Reyes, C.,
Carrillo, M., \ldots{} Carroll, E. (2017). True{'}s beaked whale
(Mesoplodon mirus) in Macaronesia. \emph{PeerJ}, \emph{5}, e3059.
\url{https://doi.org/10.7717/peerj.3059}

\leavevmode\vadjust pre{\hypertarget{ref-aguilarsoto2006}{}}%
Aguilar Soto, N., Johnson, M., Madsen, P. T., Tyack, P. L., Bocconcelli,
A., \& Fabrizio Borsani, J. (2006). Does Intense Ship Noise Disrupt
Foraging in Deep-Diving Cuvier's Beaked Whales (ziphius Cavirostris)?
\emph{Marine Mammal Science}, \emph{22}(3), 690--699.
\url{https://doi.org/10.1111/j.1748-7692.2006.00044.x}

\leavevmode\vadjust pre{\hypertarget{ref-amano2003}{}}%
Amano, M., \& Yoshioka, M. (2003). Sperm whale diving behavior monitored
using a suction-cup-attached TDR tag. \emph{Marine Ecology Progress
Series}, \emph{258}, 291--295. Retrieved from
\url{https://www.jstor.org/stable/24867054}

\leavevmode\vadjust pre{\hypertarget{ref-andruxe92011}{}}%
André, M., Schaar, M. van der, Zaugg, S., Houégnigan, L., Sánchez, A.
M., \& Castell, J. V. (2011). Listening to the Deep: Live monitoring of
ocean noise and cetacean acoustic signals. \emph{Marine Pollution
Bulletin}, \emph{63}(1), 18--26.
\url{https://doi.org/10.1016/j.marpolbul.2011.04.038}

\leavevmode\vadjust pre{\hypertarget{ref-arranz2023}{}}%
Arranz, P., Miranda, D., Gkikopoulou, K. C., Cardona, A., Alcazar, J.,
Aguilar de Soto, N., \ldots{} Marques, T. A. (2023). Comparison of
visual and passive acoustic estimates of beaked whale density off el
hierro, canary islands. \emph{The Journal of the Acoustical Society of
America}, \emph{153}(4), 2469. \url{https://doi.org/10.1121/10.0017921}

\leavevmode\vadjust pre{\hypertarget{ref-awbery2022}{}}%
Awbery, T. (2022). \emph{Spatial distribution and encounter rates of
delphinids and deep diving cetaceans in the eastern mediterranean sea of
turkey and the extent of overlap with areas of dense marine traffic}.
Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fmars.2022.860242/full}

\leavevmode\vadjust pre{\hypertarget{ref-baird2008}{}}%
Baird, R. W. (2008). Diel variation in beaked whale diving behavior.
\emph{Marine Mammal Science}.
\url{https://doi.org/10.1111/j.1748-7692.2008.00211.x}

\leavevmode\vadjust pre{\hypertarget{ref-baird2006}{}}%
Baird, R. W., Webster, D. L., McSweeney, D. J., Ligon, A. D., Schorr, G.
S., \& Barlow, J. (2006). Diving behaviour of cuvier{'}s (ziphius
cavirostris) and blainville{'}s (mesoplodon densirostris) beaked whales
in hawai{`}i. \emph{Canadian Journal of Zoology}, \emph{84}(8),
1120--1128. \url{https://doi.org/10.1139/Z06-095}

\leavevmode\vadjust pre{\hypertarget{ref-barlow2021}{}}%
Barlow, J., Cárdenas-Hinojosa, G., Henderson, E. E., Breese, D.,
Lopez-Arzate, D., Hidalgo-Pla, E., \& Taylor, B. N. (2021). Unique
morphological and acoustic characteristics of beaked whales (mesoplodon
sp.) off the west coast of baja california, mexico. \emph{Marine Mammal
Science}. \url{https://doi.org/10.1111/mms.12853}

\leavevmode\vadjust pre{\hypertarget{ref-barlow2009}{}}%
Barlow, J., Ferguson, M. C., Becker, E. A., Redfern, J. V., Forney, K.
A., Vilchis, I. L., \ldots{} Ballance, L. T. (2009). \emph{Predictive
modeling of cetacean densities in the eastern pacific ocean}. Retrieved
from
\url{https://repository.library.noaa.gov/view/noaa/3666/noaa_3666_DS1.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-barlow2007}{}}%
Barlow, J., \& Forney, K. A. (2007). Abundance and population density of
cetaceans in the california current ecosystem. \emph{Fishery Bulletin},
\emph{105}(4), 509--527. Retrieved from
\url{https://go.gale.com/ps/i.do?p=AONE\&sw=w\&issn=00900656\&v=2.1\&it=r\&id=GALE\%7CA203535836\&sid=googleScholar\&linkaccess=abs}

\leavevmode\vadjust pre{\hypertarget{ref-barlow2005}{}}%
Barlow, J., \& Gisiner, R. (2005). Mitigating, monitoring and assessing
the effects of anthropogenic sound on beaked whales. \emph{J. Cetacean
Res. Manage.}, \emph{7}(3), 239--249.
\url{https://doi.org/10.47536/jcrm.v7i3.734}

\leavevmode\vadjust pre{\hypertarget{ref-baumann-pickering2013}{}}%
Baumann-Pickering, S., McDonald, M. A., Simonis, A. E., Solsona Berga,
A., Merkens, K. P. B., Oleson, E. M., \ldots{} Hildebrand, J. A. (2013).
Species-specific beaked whale echolocation signals. \emph{The Journal of
the Acoustical Society of America}, \emph{134}(3), 2293--2301.
\url{https://doi.org/10.1121/1.4817832}

\leavevmode\vadjust pre{\hypertarget{ref-baumann-pickering2014}{}}%
Baumann-Pickering, S., Roch, M. A., Jr, R. L. B., Simonis, A. E.,
McDonald, M. A., Solsona-Berga, A., \ldots{} Hildebrand, J. A. (2014).
Spatio-Temporal Patterns of Beaked Whale Echolocation Signals in the
North Pacific. \emph{PLOS ONE}, \emph{9}(1), e86072.
\url{https://doi.org/10.1371/journal.pone.0086072}

\leavevmode\vadjust pre{\hypertarget{ref-becker2010}{}}%
Becker, E. A., Forney, K. A., Ferguson, M. C., Foley, D. G., Smith, R.
C., Barlow, J., \& Redfern, J. V. (2010). Comparing california current
cetacean{\textendash}habitat models developed using in situ and remotely
sensed sea surface temperature data. \emph{Marine Ecology Progress
Series}, \emph{413}, 163--183. Retrieved from
\url{https://www.jstor.org/stable/24875187}

\leavevmode\vadjust pre{\hypertarget{ref-bernaldodequiros2019}{}}%
Bernaldo de Quiros, Y., Fernández, A., Baird, R. W., Brownell, R. L.,
Aguilar de Soto, N., Allen, D., \ldots{} Schorr, G. (2019). Advances in
research on the impacts of anti-submarine sonar on beaked whales.
\emph{Proceedings of the Royal Society B: Biological Sciences},
\emph{286}(1895), 1--9. Retrieved from
\url{https://royalsocietypublishing.org/doi/full/10.1098/rspb.2018.2533}

\leavevmode\vadjust pre{\hypertarget{ref-borsanij.f.2023}{}}%
Borsani, J.F., Juretzek C., Klauson A., Leaper R., Le Courtois F.,
Liebschner A., Maglio A., Mueller A. , Norro A., Novellino A., Outinen
O., Popit A., Prospathopoulos A., Sigray P., Thomsen F., Tougaard J.,
Vukadin P., and Weilgart L., Borsani, J.F., Andersson M., André M.,
Azzellino A., Bou M., Castellote M., Ceyrac L., Dellong D., Folegot T.,
Hedgeland D., \& Juretzek C., Klauson A., Leaper R., Le Courtois F.,
Liebschner A., Maglio A., Mueller A. , Norro A., Novellino A., Outinen
O.,. (2023). \emph{Setting EU threshold values for continuous underwater
sound, technical group on underwater noise (TG NOISE)}.

\leavevmode\vadjust pre{\hypertarget{ref-breeze2022}{}}%
Breeze, H., Nolet, V., Thomson, D., Wright, A. J., Marotte, E., \&
Sanders, M. (2022). Efforts to advance underwater noise management in
Canada: Introduction to the Marine Pollution Bulletin Special Issue.
\emph{Marine Pollution Bulletin}, \emph{178}, 113596.
\url{https://doi.org/10.1016/j.marpolbul.2022.113596}

\leavevmode\vadjust pre{\hypertarget{ref-buxton2017}{}}%
Buxton, R. T., McKenna, M. F., Mennitt, D., Fristrup, K., Crooks, K.,
Angeloni, L., \& Wittemyer, G. (2017). Noise pollution is pervasive in
u.s. Protected areas. \emph{Science}, \emph{356}(6337), 531--533.
\url{https://doi.org/10.1126/science.aah4783}

\leavevmode\vadjust pre{\hypertarget{ref-chahouri2022}{}}%
Chahouri, A., Elouahmani, N., \& Ouchene, H. (2022). Recent progress in
marine noise pollution: A thorough review. \emph{Chemosphere},
\emph{291}, 132983.
\url{https://doi.org/10.1016/j.chemosphere.2021.132983}

\leavevmode\vadjust pre{\hypertarget{ref-cholewiak2017}{}}%
Cholewiak, D., DeAngelis, A. I., Palka, D., Corkeron, P. J., \& Van
Parijs, S. M. (2017). Beaked whales demonstrate a marked acoustic
response to the use of shipboard echosounders. \emph{Royal Society Open
Science}, \emph{4}(12), 170940.
\url{https://doi.org/10.1098/rsos.170940}

\leavevmode\vadjust pre{\hypertarget{ref-chou2021}{}}%
Chou, E., Southall, B. L., Robards, M., \& Rosenbaum, H. C. (2021).
International policy, recommendations, actions and mitigation efforts of
anthropogenic underwater noise. \emph{Ocean \& Coastal Management},
\emph{202}, 105427.
\url{https://doi.org/10.1016/j.ocecoaman.2020.105427}

\leavevmode\vadjust pre{\hypertarget{ref-coleman2008}{}}%
Coleman, K. (2008). Research review of collaborative ecosystem-based
management in the california current large marine ecosystem.
\emph{Coastal Management}, \emph{36}(5), 484--494.
\url{https://doi.org/10.1080/08920750802395541}

\leavevmode\vadjust pre{\hypertarget{ref-cooperman2022}{}}%
Cooperman, A., Duffy, P., Hall, M., Lozon, E., Shields, M., \& Musail,
W. (2022). \emph{Assessment of offshore wind energy leasing areas for
humboldt and morro bay wind energy areas, california}. Retrieved from
\url{https://www.nrel.gov/docs/fy22osti/82341.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-cox2006}{}}%
Cox, T. M., Ragen, T. J., Read, A. J., Vos, E., Baird, R. W., Balcomb,
K., \& Benner, L. (2006). Understanding the impacts of anthropogenic
sound on beaked whales. \emph{J. Cetacean Res. Manage}.

\leavevmode\vadjust pre{\hypertarget{ref-croll2001}{}}%
Croll, D. A., Clark, C. W., Calambokidis, J., Ellison, W. T., \& Tershy,
B. R. (2001). Effect of anthropogenic low-frequency noise on the
foraging ecology of Balaenoptera whales. \emph{Animal Conservation},
\emph{4}(1), 13--27. \url{https://doi.org/10.1017/S1367943001001020}

\leavevmode\vadjust pre{\hypertarget{ref-curuxe92013}{}}%
Curé, C., Antunes, R., Alves, A. C., Visser, F., Kvadsheim, P. H., \&
Miller, P. J. O. (2013). Responses of male sperm whales (Physeter
macrocephalus) to killer whale sounds: implications for anti-predator
strategies. \emph{Scientific Reports}, \emph{3}(1), 1579.
\url{https://doi.org/10.1038/srep01579}

\leavevmode\vadjust pre{\hypertarget{ref-deruiter2013}{}}%
DeRuiter, S. L., Southall, B. L., Calambokidis, J., Zimmer, W. M. X.,
Sadykova, D., Falcone, E. A., \ldots{} Tyack, P. L. (2013). First direct
measurements of behavioural responses by cuvier's beaked whales to
mid-frequency active sonar. \emph{Biology Letters}, \emph{9}(4),
20130223. \url{https://doi.org/10.1098/rsbl.2013.0223}

\leavevmode\vadjust pre{\hypertarget{ref-directiv2008}{}}%
\emph{Directive 2008/56/EC of the European Parliament and of the Council
of 17 June 2008 establishing a framework for community action in the
field of marine environmental policy (Marine Strategy Framework
Directive) (Text with EEA relevance)}. (2008). Retrieved from
\url{http://data.europa.eu/eli/dir/2008/56/oj/eng}

\leavevmode\vadjust pre{\hypertarget{ref-duarte2021}{}}%
Duarte, C. M., Chapuis, L., Collin, S. P., Costa, D. P., Devassy, R. P.,
Eguiluz, V. M., \ldots{} Juanes, F. (2021). The soundscape of the
anthropocene ocean. \emph{Science}, \emph{371}(6529), eaba4658.
\url{https://doi.org/10.1126/science.aba4658}

\leavevmode\vadjust pre{\hypertarget{ref-erbe2014}{}}%
Erbe, C., Williams, R., Sandilands, D., \& Ashe, E. (2014). Identifying
Modeled Ship Noise Hotspots for Marine Mammals of Canada's Pacific
Region. \emph{PLOS ONE}, \emph{9}(3), e89820.
\url{https://doi.org/10.1371/journal.pone.0089820}

\leavevmode\vadjust pre{\hypertarget{ref-fahlman2014}{}}%
Fahlman, A., Tyack, P., Miller, P., \& Kvadsheim, P. (2014). How
man-made interference might cause gas bubble emboli in deep diving
whales. \emph{Frontiers in Physiology}, \emph{5}. Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fphys.2014.00013}

\leavevmode\vadjust pre{\hypertarget{ref-fais2015}{}}%
Fais, A., Aguilar Soto, N., Johnson, M., Pérez-González, C., Miller, P.
J. O., \& Madsen, P. T. (2015). Sperm whale echolocation behaviour
reveals a directed, prior-based search strategy informed by prey
distribution. \emph{Behavioral Ecology and Sociobiology}, \emph{69}(4),
663--674. \url{https://doi.org/10.1007/s00265-015-1877-1}

\leavevmode\vadjust pre{\hypertarget{ref-fais2016}{}}%
Fais, A., Johnson, M., Wilson, M., Aguilar Soto, N., \& Madsen, P. T.
(2016). Sperm whale predator-prey interactions involve chasing and
buzzing, but no acoustic stunning. \emph{Scientific Reports},
\emph{6}(1), 28562. \url{https://doi.org/10.1038/srep28562}

\leavevmode\vadjust pre{\hypertarget{ref-falcone2017}{}}%
Falcone, E. A., Schorr, G. S., Watwood, S. L., DeRuiter, S. L., Zerbini,
A. N., Andrews, R. D., \ldots{} Moretti, D. J. (2017). Diving behaviour
of cuvier's beaked whales exposed to two types of military sonar.
\emph{Royal Society Open Science}, \emph{4}(8), 170629.
\url{https://doi.org/10.1098/rsos.170629}

\leavevmode\vadjust pre{\hypertarget{ref-federalregister2003}{}}%
Federal Register. (2003). \emph{Automatic identification system; vessel
carriage requirement}. Retrieved from
\url{https://www.federalregister.gov/documents/2003/07/01/03-16191/automatic-identification-system-vessel-carriage-requirement}

\leavevmode\vadjust pre{\hypertarget{ref-fernuxe1ndez2005}{}}%
Fernández, A., Edwards, J. F., Rodríguez, F., Monteros, A. E. de los,
Herráez, P., Castro, P., \ldots{} Arbelo, M. (2005). {``}Gas and Fat
Embolic Syndrome{''} Involving a Mass Stranding of Beaked Whales (Family
Ziphiidae) Exposed to Anthropogenic Sonar Signals. \emph{Veterinary
Pathology}, \emph{42}(4), 446--457.
\url{https://doi.org/10.1354/vp.42-4-446}

\leavevmode\vadjust pre{\hypertarget{ref-fiedler2023}{}}%
Fiedler, P. C., Becker, E. A., Forney, K. A., Barlow, J., \& Moore, J.
E. (2023). Species distribution modeling of deep-diving cetaceans.
\emph{Marine Mammal Science}, \emph{39}(4), 1178--1203.
\url{https://doi.org/10.1111/mms.13057}

\leavevmode\vadjust pre{\hypertarget{ref-field2006}{}}%
Field, J. C., \& Francis, R. C. (2006). Considering ecosystem-based
fisheries management in the california current. \emph{Marine Policy},
\emph{30}(5), 552--569.
\url{https://doi.org/10.1016/j.marpol.2005.07.004}

\leavevmode\vadjust pre{\hypertarget{ref-fleishman2023}{}}%
Fleishman, E., Cholewiak, D., Gillespie, D., Heble, T., Klinck, H.,
Nosal, E.-M., \& Roch, M. A. (2023). Ecological inferences about marine
mammals from passive acoustic data. \emph{Biological Reviews}.
https://doi.org/\url{https://doi.org/10.1111/brv.12969}

\leavevmode\vadjust pre{\hypertarget{ref-forney2012}{}}%
Forney, K. A., Ferguson, M. C., Becker, E. A., Fiedler, P. C., Redfern,
J. V., Barlow, J., \ldots{} Ballance, L. T. (2012). Habitat-based
spatial models of cetacean density in the eastern Pacific Ocean.
\emph{Endangered Species Research}, \emph{16}(2), 113--133.
\url{https://doi.org/10.3354/esr00393}

\leavevmode\vadjust pre{\hypertarget{ref-francis2012}{}}%
Francis, C. D., Kleist, N. J., Ortega, C. P., \& Cruz, A. (2012). Noise
pollution alters ecological services: Enhanced pollination and disrupted
seed dispersal. \emph{Proceedings of the Royal Society B: Biological
Sciences}, \emph{279}(1739), 2727--2735.
\url{https://doi.org/10.1098/rspb.2012.0230}

\leavevmode\vadjust pre{\hypertarget{ref-frisk2012}{}}%
Frisk, G. V. (2012). Noiseonomics: The relationship between ambient
noise levels in the sea and global economic trends. \emph{Scientific
Reports}, \emph{2}(1), 437. \url{https://doi.org/10.1038/srep00437}

\leavevmode\vadjust pre{\hypertarget{ref-garrett2016}{}}%
Garrett, J. K., Blondel, Ph., Godley, B. J., Pikesley, S. K., Witt, M.
J., \& Johanning, L. (2016). Long-term underwater sound measurements in
the shipping noise indicator bands 63Hz and 125Hz from the port of
Falmouth Bay, UK. \emph{Marine Pollution Bulletin}, \emph{110}(1),
438--448. \url{https://doi.org/10.1016/j.marpolbul.2016.06.021}

\leavevmode\vadjust pre{\hypertarget{ref-greig2020}{}}%
Greig, N. C., Hines, E. M., Cope, S., \& Liu, X. (2020). Using satellite
AIS to analyze vessel speeds off the coast of washington state, u.s., as
a risk analysis for cetacean-vessel collisions. \emph{Frontiers in
Marine Science}, \emph{7}. Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fmars.2020.00109}

\leavevmode\vadjust pre{\hypertarget{ref-harvey}{}}%
Harvey, C. J., Garfield, N., Williams, G. D. (Gregory. D., Tolimieri,
N., Schroeder, I., Andrews, K. S., \ldots{} Juhasz, C. (n.d.).
\emph{Ecosystem status report of the california current for 2019: A
summary of ecosystem indicators compiled by the california current
integrated ecosystem assessment team (CCEIA)}.
\url{https://doi.org/10.25923/P0ED-KE21}

\leavevmode\vadjust pre{\hypertarget{ref-hatch2008}{}}%
Hatch, L., Clark, C., Merrick, R., Van Parijs, S., Ponirakis, D.,
Schwehr, K., \ldots{} Wiley, D. (2008). Characterizing the Relative
Contributions of Large Vessels to Total Ocean Noise Fields: A Case Study
Using the Gerry E. Studds Stellwagen Bank National Marine Sanctuary.
\emph{Environmental Management}, \emph{42}(5), 735--752.
\url{https://doi.org/10.1007/s00267-008-9169-4}

\leavevmode\vadjust pre{\hypertarget{ref-haver2021}{}}%
Haver, S. M., Adams, J. D., Hatch, L. T., Van Parijs, S. M., Dziak, R.
P., Haxel, J., \ldots{} Gedamke, J. (2021). Large vessel activity and
low-frequency underwater sound benchmarks in united states waters.
\emph{Frontiers in Marine Science}, \emph{8}. Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fmars.2021.669528}

\leavevmode\vadjust pre{\hypertarget{ref-haver2019}{}}%
Haver, S. M., Fournet, M. E. H., Dziak, R. P., Gabriele, C., Gedamke,
J., Hatch, L. T., \ldots{} Van Parijs, S. M. (2019). Comparing the
underwater soundscapes of four u.s. National parks and marine
sanctuaries. \emph{Frontiers in Marine Science}, \emph{6}. Retrieved
from \url{https://www.frontiersin.org/articles/10.3389/fmars.2019.00500}

\leavevmode\vadjust pre{\hypertarget{ref-haver2020}{}}%
Haver, S. M., Rand, Z., Hatch, L. T., Lipski, D., Dziak, R. P., Gedamke,
J., \ldots{} Van Parijs, S. M. (2020). Seasonal trends and primary
contributors to the low-frequency soundscape of the cordell bank
national marine sanctuarya). \emph{The Journal of the Acoustical Society
of America}, \emph{148}(2), 845--858.
\url{https://doi.org/10.1121/10.0001726}

\leavevmode\vadjust pre{\hypertarget{ref-hawkins2017}{}}%
Hawkins, A. D., \& Popper, A. N. (2017). A sound approach to assessing
the impact of underwater noise on marine fishes and invertebrates.
\emph{ICES Journal of Marine Science}, \emph{74}(3), 635--651.
\url{https://doi.org/10.1093/icesjms/fsw205}

\leavevmode\vadjust pre{\hypertarget{ref-hermannsen2019}{}}%
Hermannsen, L., Mikkelsen, L., Tougaard, J., Beedholm, K., Johnson, M.,
\& Madsen, P. T. (2019). Recreational vessels without automatic
identification system (AIS) dominate anthropogenic noise contributions
to a shallow water soundscape. \emph{Scientific Reports}, \emph{9},
15477. \url{https://doi.org/10.1038/s41598-019-51222-9}

\leavevmode\vadjust pre{\hypertarget{ref-hildebrand2009}{}}%
Hildebrand, J. A. (2009). Anthropogenic and natural sources of ambient
noise in the ocean. \emph{Marine Ecology Progress Series}, \emph{395},
5--20. \url{https://doi.org/10.3354/meps08353}

\leavevmode\vadjust pre{\hypertarget{ref-hildebrand2015}{}}%
Hildebrand, J. A., Baumann-Pickering, S., Frasier, K. E., Trickey, J.
S., Merkens, K. P., Wiggins, S. M., \ldots{} Thomas, L. (2015). Passive
acoustic monitoring of beaked whale densities in the Gulf of Mexico.
\emph{Scientific Reports}, \emph{5}(1), 16343.
\url{https://doi.org/10.1038/srep16343}

\leavevmode\vadjust pre{\hypertarget{ref-hooker2019}{}}%
Hooker, S. K., De Soto, N. A., Baird, R. W., Carroll, E. L., Claridge,
D., Feyrer, L., \ldots{} Whitehead, H. (2019). Future directions in
research on beaked whales. \emph{Frontiers in Marine Science}, \emph{5}.
Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fmars.2018.00514}

\leavevmode\vadjust pre{\hypertarget{ref-internationalmaritimeorganization2014}{}}%
International Maritime Organization. (2014). \emph{GUIDELINES FOR THE
REDUCTION OF UNDERWATER NOISE FROM COMMERCIAL SHIPPING TO ADDRESS
ADVERSE IMPACTS ON MARINE LIFE}.

\leavevmode\vadjust pre{\hypertarget{ref-isojunno2016}{}}%
Isojunno, S., Curé, C., Kvadsheim, P. H., Lam, F.-P. A., Tyack, P. L.,
Wensveen, P. J., \& Miller, P. J. O. (2016). Sperm whales reduce
foraging effort during exposure to 1{\textendash}2 kHz sonar and killer
whale sounds. \emph{Ecological Applications}, \emph{26}(1), 77--93.
\url{https://doi.org/10.1890/15-0040}

\leavevmode\vadjust pre{\hypertarget{ref-isojunno2022}{}}%
Isojunno, S., von Benda-Beckmann, A. M., Wensveen, P. J., Kvadsheim, P.
H., Lam, F.-P. A., Gkikopoulou, K. C., \ldots{} Miller, P. J. O. (2022).
Sperm whales exhibit variation in echolocation tactics with depth and
sea state but not naval sonar exposures. \emph{Marine Mammal Science},
\emph{38}(2), 682--704. \url{https://doi.org/10.1111/mms.12890}

\leavevmode\vadjust pre{\hypertarget{ref-jarvis2022}{}}%
Jarvis, S. M., DiMarzio, N., Watwood, S., Dolan, K., \& Morrissey, R.
(2022). Automated detection and classification of beaked whale buzzes on
bottom-mounted hydrophones. \emph{Frontiers in Remote Sensing},
\emph{3}. Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/frsen.2022.941838}

\leavevmode\vadjust pre{\hypertarget{ref-johnson2009}{}}%
Johnson, M., Soto, N. A. de, \& Madsen, P. T. (2009). Studying the
behaviour and sensory ecology of marine mammals using acoustic recording
tags: A review. \emph{Marine Ecology Progress Series}, \emph{395},
55--73. Retrieved from \url{https://www.jstor.org/stable/24874241}

\leavevmode\vadjust pre{\hypertarget{ref-joyce2020}{}}%
Joyce, T. W., Durban, J. W., Claridge, D. E., Dunn, C. A., Hickmott, L.
S., Fearnbach, H., \ldots{} Moretti, D. (2020). Behavioral responses of
satellite tracked Blainville's beaked whales (Mesoplodon densirostris)
to mid-frequency active sonar. \emph{Marine Mammal Science},
\emph{36}(1), 29--46. \url{https://doi.org/10.1111/mms.12624}

\leavevmode\vadjust pre{\hypertarget{ref-kunc2016}{}}%
Kunc, H. P., McLaughlin, K. E., \& Schmidt, R. (2016). Aquatic noise
pollution: Implications for individuals, populations, and ecosystems.
\emph{Proceedings of the Royal Society B: Biological Sciences},
\emph{283}(1836), 20160839. \url{https://doi.org/10.1098/rspb.2016.0839}

\leavevmode\vadjust pre{\hypertarget{ref-kvadsheim2012}{}}%
Kvadsheim, P., Miller, P., Tyack, P., Sivle, L., Lam, F.-P., \& Fahlman,
A. (2012). Estimated tissue and blood N2 levels and risk of in vivo
bubble formation in deep-, intermediate- and shallow diving toothed
whales during exposure to naval sonar. \emph{Frontiers in Physiology},
\emph{3}. Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fphys.2012.00125}

\leavevmode\vadjust pre{\hypertarget{ref-lillis2018}{}}%
Lillis, A., Caruso, F., Mooney, T. A., Llopiz, J., Bohnenstiehl, D., \&
Eggleston, D. B. (2018). Drifting hydrophones as an ecologically
meaningful approach to underwater soundscape measurement in coastal
benthic habitats. \emph{JEA}, \emph{2}(1), 1--1.
\url{https://doi.org/10.22261/JEA.STBDH1}

\leavevmode\vadjust pre{\hypertarget{ref-madsen2002}{}}%
Madsen, P. T., Mohl, B., Nielsen, B. K., \& Wahlberg, M. (2002). Male
sperm whale behaviour during exposures to distant seismic survey pulses.
\emph{Aquatic Mammals}.

\leavevmode\vadjust pre{\hypertarget{ref-madsen2006}{}}%
Madsen, P. T., Wahlberg, M., Tougaard, J., Lucke, K., \& Tyack, P.
(2006). Wind turbine underwater noise and marine mammals: Implications
of current knowledge and data needs. \emph{Marine Ecology Progress
Series}, \emph{309}, 279--295. Retrieved from
\url{https://www.jstor.org/stable/24869996}

\leavevmode\vadjust pre{\hypertarget{ref-mannocci2015}{}}%
Mannocci, L., Monestiez, P., Spitz, J., \& Ridoux, V. (2015).
Extrapolating cetacean densities beyond surveyed regions: habitat-based
predictions in the circumtropical belt. \emph{Journal of Biogeography},
\emph{42}(7), 1267--1280. \url{https://doi.org/10.1111/jbi.12530}

\leavevmode\vadjust pre{\hypertarget{ref-manzano-roth2023}{}}%
Manzano-Roth, R., Henderson, E. E., Alongi, G. C., Martin, C. R.,
Martin, S. W., \& Matsuyama, B. (2023). Dive characteristics of Cross
Seamount beaked whales from long-term passive acoustic monitoring at the
Pacific Missile Range Facility, Kauaʻi. \emph{Marine Mammal Science},
\emph{39}(1), 22--41. \url{https://doi.org/10.1111/mms.12959}

\leavevmode\vadjust pre{\hypertarget{ref-marcoux2006}{}}%
Marcoux, M., Whitehead, H., \& Rendell, L. (2006). Coda vocalizations
recorded in breeding areas are almost entirely produced by mature female
sperm whales (physeter macrocephalus). \emph{Canadian Journal of
Zoology}, \emph{84}(4), 609--614. \url{https://doi.org/10.1139/z06-035}

\leavevmode\vadjust pre{\hypertarget{ref-markus2018}{}}%
Markus, T., \& Sánchez, P. P. S. (2018). \emph{Managing and Regulating
Underwater Noise Pollution} (M. Salomon and T. Markus, Eds.).
\url{https://doi.org/10.1007/978-3-319-60156-4_52}

\leavevmode\vadjust pre{\hypertarget{ref-marques2012}{}}%
Marques, T. A., Thomas, L., Martin, S. W., Mellinger, D. K., Jarvis, S.,
Morrissey, R. P., \ldots{} DiMarzio, N. (2012). Spatially explicit
capture{\textendash}recapture methods to estimate minke whale density
from data collected at bottom-mounted hydrophones. \emph{Journal of
Ornithology}, \emph{152}(2), 445--455.
\url{https://doi.org/10.1007/s10336-010-0535-7}

\leavevmode\vadjust pre{\hypertarget{ref-mccullough2021}{}}%
McCullough, J. L. K., Wren, J. L. K., Oleson, E. M., Allen, A. N.,
Siders, Z. A., \& Norris, E. S. (2021). An acoustic survey of beaked
whales and kogia spp. In the mariana archipelago using drifting
recorders. \emph{Frontiers in Marine Science}, \emph{8}. Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fmars.2021.664292}

\leavevmode\vadjust pre{\hypertarget{ref-mellinger2007}{}}%
Mellinger, D. K., Stafford, K. M., Moore, S. E., Dziak, R. P., \&
Matsumoto, H. (2007). An overview of fixed passive acoustic observation
methods for cetaceans. \emph{Oceanography}, \emph{20}(4), 3645.
https://doi.org/\url{https://doi.org/10.5670/oceanog.2007.03}

\leavevmode\vadjust pre{\hypertarget{ref-miksis-olds2018}{}}%
Miksis-Olds, J. L., Martin, B., \& Tyack, P. L. (2018). Exploring the
ocean through soundscapes. \emph{Acoustics Today}, \emph{14}(1), 26--34.

\leavevmode\vadjust pre{\hypertarget{ref-miller2022}{}}%
Miller, Patrick J. O., Isojunno, S., Siegal, E., Lam, F.-P. A.,
Kvadsheim, P. H., \& Curé, C. (2022). Behavioral responses to predatory
sounds predict sensitivity of cetaceans to anthropogenic noise within a
soundscape of fear. \emph{Proceedings of the National Academy of
Sciences}, \emph{119}(13), e2114932119.
\url{https://doi.org/10.1073/pnas.2114932119}

\leavevmode\vadjust pre{\hypertarget{ref-miller2009}{}}%
Miller, P. J. O., Johnson, M. P., Madsen, P. T., Biassoni, N., Quero,
M., \& Tyack, P. L. (2009). Using at-sea experiments to study the
effects of airguns on the foraging behavior of sperm whales in the Gulf
of Mexico. \emph{Deep Sea Research Part I: Oceanographic Research
Papers}, \emph{56}(7), 1168--1181.
\url{https://doi.org/10.1016/j.dsr.2009.02.008}

\leavevmode\vadjust pre{\hypertarget{ref-moore2021}{}}%
Moore, J. E. (2021). \emph{Final report of the california current
ecosystem survey (CCES) 2018: A PacMAPPS study} (p. 187). Camarillo
(CA): US Department of the Interior, Bureau of Ocean Energy Management.

\leavevmode\vadjust pre{\hypertarget{ref-mustonen2019}{}}%
Mustonen, M., Klauson, A., Andersson, M., Clorennec, D., Folegot, T.,
Koza, R., \ldots{} Sigray, P. (2019). Spatial and Temporal Variability
of Ambient Underwater Sound in the Baltic Sea. \emph{Scientific
Reports}, \emph{9}(1), 13237.
\url{https://doi.org/10.1038/s41598-019-48891-x}

\leavevmode\vadjust pre{\hypertarget{ref-new2013}{}}%
New, L. F., Moretti, D. J., Hooker, S. K., Costa, D. P., \& Simmons, S.
E. (2013). Using Energetic Models to Investigate the Survival and
Reproduction of Beaked Whales (family Ziphiidae). \emph{PLOS ONE},
\emph{8}(7), e68725. \url{https://doi.org/10.1371/journal.pone.0068725}

\leavevmode\vadjust pre{\hypertarget{ref-noaa2018}{}}%
NOAA. (2018). \emph{2018 revisions to: Technical guidance for assessing
the effects of anthropogenic sound on marine mammal hearing (version
2.0): Underwater thresholds for onset of permanent and temporary
threshold shifts}.

\leavevmode\vadjust pre{\hypertarget{ref-noaafisheries}{}}%
NOAA Fisheries, N. (n.d.). \emph{Species directory \textbar{} NOAA
fisheries}. Retrieved from
\url{https://www.fisheries.noaa.gov/species-directory?oq=\&field_species_categories_vocab=54\&field_region_vocab=1000001126\&items_per_page=350}

\leavevmode\vadjust pre{\hypertarget{ref-norris1972}{}}%
Norris, K. S., \& Harvey, G. W. (1972, January 1). \emph{A theory for
the function of the spermaceti organ of the sperm whale (physeter
catodon l.)}. Retrieved from
\url{https://ntrs.nasa.gov/citations/19720017437}

\leavevmode\vadjust pre{\hypertarget{ref-nosal2007}{}}%
Nosal, E.-M., \& Frazer, L. N. (2007). Sperm whale three-dimensional
track, swim orientation, beam pattern, and click levels observed on
bottom-mounted hydrophonesa). \emph{The Journal of the Acoustical
Society of America}, \emph{122}(4), 1969--1978.
\url{https://doi.org/10.1121/1.2775423}

\leavevmode\vadjust pre{\hypertarget{ref-officeofnationalmarinesanctuaries}{}}%
Office of National Marine Sanctuaries. (n.d.). \emph{West coast region}.
Retrieved from \url{https://sanctuaries.noaa.gov/about/westcoast.html}

\leavevmode\vadjust pre{\hypertarget{ref-oldach2022}{}}%
Oldach, E., Killeen, H., Shukla, P., Brauer, E., Carter, N., Fields, J.,
\ldots{} Fangue, N. (2022). Managed and unmanaged whale mortality in the
california current ecosystem. \emph{Marine Policy}, \emph{140}, 105039.
\url{https://doi.org/10.1016/j.marpol.2022.105039}

\leavevmode\vadjust pre{\hypertarget{ref-parsons2016}{}}%
Parsons, E. C. M. (2016). Why IUCN should replace {``}data deficient{''}
conservation status with a precautionary {``}assume threatened{''}
status{\textemdash}a cetacean case study. \emph{Frontiers in Marine
Science}, \emph{3}. Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fmars.2016.00193}

\leavevmode\vadjust pre{\hypertarget{ref-peuxf1a2007}{}}%
Peña, M. A., \& Bograd, S. J. (2007). Time series of the northeast
pacific. \emph{Progress in Oceanography}, \emph{75}(2), 115--119.
\url{https://doi.org/10.1016/j.pocean.2007.08.008}

\leavevmode\vadjust pre{\hypertarget{ref-pijanowski2011}{}}%
Pijanowski, B. C., Farina, A., Gage, S. H., Dumyahn, S. L., \& Krause,
B. L. (2011). What is soundscape ecology? An introduction and overview
of an emerging new science. \emph{Landscape Ecology}, \emph{26}(9),
1213--1232. \url{https://doi.org/10.1007/s10980-011-9600-8}

\leavevmode\vadjust pre{\hypertarget{ref-pirotta2012}{}}%
Pirotta, E., Milor, R., Quick, N., Moretti, D., Marzio, N. D., Tyack,
P., \ldots{} Hastie, G. (2012). Vessel Noise Affects Beaked Whale
Behavior: Results of a Dedicated Acoustic Response Study. \emph{PLOS
ONE}, \emph{7}(8), e42535.
\url{https://doi.org/10.1371/journal.pone.0042535}

\leavevmode\vadjust pre{\hypertarget{ref-pitman}{}}%
Pitman, R. L., \& Brownell, R. L. (n.d.). Mesoplodon carlhubbsi.
\emph{The IUCN Red List of Threatened Species 2020: E.T13243A50364109}.
https://doi.org/\url{https://dx.doi.org/10.2305/IUCN.UK.2020-3.RLTS.T13243A50364109.en.}

\leavevmode\vadjust pre{\hypertarget{ref-pitman2020}{}}%
Pitman, R. L., \& Brownell, R. L. (2020). Mesoplodon ginkgodens.
\emph{The IUCN Red List of Threatened Species 2020:
E.T127827012A127827154}.
https://doi.org/\url{https://dx.doi.org/10.2305/IUCN.UK.2020-3.RLTS.T127827012A127827154.en.}

\leavevmode\vadjust pre{\hypertarget{ref-radford2010}{}}%
Radford, C. A., Stanley, J. A., Tindle, C. T., Montgomery, J. C., \&
Jeffs, A. G. (2010). Localised coastal habitats have distinct underwater
sound signatures. \emph{Marine Ecology Progress Series}, \emph{401},
21--29. \url{https://doi.org/10.3354/meps08451}

\leavevmode\vadjust pre{\hypertarget{ref-rand2022}{}}%
Rand, Z. R., Wood, J. D., \& Oswald, J. N. (2022). Effects of duty
cycles on passive acoustic monitoring of southern resident killer whale
(orcinus orca) occurrence and behavior. \emph{The Journal of the
Acoustical Society of America}, \emph{151}(3), 1651--1660.
\url{https://doi.org/10.1121/10.0009752}

\leavevmode\vadjust pre{\hypertarget{ref-reimer2016}{}}%
Reimer, J., Gravel, C., Brown, M. W., \& Taggart, C. T. (2016).
Mitigating vessel strikes: The problem of the peripatetic whales and the
peripatetic fleet. \emph{Marine Policy}, \emph{68}, 91--99.
\url{https://doi.org/10.1016/j.marpol.2016.02.017}

\leavevmode\vadjust pre{\hypertarget{ref-richardson2013}{}}%
Richardson, W. J., Jr, C. R. G., Malme, C. I., \& Thomson, D. H. (2013).
\emph{Marine Mammals and Noise}. Academic Press.

\leavevmode\vadjust pre{\hypertarget{ref-robbins2022}{}}%
Robbins, J. R., Bell, E., Potts, J., Babey, L., \& Marley, S. A. (2022).
Likely year-round presence of beaked whales in the Bay of Biscay.
\emph{Hydrobiologia}, \emph{849}(10), 2225--2239.
\url{https://doi.org/10.1007/s10750-022-04822-y}

\leavevmode\vadjust pre{\hypertarget{ref-ryther1969}{}}%
Ryther, J. H. (1969). Photosynthesis and fish production in the sea.
\emph{Science}, \emph{166}(3901), 72--76. Retrieved from
\url{https://www.jstor.org/stable/1727735}

\leavevmode\vadjust pre{\hypertarget{ref-sebastianutto2015}{}}%
Sebastianutto, L., Fortuna, C. M., Mackelworth, P., Holcer, D., \& Rako
Gospić, N. (2015, November 27). \emph{Popper, a. N. And hawkins, a. Eds.
(2015). The effects of noise on aquatic life II. Springer
science+business media, LLC, new york.} \emph{875}.
\url{https://doi.org/10.1007/978-1-4939-2981-8_101}

\leavevmode\vadjust pre{\hypertarget{ref-shannon2016}{}}%
Shannon, G., McKenna, M. F., Angeloni, L. M., Crooks, K. R., Fristrup,
K. M., Brown, E., \ldots{} Wittemyer, G. (2016). A synthesis of two
decades of research documenting the effects of noise on wildlife.
\emph{Biological Reviews}, \emph{91}(4), 982--1005.
\url{https://doi.org/10.1111/brv.12207}

\leavevmode\vadjust pre{\hypertarget{ref-simonis2020}{}}%
Simonis, A. E., Brownell, R. L., Thayre, B. J., Trickey, J. S., Oleson,
E. M., Huntington, R., \& Baumann-Pickering, S. (2020). Co-occurrence of
beaked whale strandings and naval sonar in the mariana islands, western
pacific. \emph{Proceedings of the Royal Society B: Biological Sciences},
\emph{287}(1921), 20200070. \url{https://doi.org/10.1098/rspb.2020.0070}

\leavevmode\vadjust pre{\hypertarget{ref-simonis2020a}{}}%
Simonis, A. E., Trickey, J. S., Barlow, J. P., Rankin, S., Urban, J.,
Rojas-Bracho, L., \& Moore, J. E. (2020). \emph{Passive acoustic survey
of deep-diving odontocetes in the california current ecosystem 2018.}

\leavevmode\vadjust pre{\hypertarget{ref-sivle2012}{}}%
Sivle, L., Kvadsheim, P., Fahlman, A., Lam, F.-P., Tyack, P., \& Miller,
P. (2012). Changes in dive behavior during naval sonar exposure in
killer whales, long-finned pilot whales, and sperm whales.
\emph{Frontiers in Physiology}, \emph{3}. Retrieved from
\url{https://www.frontiersin.org/articles/10.3389/fphys.2012.00400}

\leavevmode\vadjust pre{\hypertarget{ref-southall2017}{}}%
Southall, B. L., Scholik-Schlomer, A. R., Hatch, L., Bergmann, T.,
Jasny, M., Metcalf, K., \ldots{} Wright, A. J. (2017). \emph{Underwater
Noise from Large Commercial Ships{\textemdash}International
Collaboration for Noise Reduction}.
\url{https://doi.org/10.1002/9781118476406.emoe056}

\leavevmode\vadjust pre{\hypertarget{ref-stanistreet2022}{}}%
Stanistreet, J. E., Beslin, W. A. M., Kowarski, K., Martin, S. B.,
Westell, A., \& Moors-Murphy, H. B. (2022). Changes in the acoustic
activity of beaked whales and sperm whales recorded during a naval
training exercise off eastern Canada. \emph{Scientific Reports},
\emph{12}(1), 1973. \url{https://doi.org/10.1038/s41598-022-05930-4}

\leavevmode\vadjust pre{\hypertarget{ref-symonds2011}{}}%
Symonds, M. R. E., \& Moussalli, A. (2011). A brief guide to model
selection, multimodel inference and model averaging in behavioural
ecology using Akaike{'}s information criterion. \emph{Behavioral Ecology
and Sociobiology}, \emph{65}(1), 13--21.
\url{https://doi.org/10.1007/s00265-010-1037-6}

\leavevmode\vadjust pre{\hypertarget{ref-taylor}{}}%
Taylor, B. L., Baird, R. W., Barlow, J., Dawson, S. M., Ford, J., Mead,
J. G., \ldots{} Pitman, R. L. (n.d.). Physeter macrocephalus (amended
version of 2008 assessment). \emph{The IUCN Red List of Threatened
Species 2019: E.T41755A160983555}.
https://doi.org/\url{https://dx.doi.org/10.2305/IUCN.UK.2008.RLTS.T41755A160983555.en}

\leavevmode\vadjust pre{\hypertarget{ref-tyack2008}{}}%
Tyack, P. L. (2008). Implications for marine mammals of large-scale
changes in the marine acoustic environment. \emph{Journal of Mammalogy},
\emph{89}(3), 549--558. Retrieved from
\url{https://www.proquest.com/docview/221473276/abstract/CBFFD0F9B5AA4C54PQ/1}

\leavevmode\vadjust pre{\hypertarget{ref-tyack2006}{}}%
Tyack, P. L., Aguilar Soto, N., Johnson, M., Sturlese, A., \& Madsen, P.
T. (2006). Extreme diving of beaked whales \textbar{} journal of
experimental biology \textbar{} the company of biologists. \emph{Journal
of Experimental Biology}, \emph{209}, 4238--4253.
\url{https://doi.org/10.1242/jeb.02505}

\leavevmode\vadjust pre{\hypertarget{ref-unctad2022}{}}%
UNCTAD. (2022). \emph{Review of maritime transport 2022}.

\leavevmode\vadjust pre{\hypertarget{ref-unitedstatesarmycorpsofengineersusace2018}{}}%
United States Army Corps of Engineers {[}USACE{]}. (2018). \emph{The
u.s. Coastal and inland navigation system 2021 transportation facts \&
information}.

\leavevmode\vadjust pre{\hypertarget{ref-urick1983}{}}%
Urick, R. J. (1983). \emph{Principles of underwater sound} (3rd ed.).
New York : McGraw-Hill.

\leavevmode\vadjust pre{\hypertarget{ref-vagle2021}{}}%
Vagle, S., Burnham, R. E., O'Neill, C., \& Yurk, H. (2021). Variability
in Anthropogenic Underwater Noise Due to Bathymetry and Sound Speed
Characteristics. \emph{Journal of Marine Science and Engineering},
\emph{9}(10), 1047. \url{https://doi.org/10.3390/jmse9101047}

\leavevmode\vadjust pre{\hypertarget{ref-virgili2019}{}}%
Virgili, A., Authier, M., Boisseau, O., Cañadas, A., Claridge, D., Cole,
T., \ldots{} Ridoux, V. (2019). Combining multiple visual surveys to
model the habitat of deep-diving cetaceans at the basin scale.
\emph{Global Ecology and Biogeography}, \emph{28}(3), 300--314.
\url{https://doi.org/10.1111/geb.12850}

\leavevmode\vadjust pre{\hypertarget{ref-virgili2022}{}}%
Virgili, A., Teillard, V., Dorémus, G., Dunn, T. E., Laran, S., Lewis,
M., \ldots{} Ridoux, V. (2022). Deep ocean drivers better explain
habitat preferences of sperm whales Physeter macrocephalus than beaked
whales in the Bay of Biscay. \emph{Scientific Reports}, \emph{12}(1),
9620. \url{https://doi.org/10.1038/s41598-022-13546-x}

\leavevmode\vadjust pre{\hypertarget{ref-watkins1993}{}}%
Watkins, W. A., Daher, M. A., Fristrup, K. M., Howald, T. J., \& Di
Sciara, G. N. (1993). Sperm Whales Tagged with Transponders and Tracked
Underwater by Sonar. \emph{Marine Mammal Science}, \emph{9}(1), 55--67.
\url{https://doi.org/10.1111/j.1748-7692.1993.tb00426.x}

\leavevmode\vadjust pre{\hypertarget{ref-weatherall2015}{}}%
Weatherall, P., Marks, K. M., Jakobsson, M., Schmitt, T., Tani, S.,
Arndt, J. E., \ldots{} Wigley, R. (2015). A new digital bathymetric
model of the world's oceans. \emph{Earth and Space Science},
\emph{2}(8), 331--345. \url{https://doi.org/10.1002/2015EA000107}

\leavevmode\vadjust pre{\hypertarget{ref-weilgart2018}{}}%
Weilgart, L. (2018). \emph{The impact of ocean noise pollution on fish
and invertebrates} (p. 34).

\leavevmode\vadjust pre{\hypertarget{ref-weilgart2007}{}}%
Weilgart, L. S. (2007). The impacts of anthropogenic ocean noise on
cetaceans and implications for management. \emph{Can. J. Zool.},
\emph{85}, 10911116. Retrieved from
\url{https://cdnsciencepub.com/doi/abs/10.1139/z07-101}

\leavevmode\vadjust pre{\hypertarget{ref-weiss2021}{}}%
Weiss, S. G., Cholewiak, D., Frasier, K. E., Trickey, J. S.,
Baumann-Pickering, S., Hildebrand, J. A., \& Van Parijs, S. M. (2021).
Monitoring the acoustic ecology of the shelf break of Georges Bank,
Northwestern Atlantic Ocean: New approaches to visualizing complex
acoustic data. \emph{Marine Policy}, \emph{130}, 104570.
\url{https://doi.org/10.1016/j.marpol.2021.104570}

\leavevmode\vadjust pre{\hypertarget{ref-west2017}{}}%
West, K. L., Walker, W. A., Baird, R. W., Mead, J. G., \& Collins, P. W.
(2017). Diet of cuvier{'}s beaked whales ziphius cavirostris from the
north pacific and a comparison with their diet world-wide. \emph{Marine
Ecology Progress Series}, \emph{574}, 227--242. Retrieved from
\url{https://www.jstor.org/stable/26403641}

\leavevmode\vadjust pre{\hypertarget{ref-winsor2017}{}}%
Winsor, M. H., Irvine, L. M., \& Mate, B. R. (2017). Analysis of the
spatial distribution of satellite-tagged sperm whales (physeter
macrocephalus) in close proximity to seismic surveys in the gulf of
mexico. \emph{Aquatic Mammals}, \emph{43}(4), 439--446.
\url{https://doi.org/10.1578/AM.43.4.2017.439}

\leavevmode\vadjust pre{\hypertarget{ref-yamada2019}{}}%
Yamada, T. K., Kitamura, S., Abe, S., Tajima, Y., Matsuda, A., Mead, J.
G., \& Matsuishi, T. F. (2019). Description of a new species of beaked
whale (Berardius) found in the North Pacific. \emph{Scientific Reports},
\emph{9}(1), 12723. \url{https://doi.org/10.1038/s41598-019-46703-w}

\leavevmode\vadjust pre{\hypertarget{ref-ziegenhorn2023}{}}%
Ziegenhorn, M. A., Hildebrand, J. A., Oleson, E. M., Baird, R. W.,
Wiggins, S. M., \& Baumann-Pickering, S. (2023). Odontocete spatial
patterns and temporal drivers of detection at sites in the Hawaiian
islands. \emph{Ecology and Evolution}, \emph{13}(1), e9688.
\url{https://doi.org/10.1002/ece3.9688}

\end{CSLReferences}

\end{document}

### 2023-10-20 manuscript.rmd
Update manuscript.rmd

@@
 title: "manuscript"
-format: docx
 editor: visual
 bibliography: references.bib
-fontsize: 12pt
+output:
+  word_document:
+    reference_docx: "Masters-Thesis-and-Creative-Work-Template.docx"
+csl: "marine-mammal-science.csl"
+header-includes:
+- \setlength{\parindent}{4em}
+- \usepackage{setspace}\doublespacing
 ---
@@
 Since sound travels over four times faster in water than in air and can carry information over far distances [@urick1983], the use of sound has evolved as a critical component of communication, navigation, and foraging for many invertebrates, fishes, and cetaceans [@duarte2021].
-Under the right conditions, sounds can travel across ocean basins [@miksis-olds2018].
+Under the right conditions, sounds can travel across ocean basins [@miksis-olds2018].\
 
@@
 Every region of the ocean experiences high spatio-temporal variation in sound levels and sources based on traits such as properties of water, organism diversity, human activities, and bathymetry [@andré2011].
-Because of their unique sound signatures, ocean soundscapes can be characterized, measured, and monitored as indicators of ecosystem health or species diversity [@radford2010; @pijanowski2011; @haver2019; @weiss2021].
+Because of their unique sound signatures, ocean soundscapes can be characterized, measured, and monitored as indicators of ecosystem health or species diversity [@radford2010; @pijanowski2011; @haver2019; @weiss2021].\
 
@@
 Sensitivities and responses to noise can vary among species and even individuals [@kunc2016].
-While terrestrial ecosystem-level effects have been well-documented [@francis2012; @shannon2016; @buxton2017], effects within aquatic ecosystems require further studies [@kunc2016].
+While terrestrial ecosystem-level effects have been well-documented [@francis2012; @shannon2016; @buxton2017], effects within aquatic ecosystems require further studies [@kunc2016].\
 
@@
 Ships have been increasing in terms of gross tonnage and fleet size [@hildebrand2009] and now account for over 80% of the global trade, with an estimated 1.4% expected annual increase in shipping through 2027 [@unctad2022].
-Commercial shipping now accounts for most of the low-frequency anthropogenic noise in the ocean [@hildebrand2009].
+Commercial shipping now accounts for most of the low-frequency anthropogenic noise in the ocean [@hildebrand2009].\
 
@@
 Sonar for military surveillance or research such as seafloor mapping produces mid-frequency sounds, but attenuates quicker, therefore impacting a smaller acoustic space [@hildebrand2009; @richardson2013].
-Noise has been shown to propagate differently based on static features like bathymetry and depth [@richardson2013; @vagle2021], as well as dynamic variables like temperature, salinity, and pH from a changing climate [@kunc2016; @affatati2022].
+Noise has been shown to propagate differently based on static features like bathymetry and depth [@richardson2013; @vagle2021], as well as dynamic variables like temperature, salinity, and pH from a changing climate [@kunc2016; @affatati2022].\
 
@@
 It remains uncertain how short-term effects of chronic or acute noise pollution for cetaceans translates into long-term or population level changes.
-@new2013 suggest enough exposure can cause beaked whales to cease foraging and eventually lower reproduction rates.
+@new2013 suggest enough exposure can cause beaked whales to cease foraging and eventually lower reproduction rates.\
 
@@
 
-**\[New addition as of 08/25/23\] \[\[**Two SPOT buoys sit on every buoy, so some GPS points were recorded twice.
-These duplicate tracks were eliminated.\]\] ~~Buoy tracks were similarly rounded to nearest 20 minutes.~~ Seven buoys (1,2,3,5,6,9,11) were lost at sea and two buoys had corrupted data (4,17) so these were removed from the analysis.
+New addition as of 08/25/23 -\*\* Two SPOT buoys sit on every buoy, so some GPS points were recorded twice.
+These duplicate tracks were eliminated.~~Buoy tracks were similarly rounded to nearest 20 minutes.~~ Seven buoys (1,2,3,5,6,9,11) were lost at sea and two buoys had corrupted data (4,17) so these were removed from the analysis.
 Buoys 14 and 15 were the same and therefore consolidated to 14.
@@
 
-Possible environmental and oceanographic predictors of cetacean habitat were extracted from the NOAA Environmental Research Division Data Access Program (ERDDAP) server ((https://coastwatch.pfeg.noaa.gov/erddap/) and Hybrid Coordinate Ocean Model (HYCOM) using the *matchEnvData* function in R package *PAMmisc*.
+Possible environmental and oceanographic predictors of cetacean habitat were extracted from the NOAA Environmental Research Division Data Access Program (ERDDAP) server ((<https://coastwatch.pfeg.noaa.gov/erddap/>) and Hybrid Coordinate Ocean Model (HYCOM) using the *matchEnvData* function in R package *PAMmisc*.
 Six modeled variables at depth (mixed layer depth, mixed layer temperature, temperature at 400m, thermocline temperature, thermocline depth, salinity at 400m) were calculated using a formula from @mccullough2021.
@@
 
-\[See Virgili et al 2019 for extracting slope from GEBCO Data and dynamic variables!!\]
+See Virgili et al 2019 for extracting slope from GEBCO Data and dynamic variables!!
 
@@
 
--   Buoy 18 had the most whale detections
+*Overall Soundscape*
 
--   ZC was the most frequently detected whale
+Buoy 23 had the loudest median sound levels in the TOL 2 kHz range (80.0 dB), while buoy 7 experienced the single loudest sound level (108 dB) and the widest range of acoustic conditions in this frequency (range = 52.73 dB).
+The median quietest conditions and lowest single sound level were along buoy 16 (median = 72.5 dB, min = 51.6 dB).
 
--   The loudest sound levels were only occupied by PM
+*Acoustic Conditions with Whale Detections*
 
--   Density of ship encounters
+The most number of whale detections by all species were detected by buoy 18 (n = 209).
+Cuvier's beaked whale was the most frequently detected species (n = 494).
+BWC and BW37V had very low detections and were subsequently removed from analysis (n = 2 and n = 1, respectively).
+Sperm whales experienced the loudest sound conditions in the TOL at 2kHz frequency band (median = 81.9 dB , max = 98.3 dB).
 
--   Report on environmental conditions for ZC/PM?
+TOL at 2kHz conditions more closely aligned with sperm whale and Cuvier's beaked whale detections than with TOL at 125 Hz.
+
+*Interactions with Large Vessels*
+
+The most number of unique large (\>100 m) vessels transited within the space of buoy 14, (n = 30).
+When the passage of a large vessels was known, the earliest whale detection after passing was 0.15 hours by a sperm whale, while BB's earliest detection was 28 hours.
+Sperm whales also has the lowest median time elapsed after a vessel encounter (1.47 days).
+
+*Environmental Conditions*
+
+Sperm whales are generally detected in deeper subsurface depth conditions than Cuvier's (mixed layer depth and thermocline depth).
 
@@
 
-\[Weekly/daily plot of acoustic presence of whales and ships over time from each drift? See JS Trickey et al. 2022, Fig 3\] Which buoy had the most AIS overlap and when?
+Weekly/daily plot of acoustic presence of whales and ships over time from each drift? See JS Trickey et al. 2022, Fig 3 Which buoy had the most AIS overlap and when?
 
@@
 
-\[Laws and Regulations\]
+Laws and Regulations
 
@@
 
-**\[Known limitations to this study\]\
-**Duty cycle whale detection and soundscape method knowingly omits 18 of 20 minutes of data.
+$$Known limitations to this study$$\
+Duty cycle whale detection and soundscape method knowingly omits 18 of 20 minutes of data.
 A sperm whale's true location relative to buoy is unknown so preferred habitat may not reflect the environmental variables that are associated with buoy that detects it (Barkley et al.
@@
 
-\[Add here.\]
+Add here.
 
@@
 And 3) NOAA is seeking more information on vocally active species' distributions.
-This is part of a bill introduced by Congress to "assess underwater sound in high-priority" environments (https://www.congress.gov/bill/117th-congress/house-bill/6987).
+This is part of a bill introduced by Congress to "assess underwater sound in high-priority" environments (<https://www.congress.gov/bill/117th-congress/house-bill/6987>).
 
@@
 
-**\[Recommendations?\]**
+Recommendations?
 
@@
 
-1\) Species studied:\
+1)  Species studied:
+
 -   ZC Cuvier's beaked whale *Ziphius cavirostris (IUCN status: least concern)\
--* BW43, Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed in 2021 @baumann-pickering2013 @barlow2021\
-- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)* \@simonis\
-- BB, -- Baird's beaked whale *B. bairidii* *(IUCN status: least concern)*\
-- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)* @simonis\
+    -* BW43, Perrin's beaked whale *M*. *perrini (IUCN status: endangered)* confirmed in 2021 @baumann-pickering2013 @barlow2021
+-   BW37V, possibly Hubbs' beaked whale *M. carlhubbsi (IUCN status: data deficient)* @simonis2020a
+-   BB, -- Baird's beaked whale *B. bairidii* *(IUCN status: least concern)*
+-   BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens (IUCN status: data deficient)* @simonis2020a
 -   PM, sperm whale *(IUCN status: vulnerable)*
 
-2\) Environmental covariates (group by physiographic, oceanographic, etc)
+2)  Environmental covariates (group by physiographic, oceanographic, etc)
 
 | Covariate                          | Resolution       | Source                                                                                                                                            | Justification                           |
-|------------------------------------|------------------|---------------------------------------------------------------------------------------------------------------------------------------------------|-----------------------------------------|
+|------------------|------------------|--------------------|------------------|
 |                                    |                  |                                                                                                                                                   |                                         |
@@
 
-\[Can show table of variables and justification by papers (Virgili et al. 2022 and Virgili et al 2019)\]
+Can show table of variables and justification by papers (Virgili et al. 2022 and Virgili et al 2019)
 

### 2023-10-20 07_summary_stats.Rmd
Update 07_summary_stats.Rmd
removed BWC and BW37V

@@
 
-# One detection value for BWC
-BWC <- allDrifts %>%
-  dplyr::filter(species == "BWC")
-print(paste("BWC Max TOL125 sound level: ", max(BWC$TOL_125, na.rm = TRUE)))
-print(paste("BWC Min TOL125 sound level: ", min(BWC$TOL_125, na.rm = TRUE)))
-print(paste("BWC Max TOL20000 sound level: ", max(BWC$TOL_2000, na.rm = TRUE)))
-print(paste("BWC Min TOL20000 sound level: ", min(BWC$TOL_2000, na.rm = TRUE)))
-
 BB <- allDrifts %>%
@@
 
-BW37V <- allDrifts %>%
-  dplyr::filter(species == "BW37V")
-print(paste("BW37V Max TOL125 sound level: ", max(BW37V$TOL_125, na.rm = TRUE)))
-print(paste("BW37V Min TOL125 sound level: ", min(BW37V$TOL_125, na.rm = TRUE)))
-print(paste("BW37V Max TOL20000 sound level: ", max(BW37V$TOL_2000, na.rm = TRUE)))
-print(paste("BW37V Min TOL20000 sound level: ", min(BW37V$TOL_2000, na.rm = TRUE)))
-
 BW43 <- allDrifts %>%
@@
 print(paste("ZC Min TOL20000 sound level: ", min(ZC$TOL_2000, na.rm = TRUE)))
+
 ```
@@
 print(paste("Number of BW43 detections: ", nrow(BW43)))
-print(paste("Number of BWC detections: ", nrow(BWC)))
 print(paste("Number of BB detections: ", nrow(BB)))
-print(paste("Number of BW37V detections: ", nrow(BW37V)))
 

### 2023-10-20 06_gam.Rmd
GAMs for ZC and PM

@@
 ```{r setup, include=FALSE}
-load(allDrifts, file = "data/allDrifts.rda")
+load(file = "data/allDrifts.rda")
 ```
@@
 ### GAMs
+
+# GAM by ZC
+# Top 5 env variables from RF are:depth, sst_mean, curl_mean, mldTemp, bathyslope
+
+```{r}
+ZC_filt <- allDrifts %>% 
+  filter(species %in% c("ZC", NA))
+
+# Environ only
+ZC_GAM <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + mldTemp + bathy_slope,
+              data = ZC_filt, family = binomial, method = "REML")
+summary(ZC_GAM)
+gam.check(ZC_GAM)
+
+ZC_GAM_env_TOL <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + mldTemp + bathy_slope +
+                        TOL_125 + TOL_2000 + shipin10km + timeAfterShip,
+              data = ZC_filt, family = binomial, method = "REML")
+summary(ZC_GAM_env_TOL)
+gam.check(ZC_GAM_env_TOL)
+```
+
+# GAM by PM
+# Top 5 predictors from RF: TOL_2000, dist2slope, timeaftership, ttDepth, depth
+
+```{r}
+PM_filt <- allDrifts %>% 
+  filter(species %in% c("PM", NA))
+
+# Environ only
+PM_GAM <- gam(Wpresence>0 ~ dist2slope + ttDepth + mldDepth + ssh_mean + depth,
+              data = PM_filt, family = binomial, method = "REML")
+summary(PM_GAM)
+gam.check(PM_GAM)
+
+PM_GAM_env_TOL <- gam(Wpresence>0 ~ dist2slope + ttDepth + depth + mldDepth + ssh_mean +
+                        TOL_125 + TOL_2000 + timeAfterShip,
+              data = PM_filt, family = binomial, method = "REML")
+summary(PM_GAM_env_TOL)
+gam.check(PM_GAM_env_TOL)
+```
 
@@
 Higher edfs describe more wiggly curves. Doesn't mean significant
-
-#GAM by ZC
-```{r}
-ZC_filt <- allDrifts %>% 
-  filter(species %in% c("ZC", NA))
-
-ZC_GAM <- gam(Wpresence>0 ~ depth + mldTemp + ttDepth + sal400,
-              data = ZC_filt, family = binomial, method = "REML")
-summary(ZC_GAM)
-gam.check(ZC_GAM)
-```
-
-# GAM by PM
-```{r}
-PM_filt <- allDrifts %>% 
-  filter(species %in% c("PM", NA))
-
-PM_GAM <- gam(Wpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400 +
-                TOL_125 + TOL_2000,
-              data = PM_filt, family = binomial, method = "REML")
-summary(PM_GAM)
-gam.check(PM_GAM)
-
-# Using variables from Fiedler et al. 2023 for sperm whales
-PM_GAM_fiedlerVar <- gam(Wpresence>0 ~ depth + mldDepth + curl_mean + ssh_mean + sal400,
-              data = PM_filt, family = binomial, method = "REML")
-summary(PM_GAM_fiedlerVar)
-```

### 2023-10-20 05_randomForest.R
Create 05_randomForest.R
updated rf models for ZC and PM

@@ -1,0 +1,131 @@
+#Use a Random Forest to identify most important variables
+#in beaked and sperm whale presence
+#Modified from Anne Simonis (7/11/2023)
+
+#Required Packages
+library(randomForest)
+library(dplyr)
+library(rfPermute)
+
+# Load in main dataframe
+load(file = "data/allDrifts.rda")
+
+#Define response variable (presence/absence as 0/1)
+#Create a variable for presence of each species, and then create a
+#random forest model for each species 
+
+#####################################################################
+## Cuvier's BW ####
+#####################################################################
+
+#Descriptive label for analysis
+Dep<-'CCES_Zc'
+
+# For ZC only
+allDrifts$species<-as.factor(allDrifts$species) %>%
+  droplevels(c("BB", "BW43", "PM"))
+
+allDrifts$ZcPresence<-factor(ifelse(allDrifts$species=='ZC',1,0))
+
+#List all variables for evaluation
+covariate.list<-list(c('ZcPresence','curl_mean', 'bathy_slope',
+                       'sst_mean','dist2slope','depth','dist2slope','chlorophyll_mean',
+                       'chlorophyll_mean','mldDepth','ssh_mean',
+                       'mldDepth', 'mldTemp', 'ttDepth','ttTemp', 'temp400', 'sal400'))
+
+include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
+
+#make a text string of all covariates considered for a file name 
+string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
+
+#Create a dataframe for the RF model
+DF.modelZC <- na.omit(allDrifts)
+
+# create balanced sample sizes of response for tree construction to 
+# avoid biases associated with imbalanced data
+# For ZC only
+sampsizeZC <- balancedSampsize(DF.modelZC$ZcPresence)
+
+# set seed for reproducibility
+set.seed(123)
+
+# Use random forest, not rfPermute
+RanFor.model.ZC <- randomForest(ZcPresence ~ ., data=DF.modelZC[,include.covars],
+                         replace=FALSE, ntree=4000, sampsize=sampsizeZC, 
+                         proximity=FALSE, importance = TRUE)
+RanFor.model.ZC
+
+pdf(paste(Dep,"_",string.covars.used,".pdf"), width=14, height=10)
[…]

### 2023-10-20 05_analysis.Rmd
Update 05_analysis.Rmd
new rolling averages plots

@@
 ---
-# Exploratory Analysis and Modeling for all 13 buoys
+# Exploratory Analysis for all 13 buoys
 
@@
   data <- get(col) %>% mutate(Deployment = substr(col,17,18))
+  data <- data[data$species %in% c(NA, "ZC", "PM", "BB", "BW43"),]    # Remove BW37V and BWC from analysis as there are too few detections.
   assign(paste0("GPSwhaleAISenvSS", substr(col,17,18)), data)
@@
                    GPSwhaleAISenvSS23)
+
 save(allDrifts, file = "data/allDrifts.rda")
+load(file = "data/allDrifts.rda")
+```
+
+# Which buoy was the loudest?
+```{r loudest-buoy}
+allDrifts %>%
+  ggplot(mapping = aes(x = Deployment, y = TOL_2000)) + geom_boxplot() + theme_bw()
+
+allDrifts %>%
+  group_by(Deployment) %>%
+  summarize(range(TOL_2000))
+```
+
+# Distribution of noise by drift
+```{r}
+ggplot(allDrifts, aes(x = Deployment, y = TOL_2000)) + geom_boxplot() + 
+  labs(title = "Distribution of TOL2000 by Drift") +
+  theme_classic()
+
+# Where is the loudest sound level?
+allDrifts[allDrifts$TOL_2000 == max(allDrifts$TOL_2000),]
+
+# Where is the quietest sound level?
+allDrifts[allDrifts$TOL_2000 == min(allDrifts$TOL_2000),]
 ```
@@
 ```{r fig.width = 10}
-ggplot(allDrifts, aes(UTC,TOL_2000)) + geom_point() 
+ggplot(allDrifts, aes(UTC,TOL_2000, color = Deployment)) + geom_smooth() 
 ```
@@
 ```{r}
-ZC_filt <- GPSwhaleAISenvSS23 %>% 
+ZC_filt <- GPSwhaleAISenvSS21 %>% 
   filter(species %in% c("ZC", NA)) %>%
+  # Hourly rolling averages
   mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
   mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
-  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA))
+  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA)) %>%
+  # Drop first and last NAs that are created as part of averages
+  drop_na(., rollAvg_Whl)
 
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-  labs(title = "Rolling Averages of ZC Detections alongside Soundscape Metrics - Buoy 23")
+  labs(title = "Rolling Hourly Averages of ZC Detections and Soundscape Metrics -  
+       Buoy 21")
 ```
+# rollapply for vessels
+```{r}
+ZC_filt_ship <- GPSwhaleAISenvSS08 %>% 
+  filter(species %in% c("ZC", NA)) %>%
+  # Hourly rolling averages
+  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_ship = zoo::rollapply(shipin10km, 5, mean, fill = NA)) %>%
+  # Drop first and last NAs that are created as part of averages
+  drop_na(., rollAvg_Whl)
+
+ZC_pivot_ship <- pivot_longer(ZC_filt_ship, cols = rollAvg_Whl:rollAvg_ship, names_to = "avg", values_to = "value")
+
+#Plot facet grid
+ggplot(ZC_pivot_ship, aes(x = UTC, y = value, color = avg)) + geom_smooth() +
+  facet_grid(vars(avg), scales = "free") +
+  scale_x_datetime(date_breaks = "3 days") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+  labs(title = "Rolling Hourly Averages of ZC Detections + SS Metrics + Ships -  
+       Buoy 08")
+```
+
+Vessel presence vs ZC presence
+```{r}
+ZC_filt_ship <- GPSwhaleAISenvSS23 %>% 
+  filter(species %in% c("ZC", NA)) %>%
+  # Hourly rolling averages
+  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_ship = zoo::rollapply(shipin10km, 5, mean, fill = NA)) %>%
+  # Drop first and last NAs that are created as part of averages
+  drop_na(., rollAvg_Whl)
+
+# One plot
+ggplot(ZC_filt_ship, aes(x = UTC)) +
+  geom_smooth(aes(y = rollAvg_ship, color = "ship presence")) +
+  geom_smooth(aes(y = rollAvg_Whl, color = "whale")) +  
+  scale_x_datetime(date_breaks = "3 days") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+  labs(title = "Rolling Hourly Averages of ZC Detections and Ships -  
+       Buoy 23")
+
+```
+
+# rollApply on same plot for ZC
+```{r}
+# Plot lines on same plot
+
+min_first <- min(ZC_filt$rollAvg_Whl)
+max_first <- max(ZC_filt$rollAvg_Whl)
+min_second <- min(ZC_filt$rollAvg_2000)
+max_second <- max(ZC_filt$rollAvg_2000)
+
+# scale and shift variables calculated based on desired mins and maxes
+scale = (max_second - min_second)/(max_first - min_first)
+shift = min_first - min_second
+
+# Function to scale secondary axis
+scale_function <- function(x, scale, shift){
+  return ((x)*scale - shift)
+}
+
+# Function to scale secondary variable values
+inv_scale_function <- function(x, scale, shift){
+  return ((x + shift)/scale)
+}
+
+# Plot two lines on same plot
+ggplot(ZC_filt, aes(x = UTC)) +
+  geom_smooth(aes(y = rollAvg_Whl, color = "ZC presence")) +
+  geom_smooth(aes(y = inv_scale_function(rollAvg_2000, scale, shift), color = "TOL 2000")) +
+  scale_x_datetime(date_breaks = "3 days") +
+  scale_y_continuous(limits = c(min_first, max_first), 
+                     sec.axis = sec_axis(~scale_function(., scale, shift), 
+                                         name = "TOL 2000 - hourly rolling average")) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+  labs(title = "Rolling Averages of ZC Detections alongside Soundscape Metrics - Buoy 13")
+
+########################################
+## Manual scale
+scale = 2000
+ggplot(ZC_filt, aes(x = UTC)) +
+  geom_smooth(aes(y = rollAvg_Whl, color = "ZC presence")) +
+  geom_smooth(aes(y = rollAvg_2000/scale, color = "TOL 2000")) +
+  scale_x_datetime(date_breaks = "3 days") +
+  scale_y_continuous(#limits = c(0.0, 0.05),
+                     sec.axis = sec_axis(~.* scale, name = "TOL 2000 - hourly rolling average")) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+  labs(title = "Rolling Averages of ZC Detections alongside Soundscape Metrics - Buoy 13")
+
+```
+```{r}
+ggplot(ZC_filt, aes(x = UTC)) +
+  geom_smooth(aes(y = rollAvg_Whl+0.25, color = "ZC presence")) #+
+  geom_smooth(aes(y = rollAvg_2000, color = "TOL 2000"))
+```
+
 # Using rollApply for PM
 ```{r}
-PM_filt <- GPSwhaleAISenvSS23 %>% 
+PM_filt <- GPSwhaleAISenvSS14 %>% 
   filter(species %in% c("PM", NA)) %>%
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-  labs(title = "Rolling Averages of PM Detections alongside Soundscape Metrics - Buoy 23")
+  labs(title = "Rolling Averages of PM Detections alongside Soundscape Metrics - Buoy 14")
 ```
@@
 ```{r}
-ggplot(allDrifts, aes(factor(Wpresence), TOL_125)) + geom_boxplot()
+ggplot(allDrifts, aes(factor(Wpresence), TOL_2000)) + geom_boxplot()
 ```
@@
   geom_vline(xintercept = shipOnly3$UTC, color = "green") +  # add line for ship presence
-  geom_point(aes(x = UTC, y = TOL_125, color = species), size = 2) +
+  geom_point(aes(x = UTC, y = TOL_2000, color = species), size = 2) +
   scale_x_datetime(date_breaks = "2 days") +
@@
 ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_125)) +
-  labs(title = "TOL 125 Sound levels by Species - All Buoys")
+  labs(title = "TOL 125 Sound levels by Species - All Buoys") +
+  theme_classic()
 ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_2000)) +
-    labs(title = "TOL 2000 Sound levels by Species - All Buoys")
+    labs(title = "TOL 2000 Sound levels by Species - All Buoys") +
+  theme_classic()
+
+# TOL_2000
+detsOnly3 %>%
+  group_by(species) %>%
+  summarize(median(TOL_2000))
+
+# Depth
+ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = temp400)) +
+  labs(title = "Depth by Species - All Buoys") +
+  theme_classic()
 ```
@@
 ```{r}
+allDrifts %>%
+  group_by(Deployment) %>%
+  summarize(n = sum(shipin10km))
+
 no_na <- allDrifts %>% filter(Wpresence == 1) %>%
-  mutate(HrsafterShip = timeAfterShip /3600)
-ggplot(no_na) + geom_histogram(aes(x = HrsafterShip), bins = 30) + facet_grid(species ~.) +
-  labs(title = "Hours Elapsed from Ship Event by Species - All buoys")
+  mutate(HrsafterShip = timeAfterShip /3600) %>%
+  mutate(DaysAfterShip = HrsafterShip / 24)
+
+# Hours after ship event
+ggplot(no_na) + geom_histogram(aes(x = HrsafterShip), bins = 60) + facet_grid(species ~.) +
+  labs(title = "Hours Elapsed from Ship Event by Species - All buoys") +
+  scale_x_continuous(breaks = seq(0,500, by = 50)) +
+  theme_bw()
+
+# Days after ship event
+ggplot(no_na) + geom_histogram(aes(x = DaysAfterShip), bins = 60) + facet_grid(species ~.) +
+  scale_x_continuous(breaks = seq(0,20, by = 1)) +
+  labs(title = "Days Elapsed from Ship Event by Species - All buoys") +
+  theme_bw()
+
+no_na %>%
+  drop_na(DaysAfterShip) %>%
+  group_by(species) %>%
+  summarize(median(DaysAfterShip))
 ```

### 2023-11-21 RF_ZC_presence.R
Update RF_ZC_presence.R

@@ -10,17 +10,19 @@ library(rfPermute)
 #Define response variable (presence/absence as 0/1)
 #I suggest creating a variable for presence of each species, and then create a
 #random forest model for each species 
+load("data/allDrifts.rda")
+allDrifts$newrow <- sample(10000, size = nrow(allDrifts), replace = TRUE)
 
 # For ZC only
-allDrifts$BWpresence<-as.factor(allDrifts$BWpresence) %>%
-  droplevels(c("BB", "BW43", "BW37V", "PM", "BWC"))
+allDrifts$Wpresence<-as.factor(allDrifts$Wpresence) %>%
+  droplevels(c("BB", "BW43", "PM"))
 
 
 #List all variables you want to evaluate (I only included a few here, 
 #but I know you have more environmental and AIS parameters to add)
-covariate.list<-list(c('TOL_125','TOL_2000','TOL_5000',
+covariate.list<-list(c('TOL_125','TOL_2000',
                        'sst_mean','dist2slope','depth',
-                       'chlorophyll_mean','mldDepth'))
+                       'ssh_mean', 'bathy_slope','newrow'))
 
 include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
 
@@ -31,12 +33,12 @@ string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse
 #DF.modelZC <- na.omit(allDrifts)
 
 DF.modelZC <- allDrifts %>%
-  filter(!species %in% c("BB", "BW43", "BW37V", "PM", "BWC"))
+  filter(!species %in% c("BB", "BW43", "PM"))
 
 # create balanced sample sizes of response for tree construction to 
 # avoid biases associated with imbalanced data
 # For ZC only
-sampsizeZC <- balancedSampsize(DF.modelZC$BWpresence)
+sampsizeZC <- balancedSampsize(DF.modelZC$Wpresence)
 
 # set seed for reproducibility
 set.seed(123)
@@ -44,7 +46,7 @@ set.seed(123)
 # create balanced sample size random forest model
 
 #Model for ZC
-RF.model.ZC <- rfPermute(DF.modelZC$BWpresence ~ ., DF.modelZC[,include.covars], ntree=1000) ##doesnt work!!
+RF.model.ZC <- rfPermute(DF.modelZC$Wpresence ~ ., DF.modelZC[,include.covars], ntree=100) ##doesnt work!!
[…]

### 2023-11-21 RF_example_bwpresence.R
Delete RF_example_bwpresence.R

@@ -1,74 +1,0 @@
-#Use a Random Forest to identify most important variables
-#in beaked and sperm whale presence
-#Anne Simonis 6/23/2023
-
-#Required Packages
-library(randomForest)
-library(dplyr)
-library(rfPermute)
-
-#Define response variable (presence/absence as 0/1)
-#I suggest creating a variable for presence of each species, and then create a
-#random forest model for each species 
-allDrifts$Bwpresence<-as.factor(allDrifts$BWpresence)
-
-# For ZC only
-allDrifts$BWpresence<-as.factor(allDrifts$BWpresence) %>%
-  droplevels(c("BB", "BW43", "BW37V", "PM", "BWC"))
-
-
-#List all variables you want to evaluate (I only included a few here, 
-#but I know you have more environmental and AIS parameters to add)
-covariate.list<-list(c('TOL_125','TOL_2000','TOL_5000',
-                       'sst_mean','dist2slope','depth',
-                       'chlorophyll_mean','mldDepth'))
-
-include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
-
-#make a text string of all covariates considered for a file name 
-string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
-
-#Create a dataframe for the RF model
-DF.model <- na.omit(allDrifts)
-DF.modelZC <- na.omit(allDrifts)
-
-
-# create balanced sample sizes of response for tree construction to 
-# avoid biases associated with imbalanced data
-sampsize <- balancedSampsize(DF.model$Bwpresence)
-sampsize <- balancedSampsize(DF.model$Bwpresence)
-
-# For ZC only
-sampsizeZC <- balancedSampsize(DF.modelZC$BWpresence)
-
-# set seed for reproducibility
-set.seed(123)
-
-# create balanced sample size random forest model
-RF.model <- rfPermute(DF.model$Bwpresence ~ ., DF.model[,include.covars], ntree=1000)
-RF.model
-
-#By ZC, doesn't work
-RF.model.ZC <- rfPermute(DF.modelZC$BWpresence ~ ., DF.modelZC[,include.covars], ntree=1000)
-RF.model.ZC
-
-#Evaluate RF models
-#Create a PDF of summary plots
-pdf(paste(Dep,"_",string.covars.used,".pdf"), width=14, height=10)
-round(importance(RF.model),3) #ranking of importance for each variable
[…]

### 2023-11-21 seafloor_gradi.R
Create seafloor_gradi.R

@@ -1,0 +1,157 @@
+## Adding seafloor gradient variable to drift dataframes
+
+# Required packages
+library(tidyverse)
+library(terra)
+
+# Read in buoy dataframes
+ftlist <- dir("D:/Buoy_dfs/", recursive=TRUE, full.names=TRUE, pattern="GPSwhaleAISenvSS")
+
+for (ft in ftlist) {
+  df <- readRDS(ft)
+  assign(paste0("GPSwhaleAISenvSS", substr(ft, 29, 30)), df)
+}
+
+
+# Download and attach seafloor gradient magnitude to each buoy
+grad07 <- matchEnvData(GPSwhaleAISenvSS07, nc = 'erdEtopoSeafloorGradient', 
+             fileName = "D:/Environ_variables/seafloorgradi07", 
+             var = c('magnitude_gradient')) %>%
+  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean) %>%
+  mutate(Deployment = 07) # add drift number for when I bind
+
+grad08 <- matchEnvData(GPSwhaleAISenvSS08, nc = 'erdEtopoSeafloorGradient', 
+                       fileName = "D:/Environ_variables/seafloorgradi08", 
+                       var = c('magnitude_gradient')) %>%
+  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean) %>%
+  mutate(Deployment = 08)
+
+grad10 <- matchEnvData(GPSwhaleAISenvSS10, nc = 'erdEtopoSeafloorGradient', 
+                       fileName = "D:/Environ_variables/seafloorgradi10", 
+                       var = c('magnitude_gradient')) %>%
+  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean) %>%
+  mutate(Deployment = 10)
+
+grad12 <- matchEnvData(GPSwhaleAISenvSS12, nc = 'erdEtopoSeafloorGradient', 
+                       fileName = "D:/Environ_variables/seafloorgradi12", 
+                       var = c('magnitude_gradient')) %>%
+  dplyr::select(-matchLong_mean, -matchLat_mean, -matchTime_mean) %>%
+  mutate(Deployment = 12)
+
+grad13 <- matchEnvData(GPSwhaleAISenvSS13, nc = 'erdEtopoSeafloorGradient', 
+                       fileName = "D:/Environ_variables/seafloorgradi13", 
[…]

### 2023-11-21 maps.Rmd
Update maps.Rmd
added AIS density map

@@
 
-Set Parameters
+Set Parameters and requried packages
 ```{r}
+library(tidyverse)
+library(sf)
+library(maptiles)
+library(tmap)
+library(tmaptools)
+library(crs)
+library(terra)
+
 theme_set(theme_bw())
 tmap_mode("plot")
+load("data/allDrifts.rda")
 ```
 
-Create sf objects and basemaps for maps (not needed for environmental variables)
+Establish bounding box
+Create map with inset box
 ```{r}
-whales_8_10sf <- Wsf %>%                            
-dplyr::filter(station == "8" | station == "10")
+# Make allDrifts an sf object
+allDrifts_sf <- allDrifts %>%
+  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
 
-trackssf <- st_as_sf(tracks, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_08 <- st_as_sf(tracks_SS_08, coords = c("Longitude", "Latitude"), crs=4326)
-trk_TOL_10 <- st_as_sf(tracks_SS_10, coords = c("Longitude", "Latitude"), crs=4326)
-trackBase <- get_tiles(whales_8_10sf, provider = "Esri.WorldShadedRelief")
-allTrackbase <- get_tiles(trackssf, provider = "Stamen.TonerLite", zoom = 5)
-```
+# Make detections only object
+detsOnly_sf <- allDrifts %>% filter(Wpresence == 1) %>%
+  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
+  st_make_valid()
 
-Map of whole study area
-```{r}
-# Establish bounding box
-bbox <- st_bbox(c(xmin = -110, 
-            xmax = -128, 
-            ymax = 44, 
-            ymin = 30), crs = st_crs(4326))
+# Establish bounding box of study area
+xmin <- st_bbox(allDrifts_sf)[1]
+xmax <- st_bbox(allDrifts_sf)[3]
+ymin <- st_bbox(allDrifts_sf)[2]
+ymax <- st_bbox(allDrifts_sf)[4]
 
-#Establish basemap
-allTrackbase <- get_tiles(bbox, provider = "Esri.WorldGrayCanvas", zoom = 5)
+# Add 0.5 border to the extent of drifts
+bnd <- st_bbox(c(xmin - 0.5, xmax + 0.5, 
+           ymin - 0.5, ymax + 0.5))
 
-# Establish inset map
+#Establish basemap with GreyCanvas basemap
 us_region <- st_bbox(c(xmin = -90,
@@
 
-sg <- bb_poly(bb(tracks_sf))
-inset <- tm_shape(usa) + tm_rgb() + tm_shape(sg) + tm_borders(col = "red")
+# Inset box
+sg <- bb_poly(st_bbox(allDrifts_sf))
+```
 
+Plot inset map
+```{r}
+# try tm_tiles
+
+inset <- tm_shape(usa) + tm_tiles() + tm_shape(sg) + tm_borders("red")
+inset
+
+# need to fix
+tm_basemap(leaflet::providers$Esri.WorldGrayCanvas) + tm_shape(sg) + tm_borders("red")
+
+```
+
+All Drifts Map
+```{r}
+# Establish basemap with OceanBaseMap
+allDrifts_base <- get_tiles(allDrifts_sf, provider = "Esri.OceanBasemap", zoom = 6)
+
+# Read in US state and MX boundaries
+# States boundary
+bx <- st_bbox(allDrifts_sf)
+
+states <- st_read("data/us_states_shp") %>%
+  st_transform(crs = 4326) %>%
+  st_crop(bnd) # crop to allDrifts extent + 0.5
+
+mex <- st_read("data/mexico_shp") %>%
+  st_transform(crs = 4326) %>%
+  st_crop(bnd) # crop to allDrifts extent + 0.5
+
+# US + Mex borders combined
+west <- st_union(states,mex)
+
+# Plot map - ERROR??
+p <- tm_shape(allDrifts_base) + tm_rgb() + 
+  tm_shape(states) + tm_borders() +
+  tm_shape(mex) + tm_borders() +
+  tm_shape(allDrifts_sf) + tm_dots()
+p
+```
+
+All Drifts with Ship detections
+```{r}
+# With Ship detections
+shipOnly <- allDrifts_sf %>% filter(shipin10km == 1)
+
+# Plot both
+p + tm_shape(shipOnly) + tm_dots(col = "yellow")
+```
+
+Read in AIS objects
+```{r}
+# Read in AIS_allpts
+AIS_allpts <- read_csv(file = "D:/AIS/AIS_West/AIS_allpts.csv",
+                       show_col_types = FALSE)
+
+# Make sf object
+AIS_allpts_sf <- AIS_allpts %>%
+  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
+```
+
+AIS Map with nearshore AIS points removed
+```{r}
+# Create buffer and simplify lines
+us_buff <- st_simplify(st_buffer(states, dist = 20000), dTolerance = 2000)
+mx_buff <- st_simplify(st_buffer(mex, dist = 20000), dTolerance = 2000)
+west_buff <- st_make_valid(st_union(us_buff, mx_buff))
+
+# View west coast with buffer
+tm_shape(west_buff) + tm_borders()
+
+# Only offshore AIS points
+water <- lengths(st_intersects(AIS3, west_buff)) == 0
+keep <- st_make_valid(AIS3[water,]) %>%
+  st_set_crs(4326)
+
+# View with nearshore AIS points removed
+tm_shape(keep) + tm_dots(col = "green") + 
+  tm_shape(west_buff) + tm_borders(col = "red")+
+  tm_shape(mex) + tm_borders() +
+  tm_shape(states) + tm_borders() +
+  tm_shape(allDrifts_sf) + tm_dots(size = 0.01) +
+  tmap_options(check.and.fix = TRUE)
+
+```
+
+OLD code with single day of AIS pts
+```{r}
+# Make a fishnet grid over AIS points
+grid <- st_make_grid(bnd, n = 30) %>%
+  st_set_crs(4326)
+plot(grid)
+
+# find which grid points intersect AIS points 
+# and create an index to subset from
+index <- which(lengths(st_intersects(grid, keep)) > 0)
+
+# Subset the grid to make a fishnet
+fishnet <- grid[index]
+#plot(fishnet)
+
+map <- fishnet |>  
+  st_as_sf() |> # cast to sf
+  mutate(grid_id = row_number()) |> # create unique ID
+  st_join(AIS3) |> # join the AIS dataset
+  group_by(grid_id) |> # group by the grid id 
+  summarize(n = n_distinct(MMSI)) # summarize by unique ship ID
+
+# Right now, n is total number of unique ships in each grid cell.
+# With more points, may want to change to avg daily unique ships.
+
+#ggplot
+ggplot() + 
+  geom_sf(data = map, aes(fill = n), lwd = 0.05, color = "white") +
+  geom_sf(data = states) +
+  geom_sf(data = mex) +
+  geom_sf(data = detsOnly_sf, col = 'green', size = 0.01) +
+  geom_sf(data = west_buff, col = "red", alpha = 0.05) +
+  theme_minimal()
+
+# tmap
+tm_shape(map) + tm_fill(col = 'n') +
+  #tm_shape(west_buff) + tm_borders(col = "red") +
+  tm_shape(mex) + tm_borders() +
+  tm_shape(states) + tm_borders() #+
+  #tm_shape(allDrifts_sf) + tm_dots(size = 0.01)
+
+```
+
+Points to line loop
+```{r}
+# Make a fishnet grid over AIS area
+# What size cells are these?
+grid <- st_make_grid(bnd, cellsize = 0.4, square = FALSE) %>%
+  st_set_crs(4326)
+
+shipList <- dir("D:/AIS/AIS_West", recursive=TRUE, full.names=TRUE, pattern=".rda")
+
+for (i in shipList) {
+  p <- readRDS(i) %>%
+  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
+  group_by(MMSI) %>%
+  summarize() %>%
+  filter(st_geometry_type(.) == "MULTIPOINT") %>%
+  st_cast("LINESTRING")
+
+  z <- grid %>%
+  st_as_sf() |>
+  mutate(grid_id = row_number()) |>
+  st_intersection(p) |> # join AIS line df
+  group_by(MMSI) |> # group by AIS id 
+  count()
+
+# Grid
+  m <- grid %>%
+  st_as_sf() %>%
+  mutate(grid_id = row_number()) %>%
+  st_join(z) %>%
+  group_by(grid_id) %>% # group by AIS id
+  summarize(n = n_distinct(MMSI))
+
+  assign(paste0("AIS_", substr(i,21,25), "_grid"), m)
+
+  print(paste0("Completed", substr(i,21,25)))
+}
+```
+
+Sum up all ships on all days per grid
+```{r}
+f <- bind_rows(AIS_08_06_grid,AIS_08_07_grid, AIS_08_08_grid, AIS_08_09_grid, 
+               AIS_08_10_grid, AIS_08_11_grid, AIS_08_12_grid, AIS_08_13_grid, 
+               AIS_08_14_grid, AIS_08_15_grid, AIS_08_16_grid, AIS_08_17_grid, 
+               AIS_08_18_grid, AIS_08_19_grid, AIS_08_20_grid, AIS_08_21_grid, 
+               AIS_08_22_grid, AIS_08_23_grid, AIS_08_24_grid, AIS_08_25_grid, 
+               AIS_08_26_grid, AIS_08_27_grid, AIS_08_28_grid, AIS_08_29_grid, 
+               AIS_08_30_grid, AIS_08_31_grid, AIS_09_01_grid, AIS_09_02_grid,
+               AIS_09_03_grid, AIS_09_04_grid, AIS_09_05_grid, AIS_09_06_grid, 
+               AIS_09_07_grid, AIS_09_08_grid, AIS_09_09_grid, AIS_09_10_grid, 
+               AIS_09_11_grid, AIS_09_12_grid, AIS_09_13_grid, AIS_09_14_grid, 
+               AIS_09_15_grid, AIS_09_16_grid, AIS_09_17_grid, AIS_09_18_grid, 
+               AIS_09_19_grid, AIS_09_20_grid, AIS_09_21_grid, AIS_09_22_grid, 
+               AIS_09_23_grid, AIS_09_24_grid, AIS_09_25_grid, AIS_09_26_grid, 
+               AIS_09_27_grid, AIS_09_28_grid, AIS_09_29_grid, AIS_09_30_grid, 
+               AIS_10_01_grid, AIS_10_02_grid, AIS_10_03_grid, AIS_10_04_grid, 
+               AIS_10_05_grid, AIS_10_06_grid, AIS_10_07_grid, AIS_10_08_grid, 
+               AIS_10_09_grid, AIS_10_10_grid, AIS_10_11_grid, AIS_10_12_grid, 
+               AIS_10_13_grid, AIS_10_14_grid, AIS_10_15_grid, AIS_10_16_grid, 
+               AIS_10_17_grid, AIS_10_18_grid, AIS_10_19_grid, AIS_10_20_grid, 
+               AIS_10_21_grid, AIS_10_22_grid, AIS_10_23_grid, AIS_10_24_grid, 
+               AIS_10_25_grid, AIS_10_26_grid, AIS_10_27_grid, AIS_10_28_grid,
+               AIS_10_29_grid, AIS_10_30_grid, AIS_10_31_grid, AIS_11_01_grid, 
+               AIS_11_02_grid, AIS_11_03_grid, AIS_11_04_grid, AIS_11_05_grid, 
+               AIS_11_06_grid, AIS_11_07_grid, AIS_11_08_grid, AIS_11_09_grid, 
+               AIS_11_10_grid, AIS_11_11_grid, AIS_11_12_grid, AIS_11_13_grid, 
+               AIS_11_14_grid, AIS_11_15_grid, AIS_11_16_grid, AIS_11_17_grid, 
+               AIS_11_18_grid, AIS_11_19_grid, AIS_11_20_grid, AIS_11_21_grid, 
+               AIS_11_22_grid, AIS_11_23_grid, AIS_11_24_grid, AIS_11_25_grid,
+               AIS_11_26_grid, AIS_11_27_grid, AIS_11_28_grid, AIS_11_29_grid, 
+               AIS_11_30_grid, AIS_12_01_grid, AIS_12_02_grid) %>%
+  group_by(grid_id) %>%
+  summarize(total = sum(n))
+
+saveRDS(f, file = "D:/AIS/AIS_West/AIS_grid.rda")
+```
+
+Cut out AIS on land and coastline
+```{r}
+# Read in AIS points
+f <- readRDS(file = "D:/AIS/AIS_West/AIS_grid.rda")
+
+# Only offshore AIS points
+f_pts_intersect <- lengths(st_intersects(f, west)) == 0
+AISgrd <- st_make_valid(f[f_pts_intersect,]) %>%
+  st_set_crs(4326)
+
+#What is area of these grid cells?
+st_area(grid)
+st_area(AISgrd)
+```
+Final AIS maps
+```{r fig.width=10}
+# ggplot
+ggplot(AISgrd) + 
+  geom_sf(aes(fill = total)) +
+  scale_fill_viridis_c() +
+  geom_sf(data = west) +
+  geom_sf(data = allDrifts_sf)
+
+# tmap
+#tm_basemap(leaflet::providers$Esri.WorldTopoMap) +
+  tm_shape(AISgrd) + tm_fill(fill = 'total', 
+                             col = 'lightgrey', 
+                             fill.legend = tm_legend("Ship Counts"),
+                             fill.scale = tm_scale_continuous()) + 
+  tm_shape(west) + tm_borders(col = "black", 
+                              fill = "papayawhip")+
+  tm_shape(detsOnly_sf) + tm_dots(size = 0.2, 
+                                  legend.show = TRUE) +
+  tm_layout(legend.outside = TRUE) +
+  tm_grid(alpha = 0.2, 
+          col = "grey") +
+  tm_scalebar() +
+  tm_compass(type = "arrow", 
+             position = c(0.05, 0.05), 
+             color.light = "blue") +
+  tm_add_legend('symbol',
+                fill = "black",
+                labels = "Whale Detections",
+                size = 0.5,
+                lwd = 2) +
+  tm_place_legends_right(width = 0.2) +
+  tm_title("Daily Counts of Unique AIS Vessels")
+
+```
+
+```{r}
 #Map all tracks
@@
 ```{r}
-# CCES_08 + CCES_10 Broadband
-tm_shape(trackBase) + tm_rgb() + 
-  tm_shape(trk_TOL_08) + 
-  tm_symbols(col = "BB_20-24000", 
-             border.lwd = NA, 
-             size = 0.4, 
-             sizes.legend = 0.7, 
-             n = 10) +
-  tm_shape(trk_TOL_10) + 
-  tm_symbols(col = "BB_20-24000",
-             border.lwd = NA, 
-             size = 0.4,
-             n = 10) +
-  tm_layout(legend.width = 0.5, 
-            legend.position = c("right","top"),
-            legend.bg.color = "lightblue")
 

### 2023-11-21 interpolation.Rmd
Update interpolation.Rmd

@@
 ---
+
+This document shows the process for determining whether interpolation for drift 20 was appropriate
+during the 4 days the GPS went dark. I compare environmental variables of interpolated points to
+those at the start and end of the gap. Since the interpolated environmental variables are significantly
+different, it's advised to eliminate them from environmental analysis.
 

### 2023-11-21 guest_scripts/RF_example_bwpresence.R
Create RF_example_bwpresence.R

@@ -1,0 +1,74 @@
+#Use a Random Forest to identify most important variables
+#in beaked and sperm whale presence
+#Anne Simonis 6/23/2023
+
+#Required Packages
+library(randomForest)
+library(dplyr)
+library(rfPermute)
+
+#Define response variable (presence/absence as 0/1)
+#I suggest creating a variable for presence of each species, and then create a
+#random forest model for each species 
+allDrifts$Bwpresence<-as.factor(allDrifts$BWpresence)
+
+# For ZC only
+allDrifts$BWpresence<-as.factor(allDrifts$BWpresence) %>%
+  droplevels(c("BB", "BW43", "BW37V", "PM", "BWC"))
+
+
+#List all variables you want to evaluate (I only included a few here, 
+#but I know you have more environmental and AIS parameters to add)
+covariate.list<-list(c('TOL_125','TOL_2000','TOL_5000',
+                       'sst_mean','dist2slope','depth',
+                       'chlorophyll_mean','mldDepth'))
+
+include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
+
+#make a text string of all covariates considered for a file name 
+string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
+
+#Create a dataframe for the RF model
+DF.model <- na.omit(allDrifts)
+DF.modelZC <- na.omit(allDrifts)
+
+
+# create balanced sample sizes of response for tree construction to 
+# avoid biases associated with imbalanced data
+sampsize <- balancedSampsize(DF.model$Bwpresence)
+sampsize <- balancedSampsize(DF.model$Bwpresence)
+
+# For ZC only
+sampsizeZC <- balancedSampsize(DF.modelZC$BWpresence)
+
+# set seed for reproducibility
+set.seed(123)
+
+# create balanced sample size random forest model
+RF.model <- rfPermute(DF.model$Bwpresence ~ ., DF.model[,include.covars], ntree=1000)
+RF.model
+
+#By ZC, doesn't work
+RF.model.ZC <- rfPermute(DF.modelZC$BWpresence ~ ., DF.modelZC[,include.covars], ntree=1000)
+RF.model.ZC
+
+#Evaluate RF models
+#Create a PDF of summary plots
+pdf(paste(Dep,"_",string.covars.used,".pdf"), width=14, height=10)
+round(importance(RF.model),3) #ranking of importance for each variable
[…]

### 2023-11-21 guest_scripts/PSD_Plots_by_Species.R
new folder

@@ -1,0 +1,39 @@
+
+library(PAMscapes)
+
+#Load in median power spectral density (PSD) files
+psd13 <- checkSoundscapeInput('H:/Soundscape/metrics/CCES_013/CCES_013_PSD_2min.csv')
+psd14 <- checkSoundscapeInput('H:/Soundscape/metrics/CCES_014/CCES_014_PSD_2min.csv')
+
+#To-Do
+#Marina - you'll need to define when Zc is present/absent on each drift 
+#Create separate PSD objects for the times when bw were present and absent 
+ZcTimes13<- #times when Cuvier's beaked whales were present on drift 13
+ZcTimes14<- #times when Cuvier's beaked whales were present on drift 14
+  
+psd13ZcPres<-filter(psd13,UTC %in% ZcTimes13) #Zc Present
+psd13ZcAbs<-filter(psd13,!UTC %in% ZcTimes13) #Zc Absent
+
+psd14ZcPres<-filter(psd14,UTC %in% ZcTimes14) #Zc Present
+psd14ZcAbs<-filter(psd14,!UTC %in% ZcTimes14) #Zc Absent
+
+#Combine PSD files from multiple drifts with rbind
+psdZcPres<-rbind(psd13ZcPres,psd14ZcPres)
+psdZcAbs<-rbind(psd13ZcAbs,psd14ZcAbs)
+
+
+# Plotting 100-10,000 Hz (these are ggplot objects)
+#Individual Drifts
+plotPSD(psd13ZcPres[1:10000], style='density')
+plotPSD(psd13ZcAbs[1:10000], style='density')
+
+plotPSD(psd13ZcPres[1:10000], style='quantile', q=.05)
+plotPSD(psd13ZcAbs[1:10000], style='quantile', q=.05)
+
+#Combined Drifts 
+plotPSD(psdZcPres[1:10000], style='density')
+plotPSD(psdZcAbs[1:10000], style='density')
+
+plotPSD(psdZcPres[1:10000], style='quantile', q=.05)
+plotPSD(psdZcAbs[1:10000], style='quantile', q=.05)
+

### 2023-11-21 02_AIS.Rmd
Update 02_AIS.Rmd

@@
 library(tidyverse)
+library(sf)
 ```
@@
 ```
+
+# For density plots in R
+```{r}
+#Load allDrifts object
+load("data/allDrifts.rda")
+
+# Make allDrifts an sf object
+allDrifts_sf <- allDrifts %>%
+  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
+bnd <- st_bbox(allDrifts_sf)
+
+# Make a fishnet grid over AIS points
+grid <- st_make_grid(bnd, n = 30) %>%
+  st_set_crs(4326)
+
+# find which grid points intersect AIS points 
+# and create an index to subset from
+index <- which(lengths(st_intersects(grid, keep)) > 0)
+
+# Subset the grid to make a fishnet
+fishnet <- grid[index]
+```
+
+TAKES WAY TOO LONG
+```{r}
+# One way to read in files and bind
+AISfiles <- list.files("D:/AIS/AIS_West", pattern = "*.csv", full.names = TRUE)
+AIS_dfs <- lapply(AISfiles, read.csv) # creates list
+
+# Row bind
+AIS_allpts <- bind_rows(AIS_dfs)
+
+# Filter broken up into many parts for easy completion! Not easy...
+AIS_allpts_vLength <- AIS_allpts %>%
+  filter(vesselLength >= 100)   #remove vessels under 100m in length
+
+AIS_allpts_SOG <- AIS_allpts_vLength %>%
+  filter(SOG > 0) # Filter out ships that are stopped
+
+## Alternative way to crop (using min/max latitude of bounding box)
+AIS_allpts_lat <- AIS_allpts_SOG %>%
+  filter(Latitude >= min(allDrifts$Latitude))
+
+#filter(Latitude <= max(allDrifts$Latitude)
+
+AIS_allpts_sf <- AIS_allpts_SOG %>%
+  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
+
+AIS_allpts_crop <- AIS_allpts_sf %>%
+  st_crop(bnd)
+
+# Write csv
+write_csv(AIS_allpts_crop, file = "D:/AIS/AIS_allpts.csv")
+```
+
+# ALSO TAKES WAY TOO LONG
+```{r}
+# Read in AIS from one day, convert to sf, crop to allDrifts area
+AISlist <- dir("D:/AIS/AIS_West", recursive=TRUE, full.names=TRUE)
+
+# Loop through with filters
+for (day in AISlist) {
+  dt <- read_csv(day, show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%  #remove vessels under 100m in length
+  filter(SOG > 0) %>% # Filter out ships that are stopped
+  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
+  st_crop(bnd)
+
+  assign(paste0("AIS_", substr(day, 31, 35)), dt)
+}
+```
+
+# one at a time works - ugh.
+Filter by vessels over 100m, SOG over 2.5 knots, and increase bounding box by 0.5 on all sides
+8/6 - 8/20
+```{r}
+# Read in one csv at a time.
+AIS_08_06 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_06.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_06, file = "D:/AIS/AIS_West/AIS_08_06.rda")
+#
+AIS_08_07 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_07.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+saveRDS(AIS_08_07, file = "D:/AIS/AIS_West/AIS_08_07.rda")
+#
+AIS_08_08 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_08.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+saveRDS(AIS_08_08, file = "D:/AIS/AIS_West/AIS_08_08.rda")
+#  
+AIS_08_09 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_09.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+saveRDS(AIS_08_09, file = "D:/AIS/AIS_West/AIS_08_09.rda")
+
+AIS_08_10 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_10.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+saveRDS(AIS_08_10, file = "D:/AIS/AIS_West/AIS_08_10.rda")
+
+AIS_08_11 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_11.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_11, file = "D:/AIS/AIS_West/AIS_08_11.rda")
+
+AIS_08_12 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_12.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_12, file = "D:/AIS/AIS_West/AIS_08_12.rda")
+
+AIS_08_13 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_13.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+saveRDS(AIS_08_13, file = "D:/AIS/AIS_West/AIS_08_13.rda")
+
+AIS_08_14 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_14.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_14, file = "D:/AIS/AIS_West/AIS_08_14.rda")
+
+AIS_08_15 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_15.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_15, file = "D:/AIS/AIS_West/AIS_08_15.rda")
+
+AIS_08_16 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_16.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_16, file = "D:/AIS/AIS_West/AIS_08_16.rda")
+
+AIS_08_17 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_17.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_17, file = "D:/AIS/AIS_West/AIS_08_17.rda")
+
+AIS_08_18 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_18.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_18, file = "D:/AIS/AIS_West/AIS_08_18.rda")
+
+AIS_08_19 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_19.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_19, file = "D:/AIS/AIS_West/AIS_08_19.rda")
+
+AIS_08_20 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_20.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_20, file = "D:/AIS/AIS_West/AIS_08_20.rda")
+
+```
+
+8/21 - 8/31
+```{r}
+AIS_08_21 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_21.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_21, file = "D:/AIS/AIS_West/AIS_08_21.rda")
+
+AIS_08_22 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_22.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_22, file = "D:/AIS/AIS_West/AIS_08_22.rda")
+
+AIS_08_23 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_23.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_23, file = "D:/AIS/AIS_West/AIS_08_23.rda")
+
+AIS_08_24 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_24.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_24, file = "D:/AIS/AIS_West/AIS_08_24.rda")
+
+AIS_08_25 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_25.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_25, file = "D:/AIS/AIS_West/AIS_08_25.rda")
+
+AIS_08_26 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_26.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_26, file = "D:/AIS/AIS_West/AIS_08_26.rda")
+
+AIS_08_27 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_27.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_27, file = "D:/AIS/AIS_West/AIS_08_27.rda")
+
+AIS_08_28 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_28.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_28, file = "D:/AIS/AIS_West/AIS_08_28.rda")
+
+AIS_08_29 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_29.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_29, file = "D:/AIS/AIS_West/AIS_08_29.rda")
+
+AIS_08_30 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_30.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_30, file = "D:/AIS/AIS_West/AIS_08_30.rda")
+
+AIS_08_31 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_08_31.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_08_31, file = "D:/AIS/AIS_West/AIS_08_31.rda")
+```
+
+9/1 - 9/15
+```{r}
+AIS_09_01 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_01.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_01, file = "D:/AIS/AIS_West/AIS_09_01.rda")
+
+AIS_09_02 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_02.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_02, file = "D:/AIS/AIS_West/AIS_09_02.rda")
+
+AIS_09_03 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_03.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_03, file = "D:/AIS/AIS_West/AIS_09_03.rda")
+
+AIS_09_04 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_04.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_04, file = "D:/AIS/AIS_West/AIS_09_04.rda")
+
+AIS_09_05 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_05.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_05, file = "D:/AIS/AIS_West/AIS_09_05.rda")
+
+AIS_09_06 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_06.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_06, file = "D:/AIS/AIS_West/AIS_09_06.rda")
+
+AIS_09_07 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_07.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_07, file = "D:/AIS/AIS_West/AIS_09_07.rda")
+
+AIS_09_08 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_08.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_08, file = "D:/AIS/AIS_West/AIS_09_08.rda")
+
+AIS_09_09 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_09.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_09, file = "D:/AIS/AIS_West/AIS_09_09.rda")
+
+AIS_09_10 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_10.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_10, file = "D:/AIS/AIS_West/AIS_09_10.rda")
+
+AIS_09_11 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_11.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_11, file = "D:/AIS/AIS_West/AIS_09_11.rda")
+
+AIS_09_12 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_12.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_12, file = "D:/AIS/AIS_West/AIS_09_12.rda")
+
+AIS_09_13 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_13.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_13, file = "D:/AIS/AIS_West/AIS_09_13.rda")
+
+AIS_09_14 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_14.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_14, file = "D:/AIS/AIS_West/AIS_09_14.rda")
+
+AIS_09_15 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_15.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_15, file = "D:/AIS/AIS_West/AIS_09_15.rda")
+
+```
+
+9/16 - 9/30
+```{r}
+AIS_09_16 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_16.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_16, file = "D:/AIS/AIS_West/AIS_09_16.rda")
+
+AIS_09_17 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_17.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_17, file = "D:/AIS/AIS_West/AIS_09_17.rda")
+
+AIS_09_18 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_18.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_18, file = "D:/AIS/AIS_West/AIS_09_18.rda")
+
+AIS_09_19 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_19.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_19, file = "D:/AIS/AIS_West/AIS_09_19.rda")
+
+AIS_09_20 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_20.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_20, file = "D:/AIS/AIS_West/AIS_09_20.rda")
+
+AIS_09_21 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_21.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_21, file = "D:/AIS/AIS_West/AIS_09_21.rda")
+
+AIS_09_22 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_22.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_22, file = "D:/AIS/AIS_West/AIS_09_22.rda")
+
+AIS_09_23 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_23.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_23, file = "D:/AIS/AIS_West/AIS_09_23.rda")
+
+AIS_09_24 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_24.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_24, file = "D:/AIS/AIS_West/AIS_09_24.rda")
+
+AIS_09_25 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_25.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_25, file = "D:/AIS/AIS_West/AIS_09_25.rda")
+
+AIS_09_26 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_26.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_26, file = "D:/AIS/AIS_West/AIS_09_26.rda")
+
+AIS_09_27 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_27.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_27, file = "D:/AIS/AIS_West/AIS_09_27.rda")
+
+AIS_09_28 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_28.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_28, file = "D:/AIS/AIS_West/AIS_09_28.rda")
+
+AIS_09_29 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_29.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_29, file = "D:/AIS/AIS_West/AIS_09_29.rda")
+
+AIS_09_30 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_09_30.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_09_30, file = "D:/AIS/AIS_West/AIS_09_30.rda")
+```
+
+10/1 - 10/15
+```{r}
+AIS_10_01 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_01.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_01, file = "D:/AIS/AIS_West/AIS_10_01.rda")
+
+AIS_10_02 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_02.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_02, file = "D:/AIS/AIS_West/AIS_10_02.rda")
+
+AIS_10_03 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_03.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_03, file = "D:/AIS/AIS_West/AIS_10_03.rda")
+
+AIS_10_04 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_04.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_04, file = "D:/AIS/AIS_West/AIS_10_04.rda")
+
+AIS_10_05 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_05.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_05, file = "D:/AIS/AIS_West/AIS_10_05.rda")
+
+AIS_10_06 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_06.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_06, file = "D:/AIS/AIS_West/AIS_10_06.rda")
+
+AIS_10_07 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_07.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_07, file = "D:/AIS/AIS_West/AIS_10_07.rda")
+
+AIS_10_08 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_08.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_08, file = "D:/AIS/AIS_West/AIS_10_08.rda")
+
+AIS_10_09 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_09.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_09, file = "D:/AIS/AIS_West/AIS_10_09.rda")
+
+AIS_10_10 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_10.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_10, file = "D:/AIS/AIS_West/AIS_10_10.rda")
+
+AIS_10_11 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_11.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_11, file = "D:/AIS/AIS_West/AIS_10_11.rda")
+
+AIS_10_12 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_12.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_12, file = "D:/AIS/AIS_West/AIS_10_12.rda")
+
+AIS_10_13 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_13.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_13, file = "D:/AIS/AIS_West/AIS_10_13.rda")
+
+AIS_10_14 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_14.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_14, file = "D:/AIS/AIS_West/AIS_10_14.rda")
+
+AIS_10_15 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_15.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_15, file = "D:/AIS/AIS_West/AIS_10_15.rda")
+```
+
+10/16 - 10/31
+```{r}
+AIS_10_16 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_16.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_16, file = "D:/AIS/AIS_West/AIS_10_16.rda")
+
+AIS_10_17 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_17.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_17, file = "D:/AIS/AIS_West/AIS_10_17.rda")
+
+AIS_10_18 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_18.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_18, file = "D:/AIS/AIS_West/AIS_10_18.rda")
+
+AIS_10_19 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_19.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_19, file = "D:/AIS/AIS_West/AIS_10_19.rda")
+
+AIS_10_20 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_20.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_20, file = "D:/AIS/AIS_West/AIS_10_20.rda")
+
+AIS_10_21 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_21.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_21, file = "D:/AIS/AIS_West/AIS_10_21.rda")
+
+AIS_10_22 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_22.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_22, file = "D:/AIS/AIS_West/AIS_10_22.rda")
+
+AIS_10_23 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_23.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_23, file = "D:/AIS/AIS_West/AIS_10_23.rda")
+
+AIS_10_24 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_24.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_24, file = "D:/AIS/AIS_West/AIS_10_24.rda")
+
+AIS_10_25 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_25.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_25, file = "D:/AIS/AIS_West/AIS_10_25.rda")
+
+AIS_10_26 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_26.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_26, file = "D:/AIS/AIS_West/AIS_10_26.rda")
+
+AIS_10_27 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_27.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_27, file = "D:/AIS/AIS_West/AIS_10_27.rda")
+
+AIS_10_28 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_28.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_28, file = "D:/AIS/AIS_West/AIS_10_28.rda")
+
+AIS_10_29 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_29.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_29, file = "D:/AIS/AIS_West/AIS_10_29.rda")
+
+AIS_10_30 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_30.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_30, file = "D:/AIS/AIS_West/AIS_10_30.rda")
+
+AIS_10_31 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_10_31.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_10_31, file = "D:/AIS/AIS_West/AIS_10_31.rda")
+```
+
+11/1 - 11/15
+```{r}
+AIS_11_01 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_01.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_01, file = "D:/AIS/AIS_West/AIS_11_01.rda")
+
+AIS_11_02 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_02.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_02, file = "D:/AIS/AIS_West/AIS_11_02.rda")
+
+AIS_11_03 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_03.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_03, file = "D:/AIS/AIS_West/AIS_11_03.rda")
+
+AIS_11_04 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_04.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_04, file = "D:/AIS/AIS_West/AIS_11_04.rda")
+
+AIS_11_05 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_05.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_05, file = "D:/AIS/AIS_West/AIS_11_05.rda")
+
+AIS_11_06 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_06.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_06, file = "D:/AIS/AIS_West/AIS_11_06.rda")
+
+AIS_11_07 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_07.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_07, file = "D:/AIS/AIS_West/AIS_11_07.rda")
+
+AIS_11_08 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_08.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+saveRDS(AIS_11_08, file = "D:/AIS/AIS_West/AIS_11_08.rda")
+
+AIS_11_09 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_09.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_09, file = "D:/AIS/AIS_West/AIS_11_09.rda")
+
+AIS_11_10 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_10.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_10, file = "D:/AIS/AIS_West/AIS_11_10.rda")
+
+AIS_11_11 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_11.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_11, file = "D:/AIS/AIS_West/AIS_11_11.rda")
+
+AIS_11_12 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_12.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_12, file = "D:/AIS/AIS_West/AIS_11_12.rda")
+
+AIS_11_13 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_13.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_13, file = "D:/AIS/AIS_West/AIS_11_13.rda")
+
+AIS_11_14 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_14.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_14, file = "D:/AIS/AIS_West/AIS_11_14.rda")
+
+AIS_11_15 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_15.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_15, file = "D:/AIS/AIS_West/AIS_11_15.rda")
+```
+
+11/16 - 12/02
+```{r}
+AIS_11_16 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_16.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_16, file = "D:/AIS/AIS_West/AIS_11_16.rda")
+
+AIS_11_17 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_17.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_17, file = "D:/AIS/AIS_West/AIS_11_17.rda")
+
+AIS_11_18 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_18.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_18, file = "D:/AIS/AIS_West/AIS_11_18.rda")
+
+AIS_11_19 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_19.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_19, file = "D:/AIS/AIS_West/AIS_11_19.rda")
+
+AIS_11_20 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_20.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_20, file = "D:/AIS/AIS_West/AIS_11_20.rda")
+
+AIS_11_21 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_21.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_21, file = "D:/AIS/AIS_West/AIS_11_21.rda")
+
+AIS_11_22 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_22.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_22, file = "D:/AIS/AIS_West/AIS_11_22.rda")
+
+AIS_11_23 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_23.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_23, file = "D:/AIS/AIS_West/AIS_11_23.rda")
+
+AIS_11_24 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_24.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_24, file = "D:/AIS/AIS_West/AIS_11_24.rda")
+
+AIS_11_25 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_25.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_25, file = "D:/AIS/AIS_West/AIS_11_25.rda")
+
+AIS_11_26 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_26.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_26, file = "D:/AIS/AIS_West/AIS_11_26.rda")
+
+AIS_11_27 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_27.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_27, file = "D:/AIS/AIS_West/AIS_11_27.rda")
+
+AIS_11_28 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_28.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_28, file = "D:/AIS/AIS_West/AIS_11_28.rda")
+
+AIS_11_29 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_29.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_29, file = "D:/AIS/AIS_West/AIS_11_29.rda")
+
+AIS_11_30 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_11_30.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_11_30, file = "D:/AIS/AIS_West/AIS_11_30.rda")
+
+AIS_12_01 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_12_01.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_12_01, file = "D:/AIS/AIS_West/AIS_12_01.rda")
+
+AIS_12_02 <- read_csv("D:/AIS/AIS_West/West_AIS_2018_12_02.csv", 
+                      show_col_types = FALSE) %>%
+  filter(vesselLength >= 100) %>%
+  filter(SOG > 2.5) %>%
+  filter(Latitude >= min(allDrifts$Latitude - 0.5) & Latitude <= max(allDrifts$Latitude + 0.5)) %>%
+  filter(Longitude >= min(allDrifts$Longitude - 0.5) & Longitude <= max(allDrifts$Longitude + 0.5))
+
+saveRDS(AIS_12_02, file = "D:/AIS/AIS_West/AIS_12_02.rda")
+```
+
+Load in all AIS days
+```{r}
+AISlist <- dir("D:/AIS/AIS_West", recursive=TRUE, full.names=TRUE, pattern=".rda")
+
+for (a in AISlist) {
+  df <- readRDS(a)
+  assign(paste0("AIS_", substr(a, 21, 25)), df)
+}
+```
+
+Bind Al AIS days
+```{r}
+AIS_allpts <- bind_rows(AIS_08_06,AIS_08_07, AIS_08_08, AIS_08_09, AIS_08_10, AIS_08_11, AIS_08_12,
+          AIS_08_13, AIS_08_14, AIS_08_15, AIS_08_16, AIS_08_17, AIS_08_18, AIS_08_19,
+          AIS_08_20, AIS_08_21, AIS_08_22, AIS_08_23, AIS_08_24, AIS_08_25, AIS_08_26,
+          AIS_08_27, AIS_08_28, AIS_08_29, AIS_08_30, AIS_08_31, AIS_09_01, AIS_09_02,
+          AIS_09_03, AIS_09_04, AIS_09_05, AIS_09_06, AIS_09_07, AIS_09_08, AIS_09_09,
+          AIS_09_10, AIS_09_11, AIS_09_12, AIS_09_13, AIS_09_14, AIS_09_15, AIS_09_16,
+          AIS_09_17, AIS_09_18, AIS_09_19, AIS_09_20, AIS_09_21, AIS_09_22, AIS_09_23,
+          AIS_09_24, AIS_09_25, AIS_09_26, AIS_09_27, AIS_09_28, AIS_09_29, AIS_09_30,
+          AIS_10_01, AIS_10_02, AIS_10_03, AIS_10_04, AIS_10_05, AIS_10_06, AIS_10_07,
+          AIS_10_08, AIS_10_09, AIS_10_10, AIS_10_11, AIS_10_12, AIS_10_13, AIS_10_14,
+          AIS_10_15, AIS_10_16, AIS_10_17, AIS_10_18, AIS_10_19, AIS_10_20, AIS_10_21,
+          AIS_10_22, AIS_10_23, AIS_10_24, AIS_10_25, AIS_10_26, AIS_10_27, AIS_10_28,
+          AIS_10_29, AIS_10_30, AIS_10_31, AIS_11_01, AIS_11_02, AIS_11_03, AIS_11_04,
+          AIS_11_05, AIS_11_06, AIS_11_07, AIS_11_08, AIS_11_09, AIS_11_10, AIS_11_11,
+          AIS_11_12, AIS_11_13, AIS_11_14, AIS_11_15, AIS_11_16, AIS_11_17, AIS_11_18,
+          AIS_11_19, AIS_11_20, AIS_11_21, AIS_11_22, AIS_11_23, AIS_11_24, AIS_11_25,
+          AIS_11_26, AIS_11_27, AIS_11_28, AIS_11_29, AIS_11_30, AIS_12_01, AIS_12_02)
+
+# Save as csv
+write_csv(AIS_allpts, file = "D:/AIS/AIS_West/AIS_allpts.csv")
+```

### 2023-11-21 05_analysis.Rmd
Update 05_analysis.Rmd

@@
 
+# Remove points where buoy is on the shelf - 300 meters
+allDrifts <- allDrifts %>%
+  filter(depth > 300)
+
 save(allDrifts, file = "data/allDrifts.rda")
@@
 allDrifts %>%
-  ggplot(mapping = aes(x = Deployment, y = TOL_2000)) + geom_boxplot() + theme_bw()
+  ggplot(mapping = aes(x = Deployment, y = TOL_2000)) + 
+  geom_violin() + 
+  geom_boxplot(width = 0.1) +
+  theme_bw()
 
@@
 #Plot facet grid
-ggplot(ZC_pivot, aes(x = UTC, y = value, color = avg)) + geom_smooth() +
+ggplot(ZC_pivot, aes(x = UTC, y = value, color = avg)) + geom_line() +
   facet_grid(vars(avg), scales = "free") +
-  scale_x_datetime(date_breaks = "3 days") +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-  labs(title = "Rolling Hourly Averages of ZC Detections and Soundscape Metrics -  
-       Buoy 21")
+  scale_x_datetime(date_breaks = "1 days") +
+  theme(axis.text.x = element_text(angle = 90,
+                                   vjust = 0.5, hjust=1),
+        legend.position = "none") +
+  labs(title = "Rolling Hourly Averages of ZC Detections and   
+  Soundscape Metrics - Buoy 21")
 ```
@@
 ```{r}
-ZC_filt_ship <- GPSwhaleAISenvSS08 %>% 
+ZC_filt_ship <- GPSwhaleAISSS20.alt%>% 
   filter(species %in% c("ZC", NA)) %>%
@@
   scale_x_datetime(date_breaks = "3 days") +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+  theme(axis.text.x = element_text(#angle = 90, 
+                                   vjust = 0.5, hjust=1)) +
   labs(title = "Rolling Hourly Averages of ZC Detections + SS Metrics + Ships -  
-       Buoy 08")
+       Buoy 20")
+```
+
+Original data points plotted
+```{r}
+ZC_pivot_reg <- pivot_longer(ZC_filt, cols = c(TOL_125,TOL_2000, Wpresence), names_to = "datacol", values_to = "value")
+
+# Regular plotted detections
+ggplot(ZC_pivot_reg, aes(x = UTC, y = value, color = datacol)) + geom_point(size = 0.1) +
+  facet_grid(vars(datacol), scales = "free") +
+  scale_x_datetime(date_breaks = "3 days") +
+  theme(axis.text.x = element_text(angle = 90,
+                                   vjust = 0.5, hjust=1)) +
+  labs(title = "ZC Detections + SS Metrics - Buoy 21")
+
+ggplot(ZC_pivot_reg, aes(x = UTC, y = value, color = datacol)) + geom_line() +
+  facet_grid(vars(datacol), scales = "free") +
+  scale_x_datetime(date_breaks = "3 days") +
+  theme(axis.text.x = element_text(angle = 90,
+                                   vjust = 0.5, hjust=1)) +
+  labs(title = "ZC Detections + SS Metrics + Ships - Buoy 13")
+
 ```
@@
 ```{r}
-ZC_filt_ship <- GPSwhaleAISenvSS23 %>% 
+ZC_filt_ship <- GPSwhaleAISSS20.alt %>% 
   filter(species %in% c("ZC", NA)) %>%
@@
   labs(title = "Rolling Hourly Averages of ZC Detections and Ships -  
-       Buoy 23")
+       Buoy 20")
 
@@
 ```{r}
+# Read in buoy 20 without environmental variables
+GPSwhaleAISSS20.alt <- readRDS("D:/Buoy_dfs/GPSwhaleAISSS20_alt.rda")
+
+ZC_filt_a <- GPSwhaleAISSS20.alt%>% 
+  filter(species %in% c("ZC", NA)) %>%
+  # Hourly rolling averages
+  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA)) %>%
+  # Drop first and last NAs that are created as part of averages
+  drop_na(., rollAvg_Whl)
+
 # Plot lines on same plot
-
-min_first <- min(ZC_filt$rollAvg_Whl)
-max_first <- max(ZC_filt$rollAvg_Whl)
-min_second <- min(ZC_filt$rollAvg_2000)
-max_second <- max(ZC_filt$rollAvg_2000)
+min_first <- min(ZC_filt_a$rollAvg_Whl)
+max_first <- max(ZC_filt_a$rollAvg_Whl)
+min_second <- min(ZC_filt_a$rollAvg_2000)
+max_second <- max(ZC_filt_a$rollAvg_2000)
 
@@
 # Plot two lines on same plot
-ggplot(ZC_filt, aes(x = UTC)) +
-  geom_smooth(aes(y = rollAvg_Whl, color = "ZC presence")) +
-  geom_smooth(aes(y = inv_scale_function(rollAvg_2000, scale, shift), color = "TOL 2000")) +
+ggplot(ZC_filt_a, aes(x = UTC)) +
+  geom_line(aes(y = rollAvg_Whl, color = "ZC presence")) +
+  geom_line(aes(y = inv_scale_function(rollAvg_2000, scale, shift), color = "TOL 2000")) +
   scale_x_datetime(date_breaks = "3 days") +
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-  labs(title = "Rolling Averages of ZC Detections alongside Soundscape Metrics - Buoy 13")
+  labs(title = "Rolling Averages of ZC Detections alongside   
+       Soundscape Metrics - Buoy 20")
+```
 
+```{r}
 ########################################
@@
 # Sound levels for all species
-ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_125)) +
+ggplot(detsOnly3) + geom_violin(aes(x = species, y = TOL_125)) +
   labs(title = "TOL 125 Sound levels by Species - All Buoys") +
   theme_classic()
-ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_2000)) +
+
+ggplot(detsOnly3, aes(x = species, y = TOL_2000)) + geom_violin() +
   labs(title = "TOL 2000 Sound levels by Species - All Buoys") +
-  theme_classic()
+  theme_classic() +
+  geom_boxplot(width = 0.1)
 

### 2023-11-21 05_randomForest.R
Update 05_randomForest.R

@@ -10,6 +10,13 @@ library(rfPermute)
 # Load in main dataframe
 load(file = "data/allDrifts.rda")
 
+# Create new row with random samples, make sure to add to covariates list
+allDrifts$randSamp <- sample(10000, size = nrow(allDrifts), replace = TRUE)
+
+# Load in main dataframe with seafloor gradient(?)
+#load(file = "data/allDrifts_grad.rda")
+
+
 #Define response variable (presence/absence as 0/1)
 #Create a variable for presence of each species, and then create a
 #random forest model for each species 
@@ -31,7 +38,8 @@ allDrifts$ZcPresence<-factor(ifelse(allDrifts$species=='ZC',1,0))
 covariate.list<-list(c('ZcPresence','curl_mean', 'bathy_slope',
                        'sst_mean','dist2slope','depth','dist2slope','chlorophyll_mean',
                        'chlorophyll_mean','mldDepth','ssh_mean',
-                       'mldDepth', 'mldTemp', 'ttDepth','ttTemp', 'temp400', 'sal400'))
+                       'mldDepth', 'mldTemp', 'ttDepth','ttTemp', 'temp400', 'sal400',
+                       'randSamp'))
 
 include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
 
@@ -49,9 +57,9 @@ sampsizeZC <- balancedSampsize(DF.modelZC$ZcPresence)
 # set seed for reproducibility
 set.seed(123)
 
-# Use random forest, not rfPermute
+# Use random forest
 RanFor.model.ZC <- randomForest(ZcPresence ~ ., data=DF.modelZC[,include.covars],
-                         replace=FALSE, ntree=4000, sampsize=sampsizeZC, 
+                         replace=FALSE, ntree=4400, sampsize=sampsizeZC, 
                          proximity=FALSE, importance = TRUE)
 RanFor.model.ZC
 
@@ -91,7 +99,8 @@ allDrifts$PmPresence<-factor(ifelse(allDrifts$species=='PM',1,0))
 covariate.list<-list(c('PmPresence','curl_mean', 'bathy_slope',
                        'sst_mean','dist2slope','depth','dist2slope','chlorophyll_mean',
                        'chlorophyll_mean','mldDepth','ssh_mean',
-                       'mldDepth', 'mldTemp', 'ttDepth','ttTemp', 'temp400', 'sal400'))
+                       'mldDepth', 'mldTemp', 'ttDepth','ttTemp', 'temp400', 'sal400',
[…]

### 2023-11-21 06_gam.Rmd
Update 06_gam.Rmd

@@
 library(mgcv)
+library(car)
 ```
@@
 
+# Environ + noise
+ZC_GAM_env_TOL <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + mldTemp + bathy_slope +
+                        TOL_125 + TOL_2000,
+                      data = ZC_filt, family = binomial, method = "REML")
+summary(ZC_GAM_env_TOL)
+gam.check(ZC_GAM_env_TOL)
+
+#Environ + noise + ship (** CANT MODEL SHIP)
 ZC_GAM_env_TOL <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + mldTemp + bathy_slope +
@@
               data = ZC_filt, family = binomial, method = "REML")
-summary(ZC_GAM_env_TOL)
-gam.check(ZC_GAM_env_TOL)
 ```
@@
 
+# Environ + noise
 PM_GAM_env_TOL <- gam(Wpresence>0 ~ dist2slope + ttDepth + depth + mldDepth + ssh_mean +
-                        TOL_125 + TOL_2000 + timeAfterShip,
+                        TOL_125 + TOL_2000,
               data = PM_filt, family = binomial, method = "REML")
@@
 gam.check(PM_GAM_env_TOL)
+
+# With ships - DO NOT MODEL SHIPS
+#PM_GAM_env_TOL <- gam(Wpresence>0 ~ dist2slope + ttDepth + depth + mldDepth + ssh_mean +
+#                        TOL_125 + TOL_2000 + timeAfterShip,
+#              data = PM_filt, family = binomial, method = "REML")
 ```
@@
 Higher edfs describe more wiggly curves. Doesn't mean significant
+
+Detecting Multicolinearity with Variance Inflation Factors (VIF) - Cuvier's Only
+```{r}
+ZC_filt <- allDrifts %>% 
+  filter(species %in% c("ZC", NA))
+
+# Environ only
+model1LM <- glm(Wpresence ~ depth + mldDepth
+               + sst_mean + curl_mean + dist2slope
+               + bathy_slope + dist2shore 
+               + chlorophyll_mean + ssh_mean 
+               + ttDepth + temp400 + sal400,
+               data = ZC_filt)
+
+vif(model1LM)
+
+vif_values <- tibble(vif(model1LM)) %>%
+  cbind(var = c("depth", "mldDepth", "sst_mean", "curl_mean", "dist2slope", "bathy_slope",
+          "dist2shore", "chlorophyll_mean", "ssh_mean", "ttDepth", "temp400", "sal400")) %>%
+  rename(val = "vif(model1LM)")
+# Need to remove mldTemp or ttTemp for it to work.
+
+ggplot(vif_values, aes(y = var, x = val)) + 
+  geom_col() + 
+  geom_vline(xintercept = 5, col = "red") +
+  labs(title = "VIF Values by Env. Variable - Cuvier's BW only")
+```
+
+Detecting Multicolinearity with Variance Inflation Factors (VIF) - Sperm whale only
+**IDENTICAL RESULTS TO ZC ONE***
+```{r}
+PM_filt <- allDrifts %>% 
+  filter(species %in% c("PM", NA))
+
+# Environ only
+model1LM <- glm(Wpresence ~ depth + mldDepth
+               + sst_mean + curl_mean + dist2slope
+               + bathy_slope + dist2shore 
+               + chlorophyll_mean + ssh_mean 
+               + ttDepth + temp400 + sal400,
+               data = PM_filt)
+
+vif(model1LM)
+
+vif_values <- tibble(vif(model1LM)) %>%
+  cbind(var = c("depth", "mldDepth", "sst_mean", "curl_mean", "dist2slope", "bathy_slope",
+          "dist2shore", "chlorophyll_mean", "ssh_mean", "ttDepth", "temp400", "sal400")) %>%
+  rename(val = "vif(model1LM)")
+# Need to remove mldTemp or ttTemp for it to work -- perfectly correlated!
+
+ggplot(vif_values, aes(y = var, x = val)) + 
+  geom_col() + 
+  geom_vline(xintercept = 5, col = "red") +
+  labs(title = "VIF Values by Env. Variable - Sperm whales only")
+```
+Trying GAM with correlated variables removed
+```{r}
+# Cuvier's BW
+zc.GAM.VIF <- gam(Wpresence>0 ~ ttDepth + ssh_mean + mldDepth + depth + curl_mean +
+            chlorophyll_mean + bathy_slope, data = ZC_filt)
+summary(zc.GAM.VIF)
+gam.check(zc.GAM.VIF)
+
+# Sperm whale
+pm.GAM.VIF <- gam(Wpresence>0 ~ ttDepth + ssh_mean + mldDepth + depth + curl_mean +
+            chlorophyll_mean + bathy_slope + TOL_125 + TOL_2000, data = PM_filt)
+summary(pm.GAM.VIF)
+gam.check(pm.GAM.VIF)
+
+```
+
+Load seafloor depth gradient dataset and rerun GAM
+```{r}
+load(file = "data/allDrifts_grad.rda")
+
+ZC_filt2 <- allDrifts_grad %>% 
+  filter(species %in% c("ZC", NA))
+
+# Environ only
+model2LM <- glm(Wpresence ~ depth + mldDepth
+               + sst_mean + curl_mean + dist2slope
+               + bathy_slope + dist2shore 
+               + chlorophyll_mean + ssh_mean 
+               + ttDepth + temp400 + sal400 +
+               magnitude_gradient_mean,
+               data = ZC_filt2)
+
+vif(model2LM)
+
+vif_values <- tibble(vif(model2LM)) %>%
+  cbind(var = c("depth", "mldDepth", "sst_mean", "curl_mean", "dist2slope", "bathy_slope",
+          "dist2shore", "chlorophyll_mean", "ssh_mean", "ttDepth", "temp400", "sal400",
+          "magnitude_gradient_mean")) %>%
+  rename(val = "vif(model2LM)")
+# Need to remove mldTemp or ttTemp for it to work.
+
+ggplot(vif_values, aes(y = var, x = val)) + 
+  geom_col() + 
+  geom_vline(xintercept = 5, col = "red") +
+  labs(title = "VIF Values by Env. Variable - Cuvier's BW only")
+```
+
+```{r}
+# GAM with seafloor gradient
+
+# Cuvier's
+model3GAM <- gam(Wpresence ~ depth + mldDepth
+               + curl_mean + bathy_slope
+               + chlorophyll_mean + ssh_mean +
+               magnitude_gradient_mean,
+               data = ZC_filt2)
+summary(model3GAM)
+# Deviance explained = 0.854%
+
+# Sperm whale
+PM_filt2 <- allDrifts_grad %>% 
+  filter(species %in% c("PM", NA))
+
+model4GAM <- gam(Wpresence ~ depth + mldDepth
+               + curl_mean + bathy_slope
+               + chlorophyll_mean + ssh_mean +
+               magnitude_gradient_mean,
+               data = PM_filt2)
+summary(model4GAM)
+# Deviance explained = 0.816%
+```

### 2023-11-21 07_summary_stats.Rmd
Update 07_summary_stats.Rmd

@@
   select(UTC, Wpresence, species) %>%
-  filter(species == "PM")
+  filter(species == "ZC")
 
@@
 
-g1 <- (plotTimeseries(GPSwhaleAISenvSS08, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 125 Plot + PM'))
+g1 <- (plotTimeseries(GPSwhaleAISenvSS08, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 125 Plot + zC'))
 g2 <- plotPresGrid(W.08, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS08$UTC),
@@
 #####
-t1 <- plotTimeseries(GPSwhaleAISenvSS08, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 2000 Plot + PM')
+t1 <- plotTimeseries(GPSwhaleAISenvSS08, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 8: TOL 2000 Plot + ZC')
 t2 <- plotPresGrid(W.08, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS08$UTC),
@@
 
-g1 <- (plotTimeseries(GPSwhaleAISenvSS10, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 125 Plot + PM'))
+g1 <- (plotTimeseries(GPSwhaleAISenvSS10, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 125 Plot + ZC'))
 g2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS10$UTC),
@@
 #####
-t1 <- plotTimeseries(GPSwhaleAISenvSS10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot + PM')
+t1 <- plotTimeseries(GPSwhaleAISenvSS10, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 10: TOL 2000 Plot + ZC')
 t2 <- plotPresGrid(W.10, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS10$UTC),
@@
 ```{r fig.width=12}
+# Read in buoy 20 without environmental variables
+GPSwhaleAISSS20.alt <- readRDS("D:/Buoy_dfs/GPSwhaleAISSS20_alt.rda")
+
 ## Buoy 20
-W.20 <- GPSwhaleAISenvSS20 %>%
+W.20 <- GPSwhaleAISSS20.alt %>%
   select(UTC, Wpresence, species) %>%
@@
 
-g1 <- (plotTimeseries(GPSwhaleAISenvSS20, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot + ZC'))
-g2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS20$UTC),
-                         end = max(GPSwhaleAISenvSS20$UTC), fill = NA, color = "white")
-
-combined.g <- g1
-combined.g$layers <- c(combined.g$layers, g2$layers)
-#####
-t1 <- plotTimeseries(GPSwhaleAISenvSS20, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot + ZC')
-t2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS20$UTC),
-                         end = max(GPSwhaleAISenvSS20$UTC), fill = NA, color = "white")
+g1 <- (plotTimeseries(GPSwhaleAISSS20.alt, column='TOL_125', bin='1hour', 
+                      style = 'heatmap', title='Buoy 20: TOL 125 Plot + ZC'))
+g2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISSS20.alt$UTC),
+                         end = max(GPSwhaleAISSS20.alt$UTC), fill = NA, color = "white")
+
+combined.g <- g1
+combined.g$layers <- c(combined.g$layers, g2$layers)
+#####
+t1 <- plotTimeseries(GPSwhaleAISSS20.alt, column='TOL_2000', bin='1hour', 
+                     style = 'heatmap', title='Buoy 20: TOL 2000 Plot + ZC')
+t2 <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISSS20.alt$UTC),
+                         end = max(GPSwhaleAISSS20.alt$UTC), fill = NA, color = "white")
 
@@
 
-W.20 <- GPSwhaleAISenvSS20 %>%
-  select(UTC, Wpresence, species) %>%
-  filter(species == "ZC")
-
-W.20plot <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISenvSS20$UTC),
-                         end = max(GPSwhaleAISenvSS20$UTC),  by = 'Wpresence', alpha = 0.5)
-
-plotTimeseries(GPSwhaleAISenvSS20, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot') +  W.20plot
-plotTimeseries(GPSwhaleAISenvSS20, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot') +  W.20plot
+W.20 <- GPSwhaleAISSS20.alt %>%
+  select(UTC, Wpresence, species) %>%
+  filter(species == "ZC")
+
+W.20plot <- plotPresGrid(W.20, bin = 'hour', type = 'presence', start = min(GPSwhaleAISSS20.alt$UTC),
+                         end = max(GPSwhaleAISSS20.alt$UTC),  by = 'Wpresence', alpha = 0.5)
+
+plotTimeseries(GPSwhaleAISSS20.alt, column='TOL_2000', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 2000 Plot') +  W.20plot
+plotTimeseries(GPSwhaleAISSS20.alt, column='TOL_125', bin='1hour', style = 'heatmap', title='Buoy 20: TOL 125 Plot') +  W.20plot
 

### 2023-11-21 09_psd.R
Create 09_psd.R

@@ -1,0 +1,333 @@
+# SCript from Anne Simonis 10-20-23
+library(tidyverse)
+library(PAMscapes)
+
+#Load in median power spectral density (PSD) files
+psd13 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_13_1Hz_1s_PSD_2min.csv')
+psd14 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_14_1Hz_1s_PSD_2min.csv')
+psd07 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_07_1Hz_1s_PSD_2min.csv')
+psd08 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_08_1Hz_1s_PSD_2min.csv')
+psd10 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_10_1Hz_1s_PSD_2min.csv')
+psd12 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_12_1Hz_1s_PSD_2min.csv')
+psd16 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_16_1Hz_1s_PSD_2min.csv')
+psd18 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_18_1Hz_1s_PSD_2min.csv')
+psd19 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_19_1Hz_1s_PSD_2min.csv')
+psd20 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_20_1Hz_1s_PSD_2min.csv')
+psd21 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_21_1Hz_1s_PSD_2min.csv')
+psd22 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_22_1Hz_1s_PSD_2min.csv')
+psd23 <- checkSoundscapeInput('D:/Soundscape_metrics/PSD/CCES_23_1Hz_1s_PSD_2min.csv')
+
+# Load in whale/GPS detections
+GPSwhale13 <- readRDS(file = "D:/Buoy_dfs/GPSwhale13.rda")
+GPSwhale14 <- readRDS(file = "D:/Buoy_dfs/GPSwhale14.rda")
[…]

### 2023-12-08 seafloor_gradi.R
Update seafloor_gradi.R

@@ -3,6 +3,9 @@
 # Required packages
 library(tidyverse)
 library(terra)
+library(sf)
+library(mgcv)
+library(car)
 
 # Read in buoy dataframes
 ftlist <- dir("D:/Buoy_dfs/", recursive=TRUE, full.names=TRUE, pattern="GPSwhaleAISenvSS")
@@ -97,12 +100,14 @@ allDrifts_grad <- bind_rows(grad07,grad08, grad10, grad12, grad13, grad14,
                            grad16, grad18, grad19, grad20, grad21, grad22, grad23)
 
 save(allDrifts_grad, file = "data/allDrifts_grad.rda")
-
+load(file = "data/allDrifts_grad.rda")
 
 ## Correlate TOL 2000 levels with seafloor gradient
 grad_lm <- lm(magnitude_gradient_mean ~ TOL_2000, allDrifts_grad)
 grad_glm <- glm(magnitude_gradient_mean ~ TOL_2000, data = allDrifts_grad)
-
+grad_gam <- gam(magnitude_gradient_mean ~ s(TOL_2000), data = allDrifts_grad)
+plot(grad_gam, all.terms = TRUE)
+summary(grad_gam)
 summary(grad_glm)
 
 
@@ -119,39 +124,101 @@ ggplot(allDrifts_grad, aes(x = Longitude, y = Latitude)) +
 
 seafloor <- rast("D:/Environ_variables/SeafloorGradient.tif")
 
-## from 03 ##
+load("data/allDrifts.rda")
+allDrifts_sf <- allDrifts %>%
+  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
 
-agg <- terra::aggregate(bathy, 5)  # Increase cell size by factor of 5 to be visible
 
-cont <- st_as_sf(as.contour(bathy, )) # Create contour lines from bathymetry 
-cont_slope <- cont %>% filter(level == -2000) # Establish continental shelf. Is 2000m the correct depth for continental shelf?
+# Establish bounding box of study area
+xmin <- st_bbox(allDrifts_sf)[1]
+xmax <- st_bbox(allDrifts_sf)[3]
+ymin <- st_bbox(allDrifts_sf)[2]
+ymax <- st_bbox(allDrifts_sf)[4]
 
-#Make sure to clear Global Environment of other objects with "env"
-tracklist <- grep(pattern = "env", names(.GlobalEnv), value = T)
+ext(seafloor) <- c(xmin, xmax, ymin, ymax)
+plot(seafloor)
 
[…]

### 2023-12-08 multicollin.R
Create multicollin.R

@@ -1,0 +1,77 @@
+#Testing for Multicollinearity
+
+library(tidyverse)
+library(car)
+
+
+load(file = "D:Buoy_dfs/drift_escarpment.rda")
+
+############################################################################
+
+# All covariates except mldTemp
+# Won't run due to aliased coefficients unless I remove mldTemp or ttTemp
+GLM.VIF1 <- glm(Wpresence>0 ~ dist2slope + mldDepth + curl_mean + sst_mean +
+                  depth + bathy_slope + chlorophyll_mean + dist2shore +
+                  ssh_mean + ttTemp + ttDepth + dist2escarp + sal400 + temp400,
+                data = drift_escarp)
+
+vif(GLM.VIF1)
+
+vif_values1 <- tibble(vif(GLM.VIF1)) %>%
+  cbind(var = c("dist2slope", "mldDepth", "curl_mean", "sst_mean", "depth", 
+                "bathy_slope", "chlorophyll_mean", "dist2shore", "ttTemp", 
+                "ssh_mean", "sal400", "temp400",
+                "ttDepth", "dist2escarp")) %>%
+  rename(val = "vif(GLM.VIF1)")
+
+ggplot(vif_values1, aes(y = var, x = val)) + 
+  geom_col() + 
+  geom_vline(xintercept = 5, col = "red") +
+  labs(title = "VIF Values, ALL Env. Variables")
+
+
+###############################################################################
+# Remove highest of collinear pairs.
+# Colinear pairs with removed covariate in UPPERCASE: 
+# dist2slope/DIST2SHORE, ttTemp/SST_MEAN, temp400/SAL400, TTDEPTH/mldDepth, ttTemp/TEMP400
+
+GLM.VIF2 <- glm(Wpresence>0 ~ dist2slope + mldDepth + curl_mean +
+                  depth + bathy_slope + chlorophyll_mean +
+                  ssh_mean + ttTemp + dist2escarp,
+                data = drift_escarp)
+
+vif(GLM.VIF2)
+
+vif_values2 <- tibble(vif(GLM.VIF2)) %>%
+  cbind(var = c("dist2slope", "mldDepth", "curl_mean", "depth", 
+                "bathy_slope", "chlorophyll_mean", "ssh_mean", 
+                "ttTemp", "dist2escarp")) %>%
+  rename(val = "vif(GLM.VIF2)")
+
+ggplot(vif_values2, aes(y = var, x = val)) + 
+  geom_col() + 
[…]

### 2023-12-08 manuscript.rmd
Update manuscript.rmd
added language about Tweedie distributions

@@
 
+"The Tweedie is a family of exponential type distributions, which are tolerant to large numbers of zero observations (Candy 2004)." -- Lacey and Hammond 2023
+
+(Candy, S.G. 2004. Modeling catch and effort data using Generalised Linear Models, the Tweedie Distriution, Random Vessel Effects and Random Stratum-by-Year effects. CCAMLR Sci. 11: 59--80.)
+
 ## Results
@@
 | Covariate                          | Resolution       | Source                                                                                                                                            | Justification                           |
-|------------------|------------------|--------------------|------------------|
+|------------------------------------|------------------|---------------------------------------------------------------------------------------------------------------------------------------------------|-----------------------------------------|
 |                                    |                  |                                                                                                                                                   |                                         |

### 2023-12-08 06_gam.Rmd
Update 06_gam.Rmd
testing GAMs with top 3 rf covariates

--- 2023-10-16 06_gam.Rmd
@@
 Required Packages
-```{r setup}
+```{r}
 library(tidyverse)
 library(mgcv)
+library(car)
 ```
 
-Load allDrifts object
-```{r setup, include=FALSE}
-load(allDrifts, file = "data/allDrifts.rda")
+Load allDrifts object, split out Cuvier's and Sperm whales
+```{r}
+load(file = "data/allDrifts.rda")
+
+ZC_filt <- allDrifts %>% 
+  filter(species %in% c("ZC", NA))
+
+PM_filt <- allDrifts %>% 
+  filter(species %in% c("PM", NA))
+
 ```
@@
 
+# GAM for Cuver's Beaked whale (ZC)
+
 ```{r}
-model3 <- gam(Wpresence ~ depth + TOL_125 + dist2shore + mldTemp, 
-              data = allDrifts, family = binomial)
-#plot(model3)
+# Environ only - binomial
+ZC_GAM1 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
+              data = ZC_filt, family = binomial, method = "REML")
+summary(ZC_GAM1)
+gam.check(ZC_GAM1)
 
-#try fitting linear model
+# Environ + noise - binomial
+ZC_GAM2<- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
+                      data = ZC_filt, family = binomial, method = "REML")
+summary(ZC_GAM2)
+gam.check(ZC_GAM2)
 
-lm_depth <- lm(Wpresence ~ depth, data = GPSwhaleAISenvSS10)
-ggplot(allDrifts, aes(sst_mean, chlorophyll_mean)) +
-  geom_point() +
-  geom_smooth()
-```
-Categorical term
-```{r}
-# ALL variables, GAM
-model1 <- gam(Wpresence>0 ~ s(TOL_125) + s(TOL_2000) + s(depth) + s(mldDepth) + s(Latitude)
-              + s(Longitude) + s(sst_mean) + s(curl_mean) + s(dist2slope) 
-              + s(bathy_slope) + s(dist2shore) + s(chlorophyll_mean) + s(ssh_mean) 
-              + s(mldTemp) + s(ttDepth) + s(ttTemp) + s(temp400) + s(sal400),
-              data = allDrifts, family = binomial, method = "REML")
+# Environ only - poisson
+ZC_GAM3 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
+              data = ZC_filt, family = poisson, method = "REML")
+summary(ZC_GAM3)
+gam.check(ZC_GAM3)
 
-# ALL variables, GLM
-model1GLM <- glm(Wpresence ~ TOL_125 + TOL_2000 + depth + mldDepth + Latitude
-              + Longitude + sst_mean + curl_mean + dist2slope
-              + bathy_slope + dist2shore + chlorophyll_mean + ssh_mean 
-              + mldTemp + ttDepth + ttTemp + temp400 + sal400,
-              data = allDrifts, family = binomial)
+# Environ + noise - poisson
+ZC_GAM4<- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
+                      data = ZC_filt, family = poisson, method = "REML")
+summary(ZC_GAM4)
+gam.check(ZC_GAM4)
+plot(ZC_GAM4, all.terms = TRUE, scale = 0)
 
-# By species? doesnt work, may need to create dummy variables
-model2 <- gam(Wpresence ~ s(sst_mean, by = as.factor(species)), 
-              data = allDrifts, family = binomial, method = "REML")
+# Environ only - gaussian
+ZC_GAM5 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
+                  data = ZC_filt, method = "REML")
+summary(ZC_GAM5)
+gam.check(ZC_GAM5)
 
-# Soundscape metrics only, GAM
-model4 <- gam(Wpresence ~ s(TOL_125, k = 10) + 
-                s(TOL_2000, k = 10),
-              data = allDrifts, family = binomial, method = "REML")
+# Environ + noise - gaussian
+ZC_GAM6 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
+                  data = ZC_filt, method = "REML")
+summary(ZC_GAM6)
+gam.check(ZC_GAM6)
 
-# Soundscape metrics only, GLM
-model4GLM <- glm(Wpresence ~ TOL_125 + TOL_2000,
-              data = allDrifts, family = binomial)
+# Environ only - Tweedie
+ZC_GAM7 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
+                  data = ZC_filt, family = tw(link = "log"), method = "REML")
+summary(ZC_GAM7)
+gam.check(ZC_GAM7)
 
-model125 <- gam(Wpresence ~ s(TOL_125, k = 10),
-              data = allDrifts, family = binomial, method = "REML")
-plot(model125, trans = plogis)
-summary(model125)
-gam.check(model125)
+# Environ + noise - Tweedie
+ZC_GAM8 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
+                  data = ZC_filt, family = tw(link = "log"), method = "REML")
+summary(ZC_GAM8)
+gam.check(ZC_GAM8)
 
-# Environmental var only
-model5 <- gam(Wpresence ~ s(depth) + s(mldDepth) + s(Latitude)
-              + s(Longitude) + s(sst_mean) + s(curl_mean) + s(dist2slope) 
-              + s(bathy_slope) + s(dist2shore) + s(chlorophyll_mean) + s(ssh_mean) 
-              + s(mldTemp) + s(ttDepth) + s(ttTemp) + s(temp400) + s(sal400),
-              data = allDrifts, family = binomial, method = "REML")
-
-plogis(coef(model1[1]))
-summary(model1)
-summary(model4)
-summary(model5)
-summary(model4GLM)
-plot(model4GLM)
-gam.check(model1)
-gam.check(model4)
-
-#Log odds scale (when used with family = binomial)
-plot(model1, pages = 6)
-plot(model4, pages = 4, trans = plogis)
-
-#Probability scale + centered on the intercept + adding intercept-related uncertainty with seWithMean arg
-plot(model1, pages = 14, trans = plogis,
-     shift = coef(model1)[1])#,
-     #seWithMean = TRUE)
-
-plot(model2, pages = 1, trans = plogis,
-     shift = coef(model2)[1])
-
-plot(model4, pages = 6, trans = plogis,
-     shift = coef(model4)[1])
-```
-
-Top 5 environmental variables (by RF) with one drift
-```{r}
-# Environ only
-model1 <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + ttDepth + sal400,
-              data = GPSwhaleAISenvSS23, family = binomial, method = "REML")
-
-plot(model1, trans = plogis, all.terms = TRUE)
-k.check(model1)
-summary(model1)
-gam.check(model1)
-
-# Add soundscape metrics to model 1
-model2 <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + ttDepth + sal400
-              + TOL_125 + TOL_2000,
-              data = GPSwhaleAISenvSS10, family = binomial, method = "REML")
-
-#plot(model2, trans = plogis) ## doesn't work
-summary(model2)
-gam.check(model2)
-# 18.3% deviance explained (buoy 7)
-# 1.25% deviance explained (buoy 8)
-# 8.56% deviance explained (buoy 10)
-# 1.39% deviance explained (buoy 12)
-# 11.1% deviance explained (buoy 13)
-# 28% deviance explained (buoy 14)
-# 15% deviance explained (buoy 16)
-# 9.92% deviance explained (buoy 18)
-# 8.57% deviance explained (buoy 19)
-# 7.3% deviance explained (buoy 20)
-# 38.3% deviance explained (buoy 21)
-# 15.2 % deviance explained (buoy 22)
-# 20.4% deviance explained (buoy 23)
-
-# Depth parameter only
-model_depth_only <- gam(Wpresence>0 ~ depth,
-                        data = GPSwhaleAISenvSS10, family = binomial, method = "REML")
-summary(model_depth_only)
-gam.check(model_depth_only)
-```
-GAM with FULL DATASET (not great)
-```{r}
-alldriftGAM <- gam(Wpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400
-              + TOL_125 + TOL_2000,
-              data = allDrifts, family = binomial, method = "REML")
-summary(alldriftGAM)
-gam.check(alldriftGAM)
-```
-
-GAMs with s()
-```{r}
-# Environ only, with s()
-model3 <- gam(Wpresence>0 ~ s(depth) + s(sst_mean) + s(curl_mean) + 
-                s(ttDepth) + s(sal400),
-              data = GPSwhaleAISenvSS07, family = binomial, method = "REML")
-
-plot(model3, trans = plogis)
-summary(model3)
-gam.check(model3)
-# 61.9 % deviance explained (buoy 7)
-# 1.42% deviance explained (buoy 8)
-# 12.5% deviance explained (buoy 10)
-# 1.66% deviance explained (buoy 12)
-# 12.8% deviance explained (buoy 13)
-# 41.4% deviance explained (buoy 14)
-# 39% deviance explained (buoy 16)
-# 16.6% deviance explained (buoy 18)
-# 15.5% deviance explained (buoy 19)
-# 5.8% deviance explained (buoy 20)
-## Iteration limit reached without full convergence -- 71.7% deviance explained (buoy 21) ???
-# 18.6% deviance explained (buoy 22)
-# 49.6% deviance explained (buoy 23)
-
-#Add soundscape metrics to model 3
-model4 <- gam(Wpresence>0 ~ s(depth) + s(sst_mean) + s(curl_mean) + 
-                s(ttDepth) + s(sal400) + s(TOL_125) + s(TOL_2000),
-              data = GPSwhaleAISenvSS07, family = binomial, method = "REML")
-
-plot(model4, trans = plogis)
-summary(model4)
-gam.check(model4)
-# 14.5% deviance explained
-
-# mix and match?
-model5 <- gam(BWpresence>0 ~ depth + sst_mean + s(mldTemp) + 
-                s(ttDepth) + sal400,
-              data = fulltrack21, family = binomial, method = "REML")
-
-summary(model5)
-gam.check(model5)
-# 57.4 % deviance explained
-plot(model5)
-
-```
-Higher edfs describe more wiggly curves. Doesn't mean significant
-
-#GAM by ZC
-```{r}
-ZC_filt <- allDrifts %>% 
-  filter(species %in% c("ZC", NA))
-
-ZC_GAM <- gam(Wpresence>0 ~ depth + mldTemp + ttDepth + sal400,
-              data = ZC_filt, family = binomial, method = "REML")
-summary(ZC_GAM)
-gam.check(ZC_GAM)
+plot(ZC_GAM1, all.terms = TRUE, scale = 0, pages = 1)
+plot(ZC_GAM2, all.terms = TRUE, scale = 0, pages = 2)
+plot(ZC_GAM3, all.terms = TRUE, scale = 0, pages = 1)
+plot(ZC_GAM4, all.terms = TRUE, scale = 0, pages = 2)
+plot(ZC_GAM5, all.terms = TRUE, scale = 0, pages = 1)
+plot(ZC_GAM6, all.terms = TRUE, scale = 0, pages = 2)
+plot(ZC_GAM7, all.terms = TRUE, scale = 0, pages = 1)
+plot(ZC_GAM8, all.terms = TRUE, scale = 0, pages = 2)
+AIC(ZC_GAM4, ZC_GAM8)
 ```
@@
 # GAM by PM
+
+```{r}
+# Environ only - binomial
+PM_GAM1 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + 
+                    s(sst_mean),
+              data = PM_filt, family = binomial, method = "REML")
+summary(PM_GAM1)
+gam.check(PM_GAM1)
+
+# Environ + noise - binomial
+PM_GAM2 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + 
+                 s(sst_mean) + s(TOL_125) + s(TOL_2000),
+              data = PM_filt, family = binomial, method = "REML")
+summary(PM_GAM2)
+gam.check(PM_GAM2)
+
+# Environ only - poisson
+PM_GAM3 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
+              data = PM_filt, family = poisson, method = "REML")
+summary(PM_GAM3)
+gam.check(PM_GAM3)
+
+# Environ + noise - poisson
+PM_GAM4 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) 
+               + s(sst_mean) + s(TOL_125) + s(TOL_2000),
+              data = PM_filt, family = poisson, method = "REML")
+summary(PM_GAM4)
+gam.check(PM_GAM4)
+
+# Environ only - gaussian
+PM_GAM5 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
+              data = PM_filt, method = "REML")
+summary(PM_GAM5)
+gam.check(PM_GAM5)
+
+# Environ + noise - gaussian
+PM_GAM6 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean)
+               + s(TOL_125) + s(TOL_2000),
+              data = PM_filt, method = "REML")
+summary(PM_GAM6)
+gam.check(PM_GAM6)
+
+# Environ only - Tweedie
+PM_GAM7 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
+              data = PM_filt, family = tw(link = "log"), method = "REML")
+summary(PM_GAM7)
+gam.check(PM_GAM7)
+
+# Environ + noise - Tweedie
+PM_GAM8 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean)
+               + s(TOL_125) + s(TOL_2000),
+              data = PM_filt, family = tw(link = "log"), method = "REML")
+summary(PM_GAM8)
+gam.check(PM_GAM8)
+
+plot(PM_GAM1, all.terms = TRUE, scale = 0)
+plot(PM_GAM2, all.terms = TRUE, scale = 0)
+plot(PM_GAM3, all.terms = TRUE, scale = 0)
+plot(PM_GAM4, all.terms = TRUE, scale = 0)
+plot(PM_GAM5, all.terms = TRUE, scale = 0)
+plot(PM_GAM6, all.terms = TRUE, scale = 0)
+plot(PM_GAM7, all.terms = TRUE, scale = 0)
+plot(PM_GAM8, all.terms = TRUE, scale = 0)
+AIC(PM_GAM4, PM_GAM8)
+
+```
+
+Detecting Multicollinearity with Variance Inflation Factors (VIF) - Cuvier's Only
+This appears to only work with GLMs?
+```{r}
+zc.GLM.VIF1 <- glm(Wpresence>0 ~ depth + temp400 + sst_mean + TOL_125 + TOL_2000,
+                      data = ZC_filt)
+
+vif(zc.GLM.VIF1)
+
+vif_values1 <- tibble(vif(zc.GLM.VIF1)) %>%
+  cbind(var = c("depth", "temp400", "sst_mean",
+                "TOL_125", "TOL_2000")) %>%
+  rename(val = "vif(zc.GLM.VIF1)")
+
+ggplot(vif_values1, aes(y = var, x = val)) + 
+  geom_col() + 
+  geom_vline(xintercept = 5, col = "red") +
+  labs(title = "VIF Values by Env. Variable - Cuvier's BW only")
+```
+
+Detecting Multicolinearity with Variance Inflation Factors (VIF) - Sperm whale only
 ```{r}
@@
 
-PM_GAM <- gam(Wpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400 +
-                TOL_125 + TOL_2000,
-              data = PM_filt, family = binomial, method = "REML")
-summary(PM_GAM)
-gam.check(PM_GAM)
+pm.GLM.VIF1 <- glm(Wpresence>0 ~ dist2slope + mldDepth +
+                     sst_mean + TOL_125 + TOL_2000, data = PM_filt)
 
-# Using variables from Fiedler et al. 2023 for sperm whales
-PM_GAM_fiedlerVar <- gam(Wpresence>0 ~ depth + mldDepth + curl_mean + ssh_mean + sal400,
-              data = PM_filt, family = binomial, method = "REML")
-summary(PM_GAM_fiedlerVar)
+vif(pm.GLM.VIF1)
+
+vif_values2 <- tibble(vif(pm.GLM.VIF1)) %>%
+  cbind(var = c("mldDepth", "dist2slope",
+                 "TOL_125", "TOL_2000", "sst_mean")) %>%
+  rename(val = "vif(pm.GLM.VIF1)")
+
+ggplot(vif_values2, aes(y = var, x = val)) + 
+  geom_col() + 
+  geom_vline(xintercept = 5, col = "red") +
+  labs(title = "VIF Values by Env. Variable - Sperm whales only")
 ```
+
+Load seafloor depth gradient dataset and rerun GAM
+```{r}
+load(file = "data/allDrifts_grad.rda")
+
+ZC_filt2 <- allDrifts_grad %>% 
+  filter(species %in% c("ZC", NA))
+
+# Environ only
+model2LM <- glm(Wpresence ~ depth + mldDepth
+               + sst_mean + curl_mean + dist2slope
+               + bathy_slope + dist2shore 
+               + chlorophyll_mean + ssh_mean 
+               + ttDepth + temp400 + sal400 +
+               magnitude_gradient_mean,
+               data = ZC_filt2)
+
+vif(model2LM)
+
+vif_values <- tibble(vif(model2LM)) %>%
+  cbind(var = c("depth", "mldDepth", "sst_mean", "curl_mean", "dist2slope", "bathy_slope",
+          "dist2shore", "chlorophyll_mean", "ssh_mean", "ttDepth", "temp400", "sal400",
+          "magnitude_gradient_mean")) %>%
+  rename(val = "vif(model2LM)")
+# Need to remove mldTemp or ttTemp for it to work.
+
+ggplot(vif_values, aes(y = var, x = val)) + 
+  geom_col() + 
+  geom_vline(xintercept = 5, col = "red") +
+  labs(title = "VIF Values by Env. Variable - Cuvier's BW only")
+```
+
+Should I include seafloor gradient in GAM?
+```{r}
+# GAM with seafloor gradient
+
+# Cuvier's
+model3GAM <- gam(Wpresence ~ s(depth) + s(temp400) + s(sst_mean) + 
+                   s(magnitude_gradient_mean),
+               data = ZC_filt2)
+summary(model3GAM)
+plot(model3GAM, all.terms = TRUE, scale = 0)
+# Deviance explained = 0.854%
+
+# Sperm whale
+PM_filt2 <- allDrifts_grad %>% 
+  filter(species %in% c("PM", NA))
+
+model4GAM <- gam(Wpresence ~ depth + mldDepth
+               + curl_mean + bathy_slope
+               + chlorophyll_mean + ssh_mean +
+               magnitude_gradient_mean,
+               data = PM_filt2)
+summary(model4GAM)
+# Deviance explained = 0.816%
+```
+Does distance to escarpment improve the GAM?
+```{r}
+load(file = "D:Buoy_dfs/drift_escarpment.rda")
+
+ZC_filt2 <- drift_escarp %>% 
+  filter(species %in% c("ZC", NA))
+
+PM_filt2 <- drift_escarp %>% 
+  filter(species %in% c("PM", NA))
+#####
+
+# Poisson
+ZC_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
+              data = ZC_filt2, family = poisson, method = "REML")
+summary(ZC_GAM_escarp)
+plot(ZC_GAM_escarp, all.terms = TRUE)
+
+# Tweedie (family = tw(link = "log"))
+ZC_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
+              data = ZC_filt2, family = tw(link = "log"), method = "REML")
+summary(ZC_GAM_escarp)
+plot(ZC_GAM_escarp, all.terms = TRUE)
+
+# Poisson
+PM_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(dist2slope) + s(mldDepth) + s(sst_mean)
+               + s(TOL_125) + s(TOL_2000),
+              data = PM_filt2, family = poisson, method = "REML")
+summary(PM_GAM_escarp)
+plot(PM_GAM_escarp, all.terms = TRUE)
+plot.gam(PM_GAM_escarp)
+
+# Tweedie
+PM_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(dist2slope) + s(mldDepth) + s(sst_mean)
+               + s(TOL_125) + s(TOL_2000),
+              data = PM_filt2, family = tw(link = "log"), method = "REML")
+summary(PM_GAM_escarp)
+```
+
+With TOL2000 included
+```{r}
+escarp_gam <- gam(TOL_2000 ~ s(as.numeric(dist2escarp)), 
+                  data = drift_escarp)
+summary(escarp_gam)
+plot(escarp_gam, all.terms = TRUE, scale = 0)
+```

### 2023-12-08 05_randomForest.R
Update 05_randomForest.R
could not complete with 15k trees and all covariates

@@ -95,7 +95,7 @@ allDrifts$species<-as.factor(allDrifts$species) %>%
 
 allDrifts$PmPresence<-factor(ifelse(allDrifts$species=='PM',1,0))
 
-#List all variables for evaluation
+# List all variables for evaluation
 covariate.list<-list(c('PmPresence','curl_mean', 'bathy_slope',
                        'sst_mean','dist2slope','depth','dist2slope','chlorophyll_mean',
                        'chlorophyll_mean','mldDepth','ssh_mean',
@@ -133,8 +133,6 @@ plotPredictedProbs(RanFor.model.PM)
 dev.off()
 save.image(paste(Dep,"_", string.covars.used, ".RData", sep=""))
 
-# Top 5 variables are:
-# TOL_2000, dist2slope, timeaftership, ttDepth, depth
 
 # Top 5 environmental variables only:
 # dist2slope, ttDepth, mldDepth, ssh_mean, depth
@@ -148,15 +146,31 @@ save.image(paste(Dep,"_", string.covars.used, ".RData", sep=""))
 ## Cuvier's BW
 
 # Estimate Permutation p-values for Random Forest Importance Metrics
-# takes over 1 hr to run.
+# DOES NOT COMPLETE
 RF.model.ZC <- rfPermute(ZcPresence ~ ., data=DF.modelZC[,include.covars],
-                                replace=FALSE, ntree=7000, sampsize=sampsizeZC, 
+                                replace=FALSE, ntree=11000, sampsize=sampsizeZC, 
                                 proximity=FALSE, importance = TRUE)
-RF.model.ZC
+
+
+
+#List all variables for evaluation
+covariate.list<-list(c('ZcPresence','curl_mean',
+                       'sst_mean','depth','mldTemp','ttTemp','randSamp'))
+
+include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
+
+#make a text string of all covariates considered for a file name 
+string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
+
+# Use models that Anne ran with reduced covariates
+RF.model.ZC_reducedCovariates_10kTrees <- readRDS("data/RF.model.ZC_reducedCovariates_10kTrees.rds")
+RF.model.ZC_reducedCovariates_15kTrees <- readRDS("data/RF.model.ZC_reducedCovariates_15kTrees.rds")
+
 
[…]

### 2023-12-08 05_analysis.Rmd
Update 05_analysis.Rmd
added ships to rolling plots

@@
 Original data points plotted
-```{r}
-ZC_pivot_reg <- pivot_longer(ZC_filt, cols = c(TOL_125,TOL_2000, Wpresence), names_to = "datacol", values_to = "value")
+```{r fig.width = 12}
+ZC_filt <- GPSwhaleAISenvSS10 %>% 
+  filter(species %in% c("ZC", NA)) %>%
+  # Hourly rolling averages
+  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA)) %>%
+  mutate(rollAvg_ship = zoo::rollapply(shipin10km, 5, mean, fill = NA)) %>%
+  # Drop first and last NAs that are created as part of averages
+  drop_na(., rollAvg_Whl)
+
+ZC_pivot_reg <- pivot_longer(ZC_filt, cols = c(TOL_125,TOL_2000, Wpresence, shipin10km), names_to = "datacol", values_to = "value")
 
@@
                                    vjust = 0.5, hjust=1)) +
-  labs(title = "ZC Detections + SS Metrics - Buoy 21")
+  labs(title = "ZC Detections + SS Metrics + Ships - Buoy 10")
 
@@
                                    vjust = 0.5, hjust=1)) +
-  labs(title = "ZC Detections + SS Metrics + Ships - Buoy 13")
+  labs(title = "ZC Detections + SS Metrics + Ships - Buoy 10")
 

### 2024-01-24 loud_events.Rmd
Create loud_events.Rmd

---
title: "loud_events"
author: "Marina Bozinovic"
date: "2024-01-22"
output: html_document
---

## Label and compare before/during/after 2kHz levels above 90th percentile

Required Packages
```{r}
library(tidyverse)
```

Load data
```{r}
# Individual Drifts
ftlist <- dir("D:/Buoy_dfs/", recursive=TRUE, full.names=TRUE, pattern="GPSwhaleAISenvSS")

for (ft in ftlist) {
  df <- readRDS(ft)
  assign(paste0("GPSwhaleAISenvSS", substr(ft, 29, 30)), df)
}

# Drift 20 has interpolated whale calls
GPSwhaleAISSS20.alt <- readRDS("D:Buoy_dfs/GPSwhaleAISSS20_alt.rda")

# Whale detections
load("D:/Whale_detections/CCES2018_BW_and_PM_Detections.rda")

# Combined drifts
load(file = "data/allDrifts.rda")
```

Create e_ for each drift (modified from original)
```{r}
# Manipulate raw whale call dataframe
EventInfo_mut <- EventInfo %>%
  mutate(lengCalls_in_min = as.numeric((EndTime-StartTime)/60)) %>% # convert duration of calls to minutes
  rename(UTC = StartTime) %>%
  subset(species %in% c(NA, "ZC", "PM")) %>%
  select(UTC,lengCalls_in_min, Deployment)

## Change Station 15 to Station 14 here. Drift 15 and 14 are the same (mix-up during fieldwork)
EventInfo_mut$Deployment[EventInfo_mut$Deployment == 15] <- 14

# Manipulate individual drift data and remove environmental variables for simplicity
e7 <- GPSwhaleAISenvSS07 %>%
  mutate(Deployment = as.numeric("07")) %>% #add Deployment to the GPSwhaleAIS objects
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e8 <- GPSwhaleAISenvSS08 %>%
  mutate(Deployment = as.numeric("08")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e10 <- GPSwhaleAISenvSS10 %>%
  mutate(Deployment = as.numeric("10")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e12 <- GPSwhaleAISenvSS12 %>%
  mutate(Deployment = as.numeric("12")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e13 <- GPSwhaleAISenvSS13 %>%
  mutate(Deployment = as.numeric("13")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e14 <- GPSwhaleAISenvSS14 %>%
  mutate(Deployment = as.numeric("14")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e16 <- GPSwhaleAISenvSS16 %>%
  mutate(Deployment = as.numeric("16")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e18 <- GPSwhaleAISenvSS18 %>%
  mutate(Deployment = as.numeric("18")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e19 <- GPSwhaleAISenvSS19 %>%
  mutate(Deployment = as.numeric("19")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)

# No need to remove columns from this one - it has no env variables associated with it
e20 <- GPSwhaleAISSS20.alt %>%
  mutate(Deployment = as.numeric("20")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-shipin10km, -timeAfterShip)

e21 <- GPSwhaleAISenvSS21 %>%
  mutate(Deployment = as.numeric("21")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e22 <- GPSwhaleAISenvSS22 %>%
  mutate(Deployment = as.numeric("22")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)
e23 <- GPSwhaleAISenvSS23 %>%
  mutate(Deployment = as.numeric("23")) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment")) %>%
  select(-sst_mean, -curl_mean, -dist2slope, -depth, -bathy_slope,
         -dist2shore, -chlorophyll_mean, -ssh_mean, -mldDepth, -ttDepth,
         -mldTemp, -ttTemp, -temp400, -sal400, -shipin10km, -timeAfterShip)

e_all <- rbind(e7,e8,e10,e12,e13,e14,e16,e18,e19,e20,e21,e22,e23) %>%
  left_join(EventInfo_mut, by = c("UTC", "Deployment"))

#Save edited dfs
save(e7, file = "data/e7.rda")
save(e8, file = "data/e8.rda")
save(e10, file = "data/e10.rda")
save(e12, file = "data/e12.rda")
save(e13, file = "data/e13.rda")
save(e14, file = "data/e14.rda")
save(e16, file = "data/e16.rda")
save(e18, file = "data/e18.rda")
save(e19, file = "data/e19.rda")
save(e20, file = "data/e20.rda")
save(e21, file = "data/e21.rda")
save(e22, file = "data/e22.rda")
save(e23, file = "data/e23.rda")
```

Find 90th percentiles
```{r}
# Find 90th percentile of buoy 7
q7 <- quantile(GPSwhaleAISenvSS07$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 8
q8 <- quantile(GPSwhaleAISenvSS08$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 10
q10 <- quantile(GPSwhaleAISenvSS10$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 12
q12 <- quantile(GPSwhaleAISenvSS12$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 13
q13 <- quantile(GPSwhaleAISenvSS13$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 14
q14 <- quantile(GPSwhaleAISenvSS14$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 16
q16 <- quantile(GPSwhaleAISenvSS16$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 18
q18 <- quantile(GPSwhaleAISenvSS18$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 19
q19 <- quantile(GPSwhaleAISenvSS19$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 20
q20 <- quantile(GPSwhaleAISSS20.alt$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 21
q21 <- quantile(GPSwhaleAISenvSS21$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 22
q22 <- quantile(GPSwhaleAISenvSS22$TOL_2000, probs = 0.90)

# Find 90th percentile of buoy 23
q23 <- quantile(GPSwhaleAISenvSS23$TOL_2000, probs = 0.90)

# Find 90th percentile of ALL BUOYS
q_all <- quantile(e_all$TOL_2000, probs = 0.90)

save(q7, file = "data/q7.rda")
save(q8, file = "data/q8.rda")
save(q10, file = "data/q10.rda")
save(q12, file = "data/q12.rda")
save(q13, file = "data/q13.rda")
save(q14, file = "data/q14.rda")
save(q16, file = "data/q16.rda")
save(q18, file = "data/q18.rda")
save(q19, file = "data/q19.rda")
save(q20, file = "data/q20.rda")
save(q21, file = "data/q21.rda")
save(q22, file = "data/q22.rda")
save(q23, file = "data/q23.rda")
```

DELETE?? For all drifts
```{r}
# create new column to categorize times when TOL2kHz is over 90th percentile
#d_all <- e_all %>%
#  filter(species %in% c("PM", "ZC", NA)) %>% # filter out species other than sperm and Cuvier's
#  mutate(BefDurAftLOUD = ifelse(TOL_2000 > q_all, "during", "bef/aft")) %>% #assign categories when TOL2kHz is above 90th percentile
#  mutate(lengCalls_in_min = ifelse(is.na(lengCalls_in_min), 0, lengCalls_in_min)) # turn NAs into zeros in duration of calls column

#fb <- e_all %>%
  # filter out species other than sperm and Cuvier's
#  filter(species %in% c("PM", "ZC", NA)) %>%

  #exclude first 48 hrs of every drift
#  group_by(Deployment) %>%
#  mutate(whereFirst = if_else(row_number()==1, 1, 0)) %>%
#  mutate(tmpG = cumsum(c(FALSE, as.logical(diff(whereFirst))))) %>%
#  mutate(tmp_a = c(0, diff(UTC)) * !whereFirst) %>%
#  mutate(hrsSinceStart = cumsum(tmp_a)/3600) %>% # convert to hours
#  ungroup() %>%
#  select(-c(tmp_a, tmpG, whereFirst)) %>%

  ##assign categories when TOL2kHz is above 90th percentile
#  mutate(BefDurAftLOUD = ifelse(TOL_2000 > q_all, 1, 0))  # if TOL_2000 > q_all, assign "during"
```

Assign_loud <- before/during/after function
```{r}
# create a function that assigns bef/dur/aft using drift (e_) and 90th percentile value for that drift (q_)

assign_loud <- function(drift, loud_value){

drift %>%
  # identify hours elapsed since start (in order to exclude first 48 later)
  arrange(UTC) %>%
  mutate(whereFirst = if_else(row_number()==1, 1, 0),
         tmpG = cumsum(c(FALSE, as.logical(diff(whereFirst)))),
         tmp_a = c(0, diff(UTC)) * !whereFirst,
         hrsSinceStart = cumsum(tmp_a)/3600) %>% # convert seconds to hours
  select(-c(tmp_a, tmpG, whereFirst)) %>%

  ##assign categories when TOL2kHz is above 90th percentile
  mutate(loud = ifelse(TOL_2000 > loud_value, 1, 0),
         # Find hours since last and time to "loud" event.
         tmpF = cumsum(c(FALSE, as.logical(diff(loud)))),
         tmp_af = c(0, diff(UTC)) * !loud,
         tmp_bef = as.numeric(c(diff(UTC), 0) * !loud)) %>%
  group_by(tmpF) %>%
  mutate(hrsSinceLoud = cumsum(tmp_af)/3600,
         hrsToLoud = sort(cumsum(tmp_bef), decreasing = TRUE)/3600) %>% # convert seconds to hours
  ungroup() %>%
  select(-c(tmp_af, 
            tmp_bef,
            tmpF)) %>%

  # label rows as before/during/after or NA to omit 
  mutate(BefDurAftLOUD =

                #label "during" if loud = 1
                ifelse(.$loud == 1, "during",

                    #label "after" if within 12 hrs after loud event and no upcoming loud event for 12 hrs
                    ifelse(.$hrsSinceLoud <= 12 & .$hrsToLoud >= 12, "after",

                       #label in-between loud event times as "during" if its within 6 hrs of last loud event
                       # and the next loud event is within 6 hours
                       ifelse(.$hrsSinceLoud < 6 & .$hrsToLoud < 6, "during",

                            #label "before" if within 12 hrs of loud event and if 12 hrs have passed since loud event
                            ifelse(.$hrsToLoud < 12 & .$hrsSinceLoud > 12, "before", NA))))) %>%

      #exclude first 48 hrs of every drift
      subset(., .$hrsSinceStart > 48)
}
```

Manually calculate drift 21
```{r}
labeled21 <- e21 %>%
  # identify hours elapsed since start (in order to exclude first 48 later)
  arrange(UTC) %>%
  mutate(whereFirst = if_else(row_number()==1, 1, 0),
         tmpG = cumsum(c(FALSE, as.logical(diff(whereFirst)))),
         tmp_a = c(0, diff(UTC)) * !whereFirst,
         hrsSinceStart = cumsum(tmp_a)/60) %>% # convert MIN to hours
  select(-c(tmp_a, tmpG, whereFirst)) %>%

  ##assign categories when TOL2kHz is above 90th percentile
  mutate(loud = ifelse(TOL_2000 > q21, 1, 0),
         # Find hours since last and time to "loud" event.
         tmpF = cumsum(c(FALSE, as.logical(diff(loud)))),
         tmp_af = c(0, diff(UTC)) * !loud,
         tmp_bef = as.numeric(c(diff(UTC), 0) * !loud)) %>%
  group_by(tmpF) %>%
  mutate(hrsSinceLoud = cumsum(tmp_af)/60,
         hrsToLoud = sort(cumsum(tmp_bef), decreasing = TRUE)/60) %>% # convert MIN to hours
  ungroup() %>%
  select(-c(tmp_af, 
            tmp_bef,
            tmpF)) %>%

  # label rows as before/during/after or NA to omit 
  mutate(BefDurAftLOUD =

                #label "during" if loud = 1
                ifelse(.$loud == 1, "during",

                    #label "after" if within 12 hrs after loud event and no upcoming loud event for 12 hrs
                    ifelse(.$hrsSinceLoud < 12 & .$hrsToLoud >= 12, "after",

                       #label in-between loud event times as "during" if its within 6 hrs of last loud event
                       # and the next loud event is within 6 hours
                       ifelse(.$hrsSinceLoud < 6 & .$hrsToLoud < 6, "during",

                            #label "before" if within 12 hrs of loud event and if 12 hrs have passed since loud event
                            ifelse(.$hrsToLoud < 12 & .$hrsSinceLoud > 12, "before", NA))))) %>%

      #exclude first 48 hrs of every drift
      subset(., .$hrsSinceStart > 48)
```

Assign_loud function for each drift
```{r}
labeled7 <- assign_loud(e7, q7)
labeled8 <- assign_loud(e8, q8)
labeled10 <- assign_loud(e10, q10)
labeled12 <- assign_loud(e12, q12)
labeled13 <- assign_loud(e13, q13)
labeled14 <- assign_loud(e14, q14)
labeled16 <- assign_loud(e16, q16)
labeled18 <- assign_loud(e18, q18)
labeled19 <- assign_loud(e19, q19)
labeled20 <- assign_loud(e20, q20)
#labeled21 <- assign_loud(e21, q21) # has zero rows because tmp_a calculates time elapsed since start in min, not sec.
# need to do manually (SEE ABOVE)
labeled22 <- assign_loud(e22, q22)
labeled23 <- assign_loud(e23, q23)

#saveRDS(labeled7, file = "data/labeled7.rda")
```

```{r}
# Example of where loud sound events occurred in time.
ggplot(labeled7) + geom_line(aes(x = UTC, y = loud)) + 
  scale_x_datetime(date_breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(labeled7) + geom_line(aes(x = UTC, y = TOL_2000)) + 
  scale_x_datetime(date_breaks = "1 day") +
  geom_hline(yintercept = q8, color = "yellow", linewidth = 1) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

For reference:
Drift number and duty cycle (on/off in min)
7-2/18
8-2/18
10-2/18
12-2/18
13-2/18
14-2/18
16-2/8
18-2/4
19-2/8
20-2/8
21-2/4
22-2/8
23-2/3

Calculate hourly probability for each drift
```{r}
hourly_prob_7 <- e7 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/6) # change the divisor based on the duty cycle (active min/hr)
#saveRDS(hourly_prob_7, file = "data/hourly_prob_7.rda")

hourly_prob_8 <- e8 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/6) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_10 <- e10 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/6) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_12 <- e12 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/6) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_13 <- e13 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/6) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_14 <- e14 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/6) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_16<- e16 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/15) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_18 <- e18 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/30) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_19 <- e19 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/15) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_20 <- e20 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/15) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_21 <- e21 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/30) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_22 <- e22 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/15) # change the divisor based on the duty cycle (active min/hr)

hourly_prob_23 <- e23 %>%
  group_by(UTC = floor_date(UTC, "1 hour")) %>%
  summarize(totalCallmin = sum(na.omit(lengCalls_in_min))) %>%
  add_column(hrProb = .$totalCallmin/40) # change the divisor based on the duty cycle (active min/hr)
```

Attach hourly probability of detection to before/during/after df and make preliminary plot
```{r}
# Join labeled_ with hourly_prob_ dfs
plt7 <- labeled7 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_7, by = join_by("UTC_round"=="UTC"))

plt8 <- labeled8 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_8, by = join_by("UTC_round"=="UTC"))

plt10 <- labeled10 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_10, by = join_by("UTC_round"=="UTC"))

plt12 <- labeled12 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_12, by = join_by("UTC_round"=="UTC"))

plt13 <- labeled13 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_13, by = join_by("UTC_round"=="UTC"))

plt14 <- labeled14 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_14, by = join_by("UTC_round"=="UTC"))

plt16 <- labeled16 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_16, by = join_by("UTC_round"=="UTC"))

plt18 <- labeled18 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_18, by = join_by("UTC_round"=="UTC"))

plt19 <- labeled19 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_19, by = join_by("UTC_round"=="UTC"))

plt20 <- labeled20 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_20, by = join_by("UTC_round"=="UTC"))

plt21 <- labeled21 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_21, by = join_by("UTC_round"=="UTC"))

plt22 <- labeled22 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_22, by = join_by("UTC_round"=="UTC"))

plt23 <- labeled23 %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_23, by = join_by("UTC_round"=="UTC"))
```

Trim events w/in individual drifts that don't follow before/during/after pattern
```{r, fig.width=13}
plt7 <- plt7[1:1717,] #remove final before/during period
plt8 <- plt8[35:3877,] # remove final "before" period and first "during" period
plt10 <- plt10[1:4086,]
plt12 <- plt12[1:2310,]
plt13 <- plt13[92:2665,]
plt14 <- plt14[59:2246,]
plt16 <- plt16[31:809,]
plt18 <- plt18[204:1369,]
plt19 <- plt19[109:1460,]
plt20 <- plt20[95:772,]
plt21 <- plt21[20:127,] # very short!
plt22 <- plt22[4:1022,]
plt23 <- plt23[87:751,]

# Bind
plt_all <- bind_rows(plt7, plt8, plt10, plt12, plt13, plt14, plt16,
                 plt18, plt19, plt20, plt21, plt22, plt23)

plt23 %>%
  distinct(UTC, .keep_all = TRUE) %>% #remove any duplicate UTC rows
  drop_na(BefDurAftLOUD) %>%
  ggplot() + geom_bar(aes(x = UTC, color = BefDurAftLOUD, fill = BefDurAftLOUD)) + 
  scale_x_datetime(date_breaks = "1 day") +
  scale_color_manual(values = c("red", "blue", "green")) +
  scale_fill_manual(values = c("red", "blue", "green")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Time Series with Loud Sound Events - Buoy 23")
```

## Boxplots of Cuvier's and Sperm whales
```{r}
# When whales are detected, which category of loudness are they in?

#Plot ZC AND PM
plt_no_zeros <- plt_all %>% 
  filter(hrProb != 0) %>% # remove zeros
  filter(species %in% c("PM", "ZC", NA)) %>%
  na.omit(BefDurAftLOUD)

ggplot(plt_no_zeros) + geom_boxplot(aes(x = BefDurAftLOUD, y = hrProb, fill = species)) +
  scale_x_discrete(limits = c('before', 'during', 'after')) +
  labs(title = "Hourly Probability of ZC and PM Presence Before, During,
       and After Loud Sound Event - All Buoys",
       x = "Category", y = "Hourly Probability") +
  scale_fill_discrete(labels = c("Sperm", "Cuvier's")) +
  theme_bw()

# 60 PM calls
# 196 ZC calls

```

##Testing Statistical Significance
```{r}
# Is it distributed normally? Histogram and Shapiro-Willks test to confirm.
hist(plt_no_zeros$hrProb)
shapiro.test(plt_no_zeros$hrProb)
# no -- p-value is way less than 0.05, so given data is significantly different from the normal distribution.

# Use Kruskal-Wallis test to test if hourly probability varies by period
ZC_plt_no_zeros <- plt_no_zeros %>% filter(species %in% c("ZC", NA))
kruskal.test(hrProb ~ BefDurAftLOUD, data = ZC_plt_no_zeros) #ZC only

PM_plt_no_zeros <- plt_no_zeros %>% filter(species %in% c("PM", NA))
kruskal.test(hrProb ~ BefDurAftLOUD, data = PM_plt_no_zeros) #PM only

#p-value is more than 0.05 in both, so there is no significant differences between the groups

```

DELETE?? d_all_sum
```{r}
# Group by hour and summarize minutes of calls per hour. How to incorporate back?
#d_all_sum <- d_all %>%
#  mutate(hr = floor_date(UTC, unit = "hour")) %>%
#  group_by(hr) %>%
#  summarize(tot_min = sum(lengCalls_in_min),
#            tot_dets = sum(Wpresence),
#            avg_2000 = median(TOL_2000)) %>%
#  mutate(BefDurAftLOUD = ifelse(avg_2000 > q_all, "during", "bef/aft"))

# Plot to compare animal presence during loud sound levels vs the other times.

# Loud events vs whale presence/absence for ALL DRIFTS
#ggplot(d_all, aes(Loud_event, TOL_2000)) + 
#  geom_boxplot(aes(x = as.factor(BefDurAftLOUD), y = TOL_2000, fill = as.factor(Wpresence))) +
#  guides(fill=guide_legend(title="Whale presence"))

# total whale call minutes/hr
#ggplot(d_all_sum, aes(Sum_WhaleMinPerHr, TOL_2000_median)) + 
#  geom_point(aes(x = tot_min, y = avg_2000))+
#  geom_hline(yintercept = 87.24443, color = "red")#90th percentile line

#ggplot(d_all, aes(Length_of_whl_calls_min, TOL_2000)) + 
#  geom_point(aes(x = lengCalls_in_min, y = TOL_2000))+
#  geom_hline(yintercept = 87.24443, color = "red")#90th percentile line

```

DELETE?? e_all_time
```{r}
#e_all_time <- e_all %>%
#   mutate(BefDurAftLOUD = ifelse(TOL_2000 > q_all, 1, 0)) %>%
#   arrange(Deployment, UTC) %>%
#   mutate(tmpG = cumsum(c(FALSE, as.logical(diff(BefDurAftLOUD))))) %>%
#   mutate(tmp_a = c(0, diff(UTC)) * !BefDurAftLOUD) %>%
#  mutate(tmp_b = c(diff(UTC), 0) * !BefDurAftLOUD) %>%  
#   group_by(tmpG) %>%
#   mutate(timeFromLoud = cumsum(tmp_a)) %>%
#  mutate(timeToLoud = rev(cumsum(rev(as.numeric(tmp_b))))) %>%
#   ungroup() %>%
#   select(-c(tmp_a,tmp_b, tmpG))

```

#####################################################################################

Playing with boundary conditions (zz)
```{r}
## Playing around with the following

zz <- e23 %>%
     #filter out species other than Cuvier's and PM
     filter(species %in% c("ZC", "PM", NA)) %>%

     # identify hours elapsed since start (in order to exclude first 48 later)
     arrange(UTC) %>%
     mutate(whereFirst = if_else(row_number()==1, 1, 0),
            tmpG = cumsum(c(FALSE, as.logical(diff(whereFirst)))),
            tmp_a = c(0, diff(UTC)) * !whereFirst,
            hrsSinceStart = cumsum(tmp_a)/3600) %>% # convert seconds to hours
     select(-c(tmp_a, tmpG, whereFirst)) %>%

     ##assign categories when TOL2kHz is above 90th percentile
     mutate(loud = ifelse(TOL_2000 > q23, 1, 0),
            # Find hours since last and time to "loud" event.
            tmpF = cumsum(c(FALSE, as.logical(diff(loud)))),
            tmp_af = c(0, diff(UTC)) * !loud,
            tmp_bef = as.double(c(diff(UTC), 0) * !loud)) %>%
     group_by(tmpF) %>%
     mutate(hrsSinceLoud = cumsum(tmp_af)/3600,
            hrsToLoud = sort(cumsum(tmp_bef), decreasing = TRUE)/3600) %>% # convert seconds to hours
     ungroup() %>%
     select(-c(tmp_af, 
               tmp_bef,
               tmpF)) %>%

     # label rows as before/during/after or NA to omit 
     mutate(BefDurAftLOUD =

                #label "during" if loud = 1
                ifelse(.$loud == 1, "during",

                    #label "after" if within 12 hrs after loud event and no upcoming loud event for 12 hrs
                    ifelse(.$hrsSinceLoud < 12 & .$hrsToLoud >= 12, "after",

                       #label in-between loud event times as "during" if its within 6 hrs of last loud event
                       # and the next loud event is within 6 hours
                       ifelse(.$hrsSinceLoud < 6 & .$hrsToLoud < 6, "during",

                            #label "before" if within 12 hrs of loud event and if 12 hrs have passed since loud event
                            ifelse(.$hrsToLoud < 12 & .$hrsSinceLoud > 12, "before", NA))))) %>%

      #exclude first 48 hrs of every drift
      subset(., .$hrsSinceStart > 48)
```

Bef/Dur/Aft Boxplot
```{r}
# Example zz plot
plt_zz <-  zz %>%
  group_by(UTC_round = floor_date(UTC, "1 hour")) %>%
  left_join(hourly_prob_23, by = join_by("UTC_round"=="UTC"))

plt_no_zeros <- plt_zz %>% filter(hrProb != 0) %>% # remove zeros
  na.omit(BefDurAftLOUD)

ggplot(plt_no_zeros) + 
  geom_boxplot(aes(x = as.factor(BefDurAftLOUD), y = hrProb, fill = species)) +
    scale_x_discrete(limits = c('before', 'during', 'after')) + 
    labs(title = "Hourly Probability of Whale Presence Before, During,
       and After Loud Sound Event - Buoy 23")

#7 - 
#8 - 
#10 - similar to all buoys
#12 - much higher "after" probability
#13 - slightly lower "after" prob
#14 - no plot?
#16 - "after" is MUCH higher
#18 - "after has lower prob than "during". "before" is lowest
#19 - ideal! High "before", low "after"
#20 - high "during" probability
#21 - very few detections
#22 - highest prob is "before". very few non-zero prob points
#23 - highest prob is "during". very few non-zero prob points

```

Time Series. Example of where loud events and bef/after events occurred in time
```{r, fig.width=13}
# Just zz
plt_zz %>%
  distinct(UTC, .keep_all = TRUE) %>% #remove any duplicate UTC rows
  drop_na(BefDurAftLOUD) %>%
  ggplot() + geom_bar(aes(x = UTC, color = BefDurAftLOUD, fill = BefDurAftLOUD)) + 
  scale_x_datetime(date_breaks = "1 day") +
  scale_color_manual(values = c("red", "blue", "green")) +
  scale_fill_manual(values = c("red", "blue", "green")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Time Series with Loud Sound Events - Buoy 23")
```

### 2024-01-24 maps.Rmd
Update maps.Rmd
updated AIS density map

@@
 shipList <- dir("D:/AIS/AIS_West", recursive=TRUE, full.names=TRUE, pattern=".rda")
+# be sure to remove AIS_grid.rda from this list (shipList[120])
 
@@
 
-  print(paste0("Completed", substr(i,21,25)))
+  print(paste0("Completed ", substr(i,21,25)))
 }
+
+# shipList contains AIS_grid.rda which does not have MMSI! Will generate an error, but can ignore.
 ```
@@
 st_area(grid)
+st_crs(grid)
+st_distance(grid)
+st_is_longlat(grid)
 st_area(AISgrd)
+
+#0.1385641 (area)
+
 ```
@@
 #tm_basemap(leaflet::providers$Esri.WorldTopoMap) +
-  tm_shape(AISgrd) + tm_fill(fill = 'total', 
-                             col = 'lightgrey', 
-                             fill.legend = tm_legend("Ship Counts"),
-                             fill.scale = tm_scale_continuous()) + 
-  tm_shape(west) + tm_borders(col = "black", 
-                              fill = "papayawhip")+
-  tm_shape(detsOnly_sf) + tm_dots(size = 0.2, 
+  tm_shape(AISgrd) + tm_fill(col = 'total', style = 'cont', title = "Ship Counts", palette = "Blues") +
+  tm_borders(col = 'lightgrey') +
+  tm_shape(west) + tm_fill(col = "papayawhip") + tm_borders(col = "black") +
+  tm_shape(detsOnly_sf) + tm_dots(size = 0.1, 
                                   legend.show = TRUE) +
@@
           col = "grey") +
-  tm_scalebar() +
+  tm_scale_bar(position = "center") +
   tm_compass(type = "arrow", 
@@
              color.light = "blue") +
-  tm_add_legend('symbol',
-                fill = "black",
+  tm_add_legend(type = 'symbol',
+                col = "black",
                 labels = "Whale Detections",
@@
                 lwd = 2) +
-  tm_place_legends_right(width = 0.2) +
-  tm_title("Daily Counts of Unique AIS Vessels")
+  tm_layout(main.title = "Daily Counts of Unique AIS Vessels",
+            main.title.position = 'left') +
+    tmap_style("white")
 
+# Time frame is from 8-6-18 to 12-2-18, duration of drifts in the water.
+# Hexagonal cells are 0.4 degrees across (~43 km)
+```
+Measure cell distance across
+```{r}
+measure <- function(lon1,lat1,lon2,lat2) {
+    R <- 6378.137                                # radius of earth in Km
+    dLat <- (lat2-lat1)*pi/180
+    dLon <- (lon2-lon1)*pi/180
+    a <- sin((dLat/2))^2 + cos(lat1*pi/180)*cos(lat2*pi/180)*(sin(dLon/2))^2
+    c <- 2 * atan2(sqrt(a), sqrt(1-a))
+    d <- R * c
+    return (d)                            # distance in kilometers
+}
+
+#AISgrd[25]
+measure(-128.4902, 30.02285, -128.8902, 30.25379)
+
+#AISgrd[801]
+measure(-118.8902, 30.71567, -119.2902, 30.94661)
+
+#AISgrd[400]
+measure(-125.0902, 36.60464, -125.4902, 36.83558)
+
+library(geosphere)
+p1 <- c(-128.4902, 30.02285)
+p2 <- c(-128.8902, 30.25379)
+#distance in meters
+distHaversine(p1,p2)
 ```

### 2024-01-24 03_environ_var.Rmd
Update 03_environ_var.Rmd

@@
 ```{r}
+# Remove points where buoy is on the shelf - 300 meters
+#Only occurs with buoy 7
+env07 <- env07 %>%
+  filter(depth > 300)
+
 saveRDS(env10, 'data/GPSwhaleAISenv10.rda')

### 2024-01-24 05_analysis.Rmd
Update 05_analysis.Rmd
removed loud event part, moved to different rmd

@@
 }
+
+GPSwhaleAISSS20.alt <- readRDS("D:Buoy_dfs/GPSwhaleAISSS20_alt.rda")
 ```
@@
 for (col in dlist){
-  data <- get(col) %>% mutate(Deployment = substr(col,17,18))
-  data <- data[data$species %in% c(NA, "ZC", "PM", "BB", "BW43"),]    # Remove BW37V and BWC from analysis as there are too few detections.
+  data <- get(col) %>% mutate(Deployment = as.numeric(substr(col,17,18)))
+  data <- data[data$species %in% c(NA, "ZC", "PM"),]    # Remove all but ZC and PM from analysis as there are too few detections.
   assign(paste0("GPSwhaleAISenvSS", substr(col,17,18)), data)
 }
+
+# Remove points where buoy is on the shelf - 300 meters
+#Only occurs with buoy 7
+GPSwhaleAISenvSS07 <- GPSwhaleAISenvSS07 %>%
+  filter(depth > 300)
+saveRDS(GPSwhaleAISenvSS07, file = "data/GPSwhaleAISenvSS07.rda")
+saveRDS(GPSwhaleAISenvSS07, 'D:/Buoy_dfs/GPSwhaleAISenv07.rda')
 
@@
                    GPSwhaleAISenvSS13, GPSwhaleAISenvSS14, GPSwhaleAISenvSS16, GPSwhaleAISenvSS18,
-                   GPSwhaleAISenvSS19, GPSwhaleAISenvSS20, GPSwhaleAISenvSS21, GPSwhaleAISenvSS22,
+                   GPSwhaleAISenvSS19, GPSwhaleAISSS20.alt, GPSwhaleAISenvSS21, GPSwhaleAISenvSS22,
                    GPSwhaleAISenvSS23)
-
-# Remove points where buoy is on the shelf - 300 meters
-allDrifts <- allDrifts %>%
-  filter(depth > 300)
 
@@
 
-Compare overall TOL2000sound levels when whales are absent vs. present
+Compare overall TOL2000sound levels when ALL whales are absent vs. present
 ```{r}
 ggplot(allDrifts, aes(factor(Wpresence), TOL_2000)) + geom_boxplot()
-```
-
-Scatterplots
-
-```{r}
-# SST vs TOL 2000
-ggplot(data = allDrifts, aes(x = sst_mean, y = `TOL_2000`)) +
-  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
-
-# SST over time (strange plot)
-ggplot(allDrifts, aes(x = UTC, y = sst_mean)) +
-  #geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-  #geom_smooth(method = "loess")
-
-# Depth vs TOL 125
-ggplot(data = allDrifts, aes(x = `TOL_125`, y = depth *(-1))) +
-  #geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-
-# Chlorophyll vs TOL 125
-ggplot(data = allDrifts, aes(x = chlorophyll_mean, y = `TOL_125`)) +
-  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
-
-# CHl over time
-ggplot(allDrifts, aes(x = UTC, y = chlorophyll_mean)) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-  #geom_smooth(method = "loess")
-
-# Mixed Layer Depth vs TOL125 with species
-ggplot(data = allDrifts, aes(x = mldDepth, y = `TOL_125`)) +
-  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
-
-# Mixed Layer Depth over time
-ggplot(allDrifts, aes(x = UTC, y = mldDepth*(-1))) +
-  geom_line() +
-  geom_point(aes(color = species)) +
-  scale_color_discrete(na.translate=FALSE)
-
-ggplot(data = allDrifts, aes(x = `TOL_125`, y = bathy_slope)) +
-  geom_point(aes(color = species)) + scale_color_discrete(na.translate=FALSE)
-
 ```

### 2024-01-24 05_randomForest.R
Update 05_randomForest.R
added distance to escarpment to RF model

@@ -9,12 +9,10 @@ library(rfPermute)
 
 # Load in main dataframe
 load(file = "data/allDrifts.rda")
+load(file = "D:/Buoy_dfs/drift_escarpment.rda")
 
 # Create new row with random samples, make sure to add to covariates list
-allDrifts$randSamp <- sample(10000, size = nrow(allDrifts), replace = TRUE)
-
-# Load in main dataframe with seafloor gradient(?)
-#load(file = "data/allDrifts_grad.rda")
+drift_escarp$randSamp <- sample(10000, size = nrow(drift_escarp), replace = TRUE)
 
 
 #Define response variable (presence/absence as 0/1)
@@ -85,6 +83,11 @@ save.image(paste(Dep,"_", string.covars.used, ".RData", sep=""))
 
 # Reload without ZC filter
 load(file = "data/allDrifts.rda")
+load(file = "D:Buoy_dfs/drift_escarpment.rda")
+
+# Create new row with random samples, make sure to add to covariates list
+allDrifts$randSamp <- sample(10000, size = nrow(allDrifts), replace = TRUE)
+
 
 #Descriptive label for analysis
 Dep<-'CCES_Pm'
@@ -101,6 +104,7 @@ covariate.list<-list(c('PmPresence','curl_mean', 'bathy_slope',
                        'chlorophyll_mean','mldDepth','ssh_mean',
                        'mldDepth', 'mldTemp', 'ttDepth','ttTemp', 'temp400', 'sal400',
                        'randSamp'))
+
 
 include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
 
@@ -154,13 +158,15 @@ RF.model.ZC <- rfPermute(ZcPresence ~ ., data=DF.modelZC[,include.covars],
 
 
 #List all variables for evaluation
-covariate.list<-list(c('ZcPresence','curl_mean',
-                       'sst_mean','depth','mldTemp','ttTemp','randSamp'))
+covariate.list<-list(c('ZcPresence','curl_mean', 'chlorophyll_mean', 'dist2slope',
+                       'dist2escarp', 'bathy_slope', 
+                       'sst_mean','depth','mldDepth','ttTemp','randSamp'))
 
 include.covars <- which(names(allDrifts) %in% covariate.list[[1]])
 
 #make a text string of all covariates considered for a file name 
 string.covars.used <- paste0(names(allDrifts)[include.covars], sep="+", collapse="")
+
 
[…]

### 2024-01-24 06_gam.Rmd
Update 06_gam.Rmd
added distance to escarpment to GAMs

@@
 
-Load allDrifts object, split out Cuvier's and Sperm whales
+Load drift_escarp object, split out Cuvier's and Sperm whales
 ```{r}
-load(file = "data/allDrifts.rda")
+load(file = "D:/Buoy_dfs/drift_escarpment.rda")
 
-ZC_filt <- allDrifts %>% 
+drift_escarp <- drift_escarp %>%
+  st_drop_geometry()
+
+ZC_filt <- drift_escarp %>% 
   filter(species %in% c("ZC", NA))
 
-PM_filt <- allDrifts %>% 
+PM_filt <- drift_escarp %>% 
   filter(species %in% c("PM", NA))
@@
 # Environ only - binomial
-ZC_GAM1 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
-              data = ZC_filt, family = binomial, method = "REML")
+ZC_GAM1 <- gam(Wpresence>0 ~ s(depth) + s(ttTemp) + s(ssh_mean),
+              data = ZC_filt, family = binomial, 
+              method = "REML")
 summary(ZC_GAM1)
@@
 # Environ + noise - binomial
-ZC_GAM2<- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
-                      data = ZC_filt, family = binomial, method = "REML")
+ZC_GAM2<- gam(Wpresence>0 ~ s(depth) + s(ttTemp) + s(ssh_mean) + 
+                s(TOL_125) + s(TOL_2000),
+                      data = ZC_filt, family = binomial, 
+              method = "REML")
 summary(ZC_GAM2)
@@
 
-# Environ only - poisson
-ZC_GAM3 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
-              data = ZC_filt, family = poisson, method = "REML")
-summary(ZC_GAM3)
-gam.check(ZC_GAM3)
+plot(ZC_GAM1, all.terms = TRUE, scale = 0, pages = 1)
+plot(ZC_GAM2, all.terms = TRUE, scale = 0, pages = 3)
 
-# Environ + noise - poisson
-ZC_GAM4<- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
-                      data = ZC_filt, family = poisson, method = "REML")
-summary(ZC_GAM4)
-gam.check(ZC_GAM4)
-plot(ZC_GAM4, all.terms = TRUE, scale = 0)
-
-# Environ only - gaussian
-ZC_GAM5 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
-                  data = ZC_filt, method = "REML")
-summary(ZC_GAM5)
-gam.check(ZC_GAM5)
-
-# Environ + noise - gaussian
-ZC_GAM6 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
-                  data = ZC_filt, method = "REML")
-summary(ZC_GAM6)
-gam.check(ZC_GAM6)
-
-# Environ only - Tweedie
-ZC_GAM7 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean),
-                  data = ZC_filt, family = tw(link = "log"), method = "REML")
-summary(ZC_GAM7)
-gam.check(ZC_GAM7)
-
-# Environ + noise - Tweedie
-ZC_GAM8 <- gam(Wpresence>0 ~ s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
-                  data = ZC_filt, family = tw(link = "log"), method = "REML")
-summary(ZC_GAM8)
-gam.check(ZC_GAM8)
-
-plot(ZC_GAM1, all.terms = TRUE, scale = 0, pages = 1)
-plot(ZC_GAM2, all.terms = TRUE, scale = 0, pages = 2)
-plot(ZC_GAM3, all.terms = TRUE, scale = 0, pages = 1)
-plot(ZC_GAM4, all.terms = TRUE, scale = 0, pages = 2)
-plot(ZC_GAM5, all.terms = TRUE, scale = 0, pages = 1)
-plot(ZC_GAM6, all.terms = TRUE, scale = 0, pages = 2)
-plot(ZC_GAM7, all.terms = TRUE, scale = 0, pages = 1)
-plot(ZC_GAM8, all.terms = TRUE, scale = 0, pages = 2)
-AIC(ZC_GAM4, ZC_GAM8)
+# Compare AIC scores
+AIC(ZC_GAM1, ZC_GAM2)
 ```
@@
 PM_GAM1 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + 
-                    s(sst_mean),
+                    s(dist2escarp),
               data = PM_filt, family = binomial, method = "REML")
@@
 PM_GAM2 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + 
-                 s(sst_mean) + s(TOL_125) + s(TOL_2000),
+                 s(dist2escarp) + s(TOL_125) + s(TOL_2000),
               data = PM_filt, family = binomial, method = "REML")
@@
 
-# Environ only - poisson
-PM_GAM3 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
-              data = PM_filt, family = poisson, method = "REML")
-summary(PM_GAM3)
-gam.check(PM_GAM3)
+# Plot
+plot(PM_GAM1, all.terms = TRUE, scale = 0, pages = 1)
+plot(PM_GAM2, all.terms = TRUE, scale = 0, pages = 1)
 
-# Environ + noise - poisson
-PM_GAM4 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) 
-               + s(sst_mean) + s(TOL_125) + s(TOL_2000),
-              data = PM_filt, family = poisson, method = "REML")
-summary(PM_GAM4)
-gam.check(PM_GAM4)
-
-# Environ only - gaussian
-PM_GAM5 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
-              data = PM_filt, method = "REML")
-summary(PM_GAM5)
-gam.check(PM_GAM5)
-
-# Environ + noise - gaussian
-PM_GAM6 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean)
-               + s(TOL_125) + s(TOL_2000),
-              data = PM_filt, method = "REML")
-summary(PM_GAM6)
-gam.check(PM_GAM6)
-
-# Environ only - Tweedie
-PM_GAM7 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean),
-              data = PM_filt, family = tw(link = "log"), method = "REML")
-summary(PM_GAM7)
-gam.check(PM_GAM7)
-
-# Environ + noise - Tweedie
-PM_GAM8 <- gam(Wpresence>0 ~ s(dist2slope) + s(mldDepth) + s(sst_mean)
-               + s(TOL_125) + s(TOL_2000),
-              data = PM_filt, family = tw(link = "log"), method = "REML")
-summary(PM_GAM8)
-gam.check(PM_GAM8)
-
-plot(PM_GAM1, all.terms = TRUE, scale = 0)
-plot(PM_GAM2, all.terms = TRUE, scale = 0)
-plot(PM_GAM3, all.terms = TRUE, scale = 0)
-plot(PM_GAM4, all.terms = TRUE, scale = 0)
-plot(PM_GAM5, all.terms = TRUE, scale = 0)
-plot(PM_GAM6, all.terms = TRUE, scale = 0)
-plot(PM_GAM7, all.terms = TRUE, scale = 0)
-plot(PM_GAM8, all.terms = TRUE, scale = 0)
-AIC(PM_GAM4, PM_GAM8)
+# compare AIC scores
+AIC(PM_GAM1, PM_GAM2)
 
@@
 #####
-
-# Poisson
-ZC_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
-              data = ZC_filt2, family = poisson, method = "REML")
-summary(ZC_GAM_escarp)
-plot(ZC_GAM_escarp, all.terms = TRUE)
-
-# Tweedie (family = tw(link = "log"))
-ZC_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(depth) + s(temp400) + s(sst_mean) + s(TOL_125) + s(TOL_2000),
-              data = ZC_filt2, family = tw(link = "log"), method = "REML")
-summary(ZC_GAM_escarp)
-plot(ZC_GAM_escarp, all.terms = TRUE)
-
-# Poisson
-PM_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(dist2slope) + s(mldDepth) + s(sst_mean)
-               + s(TOL_125) + s(TOL_2000),
-              data = PM_filt2, family = poisson, method = "REML")
-summary(PM_GAM_escarp)
-plot(PM_GAM_escarp, all.terms = TRUE)
-plot.gam(PM_GAM_escarp)
-
-# Tweedie
-PM_GAM_escarp <- gam(Wpresence>0 ~ s(dist2escarp) + s(dist2slope) + s(mldDepth) + s(sst_mean)
-               + s(TOL_125) + s(TOL_2000),
-              data = PM_filt2, family = tw(link = "log"), method = "REML")
-summary(PM_GAM_escarp)
 ```

### 2024-01-24 09_psd.R
Update 09_psd.R

@@ -1,4 +1,4 @@
-# SCript from Anne Simonis 10-20-23
+# Script from Anne Simonis 10-20-23
 library(tidyverse)
 library(PAMscapes)
 
@@ -106,7 +106,7 @@ save(psd23_1hr, file = "D:/Soundscape_metrics/PSD/psd23_1hr.rda")
 #Create separatdesk chaire PSD objects for the times when bw were present and absent 
 ZcTimes13 <- filter(GPSwhale13, species == "ZC") #times when Cuvier's beaked whales were present on drift 13
 ZcTimes14 <- filter(GPSwhale14, species == "ZC") #times when Cuvier's beaked whales were present on drift 14
-#ZcTimes07 <- filter(GPSwhale07, species == "ZC") #times when Cuvier's beaked whales were present on drift 07
+ZcTimes07 <- filter(GPSwhale07, species == "ZC") #times when Cuvier's beaked whales were present on drift 07
 ZcTimes08 <- filter(GPSwhale08, species == "ZC") #times when Cuvier's beaked whales were present on drift 08
 ZcTimes10 <- filter(GPSwhale10, species == "ZC") #times when Cuvier's beaked whales were present on drift 10
 ZcTimes12 <- filter(GPSwhale12, species == "ZC") #times when Cuvier's beaked whales were present on drift 12
@@ -123,6 +123,7 @@ ZcTimes23 <- filter(GPSwhale23, species == "ZC") #times when Cuvier's beaked wha
 #Extract rounded hour from Zc detections
 ZcTimes13_1hr <- mutate(ZcTimes13, UTC = floor_date(UTC, '1 hour'))
 ZcTimes14_1hr <- mutate(ZcTimes14, UTC = floor_date(UTC, '1 hour'))
+ZcTimes07_1hr <- mutate(ZcTimes07, UTC = floor_date(UTC, '1 hour'))
 ZcTimes08_1hr <- mutate(ZcTimes08, UTC = floor_date(UTC, '1 hour'))
 ZcTimes10_1hr <- mutate(ZcTimes10, UTC = floor_date(UTC, '1 hour'))
 ZcTimes12_1hr <- mutate(ZcTimes12, UTC = floor_date(UTC, '1 hour'))
[…]

### 2024-01-24 corrPlot.R
Create corrPlot.R

@@ -1,0 +1,16 @@
+# Correlation Matrix
+
+#install.packages("corrplot")
+library(corrplot)
+library(tidyverse)
+
+load(file = "D:Buoy_dfs/drift_escarpment.rda")
+
+bzz <- drift_escarp %>%
+  select(6:21, 25) %>%
+  na.omit()
+
+res <- cor(bzz)
+
+corrplot(res, type = "upper", order = "hclust", 
+          tl.col = "black", tl.srt = 45)

### 2024-01-24 manuscript.rmd
Update manuscript.rmd

--- 2023-10-20 manuscript.rmd
@@
 bibliography: references.bib
+header-includes:
+- \setlength{\parindent}{4em}
+- \usepackage{setspace}\spacing{2}
 output:
@@
 csl: "marine-mammal-science.csl"
-header-includes:
-- \setlength{\parindent}{4em}
-- \usepackage{setspace}\doublespacing
 ---
-
-# Manuscript Draft
 
@@
 
-## Results
+~~"The Tweedie is a family of exponential type distributions, which are tolerant to large numbers of zero observations (Candy 2004)." -- Lacey and Hammond 2023~~
+
+(Candy, S.G. 2004. Modeling catch and effort data using Generalised Linear Models, the Tweedie Distriution, Random Vessel Effects and Random Stratum-by-Year effects. CCAMLR Sci. 11: 59--80.)
+
+## Results (moved to Google Drive)
 
@@
 
-## Discussion
+## Discussion (moved to Google Drive)
 
@@
 
-## Acknowledgements
+## Acknowledgements (moved to Google Drive)
 
@@
 
-2)  Environmental covariates (group by physiographic, oceanographic, etc)
+2)  List of environmental predictor covariates with resolution, data source, and justification. \
 
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
 | Covariate                           | Resolution       | Source                                                                                                                                                | Justification                           |
-|------------------|------------------|--------------------|------------------|
++=====================================+==================+=======================================================================================================================================================+=========================================+
 |                                     |                  |                                                                                                                                                       |                                         |
-| Distance to shore (m)              | 0.01-Degree Grid | ERDAPP                                                                                                                                            |                                         |
-| Distance to continental slope (m)  | 15 arc sec       | Derived from General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015                                                                      | @barlow2009                             |
-| Bathymetric slope (°)              | 15 arc sec       | Derived from GEBCO tif                                                                                                                            | @becker2010 @virgili2022                |
-| Depth (m)                          | 15 arc sec       | ERDAPP                                                                                                                                            | @forney2012 @becker2010 @virgili2022    |
-| Wind curl stress                   | 1°, 6-hourly     | ERDDAP                                                                                                                                            | @mannocci2015                           |
-| Mixed layer depth (m)              | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Distance to escarpment (m)          | 30 arc sec       | (D) Calculated from : Global Seafloor Geomorphic Features Map\                                                                                        | Lacey and Hammond, 2023                 |
+|                                     |                  | P.T.
+Harris, M.
+Macmillan-Lawler, J.
+Rupp, E.K.
+Baker, Geomorphology of the oceans, #Marine Geology, Volume 352, 2014, Pages 4-24, ISSN 0025-3227,    |                                         |
+|                                     |                  |                                                                                                                                                       |                                         |
+|                                     |                  | <https://doi.org/10.1016/j.margeo.2014.01.011>.
+|                                         |
+|                                     |                  |                                                                                                                                                       |                                         |
+|                                     |                  | (<https://www.sciencedirect.com/science/article/pii/S0025322714000310>)                                                                               |                                         |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Distance to shore (m)               | 0.01-Degree Grid | (A) ERDAPP                                                                                                                                            |                                         |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Distance to continental slope (m)   | 15 arc sec       | (C) Derived from General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015                                                                      | @barlow2009                             |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Bathymetric slope (°)               | 15 arc sec       | (C) Derived from General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015                                                                      | @becker2010 @virgili2022                |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Depth (m)                           | 15 arc sec       | (A) ERDAPP                                                                                                                                            | @forney2012 @becker2010 @virgili2022    |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Wind curl stress                    | 1°, 6-hourly     | (A) ERDDAP                                                                                                                                            | @mannocci2015                           |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Mixed layer depth (m)               | 0.08°, 3-hourly  | (B) HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
 | @mccullough2021, @forney2012 @moore2021 |
-| Thermocline temperature (°C)       | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Thermocline temperature (°C)        | 0.08°, 3-hourly  | (B) HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
 | @mccullough2021                         |
-| Thermocline depth (m)              | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Thermocline depth (m)               | 0.08°, 3-hourly  | (B) HYCOM + NCODA, GLBy0.08/expt 93.0, calculated with variable representative isotherm method (Fiedler 2010), as described in McCullough et al 2021.
 |                                         |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
 | Mixed layer depth Temperature (°C)  | 0.08°, 3-hourly  | Calculated from HYCOM                                                                                                                                 | @mccullough2021                         |
-| Salinity at 400m                   | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0                                                                                                                 |                                         |
-| Temperature at 400m (°C)           | 0.08°, 3-hourly  | HYCOM + NCODA, GLBy0.08/expt 93.0                                                                                                                 |                                         |
-| SST mean                           | 0.01° , daily    | ERDAPP                                                                                                                                            | Becker 2007, @forney2012 @virgili2022   |
-| CHL                                | 4 km, monthly    | ERDAPP                                                                                                                                            | @forney2012 @virgili2019                |
-| Sea Surface Height Anomalies (m)   | daily            | ERDAPP                                                                                                                                            | @virgili2019                            |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Salinity at 400m                    | 0.08°, 3-hourly  | (B) HYCOM + NCODA, GLBy0.08/expt 93.0                                                                                                                 |                                         |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Temperature at 400m (°C)            | 0.08°, 3-hourly  | (B) HYCOM + NCODA, GLBy0.08/expt 93.0                                                                                                                 |                                         |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Sea Surface Temperature (mean) (°C) | 0.01° , daily    | (A) ERDAPP                                                                                                                                            | Becker 2007, @forney2012 @virgili2022   |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Chlorophyll-a (mean) (mg m-3)       | 4 km, monthly    | (A) ERDAPP                                                                                                                                            | @forney2012 @virgili2019                |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+| Sea Surface Height Anomalies (m)    | daily            | (A) ERDAPP                                                                                                                                            | @virgili2019                            |
++-------------------------------------+------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------+
+
+\*\* Distances were calculated in R.
+Source A: ERDAPP (add link); Source B: HYCOM + NCODA, GLBy0.08/expt 93.0 (add link); Source C: General Bathymetric Chart of the Oceans (GEBCO, add link); Source D: Global Seafloor Geomorphic Features Map (add link).
 
